---
title: "Chapter 1.2: The Causal Roadmap"
format: html
---

In this chapter, we introduce the **Causal Roadmap**, a structured approach for asking and answering causal questions with observational data. The roadmap helps researchers define their scientific question, translate it into a formal causal model, assess whether the causal effect is identifiable from data, and choose appropriate estimators to answer that question.

## Why Use a Causal Roadmap?

Traditional analysis often starts with: "Let's run a regression and see what we find." But modern causal inference begins by asking: **What is the causal question we are trying to answer?** 

A causal roadmap ensures:
- Transparency in how evidence is generated
- Separation of scientific questions from statistical methods
- Clear articulation of assumptions and their implications
- Structured thinking in both trial and non-randomized settings

We’ll walk step-by-step through each part of the roadmap, with illustrative examples.

---

## Step 1: Define the Causal Question

This includes:
- **Population**: Who are we studying?
- **Intervention**: What exposure or treatment are we manipulating?
- **Outcome**: What effect are we interested in?
- **Timeframe**: Over what duration?

Example:

> What is the 3-year risk of cardiovascular events if all postmenopausal women with osteoporosis initiated **denosumab** compared to if all initiated **zoledronic acid**?

This PICO-style framing is the foundation of causal inference.

---

## Step 2: Specify the Causal Model

We now draw out how we believe the data were generated. This includes:

- **Variables**: Treatment (A), outcome (Y), and covariates (W)
- **DAGs**: Directed acyclic graphs to encode assumptions
- **Potential Outcomes**: $Y(1)$ and $Y(0)$ for each individual

We are making an implicit **Structural Causal Model (SCM)**:

$$
W = f_W(U_W),\
A = f_A(W, U_A),\
Y = f_Y(W, A, U_Y)
$$

Where $U$ terms represent unmeasured variables. 

A simple DAG:

```
W → A → Y
 \    ↘
  →────→
```

This DAG implies:
- W (confounders) affect both A and Y
- A (treatment) affects Y
- No unmeasured confounding (U nodes omitted for simplicity)

---

## Step 3: Define the Target Causal Estimand

We often want the **Average Treatment Effect (ATE)**:

$$
ATE = E[Y(1) - Y(0)]
$$

Or equivalently, **Risk Difference**:

$$
E[Y | do(A=1)] - E[Y | do(A=0)]
$$

We might also want:
- Risk ratios
- Hazard ratios (with caveats)
- Median survival differences
- Effects in subgroups (CATEs)

---

## Step 4: Link the Causal Estimand to the Observed Data

This is the **identification step**. It answers the question: *Can we estimate the causal effect from the data we have?*

Requires three assumptions:
- **Exchangeability (No unmeasured confounding)**: $Y(a) \perp A | W$
- **Positivity**: Everyone has a nonzero probability of receiving each treatment, given W
- **Consistency**: The observed outcome equals the potential outcome under the received treatment

If satisfied, we can identify:

$$
E[Y(1) - Y(0)] = E_W[ E[Y | A = 1, W] - E[Y | A = 0, W] ]
$$

---

## Step 5: Choose a Statistical Estimator

This step translates the identified quantity into an algorithm we can apply to data.

Options include:
- **G-computation**: Predict outcomes under each treatment, average over W
- **IPTW**: Weight observations by inverse probability of treatment
- **TMLE**: Targeted learning that combines outcome and treatment models

We’ll explore these in later chapters. The choice depends on:
- Assumptions you’re willing to make
- Sample size
- Tolerance for model misspecification

---

## Step 6: Evaluate and Interpret

Once you've estimated your parameter, interpretation is **not automatic**.

- Are your assumptions plausible?
- What does your estimate mean for real-world decisions?
- Can you generalize to other populations?

Tools for evaluation:
- Sensitivity analyses
- Negative control outcomes
- Covariate balance checks
- Expert consultation

---

## A Simple Simulated Example

We'll simulate a small dataset to show how steps 1-5 look in practice.

```{r}
set.seed(42)
n <- 1000
age <- rnorm(n, 75, 6)
cvd_history <- rbinom(n, 1, plogis(0.1 * (age - 70)))
W <- data.frame(age, cvd_history)
A <- rbinom(n, 1, plogis(-1 + 0.05 * (age - 70) + 1.5 * cvd_history))
Y <- rbinom(n, 1, plogis(-2 + 0.4 * A + 0.1 * (age - 70) + 1 * cvd_history))
data <- data.frame(W, A, Y)
```

Estimate ATE using g-computation:

```{r}
model <- glm(Y ~ A + age + cvd_history, family = binomial, data = data)
newdata1 <- transform(data, A = 1)
newdata0 <- transform(data, A = 0)
p1 <- predict(model, newdata = newdata1, type = "response")
p0 <- predict(model, newdata = newdata0, type = "response")
mean(p1 - p0)
```

---

## Summary

The Causal Roadmap gives a rigorous framework for designing and evaluating real-world effect estimates. Before choosing a statistical method, start with the science:
- Define your causal question
- Specify your model and assumptions
- Decide how to express the effect
- Choose a transparent, defensible method

In the next chapter, we’ll explore different estimation strategies in detail.
