---
format: html
---

# Running Case Study: Antiretroviral Therapy, HIV Progression, and Longitudinal Confounding {#sec-running-case-study}

*Estimated reading time: 20 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

Throughout this handbook, we use a single recurring causal question to illustrate causal concepts, estimation methods, and diagnostic tools: **does initiating highly active antiretroviral therapy (HAART) reduce mortality among people living with HIV, and if so, by how much?** We answer this question using a simulated longitudinal HIV antiretroviral therapy cohort built to reflect the marginal structural model literature, distributed in the `ipw` R package [@vanderwal2011ipw]. Anchoring every chapter to the same question and dataset lets you see how each method approaches the same target, and makes it easier to compare results across approaches.


## Clinical and epidemiologic motivation

In the early years of combination antiretroviral therapy (HAART), a central clinical question was: does initiating HAART reduce mortality among people living with HIV, and if so, by how much? Although this sounds like a question about HAART "use," it is really a question about *strategies for when to start* HAART, because therapy is initiated (and then sustained) in response to evolving clinical status. The longitudinal, time-dependent nature of the initiation decision is what creates the confounding structure we have to address.

Answering this from observational data is harder than it looks. Physicians do not assign HAART randomly. They start therapy when CD4 cell counts decline and clinical status worsens, so the very biomarkers that predict death also determine who gets treated and when. This is confounding by indication. The sickest patients get the strongest drugs, and a naive comparison can therefore make the best medicines look like poison.

The problem runs deeper than baseline confounding. HAART, once initiated, improves CD4 counts. A patient's CD4 level at any given visit therefore reflects both their underlying disease trajectory and the effects of prior treatment. CD4 count is at once a confounder of future treatment decisions and a mediator of past treatment effects. This treatment-confounder feedback is the defining feature of the dataset and the reason it motivates the longitudinal g-methods developed in later chapters.

The dataset we use is a simulated cohort of 1,200 people living with HIV designed to reflect the structure of real HIV cohort studies. Because the data-generating process is known, we can compare estimation methods and understand exactly where simpler approaches fail.


## Why this example is useful

Four features make this cohort a good teaching vehicle. First, the confounding by indication is severe. HAART initiators look *worse* than non-initiators in a crude analysis, simply because they had lower CD4 and worse prognosis when they started, so the cohort shows plainly how an effective treatment can appear harmful. Second, the treatment-confounder feedback through CD4 means no single regression can both adjust for CD4 and preserve its role as a mediator, and this is what forces the longitudinal g-methods. Third, the analysis demands a clearly specified target trial and estimand before any model is fit, because vague causal questions produce numerically precise answers that cannot be interpreted against any well-defined contrast. Fourth, the functional forms are unknown. How age, sex, and the evolving CD4 jointly drive both the treatment decision and mortality is plausibly non-linear and interactive: risk might rise sharply only once CD4 falls below a threshold, or a low CD4 might weigh differently on the initiation decision at different ages. Committing to one parametric form risks the misspecification bias shown in @sec-gcomputation, which is why the nuisance models are estimated with Super Learner (@sec-superlearner) rather than a single guessed regression.


## The observed data

The dataset is distributed as `haartdat` in the `ipw` R package [@vanderwal2011ipw]. It is a *simulated* cohort (not a prospective study) built to reproduce the structure of real HIV cohorts used in the marginal structural model literature. It contains 1,200 people living with HIV observed from HIV seroconversion in person-period (long) format, with one row per patient per fixed 100-day interval. The regular interval grid is a feature of the simulation; in real-world data, visit times would be irregular and would themselves need to be modeled.

```{r}
#| code-fold: true
library(ipw)
library(tidyverse)
data(haartdat)
```

### Variable dictionary

| Variable | Description | Role |
|----------|-------------|------|
| `patient` | Patient identifier (1 to 1,200) | Unit of analysis |
| `tstart` | Start of time interval (days since seroconversion) | Time structure |
| `fuptime` | End of time interval (days since seroconversion) | Time structure |
| `haartind` | HAART indicator: 1 if HAART has been initiated by the end of this interval, 0 otherwise | Time-varying treatment (A) |
| `event` | Death indicator: 1 if patient died during this interval | Outcome (Y) |
| `sex` | Sex: 0 = male, 1 = female | Baseline confounder (W) |
| `age` | Age in years at follow-up start | Baseline confounder (W) |
| `cd4.sqrt` | Square root of CD4 cell count at the end of the interval | Time-varying confounder (L) |
| `endtime` | Patient's final observed time (days) | Administrative variable |
| `dropout` | Censoring indicator: 1 if patient dropped out at interval end | Censoring (C) |

Each patient contributes between 1 and 38 rows (median approximately 15), corresponding to 100-day intervals from seroconversion through up to day 3,700. The treatment variable `haartind` is absorbing: once a patient initiates HAART, it stays 1 for all subsequent intervals. Among the 1,200 patients, 376 (31%) initiate HAART during follow-up. The key time-varying confounder is `cd4.sqrt`, measured at the end of each interval but *before* the treatment and censoring decisions that occur at that same interval's end. This temporal ordering, with CD4 measured first and treatment and outcome following, is what makes causal identification possible under the **sequential exchangeability** assumption. Sequential exchangeability is the longitudinal version of the point-treatment exchangeability assumption from @sec-short-course: at each time point, conditional on the measured past (baseline covariates and the time-varying confounder history up to that point), the treatment decision is independent of future potential outcomes. Informally, once we account for everything measured up to the moment of each treatment decision, the patients who do and do not initiate at that visit are comparable, and no unmeasured confounder drives both the decision and the outcome.


### Example patient trajectories

```{r}
#| label: fig-patient-trajectories
#| fig-cap: "CD4 trajectories for three example patients. Vertical dashed lines mark the interval at which each patient initiates HAART (haartind switches from 0 to 1). Patient 1 initiates relatively early and has a rising CD4 trajectory afterward. Patients who initiate later show different pre-treatment CD4 patterns, illustrating the confounding-by-indication structure: treatment timing depends on the evolving CD4 trajectory."
#| fig-width: 8
#| fig-height: 4
#| code-fold: true

example_ids <- c(1, 5, 10)

# Find first HAART initiation time for each patient
haart_start <- haartdat |>
  filter(patient %in% example_ids, haartind == 1) |>
  group_by(patient) |>
  summarise(haart_time = min(fuptime), .groups = "drop")

haartdat |>
  filter(patient %in% example_ids) |>
  mutate(cd4 = cd4.sqrt^2) |>
  ggplot(aes(x = fuptime, y = cd4)) +
  geom_line(color = "grey40") +
  geom_point(aes(color = factor(haartind)), size = 1.5) +
  geom_vline(
    data = haart_start |> mutate(cd4 = NA),
    aes(xintercept = haart_time),
    linetype = "dashed", color = "steelblue", linewidth = 0.7
  ) +
  facet_wrap(~ paste("Patient", patient), scales = "free_y") +
  scale_color_manual(
    values = c("0" = "grey50", "1" = "steelblue"),
    labels = c("Pre-HAART", "On HAART"),
    name = "Treatment status"
  ) +
  labs(
    x = "Follow-up time (days since seroconversion)",
    y = "CD4 count (cells/\u00B5L)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The figure shows CD4 trajectories on the original (back-transformed) scale. Each patient initiates HAART at a different time, and CD4 counts fluctuate throughout follow-up. This is the longitudinal, time-varying structure that makes causal inference hard: treatment timing depends on the evolving CD4 trajectory, and treatment itself changes that trajectory.


### Cohort summary of baseline characteristics

Several chapters (g-computation, IPTW, point-treatment TMLE) work with a simpler point-treatment version of the question, contrasting *ever* versus *never* initiating HAART using only baseline confounders. For those chapters we collapse the longitudinal data into a baseline snapshot with one row per patient:

:::{.callout-warning}
## This point-treatment collapse induces immortal time bias by construction
Defining treatment as "ever initiated HAART during follow-up" (`ever_haart = max(haartind)`) and then analyzing it as a baseline exposure is not an unbiased analysis of the HAART question. Because `haartind` is an absorbing, time-varying event collapsed to a single baseline label, every patient counted as an "ever initiator" is guaranteed to have survived until their initiation time. That guaranteed survival is precisely the **immortal time bias** described in @sec-roadmap. The point-treatment chapters use this collapse purely to teach estimator mechanics (g-computation, IPTW, TMLE) on one familiar dataset, not to estimate the HAART effect correctly. The unbiased treatment of this question requires the longitudinal g-methods developed in Part III (@sec-longitudinal-confounding, @sec-illustrated-ltmle), which respect treatment timing.
:::


```{r}
#| code-fold: true
haart_baseline <- haartdat |>
  group_by(patient) |>
  summarise(
    sex = first(sex),
    age = first(age),
    cd4_baseline = first(cd4.sqrt)^2,
    cd4_sqrt_baseline = first(cd4.sqrt),
    ever_haart = max(haartind),
    died = max(event),
    dropout = max(dropout),
    total_fup_days = max(fuptime),
    .groups = "drop"
  )

haart_baseline |>
  summarise(
    n = n(),
    pct_female = mean(sex) * 100,
    median_age = median(age),
    median_cd4 = median(cd4_baseline),
    pct_ever_haart = mean(ever_haart) * 100,
    pct_died = mean(died) * 100,
    pct_dropout = mean(dropout) * 100
  ) |>
  knitr::kable(digits = 1, caption = "Baseline cohort summary")
```


## The causal question

The primary causal question is:

> **Among people living with HIV under longitudinal follow-up from seroconversion, what would the risk of death have been under a strategy of initiating HAART when clinically indicated, compared with a strategy of never initiating HAART during follow-up?**

"Clinically indicated" here means a CD4-driven initiation rule consistent with the guidance in force during the cohort era; for concreteness, *initiate HAART at the first interval at which CD4 falls below a fixed threshold (e.g., 350 cells/µL)* and continue thereafter.

It is worth being explicit about what this threshold is and is not. The threshold defines the **intervention regime we want to evaluate**, the counterfactual strategy whose mortality risk we are estimating. It does *not* describe the policy under which the observed data were generated. In the `haartdat` simulation, HAART initiation is not triggered by a hard CD4 cutoff but by a probabilistic model in which the hazard of initiating increases as CD4 falls (and also depends on age and sex). The whole point of the causal analysis is to estimate what would have happened under the cleaner threshold-based strategy, even though the observed initiation process is a noisier, covariate-driven one.

The comparator strategy, "never initiate HAART," is the static contrast used in the marginal structural model literature and the longitudinal analogue of the *ever vs never* contrast used in @sec-gcomputation through @sec-doubly-robust-tmle (where the point-treatment chapters collapse the longitudinal data to a baseline snapshot and ask "ever initiated vs never initiated"). The dynamic-vs-static framing here is deliberate: a single regression cannot estimate the *dynamic* arm without inducing bias, and that is the failure mode that motivates the g-methods developed in Part III. Readers interested in the clinically most relevant contrast (comparing two CD4-driven initiation strategies against each other, e.g., *initiate at CD4 < 350* vs *initiate at CD4 < 200*) will find that dynamic-vs-dynamic comparison developed in @sec-longitudinal-confounding and @sec-illustrated-ltmle.

This is a question about sustained treatment strategies, not a single-timepoint comparison of treated versus untreated patients. The causal question asks what would happen if we could assign treatment strategies to everyone, removing the influence of confounding on the treatment decision. The associational question, by contrast, compares patients who happened to initiate HAART with those who did not, and that comparison is confounded because sicker patients are more likely to start treatment.


## The target trial

To make the causal question precise, we specify the randomized trial we would ideally conduct. This target trial protocol is the design template that the observational analysis tries to emulate.

| Component | Specification |
|-----------|---------------|
| **Eligibility** | Adults living with HIV at seroconversion, antiretroviral-naive |
| **Time zero** | Date of HIV seroconversion |
| **Treatment strategies** | (1) Initiate HAART at the first interval at which CD4 falls below the contemporaneous clinical threshold (e.g., 350 cells/µL) and continue thereafter, vs. (2) never initiate HAART during follow-up |
| **Assignment** | Random assignment to strategy at time zero |
| **Outcome** | All-cause mortality |
| **Follow-up** | From seroconversion until death, dropout, or administrative end of study (up to 3,700 days) |
| **Causal contrast** | Risk difference in mortality at a specified time horizon |
| **Analysis** | Per-protocol: we estimate the effect of *adhering to* each sustained strategy, since the scientific question is about treatment strategies rather than randomized assignment. An ITT analogue is not natural in an observational target-trial emulation because there is no baseline randomization to be "intended." @sec-roadmap discusses the ITT-vs-per-protocol choice as part of the Causal Roadmap, and @sec-longitudinal-confounding walks through how the per-protocol effect is identified from observational data via g-methods. |

The target trial forces us to define the causal question before choosing a statistical method. It makes the assumptions transparent, since we are asking what would have happened under specific, well-defined treatment strategies. And it shows where the observational data fall short of the ideal, because treatment is not randomly assigned but determined by evolving clinical status. The causal roadmap (@sec-roadmap) provides the framework for bridging from this ideal trial to the data we actually have.


## Causal structure

### A simplified DAG for point-treatment chapters

For the early chapters (g-computation, IPTW, TMLE), we work with a point-treatment simplification where treatment is "ever initiated HAART" and confounders are measured at baseline:

```{r}
#| label: fig-simple-dag
#| fig-cap: "Simplified point-treatment DAG for the HAART example. Baseline confounders (age, sex, CD4) affect both treatment and mortality."
#| fig-width: 6
#| fig-height: 4
#| code-fold: true
library(dagitty)
library(ggdag)

simple_dag <- dagify(
  A ~ W_age + W_sex + W_cd4,
  Y ~ A + W_age + W_sex + W_cd4,
  exposure = "A",
  outcome = "Y",
  coords = list(
    x = c(W_age = 0, W_sex = 0, W_cd4 = 0, A = 1, Y = 2),
    y = c(W_age = 1.5, W_sex = 1, W_cd4 = 0.5, A = 0, Y = 0)
  ),
  labels = c(
    W_age = "Age", W_sex = "Sex", W_cd4 = "Baseline\nCD4",
    A = "HAART\ninitiation", Y = "Death"
  )
)

ggdag(simple_dag, text = FALSE, use_labels = "label") +
  theme_dag() +
  labs(title = "Point-treatment DAG")
```

Under the assumptions of consistency, exchangeability (conditional on baseline confounders), and positivity, the causal effect of HAART initiation on mortality is identified by the g-computation formula.

### The longitudinal DAG: treatment-confounder feedback

The real complexity becomes apparent in the longitudinal structure. The following DAG shows two time slices and illustrates the treatment-confounder feedback loop:

```{r}
#| label: fig-longitudinal-dag
#| fig-cap: "Longitudinal DAG for the HAART example showing treatment-confounder feedback. CD4 at time t is affected by prior HAART and also affects future HAART decisions."
#| fig-width: 8
#| fig-height: 5
#| code-fold: true
long_dag <- dagify(
  L1 ~ W,
  A1 ~ L1 + W,
  L2 ~ A1 + L1 + W,
  A2 ~ L2 + A1 + W,
  Y ~ A2 + A1 + L2 + L1 + W,
  C ~ L2 + A1 + W,
  exposure = "A1",
  outcome = "Y",
  coords = list(
    x = c(W = 0, L1 = 1, A1 = 2, L2 = 3, A2 = 4, Y = 5.5, C = 4.5),
    y = c(W = 1, L1 = 1, A1 = 0, L2 = 1, A2 = 0, Y = 0, C = 1.5)
  ),
  labels = c(
    W = "Baseline\n(age, sex)",
    L1 = "CD4(t)",
    A1 = "HAART(t)",
    L2 = "CD4(t+1)",
    A2 = "HAART(t+1)",
    Y = "Death",
    C = "Dropout"
  )
)

ggdag(long_dag, text = FALSE, use_labels = "label") +
  theme_dag() +
  labs(title = "Longitudinal DAG with treatment-confounder feedback")
```

The arrow from HAART(t) to CD4(t+1) is the effect of treatment on the confounder. The arrow from CD4(t+1) to HAART(t+1) is the confounder's influence on future treatment. This is treatment-confounder feedback: CD4 is at once a confounder of future treatment and a consequence of past treatment. Dropout may depend on CD4 levels and prior treatment, which raises the possibility of informative censoring.

This feedback structure is why later chapters need longitudinal g-methods. No single regression model can simultaneously adjust for CD4 as a confounder and preserve its role as a mediator of treatment effects.

### Mapping the DAG to the observed data

The person-period structure of `haartdat` maps directly onto this longitudinal DAG. Each row corresponds to one time slice. Within each interval, the temporal ordering is: CD4 is measured first, then the treatment decision and outcome/censoring occur (all at the end of the interval, with CD4 recorded before the treatment and censoring decisions). The variables map as follows: `cd4.sqrt` is $L_t$, `haartind` is $A_t$, `event` is $Y$, `dropout` is $C$, and `age` and `sex` are the baseline covariates $W$.


## Why standard approaches can mislead

The treatment-confounder feedback structure in this cohort means that naive regression approaches, whether they ignore time-varying CD4 or adjust for it, can produce misleading estimates of the HAART effect. A crude comparison of treated and untreated patients is confounded by indication, because sicker patients start HAART. Adjusting for baseline covariates alone ignores that CD4 changes over time. But adjusting for time-varying CD4 in a pooled regression blocks part of the treatment's beneficial pathway. @sec-regression-vs-causal explains these failure modes in detail and motivates the g-methods (g-computation, IPTW, TMLE) that the rest of the handbook develops.


## Other questions this cohort will answer

The mortality estimand above is the spine of the book, but the same cohort runs through the special-topics chapters, each posing a different question of the same data:

- **Realistic interventions, not all-or-nothing treatment.** Rather than "treat everyone" versus "no one," we can shift the *adherence distribution* and ask how much regimen forgiveness matters (@sec-adherence-shift), and run the full modified-treatment-policy and dynamic-rule workflow from question to estimate (@sec-applied-lmtp).
- **The best rule, and for whom.** At what CD4 threshold, and for which patients, is initiation optimal?
- **Effect modification.** Does the benefit of HAART differ by baseline CD4 or age (@sec-effect-modification)?
- **Mechanism.** How much of the effect runs through CD4 recovery rather than other pathways (@sec-mediation)?
- **Time-to-event and competing risks.** Mortality treated as a survival endpoint, with non-HIV death as a competing event (@sec-tte-competing).
- **Transportability.** Whether an effect estimated in this cohort would carry to a different patient population (@sec-transportability).

Anchoring every method to one cohort lets you see what changes when the question changes, rather than when the dataset changes.


## Sources

- Van der Wal WM, Geskus RB (2011). ipw: An R Package for Inverse Probability Weighting. *Journal of Statistical Software* 43(13):1-23.
- Robins JM, Hernan MA, Brumback B (2000). Marginal structural models and causal inference in epidemiology. *Epidemiology* 11(5):550-560.
- Hernan MA, Brumback B, Robins JM (2000). Marginal structural models to estimate the causal effect of zidovudine on the survival of HIV-positive men. *Epidemiology* 11(5):561-570.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
