---
format: html
---

# Short Course: Causal Inference with Targeted Learning {#sec-short-course}
*From roadmap to TMLE to longitudinal methods in one chapter*

*Estimated reading time: 60 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

This chapter is a self-contained short course that walks through the core ideas of the handbook in a single sitting. It covers the Causal Roadmap, point-treatment TMLE (by hand and with software), longitudinal causal inference, and the `lmtp` package for realistic interventions. Each section is concise; the full handbook chapters provide deeper treatment.

# The Causal Roadmap

It is tempting to load the data and start fitting models. The roadmap resists that, because the model you reach for silently decides which question you end up answering. Working through a fixed sequence keeps the scientific question, rather than the software, in charge. Every causal analysis in this handbook follows the same seven steps:

1. **Define the causal question** using a target trial protocol
2. **Specify the causal model** (draw a DAG)
3. **Define the estimand** (e.g., the average treatment effect on the risk difference scale)
4. **Identify** the estimand from observed data under stated assumptions
5. **Estimate** using an appropriate statistical method
6. **Sensitivity analysis** for violations of the causal assumptions
7. **Interpret** in the context of the original question

The key insight: the causal question and estimand are defined *before* any model is fit. The estimand determines which statistical quantity to target. The estimation method is chosen to match that target.

## Identification: three assumptions

Under three assumptions, the causal estimand equals a function of the observed data:

- **Consistency**: the observed outcome equals the potential outcome under the treatment actually received. Treatment must be well-defined.
- **Exchangeability**: conditional on measured covariates $W$, treatment is independent of potential outcomes. No unmeasured confounders.
- **Positivity**: every covariate stratum has a nonzero probability of receiving each treatment. $0 < P(A = a \mid W) < 1$.

Under these assumptions, the average treatment effect is identified by:

$$
E[Y(1)] - E[Y(0)] = E_W\!\left[E[Y \mid A=1, W] - E[Y \mid A=0, W]\right]
$$

This is the g-computation formula. It says: model the outcome within levels of treatment and covariates, then average over the covariate distribution.


# Point-treatment TMLE

## Setup: simulated clinical trial data

```{r}
library(tidyverse)

set.seed(2026)
n <- 1000

# Baseline covariates
age <- rnorm(n, 65, 10)
cvd <- rbinom(n, 1, plogis(-2 + 0.03 * age))

# Treatment (confounded by age and CVD history)
A <- rbinom(n, 1, plogis(-0.5 + 0.02 * age + 0.5 * cvd))

# Outcome (binary: adverse event)
Y <- rbinom(n, 1, plogis(-2.5 + 0.03 * age + 0.5 * cvd - 0.4 * A))

dat <- tibble(age, cvd, A, Y)
cat("n =", n, ", treatment rate:", round(mean(A), 2),
    ", outcome rate:", round(mean(Y), 2), "\n")
```

## G-computation in three steps

G-computation is the simplest causal estimator. It implements the identification formula directly.

```{r}
# Step 1: Fit outcome model
Qmod <- glm(Y ~ A + age + cvd, family = binomial, data = dat)

# Step 2: Predict under each treatment for everyone
Q1 <- predict(Qmod, newdata = mutate(dat, A = 1), type = "response")
Q0 <- predict(Qmod, newdata = mutate(dat, A = 0), type = "response")

# Step 3: Average (standardize)
gcomp_ate <- mean(Q1) - mean(Q0)
cat("G-computation ATE:", round(gcomp_ate, 4), "\n")
```

G-computation depends entirely on the outcome model. If that model is wrong, the estimate is biased and nothing downstream repairs it. Two ideas in the rest of this section address that: a flexible way to fit the models (Super Learner), and an estimator that stays consistent if either the outcome model or the treatment model is right (TMLE).

## TMLE by hand

TMLE adds a targeting step that corrects the outcome model using the propensity score. This makes it doubly robust: consistent if either the outcome model or the treatment model is correctly specified. The correction nudges the initial predictions hardest for patients whose treatment was least expected given their covariates, which is where the outcome model is most likely to be wrong; the propensity score enters only through that nudge, the *clever covariate* `H` below.

```{r}
logit <- function(p) log(p / (1 - p))

# Step 1: Initial outcome predictions (already done above)
Qinit <- predict(Qmod, type = "response")

# Step 2: Estimate propensity score
gmod <- glm(A ~ age + cvd, family = binomial, data = dat)
ps <- predict(gmod, type = "response")
ps <- pmax(0.01, pmin(0.99, ps))  # bound away from 0/1

# Step 3: Clever covariate
H <- dat$A / ps - (1 - dat$A) / (1 - ps)

# Step 4: Fluctuation (the targeting step)
epsilon <- glm(dat$Y ~ -1 + offset(logit(Qinit)) + H,
               family = binomial)$coef

# Update counterfactual predictions
H1 <- 1 / ps
H0 <- -1 / (1 - ps)
Q1_star <- plogis(logit(Q1) + epsilon * H1)
Q0_star <- plogis(logit(Q0) + epsilon * H0)

# TMLE estimate
tmle_ate <- mean(Q1_star) - mean(Q0_star)
cat("TMLE ATE:", round(tmle_ate, 4), "\n")
```

The fluctuation parameter $\hat{\epsilon}$ captures how much the treatment model corrects the outcome model. If the outcome model is already well-calibrated for the ATE, $\hat{\epsilon}$ is near zero.

## Inference from the efficient influence curve

Each patient contributes one number to the estimate, the patient's value of the *efficient influence curve*. The estimate behaves like the average of those numbers, so its standard error is their standard deviation divided by the square root of `n`, and a few patients with extreme values are what widen the interval.

```{r}
# Efficient influence curve
Qstar_obs <- ifelse(dat$A == 1, Q1_star, Q0_star)
H_obs <- ifelse(dat$A == 1, H1, H0)
EIC <- H_obs * (dat$Y - Qstar_obs) + Q1_star - Q0_star - tmle_ate

# Standard error and 95% CI
se <- sqrt(var(EIC) / n)
ci <- tmle_ate + c(-1.96, 1.96) * se

cat("ATE:", round(tmle_ate, 4), "\n")
cat("SE: ", round(se, 4), "\n")
cat("95% CI: [", round(ci[1], 4), ",", round(ci[2], 4), "]\n")
```

## Super Learner: flexible models for the nuisances

The code above fit the outcome and treatment models with `glm`, which assumes the relationships are linear on the logit scale. When they are not, both nuisance estimates are biased, and so is the causal estimate. Super Learner replaces a single guessed model with a library of candidate algorithms, for example a logistic regression, a spline model, and a random forest, and uses cross-validation to weight them by how well each predicts data it did not see. The ensemble is, in large samples, about as accurate as the best candidate in the library, without the analyst having to know in advance which that is. In a targeted-learning analysis Super Learner estimates the nuisance models and TMLE does the targeting on top; the software call below sets this up through its `Q.SL.library` and `g.SL.library` arguments.

## Software implementation

The `tmle` package automates all of this, including Super Learner for flexible nuisance estimation:

```{r}
#| eval: false
library(tmle)

result <- tmle(
  Y = dat$Y, A = dat$A,
  W = dat[, c("age", "cvd")],
  Q.SL.library = c("SL.glm", "SL.mean"),
  g.SL.library = c("SL.glm", "SL.mean")
)
result
```


# Longitudinal causal inference

## The problem: treatment-confounder feedback

In longitudinal settings, a time-varying confounder (like CD4 count) can be both a confounder of future treatment and a consequence of past treatment. Standard regression cannot handle both roles simultaneously: adjusting for CD4 blocks the treatment's indirect effect through CD4, while not adjusting leaves confounding.

This is treatment-confounder feedback, and it requires g-methods.

## Simulated longitudinal data

```{r}
set.seed(2026)
n <- 2000

# Baseline
W_age <- rnorm(n, 40, 10)
W_male <- rbinom(n, 1, 0.6)
W_cd4_bl <- pmax(50, rnorm(n, 350, 120))
W_vl_bl <- rnorm(n, 4.2, 0.7)
W <- 0.01 * (W_age - 40) - 0.001 * (W_cd4_bl - 350) + 0.2 * W_vl_bl - 0.1 * W_male

# Visit 1
A1 <- rbinom(n, 1, plogis(0.3 + 0.5 * W))
L1 <- rnorm(n, mean = 0.3 * W + 1.2 * A1, sd = 1)  # treatment affects confounder

# Visit 2 (confounder affects future treatment)
A2 <- rbinom(n, 1, plogis(-0.5 + 0.8 * L1 + 0.3 * A1))

# Outcome
Y <- rbinom(n, 1, plogis(-0.8 - 0.4 * A1 - 0.5 * A2 - 0.6 * L1 + 0.2 * W))
dat_long <- tibble(W, A1, L1, A2, Y)
```

## Longitudinal G-computation by hand

The backwards sequential regression resolves the feedback. At time 2, condition on $L_1$ to deconfound $A_2$. At time 1, do NOT condition on $L_1$ to preserve the mediated effect.

```{r}
# Oracle truth
set.seed(9999)
n_mc <- 100000; W_mc <- rep(W, length.out = n_mc)
L1_cf <- rnorm(n_mc, mean = 0.3 * W_mc + 1.2, sd = 1)
true_EY_11 <- mean(plogis(-0.8 - 0.4 - 0.5 - 0.6 * L1_cf + 0.2 * W_mc))

# Step 1: Model outcome at time 2, conditioning on L1
Q2_mod <- glm(Y ~ A2 + L1 + A1 + W, family = binomial, data = dat_long)
Q2_A2is1 <- predict(Q2_mod, newdata = mutate(dat_long, A2 = 1), type = "response")

# Step 2: Backwards regression: model Q2 predictions WITHOUT L1
Q1_mod <- glm(Q2_A2is1 ~ A1 + W, family = quasibinomial, data = dat_long)
Q1_A1is1 <- predict(Q1_mod, newdata = mutate(dat_long, A1 = 1), type = "response")

gcomp_long <- mean(Q1_A1is1)

# Compare to naive regression (which conditions on L1 throughout)
naive_mod <- glm(Y ~ A1 + A2 + L1 + W, family = binomial, data = dat_long)
naive_pred <- mean(predict(naive_mod,
  newdata = mutate(dat_long, A1 = 1, A2 = 1), type = "response"))

cat("True E[Y(1,1)]:     ", round(true_EY_11, 4), "\n")
cat("Naive (biased):      ", round(naive_pred, 4), "\n")
cat("G-computation (long):", round(gcomp_long, 4), "\n")
```

The naive estimate is biased because it conditions on $L_1$ (blocking part of the treatment effect). The backwards G-computation avoids this.

Longitudinal **TMLE** adds a targeting step to each of these backwards regressions, the same fluctuation idea as in the point-treatment case, applied at every time point. This buys *sequential* double robustness: the estimate is consistent if, at each step, either the outcome regression or the treatment model is correct. The mechanics are developed in @sec-illustrated-ltmle; here we turn to the kind of intervention that longitudinal data make possible.

## LMTP for realistic interventions

Static regimes ("always treat") often violate positivity. The `lmtp` package estimates modified treatment policies that shift treatment by a feasible amount.

```{r}
#| eval: false
library(lmtp)

# Shift function: increase adherence by 10 percentage points
shift_up <- function(data, trt) pmin(data[[trt]] + 0.10, 1.0)

# Natural course (no intervention)
shift_none <- function(data, trt) data[[trt]]

# Estimate under each regime
fit_shifted <- lmtp_tmle(
  data = my_data,
  trt = paste0("PDC_", 1:4),
  outcome = paste0("Y_", 1:4),
  baseline = c("W0", "A0"),
  time_vary = lapply(1:4, function(t) paste0("L_", t)),
  shift = shift_up,
  mtp = TRUE,
  outcome_type = "survival",
  learners_outcome = c("SL.glm", "SL.mean"),
  learners_trt = c("SL.glm", "SL.mean"),
  folds = 2
)

# Contrast vs natural course
contrast <- lmtp_contrast(fit_shifted, ref = fit_natural, type = "additive")
contrast
```

The shift function defines the intervention: increase each patient's adherence by 10 percentage points at each visit, capped at 1.0. This stays within the observed data support because no patient is assigned an impossible adherence level. The `mtp = TRUE` flag tells `lmtp` that the shifted value depends on the natural treatment value.


# Diagnostics and sensitivity

No causal analysis is complete without checking its assumptions and stability.

**Positivity.** Plot propensity scores. If they pile up near 0 or 1, the estimand is poorly supported. Consider restricting the population or using a shift intervention.

**Model diagnostics.** Check Super Learner weights (which learners are driving predictions?), calibration of outcome and treatment models, and whether the TMLE fluctuation parameter $\hat{\epsilon}$ is unusually large.

**Sensitivity to unmeasured confounding.** E-values quantify how strong an unmeasured confounder would need to be to explain away the result. Negative control outcomes test whether the analysis pipeline produces null effects where expected.

**Sensitivity to analytic choices.** Vary the Super Learner library, propensity score truncation threshold, and (for longitudinal analyses) the number of time points. If the estimate is unstable across these choices, report the range rather than a single number.

These tools are covered in detail in the [Positivity](03-02-positivity-overlap.qmd), [Diagnostics](03-05-advanced-diagnostics.qmd), and [Sensitivity Analysis](03-12-sensitivity-analysis.qmd) chapters.


# Where to go next

This short course covered the essentials. The full handbook develops each topic with more examples, diagnostics, and worked code:

- **Part I, Foundations:** Why regression coefficients are not causal effects; the full 7-step Causal Roadmap
- **Part II, Core Estimators:** G-computation, IPTW, TMLE, and Super Learner in detail
- **Part III, Longitudinal Causal Inference:** Time-varying confounding and treatment-confounder feedback, longitudinal TMLE (L-TMLE), and Longitudinal Modified Treatment Policies (LMTP)
- **Part IV, Diagnostics and Sensitivity Analysis:** Positivity and overlap, model diagnostics, and sensitivity analysis for unmeasured confounding
- **Part V, Putting It All Together:** An end-to-end applied workflow built around the `lmtp` package
- **Part VI, Special Topics:** Effect heterogeneity, mediation, hybrid RCT/RWD designs, and time-to-event estimands

Each part builds on the running HAART case study, so you can see how the methods connect rather than treating them in isolation.
