---
title: "Chapter 1.2: The Causal Roadmap"
format: html
---

In this chapter, we introduce the **Causal Roadmap** (Petersen & van der Laan, 2014, "Causal models and learning from data: integrating causal modeling and statistical estimation"), a structured approach for asking and answering causal questions with observational data. We begin with the **target trial protocol**, which forces you to define the scientific question precisely before choosing any statistical method, and then walk through the **six steps of the causal roadmap** that translate that question into a defensible analysis.

::: {.callout-note}
## Learning objectives
- Describe the six steps of the Causal Roadmap and their purpose
- Formulate a causal question in terms of population, intervention, comparator, and outcome (PICO)
- Draw and interpret a directed acyclic graph (DAG) encoding causal assumptions
- Distinguish a causal estimand from a statistical estimand
- Explain why pre-specifying an analysis plan before data access reduces researcher degrees of freedom
:::


```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
library(dagitty)
library(ggdag)
library(ggplot2)
```

### Why Use a Causal Roadmap?

In epidemiology, it is tempting to jump straight to fitting a model: "Let's run a regression and see what we find." But this approach conflates the scientific question with the statistical procedure, making it difficult to evaluate whether the analysis actually answers the question you care about. Modern causal inference begins by asking: **What is the causal question we are trying to answer?**

**Key point.** A causal roadmap ensures transparency in how evidence is generated, separation of scientific questions from statistical methods, clear articulation of assumptions, and structured thinking in both trial and non-randomized settings. Starting with a well-defined question, before touching data or code, is the single most important step in any causal analysis.

We'll walk step-by-step through each part of the roadmap, with illustrative examples.


## Step 1: Define the Causal Question

In traditional epidemiologic practice, the form of the outcome often dictates the analysis: binary outcomes get logistic regression, time-to-event data get Cox models, and the resulting coefficient is reported as "the effect." But this approach lets the **statistical model choose the estimand for you**. A hazard ratio from a Cox model answers a different question than a 3-year risk difference, and neither may correspond to the causal contrast that would actually inform a clinical or policy decision. Step 1 of the roadmap insists that you define the causal question *first*, before any model is fit.

### Start with the target trial protocol, then follow the roadmap

As advocated by Dang and Balzer (2023), a useful way to make the causal question precise is to sketch the **target trial protocol** --- the randomized trial you would ideally like to run, even though you only have observational data. This is a design device, not a claim that every observational study literally recreates a trial. In practice, this means specifying:

- **Eligibility criteria**: Who would be included?
- **Time zero**: When would follow-up begin?
- **Treatment strategies**: What interventions are being compared?
- **Assignment procedure**: Would treatment be randomized, or are we approximating that contrast with observed data?
- **Outcome**: What endpoint would be measured?
- **Follow-up**: Over what time horizon?
- **Causal contrast**: Risk difference, risk ratio, restricted mean survival time difference, or another estimand?
- **Analysis plan**: How will confounding, censoring, and missingness be handled?

This prevents a very common problem: starting with whatever variables happen to be available, fitting a model, and only afterward deciding what question was answered. The **causal roadmap** then translates that ideal question into a defensible analysis by making the causal model, identifiability assumptions, statistical estimand, estimator choice, diagnostics, and interpretation explicit.

> **Target trial emulation clarifies the causal question. The causal roadmap tells us how to analyze that question rigorously with the data we actually have.**

**Example:**

> Among postmenopausal women with osteoporosis who are eligible to initiate either therapy, what is the 3-year risk of cardiovascular events if all initiated **denosumab** compared with if all initiated **zoledronic acid**, starting at treatment initiation and followed for 3 years?

## Step 2: Specify the Causal Model

We now draw out how we believe the data were generated. This includes the treatment (A), outcome (Y), and covariates (W), encoded as a directed acyclic graph (DAG) with potential outcomes $Y(1)$ and $Y(0)$ for each individual.

A Structural Causal Model (SCM) is simply a formal way of writing down your story about how the world works. Think of it as a set of equations that say: "This variable is caused by these other variables, plus some factors we cannot measure." For MPH students, the SCM is the mathematical version of the causal DAG you draw on a whiteboard. Each arrow in the DAG corresponds to one of these equations. The key advantage is that it forces you to be explicit about every assumption you are making about what causes what.


The SCM for our example:

$$
W = f_W(U_W),\
A = f_A(W, U_A),\
Y = f_Y(W, A, U_Y)
$$

Where $U$ terms represent unmeasured variables. The corresponding DAG:

```{r}
#| label: fig-roadmap-dag
#| fig-cap: "Basic causal DAG: W confounds the A → Y relationship"
#| fig-width: 5
#| fig-height: 3
dag_roadmap <- dagify(
  A ~ W, Y ~ A + W,
  exposure = "A", outcome = "Y",
  coords = list(
    x = c(W = 0, A = 1, Y = 2),
    y = c(W = 1, A = 0, Y = 0)
  )
)
ggdag(dag_roadmap) + theme_dag()
```

This DAG implies W (confounders) affect both A and Y, A (treatment) affects Y, and there is no unmeasured confounding.


With the causal model in hand, we can now define exactly what quantity we want to estimate.

## Step 3: Define the Target Causal Estimand


The simplest causal estimand is the **Average Treatment Effect (ATE)**:

$$
ATE = E[Y(1) - Y(0)]
$$

Or equivalently, the **Risk Difference**: $E[Y \mid do(A=1)] - E[Y \mid do(A=0)]$.

But the ATE is only one option. Depending on the scientific question, we may want a richer contrast:

- **Conditional Average Treatment Effects (CATEs)**: treatment effects within subgroups defined by covariates (e.g., the effect among patients with prior CVD). These are useful for understanding treatment effect heterogeneity.
- **Stochastic intervention regimes**: instead of asking "what if everyone were treated?", ask "what if the probability of treatment increased by 20%?" This is more realistic when universal treatment is implausible and avoids positivity violations (see the `lmtp` R package and Chapter 3.11).
- **Dynamic treatment regimes**: rules that assign treatment based on a patient's evolving characteristics (e.g., "start treatment when blood pressure exceeds 140"). These are common in longitudinal settings.
- **Optimal individualized treatment regimes**: the rule that maximizes the expected outcome for each patient given their covariates. These can be estimated using the `tlverse` package (see [Chapter: Optimal Intervention Regimes](04-01-optimal-regimes.qmd)).

Each of these estimands requires its own identification argument and estimation strategy. This handbook covers all of them. For a roadmap of which chapters cover which estimand, see the [handbook overview](index.qmd). We also cover [outcome-blind simulation studies](04-02-outcome-blind-simulations.qmd) for evaluating estimator performance before analyzing real data.

**Key point.** Why prefer risk differences over odds ratios or hazard ratios? The risk difference directly tells you: "Treatment changes the probability of the outcome by X percentage points." It is a **marginal causal parameter** that does not depend on which covariates you condition on. Odds ratios, by contrast, are non-collapsible (they change when you marginalize over covariates even without confounding), lack a straightforward population-level causal interpretation, and are frequently misinterpreted as risk ratios. Hazard ratios share similar non-collapsibility issues and depend on the proportional hazards assumption. Additionally, the risk difference is **less sensitive to background disease prevalence** than ratio measures --- the same odds ratio can correspond to very different absolute risk changes depending on how common the outcome is, making ratio measures harder to interpret for clinical decision-making. When your goal is to inform a public health decision, the risk difference answers that question directly.

## Step 4: Link the Causal Estimand to the Observed Data

This is the **identification step** (we go into much more detail in [Chapter 1.3](01-03-identification-estimands.qmd)): *Can we estimate the causal effect from the data we have?* These assumptions are fundamentally untestable and must be justified by subject-matter knowledge.

- **Exchangeability**: $Y(a) \perp A \mid W$. You have measured every common cause of treatment and outcome.
- **Positivity**: $P(A = a \mid W = w) > 0$. No covariate stratum deterministically receives one treatment.
- **Consistency**: The observed outcome equals the potential outcome under the received treatment.

If satisfied: $E[Y(1) - Y(0)] = E_W[ E[Y \mid A = 1, W] - E[Y \mid A = 0, W] ]$.


### Step 5: Choose a Statistical Estimator

This step translates the identified quantity into an algorithm we can apply to data. You *could* use a standard regression model here (e.g., fit a logistic regression and standardize predictions via g-computation), but modern methods offer important advantages in robustness and flexibility. Options covered in upcoming chapters include:

- **G-computation** ([Chapter 2.1](02-01-gcomputation.qmd)): Predict outcomes under each treatment, average over W
- **IPTW** ([Chapter 2.2](02-02-iptw.qmd)): Weight observations by inverse probability of treatment
- **TMLE** ([Chapter 2.3](02-03-doubly-robust-tmle.qmd)): Targeted learning that combines outcome and treatment models for double robustness
- **Longitudinal TMLE (L-TMLE)** ([Chapter 3.9](03-09-illustrated-ltmle.qmd)): For time-varying treatments, time-varying confounding, and survival outcomes --- estimates causal survival curves without proportional hazards assumptions

The choice depends on the assumptions you're willing to make, the complexity of the data structure (point treatment vs. longitudinal), sample size, and tolerance for model misspecification.

### Step 6: Evaluate and Interpret

Once you've estimated your parameter, interpretation is **not automatic**. Ask yourself: Are your assumptions plausible? What does your estimate mean for real-world decisions? Can you generalize to other populations? No amount of statistical sophistication can rescue an analysis built on implausible assumptions. Use sensitivity analyses, negative control outcomes, covariate balance checks, and expert consultation to stress-test your conclusions.

We cover diagnostics in detail in [Chapter 3.5: Advanced Diagnostics](03-05-advanced-diagnostics.qmd) (covariate balance, weight distributions, influence curves) and sensitivity analyses in [Chapter 3.12: Sensitivity Analysis](03-12-sensitivity-analysis.qmd) (E-values, quantitative bias analysis, and bounding approaches for unmeasured confounding).

:::{.callout}
## Summary

The Causal Roadmap gives a rigorous framework for designing and evaluating real-world effect estimates. Before choosing a statistical method, start with the science: define your causal question, specify your model and assumptions, decide how to express the effect, and choose a transparent, defensible estimation method.
:::

## Check your understanding

<details>
<summary><strong>Q1. Why must the causal question be defined before choosing a statistical method?</strong></summary>

The causal question determines everything downstream: which variables are confounders (and therefore must be adjusted for), what the target estimand is (e.g., ATE on the risk difference scale vs. a conditional odds ratio), and which assumptions are required for identification. If you choose a statistical method first, say, logistic regression, you have implicitly committed to a conditional odds ratio as your estimand, which may not match the scientific question you actually care about. Starting with the question also prevents "researcher degrees of freedom": when the estimand is pre-specified, there is less temptation to explore multiple model specifications and selectively report favorable results.

</details>

<details>
<summary><strong>Q2. What is the difference between a causal estimand and a statistical estimand?</strong></summary>

A **causal estimand** is defined in terms of potential outcomes, for example, $E[Y(1) - Y(0)]$, the average treatment effect. It describes a contrast between hypothetical worlds and cannot be directly computed from observed data without assumptions. A **statistical estimand** is a function of the observed data distribution, for example, $E_W[E[Y \mid A=1, W] - E[Y \mid A=0, W]]$. The identification step of the Causal Roadmap shows that, *under specific assumptions* (exchangeability, positivity, consistency), the causal estimand equals a particular statistical estimand. Without those assumptions, the statistical quantity you compute may have no causal interpretation at all.

</details>

<details>
<summary><strong>Q3. Can covariate balance checks prove that exchangeability holds?</strong></summary>

No. Covariate balance checks (e.g., standardized mean differences after weighting) can only assess whether **measured** covariates are balanced between treatment groups. Exchangeability requires that there are **no unmeasured common causes** of treatment and outcome, a condition that depends on subject-matter knowledge and cannot be verified from the data. Good balance on observed covariates is necessary but not sufficient: an unmeasured confounder can induce bias even when every measured covariate is perfectly balanced. Balance checks are a useful diagnostic, but they should never be interpreted as evidence that the exchangeability assumption holds.

</details>

## What comes next

With the causal question defined and the roadmap in hand, the next chapter formalizes the identification step: the assumptions needed to bridge potential outcomes and observed data, and the statistical estimands that result.

## Sources and further reading

- Dang LE, Tager I, Petersen ML (2023). A causal roadmap for generating high-quality real-world evidence. *Journal of Clinical and Translational Science* 7(1):e127.
- Hernan MA, Robins JM (2016). Using big data to emulate a target trial when a randomized trial is not available. *Am J Epidemiol* 183(8):758-764.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Gruber S, van der Laan MJ (2010). A targeted maximum likelihood estimator of a causal effect on a bounded continuous outcome. *Int J Biostat* 6(1):Article 26.
- ICH E9 (R1) Addendum (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)
