---
title: "Chapter 1.2: The Causal Roadmap"
format: html
---

*Estimated reading time: 40 minutes*

In this chapter, we introduce the **Causal Roadmap** (Petersen & van der Laan, 2014, "Causal models and learning from data: integrating causal modeling and statistical estimation"), a structured approach for asking and answering causal questions with observational data.

::: {.callout-note}
## Learning objectives
- Describe the six steps of the Causal Roadmap and their purpose
- Formulate a causal question using a target trial protocol
- Recognize immortal time bias, prevalent user bias, and vague comparator bias as failures of target trial thinking
- Draw and interpret a directed acyclic graph (DAG) encoding causal assumptions
- State the three core identification assumptions: consistency, exchangeability, and positivity
- Distinguish a causal estimand from a statistical estimand
- Diagnose positivity violations using propensity score overlap plots
- Explain why pre-specifying an analysis plan before data access reduces researcher degrees of freedom
:::


```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
library(dagitty)
library(ggdag)
library(ggplot2)
```

### A motivating disaster: immortal time bias in statin studies

In 2006, several observational studies reported that statin users had substantially lower mortality than non-users, not just from cardiovascular disease, but from cancer, trauma, and even infections. The effect sizes were implausibly large. Were statins a miracle drug?

No. The problem was **immortal time bias**, a direct consequence of failing to align time zero with treatment assignment:

1. Researchers identified patients who filled a statin prescription at some point during follow-up.
2. They classified the entire follow-up period as "statin exposed," including time *before* the prescription was filled.
3. During that early period, the patient could not have died (otherwise they would never have filled the prescription). That time was, by definition, "immortal."
4. Non-users had no such guaranteed survival period.

The result was a spurious survival advantage that had nothing to do with the drug. A structured analytic framework (a **Causal Roadmap**) would have caught this error at the design stage, before any model was fit.

:::{.callout-important}
## The core lesson
Immortal time bias arises when time zero is not aligned with the moment of treatment assignment. A target trial protocol forces you to define time zero explicitly, preventing this error before any data are analyzed.
:::

### Why Use a Causal Roadmap?

In epidemiology, it is tempting to jump straight to fitting a model: "Let's run a regression and see what we find." But this approach conflates the scientific question with the statistical procedure, making it difficult to evaluate whether the analysis actually answers the question you care about. The statin example illustrates a more insidious version of this problem: even before worrying about confounding or model specification, the study *design* was flawed.

**Key point.** A causal roadmap ensures transparency in how evidence is generated, separation of scientific questions from statistical methods, clear articulation of assumptions, and structured thinking in both trial and non-randomized settings. Starting with a well-defined question, before touching data or code, is the single most important step in any causal analysis.

::: {.callout-note}
## Mapping to commonly cited roadmap formulations

This handbook presents the causal roadmap in **six steps** (question, model, estimand, identification, estimation, interpretation), following @petersen2014causal. Other formulations exist in the literature:

| This handbook | Petersen & van der Laan (2014) | Dang et al. (2023) high-quality RWE |
|--------------|-------------------------------|-------------------------------------|
| Step 1: Define the question | Step 1: Causal model | Step 1: Define causal question |
| Step 2: Structural causal model | Step 2: Target causal quantity | Step 2: Specify causal model |
| Step 3: Define the estimand | Step 3: Identification | Step 3: Define target quantity |
| Step 4: Identification | Step 4: Estimation | Step 4: Assess identifiability |
| Step 5: Estimation | Step 5: Interpretation | Step 5: Statistical estimation |
| Step 6: Interpretation | (see Step 5) | Step 6: Interpret, sensitivity |

The step numbers do not map one-to-one across these formulations. The table shows conceptual alignment, not step-number correspondence. The core logic is the same: separate the *scientific question* from the *statistical procedure*, and make assumptions explicit before touching data. The Dang et al. formulation [@dang2023causal] adds explicit data quality assessment and reporting requirements, which we cover in [Chapter 3.1](03-01-rwe-analysis-workflow.qmd) and [Chapter 3.8](03-08-tmle-clean-room.qmd).
:::


## Step 1: Define the Causal Question Using a Target Trial Protocol

In traditional epidemiologic practice, the form of the outcome often dictates the analysis: binary outcomes get logistic regression, time-to-event data get Cox models, and the resulting coefficient is reported as "the effect." But this approach lets the **statistical model choose the estimand for you**. A hazard ratio from a Cox model answers a different question than a 3-year risk difference, and neither may correspond to the causal contrast that would actually inform a clinical or policy decision. Step 1 of the roadmap insists that you define the causal question *first*, before any model is fit.

### Start with the target trial protocol

As advocated by @dang2023causal and @hernan2016target, a useful way to make the causal question precise is to sketch the **target trial protocol**, the randomized trial you would ideally like to run, even though you only have observational data. This is a design device, not a claim that every observational study literally recreates a trial. In practice, this means specifying:

- **Eligibility criteria**: Who would be included?
- **Time zero**: When would follow-up begin? (This must coincide with eligibility and treatment assignment; see the statin example above.)
- **Treatment strategies**: What interventions are being compared? (Vague strategies like "statin use" vs. "no statin use" lead to vague causal questions.)
- **Assignment procedure**: Would treatment be randomized, or are we approximating that contrast with observed data?
- **Outcome**: What endpoint would be measured?
- **Follow-up**: Over what time horizon?
- **Causal contrast**: Risk difference, risk ratio, restricted mean survival time difference, or another estimand?
- **Analysis plan**: How will confounding, censoring, and missingness be handled?

This prevents a very common problem: starting with whatever variables happen to be available, fitting a model, and only afterward deciding what question was answered. The **causal roadmap** then translates that ideal question into a defensible analysis by making the causal model, identifiability assumptions, statistical estimand, estimator choice, diagnostics, and interpretation explicit.

> **Target trial emulation clarifies the causal question. The causal roadmap tells us how to analyze that question rigorously with the data we actually have.**

**Example:**

> Among postmenopausal women with osteoporosis who are eligible to initiate either therapy, what is the 3-year risk of cardiovascular events if all initiated **denosumab** compared with if all initiated **zoledronic acid**, starting at treatment initiation and followed for 3 years?

### Common biases prevented by target trial thinking

Many well-known biases in observational research can be traced to violations of specific target trial protocol components:

- **Immortal time bias** (violated: time zero): Follow-up begins before treatment assignment. Prevention: align time zero with the moment of treatment assignment using a new-user design.
- **Prevalent user bias** (violated: eligibility + time zero): Including patients already on treatment at the start of follow-up. These patients are a selected subset who survived early treatment and tolerated it. Prevention: restrict to new (incident) users who initiate treatment at time zero.
- **Vague comparator definitions** (violated: treatment strategies): "Everyone not on the drug" includes a mix of people on alternatives, people with contraindications, and people who will start later. Prevention: define specific, well-characterized treatment strategies (e.g., active comparator designs).

:::{.callout-note}
## A unifying insight
Nearly every major bias in observational pharmacoepidemiology (immortal time bias, prevalent user bias, selection bias from conditioning on the future, time-window bias) is a violation of one or more target trial protocol components. If you can specify and emulate the protocol correctly, you prevent these biases by design rather than by statistical adjustment.
:::

### Simulation: immortal time bias in action

The following simulation demonstrates how failing to align time zero creates spurious treatment effects. The treatment has **no causal effect** by construction, but a naive analysis produces a large apparent benefit.

```{r}
library(tidyverse)

set.seed(2024)
n <- 5000

# --- Data-generating process (NO treatment effect) ---
baseline_hazard <- 0.10  # annual probability of event
max_follow_up <- 5       # years

sim_data <- tibble(
  id = 1:n,
  will_treat = rbinom(n, 1, 0.6),
  time_to_treat = ifelse(will_treat == 1, runif(n, 0, 3), Inf),
  event_time = rexp(n, rate = baseline_hazard),
  obs_time = pmin(event_time, max_follow_up),
  event = as.integer(event_time <= max_follow_up),
  actually_treated = as.integer(will_treat == 1 & time_to_treat < obs_time)
)

# --- WRONG analysis: ignore treatment timing ---
wrong_analysis <- sim_data |>
  group_by(actually_treated) |>
  summarise(
    n = n(), events = sum(event),
    person_years = sum(obs_time),
    rate = events / person_years, .groups = "drop"
  )

rate_ratio_wrong <- wrong_analysis$rate[wrong_analysis$actually_treated == 1] /
  wrong_analysis$rate[wrong_analysis$actually_treated == 0]

cat("True treatment effect: NONE (null by construction)\n")
cat("Biased rate ratio:", round(rate_ratio_wrong, 2),
    "(expected if no bias: 1.00)\n")
```

The biased rate ratio is well below 1.0, suggesting treatment is protective, even though the treatment has **zero** causal effect. The entire "benefit" comes from the fact that treated patients had to survive long enough to receive treatment. No statistical method can fix a fundamentally flawed study design.

## Step 2: Specify the Causal Model

We now draw out how we believe the data were generated. This includes the treatment (A), outcome (Y), and covariates (W), encoded as a directed acyclic graph (DAG) with potential outcomes $Y(1)$ and $Y(0)$ for each individual.

A Structural Causal Model (SCM) is simply a formal way of writing down your story about how the world works. Think of it as a set of equations that say: "This variable is caused by these other variables, plus some factors we cannot measure." For MPH students, the SCM is the mathematical version of the causal DAG you draw on a whiteboard. Each arrow in the DAG corresponds to one of these equations. The key advantage is that it forces you to be explicit about every assumption you are making about what causes what.


The SCM for our example:

$$
W = f_W(U_W),\
A = f_A(W, U_A),\
Y = f_Y(W, A, U_Y)
$$

Where $U$ terms represent unmeasured variables. The corresponding DAG:

```{r}
#| label: fig-roadmap-dag
#| fig-cap: "Basic causal DAG: W confounds the A → Y relationship"
#| fig-width: 5
#| fig-height: 3
dag_roadmap <- dagify(
  A ~ W, Y ~ A + W,
  exposure = "A", outcome = "Y",
  coords = list(
    x = c(W = 0, A = 1, Y = 2),
    y = c(W = 1, A = 0, Y = 0)
  )
)
ggdag(dag_roadmap) + theme_dag()
```

This DAG implies W (confounders) affect both A and Y, A (treatment) affects Y, and there is no unmeasured confounding.


With the causal model in hand, we can now define exactly what quantity we want to estimate.

## Step 3: Define the Target Causal Estimand


The simplest causal estimand is the **Average Treatment Effect (ATE)**:

$$
ATE = E[Y(1) - Y(0)]
$$

Or equivalently, the **Risk Difference**: $E[Y(1)] - E[Y(0)]$.

But the ATE is only one option. Depending on the scientific question, we may want a richer contrast:

- **Conditional Average Treatment Effects (CATEs)**: treatment effects within subgroups defined by covariates (e.g., the effect among patients with prior CVD). These are useful for understanding treatment effect heterogeneity.
- **Stochastic intervention regimes**: instead of asking "what if everyone were treated?", ask "what if the probability of treatment increased by 20%?" This is more realistic when universal treatment is implausible and avoids positivity violations (see the `lmtp` R package and [Chapter 3.3b: Adherence Shift Methods](03-03b-adherence-shift-methods.qmd)).
- **Dynamic treatment regimes**: rules that assign treatment based on a patient's evolving characteristics (e.g., "start treatment when blood pressure exceeds 140"). These are common in longitudinal settings.
- **Optimal individualized treatment regimes**: the rule that maximizes the expected outcome for each patient given their covariates. These can be estimated using the `tlverse` package (see [Chapter: Optimal Intervention Regimes](04-01-optimal-regimes.qmd)).

Each of these estimands requires its own identification argument and estimation strategy. This handbook covers all of them. For a roadmap of which chapters cover which estimand, see the [handbook overview](index.qmd). We also cover [outcome-blind simulation studies](04-02-outcome-blind-simulations.qmd) for evaluating estimator performance before analyzing real data.

**Key point.** Why prefer risk differences over odds ratios or hazard ratios? The risk difference directly tells you: "Treatment changes the probability of the outcome by X percentage points." It is a **marginal causal parameter** that does not depend on which covariates you condition on.

Odds ratios, by contrast, are non-collapsible (they change when you marginalize over covariates even without confounding), lack a straightforward population-level causal interpretation, and are frequently misinterpreted as risk ratios. Hazard ratios share similar non-collapsibility issues and depend on the proportional hazards assumption.

Additionally, the risk difference is **less sensitive to background disease prevalence** than ratio measures: the same odds ratio can correspond to very different absolute risk changes depending on how common the outcome is, making ratio measures harder to interpret for clinical decision-making. When your goal is to inform a public health decision, the risk difference answers that question directly.

## Step 4: Link the Causal Estimand to the Observed Data (Identification)

This is the **identification step**: *Can we estimate the causal effect from the data we have?*

**Identification** is the logical argument that, under stated assumptions, an unobservable causal quantity equals something we *can* compute from observed data. The three key assumptions are fundamentally untestable and must be justified by subject-matter knowledge.

:::{.callout-note}
## Assumption 1: Consistency ("You Get What You Get")

The observed outcome equals the counterfactual outcome under the treatment actually received:

$$
Y = Y(A)
$$

In practice, this means the treatment is **well-defined** and there is only one version of it. If someone took denosumab, their observed outcome equals their potential outcome under denosumab. Consistency can break down when "treatment" is vague (e.g., "exercise more") or when the drug formulation, dose, or route varies across patients in ways that matter for the outcome. The target trial protocol (Step 1) helps protect consistency by requiring well-defined treatment strategies.
:::


:::{.callout-warning}
## Assumption 2: Exchangeability, the Hardest Assumption to Defend

$$
Y(a) \perp A \mid W
$$

This is the assumption that **there are no unmeasured confounders**. After adjusting for the measured covariates $W$, the treated and untreated groups are comparable, as if treatment had been randomly assigned within strata of $W$.

- In a randomized trial, randomization guarantees exchangeability by design.
- In observational data, we must **assume** it holds, and this assumption is **untestable** from the data alone.

This is typically the most difficult assumption to defend. If there are unmeasured factors (e.g., frailty, physician preference, disease severity scores not in the data) that influence both treatment and outcome, identification fails. Always use a DAG (Step 2) to reason about which variables are sufficient to block all backdoor paths, and conduct sensitivity analyses (e.g., E-values; see [Chapter 3.12](03-12-sensitivity-analysis.qmd)) to assess robustness to potential unmeasured confounding.
:::


:::{.callout-warning}
## Assumption 3: Positivity (Watch for Practical Violations)

Everyone has a *positive probability* of receiving either treatment at each level of covariates:

$$
0 < P(A=a \mid W=w) < 1
$$

**Structural positivity violations**: Some covariate strata *never* receive a particular treatment by clinical convention. For example, patients over age 90 may never be prescribed aggressive chemotherapy, or patients with severe renal impairment may be ineligible for certain drugs. No statistical method can fix this. The causal effect is simply not defined in those strata.

**Practical positivity violations** are extremely common in real-world data: some strata have very few patients receiving one treatment, leading to near-zero propensity scores and unstable estimates. For example, if patients with very low CD4 counts are almost always started on HAART, the propensity score for those patients approaches 1, and the IPTW weight for untreated patients in that stratum becomes very large. Signs to look for: propensity scores piling up near 0 or 1, very large IPTW weights, and g-computation predictions extrapolating far outside the support of the data.
:::


:::{.callout-tip}
## The Key Insight: Replacing Counterfactuals with Observables

Under the three assumptions, the unobservable causal estimand equals a function of the observed data:

$$
E[Y(1) - Y(0)]
=
E_W\!\left[
E[Y \mid A=1, W]
-
E[Y \mid A=0, W]
\right]
$$

In plain terms, this says we can estimate the causal effect by (1) modeling the outcome within levels of covariates and treatment, then (2) averaging those predictions over the covariate distribution in the study population. This is the **g-computation identification formula**, the foundation for g-computation, IPTW, AIPW, and TMLE.
:::

:::{.callout-note}
## Causal Estimand vs. Statistical Estimand

A **causal estimand** is defined via potential outcomes (e.g., 3-year MI risk difference if all patients received denosumab vs. ZA). A **statistical estimand** is the function of the observed-data distribution that equals the causal estimand *under the identification assumptions* (e.g., the standardized mean difference from a model of $P(Y \mid A, W)$).

**Pitfall.** Regression coefficients (e.g., log-odds ratios, hazard ratios) are **not** the statistical estimand for the ATE. They are conditional measures on the wrong scale.
:::

### Diagnosing positivity violations

The propensity score overlap plot is your primary diagnostic. Look for mass piling up near 0 or 1 and poor overlap between groups.

```{r}
source("_data/load-haart-cohort.R")

ps <- predict(glm(ever_haart ~ age + sex + cd4_sqrt_baseline,
                   family = binomial, data = haart_baseline),
              type = "response")
plot(density(ps), main = "Propensity Score Distribution",
     xlab = "Estimated P(HAART = 1 | W)")
```

When you see poor overlap, consider restricting to the region of common support, trimming extreme weights, or acknowledging the limitations. See [Chapter 3.2](03-02-positivity-overlap.qmd) for a detailed treatment.


### Step 5: Choose a Statistical Estimator

This step translates the identified quantity into an algorithm we can apply to data. You *could* use a standard regression model here (e.g., fit a logistic regression and standardize predictions via g-computation), but modern methods offer important advantages in robustness and flexibility. Options covered in upcoming chapters include:

- **G-computation** ([Chapter 2.1](02-01-gcomputation.qmd)): Predict outcomes under each treatment, average over W
- **IPTW** ([Chapter 2.2](02-02-iptw.qmd)): Weight observations by inverse probability of treatment
- **TMLE** ([Chapter 2.3](02-03-doubly-robust-tmle.qmd)): Targeted learning that combines outcome and treatment models for double robustness
- **Longitudinal TMLE (L-TMLE)** ([Chapter 3.9](03-09-illustrated-ltmle.qmd)): For time-varying treatments, time-varying confounding, and survival outcomes, estimating causal survival curves without proportional hazards assumptions

The choice depends on the assumptions you're willing to make, the complexity of the data structure (point treatment vs. longitudinal), sample size, and tolerance for model misspecification.

### Step 6: Evaluate and Interpret

Once you've estimated your parameter, interpretation is **not automatic**.

Ask yourself: Are your assumptions plausible? What does your estimate mean for real-world decisions? Can you generalize to other populations? No amount of statistical sophistication can rescue an analysis built on implausible assumptions. Use sensitivity analyses, negative control outcomes, covariate balance checks, and expert consultation to stress-test your conclusions.

We cover diagnostics in detail in [Chapter 3.5: Advanced Diagnostics](03-05-advanced-diagnostics.qmd) (covariate balance, weight distributions, influence curves) and sensitivity analyses in [Chapter 3.12: Sensitivity Analysis](03-12-sensitivity-analysis.qmd) (E-values, quantitative bias analysis, and bounding approaches for unmeasured confounding).

:::{.callout}
## Summary

The Causal Roadmap gives a rigorous framework for designing and evaluating real-world effect estimates. Before choosing a statistical method, start with the science: define your causal question, specify your model and assumptions, decide how to express the effect, and choose a transparent, defensible estimation method.
:::

## Running Example: Antiretroviral Therapy Cohort

Throughout this handbook, we use a shared observational dataset to illustrate each method: the longitudinal HIV antiretroviral therapy cohort example used in the marginal structural model literature (Van der Wal & Geskus, 2011). The cohort follows 1,200 HIV-positive patients from seroconversion and asks whether initiating highly active antiretroviral therapy (HAART) reduces mortality. See the [Running Case Study](00-case-study.qmd) for the full data description, variable dictionary, and patient trajectories.

### Applying the roadmap to the HAART question

Here we walk through each roadmap step for the HAART case study. See the [Running Case Study](00-case-study.qmd) for the full target trial protocol, DAGs, and data description.

**Step 1: Target trial protocol.** The target trial compares initiating HAART when clinically indicated versus deferring HAART, among HIV-positive adults from seroconversion, with all-cause mortality as the outcome and the risk difference as the causal contrast. The full protocol table is in the [case study chapter](00-case-study.qmd#the-target-trial-we-wish-we-could-run).

**Step 2: Causal model and DAG.** The key variables are baseline covariates W (age, sex), time-varying CD4 count, treatment A (HAART initiation), and outcome Y (death). The longitudinal structure creates treatment-confounder feedback: CD4 is affected by prior HAART and also affects future HAART decisions.

Because conditioning on time-varying CD4 in a regression blocks part of the causal pathway while not conditioning on it leaves confounding unaddressed, neither standard regression approach is correct. This is the classic motivation for g-methods (g-computation, IPTW, TMLE), which we develop in Part II. See the [case study DAGs](00-case-study.qmd#the-observed-data-and-causal-structure) for the full point-treatment and longitudinal DAGs.

**Step 3: Estimand.** The marginal risk difference in mortality under the two treatment strategies.

**Step 4: Identification.** Under consistency (HAART is well-defined), sequential exchangeability (no unmeasured time-varying confounders beyond CD4, age, and sex), and positivity (at each time point, patients with any covariate profile have nonzero probability of either treatment decision), the causal effect is identified by the longitudinal g-computation formula.

**Step 5: Estimation.** Point-treatment versions of this question are addressed with g-computation (Chapter 2.1), IPTW (Chapter 2.2), and TMLE (Chapter 2.3). The full longitudinal question requires IPTW with marginal structural models (Chapter 3.7) or longitudinal TMLE (Chapter 3.9).

**Step 6: Interpretation.** Diagnostics (Chapter 3.5), sensitivity analysis for unmeasured confounding (Chapter 3.12), and consideration of informative censoring are all critical before interpreting results from this cohort.

## Check your understanding

<details>
<summary><strong>Q1. Why must the causal question be defined before choosing a statistical method?</strong></summary>

The causal question determines everything downstream: which variables are confounders (and therefore must be adjusted for), what the target estimand is (e.g., ATE on the risk difference scale vs. a conditional odds ratio), and which assumptions are required for identification. If you choose a statistical method first, say, logistic regression, you have implicitly committed to a conditional odds ratio as your estimand, which may not match the scientific question you actually care about. Starting with the question also prevents "researcher degrees of freedom": when the estimand is pre-specified, there is less temptation to explore multiple model specifications and selectively report favorable results.

</details>

<details>
<summary><strong>Q2. What is the difference between a causal estimand and a statistical estimand?</strong></summary>

A **causal estimand** is defined in terms of potential outcomes, for example, $E[Y(1) - Y(0)]$, the average treatment effect. It describes a contrast between hypothetical worlds and cannot be directly computed from observed data without assumptions. A **statistical estimand** is a function of the observed data distribution, for example, $E_W[E[Y \mid A=1, W] - E[Y \mid A=0, W]]$. The identification step of the Causal Roadmap shows that, *under specific assumptions* (exchangeability, positivity, consistency), the causal estimand equals a particular statistical estimand. Without those assumptions, the statistical quantity you compute may have no causal interpretation at all.

</details>

<details>
<summary><strong>Q3. Can covariate balance checks prove that exchangeability holds?</strong></summary>

No. Covariate balance checks (e.g., standardized mean differences after weighting) can only assess whether **measured** covariates are balanced between treatment groups. Exchangeability requires that there are **no unmeasured common causes** of treatment and outcome, a condition that depends on subject-matter knowledge and cannot be verified from the data. Good balance on observed covariates is necessary but not sufficient: an unmeasured confounder can induce bias even when every measured covariate is perfectly balanced. Balance checks are a useful diagnostic, but they should never be interpreted as evidence that the exchangeability assumption holds.

</details>

<details>
<summary><strong>Q4. A researcher defines "statin users" as anyone who filled a prescription at any time during a 5-year window, and compares their outcomes to never-users. What bias is most likely?</strong></summary>

This design introduces **immortal time bias**. The violated component is **time zero**. Statin users are classified based on a future event (filling a prescription at some point during follow-up), and their follow-up is counted from before treatment began. The period between the start of observation and the prescription fill date is "immortal" for the treated group, because dying before the fill date would have moved them to the untreated group. This creates a spurious survival advantage. The fix is to align time zero with the date of statin initiation and only compare to controls still at risk at that same time.

</details>

<details>
<summary><strong>Q5. What is a positivity violation, and what should you do when you encounter one?</strong></summary>

A positivity violation occurs when certain covariate strata have zero (or near-zero) probability of receiving one of the treatment levels. **Structural violations** arise when some patients are never treated by clinical convention (e.g., very elderly patients never receiving aggressive chemotherapy). The causal effect is simply not defined in those strata, and no statistical method can fix this. **Practical violations** occur when some strata have very few patients on one treatment, leading to propensity scores near 0 or 1 and unstable estimates. When you encounter positivity violations, you should: (1) inspect propensity score distributions for mass near the boundaries, (2) consider restricting the target population to the region of overlap, (3) use weight trimming or stabilized weights if using IPTW, and (4) be transparent about the limitations and the population to which your estimate applies.

</details>

<details>
<summary><strong>Q6. How does the target trial protocol connect to the identification assumptions?</strong></summary>

The target trial protocol makes identification assumptions concrete. The **assignment procedure** component supports **exchangeability**: in the target trial, randomization guarantees it; in the emulation, we must argue that conditional on measured covariates, treatment is independent of potential outcomes. The **treatment strategies** component supports **consistency**: well-defined strategies ensure "treatment" means the same thing for everyone, so $Y(a)$ is well-defined. The **eligibility criteria** help support **positivity**: defining the population carefully ensures that within every stratum, there are both treated and untreated individuals.

</details>

## What comes next

With the causal question defined, the roadmap laid out, and the identification framework in hand, we are ready to move from design to estimation. Part II introduces the statistical methods that deliver on the promise of this framework, beginning with g-computation ([Chapter 2.1](02-01-gcomputation.qmd)), inverse probability weighting ([Chapter 2.2](02-02-iptw.qmd)), and the doubly robust TMLE estimator ([Chapter 2.3](02-03-doubly-robust-tmle.qmd)).

## Sources and further reading

- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- Dang LE, Tager I, Petersen ML (2023). A causal roadmap for generating high-quality real-world evidence. *Journal of Clinical and Translational Science* 7(1):e127.
- Hernan MA, Robins JM (2016). Using big data to emulate a target trial when a randomized trial is not available. *Am J Epidemiol* 183(8):758-764.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Suissa S (2008). Immortal time bias in pharmacoepidemiology. *Am J Epidemiol* 167(4):492-499.
- Lund JL, Richardson DB, Sturmer T (2015). The active comparator, new user study design in pharmacoepidemiology. *Epidemiology* 26(1):105-113.
- Pearl J (2009). *Causality: Models, Reasoning, and Inference*. 2nd ed. Cambridge University Press.
- Robins JM (1986). A new approach to causal inference in mortality studies with sustained exposure periods. *Mathematical Modelling* 7:1393-1512.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.
- Gruber S, van der Laan MJ (2010). A targeted maximum likelihood estimator of a causal effect on a bounded continuous outcome. *Int J Biostat* 6(1):Article 26.
- ICH E9 (R1) Addendum (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- Kennedy EH (2019). Nonparametric causal effects based on incremental propensity score interventions. *JASA* 114(526):645-656.
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)
- `lmtp` R package: [CRAN](https://cran.r-project.org/package=lmtp)
