---
title: "Chapter 1.2: The Causal Roadmap"
format: html
---

In this chapter, we introduce the **Causal Roadmap**, a structured approach for asking and answering causal questions with observational data. The roadmap helps researchers define their scientific question, translate it into a formal causal model, assess whether the causal effect is identifiable from data, and choose appropriate estimators to answer that question.

::: {.callout-note}
## Learning objectives
- Describe the six steps of the Causal Roadmap and their purpose
- Formulate a causal question in terms of population, intervention, comparator, and outcome (PICO)
- Draw and interpret a directed acyclic graph (DAG) encoding causal assumptions
- Distinguish a causal estimand from a statistical estimand
- Explain why pre-specifying an analysis plan before data access reduces researcher degrees of freedom
:::

*Sources and scope. This chapter is educational. Causal conclusions depend on identification assumptions (e.g., consistency, exchangeability, positivity) and on diagnostic evidence that the data support the target estimand. When flexible machine learning is used for nuisance estimation, valid inference typically requires cross-fitting or a cross-validated TMLE variant, plus appropriate rate conditions.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

### Why Use a Causal Roadmap?

In epidemiology, it is tempting to jump straight to fitting a model---"Let’s run a regression and see what we find." But this approach conflates the scientific question with the statistical procedure, making it difficult to evaluate whether the analysis actually answers the question you care about. Modern causal inference begins by asking: **What is the causal question we are trying to answer?**

**Key point.** A causal roadmap ensures transparency in how evidence is generated, separation of scientific questions from statistical methods, clear articulation of assumptions, and structured thinking in both trial and non-randomized settings. Starting with a well-defined question---before touching data or code---is the single most important step in any causal analysis.

We’ll walk step-by-step through each part of the roadmap, with illustrative examples.

:::{.callout-note}
## Step 1: Define the Causal Question

This includes:
- **Population**: Who are we studying?
- **Intervention**: What exposure or treatment are we manipulating?
- **Outcome**: What effect are we interested in?
- **Timeframe**: Over what duration?

Example:

> What is the 3-year risk of cardiovascular events if all postmenopausal women with osteoporosis initiated **denosumab** compared to if all initiated **zoledronic acid**?

This PICO-style framing is the foundation of causal inference.
:::

We now draw out how we believe the data were generated. This includes the treatment (A), outcome (Y), and covariates (W), encoded as a directed acyclic graph (DAG) with potential outcomes $Y(1)$ and $Y(0)$ for each individual.

A Structural Causal Model (SCM) is simply a formal way of writing down your story about how the world works. Think of it as a set of equations that say: "This variable is caused by these other variables, plus some factors we cannot measure." For MPH students, the SCM is the mathematical version of the causal DAG you draw on a whiteboard---each arrow in the DAG corresponds to one of these equations. The key advantage is that it forces you to be explicit about every assumption you are making about what causes what.

:::{.callout-note}
## Step 2: Specify the Causal Model

The SCM for our example:

$$
W = f_W(U_W),\
A = f_A(W, U_A),\
Y = f_Y(W, A, U_Y)
$$

Where $U$ terms represent unmeasured variables. The corresponding DAG:

```
W → A → Y
 \    ↘
  →────→
```

This DAG implies W (confounders) affect both A and Y, A (treatment) affects Y, and there is no unmeasured confounding.
:::

With the causal model in hand, we can now define exactly what quantity we want to estimate.

:::{.callout-note}
## Step 3: Define the Target Causal Estimand

We often want the **Average Treatment Effect (ATE)**:

$$
ATE = E[Y(1) - Y(0)]
$$

Or equivalently, the **Risk Difference**: $E[Y \mid do(A=1)] - E[Y \mid do(A=0)]$. Other common estimands include risk ratios, hazard ratios (with caveats), median survival differences, and conditional effects in subgroups (CATEs).
:::

**Key point.** Why prefer risk differences over odds ratios? The risk difference directly tells you: "Treatment changes the probability of the outcome by X percentage points." Odds ratios, by contrast, are non-collapsible (they change when you marginalize over covariates even without confounding), lack a straightforward population-level causal interpretation, and are frequently misinterpreted as risk ratios. When your goal is to inform a public health decision, the risk difference answers that question directly.

:::{.callout-warning}
## Step 4: Link the Causal Estimand to the Observed Data

This is the **identification step**: *Can we estimate the causal effect from the data we have?* These assumptions are fundamentally untestable---they must be justified by subject-matter knowledge.

- **Exchangeability**: $Y(a) \perp A \mid W$ --- you have measured every common cause of treatment and outcome.
- **Positivity**: $P(A = a \mid W = w) > 0$ --- no covariate stratum deterministically receives one treatment.
- **Consistency**: The observed outcome equals the potential outcome under the received treatment.

If satisfied: $E[Y(1) - Y(0)] = E_W[ E[Y \mid A = 1, W] - E[Y \mid A = 0, W] ]$.
:::

### Step 5: Choose a Statistical Estimator

This step translates the identified quantity into an algorithm we can apply to data. Options include:

- **G-computation**: Predict outcomes under each treatment, average over W
- **IPTW**: Weight observations by inverse probability of treatment
- **TMLE**: Targeted learning that combines outcome and treatment models

We’ll explore these in later chapters. The choice depends on the assumptions you’re willing to make, sample size, and tolerance for model misspecification.

### Step 6: Evaluate and Interpret

Once you've estimated your parameter, interpretation is **not automatic**. Ask yourself: Are your assumptions plausible? What does your estimate mean for real-world decisions? Can you generalize to other populations?

**Pitfall.** No amount of statistical sophistication can rescue an analysis built on implausible assumptions. Use sensitivity analyses, negative control outcomes, covariate balance checks, and expert consultation to stress-test your conclusions.

## A Simple Simulated Example

We'll simulate a small dataset to show how steps 1-5 look in practice.

```{r}
set.seed(42)
n <- 1000
age <- rnorm(n, 75, 6)
cvd_history <- rbinom(n, 1, plogis(0.1 * (age - 70)))
W <- data.frame(age, cvd_history)
A <- rbinom(n, 1, plogis(-1 + 0.05 * (age - 70) + 1.5 * cvd_history))
Y <- rbinom(n, 1, plogis(-2 + 0.4 * A + 0.1 * (age - 70) + 1 * cvd_history))
data <- data.frame(W, A, Y)
```

Estimate ATE using g-computation:

```{r}
model <- glm(Y ~ A + age + cvd_history, family = binomial, data = data)
newdata1 <- transform(data, A = 1)
newdata0 <- transform(data, A = 0)
p1 <- predict(model, newdata = newdata1, type = "response")
p0 <- predict(model, newdata = newdata0, type = "response")
mean(p1 - p0)
```

:::{.callout-tip}
## Interpreting the G-Computation Result

The number produced by `mean(p1 - p0)` is the estimated **average causal risk difference**. In plain language: it tells you how much the probability of the outcome (e.g., a cardiovascular event) would change, on average, if everyone in the study population had received treatment (A = 1) compared to if everyone had received no treatment (A = 0). For example, a value of 0.07 means that treatment is estimated to increase the risk of the outcome by 7 percentage points on average across the population, after adjusting for age and CVD history. This is the core quantity that g-computation targets---a population-level "what if" comparison.
:::

---

:::{.callout-tip}
## Summary

The Causal Roadmap gives a rigorous framework for designing and evaluating real-world effect estimates. Before choosing a statistical method, start with the science: define your causal question, specify your model and assumptions, decide how to express the effect, and choose a transparent, defensible method.
:::

## Check your understanding

<details>
<summary><strong>Q1. Why must the causal question be defined before choosing a statistical method?</strong></summary>

The causal question determines everything downstream: which variables are confounders (and therefore must be adjusted for), what the target estimand is (e.g., ATE on the risk difference scale vs. a conditional odds ratio), and which assumptions are required for identification. If you choose a statistical method first---say, logistic regression---you have implicitly committed to a conditional odds ratio as your estimand, which may not match the scientific question you actually care about. Starting with the question also prevents "researcher degrees of freedom": when the estimand is pre-specified, there is less temptation to explore multiple model specifications and selectively report favorable results.

</details>

<details>
<summary><strong>Q2. What is the difference between a causal estimand and a statistical estimand?</strong></summary>

A **causal estimand** is defined in terms of potential outcomes---for example, $E[Y(1) - Y(0)]$, the average treatment effect. It describes a contrast between hypothetical worlds and cannot be directly computed from observed data without assumptions. A **statistical estimand** is a function of the observed data distribution---for example, $E_W[E[Y \mid A=1, W] - E[Y \mid A=0, W]]$. The identification step of the Causal Roadmap shows that, *under specific assumptions* (exchangeability, positivity, consistency), the causal estimand equals a particular statistical estimand. Without those assumptions, the statistical quantity you compute may have no causal interpretation at all.

</details>

<details>
<summary><strong>Q3. Can covariate balance checks prove that exchangeability holds?</strong></summary>

No. Covariate balance checks (e.g., standardized mean differences after weighting) can only assess whether **measured** covariates are balanced between treatment groups. Exchangeability requires that there are **no unmeasured common causes** of treatment and outcome---a condition that depends on subject-matter knowledge and cannot be verified from the data. Good balance on observed covariates is necessary but not sufficient: an unmeasured confounder can induce bias even when every measured covariate is perfectly balanced. Balance checks are a useful diagnostic, but they should never be interpreted as evidence that the exchangeability assumption holds.

</details>

---

## Sources and further reading

- Dang LE, Tager I, Petersen ML (2023). A causal roadmap for generating high-quality real-world evidence. *Journal of Clinical and Translational Science* 7(1):e127.
- Hernan MA, Robins JM (2016). Using big data to emulate a target trial when a randomized trial is not available. *Am J Epidemiol* 183(8):758-764.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Gruber S, van der Laan MJ (2010). A targeted maximum likelihood estimator of a causal effect on a bounded continuous outcome. *Int J Biostat* 6(1):Article 26.
- ICH E9 (R1) Addendum (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)

---

## Software Implementation (R)

This example maps the six-step Causal Roadmap to a single `tmle` call. Each comment below labels the roadmap step that the corresponding code addresses.

- **Step 1 (Question):** Average treatment effect of a binary exposure on a continuous outcome
- **Step 2 (Causal model):** Observed data $(W, A, Y)$; $W$ confounds $A \to Y$
- **Step 3 (Assumptions):** Consistency, exchangeability given $W$, positivity
- **Step 4 (Estimand):** $\psi = E[E(Y \mid A=1, W)] - E[E(Y \mid A=0, W)]$
- **Steps 5–6 (Estimation + Inference):** TMLE with Super Learner for $Q$ and $g$

```{r}
#| eval: false
set.seed(1)
n <- 500

## ── Step 2: Simulate from a causal model ──
W1 <- rnorm(n)
W2 <- rbinom(n, 1, 0.4)
A  <- rbinom(n, 1, plogis(-0.5 + 0.6 * W1 + 0.3 * W2))
Y  <- A + 0.8 * W1 - 0.5 * W2 + rnorm(n)  # true ATE = 1.0

## ── Steps 5–6: Estimation and inference via TMLE ──
if (requireNamespace("tmle", quietly = TRUE)) {
  library(tmle)
  tmle_fit <- tmle(
    Y = Y, A = A,
    W = data.frame(W1 = W1, W2 = W2),
    Q.SL.library = c("SL.glm", "SL.mean"),
    g.SL.library  = c("SL.glm", "SL.mean")
  )
  cat("Roadmap Step 5 – TMLE ATE:", round(tmle_fit$estimates$ATE$psi, 3), "\n")
  cat("Roadmap Step 6 – 95% CI: ", round(tmle_fit$estimates$ATE$CI, 3), "\n")
} else {
  message("Install the ‘tmle’ package:  install.packages(‘tmle’)")
}
```
