---
title: 'Chapter 1.1: Why We Need a Causal Roadmap'
format: html
---

*Estimated reading time: 20 minutes*

Most observational studies start with a causal question: does treatment A reduce the risk of outcome Y compared to treatment B? Researchers typically reach for a familiar tool: fit a Cox proportional hazards model, adjust for available confounders, and interpret the hazard ratio as the answer. This workflow is so common that it can feel like the *only* workflow.

But there are two distinct problems with this approach. First, the regression model may be **misspecified** (the wrong functional form, missing interactions, or omitted confounders can bias the coefficient). Second, and more fundamentally, even a perfectly specified model may answer the **wrong question**. A conditional hazard ratio from a Cox model is not the same quantity as "the causal effect of treatment on 3-year risk." The hazard ratio is a conditional, instantaneous rate measure that does not directly translate to a population-level risk difference or survival probability contrast. The causal question you want to answer may simply not correspond to any single regression coefficient.

Recognizing this gap is not a criticism of regression itself; regression remains a useful building block. The gap is a reason to adopt a more deliberate analytic workflow, a **Causal Roadmap**, that starts with a clearly defined causal question, specifies a target estimand on a meaningful scale (such as the risk difference or restricted mean survival time difference), and chooses an estimation strategy that targets it directly. Later chapters introduce **longitudinal TMLE (L-TMLE)**, which can estimate causal survival curve contrasts without relying on the proportional hazards assumption.

This chapter uses a concrete clinical example to illustrate where the standard regression-only approach falls short, introduces the counterfactual framework that makes causal questions precise, and demonstrates g-computation as a first step toward targeted estimation. Later chapters build on these ideas with inverse probability weighting (IPTW), targeted minimum loss-based estimation (TMLE), and the full Causal Roadmap. The chapter proceeds as follows:

1. A motivating example comparing osteoporosis treatments
2. What adjusted regression does and does not give you
3. The counterfactual (potential outcomes) framework
4. G-computation as a bridge from association to causation
5. Summary and practice questions

::: {.callout-note}
## Learning objectives
- Understand why regression coefficients do not, in general, equal causal effects
- Recognize confounding, collider bias, and table-2 fallacy in routine analyses
- Articulate the counterfactual (potential outcomes) framework for defining causal contrasts
- Distinguish conditional associations from marginal causal effects
- Motivate the need for the Causal Roadmap and modern estimation methods (g-computation, IPTW, TMLE)
:::

*Causal conclusions require identification assumptions (consistency, exchangeability, positivity) that no statistical method can verify from data alone.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

## Why Not Just Use Regression?

Multivariable regression has long been the standard tool in observational research for "adjusting" for confounders. But regression coefficients don't always represent causal effects. In fact, they almost never do without additional assumptions and deliberate analytic choices that go beyond what standard regression provides.

**Key point.** If you've been trained to interpret an adjusted regression coefficient as "the effect of X on Y, holding all else equal," you're in good company, but that interpretation quietly smuggles in causal language that the model itself cannot justify. The coefficient tells you about the **conditional association** within strata defined by the other variables in the model. Whether that association equals a causal effect depends on assumptions that regression alone cannot verify.

### Motivating Example: Comparing Osteoporosis Treatments

Suppose we have a large claims database and want to know whether patients who initiate **denosumab** (Prolia) have a higher 3-year risk of myocardial infarction (MI) compared to those initiating **zoledronic acid**. Both are common osteoporosis treatments, and the causal question is straightforward: if we could assign everyone to denosumab versus zoledronic acid, would the 3-year MI risk differ?

The standard approach is to fit a **Cox proportional hazards model**, adjust for confounders like age and cardiovascular disease history, and report the adjusted hazard ratio. Let's see what that looks like.


```{r, include=FALSE}
library(broom)
library(knitr)
library(survival)

# Simulate time-to-event data
set.seed(123)
n <- 2000
treatment <- rbinom(n, 1, 0.5) # 1 = denosumab, 0 = zoledronic acid
age <- rnorm(n, 75, 6)
cvd_history <- rbinom(n, 1, plogis(0.1 * (age - 70)))

# Simulate event times from an exponential model
baseline_hazard <- 0.05
log_hr <- 0.4 * treatment + 0.05 * (age - 70) + 1 * cvd_history
event_time <- rexp(n, rate = baseline_hazard * exp(log_hr))

# Administrative censoring at 3 years
censor_time <- 3
time <- pmin(event_time, censor_time)
event <- as.integer(event_time <= censor_time)
```

We adjust for age and cardiovascular disease history using a Cox model:

```{r}
cox_model <- coxph(Surv(time, event) ~ treatment + age + cvd_history)
tidy(cox_model, conf.int = TRUE, exponentiate = TRUE) |>
  kable(digits = 3, caption = "Adjusted Cox proportional hazards model (hazard ratios)")
```

:::{.callout-warning}
## The Hazard Ratio Is NOT the Causal Effect You Want

The adjusted hazard ratio for `treatment` is around 1.5. But what does a clinician or policymaker actually need to know? They need to know: **if we switched the entire eligible population from zoledronic acid to denosumab, how many additional MIs per 100 patients would we expect over 3 years?** That quantity, the **marginal risk difference** at 3 years, is the natural scale for comparing treatments and making decisions.

The Cox hazard ratio does not answer that question. It is a **conditional, instantaneous rate ratio** that:

- Assumes **proportional hazards** (that the treatment effect is constant over time), which is often violated in practice.
- Is **non-collapsible**: the conditional HR and marginal HR differ even without confounding, so the coefficient does not have a straightforward population-level interpretation.
- Does not directly translate to a risk difference or survival probability contrast without additional computation.

In later chapters, we introduce **longitudinal TMLE (L-TMLE)**, which directly estimates causal survival curves and risk differences at specific time points without relying on the proportional hazards assumption.
:::


## Enter the Counterfactual Framework

:::{.callout-note}
## Defining Causal Effects with Potential Outcomes

For each person we imagine two hypothetical worlds: **Y(1)** (outcome under denosumab) and **Y(0)** (outcome under zoledronic acid). The **average treatment effect (ATE)** is:

$$
ATE = E[Y(1) - Y(0)]
$$
:::

The fundamental problem of causal inference is that each person only receives one treatment. We observe Y(1) or Y(0), never both. Under three key identification assumptions, **exchangeability** (no unmeasured confounding), **consistency** (the observed outcome under the treatment actually received equals the potential outcome), and **positivity** (every individual has a nonzero probability of receiving either treatment given their covariates), we can identify the causal ATE from observed data using the **g-computation formula**:

$$
ATE = E_W\big[ E[Y \mid A = 1, W] - E[Y \mid A = 0, W] \big]
$$

where W represents the measured confounders and A is the treatment. This expression says: within each confounder stratum, compute the difference in expected outcomes between treated and untreated, then average over the distribution of confounders in the population.


## When Regression Fails

:::{.callout-warning}
## Model Misspecification Can Silently Bias Your Results

Misspecification bias is **silent**. A regression with the wrong functional form still produces coefficients and p-values with no warning. This motivates semiparametric estimators (like TMLE with machine learning) that relax parametric assumptions.
:::

Let's compare the Cox approach to g-computation, which directly targets the 3-year risk difference:

```{r}
# Fit a logistic model for 3-year risk (binary: event by year 3)
outcome_3yr <- as.integer(event_time <= 3)
gcomp_model <- glm(outcome_3yr ~ treatment + age + cvd_history, family = binomial)

# Predict counterfactual 3-year risks
newdata1 <- data.frame(treatment = 1, age = age, cvd_history = cvd_history)
p1 <- predict(gcomp_model, newdata = newdata1, type = "response")

newdata0 <- data.frame(treatment = 0, age = age, cvd_history = cvd_history)
p0 <- predict(gcomp_model, newdata = newdata0, type = "response")

# Marginal risk difference at 3 years
cat("G-computation risk difference:", round(mean(p1 - p0), 4), "\n")
```

:::{.callout-tip}
## Why G-Computation Gives You a Population-Level Causal Effect

G-computation **predicts** counterfactual outcomes for every person under treatment and under control, then **averages** the difference. The result is a **marginal risk difference**, a population-level quantity directly relevant to decision-making, unlike the conditional hazard ratio from the Cox model.
:::

Compare this to the Cox hazard ratio:

```{r}
# The Cox HR (exponentiating the coefficient)
exp(coef(cox_model)["treatment"])
```

The hazard ratio tells you about instantaneous rate differences, but the g-computation risk difference tells you: "X more people per 100 would have an MI within 3 years under denosumab compared to zoledronic acid." The latter is what patients and clinicians need to make decisions.

## Running Example: Antiretroviral Therapy Cohort

Let us connect the ideas in this chapter to the longitudinal HIV antiretroviral therapy cohort example used in the marginal structural model literature (see [Running Case Study](00-case-study.qmd) for the full introduction). The causal question is: **does initiating HAART reduce mortality among HIV-positive patients?**

A researcher might try a Cox model, adjusting for baseline CD4 count, age, and sex:

```{r}
source("_data/load-haart-cohort.R")

# Naive baseline-only Cox model
cox_haart <- coxph(
  Surv(total_fup_days, died) ~ ever_haart + cd4_sqrt_baseline + age + sex,
  data = haart_baseline
)
tidy(cox_haart, conf.int = TRUE, exponentiate = TRUE) |>
  kable(digits = 3, caption = "Naive Cox model: HAART and mortality (baseline adjustment only)")
```

The hazard ratio for `ever_haart` from this model suffers from at least two problems:

1. **Confounding by indication.** Patients who started HAART had lower CD4 counts (sicker), so the coefficient conflates the treatment benefit with the fact that treated patients were at higher baseline risk.
2. **The estimand mismatch.** Even if confounding were resolved, the conditional hazard ratio is not a marginal risk difference. It does not tell us how many deaths per 100 patients would be prevented by HAART.

The situation gets even worse when we consider time-varying confounding (CD4 counts change over time and are affected by prior treatment), which no standard regression can handle correctly. We return to this example in later chapters with g-computation, IPTW, and TMLE.

## Summary

:::{.callout-tip}
## Key Takeaways

- **Regression is not inherently causal.** It estimates conditional associations under strong model assumptions. An adjusted coefficient is not the same as a causal effect.
- **Causal inference starts with a question, not a model.** Define your causal question and target estimand (e.g., the ATE on the risk difference scale) *before* choosing an analytic method.
- **The counterfactual framework clarifies what we want to estimate.** Y(1), Y(0), and the ATE give us precise, model-free definitions of causal effects.
- **G-computation, IPTW, and TMLE estimate marginal causal effects.** These methods target the population-level quantity defined by the g-computation formula, rather than relying on regression coefficients that conflate the estimand with the model.
- **The hazard ratio from Cox regression is not the ATE.** Hazard ratios are conditional, instantaneous rate measures that do not directly translate to risk differences. This is perhaps the single most important distinction for epidemiologists moving from traditional regression to causal inference methods.
:::

## What comes next

Now that you understand why regression coefficients are not causal effects, the next chapter introduces the **Causal Roadmap**: a structured workflow that turns a scientific question into a defensible analysis plan, from question definition through estimation and interpretation.

## Check your understanding

<details>
<summary><strong>Q1. Why can't you interpret an adjusted logistic regression coefficient as the ATE?</strong></summary>

The adjusted coefficient from logistic regression is a **conditional log-odds ratio**. It describes the association between treatment and outcome within strata defined by the covariates in the model. This differs from the ATE (a marginal risk difference) in at least two ways. First, the odds ratio is **non-collapsible**: even without confounding, a conditional odds ratio and a marginal odds ratio will differ, so you cannot read the coefficient as a population-level summary. Second, the coefficient lives on the **log-odds scale**, not the risk-difference scale, and converting between them requires averaging over the covariate distribution (which is exactly what g-computation does). Finally, the coefficient inherits every **parametric assumption** in the model (log-linearity, no interactions, correct link function); if any assumption fails, the coefficient is biased for any causal quantity.

</details>

<details>
<summary><strong>Q2. What are the three identification assumptions needed to give the g-computation formula a causal interpretation?</strong></summary>

1. **Exchangeability (no unmeasured confounding):** Within levels of the measured covariates W, treatment assignment is independent of the potential outcomes, $Y(a) \perp A \mid W$. This means every common cause of treatment and outcome has been measured and adjusted for.
2. **Consistency:** The observed outcome for a person who received treatment $a$ equals their potential outcome $Y(a)$. This requires a well-defined intervention with no hidden variations.
3. **Positivity:** Every covariate stratum has a nonzero probability of receiving each treatment level, $P(A = a \mid W = w) > 0$ for all relevant $a$ and $w$. Without positivity, there are subgroups for which no data-based comparison is possible.

If any of these assumptions fails, the g-computation estimator may be biased regardless of how well the outcome model fits the data.

</details>

<details>
<summary><strong>Q3. What is the difference between a conditional log-odds ratio and a marginal risk difference?</strong></summary>

A **conditional log-odds ratio** is the change in the log-odds of the outcome per unit change in treatment, holding all other covariates fixed at specific values. It is a stratum-specific measure on the odds scale. A **marginal risk difference** is the difference in the probability of the outcome if everyone in the population were treated versus if everyone were untreated. It is a population-averaged measure on the probability scale. In nonlinear models (such as logistic regression), these two quantities answer different questions and are generally not numerically equal, even when the model is correctly specified. The marginal risk difference is usually more directly relevant for public health decision-making because it tells you how many additional cases per 100 people would result from a policy change.

</details>

## Sources and further reading

- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Greenland S, Pearl J, Robins JM (1999). Causal diagrams for epidemiologic research. *Epidemiology* 10(1):37-48.
- Pearl J (2009). *Causality: Models, Reasoning, and Inference*. 2nd ed. Cambridge University Press.
- Westreich D, Greenland S (2013). The table 2 fallacy: presenting and interpreting confounder and modifier coefficients. *Am J Epidemiol* 177(4):292-298.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)
- `SuperLearner` R package: [CRAN](https://cran.r-project.org/package=SuperLearner)
