---
title: "Chapter 1.2: The Causal Roadmap"
format: html
---

**NOTE: cite Petersen and van der Lann "Causal models and learning from data: integrating causal modeling and statistical estimation" 2014 here**


In this chapter, we introduce the **Causal Roadmap**, a structured approach for asking and answering causal questions with observational data. We begin with the **target trial protocol**, which forces you to define the scientific question precisely before choosing any statistical method, and then walk through the **six steps of the causal roadmap** that translate that question into a defensible analysis.

::: {.callout-note}
## Learning objectives
- Describe the six steps of the Causal Roadmap and their purpose
- Formulate a causal question in terms of population, intervention, comparator, and outcome (PICO)
- Draw and interpret a directed acyclic graph (DAG) encoding causal assumptions
- Distinguish a causal estimand from a statistical estimand
- Explain why pre-specifying an analysis plan before data access reduces researcher degrees of freedom
:::


**NOTE: this next sentence seems out of place and not yet introduced**


*Causal conclusions require identification assumptions and diagnostic evidence that the data support the target estimand.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
library(dagitty)
library(ggdag)
library(ggplot2)
```

### Why Use a Causal Roadmap?

In epidemiology, it is tempting to jump straight to fitting a model: "Let's run a regression and see what we find." But this approach conflates the scientific question with the statistical procedure, making it difficult to evaluate whether the analysis actually answers the question you care about. Modern causal inference begins by asking: **What is the causal question we are trying to answer?**

**Key point.** A causal roadmap ensures transparency in how evidence is generated, separation of scientific questions from statistical methods, clear articulation of assumptions, and structured thinking in both trial and non-randomized settings. Starting with a well-defined question, before touching data or code, is the single most important step in any causal analysis.

We'll walk step-by-step through each part of the roadmap, with illustrative examples.


## Step 1: Define the Causal Question 

**NOTE: before introducing the target trial protocol, just briefly motivate why we need to be clear about the causal question rather than jumping into the traditional analysis that fits the form of the outcome (like logistic or cox regression) **


### Start with the target trial protocol, then follow the roadmap

As advocated by Dang and Balzer (2023), a useful way to begin the roadmap is to first sketch the **target trial protocol**. This does not mean every observational study literally recreates a randomized trial. It means we use the trial protocol as a design device to make the causal question precise before we choose an estimator.

In practice, this means stating the trial you would ideally like to run:

- **Eligibility criteria**: Who would be included?
- **Time zero**: When would follow-up begin?
- **Treatment strategies**: What interventions are being compared?
- **Assignment procedure**: Would treatment be randomized, or are we approximating that contrast with observed data?
- **Outcome**: What endpoint would be measured?
- **Follow-up**: Over what time horizon?
- **Causal contrast**: Risk difference, risk ratio, restricted mean survival time difference, or another estimand?
- **Analysis plan**: How will confounding, censoring, and missingness be handled?

This target trial step helps prevent a very common problem in real-world evidence: starting with whatever variables happen to be available, fitting a model, and only afterward deciding what question was answered. Instead, the target trial protocol clarifies the question first. The **causal roadmap** then helps us move from that ideal question to a defensible analysis by making the causal model, identifiability assumptions, statistical estimand, estimator choice, diagnostics, and interpretation explicit.

A helpful way to think about the relationship is:

> **Target trial emulation clarifies the causal question. The causal roadmap tells us how to analyze that question rigorously with the data we actually have.**

So in this handbook, Step 1 is not just "state the question." It is better read as:

> **State the causal question by writing down the target trial you wish you could run.**

**NOTE: below is repetative of what's above. Rewrite for concision and to avoid repitition **


A strong causal analysis starts by specifying the trial you would ideally like to run, even if you only have observational data.

This includes:

- **Population**: Who are we studying?
- **Eligibility criteria**: Who would qualify for the study?
- **Time zero**: When does follow-up begin?
- **Intervention and comparator**: What treatment strategies are being compared?
- **Outcome**: What endpoint are we interested in?
- **Follow-up**: Over what duration?
- **Causal contrast**: What summary of effect do we want?

Example:

> Among postmenopausal women with osteoporosis who are eligible to initiate either therapy, what is the 3-year risk of cardiovascular events if all initiated **denosumab** compared with if all initiated **zoledronic acid**, starting at treatment initiation and followed for 3 years?

Writing the target trial protocol first forces you to define the scientific question before choosing a statistical model.

## Step 2: Specify the Causal Model

We now draw out how we believe the data were generated. This includes the treatment (A), outcome (Y), and covariates (W), encoded as a directed acyclic graph (DAG) with potential outcomes $Y(1)$ and $Y(0)$ for each individual.

A Structural Causal Model (SCM) is simply a formal way of writing down your story about how the world works. Think of it as a set of equations that say: "This variable is caused by these other variables, plus some factors we cannot measure." For MPH students, the SCM is the mathematical version of the causal DAG you draw on a whiteboard. Each arrow in the DAG corresponds to one of these equations. The key advantage is that it forces you to be explicit about every assumption you are making about what causes what.


The SCM for our example:

$$
W = f_W(U_W),\
A = f_A(W, U_A),\
Y = f_Y(W, A, U_Y)
$$

Where $U$ terms represent unmeasured variables. The corresponding DAG:

```{r}
#| label: fig-roadmap-dag
#| fig-cap: "Basic causal DAG: W confounds the A → Y relationship"
#| fig-width: 5
#| fig-height: 3
dag_roadmap <- dagify(
  A ~ W, Y ~ A + W,
  exposure = "A", outcome = "Y",
  coords = list(
    x = c(W = 0, A = 1, Y = 2),
    y = c(W = 1, A = 0, Y = 0)
  )
)
ggdag(dag_roadmap) + theme_dag()
```

This DAG implies W (confounders) affect both A and Y, A (treatment) affects Y, and there is no unmeasured confounding.


With the causal model in hand, we can now define exactly what quantity we want to estimate.

## Step 3: Define the Target Causal Estimand


**NOTE: instead of saying we often want, say we might want the ATE, but then discuss that we may want more complex contrasts like CATES or stochastic or dynamic or optimal intervention regimes **

**NOTE: add a placeholder QMD for optimal intervention regimes to be filled in later using the tlverse package **

**NOTE: add a placeholder QMD for outcome-blind simulation studies**


We often want the **Average Treatment Effect (ATE)**:

$$
ATE = E[Y(1) - Y(0)]
$$

Or equivalently, the **Risk Difference**: $E[Y \mid do(A=1)] - E[Y \mid do(A=0)]$. Other common estimands include risk ratios, hazard ratios (with caveats), median survival differences, and conditional effects in subgroups (CATEs).

**NOTE: add to the point below that the risk difference is more causal and less dependent on background disease prevalence**

**Key point.** Why prefer risk differences over odds ratios? The risk difference directly tells you: "Treatment changes the probability of the outcome by X percentage points." Odds ratios, by contrast, are non-collapsible (they change when you marginalize over covariates even without confounding), lack a straightforward population-level causal interpretation, and are frequently misinterpreted as risk ratios. When your goal is to inform a public health decision, the risk difference answers that question directly.

## Step 4: Link the Causal Estimand to the Observed Data

**NOTE: note that we'll go into much more detail on this in the next chapter*


This is the **identification step**: *Can we estimate the causal effect from the data we have?* These assumptions are fundamentally untestable and must be justified by subject-matter knowledge.

- **Exchangeability**: $Y(a) \perp A \mid W$. You have measured every common cause of treatment and outcome.
- **Positivity**: $P(A = a \mid W = w) > 0$. No covariate stratum deterministically receives one treatment.
- **Consistency**: The observed outcome equals the potential outcome under the received treatment.

If satisfied: $E[Y(1) - Y(0)] = E_W[ E[Y \mid A = 1, W] - E[Y \mid A = 0, W] ]$.


**NOTE: add to step 5 that we'll be covering the following options in upcoming chapters, but we could use regressions. Also add L-TMLE for longitudinal data**


### Step 5: Choose a Statistical Estimator

This step translates the identified quantity into an algorithm we can apply to data. Options include:

- **G-computation**: Predict outcomes under each treatment, average over W
- **IPTW**: Weight observations by inverse probability of treatment
- **TMLE**: Targeted learning that combines outcome and treatment models

We'll explore these in later chapters. The choice depends on the assumptions you're willing to make, sample size, and tolerance for model misspecification.

**NOTE: link below on future chapters on diagnostics and sensitivity analyses**

### Step 6: Evaluate and Interpret

Once you've estimated your parameter, interpretation is **not automatic**. Ask yourself: Are your assumptions plausible? What does your estimate mean for real-world decisions? Can you generalize to other populations? Note that mo amount of statistical sophistication can rescue an analysis built on implausible assumptions. Use sensitivity analyses, negative control outcomes, covariate balance checks, and expert consultation to stress-test your conclusions.

:::{.callout}
## Summary

The Causal Roadmap gives a rigorous framework for designing and evaluating real-world effect estimates. Before choosing a statistical method, start with the science: define your causal question, specify your model and assumptions, decide how to express the effect, and choose a transparent, defensible estimation method.
:::

## Check your understanding

<details>
<summary><strong>Q1. Why must the causal question be defined before choosing a statistical method?</strong></summary>

The causal question determines everything downstream: which variables are confounders (and therefore must be adjusted for), what the target estimand is (e.g., ATE on the risk difference scale vs. a conditional odds ratio), and which assumptions are required for identification. If you choose a statistical method first, say, logistic regression, you have implicitly committed to a conditional odds ratio as your estimand, which may not match the scientific question you actually care about. Starting with the question also prevents "researcher degrees of freedom": when the estimand is pre-specified, there is less temptation to explore multiple model specifications and selectively report favorable results.

</details>

<details>
<summary><strong>Q2. What is the difference between a causal estimand and a statistical estimand?</strong></summary>

A **causal estimand** is defined in terms of potential outcomes, for example, $E[Y(1) - Y(0)]$, the average treatment effect. It describes a contrast between hypothetical worlds and cannot be directly computed from observed data without assumptions. A **statistical estimand** is a function of the observed data distribution, for example, $E_W[E[Y \mid A=1, W] - E[Y \mid A=0, W]]$. The identification step of the Causal Roadmap shows that, *under specific assumptions* (exchangeability, positivity, consistency), the causal estimand equals a particular statistical estimand. Without those assumptions, the statistical quantity you compute may have no causal interpretation at all.

</details>

<details>
<summary><strong>Q3. Can covariate balance checks prove that exchangeability holds?</strong></summary>

No. Covariate balance checks (e.g., standardized mean differences after weighting) can only assess whether **measured** covariates are balanced between treatment groups. Exchangeability requires that there are **no unmeasured common causes** of treatment and outcome, a condition that depends on subject-matter knowledge and cannot be verified from the data. Good balance on observed covariates is necessary but not sufficient: an unmeasured confounder can induce bias even when every measured covariate is perfectly balanced. Balance checks are a useful diagnostic, but they should never be interpreted as evidence that the exchangeability assumption holds.

</details>

## What comes next

With the causal question defined and the roadmap in hand, the next chapter formalizes the identification step: the assumptions needed to bridge potential outcomes and observed data, and the statistical estimands that result.

## Sources and further reading

- Dang LE, Tager I, Petersen ML (2023). A causal roadmap for generating high-quality real-world evidence. *Journal of Clinical and Translational Science* 7(1):e127.
- Hernan MA, Robins JM (2016). Using big data to emulate a target trial when a randomized trial is not available. *Am J Epidemiol* 183(8):758-764.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Gruber S, van der Laan MJ (2010). A targeted maximum likelihood estimator of a causal effect on a bounded continuous outcome. *Int J Biostat* 6(1):Article 26.
- ICH E9 (R1) Addendum (2019). Addendum on estimands and sensitivity analyses in clinical trials.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)
