---
title: "Chapter 1.4: Target Trial Emulation for Real-World Evidence"
format: html
---

In this chapter, we formalize **target trial emulation**, the practice of designing every observational analysis as if it were emulating a specific, well-defined randomized trial. We will see how this discipline prevents some of the most common (and most damaging) biases in real-world evidence, and how it connects to the Causal Roadmap introduced in Chapter 1.2 and the identification framework in Chapter 1.3.

::: {.callout-note}
## Learning objectives
- Explain why every observational causal analysis should begin with a target trial protocol
- List and define the eight components of a target trial protocol
- Recognize immortal time bias, prevalent user bias, and vague comparator bias as failures of target trial thinking
- Simulate and quantify immortal time bias in R
- Construct a new-user cohort with aligned time zero from registry-style data
- Connect target trial protocol components to steps of the Causal Roadmap
:::

*Designing observational analyses as if they were trials.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
```

## A motivating disaster: immortal time bias in statin studies

In 2006, several observational studies reported that statin users had substantially lower mortality than non-users, not just from cardiovascular disease, but from cancer, trauma, and even infections. The effect sizes were implausibly large. Were statins a miracle drug?

No. The problem was **immortal time bias**, a direct consequence of failing to align time zero with treatment assignment. Here is what happened:

1. Researchers identified patients who filled a statin prescription at some point during follow-up.
2. They classified the entire follow-up period for these patients as "statin exposed," including the time *before* the prescription was filled.
3. During that early period, the patient could not have died (otherwise they would never have filled the prescription). That time was, by definition, "immortal."
4. Non-users had no such guaranteed survival period.

The result was a spurious survival advantage for statin users that had nothing to do with the drug. This bias can be enormous, sometimes accounting for the entire observed effect.

:::{.callout-important}
## The core lesson
Immortal time bias arises when time zero (the start of follow-up) is not aligned with the moment of treatment assignment. A target trial protocol forces you to define time zero explicitly, preventing this error before any data are analyzed.
:::

This chapter teaches you to avoid this class of mistakes by starting every analysis with a target trial protocol. We will walk through each protocol component, show how common biases map to protocol violations, simulate immortal time bias so you can see it yourself, and then work through a complete example.

## 1. Why Observational Analyses Should Start by Emulating a Target Trial

### The gap between what we want and what we have

When we ask a causal question ("Does initiating statins reduce the 5-year risk of major adverse cardiovascular events?"), the gold standard answer would come from a **randomized controlled trial** (RCT). The trial protocol would specify who is eligible, when follow-up begins, what interventions are compared, and how outcomes are measured. Randomization would handle confounding. The protocol would be written *before* enrollment begins.

With observational data, we have none of these guarantees. Treatment is not randomized. Eligibility is defined post hoc. Time zero is ambiguous. The comparison group may be poorly defined. Without discipline, the analyst can inadvertently answer a question that differs from the one they intended, or, worse, answer a question that has no coherent causal interpretation at all.

### Target trial emulation as a bridge

**Target trial emulation** (Hernan and Robins, 2016) provides that discipline. The idea is straightforward:

1. **Specify the protocol** for the ideal randomized trial you would like to run.
2. **Emulate each protocol component** using the observational data you actually have.
3. **Document each deviation** between the ideal trial and the emulation, and assess how each deviation could introduce bias.

This is not a statistical method. It is a design framework, a checklist that forces you to make explicit decisions that might otherwise remain implicit (and wrong). As we discussed in Chapter 1.2, the target trial protocol is the starting point for Step 1 of the Causal Roadmap: defining the causal question.

:::{.callout-tip}
## Think of it this way
A target trial protocol is to an observational study what an architectural blueprint is to a building. You would never start construction without a blueprint. You should never start an observational causal analysis without a target trial protocol.
:::

## 2. The Target Trial Protocol Components

A complete target trial protocol has eight components. Each one must be specified before any data manipulation or modeling begins.

### 2.1 Eligibility criteria

**What it means:** Who would be enrolled in the trial?

Eligibility criteria define the population to which causal conclusions will apply. In observational emulations, this translates directly to cohort inclusion and exclusion criteria applied at time zero.

**Example:** Adults aged 40 to 75 with newly diagnosed hyperlipidemia, no prior cardiovascular event, no prior statin use, and no contraindication to statins.

### 2.2 Treatment strategies

**What it means:** What interventions are being compared?

You must define at least two well-specified treatment strategies. Vague strategies ("statin use" vs. "no statin use") lead to vague causal questions. Precise strategies specify the drug, dose, timing, and adherence expectations.

**Example:** (A) Initiate atorvastatin within 30 days of diagnosis and remain on therapy for at least 2 years. (B) Do not initiate any statin within 30 days of diagnosis.

### 2.3 Assignment procedure

**What it means:** How would participants be assigned to treatment strategies?

In the target trial, this would be randomization. In the emulation, we observe natural treatment decisions and use statistical methods to adjust for confounding. This component reminds us that we need the exchangeability assumption (Chapter 1.3) and must identify all important confounders.

### 2.4 Time zero

**What it means:** When does follow-up begin?

This is the single most important component for preventing bias. Time zero must simultaneously satisfy three conditions:

1. The participant meets the **eligibility criteria**
2. The participant is **assigned** to (or begins) a treatment strategy
3. **Follow-up** for the outcome begins

When these three events do not coincide, biases arise. Immortal time bias, as we saw above, is a direct consequence of time zero misalignment.

:::{.callout-warning}
## Time zero alignment is non-negotiable
If eligibility, treatment assignment, and the start of follow-up do not all occur at the same moment, your analysis will be biased. There are no statistical fixes for a misaligned time zero. The only fix is to redesign the study.
:::

### 2.5 Outcome

**What it means:** What endpoint is being measured?

The outcome must be clearly defined and measured the same way across treatment groups. In observational settings, you must consider whether outcome ascertainment differs by treatment group (for example, patients on statins may have more frequent lab tests, leading to earlier detection of some outcomes).

### 2.6 Follow-up period

**What it means:** How long are participants followed?

The follow-up period begins at time zero and ends at the outcome, death, loss to follow-up, or the administrative end of the study, whichever comes first. Specifying this in advance prevents data-driven choices about follow-up length.

### 2.7 Causal contrast (estimand)

**What it means:** What summary measure compares the treatment strategies?

This is the causal estimand from Chapter 1.3. Common choices include the intention-to-treat effect, the per-protocol effect, the average treatment effect (ATE), or the restricted mean survival time difference. The choice should be driven by the scientific question, not by what is convenient to estimate.

### 2.8 Analysis plan

**What it means:** How will confounding, censoring, and model misspecification be addressed?

The analysis plan specifies the statistical approach: g-computation, inverse probability weighting, TMLE, or another estimator. It also specifies how censoring and loss to follow-up will be handled. This is where the later chapters of this handbook (Chapters 2.1 through 2.5) become relevant.

### Summary table: statin initiation example

| Protocol component | Target trial specification |
|---|---|
| **Eligibility** | Adults 40-75 with new hyperlipidemia diagnosis, no prior CVD event, no prior statin use |
| **Treatment strategies** | (A) Initiate atorvastatin within 30 days; (B) No statin initiation within 30 days |
| **Assignment** | Random assignment at time of eligibility |
| **Time zero** | Date eligibility criteria are first met |
| **Outcome** | First major adverse cardiovascular event (MACE): MI, stroke, or CV death |
| **Follow-up** | From time zero until MACE, death, loss to follow-up, or 5 years |
| **Causal contrast** | Intention-to-treat risk difference at 5 years |
| **Analysis plan** | Clone-censor-weight approach for ITT; TMLE with Super Learner for point-treatment emulation |

## 3. Common Biases Prevented by Target Trial Thinking

Many well-known biases in observational research can be traced to violations of specific target trial protocol components. Thinking in terms of the protocol makes these biases easier to diagnose and prevent.

### 3.1 Immortal time bias

**Which component is violated?** Time zero.

As described in the opening example, immortal time bias occurs when follow-up begins before treatment assignment. The period between the start of follow-up and the treatment event is "immortal" for the treated group (they had to survive to receive treatment). This artificially inflates the survival of the treated group.

**Prevention:** Align time zero with the moment of treatment assignment. Use a new-user (incident-user) design.

### 3.2 Prevalent user bias

**Which component is violated?** Eligibility criteria and time zero.

Prevalent user bias occurs when you include patients who are already on treatment at the start of follow-up. These patients are a selected subset: they have already survived the early period of treatment (when side effects and early failures are most common), and their current covariate values may be consequences of prior treatment rather than confounders.

**Prevention:** Restrict to new (incident) users who initiate treatment at time zero. This is sometimes called the "new-user design" or "active-comparator, new-user design."

### 3.3 Vague comparator definitions

**Which component is violated?** Treatment strategies.

If the comparator group is "everyone not on the drug," it includes a mix of people on alternative treatments, people with contraindications, people who will start the drug later, and people who have never been considered for treatment. This creates a poorly defined causal contrast.

**Prevention:** Define specific, well-characterized treatment strategies. An active comparator (for example, "initiate pravastatin" vs. "initiate atorvastatin") often leads to a more meaningful and less confounded comparison than a "treated vs. untreated" contrast.

### 3.4 Post-hoc causal questions

**Which component is violated?** The entire protocol.

When analysts explore the data first and formulate the causal question afterward, the question is shaped by what the data show rather than by what the science demands. This dramatically inflates the risk of false positive findings.

**Prevention:** Write the target trial protocol before accessing the data, or at minimum, before running any analyses.

:::{.callout-note}
## A unifying insight
Nearly every major bias in observational pharmacoepidemiology (immortal time bias, prevalent user bias, selection bias from conditioning on the future, time-window bias) is a violation of one or more target trial protocol components. If you can specify and emulate the protocol correctly, you prevent these biases by design rather than by statistical adjustment.
:::

### Simulation: immortal time bias in action

The following simulation demonstrates how failing to align time zero with treatment assignment creates spurious treatment effects. We generate data where the treatment has **no causal effect** on the outcome, then show that a naive analysis (which misaligns time zero) produces a large apparent benefit.

```{r}
set.seed(2024)
n <- 5000

# --- Data-generating process (NO treatment effect) ---
# Each person has a baseline hazard for the outcome (death)
# Treatment is initiated at a random time during follow-up (for those who initiate)
baseline_hazard <- 0.10  # annual probability of event
max_follow_up <- 5       # years

sim_data <- tibble(
  id = 1:n,
  # Some patients are "destined" to initiate treatment (60%)
  will_treat = rbinom(n, 1, 0.6),
  # Time to treatment initiation (uniform 0-3 years for those who initiate)
  time_to_treat = ifelse(will_treat == 1, runif(n, 0, 3), Inf),
  # Event time: exponential with rate = baseline_hazard (NO treatment effect)
  event_time = rexp(n, rate = baseline_hazard),
  # Observed event time (censored at max follow-up)
  obs_time = pmin(event_time, max_follow_up),
  event = as.integer(event_time <= max_follow_up),
  # A person actually initiates treatment only if they survive to initiation
  actually_treated = as.integer(will_treat == 1 & time_to_treat < obs_time)
)

# --- WRONG analysis: ignore treatment timing (immortal time bias) ---
# Classify entire follow-up as treated/untreated based on whether they
# ever received treatment during follow-up
wrong_analysis <- sim_data |>
  group_by(actually_treated) |>
  summarise(
    n = n(),
    events = sum(event),
    person_years = sum(obs_time),
    rate = events / person_years,
    .groups = "drop"
  )

cat("=== WRONG analysis (immortal time bias) ===\n")
cat("True treatment effect: NONE (null by construction)\n\n")
print(as.data.frame(wrong_analysis))

rate_ratio_wrong <- wrong_analysis$rate[wrong_analysis$actually_treated == 1] /
  wrong_analysis$rate[wrong_analysis$actually_treated == 0]
cat("\nBiased rate ratio (treated vs untreated):", round(rate_ratio_wrong, 2), "\n")
cat("Expected if no bias: 1.00\n")
```

The biased rate ratio is well below 1.0, suggesting that treatment is protective, even though we constructed the data so that treatment has **zero** causal effect. The entire "benefit" comes from the fact that treated patients had to survive long enough to receive treatment.

Now let us fix the analysis by aligning time zero with treatment initiation, using a **new-user design with time-matched controls**.

```{r}
# --- CORRECT analysis: align time zero with treatment initiation ---
# For each treated person, time zero = time of treatment initiation
# Match to an untreated person still at risk at that same time

treated <- sim_data |>
  filter(actually_treated == 1) |>
  mutate(
    t0 = time_to_treat,
    remaining_time = obs_time - t0,
    event_after_t0 = as.integer(event_time > t0 & event_time <= max_follow_up)
  )

# For controls: sample from those still at risk at each treatment time
set.seed(123)
matched_controls <- treated |>
  rowwise() |>
  mutate(
    # Find a control still alive and untreated at this time
    control_id = {
      eligible <- sim_data |>
        filter(actually_treated == 0, obs_time > t0)
      if (nrow(eligible) > 0) sample(eligible$id, 1) else NA_integer_
    }
  ) |>
  ungroup() |>
  filter(!is.na(control_id))

# Get control outcomes from t0 onward
controls_outcome <- matched_controls |>
  left_join(
    sim_data |> select(id, event_time_ctrl = event_time, obs_time_ctrl = obs_time),
    by = c("control_id" = "id")
  ) |>
  mutate(
    remaining_time_ctrl = obs_time_ctrl - t0,
    event_after_t0_ctrl = as.integer(
      event_time_ctrl > t0 & event_time_ctrl <= max_follow_up
    )
  )

rate_treated <- sum(controls_outcome$event_after_t0) /
  sum(controls_outcome$remaining_time)
rate_control <- sum(controls_outcome$event_after_t0_ctrl) /
  sum(controls_outcome$remaining_time_ctrl)

cat("=== CORRECT analysis (time zero aligned) ===\n")
cat("Rate in treated (from t0):", round(rate_treated, 3), "\n")
cat("Rate in controls (from t0):", round(rate_control, 3), "\n")
cat("Corrected rate ratio:", round(rate_treated / rate_control, 2), "\n")
cat("Expected if no bias: ~1.00\n")
```

:::{.callout-tip}
## What the simulation shows
When time zero is misaligned, a null treatment looks highly protective (rate ratio well below 1). When time zero is properly aligned with treatment initiation, the rate ratio is close to 1.0, as it should be. No fancy statistical method can fix a fundamentally flawed study design.
:::

## 4. Connection to the Causal Roadmap

The target trial protocol and the Causal Roadmap (Chapter 1.2) are complementary frameworks. The target trial protocol makes the causal question precise. The roadmap then provides the systematic workflow for moving from that question to a defensible estimate. The following table shows how each roadmap step maps to target trial components.

| Causal Roadmap step | Target trial component(s) | What it accomplishes |
|---|---|---|
| **Step 1:** Define the causal question | Eligibility, treatment strategies, outcome, causal contrast | Specifies the population, interventions, and estimand |
| **Step 2:** Specify the causal model | Assignment procedure, eligibility | Encodes assumptions about confounding structure (DAG) |
| **Step 3:** State identification assumptions | Assignment procedure, time zero | Justifies exchangeability, positivity, consistency |
| **Step 4:** Define the statistical estimand | Causal contrast, analysis plan | Translates the causal estimand into a function of observed data |
| **Step 5:** Choose an estimator | Analysis plan | Selects g-computation, IPTW, TMLE, or another method |
| **Step 6:** Perform sensitivity analysis | Assignment procedure, analysis plan | Assesses robustness to violations of assumptions |

:::{.callout-note}
## How the pieces fit together
The target trial protocol (this chapter) gives you Step 1. Identification (Chapter 1.3) gives you Steps 2 through 4. The estimation chapters (Chapters 2.1 through 2.5) give you Step 5. Together, they form a complete workflow for drawing causal conclusions from observational data.
:::

Notice how time zero alignment maps to the identification assumptions. If time zero is misaligned, the consistency assumption is violated: the potential outcome $Y(a)$ is not well-defined when treatment assignment and follow-up start at different times. Similarly, if prevalent users are included, the exchangeability assumption is threatened because current covariates may be affected by prior treatment.

## 5. Worked Example: Statin Initiation and MACE

We now walk through a complete target trial emulation using simulated registry-style data. The scientific question is: **Does initiating a statin (vs. not initiating) reduce 3-year MACE risk among adults with new hyperlipidemia?**

### Step 1: Define the target trial protocol

| Component | Specification |
|---|---|
| **Eligibility** | Age 40-75, new hyperlipidemia diagnosis, no prior CVD, no prior statin use |
| **Treatment strategies** | (A) Initiate statin within 30 days; (B) No statin initiation within 30 days |
| **Assignment** | Randomized at eligibility date |
| **Time zero** | Date eligibility criteria first met |
| **Outcome** | First MACE (MI, stroke, or CV death) |
| **Follow-up** | 3 years from time zero |
| **Causal contrast** | ATE on the risk difference scale |
| **Analysis plan** | G-computation with logistic regression; sensitivity analysis with IPTW |

### Step 2: Simulate registry data and construct the new-user cohort

```{r}
set.seed(42)
n <- 3000

# --- Simulate baseline covariates at time of eligibility ---
cohort <- tibble(
  id = 1:n,
  age = round(runif(n, 40, 75)),
  male = rbinom(n, 1, 0.52),
  diabetes = rbinom(n, 1, 0.25),
  smoking = rbinom(n, 1, 0.20),
  ldl = round(rnorm(n, mean = 160, sd = 30)),   # mg/dL at diagnosis
  sbp = round(rnorm(n, mean = 135, sd = 18))     # systolic BP at diagnosis
)

# --- Treatment assignment (confounded by covariates) ---
# Sicker patients more likely to receive statins
cohort <- cohort |>
  mutate(
    lp_treat = -1.5 + 0.02 * (age - 55) + 0.3 * male + 0.5 * diabetes +
      0.3 * smoking + 0.01 * (ldl - 160) + 0.01 * (sbp - 135),
    pr_treat = plogis(lp_treat),
    A = rbinom(n, 1, pr_treat)
  )

# --- Outcome: 3-year MACE risk ---
# True causal effect: statin reduces MACE risk (risk difference ~ -0.05)
cohort <- cohort |>
  mutate(
    lp_outcome = -2.5 - 0.4 * A + 0.03 * (age - 55) + 0.2 * male +
      0.6 * diabetes + 0.4 * smoking + 0.005 * (ldl - 160) +
      0.01 * (sbp - 135),
    pr_mace = plogis(lp_outcome),
    Y = rbinom(n, 1, pr_mace)
  )

# Verify confounding: crude vs. true effect
crude_rd <- mean(cohort$Y[cohort$A == 1]) - mean(cohort$Y[cohort$A == 0])
cat("Crude risk difference:", round(crude_rd, 4), "\n")
cat("(Confounded because sicker patients are more likely to get statins)\n")
```

### Step 3: Time-zero alignment

In this simulated dataset, time zero is the date of eligibility (new hyperlipidemia diagnosis). Treatment assignment (A) is determined at time zero. Follow-up begins at time zero. All three conditions for proper time-zero alignment are satisfied by construction.

In real data, you would need to verify this carefully:

```{r}
# Verify time-zero alignment
# In a real analysis, you would check:
cohort_check <- cohort |>
  summarise(
    n_total = n(),
    n_treated = sum(A),
    n_control = sum(1 - A),
    prop_treated = mean(A),
    outcome_rate_treated = mean(Y[A == 1]),
    outcome_rate_control = mean(Y[A == 0])
  )

cat("=== New-user cohort summary ===\n")
cat("Total eligible patients:", cohort_check$n_total, "\n")
cat("Statin initiators:", cohort_check$n_treated,
    sprintf("(%.1f%%)", 100 * cohort_check$prop_treated), "\n")
cat("Non-initiators:", cohort_check$n_control, "\n")
cat("3-year MACE rate (treated):",
    sprintf("%.1f%%", 100 * cohort_check$outcome_rate_treated), "\n")
cat("3-year MACE rate (untreated):",
    sprintf("%.1f%%", 100 * cohort_check$outcome_rate_control), "\n")
```

### Step 4: Estimation via g-computation

Following Chapter 2.1, we use g-computation to estimate the average treatment effect while adjusting for confounders.

```{r}
# --- G-computation ---
# Step 1: Fit outcome model
outcome_model <- glm(
  Y ~ A + age + male + diabetes + smoking + ldl + sbp,
  data = cohort,
  family = binomial
)

# Step 2: Predict potential outcomes under each treatment strategy
cohort_gcomp <- cohort |>
  mutate(
    # Predicted probability if everyone received statins
    Y1_hat = predict(outcome_model,
                     newdata = mutate(cohort, A = 1),
                     type = "response"),
    # Predicted probability if nobody received statins
    Y0_hat = predict(outcome_model,
                     newdata = mutate(cohort, A = 0),
                     type = "response")
  )

# Step 3: Compute the ATE (marginal risk difference)
ate_gcomp <- mean(cohort_gcomp$Y1_hat) - mean(cohort_gcomp$Y0_hat)
risk_treated <- mean(cohort_gcomp$Y1_hat)
risk_untreated <- mean(cohort_gcomp$Y0_hat)

cat("=== G-computation results ===\n")
cat("Estimated 3-year MACE risk under statin initiation:",
    sprintf("%.1f%%", 100 * risk_treated), "\n")
cat("Estimated 3-year MACE risk under no initiation:   ",
    sprintf("%.1f%%", 100 * risk_untreated), "\n")
cat("Estimated ATE (risk difference):",
    sprintf("%.4f", ate_gcomp),
    sprintf("(%.1f percentage points)", 100 * ate_gcomp), "\n")
cat("\nInterpretation: Initiating statins is estimated to",
    ifelse(ate_gcomp < 0, "reduce", "increase"),
    "3-year MACE risk by",
    sprintf("%.1f", abs(100 * ate_gcomp)),
    "percentage points.\n")
```

### Step 5: Bootstrap confidence interval

```{r}
# --- Bootstrap inference ---
set.seed(99)
n_boot <- 500
boot_ates <- numeric(n_boot)

for (b in seq_len(n_boot)) {
  boot_idx <- sample(nrow(cohort), replace = TRUE)
  boot_data <- cohort[boot_idx, ]

  boot_model <- glm(
    Y ~ A + age + male + diabetes + smoking + ldl + sbp,
    data = boot_data,
    family = binomial
  )

  p1 <- predict(boot_model, newdata = mutate(boot_data, A = 1), type = "response")
  p0 <- predict(boot_model, newdata = mutate(boot_data, A = 0), type = "response")
  boot_ates[b] <- mean(p1) - mean(p0)
}

ci <- quantile(boot_ates, probs = c(0.025, 0.975))

cat("=== Target trial emulation: final results ===\n")
cat("Estimand: 3-year ATE (risk difference), intention-to-treat\n")
cat("Method: G-computation with logistic regression\n")
cat("Point estimate:", sprintf("%.4f", ate_gcomp), "\n")
cat("95% bootstrap CI: [", sprintf("%.4f", ci[1]), ",",
    sprintf("%.4f", ci[2]), "]\n")
```

### Step 6: Interpret in the target trial framework

:::{.callout-tip}
## Interpreting the results
The estimated risk difference should be reported in the language of the target trial. For example: "In a hypothetical trial where adults aged 40 to 75 with new hyperlipidemia were randomized to initiate atorvastatin vs. no statin, the estimated 3-year MACE risk difference is [point estimate] (95% CI: [lower, upper])."

This framing makes the causal claim explicit, reminds the reader of the assumptions required (exchangeability given measured covariates, positivity, consistency), and connects the statistical result back to the scientific question.
:::

The validity of this analysis rests on the target trial protocol being correctly specified and emulated:

- **Time zero** is aligned: eligibility, treatment assignment, and follow-up all start at diagnosis.
- **New-user design** prevents prevalent user bias: no one is already on a statin at time zero.
- **Treatment strategies** are specific: initiate vs. do not initiate within 30 days.
- **Confounding adjustment** relies on the assumption that age, sex, diabetes, smoking, LDL, and SBP are sufficient to achieve conditional exchangeability (Chapter 1.3). In a real study, you would use a DAG to justify this assumption and perform sensitivity analyses for unmeasured confounding.

## Summary

:::{.callout-note}
## Key takeaways

1. **Start with the protocol, not the data.** Every observational causal analysis should begin by specifying the target trial you wish you could run. The eight protocol components (eligibility, treatment strategies, assignment, time zero, outcome, follow-up, causal contrast, and analysis plan) make the causal question unambiguous.

2. **Time zero alignment is the most common source of serious bias.** When eligibility, treatment assignment, and follow-up start do not coincide, biases such as immortal time bias are introduced. These biases can be large enough to fully explain an observed treatment effect, and they cannot be corrected with statistical adjustment.

3. **New-user designs prevent prevalent user bias.** Restricting to patients who are initiating (not continuing) treatment ensures that baseline covariates are not affected by prior treatment and that early treatment effects are captured.

4. **The target trial protocol is Step 1 of the Causal Roadmap.** It defines the causal question. Identification assumptions (Chapter 1.3), estimation methods (Chapters 2.1 through 2.5), and diagnostics follow from there.

5. **Document every deviation.** When the observational emulation differs from the ideal trial (as it always will), document each difference and assess how it might introduce bias. Transparency about limitations is more valuable than false precision.
:::

## Check your understanding

<details>
<summary><strong>Q1. A researcher defines "statin users" as anyone who filled a statin prescription at any time during a 5-year observation window, and compares their outcomes to people who never filled a prescription. What bias is most likely, and which target trial component was violated?</strong></summary>

This design introduces **immortal time bias**. The violated component is **time zero**. Statin users are classified based on a future event (filling a prescription at some point during follow-up), and their follow-up is counted from a time before treatment actually began. The period between the start of observation and the prescription fill date is "immortal" for the treated group, because dying before the fill date would have moved them to the untreated group. This creates a spurious survival advantage for treated patients. The fix is to align time zero with the date of statin initiation and only compare to controls who are still at risk at that same time.

</details>

<details>
<summary><strong>Q2. Why is a "new-user" (incident-user) design preferred over including patients already on treatment? Explain in terms of target trial components.</strong></summary>

Including patients already on treatment (prevalent users) violates the **eligibility criteria** and **time zero** components of the target trial. Prevalent users are a selected group: they have already survived the initiation period, tolerated the drug, and chosen to continue it. Their current covariate values (such as LDL cholesterol or blood pressure) may be consequences of prior treatment rather than pre-treatment confounders. Adjusting for covariates affected by treatment introduces collider bias (or "overadjustment"). A new-user design avoids these problems by starting follow-up at the moment of treatment initiation, just as a trial would enroll patients at the point of randomization. This ensures that baseline covariates are measured before treatment begins and that early events are captured.

</details>

<details>
<summary><strong>Q3. How does the target trial protocol connect to the identification assumptions discussed in Chapter 1.3? Give one example of a protocol component that directly supports one identification assumption.</strong></summary>

The target trial protocol makes identification assumptions concrete and verifiable. For example, the **assignment procedure** component directly supports the **exchangeability** (no unmeasured confounding) assumption. In the target trial, randomization guarantees exchangeability. In the emulation, we must argue that, conditional on measured covariates, treatment assignment is independent of potential outcomes. The target trial protocol forces us to explicitly specify what confounders we are adjusting for and why, connecting the design to the identification assumptions.

Another example: the **treatment strategies** component supports the **consistency** assumption. Well-defined treatment strategies ensure that "treatment" means the same thing for everyone, so the potential outcome $Y(a)$ is well-defined. Vague strategies ("statin use") can violate consistency because different patients may experience very different treatment versions (different drugs, doses, or durations).

A third example: the **eligibility criteria** and the scope of the target population help support the **positivity** assumption. By defining the population carefully, we can ensure that within every stratum of confounders, there are both treated and untreated individuals.

</details>

## What comes next

With the target trial protocol defining our causal question (this chapter), the identification framework linking causal and statistical estimands (Chapter 1.3), and the Causal Roadmap organizing the workflow (Chapter 1.2), we are ready to move from design to estimation. In the next part of the handbook, we will learn the statistical methods that deliver on the promise of this framework, beginning with g-computation (Chapter 2.1), inverse probability weighting (Chapter 2.2), and the doubly robust TMLE estimator (Chapter 2.3).

## Sources and further reading

- Hernan MA, Robins JM (2016). Using big data to emulate a target trial when a randomized trial is not available. *American Journal of Epidemiology* 183(8):758-764.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Hernan MA, Sauer BC, Hernandez-Diaz S, Platt R, Shrier I, VanderWeele TJ (2008). Specifying a target trial prevents immortal time bias and other self-inflicted injuries in observational analyses. *Journal of Clinical Epidemiology* 69:8-14.
- Dang LE, Tager I, Petersen ML (2023). A causal roadmap for generating high-quality real-world evidence. *Journal of Clinical and Translational Science* 7(1):e127.
- Suissa S (2008). Immortal time bias in pharmacoepidemiology. *American Journal of Epidemiology* 167(4):492-499.
- Lund JL, Richardson DB, Sturmer T (2015). The active comparator, new user study design in pharmacoepidemiology. *Epidemiology* 26(1):105-113.
- Ray WA (2003). Evaluating medication effects outside of clinical trials: new-user designs. *American Journal of Epidemiology* 158(9):915-920.

```{r}
sessionInfo()
```
