---
title: "Chapter 1.3: From Causal Questions to Analysis: Identification, Estimands, and Statistical Models"
format: html
---

**Note: the contents of this chapter should be integrated into the identification step of causal roadmap chapter and then deleted. the core contents are great and should be kept in the expanded identification step**


In this chapter, we connect the **causal estimand**, the quantity that answers our scientific question, to a **statistical estimand**, which is something we can estimate from observed data. This is the crucial middle step of the Causal Roadmap: translating what we *want* to know into what we *can* learn from the dataset at hand.




::: {.callout-note}
## Learning objectives
- Define identification and explain why it is necessary before estimation
- State the three core identification assumptions: consistency, exchangeability, and positivity
- Distinguish a causal estimand (defined via potential outcomes) from a statistical estimand (a function of the observed-data distribution)
- Derive the g-computation identification formula from the assumptions
- Recognize when identification fails and what alternative strategies exist (sensitivity analysis, bounding, instrumental variables)
:::

*Causal conclusions require identification assumptions and diagnostic evidence that the data support the target estimand.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

# 1. What Does It Mean to *Identify* a Causal Effect?

:::{.callout-note}
## Definition: Identification

**Identification** is the logical argument that, under stated assumptions, an unobservable causal quantity equals something we *can* compute from observed data. A causal estimand (e.g., ATE = E[Y(1) - Y(0)]) is defined via **potential outcomes** we never fully observe. Identification asks: can we express this causal quantity entirely in terms of the observed-data distribution? If yes, the effect is **identified**; if no, additional assumptions or data are needed.
:::

The three key assumptions required for identification are:


:::{.callout-note}
## Assumption 1: Consistency, "You Get What You Get"

The observed outcome equals the counterfactual outcome under the treatment actually received:

$$
Y = Y(A)
$$

In practice, this means the treatment is **well-defined** and there is only one version of it. If someone took denosumab, their observed outcome equals their potential outcome under denosumab. If they took ZA, their observed outcome equals their potential outcome under ZA. Consistency can break down when "treatment" is vague (e.g., "exercise more") or when the drug formulation, dose, or route of administration varies across patients in ways that matter for the outcome.
:::


:::{.callout-warning}
## Assumption 2: Exchangeability, the Hardest Assumption to Defend

Formally:

$$
Y(a) \perp A \mid W
$$

This is the assumption that **there are no unmeasured confounders**. After adjusting for the measured covariates $W$, the treated and untreated groups are comparable, as if treatment had been randomly assigned within strata of $W$.

- In a randomized trial, randomization guarantees exchangeability by design.
- In observational data, we must **assume** it holds, and this assumption is **untestable** from the data alone.

This is typically the most difficult assumption to defend in applied epidemiology. If there are unmeasured factors (e.g., frailty, physician preference, disease severity scores not in the data) that influence both treatment and outcome, identification fails. Always use a DAG to reason about which variables are sufficient to block all backdoor paths, and conduct sensitivity analyses (e.g., E-values) to assess how robust your conclusions are to potential unmeasured confounding.
:::


:::{.callout-warning}
## Assumption 3: Positivity, Watch for Practical Violations

Everyone has a *positive probability* of receiving either treatment at each level of covariates:

$$
0 < P(A=a \mid W=w) < 1
$$


**Theoretical positivity violations**: Some covariate strata *never* receive a particular treatment by clinical convention (e.g., patients over age 90 may never be prescribed aggressive chemotherapy). No statistical method can fix this. The causal effect is simply not defined in those strata.

**Practical positivity violations** are extremely common in real-world observational data and deserve careful attention: Some strata have no or very few patients receiving one treatment, leading to near-zero propensity scores, even though those patients could theoretically recieve treatment. This causes extreme inverse-probability weights and unstable estimates.

Signs to look for: propensity scores piling up near 0 or 1, very large IPTW weights, and g-computation predictions extrapolating far outside the support of the data. We will practice diagnosing this with R shortly.
:::


# 2. The Identification Formula

:::{.callout-tip}
## The Key Insight: Replacing Counterfactuals with Observables

Under the three assumptions, unobservable counterfactuals become observable conditional expectations:

$$
E[Y(1) - Y(0)]
=
E_W\!\left[
E[Y \mid A=1, W]
-
E[Y \mid A=0, W]
\right]
$$

This is the foundation for **g-computation**, **IPTW**, **AIPW**, and **TMLE**.
:::

# 3. The Statistical Estimand

:::{.callout-note}
## Definition: Causal Estimand vs. Statistical Estimand

A **causal estimand** is defined via potential outcomes (e.g., 3-year MI risk difference if all patients received denosumab vs. ZA). A **statistical estimand** is the function of the observed-data distribution that equals the causal estimand *under the identification assumptions* (e.g., the standardized mean difference from a model of $P(Y \mid A, W)$).

**Pitfall.** Regression coefficients (e.g., log-odds ratios) are **not** the statistical estimand for the ATE. They are conditional measures on the log-odds scale, not marginal risk differences.
:::



```{r, include=FALSE}
set.seed(2025)

n <- 5000
age <- rnorm(n, 75, 6)
cvd <- rbinom(n, 1, plogis(0.1 * (age - 70)))

# Treatment assignment with confounding
A <- rbinom(n, 1, plogis(-1 + 0.07 * (age - 70) + 1.2 * cvd))

# Potential outcomes generated under each treatment
risk <- plogis(-2 + 0.5*A + 0.08*(age - 70) + 0.9*cvd)
Y <- rbinom(n, 1, risk)

dat <- data.frame(age, cvd, A, Y)
mod <- glm(Y ~ A + age + cvd, family = binomial, data = dat)

```




# 4. Diagnosing Exchangeability and Positivity

## 4.1 Overlap check

:::{.callout-important}
## Diagnosing Positivity Violations in Practice

The propensity score plot below is your primary diagnostic. Look for:

- **Mass piling up near 0 or 1**: Some patients have near-certain probability of receiving (or not receiving) treatment. The data contain little information about what would happen under the other treatment.
- **Poor overlap between groups**: If the propensity score distributions for $A=1$ and $A=0$ lack common support, causal estimates in the non-overlapping region rely on extrapolation.

When you see these patterns, consider restricting to the region of common support, trimming extreme weights, or acknowledging the limitations.
:::

```{r}
plot(density(predict(glm(A ~ age + cvd, family = binomial, data = dat), type = "response")),
     main = "Propensity Score Overlap", xlab = "Estimated Propensity Score")

ps <- predict(glm(A ~ age + cvd, family = binomial, data = dat), type = "response")
```

# 5. Summary

:::{.callout-tip}
## Chapter Takeaways

- Identification bridges **potential outcomes** (causal) to **observable quantities** (statistical) via three assumptions: consistency, exchangeability, and positivity.
- A statistical estimand is not a regression coefficient. It is the observed-data function that equals the causal estimand under the identification assumptions.
:::

## Check your understanding

<details>
<summary><strong>Q1. If you have perfect covariate balance between treated and untreated groups, does that prove exchangeability holds?</strong></summary>

No. Covariate balance on *measured* covariates is necessary but not sufficient for exchangeability. Exchangeability requires that there are no unmeasured common causes of treatment and outcome, a condition that cannot be verified from the data alone. You could have perfect balance on every variable in your dataset while an unmeasured confounder (e.g., frailty, physician preference, or disease severity not captured in the data) still biases the estimate. Balance checks are useful diagnostics, but they cannot substitute for substantive knowledge and a well-reasoned DAG.

</details>

<details>
<summary><strong>Q2. A colleague runs a logistic regression and reports the coefficient on treatment as "the causal effect." What is wrong with this interpretation?</strong></summary>

A logistic regression coefficient is a *conditional log-odds ratio*. It describes the association between treatment and outcome holding covariates fixed at specific values. This is not a causal effect for at least two reasons. First, a coefficient from a regression model is only a causal effect if the identification assumptions (consistency, exchangeability, positivity) hold *and* the model is correctly specified. Simply running a regression does not guarantee either. Second, the conditional log-odds ratio is a different quantity from the marginal risk difference (ATE) that typically answers the scientific question. The log-odds ratio is non-collapsible: its value changes depending on which covariates are conditioned on, even absent confounding. To obtain a causal estimate on a meaningful scale, you need to go through the identification step and apply a method like g-computation that standardizes over the population. See Chapter 1.1 for a detailed demonstration of this distinction.

</details>

<details>
<summary><strong>Q3. What is a positivity violation, and what should you do when you encounter one?</strong></summary>

A positivity violation occurs when certain covariate strata have zero (or near-zero) probability of receiving one of the treatment levels. **Structural violations** arise when some patients are never treated by clinical convention (e.g., very elderly patients never receiving aggressive chemotherapy). The causal effect is simply not defined in those strata, and no statistical method can fix this. **Practical (random) violations** occur when some strata have very few patients on one treatment, leading to propensity scores near 0 or 1 and unstable estimates. When you encounter positivity violations, you should: (1) inspect propensity score distributions for mass near the boundaries, (2) consider restricting the target population to the region of overlap where the effect is well-defined, (3) use weight trimming or stabilized weights if using IPTW, and (4) be transparent about the limitations and the population to which your estimate applies.

</details>

## What comes next

Before moving to estimation, the next chapter introduces **target trial emulation**: the discipline of designing every observational analysis as if it were a randomized trial. Target trial thinking prevents common biases (immortal time bias, misaligned time zero, vague comparators) and connects the causal roadmap directly to real-world evidence practice. After that, Part II turns to estimation methods.

## Sources and further reading

- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapters 1-3. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Pearl J (2009). *Causality: Models, Reasoning, and Inference*. 2nd ed. Cambridge University Press.
- Robins JM (1986). A new approach to causal inference in mortality studies with sustained exposure periods. *Mathematical Modelling* 7:1393-1512.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer. Chapter 2.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- Kennedy EH (2019). Nonparametric causal effects based on incremental propensity score interventions. *JASA* 114(526):645-656.
- `lmtp` R package: [CRAN](https://cran.r-project.org/package=lmtp)
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)

