---
subtitle: "A flexible learner with a convergence guarantee targeted learning can rely on"
format: html
---

# The Highly Adaptive Lasso {#sec-hal}
*One estimator flexible enough to fit almost any relationship, yet provably fast enough for valid inference*

*Estimated reading time: 25 minutes*

Super Learner (@sec-superlearner) gives an honest way to pick the best blend of prediction algorithms, but it leaves one question unanswered. TMLE promises confidence intervals that are valid even when the nuisance models are estimated with machine learning, and that promise comes with fine print: the nuisance estimates have to approach the truth *fast enough* as the sample grows. Most flexible algorithms, including random forests and gradient boosting, come with no guarantee about how fast their error shrinks. The highly adaptive lasso (HAL) is the unusual exception. It is flexible enough to recover almost any relationship between covariates and outcome, yet it converges quickly enough, under a mild condition, to satisfy the requirement TMLE needs. That combination is why HAL keeps appearing in targeted-learning work, and why it is worth understanding on its own.

::: {.callout-note}
## Learning objectives
- Explain, without the theory, why an estimator's convergence rate matters for valid TMLE confidence intervals
- Describe what HAL fits and the one assumption it relies on
- Use HAL as a Super Learner candidate or as a standalone nuisance estimator
- Recognize when HAL is impractical and what it does not fix
:::

---

## 1. Why the speed of an estimator matters

A doubly robust estimator such as TMLE combines an outcome model and a propensity model. Its confidence intervals are trustworthy when the *product* of the two models' errors vanishes quickly as the sample grows. A useful rule of thumb is that each nuisance estimate should approach the truth faster than the fourth root of the sample size. A simple, correctly specified regression clears this bar easily. The price of that simplicity is the risk of getting the functional form wrong, which biases the estimate in a way no amount of data fixes.

Flexible machine learning protects against the wrong functional form, but most flexible methods come with no guarantee about how fast their error shrinks, so they can quietly violate the rate condition and leave the confidence interval too narrow. There are two ways out. One is to change the estimator so it no longer needs the rate condition, which is what cross-validated TMLE does through sample splitting (@sec-advanced-tmle). The other is to use a flexible learner that *does* meet the rate condition. HAL is that learner.

---

## 2. What HAL fits

The idea behind HAL is easier to picture than its name suggests. Take every covariate and, at each value it takes in the data, define a simple step that switches from zero to one. Do the same for pairs of covariates, triples, and so on. This produces a very large collection of step-shaped building blocks. Any reasonable relationship between covariates and outcome can be written as a weighted sum of these steps: a smooth curve becomes a staircase with many small steps, a sharp threshold becomes one large step, and an interaction becomes a step that only switches on when two covariates are both above their cutpoints.

HAL fits a weighted sum of all these building blocks at once, using the lasso to keep the total size of the weights under control. The single tuning quantity is that total size, which corresponds to how much the fitted function is allowed to wiggle. A small budget yields a nearly flat fit; a larger budget lets the staircase follow the data more closely. In one sentence, HAL is a lasso over a very large basis of step-function interactions, with one penalty governing flexibility.

Because the building blocks include sharp steps and interactions automatically, HAL captures the kinds of relationships that defeat a main-terms regression. A clinical example: if mortality rises steeply once CD4 falls below roughly 200 cells per microliter, and that threshold matters more in older patients, a main-terms logistic regression will smooth the threshold away and miss the age interaction, while HAL can represent both with a few of its step blocks.

---

## 3. The one assumption, and the payoff

HAL needs the true relationship to have *finite variation*, which means it does not wiggle without limit: it can rise, fall, jump, and bend, but only a controlled total amount across the covariate space. For essentially every relationship an epidemiologist would posit between patient characteristics and an outcome, this holds.

Under that mild condition, HAL converges to the truth at a rate faster than the fourth-root bar from Section 1, regardless of how many covariates are involved. That is the payoff: HAL is flexible like machine learning yet fast like a correctly specified parametric model, so it can serve as the nuisance estimator in TMLE and deliver valid confidence intervals, in many settings without requiring sample splitting [@benkeser2016hal; @vanderLaan2011targeted].

---

## 4. Using HAL in practice

There are two natural ways to use HAL, and the first is the safer default. Add it as one candidate in a Super Learner library (`SL.hal9001`). The ensemble then leans on HAL where its flexible fit helps and on simpler learners where it does not, and the analyst gets HAL's strengths without betting the whole analysis on it. The second way is to use HAL directly as the outcome or propensity estimator inside TMLE, which is the route that most cleanly inherits the rate guarantee.

The `hal9001` package implements both. The two knobs worth knowing are the maximum interaction order, which sets whether HAL considers single covariates only, pairs, or higher combinations, and the smoothness setting, which controls whether the building blocks are sharp steps or smoother pieces. The main practical cost is computation: the basis grows quickly with the number of covariates and the sample size, so analysts often cap the interaction order at two and screen covariates before fitting in high-dimensional problems.

```{r}
#| eval: false
source("_data/load-haart-cohort.R")
library(hal9001)

X <- as.matrix(haart_baseline[, c("age", "sex", "cd4_sqrt_baseline")])
Y <- haart_baseline$died

# Fit the outcome model with HAL: pairwise interactions of step-function bases.
hal_fit <- fit_hal(X = X, Y = Y, family = "binomial", max_degree = 2)

# Compare HAL's fitted mortality risks with a main-terms logistic regression.
p_hal <- predict(hal_fit, new_data = X)
p_glm <- predict(glm(died ~ age + sex + cd4_sqrt_baseline,
                     family = binomial, data = haart_baseline), type = "response")

summary(p_hal); summary(p_glm)
```

To use HAL inside an ensemble instead, list it as a learner and let Super Learner weight it against the others.

```{r}
#| eval: false
library(SuperLearner)
sl_fit <- SuperLearner(
  Y = Y, X = as.data.frame(X), family = binomial(),
  SL.library = c("SL.glm", "SL.mean", "SL.hal9001")
)
sl_fit  # the coefficient on SL.hal9001 shows how much weight the ensemble gives HAL
```

Where the true outcome surface has a threshold or interaction, HAL's fitted risks will separate patients that the main-terms model lumps together, and in a TMLE analysis those better nuisance fits translate into a less biased effect estimate with intervals that keep their nominal coverage.

---

## 5. Where HAL is the wrong tool, and what it does not fix

HAL is computationally demanding. With many covariates or a large sample, the full basis can exhaust memory, so capping the interaction order and screening covariates first are usually necessary, and at some scale a different learner is simply more practical. The finite-variation assumption, although mild, is still an assumption; a relationship that genuinely behaves erratically across the covariate space is one HAL cannot represent well.

Most importantly, HAL is a better way to estimate nuisance functions, not a way to escape the causal assumptions. It does nothing for unmeasured confounding, and it does not manufacture support where the data have none. If positivity fails, HAL will extrapolate across the gap as silently as any other flexible model (@sec-positivity-overlap). A faster, more flexible nuisance estimator makes the statistical step more reliable; it leaves the identification step exactly where the Causal Roadmap left it.

---

## Connections to other chapters

HAL is one more candidate for the Super Learner library of @sec-superlearner, distinguished by the convergence guarantee that the doubly robust estimators of @sec-doubly-robust-tmle rely on for valid inference. It offers an alternative to the sample-splitting route of cross-validated TMLE in @sec-advanced-tmle: where CV-TMLE removes the rate requirement by splitting the data, HAL satisfies it by construction. Both are reasonable defaults, and they can be combined.

## Sources and further reading

- @benkeser2016hal introduces the highly adaptive lasso and its convergence-rate guarantee.
- @vanderLaan2011targeted places HAL within the targeted-learning framework and its inference requirements.
- The `hal9001` R package implements the estimator: <https://cran.r-project.org/package=hal9001>.

## Check your understanding

<details>
<summary><strong>Q1. Why does TMLE care about how fast a nuisance estimator converges, when ordinary prediction tasks only care about accuracy?</strong></summary>

TMLE's confidence intervals are valid when the combined error of the outcome and propensity models shrinks faster than a set rate (roughly the fourth root of the sample size). If a flexible learner is accurate but converges slowly, the intervals can be too narrow and overstate certainty. A pure prediction task has no such requirement because it is not trying to support a confidence interval for a causal effect.

</details>

<details>
<summary><strong>Q2. In plain terms, what does HAL fit, and what single quantity controls its flexibility?</strong></summary>

HAL fits a weighted sum of many simple step-shaped building blocks defined at the observed covariate values and their interactions, so it can represent thresholds, bends, and interactions automatically. The lasso penalty bounds the total size of the weights, which corresponds to how much the fitted function is allowed to wiggle. That single budget is what trades a flat fit against a fit that follows the data closely.

</details>

<details>
<summary><strong>Q3. A reviewer says "you used HAL, so positivity is no longer a concern." Is that right?</strong></summary>

No. HAL improves how well the nuisance functions are estimated, which helps the statistical step, but it does not create data support where there is none. If some covariate group never received one of the treatments, HAL will extrapolate across that gap just like any flexible model. Positivity, like no unmeasured confounding, is an identification assumption that a better estimator cannot repair.

</details>
