---
title: "Chapter 2.2: Inverse Probability of Treatment Weighting (IPTW)"
format: html
---

# Chapter 2.2: Inverse Probability of Treatment Weighting  

Inverse probability of treatment weighting (IPTW) is a core method in modern causal inference. Instead of modeling the outcome directly as in g computation, IPTW uses a model for **treatment assignment** to create a pseudo population where treatment is independent of confounders.  

In this chapter we will

- Build intuition for IPTW
- Derive the weights and connect them to the causal estimand
- Show how to estimate and diagnose propensity scores
- Implement IPTW in R with tidyverse style
- Discuss stabilized weights and truncation
- Compare IPTW to g computation on the same simulated data

This is still point treatment only. Longitudinal extensions come later.

---

# 1. Intuition: Reweighting to Mimic a Trial

:::{.callout-tip}
## Key Insight: IPTW Creates a Pseudo-Population That Mimics a Randomized Trial

IPTW reweights observational data so that, in the weighted sample, treatment appears as if it were randomized given measured covariates. Patients whose treatment assignment was unlikely for their covariate pattern receive large weights; patients whose assignment was predictable receive small weights. The result is a **pseudo-population** where confounders are balanced across treatment groups --- just as in a randomized trial.
:::

:::{.callout-note}
## Definition: The Propensity Score

The **propensity score** is the probability of receiving treatment given measured covariates:

$$
e(W) = P(A = 1 \mid W)
$$

IPTW weights are constructed from the propensity score:

- For treated:  \( \displaystyle w_i = \frac{1}{e(W_i)} \)
- For control: \( \displaystyle w_i = \frac{1}{1 - e(W_i)} \)

The propensity score is a **balancing score**: conditioning on it is sufficient to remove confounding by all measured covariates $W$. This is why a single scalar summary can replace adjustment for many covariates.
:::

---

# 2. Identification Using IPTW

:::{.callout-note}
## The IPTW Identification Result

Under the same causal assumptions as before

- **Consistency**: the observed outcome equals the potential outcome under the treatment actually received
- **Exchangeability**: \( Y(a) \perp A \mid W \) --- no unmeasured confounding
- **Positivity**: every covariate stratum has a non-zero probability of receiving each treatment

the average potential outcome under treatment can be expressed as

$$
E[Y(1)] = E\left[ \frac{I(A = 1) Y}{e(W)} \right]
$$

and under control

$$
E[Y(0)] = E\left[ \frac{I(A = 0) Y}{1 - e(W)} \right]
$$

The IPTW estimator replaces the expectation with the sample average and replaces the true \( e(W) \) with an estimated one.
:::

This is called a **Horvitz-Thompson type** estimator because it has the same structure as the survey sampling estimator that reweights observations by the inverse of their selection probability. In our setting, "selection" is selection into a treatment group --- we upweight patients whose treatment assignment was unlikely given their covariates.

---

# 3. Simulated Example

We reuse a familiar setup: an osteoporosis like population with confounding by age and cardiovascular history.

```{r}
library(tidyverse)

set.seed(2025)
n <- 4000

age <- rnorm(n, 75, 6)
cvd <- rbinom(n, 1, plogis(0.12 * (age - 70)))

# Treatment assignment with strong confounding
A <- rbinom(n, 1, plogis(-1 + 0.09 * (age - 70) + 1.5 * cvd))

# Outcome model
Y <- rbinom(n, 1, plogis(-2 + 0.5*A + 0.10*(age - 70) + 1.0*cvd))

dat <- tibble(age, cvd, A, Y)
dat %>% head()
```

For simplicity

- A = 1 can be thought of as denosumab
- A = 0 can be thought of as zoledronic acid
- Y = 1 indicates occurrence of a cardiovascular outcome

---

# 4. Estimating the Propensity Score

We model the probability of receiving treatment given covariates.

In practice, one might use logistic regression, SuperLearner, or other ML. Here we start with logistic regression for clarity.

```{r}
ps_mod <- glm(A ~ age + cvd, family = binomial, data = dat)
summary(ps_mod)

dat <- dat %>%
  mutate(
    ps = predict(ps_mod, type = "response")
  )

dat %>% 
  summarize(
    min_ps = min(ps),
    max_ps = max(ps)
  )
```

We can visualize the propensity score distribution.

```{r}
ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  labs(
    x = "Estimated propensity score P(A=1 | W)",
    fill = "Treatment",
    title = "Propensity score overlap"
  ) +
  theme_minimal()
```

:::{.callout-important}
## Reading the Propensity Score Overlap Plot

**Good overlap** looks like two density curves that cover mostly the same range of propensity scores, even if their peaks differ. This means that for most covariate patterns, there are both treated and untreated patients --- exactly what positivity requires.

**Bad overlap** looks like two density curves that barely touch, with the treated group concentrated at high propensity scores and the untreated group concentrated at low ones. When this happens:

- IPTW produces extreme weights for patients in the non-overlapping tails
- Estimates become highly variable and unreliable
- You may need to restrict your target population to the region of overlap (trimming) or reconsider your study design

As a rule of thumb, be concerned if propensity scores for either group pile up near 0 or 1.
:::

---

# 5. Constructing IPTW Weights

**Definition.** Unstabilized weights are the simplest form of IPTW weights. For each individual, the weight is the inverse of the probability of receiving the treatment they actually received: for the treated ($A = 1$), $w_i = 1 / e(W_i)$; for the untreated ($A = 0$), $w_i = 1 / (1 - e(W_i))$. These weights create a pseudo-population where treatment is independent of measured confounders, but they can be highly variable when some propensity scores are close to 0 or 1.

```{r}
dat <- dat %>%
  mutate(
    w_ipw = if_else(A == 1, 1 / ps, 1 / (1 - ps))
  )

summary(dat$w_ipw)
quantile(dat$w_ipw, probs = c(0.01, 0.99))
```

Keep an eye on

- Very large weights (suggest near-violations of positivity)
- Range and extreme quantiles (the 99th percentile should not be orders of magnitude larger than the median)

:::{.callout-tip}
## Why Stabilized Weights Are Preferred in Practice

Stabilized weights multiply the standard IPTW weights by the **marginal probability of treatment**, which brings the weights closer to 1 on average and reduces their variance without introducing bias for the marginal causal effect.

For a binary treatment:

$$
SW_i = \frac{P(A_i)}{e(W_i)} \text{ if } A_i = 1
$$

and

$$
SW_i = \frac{P(A_i)}{1 - e(W_i)} \text{ if } A_i = 0
$$

Here \( P(A_i) \) is the marginal probability of treatment (estimated as the sample proportion treated).

**Why this helps:** Unstabilized weights create a pseudo-population that is larger than the original sample, which inflates standard errors. Stabilized weights preserve the original sample size in expectation, leading to:

- Narrower confidence intervals
- More stable point estimates
- Better finite-sample performance

In epidemiologic practice, **always prefer stabilized weights** unless you have a specific reason to use unstabilized ones.
:::

```{r}
pA <- mean(dat$A)

dat <- dat %>%
  mutate(
    sw_ipw = case_when(
      A == 1 ~ pA / ps,
      A == 0 ~ (1 - pA) / (1 - ps)
    )
  )

summary(dat$sw_ipw)
quantile(dat$sw_ipw, probs = c(0.01, 0.99))
```

---

# 6. Estimating the ATE via IPTW

There are several equivalent ways to use the weights.

## 6.1 Direct weighted mean of outcomes

Estimated risk under treatment

```{r}
risk1_ipw <- with(dat, sum(sw_ipw * Y * (A == 1)) / sum(sw_ipw * (A == 1)))
risk0_ipw <- with(dat, sum(sw_ipw * Y * (A == 0)) / sum(sw_ipw * (A == 0)))

ate_ipw <- risk1_ipw - risk0_ipw
c(risk1 = risk1_ipw, risk0 = risk0_ipw, ate = ate_ipw)
```

This matches the formula

$$
\hat E[Y(1)] = \frac{\sum_i SW_i Y_i I(A_i = 1)}{\sum_i SW_i I(A_i = 1)}
$$

## 6.2 Weighted regression model

We can also fit a weighted regression with treatment as the only predictor.

```{r, eval=FALSE}
library(sandwich)
library(lmtest)

# Fit a simple weighted model
fit_ipw <- glm(Y ~ A, family = binomial, weights = sw_ipw, data = dat)

# Robust standard errors
cov_ipw <- vcovHC(fit_ipw, type = "HC0")
coeftest(fit_ipw, cov_ipw)
```

Interpretation

- The coefficient of A (on the log odds scale) now estimates a marginal effect in the pseudo population
- You can compute marginal risk differences or ratios by predicting from the model at A=1 and A=0 and standardizing

---

# 7. Comparing IPTW and G Computation

We can compare the IPTW ATE to the g computation ATE from the previous chapter.

```{r}
# g computation
g_mod <- glm(Y ~ A + age + cvd, family = binomial, data = dat)
p1_g <- predict(g_mod, newdata = dat %>% mutate(A = 1), type = "response")
p0_g <- predict(g_mod, newdata = dat %>% mutate(A = 0), type = "response")
ate_g <- mean(p1_g - p0_g)

c(ate_gcomp = ate_g, ate_iptw = ate_ipw)
```

Under correct models and adequate positivity, these methods should converge to similar estimates as sample size grows.

:::{.callout-tip}
## Key Takeaway: IPTW and G-Computation Depend on Different Models

- **G-computation** depends on correctly specifying the **outcome model** $E[Y \mid A, W]$
- **IPTW** depends on correctly specifying the **treatment model** $P(A \mid W)$

Neither uses both models at once. This motivates **doubly robust** methods (next chapter), which provide consistent estimates if *either* model is correctly specified. When IPTW and g-computation give very different answers, at least one model may be misspecified.
:::

---

# 8. Diagnostics and Practical Tips

IPTW is powerful but fragile if diagnostics are ignored.

**Practical note.** Every IPTW analysis should include diagnostics. Reviewers and readers should expect to see them --- omitting diagnostics is a red flag in any manuscript.

### 8.1 Check propensity score overlap

To assess positivity, plot propensity score densities by treatment group (as in Section 4) and flag regions where one group is missing or sparse. If the treated and untreated distributions do not overlap, the ATE may not be identifiable in that region --- consider restricting inference to the overlap population or reporting the ATT instead.

### 8.2 Check weights

```{r}
dat %>%
  summarize(
    min_w = min(sw_ipw),
    max_w = max(sw_ipw),
    mean_w = mean(sw_ipw),
    sd_w = sd(sw_ipw)
  )
```

:::{.callout-warning}
## When Extreme Weights Signal Trouble

Large extreme weights are a red flag. Watch for these warning signs:

- **Maximum weight > 10--20**: a single observation is being counted as 10--20 people. Your estimate may be driven by one or two individuals.
- **Mean of stabilized weights far from 1**: stabilized weights should average approximately 1. Deviations suggest model misspecification.
- **Ratio of max to median weight > 20**: indicates near-violation of positivity for some covariate patterns.

When you see extreme weights, consider: (1) improving your propensity score model, (2) checking for data errors in the extreme-weight observations, and (3) applying weight truncation (see below).
:::

### 8.3 Consider truncation

You can truncate weights at a chosen percentile, for example the 1st and 99th percentiles.

```{r}
lower <- quantile(dat$sw_ipw, 0.01)
upper <- quantile(dat$sw_ipw, 0.99)

dat <- dat %>%
  mutate(
    sw_trunc = pmin(pmax(sw_ipw, lower), upper)
  )

# recompute ATE with truncated weights
risk1_trunc <- with(dat, sum(sw_trunc * Y * (A == 1)) / sum(sw_trunc * (A == 1)))
risk0_trunc <- with(dat, sum(sw_trunc * Y * (A == 0)) / sum(sw_trunc * (A == 0)))
ate_trunc   <- risk1_trunc - risk0_trunc

c(ate_truncated = ate_trunc)
```

:::{.callout-warning}
## The Truncation Trade-Off: Bias vs. Variance

Truncation (also called "winsorizing") reduces variance at the cost of introducing some bias. By capping extreme weights, you are effectively saying: "I am willing to accept a small amount of bias in exchange for a much more stable estimate."

**The trade-off in practice:**

- **Without truncation**: unbiased (if the propensity model is correct) but potentially high variance, with confidence intervals that may be uninformatively wide
- **With truncation**: slightly biased but often lower mean squared error, producing more useful estimates

**Common approaches:**

- Truncate at the 1st and 99th percentiles (as shown above)
- Truncate at fixed values (e.g., cap weights at 10 or 20)
- Report results with and without truncation as a sensitivity analysis

There is no single "right" truncation threshold. In your manuscripts, always report which threshold you used and show sensitivity to this choice.
:::

---

# 9. Summary

:::{.callout-tip}
## Chapter Takeaways

- IPTW reweights observational data to mimic a randomized trial using propensity scores; always prefer **stabilized weights**
- IPTW depends on a correctly specified treatment model --- never skip overlap and weight diagnostics
- Sensitive to positivity violations; extreme weights can dominate estimates
- Does not use outcome-model information, motivating **doubly robust** methods (next chapter)
:::

---

## Check your understanding

<details>
<summary><strong>Q1. Why does IPTW depend on correct specification of the treatment model rather than the outcome model?</strong></summary>

IPTW estimates causal effects by reweighting observations using the inverse of the propensity score $e(W) = P(A = 1 \mid W)$. The weights are constructed entirely from the treatment model --- the outcome model is never used. If the propensity score model is misspecified, the weights will not correctly balance confounders across treatment groups, and the pseudo-population will still be confounded. By contrast, g-computation relies on the outcome model and never uses the treatment model. This distinction is why doubly robust methods, which use both models, are attractive.

</details>

<details>
<summary><strong>Q2. A colleague reports IPTW results without checking propensity score overlap or weight diagnostics. What could go wrong?</strong></summary>

Without diagnostics, several problems can go undetected. Near-violations of positivity (propensity scores close to 0 or 1) produce extreme weights, meaning a few individuals can dominate the entire estimate. The ATE may be driven by a handful of observations rather than the full sample. Standard errors may be artificially narrow or wide, and the estimate can be highly sensitive to small perturbations in the data. Additionally, if the mean of stabilized weights deviates substantially from 1, this signals propensity score model misspecification. Reporting IPTW results without overlap plots and weight summaries makes it impossible for reviewers to assess the credibility of the analysis.

</details>

<details>
<summary><strong>Q3. What is the difference between stabilized and unstabilized weights, and why do we prefer stabilized weights?</strong></summary>

Unstabilized weights are simply $1 / e(W)$ for the treated and $1 / (1 - e(W))$ for the untreated. Stabilized weights multiply these by the marginal probability of treatment: $P(A) / e(W)$ for the treated and $(1 - P(A)) / (1 - e(W))$ for the untreated. This rescaling brings the weights closer to 1 on average and preserves the original sample size in expectation. Stabilized weights produce narrower confidence intervals and more stable point estimates without introducing bias for the marginal causal effect. In practice, stabilized weights should be the default choice.

</details>

```{r}
sessionInfo()
```
