---
title: "Chapter 2.3: Doubly Robust Estimators and Targeted Learning (AIPW + TMLE)"
format: html
---

# Chapter 2.3: Doubly Robust Estimation and Targeted Learning
*Bridging outcome modeling and weighting for more robust causal effect estimation*

In previous chapters, you learned:

- **G-computation** depends on correctly modeling the *outcome*
- **IPTW** depends on correctly modeling the *treatment mechanism*

But what if you could use **both models**, and as long as **either one is correct**, your estimator is still consistent?

This is exactly what **doubly robust estimators** provide.

We explore:

- Why doubly robust estimators matter
- AIPW (Augmented IPTW)
- TMLE (Targeted Maximum Likelihood Estimation)
- How to implement both with tidyverse-friendly R code
- Why TMLE is preferred in modern causal inference

---

# 1. Why Doubly Robust Estimators?

:::{.callout-tip}
## Key Idea: Why Doubly Robust Estimators?

A doubly robust estimator is consistent if **either** the treatment model or the outcome model is correctly specified. By combining both sources of information, doubly robust estimators provide **two chances to get it right** and often outperform G-computation or IPTW used alone.
:::

---

# 2. AIPW: Augmented Inverse Probability Weighting

:::{.callout-note}
## What Makes AIPW "Doubly Robust"?

AIPW adds a correction term to IPTW using the outcome regression. The key insight is that it is constructed so that:

- If the **treatment model** is right, the IPTW component produces an unbiased estimate
- If the **outcome model** is right, the augmentation (G-computation) component corrects any bias in the weights
- If **both** are right, the estimator achieves the **semiparametric efficiency bound**, meaning it has the smallest possible variance among all regular asymptotically linear estimators

In other words, AIPW "augments" the weighted estimator with a bias-correction term derived from the outcome model. This makes the estimator robust to misspecification of either (but not both) nuisance models, hence the name *doubly robust*.
:::

## 2.1 Formula

Let:

- $e(W) = P(A = 1 \mid W)$ be the propensity score
- $m(a, W) = E[Y \mid A = a, W]$ be the outcome regression

Then:

$$
\hat{\mu}_1 = \frac{1}{n} \sum_i \left[ \frac{A_i Y_i}{e(W_i)} - \frac{A_i - e(W_i)}{e(W_i)} m(1, W_i) \right]
$$

$$
\hat{\mu}_0 = \frac{1}{n} \sum_i \left[ \frac{(1-A_i) Y_i}{1 - e(W_i)} + \frac{A_i - e(W_i)}{1 - e(W_i)} m(0, W_i) \right]
$$

ATE = $\hat{\mu}_1 - \hat{\mu}_0$

---

# 3. Simulated Osteoporosis Dataset

```{r}
library(tidyverse)

set.seed(2026)
n <- 4000

age <- rnorm(n, 75, 6)
cvd <- rbinom(n, 1, plogis(0.12 * (age - 70)))

# Treatment model
A <- rbinom(n, 1, plogis(-1 + 0.08*(age - 70) + 1.4*cvd))

# Outcome model
Y <- rbinom(n, 1, plogis(-2 + 0.6*A + 0.10*(age - 70) + 1.0*cvd))

dat <- tibble(age, cvd, A, Y)
```

---

# 4. Fit Outcome and Propensity Score Models

```{r}
Qmod <- glm(Y ~ A + age + cvd, family = binomial, data = dat)
gmod <- glm(A ~ age + cvd, family = binomial, data = dat)

dat <- dat %>%
  mutate(
    ps = predict(gmod, type = "response"),
    Q1 = predict(Qmod, newdata = mutate(dat, A=1), type = "response"),
    Q0 = predict(Qmod, newdata = mutate(dat, A=0), type = "response")
  )
```

---

# 5. Compute AIPW Estimator

```{r}
with(dat, {
  mu1 <- mean( Q1 + A * (Y - Q1) / ps )
  mu0 <- mean( Q0 + (1 - A) * (Y - Q0) / (1 - ps) )
  ate <- mu1 - mu0
  c(mu1 = mu1, mu0 = mu0, ate = ate)
})
```

:::{.callout-tip}
## Interpreting the AIPW Estimate

Each AIPW component starts with the outcome-model prediction and adds a weighted residual correction. For instance, `Q1 + A * (Y - Q1) / ps` corrects $m(1, W)$ by upweighting the residual $(Y - Q1)$ for treated individuals. This structure gives AIPW its doubly robust property: even if $Q$ is wrong, the IPW correction fixes the bias (and vice versa).
:::

---

# 6. TMLE: Targeted Maximum Likelihood Estimation

:::{.callout-tip}
## Key Idea: Why TMLE Over AIPW?

TMLE offers several advantages over AIPW: predicted probabilities always stay in $[0, 1]$ (AIPW can fall outside), it integrates naturally with ensemble machine learning (SuperLearner), and the fluctuation step solves the efficient influence curve equation by construction, providing a foundation for valid inference.
:::

---

# 7. TMLE Step-by-Step

**Step 1: Fit the initial outcome model (Q).** Fit a model for $E[Y \mid A, W]$ and generate initial predictions under both treatment values for each individual: $Q_n^0(1, W_i)$ (predicted outcome if treated) and $Q_n^0(0, W_i)$ (predicted outcome if untreated). These predictions are "initial" because they have not yet been targeted toward the causal parameter of interest.

```{r}
logit <- function(p) log(p/(1-p))
expit <- function(x) 1/(1+exp(-x))

# Step 1: Initial Q predictions at observed A
Qinit <- predict(Qmod, type="response")
```

**Step 2: Estimate the propensity score.** The propensity score $e(W) = P(A = 1 \mid W)$ is needed to construct the clever covariate in the next step. We already fit this model in Section 4.

```{r}
# Step 2: Propensity scores (already estimated)
ps <- dat$ps
```

:::{.callout-note}
## Definition: The Clever Covariate

The **clever covariate** $H(A, W)$ is the key ingredient that makes TMLE "targeted." It is derived from the efficient influence curve for the ATE:

$$
H(A, W) = \frac{A}{e(W)} - \frac{1 - A}{1 - e(W)}
$$

For computing the counterfactual means, we also need the treatment-specific components:

- $H_1(W) = \frac{1}{e(W)}$ (for the treated potential outcome)
- $H_0(W) = \frac{-1}{1 - e(W)}$ (for the control potential outcome)

The clever covariate directs the fluctuation to reduce bias for the specific causal estimand we care about, the ATE.
:::

```{r}
# Step 3: Clever covariate at observed treatment
H <- with(dat, A/ps - (1-A)/(1-ps))
```

:::{.callout-important}
## Step 4: Fit the Fluctuation Model and Update Predictions

This is the critical "targeting" step. We fit a logistic regression of $Y$ on $H$ with the initial logit-predictions as an offset:

$$
\text{logit}(E[Y \mid A, W]) = \text{logit}(Q_n^0(A, W)) + \epsilon \cdot H(A, W)
$$

The estimated $\hat{\epsilon}$ is then used to **update** (fluctuate) the initial predictions. Crucially, we must apply this fluctuation to the counterfactual predictions $Q_n^0(1, W)$ and $Q_n^0(0, W)$ using their respective clever covariates $H_1$ and $H_0$.

This step ensures the final estimates solve the efficient influence curve equation, which is what gives TMLE its optimal statistical properties.
:::

```{r}
# Step 4: Fluctuation model
epsilon <- glm(dat$Y ~ -1 + offset(logit(Qinit)) + H,
               family = binomial)$coef
```

---

# 8. Counterfactual Predictions for TMLE

```{r}
# Get initial (pre-fluctuation) predictions under each treatment
Q1_init <- predict(Qmod, newdata = mutate(dat, A=1), type="response")
Q0_init <- predict(Qmod, newdata = mutate(dat, A=0), type="response")

# Clever covariates for counterfactual treatment values
H1 <- 1 / ps
H0 <- -1 / (1 - ps)

# Apply the fluctuation to counterfactual predictions
Q1_star <- plogis(logit(Q1_init) + epsilon * H1)
Q0_star <- plogis(logit(Q0_init) + epsilon * H0)

# TMLE ATE estimate
tmle_ate <- mean(Q1_star) - mean(Q0_star)
tmle_ate
```

---

# 8.1 TMLE Inference: Standard Errors and Confidence Intervals

A major advantage of TMLE is that valid inference follows directly from the **efficient influence curve (EIC)**. The EIC for the ATE at the TMLE fit is:

$$
D_i = H_1(W_i)(Y_i - Q_1^*(W_i)) - H_0(W_i)(Y_i - Q_0^*(W_i)) + Q_1^*(W_i) - Q_0^*(W_i) - \hat{\psi}
$$

where $\hat{\psi}$ is the TMLE ATE estimate. The variance of the ATE is estimated by the sample variance of the EIC divided by $n$.

```{r}
# Observed outcomes
Y_obs <- dat$Y

# EIC (efficient influence curve) components
# For treated: use Q1_star; for control: use Q0_star
# But we need Q* at the OBSERVED A for the residual terms
Qstar_obs <- ifelse(dat$A == 1, Q1_star, Q0_star)
H_obs <- ifelse(dat$A == 1, H1, H0)

# Efficient influence curve for each observation
EIC <- H_obs * (Y_obs - Qstar_obs) + Q1_star - Q0_star - tmle_ate

# Standard error and 95% CI
tmle_se <- sqrt(var(EIC) / n)
ci_lower <- tmle_ate - 1.96 * tmle_se
ci_upper <- tmle_ate + 1.96 * tmle_se

tibble(
  Estimate = tmle_ate,
  SE = tmle_se,
  CI_lower = ci_lower,
  CI_upper = ci_upper
)
```

:::{.callout-tip}
## Interpreting TMLE Inference

The EIC-based standard error is a key output of TMLE. Because the fluctuation step solves the EIC estimating equation, the resulting confidence interval has correct coverage under standard regularity conditions, even when machine learning is used for the nuisance models (provided cross-fitting or other sample-splitting techniques are employed).

If the confidence interval excludes zero, we have evidence that the treatment has a non-null causal effect on the outcome at the 95% confidence level.
:::

---

# 9. TMLE + SuperLearner (Recommended)

```{r, eval=FALSE}
library(SuperLearner)

SL_lib <- c("SL.glm", "SL.glmnet", "SL.ranger", "SL.mean")

Q_SL <- SuperLearner(
  Y = dat$Y,
  X = dat %>% select(A, age, cvd),
  family = binomial(),
  SL.library = SL_lib
)

g_SL <- SuperLearner(
  Y = dat$A,
  X = dat %>% select(age, cvd),
  family = binomial(),
  SL.library = SL_lib
)
```

You can plug these flexible nuisance models directly into TMLE. SuperLearner builds an ensemble (weighted combination) of multiple algorithms, reducing the risk that any single model is badly misspecified. When combined with TMLE's fluctuation step, this provides a powerful, data-adaptive approach to causal inference that maintains valid statistical inference.

---

# 10. Comparison: AIPW vs TMLE

### AIPW vs TMLE: Head-to-Head

| Property | AIPW | TMLE |
|---|---|---|
| Doubly robust | Yes | Yes |
| Semiparametrically efficient | Yes (asymptotically) | Yes |
| Handles ML well | Requires cross-fitting | Yes, with SuperLearner |
| Predictions bounded in $[0,1]$ | No | Yes |
| Solves EIC equation | Not guaranteed | Yes, by construction |
| Implementation complexity | Simple | Moderate |

**Key point.** Both AIPW and TMLE are valid doubly robust estimators. TMLE is generally preferred because it respects the bounds of the parameter space (e.g., probabilities stay in $[0,1]$), integrates naturally with ensemble machine learning, and solves the efficient influence curve equation by construction, giving it better finite-sample performance and more trustworthy inference.

---

# Running Example: Antiretroviral Therapy Cohort

In this section we apply TMLE to the HIV antiretroviral therapy cohort that serves as the running example throughout this handbook. The goal is to estimate the causal effect of ever initiating HAART (`ever_haart`) on mortality (`died`), adjusting for baseline covariates `age`, `sex`, and `cd4_sqrt_baseline`.

This is a **point-treatment** analysis using a single baseline snapshot per patient. Longitudinal TMLE, which accounts for time-varying treatment and confounding in the full panel data, is covered in Chapter 3.9.

```{r}
source("_data/load-haart-cohort.R")

glimpse(haart_baseline)
```

### Step 1: Fit the initial outcome model

```{r}
Q_fit <- glm(died ~ ever_haart + age + sex + cd4_sqrt_baseline,
             family = binomial, data = haart_baseline)

haart_baseline <- haart_baseline |>
  mutate(
    Q1 = predict(Q_fit,
                 newdata = mutate(haart_baseline, ever_haart = 1),
                 type = "response"),
    Q0 = predict(Q_fit,
                 newdata = mutate(haart_baseline, ever_haart = 0),
                 type = "response")
  )
```

### Step 2: Fit the propensity score model

```{r}
g_fit <- glm(ever_haart ~ age + sex + cd4_sqrt_baseline,
             family = binomial, data = haart_baseline)

haart_baseline <- haart_baseline |>
  mutate(ps = predict(g_fit, type = "response"))

# Check overlap
summary(haart_baseline$ps)
```

### Step 3: Construct the clever covariate and fluctuate

```{r}
logit <- function(p) log(p / (1 - p))

Q_init <- predict(Q_fit, type = "response")
H_clever <- with(haart_baseline, ever_haart / ps - (1 - ever_haart) / (1 - ps))

eps <- coef(glm(haart_baseline$died ~ -1 + offset(logit(Q_init)) + H_clever,
                family = binomial))
```

### Step 4: Update predictions and compute the ATE

```{r}
H1 <-  1 / haart_baseline$ps
H0 <- -1 / (1 - haart_baseline$ps)

Q1_star <- plogis(logit(haart_baseline$Q1) + eps * H1)
Q0_star <- plogis(logit(haart_baseline$Q0) + eps * H0)

ate <- mean(Q1_star) - mean(Q0_star)

# EIC-based standard error
n_obs <- nrow(haart_baseline)
Qstar_obs <- ifelse(haart_baseline$ever_haart == 1, Q1_star, Q0_star)
H_obs     <- ifelse(haart_baseline$ever_haart == 1, H1, H0)
EIC <- H_obs * (haart_baseline$died - Qstar_obs) +
       Q1_star - Q0_star - ate

se  <- sqrt(var(EIC) / n_obs)

tibble(
  ATE      = round(ate, 4),
  SE       = round(se, 4),
  CI_lower = round(ate - 1.96 * se, 4),
  CI_upper = round(ate + 1.96 * se, 4)
)
```

:::{.callout-note}
## Interpreting the HAART TMLE Result

A positive ATE would suggest that HAART initiation increases the probability of death in this simple point-treatment model. However, this likely reflects **confounding by indication**: patients with more advanced disease (lower CD4 counts) are both more likely to initiate treatment and more likely to die. The baseline snapshot cannot fully capture the time-varying nature of CD4 decline, treatment initiation, and mortality. Chapter 3.9 addresses this with longitudinal TMLE and marginal structural models that properly handle time-varying confounding.
:::

---

# 11. Summary

:::{.callout-tip}
## Key Takeaways

- Doubly robust estimators (AIPW, TMLE) combine outcome and treatment models and are consistent if **either** model is correct
- TMLE is preferred: bounded predictions, plug-in estimation, natural ML integration, and EIC-based inference
- Doubly robust estimators are the recommended default for point-treatment causal inference in modern epidemiology
:::

```{r}
sessionInfo()
```

---

## Check your understanding

<details>
<summary><strong>Q1. What does "doubly robust" mean in practice? Does it mean you can ignore model specification entirely?</strong></summary>

No. "Doubly robust" means the estimator is consistent if *either* the treatment model or the outcome model is correctly specified, but not necessarily if both are wrong. It does not give you license to ignore model specification. If both nuisance models are badly misspecified, the estimator can be biased. In practice, you should still try to specify both models as carefully as possible. The double robustness property is an insurance policy, not a substitute for thoughtful modeling.

</details>

<details>
<summary><strong>Q2. Why does TMLE use a "clever covariate" derived from the propensity score, and what happens if propensity scores are extreme?</strong></summary>

The clever covariate $H(A, W) = A / e(W) - (1 - A) / (1 - e(W))$ is derived from the efficient influence curve for the ATE. It directs the fluctuation of the initial outcome model toward the specific causal parameter of interest, ensuring that the final estimate solves the EIC estimating equation. If propensity scores are extreme (near 0 or 1), the clever covariate becomes very large, which can destabilize the fluctuation step. This is why practical positivity (adequate overlap in propensity score distributions) is important for TMLE just as it is for IPTW.

</details>

<details>
<summary><strong>Q3. Your TMLE estimate of a risk probability is 0.85. Could AIPW have produced an estimate above 1.0 for the same data?</strong></summary>

Yes. AIPW is a linear estimator that does not enforce the bounds of the parameter space. Because it adds a weighted residual correction to the outcome-model prediction, the resulting estimate can fall outside $[0, 1]$, especially when propensity scores are extreme and the correction term is large. TMLE avoids this by performing its update on the logistic (log-odds) scale and then back-transforming, which guarantees that predicted probabilities remain in $[0, 1]$. This is one of the key practical advantages of TMLE over AIPW.

</details>

---

## What comes next

This chapter presented the theory and mechanics of AIPW and TMLE. Both estimators rely on nuisance models, and the quality of those models directly affects performance. The next chapter introduces Super Learner, the ensemble machine learning tool that makes nuisance estimation flexible and data-adaptive.

---

## Software Implementation (R)

This example demonstrates the **double robustness** property using the `tmle` package. We deliberately misspecify one nuisance model at a time to show that the TMLE estimate remains consistent as long as at least one model is correct.

- Simulate data with a nonlinear confounder–outcome relationship
- Scenario A: correct $Q$, wrong $g$ → TMLE consistent
- Scenario B: wrong $Q$, correct $g$ → TMLE consistent
- Scenario C: both correct → TMLE efficient
- Falls back to a manual demonstration if `tmle` is unavailable

```{r}
#| eval: false
set.seed(1)
n <- 1000
W <- rnorm(n)
A <- rbinom(n, 1, plogis(0.3 * W + 0.2 * W^2))
Y <- 0.5 * A + sin(W) + rnorm(n, sd = 0.3)  # true ATE = 0.5

if (requireNamespace("tmle", quietly = TRUE)) {
  library(tmle)

  ## Scenario A: correct Q model, misspecified g (intercept only)
  fit_a <- tmle(Y = Y, A = A, W = data.frame(W = W),
                Q.SL.library = c("SL.glm", "SL.earth"),
                g.SL.library  = "SL.mean")
  cat("Scenario A (correct Q, wrong g):", round(fit_a$estimates$ATE$psi, 3), "\n")

  ## Scenario B: misspecified Q (intercept only), correct g
  fit_b <- tmle(Y = Y, A = A, W = data.frame(W = W),
                Q.SL.library = "SL.mean",
                g.SL.library  = c("SL.glm", "SL.earth"))
  cat("Scenario B (wrong Q, correct g):", round(fit_b$estimates$ATE$psi, 3), "\n")

  ## Scenario C: both reasonably specified
  fit_c <- tmle(Y = Y, A = A, W = data.frame(W = W),
                Q.SL.library = c("SL.glm", "SL.earth"),
                g.SL.library  = c("SL.glm", "SL.earth"))
  cat("Scenario C (both correct):      ", round(fit_c$estimates$ATE$psi, 3), "\n")
  cat("95% CI:", round(fit_c$estimates$ATE$CI, 3), "\n")
} else {
  message("Install the 'tmle' package:  install.packages('tmle')")
}
```
