---
title: "Chapter 2.1: Outcome Modeling and Standardization (G-Computation)"
format: html
---

# Chapter 2.1: Outcome Modeling and Standardization  
*G-computation as a foundation for causal estimation*

Outcome modeling and standardization, often referred to as **g-computation**, is one of the oldest and most intuitive approaches to causal inference. In this chapter, we'll build intuition, walk carefully through why the method works, show where it fails, and provide fully reproducible examples in R (using tidyverse style).  

This chapter is intentionally thorough, designed for students new to causal inference but with working knowledge of regression.

---

:::{.callout-tip}
## Why Start With G-Computation?

G-computation estimates causal effects by predicting what *would* happen to everyone under each treatment, then comparing those predictions. No reliance on a single regression coefficient required. It directly implements the identification formula from Chapter 1.3 ($E[Y(a)] = E_W[E[Y \mid A=a, W]]$), making it the most concrete entry point for learning causal estimation.
:::

**Key point.** G-computation reconstructs what would have happened if *everyone* received treatment ($A=1$) and, separately, if *everyone* received control ($A=0$). The recipe is three steps: (1) fit an outcome model $E[Y \mid A, W]$, (2) predict outcomes for each individual under both treatments, and (3) average those predictions (standardization). That is the entire method.

# 3. Implementation Example: Simulated Osteoporosis Cohort

We simulate a small cohort similar to the denosumab vs zoledronic acid motivating example.

```{r}
library(tidyverse)

set.seed(2025)
n <- 3000

# baseline covariates
age <- rnorm(n, 75, 6)
cvd <- rbinom(n, 1, plogis(0.12 * (age - 70)))

# treatment assignment
A <- rbinom(n, 1, plogis(-1 + 0.07 * (age - 70) + 1.4 * cvd))

# outcome
Y <- rbinom(n, 1, plogis(-2 + 0.6*A + 0.12*(age - 70) + 1.2*cvd))

dat <- tibble(age, cvd, A, Y)
```

The treatment is strongly confounded by cardiovascular history, making this a good scenario for demonstrating g-computation.

---

# 4. Step-by-Step G-Computation

### Step 1: Fit an Outcome Model

We fit a logistic regression model predicting the outcome from treatment and confounders:

```{r}
mod <- glm(Y ~ A + age + cvd, family = binomial, data = dat)
summary(mod)
```

**Pitfall.** The coefficient on `A` from this logistic regression is a *conditional log-odds ratio*. It tells you the association between treatment and outcome *holding confounders fixed*. That is **not** a causal effect on its own. We need the next two steps (prediction + averaging) to turn this model into a causal estimate.

### Step 2: Predict Counterfactual Outcomes (The "What If" Step)

This is where g-computation earns its name. We ask two "what if" questions for *every single person* in the dataset, regardless of what treatment they actually received: what if this person had been treated ($A=1$)? What if untreated ($A=0$)?

```{r}
# create counterfactual datasets
dat1 <- dat %>% mutate(A = 1)
dat0 <- dat %>% mutate(A = 0)

# predict potential outcomes
p1 <- predict(mod, newdata = dat1, type = "response")
p0 <- predict(mod, newdata = dat0, type = "response")
```

Here `p1[i]` and `p0[i]` are the predicted outcomes for person *i* under treatment and control, respectively. Notice we are *not* comparing treated vs. untreated groups. We are comparing two hypothetical worlds for the *same* people. This is the counterfactual reasoning that separates causal inference from plain regression.

:::{.callout-important}
## Step 3: Standardize. This Is the Step That Makes It Causal

Averaging the counterfactual predictions over the *entire population* converts model-based predictions into a **causal** estimate: a standardized (marginal) risk difference, the same quantity a perfectly executed randomized trial would yield.

```{r}
risk1 <- mean(p1)
risk0 <- mean(p0)
ate  <- risk1 - risk0

list(risk1 = risk1, risk0 = risk0, ate = ate)
```

Interpret the result as: “Initiating denosumab rather than ZA is estimated to change 3-year MI/stroke risk by X percentage points.”
:::

# 5. Why G-Computation Works

It directly implements the identification formula:

$$
E_W[ E[Y | A=a, W] ].
$$

This contrasts with regression coefficients, which estimate:

$$
	ext{conditional log-odds ratios given W}
$$

These are completely different from the marginal risk difference.

Standardization always yields marginal (population-level) effects.

:::{.callout-tip}
## Key Distinction: Marginal Risk Difference vs. Conditional Log-Odds Ratio

This is one of the most commonly confused points in epidemiology:

- **Conditional log-odds ratio** (what logistic regression gives you): "Among people with the *same* age and CVD status, how do the odds of the outcome differ between treated and untreated?" This is a measure of *association within strata*. It does not tell you what would happen if you intervened on the whole population. It is also non-collapsible, meaning it changes depending on which covariates you condition on, even without confounding.

- **Marginal risk difference** (what g-computation gives you): "If we could treat *everyone* versus *no one*, how much would the overall population risk change?" This is a *causal* contrast on the risk scale, the kind of number a clinician or policy maker can act on.

G-computation bridges the gap: it uses the conditional model internally but **standardizes** over the population to produce the marginal (causal) effect.
:::

---

# 6. Diagnostics: When Can G-Computation Fail?

:::{.callout-warning}
## Model Misspecification

G-computation rests entirely on the outcome model $E[Y \mid A, W]$. If that model is wrong (e.g., omits interactions, assumes linearity when the truth is non-linear), bias propagates directly into the causal estimate. Always check residuals and consider flexible alternatives (see Chapter 2.3).
:::

---

:::{.callout-warning}
## Poor Positivity

If some strata almost never receive a treatment, g-computation *extrapolates* into regions with little observed data. R will return a number, but it may be meaningless. Check propensity scores for values near 0 or 1:

```{r}
ps <- predict(glm(A ~ age + cvd, family = binomial, data = dat), type = "response")
summary(ps)
```
:::

---

**Pitfall.** No modeling strategy fixes missing confounders. If an important common cause of treatment and outcome is absent from your dataset, g-computation (like every other method in this handbook) will produce a biased estimate. However, g-computation makes its assumptions explicit (you can see exactly which variables $W$ enter the model), which aids transparent reporting and sensitivity analysis.

---

**Practical note.** The biggest limitation of g-computation with logistic regression is that you must get the functional form right. A powerful alternative is to replace the parametric model with a data-adaptive learner (random forests, gradient boosting, GAMs, or a SuperLearner ensemble), reducing reliance on parametric assumptions. In later chapters we systematically integrate SuperLearner with **TMLE**.

```{r, eval=F}
library(SuperLearner)

sl_mod <- SuperLearner(
  Y = dat$Y,
  X = dat %>% select(A, age, cvd),
  family = binomial(),
  SL.library = c("SL.glm", "SL.ranger", "SL.gam")
)

# predict for counterfactuals
p1_sl <- predict(sl_mod, newdata = dat1)$pred
p0_sl <- predict(sl_mod, newdata = dat0)$pred

mean(p1_sl - p0_sl)
```

---

:::{.callout-tip}
## Summary

G-computation = fit a model, predict under both treatments for everyone, then average. That three-step standardization is what converts a conditional model into a marginal causal estimate. Its main vulnerability is outcome-model misspecification; flexible learners and doubly robust methods (IPTW, TMLE) address this in subsequent chapters.
:::

---

## Check your understanding

<details>
<summary><strong>Q1. You run g-computation using a logistic regression model. A colleague asks: "Isn't this just a regression coefficient?" How do you explain the difference?</strong></summary>

No. A logistic regression coefficient is a *conditional log-odds ratio* that describes how the log-odds of the outcome change with treatment, holding covariates fixed at particular values. G-computation uses that same model but adds two critical steps: it predicts outcomes for every individual under both $A=1$ and $A=0$, then *averages* those predictions over the entire population. The result is a **marginal risk difference** (the ATE on the risk scale), which answers a different question: "How much would population-level risk change if we could treat everyone versus no one?" The regression coefficient is conditional and on the log-odds scale; the g-computation estimate is marginal and on the risk scale. These are different quantities even when the underlying model is the same.

</details>

<details>
<summary><strong>Q2. What happens to the g-computation estimate if the outcome model is misspecified?</strong></summary>

G-computation is only as good as its outcome model. If $E[Y \mid A, W]$ is misspecified (for example, it assumes a linear relationship when the truth involves interactions or non-linearities), the counterfactual predictions will be systematically wrong, and the ATE estimate will be biased. Unlike doubly robust methods (AIPW, TMLE), g-computation has no second chance: there is no treatment model to compensate for outcome-model errors. This is why diagnostics (residual checks, comparison across model specifications) and flexible modeling strategies (SuperLearner) are important when using g-computation.

</details>

<details>
<summary><strong>Q3. Why does g-computation produce a marginal (population-level) effect even though it uses a conditional model internally?</strong></summary>

The key is the standardization (averaging) step. The outcome model $E[Y \mid A, W]$ is conditional: it gives a predicted outcome for a specific individual with covariates $W=w$ under treatment $A=a$. But g-computation does not report these individual-level predictions. Instead, it predicts for *every person* under $A=1$ and under $A=0$, then takes the population average of the difference. This averaging over the empirical covariate distribution $P(W)$ is exactly the operation $E_W[E[Y \mid A=a, W]]$ from the identification formula, and it produces a marginal effect. The conditional model is an intermediate tool; the final quantity is a population-level causal contrast.

</details>

---

## What comes next

G-computation depends entirely on the outcome model being correctly specified. What if we could instead model the treatment mechanism? The next chapter introduces IPTW, which takes exactly that approach: reweighting observations to create a pseudo-population where treatment appears randomized.

---

## Software Implementation (R)

This example implements **g-computation** (parametric standardization) by hand, then shows how the `tmle` package automates and improves on this approach with targeting.

- Simulate a confounded point-treatment dataset
- Fit an outcome model $\hat{Q}(A,W) = E[Y \mid A, W]$
- Predict under $A=1$ and $A=0$ for everyone, then average (g-computation)
- Compare with the TMLE estimate, which adds a targeting step for bias correction

```{r}
#| eval: false
set.seed(1)
n <- 500
W <- rnorm(n)
A <- rbinom(n, 1, plogis(0.5 * W))
Y <- rbinom(n, 1, plogis(-1 + 0.8 * A + 1.0 * W))  # true RD ≈ 0.16

## ── Manual g-computation ──
Q_mod <- glm(Y ~ A + W, family = binomial, data = data.frame(Y, A, W))

Q1 <- predict(Q_mod, newdata = data.frame(A = 1, W = W), type = "response")
Q0 <- predict(Q_mod, newdata = data.frame(A = 0, W = W), type = "response")
gcomp_ate <- mean(Q1) - mean(Q0)
cat("G-computation ATE:", round(gcomp_ate, 3), "\n")

## ── TMLE (adds targeting to g-computation) ──
if (requireNamespace("tmle", quietly = TRUE)) {
  library(tmle)
  fit <- tmle(Y = Y, A = A, W = data.frame(W = W),
              family = "binomial",
              Q.SL.library = "SL.glm", g.SL.library = "SL.glm")
  cat("TMLE ATE:", round(fit$estimates$ATE$psi, 3), "\n")
  cat("95% CI: ", round(fit$estimates$ATE$CI, 3), "\n")
} else {
  message("Install the 'tmle' package:  install.packages('tmle')")
}
```
