---
title: "Chapter 2.5: From Question to Estimate: TMLE in Practice"
format: html
---

*Estimated reading time: 75 minutes*

# Chapter 2.5: From Question to Estimate: TMLE in Practice
*A progressive tutorial for pharmacoepidemiologists*

This chapter walks through a complete causal analysis from start to finish, building each estimator step-by-step so you can see exactly where bias enters, how each method addresses it, and why TMLE provides a principled solution.

We use a realistic post-marketing safety scenario: **evaluating the cardiovascular safety of a new diabetes drug**.

By the end of this chapter, you will have implemented:

1. A naive (unadjusted) model
2. G-computation (outcome modeling)
3. IPTW (treatment modeling)
4. AIPW (combining both)
5. TMLE (targeting the estimate)

Each method builds on what came before. Every step includes code, diagnostics, and interpretation.

::: {.callout-note}
## Learning objectives
- Implement five estimation approaches (naive, g-computation, IPTW, AIPW, TMLE) on the same dataset and compare their behavior
- Diagnose propensity score overlap and covariate balance for each method
- Explain how TMLE's targeting step corrects the initial outcome model using the efficient influence curve
- Integrate Super Learner into TMLE for flexible nuisance estimation
- Interpret the efficient influence curve and use it for inference (confidence intervals, hypothesis tests)
:::

*This chapter is educational. Causal conclusions depend on identification assumptions (e.g., consistency, exchangeability, positivity) and on diagnostic evidence that the data support the target estimand. When flexible machine learning is used for nuisance estimation, valid inference typically requires cross-fitting or a cross-validated TMLE variant, plus appropriate rate conditions.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

---

# Clinical Motivation

## The Regulatory Question

A new sodium-glucose co-transporter 2 (SGLT2) inhibitor, **dapagliflozin**, has been approved for type 2 diabetes. Post-marketing surveillance data from insurance claims suggest a potential cardiovascular safety signal. The FDA's Office of Surveillance and Epidemiology asks:

> **Does initiation of dapagliflozin (vs. a sulfonylurea comparator) increase the 1-year risk of major adverse cardiovascular events (MACE) in adults with type 2 diabetes?**

This is a classic **active comparator, new user** design question.

### Key elements

- **Target population:** Adults aged 40-85 with type 2 diabetes initiating a new oral glucose-lowering agent
- **Treatment strategies:** Initiate dapagliflozin (A = 1) vs. initiate sulfonylurea (A = 0)
- **Outcome:** 1-year MACE (composite of MI, stroke, CV death), binary
- **Intercurrent events:** Treatment discontinuation, switching, death from non-CV causes
- **Decision:** Should the label carry a cardiovascular warning? Is a formal safety trial needed?

### Why standard regression falls short

As introduced in [Chapter 1.1](01-01-regression-vs-causal.qmd), regression coefficients from observational data reflect both the treatment effect and confounding. A logistic regression of MACE on treatment will conflate the drug's effect with differences in baseline health, because patients prescribed newer, more expensive agents differ systematically from those prescribed sulfonylureas in age, comorbidity burden, prior cardiovascular history, renal function, and concomitant medications. We need methods that explicitly separate confounding from causal effects.

---

# Setup: Question, Assumptions, and Estimand

This chapter applies the Causal Roadmap from [Chapter 1.2](01-02-causal-roadmap.qmd) to the dapagliflozin safety question. We briefly state each roadmap step here; see Chapters 1.2 and 1.3 for the full framework.

**Causal question.** What would happen to 1-year MACE risk if every eligible patient started dapagliflozin, compared to if every eligible patient started a sulfonylurea?

**Estimand.** The ATE on the risk difference scale: $\psi = E[Y(1)] - E[Y(0)]$.

**Causal model.** Confounders $W$ (age, BMI, HbA1c, prior CVD, eGFR, statin use) affect both treatment $A$ and outcome $Y$. The DAG is the standard point-treatment structure: $W \to A$, $W \to Y$, $A \to Y$.

**Identification.** We assume consistency, exchangeability, and positivity as defined in [Chapter 1.2](01-02-causal-roadmap.qmd). In this application, consistency requires that "initiate dapagliflozin" is well-defined across patients. Exchangeability is the most vulnerable assumption: we cannot test it, but we strengthen it by adjusting for a rich set of confounders. Positivity can be violated when, for example, patients with severe renal impairment are never prescribed dapagliflozin.

**Statistical estimand.** Under identification, the causal ATE equals $E_W[E[Y \mid A=1, W]] - E_W[E[Y \mid A=0, W]]$.

**Estimation strategy.** We implement five estimators progressively, from naive to TMLE, so you can see exactly what each adds. The table below summarizes the trade-offs developed in Chapters 2.1 through 2.3:

| Method | Relies on | Robust to | Weakness |
|--------|-----------|-----------|----------|
| G-computation | Outcome model $E[Y \mid A, W]$ | Treatment model misspecification | Sensitive to outcome model misspecification |
| IPTW | Treatment model $P(A \mid W)$ | Outcome model misspecification | Sensitive to positivity violations and weight instability |
| AIPW | Both models | Misspecification of either one model | Not targeted; can produce out-of-bounds estimates |
| TMLE | Both models | Misspecification of either one model | More complex to implement; requires understanding the targeting step |

We now implement each estimator progressively.

---

# Data Simulation

We simulate a realistic claims-like dataset with multiple confounders, nonlinear effects, and treatment-confounder interactions.

```{webr-r}
library(tidyverse)

set.seed(2026)
n <- 5000

# --- Baseline confounders ---
age    <- rnorm(n, mean = 62, sd = 10)
bmi    <- rnorm(n, mean = 31, sd = 5)
hba1c  <- rnorm(n, mean = 8.2, sd = 1.3)
# Prior CVD: more likely with age and high BMI
cvd    <- rbinom(n, 1, plogis(-3.5 + 0.04 * age + 0.03 * bmi))
# eGFR: lower with age, higher with lower BMI
egfr   <- pmax(15, rnorm(n, mean = 95 - 0.5 * (age - 60) - 0.3 * bmi, sd = 15))
# Statin use: more likely with CVD and age
statin <- rbinom(n, 1, plogis(-1.5 + 1.8 * cvd + 0.02 * age))

# --- Treatment model (propensity score) ---
# Dapagliflozin is preferentially prescribed to:
# - younger patients, higher eGFR (renal eligibility), higher BMI, lower HbA1c
# Includes nonlinear terms and interactions
lp_trt <- -0.8 +
  -0.03 * (age - 62) +
  0.04 * (bmi - 31) +
  -0.15 * (hba1c - 8.2) +
  -0.6 * cvd +
  0.02 * (egfr - 80) +
  -0.3 * statin +
  0.01 * (age - 62) * cvd +         # interaction
  -0.002 * (age - 62)^2             # nonlinear age effect

A <- rbinom(n, 1, plogis(lp_trt))

# --- Outcome model ---
# MACE risk depends on treatment, confounders, with nonlinearities
# True treatment effect: dapagliflozin REDUCES MACE risk (protective)
lp_out <- -3.2 +
  -0.40 * A +                       # true causal effect (protective)
  0.04 * (age - 62) +
  0.02 * (bmi - 31) +
  0.10 * (hba1c - 8.2) +
  1.00 * cvd +
  -0.01 * (egfr - 80) +
  0.30 * statin +
  0.015 * (age - 62) * cvd +        # interaction
  0.0008 * (age - 62)^2 +           # nonlinear age effect
  -0.10 * A * cvd                   # treatment-confounder interaction

Y <- rbinom(n, 1, plogis(lp_out))

dat <- tibble(age, bmi, hba1c, cvd, egfr, statin, A, Y)
```

### True causal effect

We compute the true ATE from the data-generating process:

```{webr-r}
# True potential outcomes from the structural model
p1_true <- plogis(-3.2 + -0.40 * 1 + 0.04 * (age - 62) + 0.02 * (bmi - 31) +
                    0.10 * (hba1c - 8.2) + 1.00 * cvd + -0.01 * (egfr - 80) +
                    0.30 * statin + 0.015 * (age - 62) * cvd +
                    0.0008 * (age - 62)^2 + -0.10 * 1 * cvd)

p0_true <- plogis(-3.2 + -0.40 * 0 + 0.04 * (age - 62) + 0.02 * (bmi - 31) +
                    0.10 * (hba1c - 8.2) + 1.00 * cvd + -0.01 * (egfr - 80) +
                    0.30 * statin + 0.015 * (age - 62) * cvd +
                    0.0008 * (age - 62)^2 + -0.10 * 0 * cvd)

true_ate <- mean(p1_true - p0_true)
cat("True ATE (risk difference):", round(true_ate, 4), "\n")
cat("True risk under dapagliflozin:", round(mean(p1_true), 4), "\n")
cat("True risk under sulfonylurea: ", round(mean(p0_true), 4), "\n")
```

### Inspect the data

```{webr-r}
dat %>%
  group_by(A) %>%
  summarise(
    n          = n(),
    mean_age   = mean(age),
    mean_bmi   = mean(bmi),
    mean_hba1c = mean(hba1c),
    pct_cvd    = mean(cvd),
    mean_egfr  = mean(egfr),
    pct_statin = mean(statin),
    mace_rate  = mean(Y),
    .groups = "drop"
  )
```

Notice the imbalances: dapagliflozin patients are younger, have higher eGFR, and less prior CVD. These differences will bias a naive comparison.

---

# Progressive Estimation

## 4A. Naive (Unadjusted) Model: What Happens Without Adjustment?

We start with what many analysts would try first: a simple comparison of MACE rates by treatment group. This establishes the baseline from which all other methods improve.

```{webr-r}
# Unadjusted risk difference
risk_A1 <- mean(dat$Y[dat$A == 1])
risk_A0 <- mean(dat$Y[dat$A == 0])
naive_rd <- risk_A1 - risk_A0

cat("Naive risk (dapagliflozin):", round(risk_A1, 4), "\n")
cat("Naive risk (sulfonylurea): ", round(risk_A0, 4), "\n")
cat("Naive risk difference:     ", round(naive_rd, 4), "\n")
cat("True ATE:                  ", round(true_ate, 4), "\n")
```

```{webr-r}
# Unadjusted logistic regression
naive_mod <- glm(Y ~ A, family = binomial, data = dat)
broom::tidy(naive_mod, conf.int = TRUE) |> knitr::kable(digits = 3, caption = "Unadjusted logistic regression coefficients")
```

:::{.callout-warning}
## What Went Wrong?

The naive estimate is biased because dapagliflozin patients differ systematically from sulfonylurea patients. The treatment groups are not exchangeable (see [Chapter 1.2](01-02-causal-roadmap.qmd)), so the naive comparison conflates the drug effect with confounding by indication. The difference in MACE rates reflects both the drug's effect AND the differences in who gets each drug. This is why unadjusted treatment comparisons cannot be interpreted as causal effects in observational data.
:::

---

## 4B. G-Computation (Outcome Modeling)

G-computation directly implements the identification formula by modeling $E[Y \mid A, W]$, then standardizing.

**Step 1: Fit the Outcome Model.** Fit a model for $E[Y \mid A, W]$, the expected outcome given treatment and confounders. This model will be used to predict what would happen to each patient under each treatment scenario.

```{webr-r}
# Parametric outcome model (logistic regression)
# We include main effects and key interactions
q_mod <- glm(Y ~ A + age + bmi + hba1c + cvd + egfr + statin +
               A:cvd + I(age^2),
             family = binomial, data = dat)
```

**Step 2: Predict Counterfactual Outcomes.** Ask: "What would this patient's outcome be if they received dapagliflozin? And if they received sulfonylurea?" We generate these predictions for *every* patient, regardless of what they actually received.

```{webr-r}
# Create counterfactual datasets
dat1 <- dat %>% mutate(A = 1)
dat0 <- dat %>% mutate(A = 0)

# Predict potential outcomes for each individual
Q1 <- predict(q_mod, newdata = dat1, type = "response")
Q0 <- predict(q_mod, newdata = dat0, type = "response")
```

**Step 3: Standardize (Average Over the Population).** Average the predicted outcomes across all patients. This implements the G-computation formula: $E_W[E[Y \mid A=a, W]]$.

```{webr-r}
gcomp_risk1 <- mean(Q1)
gcomp_risk0 <- mean(Q0)
gcomp_ate   <- gcomp_risk1 - gcomp_risk0

cat("G-comp risk (dapagliflozin):", round(gcomp_risk1, 4), "\n")
cat("G-comp risk (sulfonylurea): ", round(gcomp_risk0, 4), "\n")
cat("G-comp ATE:                 ", round(gcomp_ate, 4), "\n")
cat("True ATE:                   ", round(true_ate, 4), "\n")
```

### Bootstrap confidence interval for G-computation

```{webr-r}
set.seed(42)
n_boot <- 500
boot_ate <- numeric(n_boot)

for (b in 1:n_boot) {
  idx <- sample(1:n, n, replace = TRUE)
  dat_b <- dat[idx, ]
  mod_b <- glm(Y ~ A + age + bmi + hba1c + cvd + egfr + statin +
                  A:cvd + I(age^2),
                family = binomial, data = dat_b)
  Q1_b <- predict(mod_b, newdata = dat_b %>% mutate(A = 1), type = "response")
  Q0_b <- predict(mod_b, newdata = dat_b %>% mutate(A = 0), type = "response")
  boot_ate[b] <- mean(Q1_b - Q0_b)
}

gcomp_se <- sd(boot_ate)
gcomp_ci <- quantile(boot_ate, c(0.025, 0.975))

cat("G-comp ATE:", round(gcomp_ate, 4),
    " SE:", round(gcomp_se, 4),
    " 95% CI: [", round(gcomp_ci[1], 4), ",", round(gcomp_ci[2], 4), "]\n")
```

**Pitfall.** G-computation gave a much better estimate than the naive approach by adjusting for confounders. It works by predicting what would happen to each patient under each treatment, then averaging. But G-computation depends entirely on the outcome model being correct. If we misspecified the functional form (missed interactions, nonlinearities), the estimate would be biased, and we would have no way to know. This motivates IPTW, which adjusts for confounding using the *treatment* model instead.

---

## 4C. IPTW (Treatment Modeling)

IPTW takes a different approach: instead of modeling the outcome, it models treatment assignment and reweights the data to create a pseudo-population where confounders are balanced.

**Step 1: Estimate the Propensity Score.** The propensity score $g(W) = P(A = 1 \mid W)$ answers: given a patient's covariates, how likely were they to receive dapagliflozin? We use this to reweight the data so that treatment looks "as-if-randomized."

```{webr-r}
# Propensity score model
g_mod <- glm(A ~ age + bmi + hba1c + cvd + egfr + statin +
               I(age^2) + age:cvd,
             family = binomial, data = dat)

dat <- dat %>%
  mutate(ps = predict(g_mod, type = "response"))

# Summary of propensity scores
summary(dat$ps)
```

### Step 2: Assess propensity score overlap

```{webr-r}
ggplot(dat, aes(x = ps, fill = factor(A, labels = c("Sulfonylurea", "Dapagliflozin")))) +
  geom_density(alpha = 0.45) +
  labs(
    x = "Estimated propensity score P(A=1 | W)",
    y = "Density",
    fill = "Treatment",
    title = "Propensity Score Overlap"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#4477AA", "#EE6677"))
```

Good overlap means every type of patient has some chance of getting either drug. If the distributions barely overlap, IPTW will produce extreme weights and unstable estimates.

### Step 3: Construct stabilized weights

```{webr-r}
# Marginal treatment probability
p_A <- mean(dat$A)

dat <- dat %>%
  mutate(
    # Unstabilized weights
    w_unstab = ifelse(A == 1, 1 / ps, 1 / (1 - ps)),
    # Stabilized weights
    sw = ifelse(A == 1, p_A / ps, (1 - p_A) / (1 - ps))
  )
```

### Step 4: Diagnose weights

```{webr-r}
# Weight diagnostics
dat %>%
  group_by(A) %>%
  summarise(
    min_w  = min(sw),
    p01_w  = quantile(sw, 0.01),
    median = median(sw),
    p99_w  = quantile(sw, 0.99),
    max_w  = max(sw),
    mean_w = mean(sw),
    sd_w   = sd(sw),
    .groups = "drop"
  )
```

```{webr-r}
# Weight distribution plot
ggplot(dat, aes(x = sw, fill = factor(A, labels = c("Sulfonylurea", "Dapagliflozin")))) +
  geom_histogram(bins = 50, alpha = 0.6, position = "identity") +
  labs(
    x = "Stabilized IPTW weight",
    y = "Count",
    fill = "Treatment",
    title = "Distribution of Stabilized IPTW Weights"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#4477AA", "#EE6677")) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "grey40")
```

**What to look for:**

- Weights should be centered around 1 (for stabilized weights)
- Very large weights (> 10) indicate positivity problems
- The mean of stabilized weights should be approximately 1

### Step 5: Check covariate balance after weighting

```{webr-r}
# Standardized mean differences: unweighted vs weighted
compute_smd <- function(data, var, trt, weights = NULL) {
  if (is.null(weights)) weights <- rep(1, nrow(data))
  d1 <- data[[var]][data[[trt]] == 1]
  d0 <- data[[var]][data[[trt]] == 0]
  w1 <- weights[data[[trt]] == 1]
  w0 <- weights[data[[trt]] == 0]
  m1 <- weighted.mean(d1, w1)
  m0 <- weighted.mean(d0, w0)
  s1 <- sqrt(sum(w1 * (d1 - m1)^2) / sum(w1))
  s0 <- sqrt(sum(w0 * (d0 - m0)^2) / sum(w0))
  pooled_sd <- sqrt((s1^2 + s0^2) / 2)
  (m1 - m0) / pooled_sd
}

covariates <- c("age", "bmi", "hba1c", "cvd", "egfr", "statin")

smd_raw <- sapply(covariates, function(v) compute_smd(dat, v, "A"))
smd_wt  <- sapply(covariates, function(v) compute_smd(dat, v, "A", dat$sw))

smd_df <- tibble(
  variable = covariates,
  Unadjusted = abs(smd_raw),
  Weighted   = abs(smd_wt)
) %>%
  pivot_longer(-variable, names_to = "Method", values_to = "SMD")

ggplot(smd_df, aes(x = SMD, y = reorder(variable, SMD), color = Method, shape = Method)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "grey50") +
  labs(
    x = "Absolute Standardized Mean Difference",
    y = "",
    title = "Covariate Balance: Before and After IPTW"
  ) +
  theme_minimal() +
  scale_color_manual(values = c("#EE6677", "#228833"))
```

The dashed line at 0.1 is a common threshold. After weighting, all covariates should be below this line.

### Step 6: Estimate the ATE

```{webr-r}
# Hajek (ratio) estimator with stabilized weights
risk1_iptw <- with(dat, sum(sw * Y * (A == 1)) / sum(sw * (A == 1)))
risk0_iptw <- with(dat, sum(sw * Y * (A == 0)) / sum(sw * (A == 0)))
iptw_ate   <- risk1_iptw - risk0_iptw

cat("IPTW risk (dapagliflozin):", round(risk1_iptw, 4), "\n")
cat("IPTW risk (sulfonylurea): ", round(risk0_iptw, 4), "\n")
cat("IPTW ATE:                 ", round(iptw_ate, 4), "\n")
cat("True ATE:                 ", round(true_ate, 4), "\n")
```

### Weight truncation

If weights are extreme, truncation can improve stability at the cost of a small amount of bias:

```{webr-r}
# Truncate at 1st and 99th percentiles
trunc_lower <- quantile(dat$sw, 0.01)
trunc_upper <- quantile(dat$sw, 0.99)

dat <- dat %>%
  mutate(sw_trunc = pmin(pmax(sw, trunc_lower), trunc_upper))

risk1_trunc <- with(dat, sum(sw_trunc * Y * (A == 1)) / sum(sw_trunc * (A == 1)))
risk0_trunc <- with(dat, sum(sw_trunc * Y * (A == 0)) / sum(sw_trunc * (A == 0)))
iptw_ate_trunc <- risk1_trunc - risk0_trunc

cat("IPTW ATE (truncated weights):", round(iptw_ate_trunc, 4), "\n")
```

**Pitfall.** IPTW adjusts for confounding through the treatment model alone. It does not use information about the outcome-confounder relationship. If the propensity score model is wrong, IPTW is biased. If positivity is limited, weights become extreme and the estimate becomes unstable. This motivates doubly robust methods, which use BOTH the outcome model and the treatment model, so that if either one is correct, the estimate is consistent.

---

## 4D. AIPW (Augmented Inverse Probability Weighting)

:::{.callout-tip}
## The Doubly Robust Idea

What if we could use BOTH the outcome model and the treatment model, and as long as EITHER one is correct, get a consistent estimate? This is exactly what AIPW (and TMLE) provide.
:::

AIPW combines the outcome model and the treatment model. It is **doubly robust**: consistent if either model is correct. The formula starts with the G-computation prediction and adds a weighted residual correction from the propensity score. If the outcome model is already good, the correction is small; if the outcome model is poor but the propensity score is accurate, the correction rescues the estimate.

```{webr-r}
# Using models from previous steps:
# Q1, Q0 = predicted outcomes from g-computation
# ps     = propensity scores from IPTW

# AIPW components
mu1_aipw <- with(dat, mean(
  Q1 + A * (Y - Q1) / ps
))
mu0_aipw <- with(dat, mean(
  Q0 + (1 - A) * (Y - Q0) / (1 - ps)
))
aipw_ate <- mu1_aipw - mu0_aipw

cat("AIPW risk (dapagliflozin):", round(mu1_aipw, 4), "\n")
cat("AIPW risk (sulfonylurea): ", round(mu0_aipw, 4), "\n")
cat("AIPW ATE:                 ", round(aipw_ate, 4), "\n")
cat("True ATE:                 ", round(true_ate, 4), "\n")
```

### Influence curve and inference

The efficient influence curve (EIC) for the ATE gives us individual-level "scores" that measure each observation's contribution to the estimate. The variance of these scores provides valid standard errors.

```{webr-r}
# Efficient influence curve for ATE
eic <- with(dat,
  (A / ps) * (Y - Q1) + Q1 - mu1_aipw -
  ((1 - A) / (1 - ps)) * (Y - Q0) - Q0 + mu0_aipw
)

aipw_se <- sqrt(var(eic) / n)
aipw_ci <- aipw_ate + c(-1.96, 1.96) * aipw_se

cat("AIPW ATE:", round(aipw_ate, 4),
    " SE:", round(aipw_se, 4),
    " 95% CI: [", round(aipw_ci[1], 4), ",", round(aipw_ci[2], 4), "]\n")
```

**Key insight:** AIPW improves on both g-computation and IPTW by using both models. However, AIPW does not "target" its initial estimates. It can produce predicted probabilities outside [0, 1], and it does not solve the efficient influence curve equation exactly when machine learning is used. TMLE addresses both issues.

---

## 4E. TMLE (Targeted Maximum Likelihood Estimation)

:::{.callout-tip}
## Why TMLE?

TMLE combines the best properties of all previous methods: it uses both outcome and treatment models (doubly robust), targets the specific estimand of interest, respects natural parameter bounds (probabilities stay in [0, 1]), solves the efficient influence curve equation for valid inference, and integrates naturally with machine learning (Super Learner).
:::

### Conceptual overview

:::{.callout-note}
## TMLE in Three Stages

1. **Initial estimate:** Fit an outcome model $\hat{Q}^0(A, W) = \hat{E}[Y \mid A, W]$ (like g-computation)
2. **Targeting step:** Update $\hat{Q}^0$ using information from the propensity score to reduce bias for the specific estimand
3. **Substitution estimate:** Plug the updated $\hat{Q}^*$ into the G-computation formula

The targeting step is what makes TMLE special. It uses the **clever covariate**, a function of the propensity score, to fluctuate the initial outcome model in the direction that solves the efficient influence curve equation.
:::

### Step-by-step implementation

**Step 1: Initial Outcome Model.** This is the same as G-computation: fit $E[Y \mid A, W]$. Think of this as your "first draft" of the outcome model. It does not need to be perfect; the targeting step will correct it.

```{webr-r}
# Same as g-computation: fit E[Y | A, W]
q_init <- glm(Y ~ A + age + bmi + hba1c + cvd + egfr + statin +
                A:cvd + I(age^2),
              family = binomial, data = dat)

# Initial predictions
Q1_init <- predict(q_init, newdata = dat %>% mutate(A = 1), type = "response")
Q0_init <- predict(q_init, newdata = dat %>% mutate(A = 0), type = "response")
QA_init <- predict(q_init, type = "response")  # predictions at observed A
```

**Step 2: Estimate the Propensity Score.** Same as IPTW: model the probability of treatment given confounders. We bound the scores away from 0 and 1 to prevent numerical instability.

```{webr-r}
# Same model as IPTW
g_fit <- glm(A ~ age + bmi + hba1c + cvd + egfr + statin +
               I(age^2) + age:cvd,
             family = binomial, data = dat)

ps_tmle <- predict(g_fit, type = "response")

# Bound propensity scores away from 0 and 1 for stability
ps_tmle <- pmax(0.01, pmin(0.99, ps_tmle))
```

:::{.callout-note}
#### Step 3: Compute the Clever Covariate

The clever covariate $H(A, W)$ is the bridge between the treatment model and the outcome model. It tells the targeting step *which observations are most informative* about the causal effect.
:::

```{webr-r}
# Clever covariate for each observation at their OBSERVED treatment
H_A <- dat$A / ps_tmle - (1 - dat$A) / (1 - ps_tmle)

# Clever covariate under A=1 and A=0 (for prediction)
H_1 <- 1 / ps_tmle
H_0 <- -1 / (1 - ps_tmle)
```

The clever covariate is large when a patient's treatment was unlikely given their covariates, which are exactly the patients who are most informative about the causal effect.

:::{.callout-important}
#### Step 4: Fluctuation (The Targeting Step)

This is **the key step** that makes TMLE different from every other estimator. We fit a logistic regression of $Y$ on $H(A, W)$ with the initial $\hat{Q}^0$ as an offset. The coefficient $\epsilon$ tells us how much to "nudge" the initial estimate toward solving the efficient influence curve equation.
:::

```{webr-r}
logit <- function(p) log(p / (1 - p))

# Fluctuation model: logit(Y) ~ offset(logit(Q_init)) + H
fluc <- glm(Y ~ -1 + offset(logit(QA_init)) + H_A,
            family = binomial, data = dat)

epsilon <- coef(fluc)
cat("Epsilon (fluctuation parameter):", round(epsilon, 5), "\n")
```

**Key point.** A small $\epsilon$ means the initial model was already close to solving the EIC equation. A large $\epsilon$ means the targeting step made a substantial correction.

**Step 5: Update Predictions.** Apply the fluctuation to get the **targeted** predictions. Because we use the inverse-logit function (`plogis`), predictions are guaranteed to stay in [0, 1], the natural bounds for a probability.

```{webr-r}
# Apply the fluctuation to counterfactual predictions
Q1_star <- plogis(logit(Q1_init) + epsilon * H_1)
Q0_star <- plogis(logit(Q0_init) + epsilon * H_0)
```

**Step 6: Compute the TMLE Estimate.** Average the targeted predictions across the population, just like the final step of G-computation, but now using the **updated** $\hat{Q}^*$ that has been targeted for our specific estimand.

```{webr-r}
tmle_risk1 <- mean(Q1_star)
tmle_risk0 <- mean(Q0_star)
tmle_ate   <- tmle_risk1 - tmle_risk0

cat("TMLE risk (dapagliflozin):", round(tmle_risk1, 4), "\n")
cat("TMLE risk (sulfonylurea): ", round(tmle_risk0, 4), "\n")
cat("TMLE ATE:                 ", round(tmle_ate, 4), "\n")
cat("True ATE:                 ", round(true_ate, 4), "\n")
```

:::{.callout-important}
#### Step 7: Inference via the Efficient Influence Curve

The EIC gives each observation a "score" measuring its contribution to the estimate. The variance of these scores provides asymptotically valid standard errors, with no bootstrap needed.
:::

```{webr-r}
# EIC evaluated at the TMLE estimates
eic_tmle <- with(dat,
  (A / ps_tmle) * (Y - Q1_star) + Q1_star - tmle_risk1 -
  ((1 - A) / (1 - ps_tmle)) * (Y - Q0_star) - Q0_star + tmle_risk0
)

tmle_se <- sqrt(var(eic_tmle) / n)
tmle_ci <- tmle_ate + c(-1.96, 1.96) * tmle_se

cat("TMLE ATE:", round(tmle_ate, 4),
    " SE:", round(tmle_se, 4),
    " 95% CI: [", round(tmle_ci[1], 4), ",", round(tmle_ci[2], 4), "]\n")
```

### Verify: Is the EIC Mean Zero?

**Key point.** A correctly implemented TMLE solves the efficient influence curve equation, meaning the sample mean of the EIC should be approximately zero. This is how you know the targeting step worked correctly.

```{webr-r}
cat("Mean of EIC:", round(mean(eic_tmle), 8), "\n")
cat("(Should be very close to zero)\n")
```

---

**Putting It All Together.** We now have five estimators of the same causal quantity. Let us see how they compare. If our models are reasonable, the doubly robust methods (AIPW and TMLE) should be closest to the truth.

# Comparing All Estimators

```{webr-r}
results <- tibble(
  Method  = c("Naive", "G-computation", "IPTW", "IPTW (truncated)", "AIPW", "TMLE"),
  ATE     = c(naive_rd, gcomp_ate, iptw_ate, iptw_ate_trunc, aipw_ate, tmle_ate),
  SE      = c(NA, gcomp_se, NA, NA, aipw_se, tmle_se),
  CI_low  = c(NA, gcomp_ci[1], NA, NA, aipw_ci[1], tmle_ci[1]),
  CI_high = c(NA, gcomp_ci[2], NA, NA, aipw_ci[2], tmle_ci[2])
)

results %>%
  mutate(across(where(is.numeric), ~round(., 4)))
```

```{webr-r}
# Visualize the comparison
plot_df <- results %>%
  filter(!is.na(SE)) %>%
  mutate(Method = factor(Method, levels = c("G-computation", "AIPW", "TMLE")))

ggplot(plot_df, aes(x = ATE, y = Method)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = CI_low, xmax = CI_high), height = 0.2) +
  geom_vline(xintercept = true_ate, linetype = "dashed", color = "red") +
  annotate("text", x = true_ate, y = 0.5, label = "True ATE",
           hjust = -0.1, color = "red", size = 3.5) +
  labs(
    x = "Estimated ATE (Risk Difference)",
    y = "",
    title = "Comparison of Causal Estimators",
    subtitle = "Red dashed line = true causal effect"
  ) +
  theme_minimal()
```

---

# Diagnostics

:::{.callout-warning}
## Diagnostics Are Non-Negotiable

Good estimation is nothing without good diagnostics. **Never report a causal estimate without checking these diagnostics.** They tell you whether your assumptions are plausible and whether the estimator is well-behaved.
:::

### Propensity Score Overlap

If the propensity score distributions for treated and untreated overlap well, the positivity assumption is plausible. **Gaps** indicate regions where one treatment is rarely or never prescribed. All methods will struggle there, but IPTW is especially vulnerable.

```{webr-r}
ggplot(dat, aes(x = ps_tmle, fill = factor(A, labels = c("Sulfonylurea", "Dapagliflozin")))) +
  geom_histogram(bins = 40, alpha = 0.55, position = "identity") +
  labs(
    x = "Propensity score",
    y = "Count",
    fill = "Treatment",
    title = "Propensity Score Distribution by Treatment Group"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#4477AA", "#EE6677"))
```

### Weight Distribution

**Key point.** Stabilized weights should center around 1. Very large weights (> 10) indicate practical positivity violations, meaning patients who almost never receive their observed treatment. These patients dominate IPTW estimates and destabilize the clever covariate in TMLE.

```{webr-r}
cat("--- Unstabilized weight summary ---\n")
summary(dat$w_unstab)
cat("\n--- Stabilized weight summary ---\n")
summary(dat$sw)
cat("\nProportion of stabilized weights > 5:", mean(dat$sw > 5), "\n")
cat("Proportion of stabilized weights > 10:", mean(dat$sw > 10), "\n")
```

:::{.callout-important}
## Clever Covariate Diagnostics

The clever covariate drives the TMLE targeting step. Extreme values indicate potential instability. If $H(A, W)$ has heavy tails, the targeting step may overfit to a few influential observations.
:::

```{webr-r}
clever_cov_df <- tibble(
  H = H_A,
  treatment = factor(dat$A, labels = c("Sulfonylurea", "Dapagliflozin"))
)

ggplot(clever_cov_df, aes(x = H, fill = treatment)) +
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  labs(
    x = "Clever covariate H(A, W)",
    y = "Count",
    fill = "Treatment",
    title = "Distribution of the Clever Covariate"
  ) +
  theme_minimal() +
  scale_fill_manual(values = c("#4477AA", "#EE6677"))
```

:::{.callout-warning}
## Influence Curve Diagnostics

The EIC values reveal **influential observations**. A well-behaved TMLE has EIC values centered near zero with no extreme outliers. If you see long tails or a non-zero mean, something may be wrong with the targeting step or the propensity score model.
:::

```{webr-r}
eic_df <- tibble(eic = eic_tmle)

ggplot(eic_df, aes(x = eic)) +
  geom_histogram(bins = 50, fill = "#228833", alpha = 0.7) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x = "Efficient influence curve value",
    y = "Count",
    title = "Distribution of TMLE Influence Curve Values"
  ) +
  theme_minimal()

cat("EIC summary:\n")
summary(eic_tmle)
cat("EIC mean:", round(mean(eic_tmle), 6), "\n")
```

### Extreme Prediction Check

**Key point.** One of TMLE's advantages: because the targeting step uses a logistic fluctuation, predictions are guaranteed to stay in [0, 1]. Compare the initial and targeted ranges. The targeting step may shift predictions, but never outside the natural bounds.

```{webr-r}
cat("--- Predicted outcome range (initial model) ---\n")
cat("Q1 range:", round(range(Q1_init), 4), "\n")
cat("Q0 range:", round(range(Q0_init), 4), "\n")
cat("\n--- Predicted outcome range (after targeting) ---\n")
cat("Q1* range:", round(range(Q1_star), 4), "\n")
cat("Q0* range:", round(range(Q0_star), 4), "\n")
```

---

# Interpretation

### Translating to Regulatory Language

**Key point.** Based on the TMLE analysis:

- **Point estimate:** Dapagliflozin is estimated to reduce 1-year MACE risk by several percentage points compared to sulfonylurea (see the results table above for exact values)
- **Confidence interval:** The 95% CI from the EIC-based variance estimate is reported in the results table above
- **Risk difference vs. odds ratio:** We report on the risk difference scale because it is directly interpretable. An odds ratio would obscure the magnitude of absolute risk change, which is what matters for labeling and clinical decisions.

**Regulatory Decision.** If the confidence interval excludes zero and the point estimate is negative (protective), the data do not support adding a cardiovascular warning. If the CI includes zero, the evidence is inconclusive and further monitoring or a dedicated outcomes trial may be warranted.

:::{.callout-warning}
## Sensitivity to Assumptions: Which Are Most Fragile?

1. **No unmeasured confounding (exchangeability):** This is always the weakest link in observational data. Lifestyle factors (diet, exercise, smoking) may confound and are poorly captured in claims data.

2. **Positivity:** If some patient types never receive one drug (e.g., patients with very low eGFR are ineligible for SGLT2 inhibitors), the ATE is not identifiable for the full population. Consider restricting to a subpopulation where positivity holds.

3. **Consistency:** If treatment initiation is not well-defined (e.g., varying doses), the potential outcomes are ambiguous.

**What if unmeasured confounding exists?** A sensitivity analysis (E-value or bias analysis) can quantify how strong unmeasured confounding would need to be to explain away the observed effect.
:::

---

# Why Super Learner Improves Nuisance Estimation

In the examples above, we used parametric logistic regression for both the outcome and treatment models. In practice, these models may be misspecified: missing interactions, wrong functional forms, or overly rigid assumptions. **Super Learner** eliminates the need to guess the right model.

**Definition.** Super Learner is an ensemble machine learning method that: (1) takes a library of candidate learners (logistic regression, LASSO, random forest, etc.), (2) uses cross-validation to evaluate each learner's performance, (3) combines them using optimally weighted stacking, and (4) guarantees performance at least as good as the best individual learner (asymptotically).

:::{.callout-tip}
## Why This Matters for TMLE

TMLE is doubly robust (consistent if either nuisance model is correct) and **efficient** when both models converge at reasonable rates. Super Learner reduces the risk of misspecifying either model, while its internal cross-validation protects against overfitting. **Bottom line:** Super Learner + TMLE = data-adaptive, doubly robust, efficient estimation with valid inference.
:::

### Conceptual demonstration

The following shows how Super Learner would be integrated (this requires the SuperLearner package, which may not run in all webR environments):

```{webr-r, eval=FALSE}
library(SuperLearner)

# Define learner library
SL_lib <- c("SL.glm", "SL.glmnet", "SL.gam", "SL.mean")

# Outcome model via Super Learner
Q_SL <- SuperLearner(
  Y = dat$Y,
  X = dat %>% select(A, age, bmi, hba1c, cvd, egfr, statin),
  family = binomial(),
  SL.library = SL_lib
)

# Treatment model via Super Learner
g_SL <- SuperLearner(
  Y = dat$A,
  X = dat %>% select(age, bmi, hba1c, cvd, egfr, statin),
  family = binomial(),
  SL.library = SL_lib
)

# View the cross-validated weights
Q_SL$coef  # how much each learner contributes to the outcome model
g_SL$coef  # how much each learner contributes to the treatment model
```

The Super Learner automatically determines which learners are most useful and how to combine them, so no manual model selection is required.

---

# Running Example: Antiretroviral Therapy Cohort

See the [Running Case Study](00-case-study.qmd) for the full data description.

The HIV antiretroviral therapy cohort illustrates the same progressive estimation strategy developed in this chapter. The baseline snapshot provides a clean point-treatment structure: `ever_haart` is the treatment, `died` is the outcome, and baseline covariates include `age`, `sex`, and `cd4_sqrt_baseline`. This point-treatment simplification collapses each patient's longitudinal record into a single row, setting aside time-varying confounding for now (see Chapter 3.9 for the longitudinal extension).

The five estimators from this chapter (naive, g-computation, IPTW, AIPW, TMLE) can each be applied to this cohort. The progression from naive to TMLE is especially instructive here, because the HAART cohort has strong confounding by indication: sicker patients, with lower CD4 counts, are more likely to receive HAART and also more likely to die. A naive comparison would be misleading. Each successive estimator addresses this confounding more robustly, culminating in TMLE.

```{r}
#| eval: false
download.file("https://raw.githubusercontent.com/amertens/causal-learning-handbook/main/_data/load-haart-cohort.R", "load-haart-cohort.R")
source("load-haart-cohort.R")

# Baseline snapshot: one row per patient
glimpse(haart_baseline)

# A TMLE analysis mirrors the diabetes-drug example above:
# - Treatment model: P(ever_haart = 1 | age, sex, cd4_sqrt_baseline)
# - Outcome model:   E[died | ever_haart, age, sex, cd4_sqrt_baseline]
# The TMLE targeting step and EIC-based inference apply identically.
# See Chapter 2.3 for the full TMLE estimate on this cohort.
```

Because the same causal roadmap applies (define the question, specify the DAG, check assumptions, identify the estimand, and estimate), readers can practice each method on the HAART data after working through the simulated diabetes-drug scenario above.

---

# Summary of Estimator Trade-offs

The table below summarizes the key properties of each estimator. TMLE stands out as the only method that is doubly robust, respects parameter bounds, solves the EIC, and integrates naturally with machine learning.

| Property | G-comp | IPTW | AIPW | TMLE |
|----------|--------|------|------|------|
| Uses outcome model | Yes | No | Yes | Yes |
| Uses treatment model | No | Yes | Yes | Yes |
| Doubly robust | No | No | Yes | Yes |
| Respects bounds | Depends | N/A | No | Yes |
| Efficient (solves EIC) | No | No | Asymptotically | Yes |
| Works with ML | Partially | Partially | Needs cross-fitting | Naturally |
| Inference method | Bootstrap | Sandwich | Influence curve | Influence curve |

---

:::{.callout-tip}
# Key Takeaways

1. **Naive estimates are biased** by confounding by indication. Never interpret unadjusted comparisons as causal effects.
2. **G-computation** models the outcome directly but is sensitive to outcome-model misspecification.
3. **IPTW** reweights by the propensity score but is sensitive to positivity violations and weight instability.
4. **AIPW** is doubly robust but can produce out-of-bounds predictions and does not solve the EIC exactly.
5. **TMLE** targets the estimand, respects parameter bounds, and provides efficient influence-curve-based inference.
6. **Super Learner** reduces model misspecification risk and pairs naturally with TMLE.
7. **Diagnostics are non-negotiable:** always check overlap, weights, covariate balance, and influence curve behavior.
:::

---

# Where to Go Next

## What comes next

With the core estimators in hand, the next part of the handbook turns to diagnostics and heterogeneity. How do you know if your analysis is trustworthy? And does the treatment work differently for different patients? Part III addresses both questions.

The framework in this chapter also extends to several important settings covered in later chapters, including dynamic treatment regimes, stochastic interventions, mediation analysis, longitudinal TMLE, and conditional average treatment effects. Each extension follows the same causal roadmap: define the question, specify the causal model, verify assumptions, identify the statistical estimand, and choose an appropriate estimator.

```{webr-r}
sessionInfo()
```

---

## Check your understanding

<details>
<summary><strong>Q1. A naive comparison of MACE rates between dapagliflozin and sulfonylurea users shows dapagliflozin has lower MACE. Does this prove dapagliflozin is cardioprotective? What could explain this finding even if the drug has no true effect?</strong></summary>

No. A naive comparison does not account for confounding by indication. Physicians prescribe dapagliflozin preferentially to younger, healthier patients with higher eGFR and less prior cardiovascular disease. These patients would have had lower MACE risk regardless of which drug they received. The apparent protective effect could be entirely due to these baseline differences between treatment groups rather than any causal effect of the drug. This is exactly why causal methods (g-computation, IPTW, TMLE) that adjust for measured confounders are necessary, and why the exchangeability assumption (no unmeasured confounding) is the most critical and untestable requirement.

</details>

<details>
<summary><strong>Q2. TMLE uses both an outcome model and a treatment model. If both models are somewhat misspecified, does double robustness still protect you?</strong></summary>

Not fully. Double robustness means the estimator is consistent if *at least one* of the two nuisance models is correctly specified. If both models are misspecified, the estimate can be biased. However, TMLE still has an advantage in this scenario: the bias depends on the *product* of the errors in the two models, so moderate misspecification of both may produce smaller bias than severe misspecification of a single model in g-computation or IPTW. This is why pairing TMLE with Super Learner is recommended. By using a flexible, data-adaptive ensemble for both nuisance functions, you reduce the chance that either model is badly wrong.

</details>

<details>
<summary><strong>Q3. You run TMLE and find that the mean of the efficient influence curve is not close to zero. What does this suggest about your analysis?</strong></summary>

A correctly implemented TMLE solves the efficient influence curve (EIC) equation, so the sample mean of the EIC should be numerically zero (or very close to it). If the mean is appreciably different from zero, this suggests a problem with the implementation or with the models feeding into it. Common causes include: (1) a programming error in the targeting step, (2) propensity scores that are too close to 0 or 1, causing numerical instability in the clever covariate, or (3) convergence failure in the fluctuation model. The EIC mean serves as a built-in diagnostic. It is one of the key advantages TMLE has over other doubly robust estimators like AIPW, where the EIC equation is only solved asymptotically.

</details>

---

## Sources and further reading

- Gruber S, van der Laan MJ (2010). A targeted maximum likelihood estimator of a causal effect on a bounded continuous outcome. *Int J Biostat* 6(1):Article 26.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.
- Schuler MS, Rose S (2017). Targeted maximum likelihood estimation for causal inference in observational studies. *Am J Epidemiol* 185(1):65-73.
- Hoffman K (2021). An illustrated guide to TMLE. [khstats.com](https://www.khstats.com/blog/tmle/tutorial)
- Luque-Fernandez MA et al. (2018). Targeted maximum likelihood estimation for a binary treatment: a tutorial. *Statistics in Medicine* 37(16):2530-2546.
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)
- `SuperLearner` R package: [CRAN](https://cran.r-project.org/package=SuperLearner)
- `drtmle` R package: [CRAN](https://cran.r-project.org/package=drtmle)

---

## Software Implementation (R)

This example brings together the progressive estimation sequence from this chapter (naive → g-computation → IPTW → TMLE) using the `tmle` package with Super Learner.

- Simulate a pharmacoepi-style dataset with multiple confounders
- Run the `tmle()` function with a diverse Super Learner library
- Extract and interpret: ATE, risk difference, confidence interval, and influence curve diagnostics

```{r}
#| eval: false
set.seed(1)
n <- 800
W1 <- rnorm(n)                                      # age (standardized)
W2 <- rbinom(n, 1, 0.3)                             # diabetes indicator
W3 <- rnorm(n, mean = 0.5 * W1)                     # BMI (standardized)
A  <- rbinom(n, 1, plogis(-0.3 + 0.4 * W1 + 0.5 * W2 - 0.2 * W3))
Y  <- rbinom(n, 1, plogis(-2 + 0.6 * A + 0.5 * W1 + 0.8 * W2 + 0.3 * W3))

if (requireNamespace("tmle", quietly = TRUE)) {
  library(tmle)
  W <- data.frame(W1 = W1, W2 = W2, W3 = W3)

  tmle_fit <- tmle(
    Y = Y, A = A, W = W,
    family = "binomial",
    Q.SL.library = c("SL.glm", "SL.step", "SL.mean"),
    g.SL.library  = c("SL.glm", "SL.step", "SL.mean")
  )

  cat("── TMLE results ─────────────────────\n")
  cat("ATE (risk difference):", round(tmle_fit$estimates$ATE$psi, 4), "\n")
  cat("95% CI:", round(tmle_fit$estimates$ATE$CI, 4), "\n")
  cat("p-value:", format.pval(tmle_fit$estimates$ATE$pvalue, digits = 3), "\n")

  ## Influence curve diagnostic (mean should be ≈ 0)
  ic <- tmle_fit$estimates$ATE$IC
  cat("\nInfluence curve mean:", round(mean(ic), 6),
      " (should be ≈ 0)\n")
} else {
  message("Install the 'tmle' package:  install.packages('tmle')")
}
```
