---
format: html
---

# Outcome Modeling and Standardization (G-Computation) {#sec-gcomputation}

*Estimated reading time: 25 minutes*

*G-computation as a foundation for causal estimation*

Outcome modeling and standardization, often referred to as **g-computation**, is one of the oldest and most intuitive approaches to causal inference [@robins1994estimation; @petersen2014causal].

---

G-computation estimates causal effects by predicting what *would* happen to everyone under each treatment, then comparing those predictions [@hernan2025whatif]. No reliance on a single regression coefficient required. It directly implements the identification formula from @sec-roadmap ($E[Y(a)] = E_W[E[Y \mid A=a, W]]$, where the potential outcome $Y(a)$ is written $Y(1)$ and $Y(0)$ for the two treatment levels), making it the most concrete entry point for learning causal estimation, and mapping the causal estimand to a statistical estimand.

G-computation reconstructs what would have happened if *everyone* received treatment ($A=1$) and, separately, if *everyone* received control ($A=0$). The recipe is three steps: (1) fit an outcome model $E[Y \mid A, W]$, (2) predict outcomes for each individual under both treatments, and (3) average those predictions (standardization).

# Implementation Example: HIV Antiretroviral Therapy Cohort

We use the shared HIV antiretroviral therapy cohort introduced in the case study [@vanderwal2011ipw]. This dataset follows HIV-positive patients from seroconversion, tracking HAART initiation and mortality. Here we collapse the longitudinal data to a point-treatment snapshot: A = `ever_haart` (1 = initiated HAART at any point, 0 = never initiated), Y = `died` (1 = died, 0 = alive at end of follow-up), with confounders age, sex, and baseline CD4 count (square-root transformed).

```{r}
#| interactive: true
#| code-fold: true
library(tidyverse)

source("_data/load-haart-cohort.R")
dat <- as_tibble(haart_baseline)
```

Treatment is strongly confounded by baseline CD4 count and age, making this a good scenario for demonstrating g-computation.

---

# Step-by-Step G-Computation

### Step 1: Fit an Outcome Model

We fit a logistic regression model predicting the outcome from treatment and confounders:

```{r}
#| interactive: true
#| code-fold: true
mod <- glm(died ~ ever_haart + age + sex + cd4_sqrt_baseline,
           family = binomial, data = dat)
broom::tidy(mod, conf.int = TRUE) |> knitr::kable(digits = 3, caption = "Outcome model coefficients")
```

**Pitfall.** The coefficient on `ever_haart` from this logistic regression is a *conditional log-odds ratio*. It tells you the association between HAART initiation and mortality *holding confounders fixed*. That is not a causal effect on its own. We need the next two steps (prediction + averaging) to turn this model into a causal estimate.

### Step 2: Predict Counterfactual Outcomes (The "What If" Step)

We ask two "what if" questions for *every single person* in the dataset, regardless of what treatment they actually received: what if this person had been treated ($A=1$)? What if untreated ($A=0$)?

```{r}
#| interactive: true
# create counterfactual datasets
dat1 <- dat %>% mutate(ever_haart = 1)
dat0 <- dat %>% mutate(ever_haart = 0)

# predict potential outcomes
p1 <- predict(mod, newdata = dat1, type = "response")
p0 <- predict(mod, newdata = dat0, type = "response")
```

Here `p1[i]` and `p0[i]` are the predicted outcomes for person *i* under treatment and control, respectively. Notice we are *not* comparing treated vs. untreated groups. We are comparing two hypothetical worlds for the *same* people. This is the counterfactual reasoning that separates causal inference from plain regression.

### Step 3: Standardize (Averaging Over the Population)

Averaging the counterfactual predictions over the *entire population* converts model-based predictions into a causal estimate: a standardized (marginal) risk difference, the same quantity a well-randomized trial would yield.

```{r}
#| interactive: true
#| code-fold: true
risk1 <- mean(p1)
risk0 <- mean(p0)
ate  <- risk1 - risk0

knitr::kable(data.frame(risk1 = risk1, risk0 = risk0, ate = ate))
```

```{r}
#| label: fig-gcomp-counterfactual
#| fig-cap: "Predicted mortality risk for every individual under universal HAART initiation ($A=1$) versus no initiation ($A=0$). G-computation averages each distribution down to a single standardized risk (dashed lines); the gap between those averages is the ATE."
#| code-fold: true
library(ggplot2)
tibble::tibble(
  risk = c(p1, p0),
  arm  = rep(c("Under HAART (A = 1)", "No HAART (A = 0)"), each = length(p1))
) |>
  ggplot(aes(risk, fill = arm)) +
  geom_density(alpha = 0.5) +
  geom_vline(xintercept = c(risk1, risk0), linetype = "dashed") +
  labs(x = "Predicted mortality risk", y = "Density", fill = NULL,
       title = "Counterfactual risk distributions g-computation averages over") +
  theme_minimal()
```

**How to interpret these results:** `ate` is the standardized (marginal) risk difference, and it is the causal number, not either regression coefficient. Read it as: if the entire population had initiated HAART rather than not, the estimated mortality risk would change by `ate` (about `r round(ate * 100, 1)` percentage points), with the measured confounders' distribution held fixed. `risk1` and `risk0` are the standardized risks under universal treatment and under no treatment; the contrast between them is what a well-randomized trial would have estimated.

### Beyond the Risk Difference: Relative Risk and Population Attributable Fraction

One of the strengths of g-computation is that once you have the counterfactual predictions `p1` and `p0`, you can compute *any* contrast, not just the risk difference. This flexibility is difficult to achieve with a single regression coefficient.

```{r}
#| interactive: true
#| code-fold: true
# Relative risk (risk ratio)
rr <- risk1 / risk0

# Causal prevented (excess) fraction, relative to a "no one treated" world:
# (P(Y=1) - P(Y(0)=1)) / P(Y=1)
# For a PROTECTIVE exposure (risk0 > overall_risk) this comes out NEGATIVE,
# meaning the current treatment distribution PREVENTS that fraction of deaths
# relative to no one being treated.
overall_risk <- mean(dat$died)
prevented_frac <- (overall_risk - risk0) / overall_risk

knitr::kable(data.frame(
  Measure = c("Risk difference (ATE)", "Relative risk", "Prevented fraction (vs no one treated)"),
  Estimate = round(c(ate, rr, prevented_frac), 4)
))
```

The **relative risk** tells you how many times more likely the outcome is under treatment versus control. The **prevented fraction** compares the observed (current) mortality risk to the risk in a counterfactual world where *no one* initiated HAART: $(P(Y=1) - P(Y(0)=1)) / P(Y=1)$. The familiar "population attributable fraction" reading, the fraction of cases that would be *eliminated* by removing exposure, applies to a *harmful* exposure. HAART is protective, so the sign flips: $\text{risk0} > \text{overall risk}$, the quantity comes out **negative**, and a negative value means the current treatment distribution *prevents* that fraction of deaths relative to a world where no one is treated. Read literally, moving everyone to no-HAART would *add* deaths, not eliminate them. Under the identification assumptions (exchangeability, positivity, consistency), this contrast has a causal interpretation, unlike the traditional epidemiologic PAF formula, which uses observed prevalence and does not require causal assumptions but also does not have a causal interpretation.

<!-- editor note: implemented Option A from the editorial spec. The original code/prose labeled (overall_risk - risk0)/overall_risk as a "population attributable fraction" eliminated by removing exposure, which is sign-inverted for a protective treatment. Kept the formula, renamed to "prevented fraction", and stated the sign direction explicitly. -->


Neither of these measures is directly available from a single logistic regression coefficient; you would need to exponentiate and back-transform, making strong assumptions about the scale. With g-computation, you compute them directly from the predicted counterfactual risks.


# Why G-Computation Works

It directly implements the identification formula:

$$
E_W[ E[Y | A=a, W] ].
$$

In plain terms, this averages the predicted outcomes under treatment (or control) across the full study population, weighting each prediction by the population's covariate distribution. The goal remains the same: estimate $E[Y(1)] - E[Y(0)]$, the population-level risk difference.



:::{.callout-tip}
## Key Distinction: Marginal Risk Difference vs. Conditional Log-Odds Ratio

This is one of the most commonly confused points in epidemiology:

- **Conditional log-odds ratio** (what logistic regression gives you): "Among people with the *same* age, sex, and baseline CD4, how do the odds of death differ between those who initiated HAART and those who did not?" This is a measure of *association within strata*. It does not tell you what would happen if you intervened on the whole population. It is also non-collapsible, meaning it changes depending on which covariates you condition on, even without confounding.

- **Marginal risk difference** (what g-computation gives you): "If we could treat *everyone* versus *no one*, how much would the overall population risk change?" This is a *causal* contrast on the risk scale, the kind of number a clinician or policy maker can act on.

G-computation bridges the gap: it uses the conditional model internally but standardizes over the population to produce the marginal (causal) effect.
:::

---

# Diagnostics: When Can G-Computation Fail?

:::{.callout-warning}
## Model Misspecification

G-computation rests entirely on the outcome model $E[Y \mid A, W]$. If that model is wrong (e.g., omits interactions, assumes linearity when the truth is non-linear), bias propagates directly into the causal estimate. For example, suppose the true risk of virologic failure depends on a nonlinear interaction between baseline CD4 count and age, but the fitted model includes only main effects. The model will generate systematically wrong counterfactual predictions for patients with unusual combinations of CD4 and age, and the resulting ATE estimate will be biased. Unlike IPTW, where positivity failures produce visible extreme weights, g-computation's extrapolation errors are silent: the model returns a number with no warning that it is wrong. Always check residuals and consider flexible alternatives (see @sec-doubly-robust-tmle).
:::

## Watch it fail: a misspecified model biases the estimate

The warning above is easy to state and easy to ignore, so here it is as a number. We simulate a world where the truth is nonlinear [@sofrygin2017simcausal], then estimate the effect two ways: with a main-terms logistic regression that misses the nonlinearity, and with a model that includes it. The risk difference is known by construction, so we can see exactly how far the misspecified model lands from the truth.

```{r}
#| code-fold: show
set.seed(2024)
n   <- 4000
age <- runif(n, 30, 80)
A   <- rbinom(n, 1, plogis(-2 + 0.04 * age))             # older patients treated more often

# True risk: curved in age, and the treatment effect is larger above age 65
true_risk <- function(a, w) plogis(-1 + 0.8 * a - 0.0009 * (w - 55)^2 + 0.6 * a * (w > 65))
Y <- rbinom(n, 1, true_risk(A, age))
d <- data.frame(Y, A, age)

# Truth: population risk difference under the known model
true_ate <- mean(true_risk(1, age) - true_risk(0, age))

# Misspecified g-computation: main terms only (no curvature, no interaction)
m_wrong   <- glm(Y ~ A + age, family = binomial, data = d)
ate_wrong <- mean(predict(m_wrong, transform(d, A = 1), type = "response") -
                  predict(m_wrong, transform(d, A = 0), type = "response"))

# Correctly flexible g-computation: include the curvature and interaction
m_flex   <- glm(Y ~ A * age + I((age - 55)^2) + A:I(age > 65), family = binomial, data = d)
ate_flex <- mean(predict(m_flex, transform(d, A = 1), type = "response") -
                 predict(m_flex, transform(d, A = 0), type = "response"))

round(c(truth = true_ate, main_terms = ate_wrong, flexible = ate_flex), 3)
```

The main-terms estimate misses the true risk difference because it cannot represent the curvature and the above-65 interaction; the flexible model, which happens to include exactly those terms, recovers the truth. The catch in real data is that no one tells you which terms to add. That is the problem Super Learner (@sec-superlearner) solves: rather than guessing the functional form, it lets a library of algorithms and cross-validation choose it [@vanderLaan2007superlearner], which is why the next chapters build g-computation and TMLE on top of machine learning rather than a single hand-specified regression.

---

:::{.callout-warning}
## Poor Positivity

If some strata almost never receive a treatment, g-computation *extrapolates* into regions with little observed data. R will return a number, but it may be meaningless. Check propensity scores for values near 0 or 1:

```{r}
#| interactive: true
ps <- predict(glm(ever_haart ~ age + sex + cd4_sqrt_baseline,
                   family = binomial, data = dat), type = "response")
summary(ps)
```

G-computation handles near-positivity violations differently from IPTW. Where IPTW produces extreme weights for observations with propensity scores near 0 or 1, g-computation instead extrapolates from the fitted outcome model. This may be more or less reliable depending on how well the model captures the true outcome surface in sparse regions of the covariate space. The extrapolation is implicit and may not produce visible warnings. Positivity and overlap receive a dedicated treatment, with diagnostics and remedies, in @sec-positivity-overlap.
:::

---

**Pitfall.** No modeling strategy fixes missing confounders. If an important common cause of treatment and outcome is absent from your dataset, g-computation (like every other method in this handbook) will produce a biased estimate. However, g-computation makes its assumptions explicit (you can see exactly which variables $W$ enter the model), which aids transparent reporting and sensitivity analysis.

---

**Practical note.** The biggest limitation of g-computation with logistic regression is that you must get the functional form right. A powerful alternative is to replace the parametric model with a data-adaptive learner (random forests, gradient boosting, GAMs, or a SuperLearner ensemble) [@vanderLaan2007superlearner], reducing reliance on parametric assumptions. In later chapters we systematically integrate SuperLearner with **TMLE** [@vanderLaan2006tmle].

::: {.callout-note}
## Super Learner in one sentence (for now)
Super Learner is an ensemble that uses cross-validation to combine many candidate models into a single prediction, formally introduced in @sec-superlearner. Until then you can read the `SuperLearner(...)` call below as a flexible stand-in for the single GLM: same role (predict the outcome given treatment and covariates), but it lets the data choose the functional form instead of you guessing it.
:::

```{r}
#| interactive: true
#| warning: false
#| message: false
library(SuperLearner)
set.seed(2026)

# SuperLearner expects a plain data.frame of predictors
X_all <- as.data.frame(dat[, c("ever_haart", "age", "sex", "cd4_sqrt_baseline")])

sl_mod <- SuperLearner(
  Y = dat$died,
  X = X_all,
  family = binomial(),
  SL.library = c("SL.glm", "SL.ranger", "SL.gam")
)

# predict for counterfactuals (predictor columns only)
p1_sl <- predict(sl_mod, newdata = transform(X_all, ever_haart = 1))$pred
p0_sl <- predict(sl_mod, newdata = transform(X_all, ever_haart = 0))$pred

# Super Learner g-computation ATE (compare to the GLM estimate above)
c(ate_sl = mean(p1_sl - p0_sl))
```

---

:::{.callout-warning}
## Why G-Computation Is Less Common in Practice: No Built-In Inference

The simple plug-in g-computation implementation shown here produces a point estimate but does not automatically provide standard errors or confidence intervals. This is a practical limitation of the plug-in approach, not a fundamental property of g-computation: influence-function-based variance estimators exist under regularity conditions and are used by TMLE (covered in @sec-doubly-robust-tmle).

In practice, the most common approach for inference with plug-in g-computation is the **nonparametric bootstrap**: repeat the entire procedure (fit model, predict counterfactuals, average) on many resampled datasets and take percentiles of the resulting distribution.

```{r}
#| interactive: true
# Bootstrap confidence interval for g-computation
set.seed(42)
n_boot <- 500
boot_ates <- numeric(n_boot)

for (b in seq_len(n_boot)) {
  boot_dat <- dat[sample(nrow(dat), replace = TRUE), ]
  boot_mod <- glm(died ~ ever_haart + age + sex + cd4_sqrt_baseline,
                   family = binomial, data = boot_dat)
  bp1 <- predict(boot_mod, newdata = mutate(boot_dat, ever_haart = 1), type = "response")
  bp0 <- predict(boot_mod, newdata = mutate(boot_dat, ever_haart = 0), type = "response")
  boot_ates[b] <- mean(bp1) - mean(bp0)
}

quantile(boot_ates, c(0.025, 0.975))
```

This works but is computationally expensive and does not benefit from efficiency theory. The methods in the next two chapters (**IPTW** and especially TMLE) provide analytic standard errors and confidence intervals based on influence function theory, making them more practical for routine use [@vanderLaan2006tmle]. TMLE in particular combines the outcome-modeling strengths of g-computation with formal inference, which is one reason it is the recommended approach in later chapters.
:::

:::{.callout-tip}
## Summary

G-computation = fit a model, predict under both treatments for everyone, then average. That three-step standardization is what converts a conditional model into a marginal causal estimate. Its main vulnerability is outcome-model misspecification; flexible learners and doubly robust methods (IPTW, TMLE) address this in subsequent chapters.
:::

---

## Check your understanding

<details>
<summary><strong>Q1. You run g-computation using a logistic regression model. A colleague asks: "Isn't this just a regression coefficient?" How do you explain the difference?</strong></summary>

No. A logistic regression coefficient is a *conditional log-odds ratio* that describes how the log-odds of the outcome change with treatment, holding covariates fixed at particular values. G-computation uses that same model but adds two critical steps: it predicts outcomes for every individual under both $A=1$ and $A=0$, then *averages* those predictions over the entire population. The result is a marginal risk difference (the ATE on the risk scale), which answers a different question: "How much would population-level risk change if we could treat everyone versus no one?" The regression coefficient is conditional and on the log-odds scale; the g-computation estimate is marginal and on the risk scale. These are different quantities even when the underlying model is the same.

</details>

<details>
<summary><strong>Q2. What happens to the g-computation estimate if the outcome model is misspecified?</strong></summary>

G-computation is only as good as its outcome model. If $E[Y \mid A, W]$ is misspecified (for example, it assumes a linear relationship when the truth involves interactions or non-linearities), the counterfactual predictions will be systematically wrong, and the ATE estimate will be biased. Unlike doubly robust methods (AIPW, TMLE), g-computation has no second chance: there is no treatment model to compensate for outcome-model errors. This is why diagnostics (residual checks, comparison across model specifications) and flexible modeling strategies (SuperLearner) are important when using g-computation.

</details>

<details>
<summary><strong>Q3. Why does g-computation produce a marginal (population-level) effect even though it uses a conditional model internally?</strong></summary>

The key is the standardization (averaging) step. The outcome model $E[Y \mid A, W]$ is conditional: it gives a predicted outcome for a specific individual with covariates $W=w$ under treatment $A=a$. But g-computation does not report these individual-level predictions. Instead, it predicts for *every person* under $A=1$ and under $A=0$, then takes the population average of the difference. This averaging over the empirical covariate distribution $P(W)$ is exactly the operation $E_W[E[Y \mid A=a, W]]$ from the identification formula, and it produces a marginal effect. The conditional model is an intermediate tool; the final quantity is a population-level causal contrast.

</details>

---

## What comes next

G-computation depends entirely on the outcome model being correctly specified. What if we could instead model the treatment mechanism? The next chapter introduces IPTW, which takes exactly that approach: reweighting observations to create a pseudo-population where treatment appears randomized.

---
