# Advanced TMLE: Cross-Validated, Collaborative, and Adaptive {#sec-advanced-tmle}

*Estimated reading time: 30 minutes*

::: {.callout-tip}
## Learning Objectives

After completing this chapter, you will be able to:

1. Explain why standard TMLE can break down when nuisance models are fit with aggressive machine learning, and how cross-validation fixes this
2. Describe the non-technical intuition behind CV-TMLE, C-TMLE, and A-TMLE
3. Identify practical scenarios where each variant is preferred over standard TMLE
4. Implement CV-TMLE and C-TMLE in R using available packages
5. Distinguish CV-TMLE from double/debiased machine learning (DML) and articulate why the substitution property matters
:::

Chapter 2.3 introduced TMLE as a doubly robust, semiparametrically efficient estimator that combines flexible outcome modeling with a targeting step. In many applications, standard TMLE works well. But three practical problems can arise, each addressed by a different TMLE variant:

| Problem | Variant | Core idea |
|:--------|:--------|:----------|
| Aggressive ML violates regularity conditions | **CV-TMLE** | Use sample splitting so the targeting step sees "fresh" data |
| Propensity score overfitting inflates variance | **C-TMLE** | Let the outcome model guide how much to adjust the propensity score |
| External data could boost precision but may be biased | **A-TMLE** | Data-adaptively borrow from real-world data while guaranteeing no harm |

This chapter explains the non-technical intuition for each, when and why to use it, and how to implement it.

```{r}
#| label: setup-advanced-tmle
#| include: false

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```


## The Problem That Motivates All Three Variants {#sec-why-advanced}

Recall the standard TMLE recipe from Chapter 2.3:

1. **Initial estimation**: Fit an outcome model $\hat{Q}(A, W)$ and a propensity score $\hat{g}(W)$, typically using Super Learner.
2. **Targeting step**: Update $\hat{Q}$ along a least-favorable parametric submodel, using $\hat{g}$ as the "clever covariate."
3. **Plug in**: Compute the causal effect estimate from the updated $\hat{Q}^*$.

The asymptotic guarantees of TMLE (consistency, efficiency, valid confidence intervals) depend on regularity conditions for the nuisance estimators $\hat{Q}$ and $\hat{g}$. The most important of these is the **Donsker condition**, which roughly requires that the nuisance estimators do not overfit the data too aggressively.

::: {.callout-important}
## When Standard TMLE Can Fail

Standard TMLE's theoretical guarantees assume the nuisance estimators fall in a **Donsker class**---a set of functions with controlled complexity. This is satisfied by parametric models and many regularized estimators, but can be violated by:

- Highly adaptive machine learning (deep ensembles, unregularized random forests, neural networks)
- Super Learners with very large candidate libraries
- High-dimensional settings where the number of covariates grows with sample size

When the Donsker condition fails, the influence-curve-based variance estimator may be anti-conservative, confidence intervals can have below-nominal coverage, and the estimator may not be asymptotically linear.
:::

Each of the three variants addresses a different facet of this problem.


## CV-TMLE: Cross-Validated TMLE {#sec-cv-tmle}

### The intuition

CV-TMLE solves the Donsker problem through **sample splitting**. The idea is simple: if you fit your nuisance models on one part of the data and evaluate the targeting step on a held-out part, the nuisance estimator is a fixed function when applied to the validation data. Fixed functions are trivially Donsker. This is the same insight that powers cross-validation and, more recently, double/debiased machine learning (DML).

In practice, CV-TMLE uses V-fold cross-validation:

1. Split the data into V folds.
2. For each fold $v$, fit $\hat{Q}^{(-v)}$ and $\hat{g}^{(-v)}$ on the *training* data (all folds except $v$).
3. Generate predictions for fold $v$ using the models fit without fold $v$.
4. Stack all V sets of predictions together.
5. Run the TMLE targeting step on the full (stacked) dataset using the cross-validated predictions.
6. Compute the estimate and influence-curve-based inference.

::: {.callout-tip}
## The Key Insight

By fitting nuisance models on training data and targeting on validation data, CV-TMLE **completely removes the Donsker requirement**. You can use arbitrarily complex machine learning---deep Super Learners, HAL, neural networks---without worrying that overfitting will invalidate your inference. The price is a modest increase in computational cost (fitting V models instead of one).
:::

### CV-TMLE vs. Double Machine Learning

CV-TMLE and DML (debiased machine learning) both use sample splitting and both achieve semiparametric efficiency. The key difference is the estimator type:

| Property | CV-TMLE | DML |
|:---------|:--------|:----|
| **Estimator type** | Substitution (plug-in) | Estimating equation (one-step) |
| **Respects parameter bounds** | Yes (e.g., probabilities stay in [0,1]) | Not guaranteed |
| **Targeting step** | Updates the outcome model directly | Adds a bias-correction term |
| **Finite-sample behavior** | Generally more stable | Can produce out-of-range estimates |

In practice, the substitution property of CV-TMLE matters when the outcome is rare (estimated probabilities near 0) or when the treatment effect is small. DML can produce negative probability estimates; CV-TMLE cannot.

### When to use CV-TMLE

- **Always a safe default**: CV-TMLE is never worse than standard TMLE asymptotically, and it is strictly better when using aggressive ML. Many practitioners now use CV-TMLE as their default.
- **Large Super Learner libraries**: If your library includes dozens of candidates or highly adaptive algorithms (HAL, deep learners), CV-TMLE is strongly recommended.
- **High-dimensional covariates**: When the number of potential confounders is large relative to sample size.

### Implementation in R

The `tmle3` package in the `tlverse` ecosystem implements CV-TMLE natively. The older `tmle` package does not support cross-validation by default, but you can implement it manually by cross-validating the initial fits.

```{r}
#| label: cv-tmle-example
#| code-fold: show
#| message: false
#| warning: false

library(ipw)
library(dplyr)
library(SuperLearner)
library(tmle)

# Load the ART cohort
data(haartdat)
haart_bl <- haartdat |>
  group_by(patient) |>
  summarise(
    sex = first(sex),
    age = first(age),
    cd4_bl = first(cd4.sqrt)^2,
    cd4_sqrt_bl = first(cd4.sqrt),
    A = max(haartind),
    Y = max(event),
    .groups = "drop"
  )

W <- haart_bl[, c("age", "sex", "cd4_sqrt_bl")]

# --- Standard TMLE (no cross-validation) ---
set.seed(42)
fit_standard <- tmle(
  Y = haart_bl$Y,
  A = haart_bl$A,
  W = W,
  Q.SL.library = c("SL.glm", "SL.glmnet", "SL.mean"),
  g.SL.library = c("SL.glm", "SL.glmnet", "SL.mean"),
  family = "binomial"
)

cat("=== Standard TMLE ===\n")
cat(sprintf("ATE: %.4f (95%% CI: %.4f, %.4f)\n",
    fit_standard$estimates$ATE$psi,
    fit_standard$estimates$ATE$CI[1],
    fit_standard$estimates$ATE$CI[2]))
```

```{r}
#| label: cv-tmle-manual
#| code-fold: show
#| message: false
#| warning: false

# --- Manual CV-TMLE (5-fold) ---
# Demonstrates the cross-validation logic explicitly

set.seed(42)
n <- nrow(haart_bl)
V <- 5
folds <- sample(rep(1:V, length.out = n))

# Storage for cross-validated predictions
Qbar_cv <- matrix(NA, n, 2)  # columns: Q(0,W), Q(1,W)
g_cv <- rep(NA, n)

for (v in 1:V) {
  train <- folds != v
  valid <- folds == v

  # Fit outcome model on training data
  Q_fit <- SuperLearner(
    Y = haart_bl$Y[train],
    X = data.frame(A = haart_bl$A[train], W[train, ]),
    SL.library = c("SL.glm", "SL.glmnet", "SL.mean"),
    family = binomial()
  )

  # Predict on validation data under A=0 and A=1
  newX0 <- data.frame(A = 0, W[valid, ])
  newX1 <- data.frame(A = 1, W[valid, ])
  Qbar_cv[valid, 1] <- predict(Q_fit, newdata = newX0)$pred
  Qbar_cv[valid, 2] <- predict(Q_fit, newdata = newX1)$pred

  # Fit propensity score on training data
  g_fit <- SuperLearner(
    Y = haart_bl$A[train],
    X = W[train, ],
    SL.library = c("SL.glm", "SL.glmnet", "SL.mean"),
    family = binomial()
  )
  g_cv[valid] <- predict(g_fit, newdata = W[valid, ])$pred
}

# --- Targeting step on full dataset using CV predictions ---
# Clever covariate
H1 <- haart_bl$A / g_cv
H0 <- -(1 - haart_bl$A) / (1 - g_cv)

# Fluctuation model (logistic submodel)
Q_logit_A <- qlogis(ifelse(haart_bl$A == 1, Qbar_cv[, 2], Qbar_cv[, 1]))
eps_fit <- glm(haart_bl$Y ~ -1 + offset(Q_logit_A) + H1 + H0,
               family = binomial)
eps <- coef(eps_fit)

# Update predictions
Q1_star <- plogis(qlogis(Qbar_cv[, 2]) + eps["H1"] / g_cv)
Q0_star <- plogis(qlogis(Qbar_cv[, 1]) - eps["H0"] / (1 - g_cv))

# Plug in
mu1 <- mean(Q1_star)
mu0 <- mean(Q0_star)
ate_cv <- mu1 - mu0

# Influence curve-based SE
IC <- (haart_bl$A / g_cv) * (haart_bl$Y - Q1_star) + Q1_star - mu1 -
      ((1 - haart_bl$A) / (1 - g_cv)) * (haart_bl$Y - Q0_star) - Q0_star + mu0
se_cv <- sqrt(var(IC) / n)

cat("=== CV-TMLE (manual, 5-fold) ===\n")
cat(sprintf("ATE: %.4f (95%% CI: %.4f, %.4f)\n",
    ate_cv, ate_cv - 1.96 * se_cv, ate_cv + 1.96 * se_cv))
```

::: {.callout-tip}
## Implementation Note

In the `tlverse` ecosystem, `tmle3` handles CV-TMLE automatically. For the `tmle` package, you can approximate CV-TMLE by passing pre-computed cross-validated predictions. The manual implementation above shows the core logic: fit on training folds, predict on validation folds, target on the stacked predictions.
:::


## C-TMLE: Collaborative TMLE {#sec-c-tmle}

### The intuition

Standard TMLE fits the propensity score $\hat{g}(W)$ independently of the outcome model $\hat{Q}(A, W)$. This makes sense from a pure prediction standpoint: the propensity score should predict treatment assignment as well as possible. But from the standpoint of *estimating the causal effect*, an overly flexible propensity score can be counterproductive.

Here is the key insight: **the propensity score enters TMLE only through the clever covariate $H = A/g(W)$**. If $g(W)$ is very close to 0 or 1 for some patients (near-positivity violations), the clever covariate explodes, and the targeting step becomes unstable---even if the propensity score fits the data beautifully.

C-TMLE flips the script: instead of fitting $g$ to predict treatment well, it builds a sequence of increasingly complex propensity score models and selects the one that produces the best *outcome model fit* for the target parameter. In other words, the propensity score *collaborates* with the outcome model.

::: {.callout-important}
## The Collaboration Principle

In standard TMLE, the propensity score and outcome model are fit independently. C-TMLE breaks this independence:

1. Start with a simple propensity score (e.g., intercept only: $g(W) = P(A=1)$).
2. Build a sequence of increasingly complex models: $g_1, g_2, \ldots, g_K$.
3. For each $g_k$, compute the targeted update $\hat{Q}^*_k$ and evaluate a cross-validated loss.
4. Select the $g_k$ that minimizes the loss for $\hat{Q}^*_k$.

This is **not** choosing the propensity score that predicts treatment best. It is choosing the propensity score that, when plugged into the targeting step, produces the best estimate of the *outcome under intervention*. Variables that improve treatment prediction but destabilize the causal estimate are excluded.
:::

### When to use C-TMLE

- **Near-positivity violations**: When some patients have propensity scores very close to 0 or 1, standard TMLE can be unstable. C-TMLE automatically avoids overfitting the propensity score in ways that amplify these near-violations.
- **High-dimensional propensity score models**: When you have hundreds or thousands of potential confounders (e.g., claims databases), C-TMLE can perform variable selection in the propensity score guided by the target estimand.
- **Practical positivity problems**: Even without theoretical violations, small samples or rare treatment patterns can cause practical instability that C-TMLE mitigates.

### Scalable C-TMLE

The original greedy C-TMLE algorithm considers adding one variable at a time and evaluates the targeting step at each step. With $p$ covariates, this is $O(p^2)$, which is prohibitive for high-dimensional data. Scalable C-TMLE uses **pre-ordering strategies** to reduce complexity to $O(p)$:

- **LASSO-C-TMLE**: Order covariates by the LASSO regularization path for the propensity score. Evaluate the targeting step at each point along the path.
- **Univariate pre-ordering**: Rank covariates by their univariate association with treatment, then add them in that order.

### Implementation in R

The `ctmle` R package implements both discrete (greedy) and LASSO-based C-TMLE.

```{r}
#| label: ctmle-example
#| code-fold: show
#| message: false
#| warning: false

# install.packages("ctmle")
library(ctmle)

# Prepare data
Y <- haart_bl$Y
A <- haart_bl$A
W_mat <- as.matrix(W)

# --- LASSO-C-TMLE ---
# Step 1: Fit initial outcome model Q(A, W)
Q_init <- SuperLearner(
  Y = Y, X = data.frame(A = A, W_mat),
  SL.library = c("SL.glm", "SL.mean"), family = binomial()
)
Q0 <- predict(Q_init, newdata = data.frame(A = 0, W_mat))$pred
Q1 <- predict(Q_init, newdata = data.frame(A = 1, W_mat))$pred
Q_mat <- cbind(QAW = ifelse(A == 1, Q1, Q0), Q0W = Q0, Q1W = Q1)

# Step 2: Get LASSO regularization path for the propensity score
glmnet_fit <- glmnet::glmnet(x = W_mat, y = A, family = "binomial", nlambda = 20)

# Step 3: C-TMLE selects the lambda (propensity score complexity) that
#         produces the best TMLE for the causal estimate
set.seed(42)
fit_ctmle <- ctmleGlmnet(
  Y = Y, A = A, W = W_mat, Wg = W_mat,
  Q = Q_mat,
  lambdas = glmnet_fit$lambda,
  ctmletype = 1,         # forward selection along LASSO path
  family = "gaussian",
  V = 5                  # folds for cross-validated selection
)

cat("=== C-TMLE (LASSO path) ===\n")
cat(sprintf("ATE:  %.4f\n", fit_ctmle$est))
cat(sprintf("Best lambda index: %d (of %d along path)\n",
    fit_ctmle$best_k, length(glmnet_fit$lambda)))
cat("Compare to standard TMLE above---C-TMLE may differ when\n")
cat("it selects a simpler propensity score to stabilize estimation.\n")
```

::: {.callout-tip}
## Comparing Standard TMLE and C-TMLE

The difference between standard TMLE and C-TMLE is most visible when there are near-positivity violations. In the ART cohort, patients with very low CD4 counts are almost always treated, creating near-violations. C-TMLE automatically finds the propensity score complexity that balances confounding adjustment against instability from extreme weights.
:::


## A-TMLE: Adaptive TMLE {#sec-a-tmle}

### The intuition

A-TMLE addresses a different problem: **how to safely borrow strength from external data**. Suppose you have a small randomized trial and a large observational database. Can you use the observational data to improve precision without introducing bias?

The challenge is that the observational data may differ from the trial population in unmeasured ways. Naive pooling can introduce bias. Ignoring the external data wastes information.

A-TMLE takes a principled middle path:

1. **Decompose the estimand**: Write the ATE as the difference between a "pooled ATE" (using both RCT and external data) and a "bias term" (the effect of being in the RCT on the outcome, conditional on treatment and covariates).
2. **Data-adaptively estimate the bias**: Use machine learning to learn a working model for the bias term. If the bias is simple (e.g., explained by measured covariates), large efficiency gains are possible.
3. **Guarantee no harm**: By construction, A-TMLE is always at least as efficient as using the RCT data alone. If the external data are badly biased, A-TMLE learns to ignore them.

::: {.callout-important}
## The "No Harm" Guarantee

A-TMLE's defining property is that **borrowing external data can help but never hurt**. The asymptotic variance of the A-TMLE estimator is at most equal to the variance of the TMLE using RCT data alone. This is because the bias working model is selected by cross-validation: if no model improves on the intercept-only model (which corresponds to ignoring external data), the estimator falls back to the RCT-only analysis.

This makes A-TMLE particularly attractive for regulatory settings, where borrowing from external controls must be justified.
:::

### When to use A-TMLE

- **Single-arm trials with external controls**: A-TMLE can augment a single-arm trial with a real-world control arm while guaranteeing that the augmentation does not inflate type I error.
- **Underpowered RCTs**: When the trial is too small for adequate precision, external data can help---but only if the bias is manageable.
- **Hybrid trial designs**: More generally, whenever you want to combine experimental and observational data with formal efficiency guarantees.

### Software

A-TMLE is a relatively recent development. Implementation is available in the `tlverse` ecosystem and in custom code accompanying the original publications. As of 2026, no standalone CRAN package exists, but the methodology is described in detail with reproducible code in the original paper.

```{r}
#| label: atmle-concept
#| eval: false
#| code-fold: show

# Conceptual pseudocode for A-TMLE
# (Full implementation requires the tlverse ecosystem or custom code)

# 1. Combine RCT and external data
# pooled_data <- rbind(rct_data, external_data)
# pooled_data$S <- c(rep(1, nrow(rct_data)), rep(0, nrow(external_data)))

# 2. Fit nuisance models on pooled data
# Q_pooled <- SuperLearner(Y ~ A + W, data = pooled_data)
# g_pooled <- SuperLearner(A ~ W, data = pooled_data)

# 3. Estimate bias working model: E[Y | A, W, S=1] - E[Y | A, W, S=0]
# Use cross-validation to select the best bias model
# If bias is zero (external data unbiased), maximum precision gain
# If bias is complex, falls back to RCT-only estimate

# 4. Construct A-TMLE estimator
# The targeting step adjusts for both confounding and bias simultaneously
```


## Putting It All Together: When to Use Which {#sec-when-to-use}

::: {.callout-tip}
## Decision Guide

**Start with CV-TMLE as your default.** It is never worse than standard TMLE and provides valid inference with aggressive machine learning. Then consider:

| Scenario | Recommended variant | Why |
|:---------|:-------------------|:----|
| Standard analysis with Super Learner | **CV-TMLE** | Removes Donsker condition; safe default |
| Near-positivity violations or extreme weights | **C-TMLE** | Propensity score guided by outcome model; avoids instability |
| High-dimensional covariates (100s+) | **Scalable C-TMLE** (LASSO-C-TMLE) | $O(p)$ complexity; automatic variable selection in propensity score |
| Small RCT + large external database | **A-TMLE** | Data-adaptive borrowing with no-harm guarantee |
| Simple analysis, parametric models only | **Standard TMLE** | Donsker condition satisfied by construction; simplest implementation |

These variants are not mutually exclusive. CV-C-TMLE (cross-validated collaborative TMLE) combines the sample-splitting of CV-TMLE with the collaborative propensity score selection of C-TMLE.
:::


## Comparison Table: Standard TMLE, CV-TMLE, C-TMLE, A-TMLE

| Feature | Standard TMLE | CV-TMLE | C-TMLE | A-TMLE |
|:--------|:-------------|:--------|:-------|:-------|
| **Donsker condition required?** | Yes | No | Yes (but less critical) | No (uses CV) |
| **Handles aggressive ML?** | With care | Yes | With care | Yes |
| **Addresses positivity problems?** | No | No | Yes | No |
| **Borrows external data?** | No | No | No | Yes |
| **Substitution estimator?** | Yes | Yes | Yes | Yes |
| **Computational cost** | Low | Moderate (V-fold) | Moderate--High | High |
| **R packages** | `tmle` | `tmle3` (tlverse) | `ctmle` | Custom / tlverse |
| **Key reference** | van der Laan & Rose 2011 | Zheng & van der Laan 2010 | van der Laan & Gruber 2010 | van der Laan et al. 2024 |


## Check Your Understanding

::: {.callout-caution}
## Questions

1. **CV-TMLE vs. standard TMLE**: You are using a Super Learner with 20 candidate algorithms including neural networks and boosted trees. Should you use standard TMLE or CV-TMLE? Why?

2. **C-TMLE intuition**: A colleague says "I should always make my propensity score model as flexible as possible to remove confounding bias." Explain why this can backfire, and how C-TMLE addresses the problem.

3. **CV-TMLE vs. DML**: Both CV-TMLE and double machine learning use cross-fitting. What is the key structural difference, and when does it matter?

4. **A-TMLE**: You have a 200-patient single-arm trial and a 50,000-patient claims database. Under what conditions would A-TMLE provide large efficiency gains? Under what conditions would it fall back to the trial-only estimate?
:::

::: {.callout-tip collapse="true"}
## Answers

1. **CV-TMLE**. With 20 candidates including neural networks and boosted trees, the Super Learner ensemble is highly data-adaptive and likely violates the Donsker condition. Standard TMLE's influence-curve-based variance estimator may be anti-conservative, leading to confidence intervals that are too narrow. CV-TMLE eliminates this concern by fitting nuisance models on training folds and targeting on validation folds.

2. A maximally flexible propensity score will fit the data perfectly, but estimated propensity scores near 0 or 1 create enormous clever covariates $H = A/g(W)$, making the targeting step unstable. The resulting TMLE estimate may have enormous variance even if the propensity score model is "better" in terms of treatment prediction. C-TMLE addresses this by selecting propensity score complexity based on the *causal estimate's performance*, not treatment prediction performance. It may choose a simpler propensity score that produces a better-behaved TMLE estimate.

3. CV-TMLE is a **substitution estimator**: it updates the outcome model and plugs it in, so estimates always respect natural bounds (e.g., probabilities in [0,1]). DML is an **estimating equation estimator**: it adds a bias-correction term to a plug-in estimate, which can produce values outside the parameter space. This matters most when outcomes are rare (probabilities near 0), when sample sizes are small, or when the treatment effect is bounded.

4. **Large gains**: When the observational data come from a population similar to the trial, with the same treatment-outcome relationship conditional on measured covariates (i.e., the bias term is small or zero). In the extreme, if the external data are unbiased, A-TMLE effectively adds 50,000 controls to the 200-patient trial. **Falls back**: When the observational data have large, complex unmeasured confounding that cannot be captured by the bias working model. A-TMLE's cross-validated model selection will find that no bias model improves on the intercept (which corresponds to ignoring external data), and the estimator reduces to the trial-only analysis.
:::


## Sources and Further Reading

- @zheng2010cvtmle: Foundational theory for CV-TMLE; proves asymptotic linearity and efficiency without the Donsker condition
- @levy2018easy: Simplified implementation of CV-TMLE by stacking cross-validated predictions
- @vanderlaan2010ctmle: Original C-TMLE paper; collaborative propensity score estimation guided by the outcome model
- @ju2017scalable: Scalable C-TMLE with pre-ordering strategies for high-dimensional data; the `ctmle` R package
- @vanderlaan2024atmle: A-TMLE for integrating RCT and real-world data with the no-harm guarantee
- @chernozhukov2018dml: Double/debiased machine learning; the estimating-equation counterpart to CV-TMLE
- @vanderLaan2011targeted: Targeted Learning textbook; comprehensive treatment of all TMLE variants
