---
subtitle: "Why standard regression fails when treatment and confounders co-evolve, and what to do about it"
format: html
---

# Longitudinal Causal Inference and Time-Dependent Confounding {#sec-longitudinal-confounding}

*Estimated reading time: 30 minutes*

::: {.callout-note}
## Learning objectives
- Explain why time-varying confounders affected by prior treatment break standard regression, GEE, and mixed models
- Identify the treatment--confounder feedback loop in a DAG and distinguish confounding paths from causal mediated paths
- State the sequential versions of consistency, exchangeability, and positivity
- Describe why G-methods (G-computation, IPTW, L-TMLE) are needed and how they differ from regression
- Understand why static interventions may violate positivity and how stochastic interventions address this
:::

```{r}
#| include: false
#| code-fold: true
library(dagitty)
library(ggdag)
library(ggplot2)
library(tidyverse)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

## Why Longitudinal Treatments Are Different

Parts I and II dealt with point-treatment settings. Real clinical care is not so tidy. Patients start medications, clinicians monitor their health, and treatment gets adjusted in response to what the monitoring shows. That loop, treatment shaping health and health shaping the next treatment decision, repeats at every clinic visit, and it is what makes the longitudinal problem harder than the point-treatment case. This chapter explains why time-varying confounding affected by prior treatment breaks standard regression, lays out the identification assumptions, and previews the two estimation approaches developed in the next two chapters, L-TMLE and LMTP.

## Treatment-confounder feedback in the HIV cohort

Suppose we want the effect of sustained antiretroviral therapy (ART) on virologic failure over two years in an HIV cohort. At each visit, the clinician measures CD4 cell count and weighs it against adherence support and side-effect tolerance before deciding whether to continue or modify therapy. CD4 wears two hats at once. It is a **confounder** for future treatment, since low CD4 prompts more aggressive intervention and also predicts virologic failure, and it is a **mediator** of past treatment, since ART raises CD4 by suppressing viral replication. The DAG makes this explicit.

```{r}
#| label: fig-hiv-feedback-intro
#| fig-cap: "Time-varying confounding in a two-year HIV/ART study. CD4 at the end of year 1 is both a confounder of year-2 treatment and a mediator of year-1 treatment."
#| fig-width: 8
#| fig-height: 3.5
#| code-fold: true
dag_hiv_intro <- dagify(
  A1 ~ W, L1 ~ A1 + W, A2 ~ L1 + W, Y ~ A1 + A2 + L1 + W,
  exposure = "A1", outcome = "Y",
  labels = c(W = "Baseline\n(age, sex, CD4_0)", A1 = "ART\nyear 1",
             L1 = "CD4\nend of year 1", A2 = "ART\nyear 2",
             Y = "Virologic\nfailure by year 2"),
  coords = list(
    x = c(W = 0, A1 = 1, L1 = 2, A2 = 3, Y = 4),
    y = c(W = 0.8, A1 = 0, L1 = 0.8, A2 = 0, Y = 0)
  )
)
ggdag(dag_hiv_intro, text = FALSE, use_labels = "label", text_size = 2.8) +
  theme_dag()
```

Adjusting for CD4 in a regression blocks the causal pathway $A_1 \to L_1 \to Y$. Leaving it out leaves $A_2$ confounded. A time-varying covariate that is both a confounder for future treatment and a mediator of past treatment simply cannot be handled by standard regression. The simulation below shows this with a stylised two-period HIV cohort.

```{r}
set.seed(2025)
n <- 5000
W  <- rnorm(n)                         # baseline severity (e.g., baseline CD4, lower = sicker)
A1 <- rbinom(n, 1, plogis(0.3 * W))    # ART adherence year 1
L1 <- rnorm(n, 0.5 * A1 + 0.4 * W)     # CD4 end of year 1: improved by ART
A2 <- rbinom(n, 1, plogis(0.5 * L1))   # ART adherence year 2 driven by current CD4
Y  <- rbinom(n, 1, plogis(-1.5 + 0.4 * A1 + 0.4 * A2 + 0.3 * L1 + 0.3 * W))

# The quantity we want is the TOTAL effect of sustained adherence on the risk
# scale: failure risk if everyone always adhered (A1 = A2 = 1) minus if no one
# did (A1 = A2 = 0). Simulate it from the known mechanism, letting CD4 (L1)
# respond to year-1 treatment the way it does in the data-generating process.
risk_under <- function(a) {
  L1a <- rnorm(n, 0.5 * a + 0.4 * W)   # CD4 responds to the assigned regime
  mean(plogis(-1.5 + 0.4 * a + 0.4 * a + 0.3 * L1a + 0.3 * W))
}
true_total_rd <- risk_under(1) - risk_under(0)

# A regression that "controls for CD4" conditions on L1, which sits on the
# A1 -> L1 -> Y pathway. Standardizing its predictions holds CD4 at its observed
# values instead of letting treatment move it, so the mediated part of the
# effect is discarded.
naive_mod <- glm(Y ~ A1 + A2 + L1 + W, family = binomial)
dat_naive <- data.frame(A1, A2, L1, W)
p_treat   <- predict(naive_mod, transform(dat_naive, A1 = 1, A2 = 1), type = "response")
p_control <- predict(naive_mod, transform(dat_naive, A1 = 0, A2 = 0), type = "response")
naive_rd  <- mean(p_treat) - mean(p_control)

cat("True total effect of sustained ART (risk difference):", round(true_total_rd, 3), "\n")
cat("CD4-adjusted regression estimate:                    ", round(naive_rd, 3), "\n")
cat("Direct-effect-only A1 coefficient (logit scale):     ", round(coef(naive_mod)["A1"], 3), "\n")
```

The regression that conditions on CD4 recovers only the *direct* effect of ART: its $A_1$ coefficient sits near 0.4, the direct structural parameter, and the standardized risk difference it implies falls short of the true total effect. Conditioning on $L_1$ holds the mediator fixed, so the part of the effect that runs through CD4 recovery is discarded. This is structural bias, not sampling noise. It persists at any sample size, because the problem is which pathway the model blocks, not how much data we have.


# The HIV Cohort: Our Running Example

CD4 count is the canonical time-dependent confounder in the HIV literature [@petersen2007msm; @vanderwal2011ipw], and we will return to it throughout the next several chapters. Building on the toy two-period DAG above, here is the clinical question we are working toward:

> What is the probability of virologic failure by 12 months under sustained high adherence to ART, compared to observed adherence patterns, adjusting for time-varying CD4 count?

The data structure at each clinic visit looks like this:

| Visit | What is measured | Variable role |
|-------|-----------------|---------------|
| Baseline | Age, sex, baseline CD4, baseline viral load | Fixed covariates $W$ |
| Visit $t$ | CD4 count | Time-varying confounder $L_t$ |
| Visit $t$ | Adherence to ART | Treatment $A_t$ |
| Visit $t$ | Still in study? | Censoring $C_t$ |
| End of follow-up | Virologic failure (HIV RNA $\geq$ 200 copies/mL) | Outcome $Y$ |

At each visit, CD4 is measured first, then adherence is recorded, then we check whether the patient is still under observation. The ordering is not arbitrary. It encodes our assumptions about the temporal sequence, and everything downstream depends on getting it right.

ART adherence at visit $t-1$ improves CD4 at visit $t$. Higher CD4 may in turn reduce clinical urgency and change adherence decisions, and CD4 also directly affects virologic failure risk. This is the same feedback structure as the toy two-period example above, scaled up to many time points.


# Why Standard Regression Fails

We have already seen the problem in the two-period HIV example, but it is worth spelling out precisely what goes wrong and why.

## The treatment--confounder feedback loop

When a time-varying covariate $L_t$ is (1) affected by prior treatment $A_{t-1}$ and (2) confounds the relationship between $A_t$ and $Y$, we say there is **treatment--confounder feedback**. In HIV, CD4 count satisfies both conditions: ART raises CD4, and CD4 predicts both future adherence decisions and virologic failure.

```{r}
#| label: fig-hiv-feedback
#| fig-cap: "Treatment--confounder feedback in an HIV cohort. CD4 count at each visit is affected by prior ART adherence and also confounds the relationship between future adherence and the outcome."
#| fig-width: 9
#| fig-height: 4
#| code-fold: true
dag_hiv <- dagify(
  A1 ~ W, L1 ~ A1 + W, A2 ~ L1 + A1 + W, Y ~ A1 + A2 + L1 + W,
  exposure = "A1", outcome = "Y",
  labels = c(W = "Baseline\n(age, sex, CD4_0)", A1 = "ART adherence\nvisit 1",
             L1 = "CD4 count\nvisit 1", A2 = "ART adherence\nvisit 2",
             Y = "Virologic\nfailure"),
  coords = list(
    x = c(W = 0, A1 = 1, L1 = 2, A2 = 3, Y = 4),
    y = c(W = 1, A1 = 0, L1 = 0.8, A2 = 0, Y = 0)
  )
)
ggdag(dag_hiv, text = FALSE, use_labels = "label", text_size = 2.8) +
  theme_dag()
```

## The dilemma

A standard regression model, whether logistic, Cox, or GEE, faces a lose-lose choice. Adjust for $L_t$ and you block the causal pathway $A_{t-1} \to L_t \to Y$. Omit $L_t$ and you leave $A_t$ confounded. No regression specification manages to deconfound future treatment and preserve the mediated effect of past treatment at the same time.

## Seeing the dilemma step by step

The following DAG panels isolate each component using a simplified two-period version of the HIV DAG.

```{r}
#| label: dag-step-setup
#| include: false
#| code-fold: true

# Tableau 10 colour palette
tab10 <- c(
  blue   = "#4E79A7", orange = "#F28E2B", red    = "#E15759",
  teal   = "#76B7B2", green  = "#59A14F", yellow = "#EDC948",
  purple = "#B07AA1", pink   = "#FF9DA7", brown  = "#9C755F",
  grey   = "#BAB0AC"
)

# Two-period DAG: W -> A1 -> L -> A2 -> Y, with feedback
library(dagitty)
library(ggdag)

dag_ch <- dagitty('dag {
  W   [pos="0,0.8"]
  A1  [pos="1,0"]
  L2  [pos="2,0.8"]
  A2  [pos="3,0"]
  Y   [pos="4,0"]

  W -> A1
  W -> L2
  W -> A2
  W -> Y
  A1 -> L2
  A1 -> Y
  L2 -> A2
  L2 -> Y
  A2 -> Y
}')

dag_ch_tidy <- tidy_dagitty(dag_ch)
dag_ch_df   <- dag_ch_tidy$data

# Node roles for fill colour
node_roles_ch <- c(W = "baseline", A1 = "treatment", L2 = "confounder",
                   A2 = "treatment", Y = "outcome")
dag_ch_df$role <- node_roles_ch[dag_ch_df$name]

role_fills_ch <- c("baseline"   = "#F0F0F0",
                   "confounder" = "#FEE0D2",
                   "treatment"  = "#DEEBF7",
                   "outcome"    = "#E5F5E0")

# Helper: draw the DAG with optional edge highlighting
draw_dag_step <- function(highlight_pairs = list(),
                          highlight_colour = tab10[["red"]],
                          df = dag_ch_df) {
  edges_all <- df[!is.na(df$to), ]
  edges_all$highlighted <- FALSE
  for (hp in highlight_pairs) {
    idx <- edges_all$name == hp[1] & edges_all$to == hp[2]
    edges_all$highlighted[idx] <- TRUE
  }
  has_hl <- length(highlight_pairs) > 0

  xrng <- range(df$x, na.rm = TRUE)
  yrng <- range(df$y, na.rm = TRUE)
  pad  <- 0.6

  ggplot(df, aes(x = x, y = y)) +
    geom_segment(data = edges_all,
                 aes(xend = xend, yend = yend),
                 colour = if (has_hl) "grey70" else "grey30",
                 linewidth = if (has_hl) 0.4 else 0.6,
                 arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
    {if (has_hl)
      geom_segment(data = edges_all[edges_all$highlighted, ],
                   aes(xend = xend, yend = yend),
                   colour = highlight_colour, linewidth = 1.0,
                   arrow = arrow(length = unit(0.12, "inches"), type = "closed"))
    } +
    geom_point(aes(fill = role), size = 16, shape = 21,
               colour = "grey30", stroke = 0.6) +
    geom_text(aes(label = name), size = 3.5, fontface = "bold") +
    scale_fill_manual(values = role_fills_ch, guide = "none") +
    coord_cartesian(xlim = c(xrng[1] - pad, xrng[2] + pad),
                    ylim = c(yrng[1] - pad, yrng[2] + pad),
                    clip = "off") +
    theme_dag(base_size = 14)
}
```

### Step 1: The full causal structure

```{r}
#| label: fig-dag-step1
#| fig-cap: "Step 1: Full causal structure. All paths shown, no highlighting."
#| fig-width: 8
#| fig-height: 3.5
#| code-fold: true
draw_dag_step(highlight_pairs = list())
```

### Step 2: Why adjusting for $L_2$ fails

Conditioning on $L_2$ (shown by the box around the node) blocks the confounding paths $L_2 \to A_2$ and $L_2 \to Y$, which is what we want. But it *also* blocks the causal mediated path $A_1 \to L_2 \to Y$, which we wanted to keep open. The edge $L_2 \to Y$ sits on *both* paths at once: it is part of the confounding problem and part of the causal effect we are trying to estimate. That double duty is why no regression specification can solve both problems at once.

```{r}
#| label: fig-dag-step2
#| fig-cap: "Step 2: Why adjusting for L2 fails. Red = confounding paths we want to block. Blue = causal mediated path also blocked by conditioning on L2. Purple = L2 -> Y is on BOTH paths. The box around L2 represents conditioning."
#| fig-width: 8
#| fig-height: 4.5
#| code-fold: true

edges_step4 <- dag_ch_df[!is.na(dag_ch_df$to), ]
edges_step4$step4_type <- "dim"

# Confounding: L2 -> A2
for (hp in list(c("L2", "A2"))) {
  idx <- edges_step4$name == hp[1] & edges_step4$to == hp[2]
  edges_step4$step4_type[idx] <- "confounding"
}
# Causal mediated: A1 -> L2
for (hp in list(c("A1", "L2"))) {
  idx <- edges_step4$name == hp[1] & edges_step4$to == hp[2]
  edges_step4$step4_type[idx] <- "causal_mediated"
}
# L2 -> Y is on BOTH paths
idx_both <- edges_step4$name == "L2" & edges_step4$to == "Y"
edges_step4$step4_type[idx_both] <- "both"

xrng <- range(dag_ch_df$x, na.rm = TRUE)
yrng <- range(dag_ch_df$y, na.rm = TRUE)

ggplot(dag_ch_df, aes(x = x, y = y)) +
  # Layer 1: All edges in grey
  geom_segment(data = edges_step4,
               aes(xend = xend, yend = yend),
               colour = "grey70", linewidth = 0.4,
               arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  # Layer 2: Confounding paths (red)
  geom_segment(data = edges_step4[edges_step4$step4_type == "confounding", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["red"]], linewidth = 1.0,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed")) +
  # Layer 3: Causal mediated paths (blue)
  geom_segment(data = edges_step4[edges_step4$step4_type == "causal_mediated", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["blue"]], linewidth = 1.0,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed")) +
  # Layer 4: Both paths (purple -- L2 -> Y)
  geom_segment(data = edges_step4[edges_step4$step4_type == "both", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["purple"]], linewidth = 1.1,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed")) +
  # Nodes
  geom_point(aes(fill = role), size = 16, shape = 21,
             colour = "grey30", stroke = 0.6) +
  geom_text(aes(label = name), size = 3.5, fontface = "bold") +
  scale_fill_manual(values = role_fills_ch, guide = "none") +
  # Conditioning box around L2
  annotate("rect",
           xmin = dag_ch_df$x[dag_ch_df$name == "L2"] - 0.45,
           xmax = dag_ch_df$x[dag_ch_df$name == "L2"] + 0.45,
           ymin = dag_ch_df$y[dag_ch_df$name == "L2"] - 0.35,
           ymax = dag_ch_df$y[dag_ch_df$name == "L2"] + 0.35,
           fill = NA, colour = "black", linewidth = 1.2, linetype = "solid") +
  # Legend annotations
  annotate("segment", x = 0.3, xend = 0.9, y = -0.9, yend = -0.9,
           colour = tab10[["red"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.0, y = -0.9,
           label = "Confounding path (want to block)", hjust = 0, size = 3) +
  annotate("segment", x = 0.3, xend = 0.9, y = -1.15, yend = -1.15,
           colour = tab10[["blue"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.0, y = -1.15,
           label = "Causal path through L2 (also blocked!)", hjust = 0, size = 3) +
  annotate("segment", x = 0.3, xend = 0.9, y = -1.4, yend = -1.4,
           colour = tab10[["purple"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.0, y = -1.4,
           label = "L2 -> Y: on BOTH paths (the dilemma)", hjust = 0, size = 3) +
  coord_cartesian(xlim = c(xrng[1] - 0.6, xrng[2] + 0.6),
                  ylim = c(-1.7, yrng[2] + 0.6),
                  clip = "off") +
  theme_dag(base_size = 14)
```

### Step 3: The feedback loop

The reason $L_2$ is both a confounder and a mediator is the **feedback loop**: $A_1 \to L_2 \to A_2$. Prior treatment affects the time-varying confounder, which in turn affects future treatment.

```{r}
#| label: fig-dag-step3
#| fig-cap: "Step 3: The feedback loop. A1 -> L2 -> A2 is the path that makes L2 simultaneously a confounder and a mediator."
#| fig-width: 8
#| fig-height: 3.5
#| code-fold: true
draw_dag_step(
  highlight_pairs = list(
    c("A1", "L2"),   # treatment affects future confounder
    c("L2", "A2"),   # confounder affects future treatment
    c("L2", "Y")     # confounder affects outcome
  ),
  highlight_colour = tab10[["orange"]]
)
```

G-methods (G-computation, IPTW, L-TMLE) get around this by integrating over the distribution of $L_2$ under the hypothetical intervention rather than conditioning on it.

## What about GEE, mixed models, or covariate-adjusted regression?

GEE, mixed models, and covariate-adjusted regression all condition on time-varying covariates the same way standard regression does, so none of them escape the treatment--confounder feedback problem. GEE and mixed models handle within-subject correlation, but correlation is a separate issue from the causal structure of feedback. A regression that omits time-varying covariates sidesteps the mediator-blocking problem, but it then targets a different estimand: a regression oracle that never integrates over the interventional distribution of $L_t$. When feedback is weak, that oracle can happen to agree with the causal truth, but the agreement falls apart once feedback is strong. One useful diagnostic is to ask what each method converges to under infinite data. If the regression oracle differs from the G-computation oracle, the two methods are estimating fundamentally different quantities.


# Counterfactual Notation for Longitudinal Interventions

To formalize the causal question, we need notation for hypothetical treatment sequences.

Let $\bar{A} = (A_1, A_2, \ldots, A_T)$ denote the observed treatment history across $T$ visits, and let $\bar{L} = (L_1, \ldots, L_T)$ denote the observed time-varying confounders. Let $\bar{a} = (a_1, \ldots, a_T)$ denote a specific hypothetical treatment sequence. The potential outcome $Y(\bar{a})$ is the outcome that would have been observed if the patient had followed treatment sequence $\bar{a}$, regardless of what actually happened.

A **static regime** assigns the same treatment at every visit regardless of patient history. For instance, $\bar{a} = (1, 1, \ldots, 1)$ means "always maintain high adherence." A **dynamic regime** lets the treatment at each visit depend on the evolving covariate history: $a_t = d_t(\bar{L}_t, \bar{A}_{t-1}, W)$. A **stochastic regime** does not fix a specific treatment value at all. Instead it modifies the probability of treatment, for example raising the probability of high adherence by 20 percentage points at each visit.

The canonical estimand in the HIV example is the average counterfactual outcome under sustained high adherence versus the observed adherence pattern:

$$
\psi = E\big[Y(\bar{a} = \bar{1})\big] - E\big[Y^{\text{obs}}\big]
$$

where $Y(\bar{1})$ is the virologic failure probability if everyone had maintained high adherence at every visit.


# Identification Assumptions

The potential outcome $Y(\bar{a})$ is a counterfactual quantity. To estimate it from observed data, we need assumptions that bridge what we observe and what would have happened under a hypothetical intervention. In the longitudinal setting, the familiar point-treatment assumptions have to hold at every time point rather than just once.

**Consistency.** If a patient's observed treatment history matches the hypothetical sequence $\bar{a}$, then their observed outcome equals their potential outcome: $Y = Y(\bar{a})$ when $\bar{A} = \bar{a}$. This requires that the treatment is well-defined and that there are no hidden versions of treatment.

**Sequential exchangeability.** At each time $t$, treatment is unconfounded given the observed history:
$$
Y(\bar{a}) \perp\!\!\!\perp A_t \mid \bar{L}_t, \bar{A}_{t-1}, W \quad \text{for all } t.
$$
This says that, conditional on everything we have measured up to visit $t$, treatment assignment at visit $t$ is as good as random. In practice, this requires measuring every confounder that influences both the treatment decision and the outcome at every visit.

Unmeasured confounders (substance use severity, housing instability, clinic-specific prescribing culture) that affect both adherence and virologic failure would violate this assumption.

**Sequential positivity.** Every treatment level must be possible at every visit for every observed covariate history:
$$
P(A_t = a \mid \bar{L}_t, \bar{A}_{t-1}, W) > 0 \quad \text{for all } t, a.
$$
If some patient subgroups essentially never achieve high adherence (for example, patients with severe side effects on a particular regimen), then the counterfactual $Y(\bar{1})$ is not well-defined for those subgroups without extrapolation.

All three assumptions are harder to satisfy in the longitudinal setting. [Sequential exchangeability](glossary.qmd) must hold at every visit, and positivity must hold for every covariate history. This is a large part of why stochastic regimes, which relax the positivity requirement, are often preferred over deterministic "always treat" regimes.


# Why G-Methods Are Needed

Under the assumptions above, the causal quantity $E[Y(\bar{a})]$ can be expressed as a function of the observed data distribution through the **G-computation formula** [@hernan2025whatif]:

$$
E[Y(\bar{a})] = \sum_{\bar{l}} E[Y \mid \bar{A} = \bar{a}, \bar{L} = \bar{l}, W] \;\prod_t P(L_t = l_t \mid \bar{A}_{t-1} = \bar{a}_{t-1}, \bar{L}_{t-1} = \bar{l}_{t-1}, W)
$$

The product term generates the interventional distribution of covariate trajectories, meaning what CD4 counts would look like under regime $\bar{a}$, and the outer expectation gives the outcome under those trajectories. Standard regression conditions on time-varying covariates instead of integrating over their interventional distribution, which is exactly the difference that matters here. The class of methods that can compute this quantity, namely G-computation, IPTW, and TMLE, are collectively called **g-methods**.

The three main g-methods were each introduced for the point-treatment case in @sec-gcomputation, @sec-iptw, and @sec-doubly-robust-tmle. Each brings its own strengths and limitations to the longitudinal setting:

| Method | Core idea | Main limitation |
|--------|-----------|----------------|
| G-computation | Sequential outcome regressions, working backwards | Relies on correct outcome model specification |
| IPTW (marginal structural models) [@robins1994estimation; @petersen2007msm] | Reweight by inverse of cumulative treatment probability | Cumulative weights become extreme with many time points |
| Longitudinal TMLE | Sequential regressions plus targeting steps | More complex to implement, but doubly robust |

IPTW was historically the first approach applied to HIV data [@robins1994estimation], but it grows unstable as the number of visits climbs, because the weights are products of time-specific inverse probabilities and those products can become enormous. Longitudinal TMLE and LMTP tame this instability by using substitution estimators with targeting steps.


# Preview of L-TMLE and LMTP

The next two chapters develop two estimation approaches in detail, each suited to a different type of causal question.

## L-TMLE for standard longitudinal interventions

@sec-illustrated-ltmle introduces **longitudinal TMLE (L-TMLE)**, which runs the TMLE "model, then target" cycle backwards through time. At each step it conditions on $L_t$ to deconfound future treatment, then integrates $L_t$ out in the backwards regression so the mediated effect of past treatment survives. The `ltmle` R package implements this procedure.

## LMTP for realistic adherence interventions

@sec-adherence-shift presents the **LMTP framework** (longitudinal modified treatment policies), which defines interventions in which the hypothetical treatment depends on the patient's natural treatment value. This includes stochastic shifts that sidestep the positivity problems baked into deterministic regimes. The `lmtp` R package implements this framework.

Both approaches build on the identification assumptions in this chapter. L-TMLE handles standard static or dynamic regimes, while LMTP accommodates more flexible modified treatment policies.

For a detailed treatment of realistic adherence interventions and how LMTP addresses the positivity limitations of static regimes, see the [adherence chapter](03-03b-adherence-shift-methods.qmd).

## What comes next

The next chapter walks through L-TMLE step by step using a two-time-point HIV cohort example with the `ltmle` package.

---

## Check your understanding

<details>
<summary><strong>Q1. Why can't you simply add time-varying confounders as covariates in a Cox regression when those confounders are affected by prior treatment?</strong></summary>

Because a time-varying confounder that is affected by prior treatment sits on the causal pathway between earlier treatment and the outcome. Adjust for it in a standard regression and you block part of the treatment effect (overadjustment bias). Omit it and you leave uncontrolled confounding. This is the treatment-confounder feedback problem, and G-methods get around it by integrating over the interventional distribution of the time-varying confounder rather than conditioning on it.

</details>

<details>
<summary><strong>Q2. What does "sequential exchangeability" require that ordinary exchangeability does not?</strong></summary>

Ordinary (point-treatment) exchangeability requires that treatment is unconfounded conditional on baseline covariates measured at a single time point. Sequential exchangeability requires the same thing at every time point, conditional on the entire covariate and treatment history up to that point. In practice that means measuring every confounder that influences both treatment decisions and the outcome *at every visit*, a much harder bar to clear, especially when physician judgment, symptom severity, or patient preferences evolve over time and may never make it into the data.

</details>


---

## Sources and further reading

- [@robins1994estimation]: foundational paper on marginal structural models and IPTW for time-varying treatments.
- [@petersen2007msm]: targeted TMLE estimation for dynamic and static longitudinal marginal structural working models.
- [@hernan2025whatif]: textbook treatment of longitudinal causal inference and g-methods.
- [@diaz2021lmtp]: lmtp R package for modified treatment policies.
- [@vanderwal2011ipw]: HAART observational cohort used as the running example.
