---
subtitle: "How backwards sequential regression solves treatment-confounder feedback"
format: html
---

# Longitudinal TMLE: Concepts and Intuition {#sec-illustrated-ltmle}

*Estimated reading time: 30 minutes*

::: {.callout-note}
## Learning objectives
- Understand the backwards sequential regression logic that underpins L-TMLE
- Explain why conditioning on the time-varying confounder is needed at one time step but harmful at another
- Describe the role of targeting (fluctuation steps) in the longitudinal setting
- Recognize when static regimes violate positivity and when LMTP offers a better framework
:::

```{r}
#| include: false
#| code-fold: true
library(dagitty)
library(ggdag)
library(ggplot2)
library(tidyverse)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
logit <- function(p) log(p / (1 - p))
```


# Where this chapter fits

The previous chapter explained why treatment-confounder feedback breaks standard regression. This chapter explains how L-TMLE [@vanderLaan2012ltmle; @petersen2014causal] solves the problem conceptually. The focus is on the logic of the algorithm, not a full code walkthrough. Readers who want the complete by-hand implementation can consult the [technical companion chapter](03-09b-ltmle-by-hand.qmd). For practical applied analysis of realistic longitudinal interventions, the LMTP chapters that follow are the recommended pathway.


# The causal question

We use a two-time-point HIV cohort [@vanderwal2011ipw] to make the mechanics visible. The question: what is the probability of virologic failure if every patient maintained high ART adherence at each visit, compared to never maintaining high adherence?

This is a comparison of two static regimes, $\bar{a} = (1,1)$ versus $\bar{a} = (0,0)$, with treatment-confounder feedback through CD4 count. We use two time points for clarity; the same logic extends to any number of visits.

```{r}
#| label: fig-hiv-node-order
#| fig-cap: "Node ordering for the two-time-point HIV cohort."
#| fig-width: 9
#| fig-height: 3.5
#| code-fold: true
dag_hiv <- dagify(
  A1 ~ W, L1 ~ A1 + W, A2 ~ L1 + A1 + W, Y ~ A1 + A2 + L1 + W,
  exposure = "A1", outcome = "Y",
  labels = c(W = "Baseline\n(age, sex, CD4_0, VL_0)",
             A1 = "ART adherence\nvisit 1",
             L1 = "CD4 count\nvisit 1",
             A2 = "ART adherence\nvisit 2",
             Y = "Virologic\nfailure"),
  coords = list(
    x = c(W = 0, A1 = 1, L1 = 2, A2 = 3, Y = 4),
    y = c(W = 1, A1 = 0, L1 = 0.8, A2 = 0, Y = 0)
  )
)
ggdag(dag_hiv, text = FALSE, use_labels = "label", text_size = 2.8) +
  theme_dag()
```


# Simulated data

```{r}
set.seed(2026)
# n = 1000 is large enough to reproduce the longitudinal-confounding
# behaviour that motivates LTMLE while keeping the chapter fast in the
# WebR (interactive) build. Production analyses can use any n.
n <- 1000

W_age    <- rnorm(n, mean = 40, sd = 10)
W_male   <- rbinom(n, 1, 0.6)
W_cd4_bl <- pmax(50, rnorm(n, mean = 350, sd = 120))
W_vl_bl  <- rnorm(n, mean = 4.2, sd = 0.7)
W <- 0.01 * (W_age - 40) - 0.001 * (W_cd4_bl - 350) + 0.2 * W_vl_bl - 0.1 * W_male

A1 <- rbinom(n, 1, plogis(0.3 + 0.5 * W))
L1 <- rnorm(n, mean = 0.3 * W + 1.2 * A1, sd = 1)  # <1>
A2 <- rbinom(n, 1, plogis(-0.5 + 0.8 * L1 + 0.3 * A1))  # <2>
Y  <- rbinom(n, 1, plogis(-0.8 - 0.4 * A1 - 0.5 * A2 - 0.6 * L1 + 0.2 * W))  # strong L1 effect

dat <- tibble(W, A1, L1, A2, Y)
cat("Failure rate:", round(mean(Y), 3), "\n")
```
1. A1 causes L1 (coefficient 1.2): prior adherence improves the intermediate biomarker. This is the *first half* of treatment-confounder feedback — L1 is a consequence of past treatment.
2. L1 strongly predicts A2 (coefficient 0.8 on L1): a worse biomarker at visit 1 reduces the probability of adhering at visit 2. This closes the loop — L1 is simultaneously a *consequence* of A1 and a *confounder* of A2, making a single regression on all time points invalid.


# The backwards regression idea

We solve the maze from the exit. Start at the final outcome and work back toward the first treatment decision, so each step already knows where it needs to end up. Concretely, L-TMLE estimates the causal effect through three ideas working together.

**Sequential regressions, working backwards.** Start at the final outcome and model $E[Y \mid A_2, L_1, A_1, W]$. This conditions on $L_1$ to deconfound $A_2$. Then move backwards: take the predicted outcomes under $A_2 = 1$ and regress them on $(A_1, W)$ only, without conditioning on $L_1$. This integrates $L_1$ out, preserving the mediated effect of $A_1$ through $L_1$.

**Treatment mechanism models.** Estimate the probability of receiving treatment at each visit conditional on history: $\hat{g}_1(W)$ and $\hat{g}_2(A_1, L_1, W)$.

**Targeting steps.** Use the treatment mechanism to construct clever covariates that correct the sequential regressions for any remaining bias. At each time point, the final estimate is consistent if either the outcome regression or the treatment model is correctly specified (sequential double robustness).

The key insight: conditioning on $L_1$ at time 2 deconfounds $A_2$, while NOT conditioning on $L_1$ at time 1 preserves the indirect effect $A_1 \to L_1 \to Y$. Standard regression cannot do both in a single model.

@fig-backwards-regression shows the two-step logic. The same variable $L_1$ is handled differently at each step: it is included in the visit-2 regression, where it confounds $A_2$, and omitted from the visit-1 regression, where it carries part of the effect of $A_1$.

```{mermaid}
%%| label: fig-backwards-regression
%%| fig-cap: "Backwards sequential regression for two time points. The time-varying covariate L1 is included where it confounds (visit 2) and omitted where it mediates (visit 1). Including L1 at visit 1 would block the pathway A1 to L1 to Y and bias the estimate, which is the error a single standard regression makes."
flowchart TD
    Y["Observed outcome Y"] --> Q2["<b>Step 1: visit-2 regression</b><br/>Model E[Y | A2, L1, A1, W]<br/><i>include L1</i> to deconfound A2"]
    Q2 --> P2["Predict under A2 = 1<br/>= targeted prediction Q-hat 2"]
    P2 --> Q1["<b>Step 2: visit-1 regression</b><br/>Regress Q-hat 2 on A1 and W only<br/><i>omit L1</i> to keep the path A1 &rarr; L1 &rarr; Y"]
    Q1 --> P1["Predict under A1 = 1, then average<br/>= E[Y(1,1)]"]
    note["L1 confounds A2 but mediates A1.<br/>Include it at visit 2; omit it at visit 1."]
    note -. "include" .-> Q2
    note -. "omit" .-> Q1
```


# How the recipe works in plain language

**Step 1.** At each visit, model how likely each patient was to actually adhere. Sicker patients are less likely to adhere; patients who already missed doses are less likely to catch up.

**Step 2.** Starting at visit 2, predict each patient's failure risk as if they had high adherence. Adjust using the treatment model so that the prediction is calibrated for the causal estimand, not just for prediction accuracy. This is the targeting step.

**Step 3.** Take the adjusted visit-2 predictions and use them as the "outcome" for visit 1. Predict under high adherence at visit 1, and target again. The visit-1 model does not condition on CD4, because CD4 is a consequence of visit-1 adherence.

**Step 4.** Average the final targeted predictions across all patients. This is the L-TMLE estimate of $E[Y(1,1)]$.


# Naive regression vs L-TMLE

A quick demonstration of why the backwards structure matters:

```{r}
# Oracle truth
set.seed(9999)
n_mc <- 100000; W_mc <- rep(W, length.out = n_mc)
L1_cf11 <- rnorm(n_mc, mean = 0.3 * W_mc + 1.2, sd = 1)
true_EY_11 <- mean(plogis(-0.8 - 0.4 - 0.5 - 0.6 * L1_cf11 + 0.2 * W_mc))

# Naive: condition on L1 (blocks A1 -> L1 -> Y)
naive_mod <- glm(Y ~ A1 + A2 + L1 + W, family = binomial, data = dat)
naive_pred <- mean(predict(naive_mod, newdata = dat %>% mutate(A1 = 1, A2 = 1), type = "response"))

cat("True E[Y(1,1)]:", round(true_EY_11, 4), "\n")
cat("Naive (conditions on L1):", round(naive_pred, 4), "\n")
```

The naive estimate is biased because conditioning on $L_1$ blocks part of the causal pathway. L-TMLE avoids this by conditioning on $L_1$ only where it confounds ($A_2$) and integrating it out where it mediates ($A_1$).

::: {.callout-note}
## Full by-hand implementation
The complete L-TMLE algorithm, with code for treatment mechanism estimation, sequential outcome regression, clever covariates, fluctuation models, and estimator comparison, is in the [technical companion chapter](03-09b-ltmle-by-hand.qmd). That chapter walks through every line of the two-time-point implementation.
:::


# When static regimes break down

The "always adhere" regime is pedagogically useful, but it has a practical limitation: no real intervention achieves 100% adherence. When the static regime assigns treatment values that are rare in the observed data, positivity breaks down: propensity scores approach zero, clever covariates explode, and the estimate becomes unstable.

The instability is a property of the estimand, not of L-TMLE. Asking "what if everyone achieved perfect adherence?" requires extrapolating far outside the data for patients who are very unlikely to adhere, which produces high variance and potential bias.

The solution is to define more realistic interventions that stay within the observed data support. The LMTP framework provides this [@diaz2021lmtp; @williams2025lmtp]: instead of forcing a specific treatment value, it shifts the treatment process by a feasible amount. The [LMTP adherence-shift chapter](03-03b-adherence-shift-methods.qmd) develop this approach.


# Package implementation

The `ltmle` R package automates the full L-TMLE procedure. It accepts a Super Learner library [@vanderLaan2007superlearner] for nuisance estimation. For static regimes, the interface is straightforward:

```{r}
#| code-fold: true
#| warning: false
#| message: false
library(ltmle)

dat_ltmle <- data.frame(W = dat$W, A1 = dat$A1, L1 = dat$L1, A2 = dat$A2, Y = dat$Y)

result <- ltmle(
  data = dat_ltmle,
  Anodes = c("A1", "A2"),
  Lnodes = c("L1"),  # <1>
  Ynodes = "Y",
  abar = list(treatment = c(1, 1), control = c(0, 0)),  # <2>
  SL.library = c("SL.glm", "SL.mean")
)

summary(result)
```
1. `Lnodes` declares which columns are time-varying confounders sitting *between* treatment nodes in the causal graph. Omitting this argument makes ltmle treat L1 as a baseline covariate, breaking the sequential g-computation logic and invalidating the estimate under treatment-confounder feedback.
2. `abar` specifies the joint static intervention at every treatment time point. The named list is optional but clarifying: `treatment = c(1, 1)` sets $(A_1, A_2) = (1, 1)$; `control = c(0, 0)` sets both to 0. ltmle returns $E[Y^{(1,1)}] - E[Y^{(0,0)}]$ with a 95% confidence interval.

For realistic interventions estimated with `lmtp`, see the [adherence-shift methods chapter](03-03b-adherence-shift-methods.qmd) and the [applied workflow chapter](05-01-applied-lmtp-workflow.qmd).


## Check your understanding

<details>
<summary><strong>Q1. Why does L-TMLE work backwards through time?</strong></summary>

Working backwards allows L-TMLE to condition on $L_t$ where it confounds future treatment and integrate it out where it mediates past treatment. A forward approach cannot do both simultaneously.

</details>

<details>
<summary><strong>Q2. What is sequential double robustness?</strong></summary>

At each time point, L-TMLE is consistent if either the outcome model or the treatment model is correctly specified. Different models can fail at different time points, as long as at each point at least one is correct.

</details>

<details>
<summary><strong>Q3. When should you use LMTP instead of static L-TMLE?</strong></summary>

When the static regime (e.g., "always adhere") assigns treatment values that are rare in the observed data. LMTP defines feasible shifts that stay within the data support, producing more stable estimates and answering more clinically relevant questions.

</details>

---

## Sources and further reading

- [@petersen2014causal] — targeted TMLE for dynamic and static longitudinal marginal structural working models.
- [@diaz2021lmtp; @williams2025lmtp] — lmtp R package for modified treatment policies and LMTP.
- [@vanderLaan2007superlearner] — Super Learner ensemble for nuisance function estimation.
- [@vanderwal2011ipw] — HAART observational cohort used as the running example.
