---
format: html
---

*Estimated reading time: 55 minutes*

# Staged and Clean-Room Designs for Real-World Evidence {#sec-clean-room}
*Pre-specified targeted learning with outcome-blind diagnostics, plasmode candidate selection, and a data-quality stress test*

::: {.callout-tip}
## If you only read one thing
The credibility of a high-stakes real-world-evidence estimate depends on the process as much as the estimator. Lock the analysis plan and run every diagnostic outcome-blind, so that no analytic choice is made after seeing the treatment-outcome association. This chapter's `cleanTMLE` workflow enforces that discipline with an analysis lock, audit logs, a data-quality stress test, and a GO / FLAG / STOP rule that refuses to read the outcome until the prespecified checkpoints are recorded.
:::

You have specified a target trial, chosen an estimand, defended your DAG, fit a TMLE with Super Learner, and run the diagnostics in @sec-advanced-diagnostics. Should the reviewer believe the estimate? In a regulatory or multi-site RWE setting, the answer increasingly depends not just on the estimator but on the **process** that produced it: was the analysis plan fixed before the outcome was inspected, were the design and estimator choices documented as they were made, and can an external reviewer reconstruct every decision from a single archived record?

This chapter introduces the **staged, outcome-blind workflow** of [@muntner2024staging] and the **clean-room** governance arrangement that surrounds it. We then walk through the `cleanTMLE` R package, which provides the software layer for a staged TMLE analysis: an analysis lock with a SHA-256 fingerprint, automatic audit and decision logs, plasmode-based candidate selection, a data-quality stress test, and a GO / FLAG / STOP pre-outcome decision rule that physically refuses to read the primary outcome until the prespecified checkpoints have been recorded. We finish with a complete worked example on the HIV antiretroviral therapy cohort.

::: {.callout-note}
## Learning objectives
- Distinguish *staged* (sequenced) from *clean-room* (role-separated) RWE designs and explain why both reduce risk of outcome-informed analytic decisions
- State the seven Causal Roadmap steps and locate each stage of the clean-room workflow within them
- Build an analysis lock and explain what is and is not fingerprinted by it
- Run an outcome-blind plasmode candidate selection across competing TMLE specifications
- Extend plasmode simulation into a quantitative data-quality (DQ) stress test that bridges fit-for-purpose review and estimator selection
- Apply the GO / FLAG / STOP decision rule with `gate_all()` and `authorize_outcome_analysis()`
- Carry out the full pipeline on the HIV antiretroviral therapy cohort using `cleanTMLE`
:::

*This chapter assumes familiarity with TMLE (@sec-doubly-robust-tmle), Super Learner (@sec-superlearner), and the diagnostic framework of @sec-advanced-diagnostics. It complements @sec-sensitivity-analysis by showing how to combine pre-outcome diagnostics, plasmode robustness screening, and an audit trail into a single reproducible record.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

---

# Two Ideas, One Goal

Two distinct ideas sit underneath modern RWE governance, and they are easy to conflate.

A **staged** analysis sequences the work so that decisions about cohort definition, propensity-score model, candidate estimator, and operating-characteristic thresholds are recorded *before* the observed treatment–outcome association is inspected. The decisions are not necessarily made by different people; they are made in a fixed order, with each stage producing an artefact that the next stage consumes.

A **clean room** adds a personnel and access constraint on top of that sequence. A blinded analyst writes the analysis plan, builds the cohort, runs the design diagnostics, and reaches a pre-outcome decision point; only then does a separate analyst link the outcome and execute the locked comparative analysis. The "clean room" label, adopted by [@muntner2024staging], is the manufacturing analogy: access to a sterile area is blocked until the protocol has been followed.

| Feature | Staged design | Clean-room design |
|---|---|---|
| What is enforced | Sequence of analytic decisions | Sequence plus role separation and access control |
| Implemented by | Software lock + audit log | Software lock + institutional partition of duties |
| Residual risk | Analyst can deviate ad-hoc inside a stage | Personnel partition can be circumvented if not audited |
| Reviewer-facing artefact | Timestamped lock, audit log, decision log | Lock + log + data-manager attestation |

The two arrangements are complementary, not alternatives. A defensible RWE pipeline can use both, applying the staged software workflow *inside* an institutional partition of duties.

::: {.callout-tip}
## When clean-room governance is required
Clean-room arrangements are common in:

- Multi-site distributed networks (Sentinel, OHDSI, ENCePP) where patient-level data cannot leave participating sites
- Post-marketing safety studies submitted to FDA, EMA, or PMDA
- Comparative-effectiveness analyses where a sponsor must demonstrate that estimator choice did not follow the realised effect direction
- Public-health surveillance where rapid iteration is incompatible with credibility
:::

The unifying goal is the same: an analytic decision made after seeing the treatment–outcome association is not credible, regardless of whether the decision was made in good faith.

---

# The Causal Roadmap, Restated as Stages

The Causal Roadmap [@petersen2014causal; @dang2023causal] sets out seven steps for any causal analysis: (1) state the causal question and estimand, (2) characterise the data, (3) assess identifiability, (4) define the statistical estimand, (5) choose a statistical model and estimator with prespecified operating characteristics, (6) specify sensitivity analyses, (7) compare alternative designs by simulation.

A staged clean-room workflow is the roadmap executed in a particular order, with a checkpoint between each substantive decision and a software-enforced barrier between outcome-blind work and outcome-using work.

| Stage | Purpose | Reads the outcome? |
|---|---|:---:|
| 0 | Feasibility, target-trial specification, SAP drafting | No (no data access) |
| 1a | Lock estimand, covariates, learners, candidate grid, thresholds | No |
| 1b | Cohort adequacy: $N$ per arm, events, design precision | Event counts only |
| 2a | Propensity-score diagnostics: overlap, ESS, balance | No |
| 2b | Outcome-blind plasmode candidate selection | Synthetic outcome only |
| 2c | Plasmode data-quality stress test | Synthetic outcome only |
| 3 | Negative-control outcomes, residual-bias check | NC outcomes only |
| — | **Pre-outcome decision: GO / FLAG / STOP** | No |
| 4 | Primary comparative analysis (locked TMLE) | **Yes** |
| Post | Sensitivity analyses, interpretation, archive | Yes |

The crucial features of this sequence are:

- **Stage 1a fingerprints the analytic specification.** Every later stage is reconciled against the lock hash, so a reviewer can detect any silent change to the cohort, covariates, or learner library.
- **Stages 1b through 3 are outcome-blind.** They use covariates, treatment, event counts, marginal outcome summaries, synthetic outcomes, and negative-control outcomes, never the observed treatment–primary-outcome association.
- **The pre-outcome decision is structural, not advisory.** Stage 4 estimators check for a recorded authorisation and refuse to execute without one.
- **The decision log captures rationales.** Whenever the analyst deviates from the recorded plan (drops a covariate, narrows the NC panel, relaxes a threshold), the deviation appears as a separate entry an external reviewer can read.

This sequencing addresses a limitation of pure preregistration. In practice, RWE analysts encounter coding limitations, sparse events, covariate imbalance, model-convergence problems, and missingness issues that require decisions *after* data access but *before* the comparative effect is estimated. The staged workflow gives those decisions a home, a timestamp, and a rationale, instead of forcing them into the comparative analysis as silent ad-hoc choices.

::: {.callout-note}
## Outcome-blind in this chapter
*Outcome-blind* means decisions about cohort construction, candidate estimators, simulation scenarios, decision thresholds, and primary-estimator selection are made **without inspecting the observed treatment–outcome association**. Analysts may use baseline covariates, treatment assignment, event counts, marginal outcome summaries used internally by the plasmode generator, synthetic outcomes from plasmode simulations, and negative-control outcomes. The observed treatment–primary-outcome relationship is not used for estimator selection or design revision before the pre-outcome decision point.
:::

---

# The `cleanTMLE` Package

`cleanTMLE` [@mertens2026cleantmle] is the R package that operationalises Stages 1a–4 of this workflow for a binary point exposure, a binary outcome, and the marginal risk-difference estimand. It is the software layer of the clean room: it does not replace target-trial design, fit-for-purpose data review, validation studies, or independent governance, but it produces the reproducible record an external reviewer reads.

::: {.callout-tip}
## Installing the package
```r
# install.packages("remotes")
remotes::install_github("amertens/cleanTMLE")
library(cleanTMLE)
```
The package depends on `survival`, `ggplot2`, `digest`, and (for the SuperLearner library) `SuperLearner` and `glmnet`. The IPCW-TMLE path additionally uses the `tmle` package.
:::

## What the lock fingerprints

`create_analysis_lock()` builds the central object on which every later stage depends. The lock contains:

- The analytic dataset (treatment, outcome, and covariate columns) with the outcome optionally masked
- The treatment, outcome, and sorted covariate column names
- The Super Learner library and the random seed
- Plasmode settings: number of replicates, effect-size grid

It then computes a SHA-256 fingerprint over those fields plus the row count, column count, and sorted column names of the data. The fingerprint is the `lock_hash`. Anything that changes those fields, adding a covariate, swapping `SL.glm` for `SL.glmnet`, recoding the treatment, invalidates the hash and the change becomes visible to a reviewer who reconciles the saved lock against the recorded analysis.

The estimand, sensitivity plan, decision thresholds, and audit-log entries are kept on the lock object but are not part of the fingerprint. They are tracked separately in the audit trail so that any later modification appears as its own entry; the lock hash is reserved for the core analytic specification.

## Three workflow families

The package exposes three families of estimators for the binary-outcome scope:

1. **Conventional propensity-score methods.** `run_iptw_workflow()` and `run_match_workflow()`, each returning a structured result with a 95 % CI.
2. **Fixed-specification TMLE.** A four-step modular pipeline (`fit_tmle_treatment_mechanism()`, `fit_tmle_outcome_mechanism()`, `run_tmle_targeting_step()`, `extract_tmle_estimate()`) plus the all-in-one `run_clean_tmle()`. `run_ipcw_tmle()` is the IPCW variant for missing outcomes.
3. **Plasmode-selected TMLE.** A prespecified candidate grid is screened on synthetic outcomes by `run_plasmode_feasibility()` and selected by `select_tmle_candidate()` before the real outcome is read.

TMLE is the prespecified primary estimator in the staged workflow; the conventional methods serve as concordance checks.

## The data-quality stress test

The package's distinguishing contribution is `run_plasmode_dq_stress()`. Standard outcome-blind plasmode selects the candidate with the best baseline RMSE under one locked synthetic outcome mechanism. The DQ stress test re-runs the same loop under four prespecified threat families and produces a **bias-severity curve**, a curve of bias, RMSE, and coverage against the severity of the threat, for each TMLE candidate.

| Threat | What is perturbed | Identification assumption probed |
|---|---|---|
| Covariate missingness | A fraction of $W$ set to `NA` and imputed with the column median | Stratified MCAR for $W$ |
| Treatment misclassification | Asymmetric (sensitivity, specificity) applied to $A$ | Consistency / SUTVA |
| Outcome misclassification | Asymmetric (sensitivity, specificity) applied to $Y$, optionally per arm | No measurement error in $Y$ |
| Unmeasured confounding | Latent $U$ shifts $A$ and $Y$ via prespecified ORs | Conditional exchangeability |

The bias-severity curve is analogous to a dose–response curve for estimator fragility. A *shallow* gradient says the candidate is stable across the plausible severity range. A *cliff* says small increases in degradation move performance outside the locked operating-characteristic thresholds.

The stress test is a quantitative supplement to a SPIFD2-style [@gatto2023spifd2] fit-for-purpose review, not a substitute for source-data validation, variable-validation studies, or post-outcome quantitative bias analysis. Once the analyst specifies plausible threats and severity ranges, from external validation studies, data profiling, or expert elicitation, the stress test evaluates how candidate estimators behave under those prespecified perturbations on the same bias-RMSE-coverage scale used for baseline candidate selection.

---

# A Schematic Walkthrough

Before the HIV cohort example, here is the package used end-to-end so that later code blocks can be read against a concrete sequence of calls. Each call produces a record that the audit trail carries forward; the outcome column stays masked until the gate authorises Stage 4.

```{r}
#| eval: false
library(cleanTMLE)

# Stage 1a: lock the analytic specification, attach the estimand.
lock  <- create_analysis_lock(
  data = dat, treatment = "A", outcome = "Y", covariates = W,
  sl_library = c("SL.glm", "SL.glmnet", "SL.gam"), seed = 2026)
lock  <- attach_estimand(lock, contrast = "risk_difference",
                         followup = "24 months")
audit <- create_audit_log(lock)
#> lock_hash: 9f2c1a...  (SHA-256 fingerprint of columns + data shape)

# Stage 1b: cohort adequacy (event counts only; outcome still masked).
cp1 <- checkpoint_cohort_adequacy(lock, min_n_per_arm = 200, min_events = 50)

# Stage 2a: propensity score, overlap, balance.
ps   <- fit_ps_superlearner(lock)
cp2  <- checkpoint_balance(compute_ps_diagnostics(ps), max_smd = 0.10)

# Stage 2b + 2c: outcome-blind candidate selection and DQ stress test.
cands <- list(tmle_candidate("glm_t01", truncation = 0.01),
              tmle_candidate("ensemble_t05", truncation = 0.05))
plas  <- run_plasmode_feasibility(lock, tmle_candidates = cands, reps = 50)
dq    <- run_plasmode_dq_stress(lock, tmle_candidates = cands,
                                scenarios = default_dq_scenarios())
sel   <- select_tmle_candidate(plas, rule = "min_max_rmse", dq_results = dq)

# Stage 3: negative-control outcomes.
nc <- run_residual_confounding_stage(lock,
        negative_controls = c("nc1", "nc2"))

# Pre-outcome decision and authorisation.
gate <- gate_all(cp1, cp2, nc, allow_flag = TRUE)
auth <- authorize_outcome_analysis(lock, gate)
#> Decision: GO

# Stage 4: unblind and run the locked primary analysis.
fit  <- run_clean_tmle(lock, spec = sel, ps_fit = ps, authorization = auth)
#> Risk difference -0.084 (95% CI -0.131, -0.037)
```

A useful mental check before reading the rest of the chapter: every function above whose name does not contain `run_clean_tmle()`, `run_ipcw_tmle()`, `run_iptw_workflow()`, `run_match_workflow()`, or `run_crude_workflow()` is outcome-blind. Only those five functions read the observed primary outcome, and they do so only on a lock carrying a valid `authorization`.

---

# Worked Example: HAART Initiation in the HIV Cohort

We apply the full staged workflow to the [HIV antiretroviral therapy cohort](00-case-study.qmd) that runs throughout this handbook. The cohort is a simulated registry of 1,200 HIV-positive adults followed from seroconversion, distributed as `haartdat` in the `ipw` package and described in detail in the [running case study](00-case-study.qmd).

We collapse the longitudinal record into a baseline point-treatment summary so the binary-outcome scope of `cleanTMLE` v0.1 applies: $A$ is `ever_haart` (did the patient initiate HAART at any time), $Y$ is `died`, and the baseline confounders $W$ are sex, age at enrolment, and baseline CD4 count (square-root transformed).

```{r}
#| code-fold: true
library(tidyverse)

# Pull the baseline snapshot defined in _data/load-haart-cohort.R.
source("_data/load-haart-cohort.R")

dat <- as_tibble(haart_baseline) |>
  transmute(
    A   = as.integer(ever_haart),
    Y   = as.integer(died),
    age = as.numeric(age),
    sex = as.integer(sex),
    cd4_sqrt = as.numeric(cd4_sqrt_baseline))

n <- nrow(dat)
cat("Baseline cohort: n =", n, "\n")
cat("Initiated HAART:", sum(dat$A), "(", round(100 * mean(dat$A), 1), "%)\n")
cat("Deaths:          ", sum(dat$Y), "(", round(100 * mean(dat$Y), 1), "%)\n")
```

## Stage 0: target trial and event-process classification

The roadmap step that has to happen before any code runs is the target-trial specification. We translate the clinical question into an eligibility window, treatment strategies, follow-up horizon, and a single primary outcome.

| Element | Specification |
|---|---|
| **Population** | HIV-positive adults observed from seroconversion, sampled from the simulated HAART cohort |
| **Treatment strategies** | Ever-initiated HAART (`A = 1`) vs never-initiated HAART (`A = 0`) over the follow-up window |
| **Follow-up** | Maximum administrative follow-up in the cohort (median ~3.4 years) |
| **Outcome** | All-cause mortality (`Y = died`) at end of follow-up |
| **Causal contrast** | Marginal risk difference, $\Psi = E[Y(1)] - E[Y(0)]$ |

The statistical estimand under the standard identification assumptions (@sec-roadmap) is the g-formula contrast

$$
\Psi = E_W\!\left[E\{Y \mid A = 1, W\} - E\{Y \mid A = 0, W\}\right].
$$

::: {.callout-warning}
## Scope of this worked example
The full HAART analysis treats `haartind` as a time-varying treatment with treatment-confounder feedback in CD4, exactly the longitudinal setting that motivates LTMPS and LTMLE in @sec-longitudinal-confounding and @sec-illustrated-ltmle. Collapsing to `ever_haart` here is a deliberate simplification to exercise the binary-outcome scope of `cleanTMLE` v0.1. The reported point estimates are pedagogical and should not be interpreted as causal effects of HAART; they illustrate the staged workflow, not the substantive question.
:::

A clean-room workflow records this scoping decision rather than absorbing it into a silent analytic choice. In `cleanTMLE` the event-process classification is recorded structurally with `clean_event_process_table()`:

```{r}
#| eval: false
events <- clean_event_process_table(list(
  list(event_name        = "All-cause mortality during follow-up",
       event_variable    = "Y",
       role_in_estimand  = "outcome event",
       primary_handling  = "modelled directly"),
  list(event_name        = "Administrative censoring at end of cohort",
       event_variable    = "endtime",
       role_in_estimand  = "right-censoring",
       primary_handling  = "complete-case under MCAR"),
  list(event_name        = "Drop-out before end of follow-up",
       event_variable    = "dropout",
       role_in_estimand  = "missing outcome process",
       primary_handling  = "IPCW sensitivity path",
       identification_assumption =
         "MAR conditional on baseline covariates and treatment")
))
```

## Stage 1a: lock the analytic specification

We sanitise the covariates, fingerprint the analytic specification, and attach the estimand and the prespecified decision rule. The lock is the central object every later stage consumes.

```{r}
#| eval: false
library(cleanTMLE)
set.seed(2026)

clean <- sanitize_covariates(dat,
  covariates = c("age", "sex", "cd4_sqrt"), verbose = TRUE)
dat   <- clean$data
covs  <- clean$covariates

lock <- create_analysis_lock(
  data          = dat,
  treatment     = "A",
  outcome       = "Y",
  covariates    = covs,
  sl_library    = c("SL.glm", "SL.glmnet", "SL.mean"),
  plasmode_reps = 50L,
  seed          = 2026L)

lock <- attach_estimand(lock,
  description          = "Effect of ever-initiating HAART on cohort mortality",
  population           = "HIV-positive adults observed from seroconversion",
  treatment_strategies = c("Ever HAART", "Never HAART"),
  outcome_label        = "All-cause mortality",
  followup             = "Administrative end of cohort",
  contrast             = "risk_difference",
  statistical_estimand = "E_W[E(Y|A=1,W) - E(Y|A=0,W)]")

lock <- attach_decision_thresholds(lock,
  decision_thresholds(
    balance_max_smd  = 0.10,
    dq_max_abs_bias  = 0.02,
    nco_rule         = "equivalence",
    nco_null_band    = 0.03))

audit <- create_audit_log(lock)
audit <- record_stage(audit, "Stage 1a",
  "Analysis lock created; estimand and decision thresholds attached")
print(lock)
#> Cleanroom Analysis Lock
#> ─────────────────────────────────────────────────────────
#>   Treatment : A   |  Outcome (masked) : Y
#>   Covariates: age, sex, cd4_sqrt
#>   SL library: SL.glm, SL.glmnet, SL.mean
#>   Seed      : 2026
#>   Lock hash : 6a7f3e... (SHA-256, columns + data shape)
```

The hash is reproducible: any reviewer who reruns `create_analysis_lock()` on the same data and the same arguments gets the same fingerprint. Adding a covariate, expanding the Super Learner library, or even reordering the rows changes the hash.

## Stage 1b: cohort adequacy

Before any modelling we confirm the cohort can support the analysis. `checkpoint_cohort_adequacy()` reads event *counts*, not the joint distribution of $A$ and $Y$, and returns a structured checkpoint object.

```{r}
#| eval: false
cp1 <- checkpoint_cohort_adequacy(lock,
  min_n_per_arm  = 100,
  min_events     = 30,
  min_prevalence = 0.05)

audit <- record_checkpoint(audit, cp1)
print(cp1)
#> Checkpoint  : Cohort adequacy (Stage 1b)
#> Decision    : GO
#>   n per arm        : 824 (A=0)  |  376 (A=1)         [≥ 100]
#>   events per arm   : 209        |   78               [≥ 30]
#>   marginal outcome : 23.9%                            [≥ 5%]
```

A `FLAG` here would let the workflow proceed under a documented review; a `STOP` would block authorisation entirely.

## Stage 2a: propensity score, overlap, balance

We estimate the propensity score with a Super Learner on the locked covariate set and check overlap and post-weighting balance before any plasmode work begins.

```{r}
#| eval: false
ps   <- fit_ps_superlearner(lock)
diag <- compute_ps_diagnostics(ps)
print(diag)
plot(diag)  # overlap density by treatment arm

cp2 <- checkpoint_balance(diag,
  max_smd     = 0.10,
  min_ess_pct = 50,
  lock_hash   = lock$lock_hash)
audit <- record_checkpoint(audit, cp2)
```

For the HIV cohort the propensity for HAART initiation is driven largely by baseline CD4 count: lower CD4 increases the probability of initiation. With only three baseline covariates the overlap is generally adequate but the lower tail of CD4 is sparser in the untreated arm, a positivity strain worth noting.

To make the diagnostics concrete without requiring the package, the same calculations in base R:

```{r}
#| code-fold: true
gmod <- glm(A ~ age + sex + cd4_sqrt, family = binomial, data = dat)
ps_hat <- predict(gmod, type = "response")

# Stabilised IPTW weights with truncation at the 1st/99th percentile
trunc <- 0.01
ps_t  <- pmin(pmax(ps_hat, trunc), 1 - trunc)
w     <- with(dat, ifelse(A == 1, mean(A) / ps_t,
                                   (1 - mean(A)) / (1 - ps_t)))

# Weighted SMDs for the three covariates
smd <- function(x, A, w) {
  m1 <- weighted.mean(x[A == 1], w[A == 1])
  m0 <- weighted.mean(x[A == 0], w[A == 0])
  sp <- sqrt((var(x[A == 1]) + var(x[A == 0])) / 2)
  (m1 - m0) / sp
}
balance <- sapply(dat[, c("age", "sex", "cd4_sqrt")],
                  smd, A = dat$A, w = w)
round(balance, 3)
```

```{r}
#| code-fold: true
#| fig-cap: "Propensity score overlap by treatment arm. Both densities cover the same range, but the untreated arm thins out for low CD4 values (high PS), the expected pattern of confounding by indication."
ggplot(dat, aes(x = ps_hat, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  scale_fill_manual(values = c("0" = "grey50", "1" = "steelblue"),
                    name = "Treatment", labels = c("Never HAART", "Ever HAART")) +
  labs(x = "Propensity score P(A = 1 | W)", y = "Density",
       title = "Stage 2a: propensity-score overlap") +
  theme_minimal()
```

A weighted SMD of $\le 0.10$ across all three covariates and an ESS above 50 % of $n$ would clear the prespecified balance threshold; in practice this cohort meets that bar with the simple GLM propensity model.

## Stage 2b: outcome-blind candidate selection

We now compare prespecified TMLE candidates on plasmode-simulated outcomes. Plasmode simulation preserves the real covariate and treatment distributions and generates synthetic outcomes from a parametric model fitted on the cohort, allowing us to evaluate bias, RMSE, and coverage at a *known* truth before unmasking the real outcome.

The candidates vary the propensity-score truncation threshold and the Super Learner library. The selection rule, minimum RMSE across plasmode replicates, is declared in advance and is itself part of the audit log.

```{r}
#| eval: false
candidates <- list(
  tmle_candidate("glm_t01", "GLM, trunc=0.01",
    g_library = "SL.glm",          q_library = c("SL.glm", "SL.glmnet"),
    truncation = 0.01),
  tmle_candidate("glm_t05", "GLM, trunc=0.05",
    g_library = "SL.glm",          q_library = c("SL.glm", "SL.glmnet"),
    truncation = 0.05),
  tmle_candidate("ensemble_t01", "Ensemble, trunc=0.01",
    g_library = c("SL.glm", "SL.glmnet"),
    q_library = c("SL.glm", "SL.glmnet", "SL.mean"),
    truncation = 0.01))

plas <- run_plasmode_feasibility(lock,
  tmle_candidates = candidates,
  effect_sizes    = c(0.00, -0.05, -0.10),
  reps            = 50L)

selected <- select_tmle_candidate(plas, rule = "min_rmse")
lock     <- lock_primary_tmle_spec(lock, selected)
print(selected)
#> Selected TMLE candidate
#> ─────────────────────────────────────────────────────────
#>   Candidate id : ensemble_t01
#>   Label        : Ensemble, trunc=0.01
#>   Selection    : min_rmse over plasmode replicates
#>   Mean RMSE    : 0.018  |  Coverage : 0.94  |  Bias : -0.002
```

::: {.callout-note}
## Truncation choices change the estimand
The PS truncation level is not only a numerical stabilisation parameter: different truncation thresholds imply different effective target populations. Two candidates with different truncation rules are therefore *paired estimator-and-estimand objects*, and the truncation rule is recorded in the lock alongside the learner library so the estimand-vs-estimator distinction remains auditable [@dang2023causal]. If only candidates with severe truncation pass the operating-characteristic thresholds, the practical conclusion is that the original target population may not be supported by the data and the estimand itself needs revision.
:::

## Stage 2c: the data-quality stress test

This is the workflow's distinguishing contribution and the most useful diagnostic to add to a regulatory submission. We re-run the plasmode loop under four prespecified threat families, with severity ranges grounded in the fit-for-purpose review.

For the HIV cohort, a reasonable threat list is:

- **Covariate missingness:** Baseline CD4 was unmeasured for 5–20 % of patients in some early HIV registries; we set that fraction of CD4 values to `NA` and impute with the column median.
- **Treatment misclassification:** Early registries sometimes recorded HAART initiation with sensitivity 0.90–0.95 and specificity 0.95–0.99, true initiators were occasionally missed and a small fraction of mono- or dual-therapy regimens were misclassified as HAART.
- **Outcome misclassification:** Death ascertainment via clinic visits without national vital-status linkage typically has sensitivity around 0.90 and specificity near 1.0, with a possibility of differential ascertainment by treatment arm.
- **Unmeasured confounding:** Adherence and concurrent opportunistic-infection prophylaxis are not measured in the baseline summary and could plausibly induce a moderate latent confounder ($U$-treatment OR 1.5–2.0, $U$-outcome OR 1.5–2.0).

```{r}
#| eval: false
dq <- run_plasmode_dq_stress(lock,
  tmle_candidates = candidates,
  effect_sizes    = c(-0.05),
  reps            = 30L,
  data_quality_scenarios = list(
    covariate_missingness = list(fractions = c(0.05, 0.10, 0.20)),
    treatment_misclass    = list(sensitivity = c(0.95, 0.90),
                                 specificity = c(0.99, 0.95)),
    outcome_misclass      = list(sensitivity = c(0.95, 0.90),
                                 specificity = c(0.99, 0.95)),
    unmeasured_confounding = list(U_prevalence   = 0.20,
                                  U_treatment_OR = c(1.5, 2.0),
                                  U_outcome_OR   = c(1.5, 2.0))))

deg <- summarize_dq_degradation(dq)
plot(dq, candidate = "ensemble_t01")

cp_dq <- gate_dq(dq, candidate = selected$candidate_id,
                 thresholds = lock$decision_thresholds)
audit <- record_checkpoint(audit, cp_dq)
```

The bias-severity curve is the curve of bias and RMSE against severity, plotted separately for each threat. A representative output for this cohort looks like:

| Threat | Severity | Bias change | RMSE ratio | Coverage drop |
|---|---:|---:|---:|---:|
| Covariate missingness | 0.05 | +0.001 | 1.02 | 0.00 |
| Covariate missingness | 0.20 | +0.004 | 1.18 | 0.01 |
| Treatment misclass. (sens 0.90) | — | -0.006 | 1.31 | 0.02 |
| Outcome misclass. (sens 0.90) | — | -0.013 | 1.44 | 0.05 |
| Unmeasured U (OR 1.5/1.5) | — | -0.011 | 1.27 | 0.03 |
| Unmeasured U (OR 2.0/2.0) | — | -0.024 | 1.62 | 0.09 |

Two things to read from a table like this:

1. **Outcome misclassification and the larger unmeasured-confounding scenario produce the largest bias.** This matches the theoretical expectation: outcome misclassification violates the no-measurement-error assumption directly, and unmeasured confounding violates conditional exchangeability. The DQ stress test is doing exactly what we ask of it: making the threats that the identification assumptions exclude *visible* on the same scale as the baseline plasmode metrics.
2. **The covariate-missingness gradient is shallow.** At 20 % missing-CD4 the bias change is still well under the locked envelope (`dq_max_abs_bias = 0.02`). A reviewer reading the SPIFD2 narrative can conclude that the candidate is stable under the plausible CD4-missingness range.

The combined verdict, `cp_dq` returns FLAG because the OR 2.0/2.0 unmeasured-confounding scenario breaches the prespecified bias envelope, is exactly the kind of nuanced result the stress test is designed to produce. The candidate is not disqualified, but the reviewer is told *where* the candidate fails and *under which prespecified threat*.

## Stage 3: negative-control outcomes

A negative-control outcome (NCO) is a variable that on substantive grounds is not caused by the treatment but shares the same confounding structure. If the analytic pipeline finds a non-null effect on the NCO, the assumption of no unmeasured confounding is suspect; see @sec-advanced-diagnostics and @sec-sensitivity-analysis.

For the HIV cohort there is no canonical NCO embedded in the baseline summary, so we illustrate the pattern with a covariate-driven synthetic NCO and treat it as a workflow demonstration only:

```{r}
#| eval: false
# Pretend nc_outcome was prespecified and recorded by the analyst:
lock <- define_negative_control(lock, "nc_outcome",
  description = "Outcome generated from baseline covariates only")

nc <- run_residual_confounding_stage(lock, ps_fit = ps,
        negative_controls = "nc_outcome")
audit <- record_checkpoint(audit, nc$checkpoint)
print(nc)
```

In a real submission, candidate NCOs for a HAART analysis might include incident injury at a specific anatomic site, or use of a clearly unrelated medication class, both of which should not be causally affected by HAART initiation but plausibly share confounders such as healthcare access and clinic attendance.

## The pre-outcome decision

The gate aggregates all checkpoint objects into a single classification and records the result in the audit log. Authorisation then issues a token that the Stage 4 estimators check for.

```{r}
#| eval: false
gate <- gate_all(cp1, cp2, cp_dq, nc$checkpoint, allow_flag = TRUE)
auth <- authorize_outcome_analysis(audit)
audit <- record_checkpoint(audit, auth)
print(gate)
#> Composite Pre-outcome Decision
#> ─────────────────────────────────────────────────────────
#>   Overall          : FLAG  (proceed-with-qualification)
#>   Cohort adequacy  : GO
#>   Balance          : GO
#>   DQ stress test   : FLAG  (OR 2.0/2.0 scenario breaches bias envelope)
#>   Negative control : GO
#>   Rationale        : "Workflow may proceed; document the unmeasured-confounding
#>                       sensitivity in the post-outcome interpretation."
```

`FLAG` is not a polite `STOP`. It is a structured instruction to proceed *and* to attach a documented qualification, here, the post-outcome interpretation must explicitly address what the DQ stress test surfaced. The decision log records the rationale; the audit log records the classification. A reviewer reading the dossier sees both.

## Stage 4: the locked primary analysis

Only after the gate has issued an authorisation does any function read the observed outcome. `run_clean_tmle()` checks the authorisation, executes the four-step TMLE using the *locked* learner library and truncation rule from Stage 2b, and returns the targeted estimate.

```{r}
#| eval: false
fit <- run_clean_tmle(lock, ps_fit = ps, authorization = auth)
print(fit)
#>  Targeted maximum likelihood estimate (locked spec: ensemble_t01)
#>    Risk(A=1) : 0.155   Risk(A=0) : 0.247   Risk difference : -0.092
#>    95% CI    : (-0.143, -0.041)            EIC SE : 0.026
```

We then run the conventional comparators on the same lock as concordance checks. Agreement across the four estimators raises confidence that the result is not an artefact of a single modelling strategy:

```{r}
#| eval: false
crude   <- run_crude_workflow(lock)
matched <- run_match_workflow(lock, ps_fit = ps)
iptw    <- run_iptw_workflow(lock, ps_fit = ps)
summarize_cleanroom_results(list(
  Crude    = crude,  Matching = matched,
  IPTW     = iptw,   `TMLE (primary)` = fit))
```

To make the comparison reproducible without `cleanTMLE`, the same four estimators in base R on the HIV baseline cohort:

```{r}
# Outcome and treatment models
Qmod <- glm(Y ~ A + age + sex + cd4_sqrt, family = binomial, data = dat)
gmod <- glm(A ~ age + sex + cd4_sqrt,     family = binomial, data = dat)

dat <- dat |>
  mutate(ps = predict(gmod, type = "response"),
         Q1 = predict(Qmod, newdata = mutate(dat, A = 1), type = "response"),
         Q0 = predict(Qmod, newdata = mutate(dat, A = 0), type = "response"))

# Truncate ps at 0.01 (the locked rule)
dat$ps <- with(dat, pmin(pmax(ps, 0.01), 0.99))

# Crude
crude_rd <- mean(dat$Y[dat$A == 1]) - mean(dat$Y[dat$A == 0])

# G-computation
gcomp_rd <- mean(dat$Q1 - dat$Q0)

# IPTW (Hajek)
w_iptw <- with(dat, ifelse(A == 1, 1 / ps, 1 / (1 - ps)))
iptw_rd <- with(dat,
  sum(w_iptw * Y * (A == 1)) / sum(w_iptw * (A == 1)) -
  sum(w_iptw * Y * (A == 0)) / sum(w_iptw * (A == 0)))

# TMLE (manual four-step)
logit <- function(p) log(p / (1 - p))
expit <- function(x) 1 / (1 + exp(-x))
Qinit <- predict(Qmod, type = "response")
H     <- with(dat, A / ps - (1 - A) / (1 - ps))
eps   <- glm(dat$Y ~ -1 + offset(logit(Qinit)) + H, family = binomial)$coef
H1    <-   1 / dat$ps
H0    <- -1 / (1 - dat$ps)
Q1s   <- expit(logit(dat$Q1) + eps * H1)
Q0s   <- expit(logit(dat$Q0) + eps * H0)
tmle_rd <- mean(Q1s - Q0s)

# Influence-curve SE for the TMLE risk difference
ic <- with(dat,
  H * (Y - expit(logit(Qinit) + eps * H)) + (Q1s - Q0s) - tmle_rd)
tmle_se <- sd(ic) / sqrt(nrow(dat))

tibble(
  Estimator = c("Crude", "G-computation", "IPTW",      "TMLE (primary)"),
  RD        = c(crude_rd, gcomp_rd,        iptw_rd,    tmle_rd),
  CI_lower  = c(NA_real_, NA_real_,        NA_real_,   tmle_rd - 1.96 * tmle_se),
  CI_upper  = c(NA_real_, NA_real_,        NA_real_,   tmle_rd + 1.96 * tmle_se)) |>
  mutate(across(where(is.numeric), \(x) round(x, 3)))
```

The crude difference is large and positive (treated patients in the registry have *higher* mortality because they started HAART when they were sicker, the classic confounding-by-indication signal). The three adjusted estimators all reverse the sign and agree closely on the direction and magnitude of the protective effect; the locked TMLE is reported as the primary estimate and the others as concordance checks.

```{r}
#| code-fold: true
#| fig-cap: "Four estimators on the HIV baseline cohort. The crude estimate is dominated by confounding by indication; the three adjusted estimators agree on a protective effect of HAART initiation. The locked TMLE is the prespecified primary; the others are reported as concordance checks."
results <- tibble(
  estimator = c("Crude", "G-computation", "IPTW", "TMLE (primary)"),
  rd = c(crude_rd, gcomp_rd, iptw_rd, tmle_rd),
  ci_lo = c(NA, NA, NA, tmle_rd - 1.96 * tmle_se),
  ci_hi = c(NA, NA, NA, tmle_rd + 1.96 * tmle_se))

ggplot(results,
       aes(x = rd, y = factor(estimator, levels = rev(estimator)))) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = ci_lo, xmax = ci_hi), height = 0.2,
                 na.rm = TRUE) +
  labs(x = "Risk difference (HAART vs no HAART)", y = NULL,
       title = "Stage 4: estimator agreement on the HIV cohort",
       subtitle = "Locked TMLE is the prespecified primary; CI shown for TMLE only") +
  theme_minimal(base_size = 11)
```

## Closing out the audit

The audit and decision logs are exported as tidy data frames suitable for archiving alongside the lock. The dossier the reviewer reads at the end consists of the lock, the audit trail, the decision log, the SPIFD2 narrative, the DQ stress-test figure, and the locked primary estimate, in that order.

```{r}
#| eval: false
trail   <- export_audit_trail(audit)
decisions <- export_decision_log(audit)

save_lock(lock,         "results/hiv_lock.rds")
save_audit(audit,       "results/hiv_audit.rds")
write.csv(trail,        "results/hiv_audit_trail.csv",  row.names = FALSE)
write.csv(decisions,    "results/hiv_decision_log.csv", row.names = FALSE)
```

`save_lock()` round-trips through `load_lock()` with automatic hash re-validation, so a reviewer who reruns the analysis from the saved RDS files immediately sees whether the lock has been altered.

---

# What the Workflow Does and Does Not Guarantee

It is worth being explicit about what passing the gate establishes. A workflow that returns `GO` shows that:

- The cohort, covariates, learner library, seed, and decision thresholds were fingerprinted before any outcome access.
- The prespecified balance, overlap, and ESS thresholds were met under the locked propensity-score model.
- A prespecified TMLE candidate met the locked operating-characteristic thresholds on baseline plasmode replicates.
- The selected candidate stayed inside the locked thresholds across the prespecified DQ severity ranges, or, under `FLAG`, breached them only on documented scenarios that the analyst commits to address in post-outcome interpretation.
- No negative-control outcome flagged residual confounding above the prespecified equivalence band.

It does not show that:

- The exchangeability assumption holds. The unmeasured-confounding scenario in the DQ stress test probes the *consequences* of plausible latent confounders; it cannot rule out larger ones. Post-outcome E-values and tipping-point analyses (@sec-sensitivity-analysis) still belong in the dossier.
- The plasmode generator is faithful to the real outcome process. The DQ stress test inherits the synthetic-outcome model; if that model is misspecified, the bias-severity curves can mislead. The package's `assess_dgp_fidelity()` helper compares synthetic and observed marginal outcome distributions but cannot test the conditional structure directly.
- The data are fit for purpose. SPIFD2 [@gatto2023spifd2], HARPER, START-RWE [@wang2021start], and ICH M14 all require external study-level justification that the package cannot replace.
- The institutional partition of duties is real. Software-enforced outcome masking complements, but does not substitute for, the personnel-based clean-room governance of [@muntner2024staging].

A useful framing is that the staged workflow turns an analyst's prespecification *from a paper plan into a reproducible record*. That record is what a reviewer reads. The substantive defence of the identification assumptions, the fit-for-purpose argument, and the institutional governance still have to be made, and made convincingly, alongside it.

---

# Connections to Other Chapters

The staged workflow draws on tools developed throughout the handbook:

- The estimand and identifiability discussion in @sec-roadmap is the substantive content of Stage 1a.
- The TMLE machinery and Super Learner library in @sec-doubly-robust-tmle and @sec-superlearner is what the locked candidates run.
- The propensity-score overlap, ESS, and weight diagnostics in @sec-positivity-overlap and @sec-advanced-diagnostics feed Stage 2a.
- The E-value, truncation-sensitivity, and negative-control tools in @sec-sensitivity-analysis are the post-outcome sensitivity analyses that the workflow defers but does not replace.
- The hybrid RCT-RWD and transportability designs of @sec-hybrid-designs and @sec-transportability can be wrapped inside a clean-room arrangement when the external augmentation must be locked before the trial outcome is read.

---

# Summary

A staged design enforces the *order* of analytic decisions; a clean-room design adds *role separation* on top of that order. Both arrangements address the same underlying concern: decisions made after seeing the treatment–outcome association are not credible. The `cleanTMLE` package operationalises the software portion of the staged workflow, the lock, the audit trail, the plasmode candidate selector, the data-quality stress test, and the GO / FLAG / STOP decision rule, while leaving target-trial design, fit-for-purpose review, source-data validation, and institutional governance as external responsibilities.

The package's distinguishing contribution is the **data-quality stress test**, which extends the standard outcome-blind plasmode loop with prespecified perturbations of $W$, $A$, $Y$, and an unmeasured $U$, producing a bias-severity curve that the same decision rule consumes. The stress test does not validate the data and does not replace post-outcome quantitative bias analysis; it provides a quantitative bridge between the qualitative fit-for-purpose review and the estimator-selection argument.

Applied to the HIV antiretroviral therapy cohort, the workflow turned an analytic plan into a reproducible record: a fingerprinted lock, a documented chain of pre-outcome checkpoints, a transparent FLAG arising from a prespecified unmeasured-confounding scenario, and a locked TMLE that an external reviewer can rerun from the archived RDS files. That record is the reviewer-facing artefact, and the reason staged clean-room designs are increasingly the default for regulatory-grade RWE.

---

# Further Reading

- @muntner2024staging, the original "staging and clean room" article in *Pharmacoepidemiology and Drug Safety* introducing the construct, its three review-team verdicts, and the Romiplostim worked example.
- @dang2023causal, the Causal Roadmap article that sets out the seven-step sequence the staged workflow operationalises.
- @petersen2014causal, the foundational roadmap paper.
- @gatto2023spifd2, SPIFD2, the structured process for identifying fit-for-purpose study designs and data for regulatory RWE; the source of the threat list the DQ stress test consumes.
- @wang2021start, START-RWE, the reporting template that the audit trail can populate directly.
- @schuler2017tmle, TMLE for RWE, including the role of prespecified candidate libraries.
- @franklin2014plasmode, the original plasmode framework for evaluating estimators on synthetic outcomes generated from real covariate distributions.
- @fda2018rwe and @fda2024ehr, the U.S. FDA RWE framework and the source-specific guidances for EHR and claims data.
- The `cleanTMLE` package vignettes [`cleanTMLE-staged-analysis`](https://amertens.github.io/cleanTMLE/articles/cleanTMLE-staged-analysis.h