---
title: "Chapter 3.5: Diagnostics and Sensitivity Analyses in Targeted Learning"
format: html
---

# Chapter 3.5
## Diagnostics and Sensitivity Analyses in Targeted Learning
*A systematic framework for credible causal inference with modern estimators*

::: {.callout-note}
## Learning objectives
- Map diagnostic procedures to specific steps of the Causal Roadmap
- Distinguish among assumption diagnostics, model diagnostics, estimator diagnostics, and design diagnostics
- Evaluate exchangeability using DAG review, negative controls, and quantitative bias analysis
- Diagnose positivity violations through propensity score analysis, clever covariate inspection, and effective sample size calculations
- Assess consistency and intervention well-definedness for static, dynamic, and stochastic regimes
- Evaluate nuisance model performance for both the outcome regression and the treatment mechanism using Super Learner diagnostics, calibration, and covariate balance
- Interpret efficient influence function diagnostics for TMLE and AIPW
- Apply simulation-based design diagnostics and sensitivity analysis frameworks
- Implement estimand adaptation strategies when identifying assumptions fail
- Report diagnostics transparently following recommended standards for regulatory and scientific audiences
:::

::: {.callout-important}
## Sources and scope
This chapter is educational. It provides a comprehensive framework for diagnostics and sensitivity analyses in the context of targeted learning and modern semiparametric causal inference. Causal conclusions depend on identification assumptions---consistency, exchangeability, and positivity---and on diagnostic evidence that the data and models support the target estimand. The framework presented here applies across TMLE, AIPW, IPTW, longitudinal TMLE, and modified treatment policy estimators. When flexible machine learning is used for nuisance estimation, valid inference typically requires cross-fitting or a cross-validated TMLE variant, plus appropriate rate conditions on nuisance function estimation.
:::

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

---

## 1. Diagnostics Within the Causal Roadmap Framework

The Causal Roadmap provides a structured sequence of decisions that connect a scientific question to a credible causal answer. At each step of this roadmap, specific diagnostic procedures serve as quality checks that determine whether the analysis can proceed on solid ground or whether modifications are required. This section establishes how diagnostics map onto the roadmap and introduces a taxonomy of diagnostic types that will organize the remainder of the chapter.

### 1.1 The Roadmap Steps and Their Diagnostic Counterparts

The Causal Roadmap, as formalized by Petersen and van der Laan, proceeds through a sequence of steps that translate a causal question into a statistical answer. Each step carries its own assumptions, and each assumption admits its own class of diagnostics.

**Steps 1--2: Causal model specification and assumption articulation.** The analyst begins by specifying the structural causal model (SCM) that encodes the data-generating process and the causal question of interest. This involves defining the endogenous variables $(W, A, Y)$, the background exogenous variables $U$, and the structural equations $F$ that relate them. The causal model also encodes the set of assumptions the analyst is willing to defend---for example, that certain variables are not caused by others, or that specific exclusion restrictions hold. Diagnostics at this stage are primarily conceptual: DAG review with domain experts, assessment of whether the causal model is compatible with the observed data structure, and verification that the temporal ordering encoded in the model matches the data collection process. These diagnostics cannot be automated; they require substantive scientific judgment.

**Step 3: Identifiability.** Given the causal model, the analyst must establish that the causal parameter of interest---the target causal estimand---is identified from the observed data distribution under the articulated assumptions. The three core identification assumptions are exchangeability (no unmeasured confounding), positivity (adequate support for all treatment levels across the covariate distribution), and consistency (well-defined interventions). Diagnostics for identifiability form the subject of Section 2 of this chapter. These diagnostics are necessarily indirect: the assumptions themselves are not testable from observed data alone, but their empirical implications can be evaluated, and their plausibility can be assessed through sensitivity analyses and falsification tests.

**Steps 4--5: Statistical estimand and estimator.** Once the causal parameter is identified, the analyst defines the corresponding statistical estimand---a functional of the observed data distribution that equals the causal parameter under the identification assumptions. The analyst then selects an estimator: TMLE, AIPW, IPTW, or another procedure. Diagnostics at this stage concern the quality of nuisance function estimation (Section 3), the behavior of the efficient influence function (Section 4), and the finite-sample performance of the chosen estimator (Section 5). These diagnostics assess whether the statistical machinery is performing adequately, conditional on the identification assumptions holding.

**Step 6: Inference.** Statistical inference---confidence intervals and hypothesis tests---relies on asymptotic approximations that may or may not hold in finite samples. Diagnostics for inference include assessment of whether the influence function is approximately normally distributed, whether variance estimates are stable, and whether small-sample corrections are needed. These diagnostics are particularly important when the effective sample size is reduced by extreme weights or near-positivity violations.

**Step 7: Interpretation.** The final step returns from the statistical answer to the causal question. Diagnostics at this stage concern the sensitivity of conclusions to violations of the identification assumptions (Section 9) and the appropriate communication of both statistical uncertainty and causal uncertainty (Section 10). A treatment effect estimate of $\hat{\psi} = 0.15$ with a 95% confidence interval of $(0.05, 0.25)$ conveys statistical precision, but it says nothing about the robustness of the causal interpretation to unmeasured confounding, positivity violations, or model misspecification. The diagnostics described throughout this chapter provide the evidence needed to evaluate that robustness.

### 1.2 A Taxonomy of Diagnostic Types

Across these roadmap steps, four distinct classes of diagnostics emerge. Recognizing which class a given diagnostic belongs to clarifies what it can and cannot tell the analyst.

**Assumption diagnostics** evaluate the plausibility of the identification assumptions that link the causal parameter to the statistical estimand. These include checks for exchangeability (negative controls, sensitivity analyses, E-values), positivity (propensity score distributions, effective sample sizes), and consistency (intervention well-definedness assessments). Assumption diagnostics operate at the boundary between causal and statistical reasoning. A failure detected by an assumption diagnostic does not mean the estimator is broken---it means the causal interpretation of the estimate may be invalid, regardless of how well the estimator performs statistically.

**Model diagnostics** evaluate the quality of the fitted nuisance functions: the outcome regression $\bar{Q}(A, W) = E[Y \mid A, W]$ and the treatment mechanism $g(A \mid W) = P(A \mid W)$. These include cross-validated risk, calibration, covariate balance, and Super Learner weight distributions. Model diagnostics operate entirely within the statistical domain. They assess whether the nuisance functions are estimated well enough to support the asymptotic properties (consistency, efficiency, valid inference) of the downstream causal estimator.

**Estimator diagnostics** evaluate the behavior of the causal estimator itself, given the fitted nuisance functions. These include influence function summaries, convergence checks for the TMLE targeting step, comparison of plug-in versus targeted estimates, and finite-sample performance assessments. Estimator diagnostics reveal whether the estimator is functioning as its asymptotic theory predicts, or whether finite-sample pathologies---instability, bias, poor coverage---are present.

**Design diagnostics** are conducted before the outcome data are accessed, using only the treatment and covariate information. These include simulation-based assessments of estimator performance under plausible data-generating processes, support diagnostics for the treatment mechanism, and sample size calculations that account for the specific estimator to be used. Design diagnostics are particularly relevant for regulatory applications and pre-registered analyses, where the credibility of the analysis depends on demonstrating that the study design can support the intended estimand before the results are known.

::: {.callout-important}
## The Critical Distinction: Causal Parameter, Statistical Estimand, Estimator, and Estimate

Throughout this chapter, it is essential to maintain the distinction among four objects that are often conflated:

The **causal parameter** $\Psi^F$ is a functional of the structural causal model---for example, the average treatment effect $E[Y(1) - Y(0)]$ defined through potential outcomes under the SCM. It exists in the causal world and requires no data to define.

The **statistical estimand** $\Psi(P_0)$ is a functional of the observed data distribution $P_0$ that equals the causal parameter under the identification assumptions. For example, $E_W[E[Y \mid A=1, W] - E[Y \mid A=0, W]]$ is a statistical estimand that equals the ATE under exchangeability, positivity, and consistency.

The **estimator** $\hat{\Psi}$ is a mapping from the observed data to a real number---a specific algorithm such as TMLE, AIPW, or IPTW that targets the statistical estimand.

The **estimate** $\hat{\psi}_n$ is the specific numerical value produced by applying the estimator to the observed sample of size $n$.

Diagnostics apply at each level. Assumption diagnostics concern the link between the causal parameter and the statistical estimand. Model and estimator diagnostics concern the link between the statistical estimand and the estimator. Inference diagnostics concern the reliability of the estimate as a summary of the estimator's sampling distribution.
:::

### 1.3 When Diagnostics Should Alter the Analysis

Diagnostics are not merely a reporting exercise to be appended after the analysis is complete. They are decision points that can and should alter the course of the analysis. The Causal Roadmap is explicitly iterative: when a diagnostic reveals a problem, the analyst returns to an earlier step and modifies the approach.

If assumption diagnostics reveal severe positivity violations, the analyst may need to redefine the estimand---for example, by restricting the target population, shifting from a static intervention to a modified treatment policy, or targeting a conditional rather than marginal effect. If model diagnostics reveal poor nuisance function estimation, the analyst may need to expand the Super Learner library, adjust the cross-validation scheme, or reconsider the functional form of the models. If estimator diagnostics reveal instability, the analyst may need to apply truncation, switch estimators, or acknowledge that the data cannot credibly support the intended inference.

Section 8 of this chapter provides a structured decision framework for these adaptations. The key principle is that the diagnostics guide principled adaptation of the analysis, rather than ad hoc modifications that compromise the pre-specified nature of the investigation.

---

## 2. Diagnosing Identifiability Assumptions

The identification assumptions---exchangeability, positivity, and consistency---are the bridge between the causal parameter defined under a structural causal model and the statistical estimand defined as a functional of the observed data distribution. If any of these assumptions fails, the statistical estimand no longer equals the causal parameter, and the estimate produced by any estimator, no matter how sophisticated, does not have a causal interpretation. Diagnostics for identifiability assumptions are therefore the most consequential diagnostics in any causal analysis: they determine whether the analysis answers a causal question or merely a predictive one.

A fundamental epistemological constraint governs this section. None of the three identification assumptions is testable from observed data alone. Exchangeability asserts something about variables that are, by definition, unmeasured. Positivity can be checked empirically in finite samples but may fail in the population. Consistency concerns the relationship between a hypothetical intervention and the observed treatment, which cannot be verified without experimental manipulation. The diagnostics described here are therefore indirect: they assess the plausibility of the assumptions through their observable implications, provide quantitative summaries of how sensitive conclusions are to assumption violations, and identify empirical patterns that would be unlikely if the assumptions held.

### 2.1 Exchangeability (No Unmeasured Confounding)

Exchangeability, stated formally as $Y(a) \perp\!\!\!\perp A \mid W$ for all treatment levels $a$ in the support of $A$, asserts that the potential outcomes are independent of treatment assignment conditional on the measured covariates $W$. This is the assumption that permits interpreting associations in the adjusted observed data as causal effects. It is also the assumption most frequently questioned by reviewers, regulators, and scientific audiences, because it can never be verified from the data at hand.

#### 2.1.1 Conceptual DAG Diagnostics

The directed acyclic graph (DAG) is the primary tool for reasoning about exchangeability. A DAG encodes the qualitative causal structure assumed by the analyst: which variables cause which others, which variables are unmeasured, and which causal pathways connect treatment to outcome. The DAG makes the exchangeability assumption explicit by revealing the adjustment set---the set of variables that, when conditioned upon, blocks all backdoor paths from $A$ to $Y$.

DAG diagnostics are inherently qualitative and require domain expertise. The analyst should construct the DAG in collaboration with subject-matter experts and then systematically interrogate it by asking the following questions. First, are there common causes of $A$ and $Y$ that are not included in $W$? Each such variable represents a potential source of unmeasured confounding. Second, does the DAG include variables that are consequences of both $A$ and $Y$ (colliders on the path between them), and if so, are any of these variables included in $W$? Conditioning on a collider opens a non-causal path and introduces collider bias. Third, does $W$ include any mediators---variables on the causal pathway from $A$ to $Y$? Conditioning on a mediator blocks part of the causal effect and changes the estimand from the total effect to a direct effect.

The DAG should be drawn before any statistical analysis is conducted, and it should be revisited whenever the analysis plan changes. Changes to the covariate adjustment set $W$---for example, adding or removing a variable---should be justified by reference to the DAG, not by statistical criteria such as significance in a regression model.

::: {.callout-tip}
## Practical DAG Review Protocol

A structured DAG review with clinical collaborators should proceed as follows. Present the DAG with all proposed nodes and edges. For each edge, ask whether the collaborator agrees that the causal relationship exists and whether its direction is correct. For each variable not included in the DAG, ask whether it could plausibly cause both $A$ and $Y$. For each variable included in $W$, ask whether it could be a consequence of $A$ (which would make it a mediator or collider). Document disagreements and note which unmeasured variables are most concerning. This record becomes the foundation for the sensitivity analyses described below.
:::

#### 2.1.2 Negative Control Exposures and Outcomes

Negative controls are among the most powerful empirical tools for detecting violations of exchangeability. The logic is based on falsification: if the identification assumptions hold, then certain associations should be absent in the data. Finding those associations present provides evidence that the assumptions are violated.

A **negative control outcome** (NCO) is an outcome variable $Y^{nc}$ that is known, on substantive grounds, not to be causally affected by $A$, but that shares the same confounding structure as the primary outcome $Y$. If the analyst estimates the effect of $A$ on $Y^{nc}$ using the same adjustment set $W$ and the same estimation procedure and finds a non-null association, this constitutes evidence that unmeasured confounding is present. The unmeasured confounder that biases the primary analysis also biases the negative control analysis, but because $A$ has no true effect on $Y^{nc}$, any detected association must be attributable to bias.

A **negative control exposure** (NCE) is a treatment variable $A^{nc}$ that is known not to causally affect $Y$, but that shares the same confounding structure as the primary treatment $A$. Estimating the effect of $A^{nc}$ on $Y$ using the same methods provides an analogous falsification test.

```r
## Negative control outcome analysis
## Y_nc: an outcome known to be unaffected by treatment A
## If the estimated effect is non-null, residual confounding is likely present
if (requireNamespace("tmle", quietly = TRUE)) {
  nco_fit <- tmle::tmle(
    Y = dat$Y_negative_control,
    A = dat$A,
    W = dat[, covariates],
    Q.SL.library = c("SL.glm", "SL.glmnet", "SL.ranger"),
    g.SL.library = c("SL.glm", "SL.glmnet", "SL.ranger")
  )
  cat("NCO effect estimate:", round(nco_fit$estimates$ATE$psi, 4), "\n")
  cat("95% CI:", round(nco_fit$estimates$ATE$CI, 4), "\n")
  cat("If this interval excludes zero, unmeasured confounding is likely.\n")
}
```

The selection of negative controls requires careful scientific reasoning. The NCO must be plausibly unaffected by $A$ but must share the confounding structure of the primary outcome. In pharmacoepidemiology, a common NCO is an outcome that occurs before treatment initiation in a new-user design, or an outcome in an organ system unrelated to the drug's mechanism of action but related to the same health-seeking behaviors that drive treatment selection. The credibility of the negative control analysis depends entirely on the validity of the assumption that $A$ does not affect $Y^{nc}$; if this assumption is wrong, a null finding on the NCO does not validate the primary analysis, and a non-null finding does not necessarily indicate confounding.

Negative controls can also be embedded within a formal bias-correction framework. Tchetgen Tchetgen and colleagues have developed methods that use negative control variables not merely as falsification tests but as instruments for estimating and correcting the bias due to unmeasured confounding, under assumptions about the structure of the bias. These methods require additional assumptions---specifically, that the negative control variable satisfies certain conditional independence relations with respect to the unmeasured confounder---but when applicable, they provide a more powerful diagnostic than a simple falsification test.

#### 2.1.3 Sensitivity Analysis for Unmeasured Confounding

Because exchangeability cannot be verified from observed data, sensitivity analysis provides a formal framework for assessing how robust causal conclusions are to potential violations. The fundamental question is: how strong would unmeasured confounding need to be to explain away the observed effect, or to change the conclusion in a meaningful way?

**E-values.** The E-value, introduced by VanderWeele and Ding, provides a simple summary of the minimum strength of association that an unmeasured confounder $U$ would need to have with both the treatment $A$ and the outcome $Y$, conditional on the measured covariates $W$, to fully explain away the observed treatment-outcome association. For an observed risk ratio $\text{RR}$, the E-value is defined as:

$$
E\text{-value} = \text{RR} + \sqrt{\text{RR} \times (\text{RR} - 1)}
$$

An E-value of 3.5 means that an unmeasured confounder would need to be associated with at least a 3.5-fold increase in both the probability of treatment and the probability of the outcome, above and beyond the measured covariates, to reduce the observed effect to the null. The E-value can also be computed for the confidence interval limit closest to the null, which provides a measure of how robust the statistical significance (not just the point estimate) is to unmeasured confounding.

```r
## E-value computation for a risk ratio estimate
## No external package required
compute_evalue <- function(RR) {
  if (RR >= 1) {
    e_val <- RR + sqrt(RR * (RR - 1))
  } else {
    ## For RR < 1, compute on the inverse scale
    RR_inv <- 1 / RR
    e_val <- RR_inv + sqrt(RR_inv * (RR_inv - 1))
  }
  return(e_val)
}

RR_point <- 1.8
RR_lower <- 1.2  ## lower bound of 95% CI for RR

cat("E-value for point estimate (RR =", RR_point, "):",
    round(compute_evalue(RR_point), 2), "\n")
cat("E-value for CI lower bound (RR =", RR_lower, "):",
    round(compute_evalue(RR_lower), 2), "\n")
cat("\nInterpretation: an unmeasured confounder would need associations\n")
cat("of at least these magnitudes with both A and Y to explain away\n")
cat("the observed effect (or its statistical significance).\n")
```

E-values have several important limitations that the analyst should communicate transparently. They assume a specific form for the confounding bias (the bounding factor of Ding and VanderWeele), which may not hold in all settings. They provide a worst-case bound, not a point estimate of the bias. They do not account for multiple unmeasured confounders acting jointly. And they require the effect to be expressed on the risk ratio scale, which may involve transformation from other scales.

**Quantitative bias analysis.** A more flexible framework for assessing sensitivity to unmeasured confounding is quantitative bias analysis (QBA), which directly models the potential bias as a function of parameters that describe the unmeasured confounder. The analyst specifies a set of bias parameters---typically the prevalence of the unmeasured confounder $U$ in the exposed and unexposed groups, and the association between $U$ and the outcome---and then computes the bias-adjusted estimate under each combination of parameter values.

Probabilistic QBA extends this framework by placing prior distributions over the bias parameters, reflecting the analyst's uncertainty about the unmeasured confounding structure. The result is a distribution of bias-adjusted estimates that integrates both statistical uncertainty (from sampling variability) and systematic uncertainty (from unmeasured confounding). The R packages **episensr** and **causalsens** provide tools for implementing QBA in practice.

```r
## Probabilistic quantitative bias analysis using episensr
if (requireNamespace("episensr", quietly = TRUE)) {
  library(episensr)
  ## Example: 2x2 table from an observational study
  ## Exposed cases, exposed non-cases, unexposed cases, unexposed non-cases
  result <- episensr::probsens(
    matrix(c(120, 80, 60, 140), nrow = 2, byrow = TRUE),
    type = "exposure",
    rrdist = list("trapezoidal", c(1.5, 2.0, 2.5, 3.0)),
    orprior = list("trapezoidal", c(0.3, 0.4, 0.5, 0.6)),
    corr.p = 0.8,
    discard = FALSE,
    seed = 12345
  )
  summary(result)
}
```

#### 2.1.4 Structural Versus Random Confounding

An important conceptual distinction that bears on the choice of diagnostic strategy is the difference between structural confounding and random confounding. Structural confounding arises when the data-generating process includes a common cause of $A$ and $Y$ that is not measured and cannot be addressed by the study design. This is the classical unmeasured confounding problem, and it produces systematic bias that does not diminish with increasing sample size. Sensitivity analyses and negative controls are the appropriate diagnostic tools for structural confounding.

Random confounding, by contrast, arises in finite samples when the random assignment of treatment happens to produce imbalances in prognostic factors, even though no systematic confounding mechanism exists. In a randomized trial, random confounding is addressed by randomization inference or covariate adjustment that improves precision. In observational studies, the distinction is less clean, but it remains relevant: if the analyst's adjustment set $W$ is rich enough to capture all common causes of $A$ and $Y$, then any residual imbalance is random rather than structural, and the bias vanishes asymptotically.

The practical implication is that the analyst should not treat every covariate imbalance as evidence of confounding bias. Standardized mean differences between treatment groups, for example, measure the degree of imbalance in the observed sample, but nonzero standardized differences are expected even when exchangeability holds, particularly in small samples. The relevant question is whether the imbalances are consistent with random variation or whether they reflect systematic confounding that the adjustment set does not address.

#### 2.1.5 Measurement Error Versus Omitted Confounders

Another distinction relevant to exchangeability diagnostics is between unmeasured confounders (variables that are entirely absent from the data) and mismeasured confounders (variables that are present but recorded with error). Both can produce bias, but the mechanisms and remedial strategies differ.

Measurement error in a confounder $W_j$ attenuates the degree to which conditioning on $W_j$ removes confounding. Even if all common causes of $A$ and $Y$ are present in the data, substantial measurement error can leave residual confounding that mimics the effect of an omitted variable. Classical (non-differential) measurement error in a binary confounder, for example, produces bias toward the null of the confounder-adjusted association, which in turn leaves residual confounding bias in the treatment effect estimate.

Diagnostics for measurement error are limited in the absence of validation data. When validation data are available (e.g., a gold-standard measurement on a subsample), regression calibration or multiple imputation for measurement error can be applied. When validation data are not available, the analyst should report the likely direction and magnitude of measurement-error-induced bias, using sensitivity analyses that vary the assumed misclassification rates.

#### 2.1.6 Time-Varying Confounding in Longitudinal Settings

In longitudinal settings with time-varying treatment, the exchangeability assumption takes a sequential form: at each time point $t$, the potential outcomes must be independent of the treatment at time $t$, conditional on the measured covariate and treatment history up to that point. This sequential exchangeability assumption is both stronger and harder to assess than its point-treatment counterpart, because the set of potential confounders grows with the number of time points, and because time-varying confounders may themselves be affected by prior treatment.

Diagnostics for time-varying confounding must address the specific challenge that a variable $L_t$ can simultaneously be a confounder of the treatment-outcome relationship at time $t+1$ and a mediator of the treatment effect at time $t$. Standard regression adjustment cannot handle this dual role without introducing bias; methods such as LTMLE, IPTW with stabilized weights, and the g-computation formula are designed to address it. The relevant diagnostic question is whether the measured time-varying covariates are sufficient to capture the time-varying confounding structure. This can be assessed by examining whether the DAG at each time point includes all relevant common causes, and by conducting negative control analyses that use time-varying negative control outcomes to detect residual time-varying confounding.

### 2.2 Positivity and Overlap

The positivity assumption requires that every unit in the target population has a nonzero probability of receiving each treatment level, conditional on its covariates: $P(A = a \mid W = w) > 0$ for all $a$ in the support of $A$ and all $w$ in the support of $W$. Positivity is the assumption that permits the counterfactual contrast to be estimated from observed data: if certain types of individuals never receive a particular treatment, the data contain no information about what would happen to them under that treatment, and the statistical estimand is undefined.

Unlike exchangeability, positivity has direct empirical implications that can be assessed from observed data. However, finite-sample diagnostics can only detect practical positivity violations (regions where the estimated probability of treatment is near zero), not structural positivity violations (regions where the true probability is exactly zero). The diagnostics described in this section go substantially beyond simple propensity score histograms, providing a comprehensive toolkit for assessing whether the data can support the intended estimand.

#### 2.2.1 Propensity Score Distribution Diagnostics

The most fundamental positivity diagnostic is the distribution of estimated propensity scores $\hat{g}(1 \mid W_i) = \hat{P}(A = 1 \mid W_i)$ across the study population. In the ideal case, the propensity score distributions in the treated and untreated groups overlap substantially, with neither group having mass concentrated near 0 or 1. Separation of these distributions indicates regions of the covariate space where one treatment is strongly preferred, making counterfactual estimation unreliable.

```r
## Propensity score overlap diagnostics
ps_model <- glm(A ~ W1 + W2 + W3 + W4 + W5, data = dat, family = binomial)
dat$ps <- predict(ps_model, type = "response")

## Density plot by treatment group
library(ggplot2)
ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  labs(x = "Estimated propensity score",
       y = "Density",
       fill = "Treatment") +
  theme_minimal() +
  ggtitle("Propensity Score Overlap by Treatment Group")

## Quantile summary
tapply(dat$ps, dat$A, function(x) {
  round(quantile(x, probs = c(0.01, 0.05, 0.25, 0.50, 0.75, 0.95, 0.99)), 4)
})
```

Beyond density plots, the analyst should examine the tails of the propensity score distribution with particular care. The tails are where positivity violations concentrate, and they exert disproportionate influence on weighted estimators. A propensity score of 0.01 produces an inverse probability weight of 100, meaning that a single observation effectively represents 100 individuals in the pseudo-population. The fraction of observations with propensity scores below a threshold (e.g., 0.025 or 0.05) or above its complement provides a direct summary of the severity of practical positivity violations.

#### 2.2.2 Effective Sample Size Under Weighting

Extreme inverse probability weights reduce the effective sample size---the number of independent observations that contribute meaningfully to the estimate. Even if the nominal sample size is large, the effective sample size under weighting can be much smaller, leading to high variance and unreliable inference.

The effective sample size for IPTW can be computed as:

$$
n_{\text{eff}} = \frac{\left(\sum_{i=1}^{n} w_i\right)^2}{\sum_{i=1}^{n} w_i^2}
$$

where $w_i$ are the inverse probability weights. When all weights are equal, $n_{\text{eff}} = n$. When weights are highly variable, $n_{\text{eff}}$ can be much less than $n$. An effective sample size that is a small fraction of the nominal sample size---say, less than 50\%---is a strong signal that positivity violations are degrading the precision of the analysis.

```r
## Effective sample size computation
weights_treated <- 1 / dat$ps[dat$A == 1]
weights_control <- 1 / (1 - dat$ps[dat$A == 0])

ess_treated <- sum(weights_treated)^2 / sum(weights_treated^2)
ess_control <- sum(weights_control)^2 / sum(weights_control^2)

cat("Nominal n (treated):", sum(dat$A == 1), "\n")
cat("Effective n (treated):", round(ess_treated, 1), "\n")
cat("ESS ratio (treated):", round(ess_treated / sum(dat$A == 1), 3), "\n\n")
cat("Nominal n (control):", sum(dat$A == 0), "\n")
cat("Effective n (control):", round(ess_control, 1), "\n")
cat("ESS ratio (control):", round(ess_control / sum(dat$A == 0), 3), "\n")
```

#### 2.2.3 Clever Covariate Diagnostics for TMLE

In TMLE, positivity violations manifest directly through the clever covariate (also known as the fluctuation covariate). For the average treatment effect, the clever covariate for observation $i$ takes the form:

$$
H(A_i, W_i) = \frac{I(A_i = 1)}{\hat{g}(1 \mid W_i)} - \frac{I(A_i = 0)}{1 - \hat{g}(1 \mid W_i)}
$$

When propensity scores are near the boundaries, the clever covariate grows without bound, and a small number of observations can dominate the TMLE targeting step. The analyst should examine the distribution of $|H(A_i, W_i)|$ and flag observations where the magnitude exceeds a threshold that indicates practical instability. A common heuristic is to examine observations where $|H| > 20$, though the appropriate threshold depends on the sample size and the scale of the outcome.

```r
## Clever covariate diagnostics for TMLE
H1 <- dat$A / dat$ps                    ## treated component
H0 <- (1 - dat$A) / (1 - dat$ps)       ## control component
H  <- H1 - H0                           ## clever covariate for ATE

cat("Clever covariate summary:\n")
cat("  Range:", round(range(H), 2), "\n")
cat("  Mean:", round(mean(H), 4), "\n")
cat("  SD:", round(sd(H), 2), "\n")
cat("  |H| > 10:", sum(abs(H) > 10), "observations\n")
cat("  |H| > 20:", sum(abs(H) > 20), "observations\n")
cat("  |H| > 50:", sum(abs(H) > 50), "observations\n")
cat("  Max |H|:", round(max(abs(H)), 1), "\n")

## Identify and inspect observations with extreme clever covariates
extreme_idx <- which(abs(H) > 20)
if (length(extreme_idx) > 0) {
  cat("\nObservations with |H| > 20:\n")
  print(dat[extreme_idx, c("A", "ps", "W1", "W2")])
}
```

The clever covariate diagnostic is more informative than a propensity score histogram alone because it directly reflects the influence that each observation exerts on the TMLE update. An observation with a propensity score of 0.02 in the treated group produces $H = 1/0.02 = 50$, which means the TMLE targeting step effectively treats this observation as 50 times more informative than an observation with a propensity score of 0.50. If only a handful of such observations exist, the TMLE estimate is effectively determined by those observations rather than by the full sample.

#### 2.2.4 Diagnostics Based on EIF Magnitude

The efficient influence function (EIF) provides an estimator-independent diagnostic for positivity. For the ATE, the EIF evaluated at the estimated nuisance functions takes the form:

$$
\widehat{D}(O_i) = \frac{I(A_i = 1)}{\hat{g}(1 \mid W_i)}\left(Y_i - \hat{\bar{Q}}(1, W_i)\right) - \frac{I(A_i = 0)}{1 - \hat{g}(1 \mid W_i)}\left(Y_i - \hat{\bar{Q}}(0, W_i)\right) + \hat{\bar{Q}}(1, W_i) - \hat{\bar{Q}}(0, W_i) - \hat{\psi}
$$

Observations with large $|\widehat{D}(O_i)|$ are exerting disproportionate influence on the estimate. When the EIF magnitude is large because of near-zero propensity scores (i.e., the $1/\hat{g}$ terms dominate), this directly indicates a positivity problem. When the EIF magnitude is large because of poor outcome model fit (i.e., the residual terms $Y_i - \hat{\bar{Q}}$ are large), this indicates a model diagnostics problem. Decomposing the source of large EIF values helps the analyst distinguish between positivity failures and model misspecification.

```r
## EIF-based positivity diagnostics
## After fitting TMLE, inspect the influence curve
IC <- tmle_fit$estimates$ATE$IC

cat("Influence curve diagnostics:\n")
cat("  Mean:", round(mean(IC), 6), "(should be near 0)\n")
cat("  SD:", round(sd(IC), 4), "\n")
cat("  Skewness:", round(moments::skewness(IC), 3), "\n")
cat("  Kurtosis:", round(moments::kurtosis(IC), 3), "\n")
cat("  Max |IC|:", round(max(abs(IC)), 2), "\n")
cat("  |IC| > 3*SD:", sum(abs(IC) > 3 * sd(IC)), "observations\n")

## Ratio of max IC to mean IC --- a large ratio signals instability
cat("  Max/SD ratio:", round(max(abs(IC)) / sd(IC), 1), "\n")
```

#### 2.2.5 Structural Versus Practical Positivity Violations

A structural positivity violation occurs when the true probability of treatment is exactly zero for some subpopulation: $P(A = 1 \mid W = w) = 0$ for a set of $w$ values that has positive probability. This arises, for example, when a medication is contraindicated for a specific patient subgroup, or when a treatment did not exist during part of the study period. Structural violations cannot be resolved by increasing the sample size; the counterfactual quantity is undefined for the affected subgroup.

A practical positivity violation occurs when the true probability of treatment is nonzero but very small, making estimation unreliable in finite samples. The estimated propensity scores may be near zero not because treatment is impossible but because it is rare in certain covariate strata. Practical violations can sometimes be mitigated by simplifying the propensity score model (using fewer covariates), but this trades positivity for potential confounding bias.

The distinction matters for the choice of remedial action. Structural violations require changing the estimand---restricting the target population to exclude the affected subgroup, or switching to an estimand that does not require extrapolation into unsupported regions. Practical violations may be addressed by truncation, stabilized weights, or more flexible modeling, though these approaches introduce their own biases and should be accompanied by sensitivity analyses.

#### 2.2.6 Weight Truncation Rules and Their Implications

Weight truncation (also called weight clipping or trimming) is the most common approach to mitigating practical positivity violations. The analyst caps the inverse probability weights at some upper bound, or equivalently, restricts the estimated propensity scores to lie within $[\delta, 1 - \delta]$ for some small $\delta > 0$. In TMLE, this is implemented through the `gbound` parameter; in IPTW, it is applied directly to the computed weights.

Truncation reduces variance at the cost of introducing bias. The bias arises because truncation effectively changes the estimand: the analyst is no longer targeting the ATE for the full population but rather for a population where extreme propensity scores have been modified. The magnitude of this bias depends on the proportion of observations affected by truncation and on the treatment effect heterogeneity in the truncated region.

```r
## Truncation sensitivity analysis
## Compare estimates across different truncation levels
truncation_levels <- c(0.005, 0.01, 0.025, 0.05, 0.10)

results <- data.frame(
  gbound = truncation_levels,
  ATE = NA_real_,
  SE = NA_real_,
  CI_lower = NA_real_,
  CI_upper = NA_real_,
  n_truncated = NA_integer_
)

for (i in seq_along(truncation_levels)) {
  gb <- truncation_levels[i]
  ps_trunc <- pmin(pmax(dat$ps, gb), 1 - gb)
  results$n_truncated[i] <- sum(dat$ps < gb | dat$ps > (1 - gb))

  fit_trunc <- tmle::tmle(
    Y = dat$Y, A = dat$A, W = dat[, covariates],
    Q.SL.library = "SL.glm",
    g.SL.library = "SL.glm",
    gbound = gb
  )
  results$ATE[i] <- fit_trunc$estimates$ATE$psi
  results$SE[i]  <- sqrt(fit_trunc$estimates$ATE$var.psi)
  results$CI_lower[i] <- fit_trunc$estimates$ATE$CI[1]
  results$CI_upper[i] <- fit_trunc$estimates$ATE$CI[2]
}

print(results)
cat("\nIf estimates vary substantially across truncation levels,\n")
cat("the data poorly support the full-population ATE.\n")
```

A truncation sensitivity analysis---reporting the point estimate and confidence interval across a range of truncation thresholds---provides a transparent summary of how much the results depend on the most extreme observations. If the estimate is stable across truncation levels, positivity is not a major concern for the specific dataset. If the estimate changes substantially, the analyst should report the range of estimates and consider whether the full-population ATE is a credible target.

#### 2.2.7 Estimand Redefinition and Alternative Interventions

When positivity diagnostics reveal severe violations, the most principled response is often to redefine the estimand rather than to apply truncation heuristics. Several alternatives exist, each of which naturally avoids or mitigates positivity problems.

**Restricted population ATE.** The analyst restricts the target population to the subset where positivity holds, such as the region where $\delta \leq \hat{g}(1 \mid W) \leq 1 - \delta$. This changes the causal question from "what is the effect for the whole population?" to "what is the effect for the subpopulation where both treatments are plausible?" The restricted ATE is often more credible than a truncated full-population estimate, because it explicitly acknowledges the limitation of the data.

**Modified treatment policies (MTPs).** Rather than comparing the counterfactual outcomes under treatment versus no treatment (a static intervention), modified treatment policies define the intervention as a shift in the natural value of treatment. For a continuous exposure $A$, an MTP might define the intervention as $d(A, W) = A - \delta$, reducing each individual's exposure by a fixed amount $\delta$. For a binary treatment, an MTP might define the intervention as increasing or decreasing the probability of treatment by a specified amount. MTPs naturally respect the support of the data because the intervention is defined relative to the observed treatment rather than requiring extrapolation to a treatment level that may be implausible for some individuals. The `lmtp` R package implements estimation for MTPs.

**Stochastic interventions.** Instead of deterministic interventions that assign each individual to a specific treatment level, stochastic interventions assign each individual a probability distribution over treatment levels. For example, the intervention "increase the probability of treatment by 10 percentage points for everyone" is a stochastic intervention. Stochastic interventions can be defined to remain within the support of the observed treatment mechanism, avoiding positivity violations by construction.

::: {.callout-important}
## Positivity Diagnostics Across Estimators

Different estimators are affected by positivity violations in different ways.

**IPTW** is the most sensitive to positivity violations because it relies entirely on the treatment model $g$. Extreme weights directly destabilize the estimate, and the effective sample size can become very small. IPTW with unstabilized weights is particularly vulnerable; stabilized weights $\text{sw}_i = P(A = a_i) / P(A = a_i \mid W_i)$ reduce variance but do not eliminate the fundamental problem.

**AIPW** (augmented IPW) is less sensitive than IPTW because the augmentation term $\hat{\bar{Q}}(a, W) - \hat{\psi}$ provides a "bias correction" that partially compensates for extreme weights when the outcome model is well-specified. However, when both $g$ and $\bar{Q}$ are poorly estimated in the low-support region, AIPW can be even more unstable than IPTW.

**TMLE** mitigates positivity violations through its use of a bounded loss function (logistic fluctuation for bounded outcomes), which constrains the update step and prevents the estimate from being dominated by extreme clever covariate values. However, TMLE is not immune to positivity violations: when propensity scores are near the boundary, the targeting step may fail to converge, or the resulting estimate may have high variance.

**LTMLE** in longitudinal settings faces compounded positivity challenges because the cumulative treatment mechanism is the product of time-specific mechanisms. Even moderate positivity violations at each time point can produce extreme cumulative weights that destabilize the estimate.

In all cases, the diagnostic approach is the same: examine the distribution of the relevant weights or clever covariates, compute effective sample sizes, and assess the sensitivity of the estimate to truncation. The thresholds for concern differ across estimators, but the logic is uniform.
:::

### 2.3 Consistency and Intervention Definition

The consistency assumption states that the observed outcome for an individual who received treatment $a$ equals that individual's potential outcome under treatment $a$: if $A_i = a$, then $Y_i = Y_i(a)$. This assumption appears tautological but encodes a substantive requirement: the intervention corresponding to "setting $A = a$" must be well-defined, and the treatment as recorded in the data must correspond to this intervention without ambiguity.

#### 2.3.1 Multiple Versions of Treatment

Consistency fails when the treatment variable $A$ conflates multiple distinct interventions that may have different effects on the outcome. This is the problem of "multiple versions of treatment" or "treatment variation irrelevance." For example, if $A$ indicates whether a patient received a particular drug class, but different drugs within the class have different efficacies, then the potential outcome $Y(A = 1)$ is not uniquely defined: it depends on which specific drug the patient received.

Diagnostics for treatment versioning begin with a careful examination of the treatment definition. The analyst should ask: conditional on $A = a$ and $W = w$, is the treatment received homogeneous across individuals, or does it vary in ways that could affect the outcome? If treatment is heterogeneous, the analyst should consider whether the heterogeneity can be captured by refining the treatment variable (e.g., separating drug classes into individual drugs) or whether it should be treated as a source of residual variability.

Empirically, the analyst can examine whether the treatment effect estimate changes when the treatment definition is refined or coarsened. If subdividing $A = 1$ into $A = 1a$ and $A = 1b$ produces substantially different effect estimates, this suggests that the original treatment variable conflates distinct interventions, and the consistency assumption for the coarser variable may not hold in a meaningful sense.

#### 2.3.2 Static, Dynamic, and Stochastic Interventions

The nature of the intervention implicit in the estimand has important implications for both consistency and positivity. A **static intervention** assigns a fixed treatment level to all individuals (e.g., "set $A = 1$ for everyone"). A **dynamic intervention** assigns treatment as a function of individual covariates (e.g., "treat if $W_1 > c$, do not treat otherwise"). A **stochastic intervention** assigns a probability distribution over treatment levels that may depend on covariates.

Each type of intervention carries different requirements for consistency. Static interventions require that the single treatment level $a$ be well-defined and implementable for all individuals in the target population. Dynamic interventions require that the rule $d(W)$ be well-defined and that the treatment levels it assigns are feasible for each individual given their covariates. Stochastic interventions and modified treatment policies have the weakest consistency requirements because they do not require any individual to receive a treatment level that would never occur naturally.

The analyst should verify that the chosen intervention type is compatible with the data structure and the scientific question. If the estimand involves a static intervention but the data contain individuals for whom that treatment level would never be prescribed (violating both consistency and positivity), the analyst should consider switching to a dynamic or stochastic intervention that respects the natural constraints of the data-generating process.

#### 2.3.3 Alignment Between Estimand and Data Structure

A final consistency diagnostic concerns the alignment between the formal estimand and the actual data structure. The analyst should verify that the treatment variable in the data corresponds to the intervention described in the estimand, that the outcome is measured at the time point specified by the estimand, and that the covariates $W$ are measured before treatment assignment. Misalignment between the estimand and the data structure---for example, conditioning on post-treatment variables, or measuring the outcome before the treatment could plausibly affect it---violates the causal framework and invalidates the analysis regardless of the estimation method used.

In longitudinal settings, alignment diagnostics require verifying the temporal ordering at each time point: covariates $L_t$ must be measured after treatment $A_{t-1}$ and before treatment $A_t$, and the outcome $Y$ must be measured after the final treatment. Violations of this ordering can introduce immortal time bias, reverse causation, or conditioning on post-treatment variables, each of which undermines the causal interpretation.

---

## 3. Nuisance Model Diagnostics {#sec-nuisance}

The nuisance functions---the outcome regression $\bar{Q}(A, W) = E[Y \mid A, W]$ and the treatment mechanism $g(A \mid W) = P(A \mid W)$---are the workhorses of semiparametric causal inference. They are called "nuisance" functions not because they are unimportant but because they are not the parameters of primary scientific interest; rather, they are intermediate quantities that must be estimated on the way to estimating the causal parameter. Their quality directly determines the finite-sample performance of TMLE, AIPW, and IPTW estimators. Poor nuisance estimation can produce biased causal estimates, inflated variance, invalid confidence intervals, and misleading conclusions, even when the identification assumptions hold perfectly.

The relationship between nuisance function quality and causal estimator performance depends on the specific estimator. For IPTW, only the treatment mechanism $g$ matters: if $g$ is consistently estimated, IPTW is consistent for the causal parameter regardless of how the outcome regression is specified (or whether it is specified at all). For parametric g-computation, only the outcome regression $\bar{Q}$ matters. For doubly robust estimators---TMLE and AIPW---consistency requires that at least one of $g$ or $\bar{Q}$ be consistently estimated, and efficiency is achieved when both are estimated at sufficiently fast rates. This double robustness property does not, however, eliminate the need for nuisance function diagnostics. In finite samples, the quality of both nuisance functions affects the bias, variance, and coverage of doubly robust estimators, and diagnostics provide the evidence needed to assess whether the asymptotic theory is a reliable guide to finite-sample performance.

When machine learning methods are used for nuisance estimation---as is increasingly standard---valid inference typically requires either cross-fitting (sample splitting where different folds are used to estimate nuisance functions and to construct the final estimator) or a cross-validated TMLE variant. Without cross-fitting, the overfitting behavior of flexible learners can invalidate the asymptotic linearity of the causal estimator, producing confidence intervals with incorrect coverage even when the nuisance functions are estimated consistently. The diagnostics described in this section address both the quality of the nuisance function estimates and the conditions under which the cross-fitting or cross-validation procedure is functioning as intended.

### 3.1 Super Learner Diagnostics

The Super Learner is an ensemble machine learning method that combines predictions from a library of candidate algorithms using cross-validation to select the optimal convex combination. It is the recommended approach for nuisance function estimation in the targeted learning framework because it provides an asymptotic optimality guarantee: the cross-validated risk of the Super Learner converges to the oracle risk (the risk of the best possible convex combination of the candidate learners) as the sample size grows. However, this asymptotic guarantee does not eliminate the need for diagnostics in finite samples. The Super Learner's performance depends critically on the composition of the candidate library, the cross-validation scheme, and the properties of the data.

#### 3.1.1 Cross-Validated Risk Comparisons

The most fundamental Super Learner diagnostic is the comparison of cross-validated risk across candidate learners and the ensemble. For each candidate algorithm $k$ in the library, the cross-validated risk $\hat{R}_{CV}^{(k)}$ is the average loss computed on held-out validation folds. For continuous outcomes, the loss is typically mean squared error; for binary outcomes, it is the negative log-likelihood (cross-entropy). The Super Learner ensemble risk $\hat{R}_{CV}^{SL}$ should be reported alongside the individual learner risks.

```r
## Super Learner diagnostics for the outcome regression
library(SuperLearner)

## Define a diverse library
SL.library <- c("SL.glm", "SL.glmnet", "SL.ranger",
                 "SL.xgboost", "SL.earth", "SL.nnet")

## Fit Super Learner for Q(A,W) = E[Y | A, W]
sl_Q <- SuperLearner(
  Y = dat$Y,
  X = dat[, c("A", "W1", "W2", "W3")],
  SL.library = SL.library,
  family = gaussian(),
  cvControl = list(V = 10)
)

## Cross-validated risk for each learner
cat("== Cross-Validated Risk (MSE) ==\n")
cv_risks <- data.frame(
  Learner = c(SL.library, "SuperLearner"),
  CV_Risk = c(sl_Q$cvRisk, min(sl_Q$cvRisk)),
  Coefficient = c(sl_Q$coef, NA)
)
cv_risks <- cv_risks[order(cv_risks$CV_Risk), ]
print(cv_risks, row.names = FALSE, digits = 4)

## Ratio of ensemble risk to best single learner risk
discrete_SL_risk <- min(sl_Q$cvRisk)
cat("\nDiscrete SL risk:", round(discrete_SL_risk, 4), "\n")
cat("Ensemble improvement:",
    round((1 - sl_Q$cvRisk[which.min(sl_Q$cvRisk)] / discrete_SL_risk) * 100, 1),
    "%\n")
```

Several patterns in the cross-validated risk comparison are diagnostically informative. When the ensemble risk is substantially lower than that of any individual learner, this indicates that the library captures complementary aspects of the data-generating process---different learners are performing well in different regions of the covariate space, and the ensemble benefits from combining them. When a single learner dominates (receiving nearly all the weight in the convex combination), the ensemble reduces to approximately the discrete Super Learner, which may indicate either that the dominant learner is clearly superior or that the library lacks sufficient diversity. When all learners have similar cross-validated risk, the outcome surface may be relatively simple and well-captured by any reasonable model, or the sample size may be too small to discriminate among learners.

The analyst should also compare the continuous Super Learner (the convex combination with optimized weights) to the discrete Super Learner (the single best learner selected by cross-validation). When the continuous Super Learner offers a meaningful improvement over the discrete Super Learner, this provides evidence that the ensemble is achieving genuine gains from combining predictions. When the two are essentially equivalent, the analyst may choose to report results using the simpler discrete Super Learner for transparency.

#### 3.1.2 Learner Weight Distributions

The Super Learner coefficients---the weights assigned to each candidate algorithm in the convex combination---provide a window into which modeling strategies are contributing to the ensemble. Reporting these coefficients is a minimum diagnostic requirement for any analysis that uses the Super Learner for nuisance estimation.

```r
## Learner weight diagnostics
cat("== Super Learner Coefficients (Q model) ==\n")
coef_df <- data.frame(
  Learner = SL.library,
  Weight = round(sl_Q$coef, 4)
)
coef_df <- coef_df[order(-coef_df$Weight), ]
print(coef_df, row.names = FALSE)

## Effective number of learners (analogous to effective sample size)
weights <- sl_Q$coef
eff_n_learners <- sum(weights)^2 / sum(weights^2)
cat("\nEffective number of learners:", round(eff_n_learners, 1),
    "out of", length(SL.library), "\n")
```

An effective number of learners close to 1 indicates that the ensemble is dominated by a single algorithm. This is not necessarily problematic---it may simply mean that one learner is clearly best for this problem---but it warrants attention. If the dominant learner is a highly flexible algorithm (such as a deep neural network or a heavily tuned gradient boosting model), the analyst should verify that the cross-validated risk is not overly optimistic due to information leakage or insufficient regularization. If the dominant learner is a simple algorithm (such as a main-terms logistic regression), this may indicate that the outcome surface is genuinely simple, or that the sample size is too small for more flexible learners to outperform simpler ones.

The coefficient distribution should be examined separately for the outcome regression $\bar{Q}$ and the treatment mechanism $g$, as the two functions may have different complexity. It is common for the propensity score model to be well-captured by relatively simple learners (because treatment assignment mechanisms in many studies are driven by a moderate number of strong predictors), while the outcome regression requires more flexible modeling (because the relationship between covariates and outcomes is often more complex).

#### 3.1.3 Library Sensitivity Analysis

A library sensitivity analysis assesses whether the causal estimate is robust to changes in the Super Learner library composition. The analyst fits the full causal estimator (TMLE or AIPW) using several different library specifications---for example, a library with only parametric learners, a library with only tree-based learners, the full library, and the full library with the addition of highly adaptive lasso---and compares the resulting causal estimates. If the estimates agree closely across library specifications, this provides evidence that the nuisance function estimates are sufficiently accurate across a range of modeling strategies. If the estimates diverge, this indicates that the causal conclusion depends on modeling choices for the nuisance functions, which undermines confidence in any single estimate.

```r
## Library sensitivity analysis
libraries <- list(
  "Parametric only" = c("SL.glm", "SL.glmnet"),
  "Tree-based only" = c("SL.ranger", "SL.xgboost"),
  "Full library"    = c("SL.glm", "SL.glmnet", "SL.ranger",
                         "SL.xgboost", "SL.earth", "SL.nnet"),
  "Full + HAL"      = c("SL.glm", "SL.glmnet", "SL.ranger",
                         "SL.xgboost", "SL.earth", "SL.nnet", "SL.hal9001")
)

lib_results <- data.frame(
  Library = names(libraries),
  ATE = NA_real_,
  SE = NA_real_,
  CI_lower = NA_real_,
  CI_upper = NA_real_
)

for (i in seq_along(libraries)) {
  fit_lib <- tmle::tmle(
    Y = dat$Y, A = dat$A, W = dat[, c("W1", "W2", "W3")],
    Q.SL.library = libraries[[i]],
    g.SL.library = libraries[[i]]
  )
  lib_results$ATE[i]      <- fit_lib$estimates$ATE$psi
  lib_results$SE[i]       <- sqrt(fit_lib$estimates$ATE$var.psi)
  lib_results$CI_lower[i] <- fit_lib$estimates$ATE$CI[1]
  lib_results$CI_upper[i] <- fit_lib$estimates$ATE$CI[2]
}

cat("== Library Sensitivity Analysis ==\n")
print(lib_results, row.names = FALSE, digits = 4)
cat("\nRange of ATE estimates:",
    round(diff(range(lib_results$ATE)), 4), "\n")
```

The library sensitivity analysis serves a fundamentally different purpose from the cross-validated risk comparison. The cross-validated risk comparison identifies which learners fit the data best. The library sensitivity analysis identifies whether the causal estimate is sensitive to the choice of learners. It is possible for the cross-validated risk to be similar across libraries (because all libraries fit the data reasonably well) while the causal estimates diverge (because the libraries differ in their extrapolation behavior in regions of the covariate space that are important for the causal contrast). It is also possible for the cross-validated risks to differ substantially while the causal estimates agree (because the differences in model fit occur in regions that are irrelevant to the causal parameter).

#### 3.1.4 Overfitting Detection

Overfitting occurs when a learner captures noise in the training data rather than the true underlying function, producing predictions that perform well on the training set but poorly on new data. In the Super Learner framework, cross-validation provides a natural safeguard against overfitting, but it does not eliminate the risk entirely, particularly when the sample size is small relative to the complexity of the learners.

The primary diagnostic for overfitting is the gap between training risk and cross-validated risk for each candidate learner. A large gap indicates that the learner is fitting noise in the training data. For parametric models such as logistic regression, this gap is typically small. For highly flexible learners such as random forests with very deep trees or neural networks with many hidden units, the gap can be substantial.

```r
## Overfitting diagnostic: training risk vs CV risk
## Requires refitting each learner on the full training set
train_risks <- numeric(length(SL.library))
for (k in seq_along(SL.library)) {
  pred_train <- predict(sl_Q$fitLibrary[[k]],
                        newdata = dat[, c("A", "W1", "W2", "W3")])
  if (is.list(pred_train)) pred_train <- pred_train$pred
  train_risks[k] <- mean((dat$Y - pred_train)^2)
}

overfit_df <- data.frame(
  Learner = SL.library,
  Train_Risk = round(train_risks, 4),
  CV_Risk = round(sl_Q$cvRisk, 4),
  Ratio = round(sl_Q$cvRisk / train_risks, 2)
)
cat("== Overfitting Diagnostics ==\n")
print(overfit_df, row.names = FALSE)
cat("\nRatio >> 1 indicates substantial overfitting.\n")
```

When overfitting is detected, the analyst has several options. The most direct is to add regularized variants of the offending learner to the library (e.g., replacing a random forest with default settings by a random forest with a minimum node size constraint). Another option is to remove the overfitting learner from the library, though this may reduce the library's capacity to capture complex patterns. The Super Learner's cross-validation procedure will naturally downweight overfitting learners in the ensemble, but the analyst should verify that this is occurring by examining the learner coefficients.

#### 3.1.5 Cross-Validation Scheme and Screening

The number of cross-validation folds $V$ affects both the bias and variance of the Super Learner. With $V = 10$ (the default), each learner is trained on 90\% of the data and evaluated on 10\%, providing a reasonably unbiased estimate of out-of-sample risk. Smaller values of $V$ (e.g., $V = 5$) reduce computational cost but may produce more biased risk estimates, because each learner is trained on a smaller fraction of the data. Larger values (e.g., $V = 20$ or leave-one-out) reduce bias but increase variance and computational burden.

For small samples (e.g., $n < 200$), the analyst should consider whether $V = 10$ leaves too few observations in each validation fold for stable risk estimation. In such cases, $V = 5$ or even $V = 3$ may be more appropriate, with the understanding that the cross-validated risk estimates will be noisier. Conversely, for very large samples, increasing $V$ beyond 10 provides minimal benefit and is rarely necessary.

Variable screening within the Super Learner---using a data-adaptive procedure to select a subset of covariates before fitting each candidate learner---can improve performance when the covariate space is high-dimensional. However, screening introduces an additional source of variability that the cross-validation procedure must account for. The screening step should be performed within each cross-validation fold (not on the full dataset before cross-validation), or else the cross-validated risk estimate will be optimistically biased. The `SuperLearner` package supports this through the `method` argument, which allows screening wrappers to be applied within the cross-validation loop.

### 3.2 Outcome Regression Diagnostics

The outcome regression $\bar{Q}(A, W) = E[Y \mid A, W]$ is used directly by TMLE (as the initial estimate that is subsequently targeted) and by AIPW (as the augmentation term). Even for IPTW, which does not use the outcome regression in its estimating equation, the outcome regression plays an implicit role: when the propensity score model is correct but the outcome regression is misspecified, IPTW remains consistent, but augmenting IPTW with the outcome regression (yielding AIPW) can improve efficiency. Diagnostics for the outcome regression therefore matter for all the major causal estimators, albeit in different ways.

#### 3.2.1 Calibration

Calibration measures whether the predicted values from the outcome model align with the observed outcomes. A well-calibrated model satisfies $E[Y \mid \hat{\bar{Q}}(A, W) = q] = q$ for all values of $q$ in the range of the predictions. Calibration is a weaker requirement than overall model fit (a well-calibrated model can still have high variance), but miscalibration is a particularly concerning form of misspecification because it indicates systematic bias in the predicted values.

For binary outcomes, calibration is assessed by dividing the predicted probabilities into deciles, computing the observed event rate within each decile, and plotting the observed rates against the predicted rates. A perfectly calibrated model produces points along the 45-degree line. Systematic departure from this line indicates miscalibration: predictions that are too high (overconfident) or too low (underconfident) in specific risk strata.

```r
## Calibration plot for binary outcome
## Assumes Q_pred contains predicted probabilities E[Y|A,W]
calibration_diagnostic <- function(Y, Q_pred, n_bins = 10) {
  bins <- cut(Q_pred, breaks = quantile(Q_pred, probs = seq(0, 1, 1/n_bins)),
              include.lowest = TRUE)
  cal_df <- data.frame(
    bin = levels(bins),
    predicted = tapply(Q_pred, bins, mean),
    observed  = tapply(Y, bins, mean),
    n         = tapply(Y, bins, length)
  )

  ## Calibration slope and intercept
  cal_model <- glm(Y ~ Q_pred, family = binomial(link = "logit"),
                   data = data.frame(Y = Y, Q_pred = Q_pred))
  cal_intercept <- coef(cal_model)[1]
  cal_slope     <- coef(cal_model)[2]

  cat("Calibration intercept:", round(cal_intercept, 3),
      "(ideal = 0)\n")
  cat("Calibration slope:", round(cal_slope, 3),
      "(ideal = 1)\n")

  library(ggplot2)
  ggplot(cal_df, aes(x = predicted, y = observed)) +
    geom_point(aes(size = n)) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    labs(x = "Predicted probability",
         y = "Observed proportion",
         title = "Calibration Plot for Outcome Model") +
    theme_minimal()
}
```

For continuous outcomes, calibration is assessed by plotting predicted values against observed values, with a smoothed regression line. The slope of the calibration line should be approximately 1, and the intercept should be approximately 0. Substantial departure from these values indicates that the outcome model is systematically over- or under-predicting in certain regions.

Calibration diagnostics are particularly important for TMLE because the TMLE targeting step corrects for bias in the initial outcome regression estimate, but only along the direction of the efficient influence function. Miscalibration that is orthogonal to the EIF direction will not be corrected by the targeting step and will persist in the final causal estimate. While the double robustness property provides protection against misspecification of $\bar{Q}$ when $g$ is correctly specified, the finite-sample performance of the estimator improves substantially when $\bar{Q}$ is well-calibrated.

#### 3.2.2 Residual Analysis

Residual analysis for the outcome regression follows the same principles as residual analysis for any regression model, with the important caveat that the goal is not to improve the prediction of $Y$ per se but to ensure that the outcome model is sufficiently well-specified to support the causal estimator. The residuals $e_i = Y_i - \hat{\bar{Q}}(A_i, W_i)$ should be examined for systematic patterns that indicate misspecification.

The analyst should plot the residuals against each covariate $W_j$, against the treatment variable $A$, and against the predicted values $\hat{\bar{Q}}(A_i, W_i)$. Systematic trends in these plots indicate that the outcome model is missing important features of the data-generating process. Of particular concern for causal inference are residual patterns that differ between treatment groups: if the residuals show a different relationship with a covariate $W_j$ in the treated group than in the control group, this suggests that the outcome model is misspecified in a way that directly affects the estimated treatment effect.

```r
## Residual diagnostics for outcome regression
residuals <- dat$Y - Q_pred
dat$resid <- residuals

## Residuals by treatment group
cat("== Residual Summary by Treatment Group ==\n")
cat("Treated:  mean =", round(mean(residuals[dat$A == 1]), 4),
    " SD =", round(sd(residuals[dat$A == 1]), 4), "\n")
cat("Control:  mean =", round(mean(residuals[dat$A == 0]), 4),
    " SD =", round(sd(residuals[dat$A == 0]), 4), "\n")

## Residuals vs each covariate, faceted by treatment
library(ggplot2)
p_resid <- ggplot(dat, aes(x = W1, y = resid, color = factor(A))) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~ A, labeller = labeller(A = c("0" = "Control", "1" = "Treated"))) +
  labs(x = "W1", y = "Residual", title = "Residuals vs W1 by Treatment Group") +
  theme_minimal() +
  theme(legend.position = "none")
print(p_resid)
```

For continuous outcomes, the analyst should also examine the residual distribution for non-normality and heteroscedasticity. While the asymptotic theory for TMLE and AIPW does not require normally distributed residuals, heavy-tailed residuals can degrade finite-sample performance by inflating the variance of the efficient influence function. Heteroscedasticity---variance that depends on the covariates---is less concerning for the point estimate (which is not affected by heteroscedasticity under correct mean specification) but can affect the efficiency of the estimator and the accuracy of variance estimates.

#### 3.2.3 Cross-Validated Prediction Performance

The cross-validated risk of the outcome regression provides a global summary of model fit that accounts for overfitting. For continuous outcomes, the cross-validated mean squared error $\hat{R}_{CV} = n^{-1} \sum_{i=1}^{n} (Y_i - \hat{\bar{Q}}_{-i}(A_i, W_i))^2$---where $\hat{\bar{Q}}_{-i}$ denotes the fit obtained from the training fold that excludes observation $i$---should be compared to the variance of the outcome $\text{Var}(Y)$. The ratio $R^2_{CV} = 1 - \hat{R}_{CV} / \text{Var}(Y)$ provides a cross-validated $R^2$ that summarizes the proportion of outcome variance explained by the model. For the causal estimator, a higher cross-validated $R^2$ generally translates to a more efficient estimate, because the outcome model explains more of the variation in $Y$ and leaves less to be captured by the weighting term.

However, a low cross-validated $R^2$ does not necessarily indicate a problem for the causal analysis. If the outcome is inherently noisy (high residual variance), the outcome model may be well-specified but unable to explain much of the total variation. In such cases, the causal estimator will have higher variance but need not be biased. The analyst should distinguish between a low $R^2$ due to irreducible noise and a low $R^2$ due to model misspecification by examining whether more flexible learners can achieve a substantially higher $R^2$ than simpler learners. If they cannot, the residual variance is likely irreducible; if they can, the simpler model is misspecified.

#### 3.2.4 Subgroup Predictive Performance

The global cross-validated risk may mask important heterogeneity in model performance across subgroups. The outcome model may fit well on average but poorly in a subgroup that is critical for the causal question---for example, a subgroup with low propensity scores where the weighting terms in the EIF are large. Poor outcome model performance in high-weight subgroups will disproportionately affect the causal estimate.

The analyst should compute the cross-validated risk separately for clinically relevant subgroups, for treatment groups, and---most importantly---for strata defined by the propensity score. If the outcome model performs substantially worse in the tails of the propensity score distribution (where weights are largest), this indicates that the outcome model's misspecification is concentrated precisely where it matters most for the causal estimate.

```r
## Subgroup prediction performance
## Assess outcome model fit across propensity score strata
dat$ps_quintile <- cut(dat$ps, breaks = quantile(dat$ps, probs = 0:5/5),
                        include.lowest = TRUE, labels = 1:5)

subgroup_perf <- do.call(rbind, lapply(1:5, function(q) {
  idx <- dat$ps_quintile == q
  data.frame(
    PS_quintile = q,
    n = sum(idx),
    MSE = round(mean(residuals[idx]^2), 4),
    Mean_abs_resid = round(mean(abs(residuals[idx])), 4),
    Mean_PS = round(mean(dat$ps[idx]), 3)
  )
}))

cat("== Outcome Model Performance by PS Quintile ==\n")
print(subgroup_perf, row.names = FALSE)
cat("\nLarger MSE in extreme PS quintiles suggests misspecification\n")
cat("where weights are largest and the causal estimate is most sensitive.\n")
```

#### 3.2.5 Prediction Stability Across Cross-Validation Folds

A final outcome regression diagnostic concerns the stability of predictions across cross-validation folds. In cross-fitted estimation, the predicted values $\hat{\bar{Q}}_{-k}(A_i, W_i)$ for observations in fold $k$ are produced by a model trained on all folds except $k$. If the model is stable, these predictions should be similar regardless of which fold is held out. If the predictions vary substantially across folds---that is, if $\hat{\bar{Q}}_{-k}(A_i, W_i)$ for a fixed observation $i$ changes appreciably depending on $k$---this indicates that the outcome model is sensitive to the specific training sample, which is a form of instability that can degrade the performance of the causal estimator.

This diagnostic can be implemented by computing, for each observation, the standard deviation of its predicted values across leave-one-fold-out fits, and examining the distribution of these standard deviations. Large prediction variability, particularly for observations with extreme propensity scores, signals potential instability in the causal estimate.

### 3.3 Treatment Mechanism Diagnostics

The treatment mechanism $g(A \mid W) = P(A \mid W)$ enters the causal estimator through the inverse probability weights (in IPTW), the clever covariate (in TMLE), and the augmentation term (in AIPW). Diagnostics for the treatment mechanism are fundamentally different from diagnostics for the outcome regression, because the goal of the treatment model is not prediction per se but confounding control. A treatment model that predicts treatment assignment with high accuracy may produce extreme weights that destabilize the estimator. A treatment model that sacrifices some predictive accuracy for smoother weights may produce a more reliable causal estimate. The diagnostics in this section reflect this distinction.

#### 3.3.1 Propensity Score Distribution and Tail Behavior

The distribution of estimated propensity scores has already been discussed in the context of positivity diagnostics (Section 2.2). Here, the focus shifts from positivity as an identification assumption to the practical implications of propensity score distributions for nuisance function estimation quality. Specifically, the analyst should examine whether the estimated propensity scores are well-separated from the boundaries (0 and 1), whether the distribution is smooth and unimodal, and whether there are systematic differences in the distribution across subgroups defined by key covariates.

A multimodal propensity score distribution---with distinct clusters of high and low propensity scores and few observations in between---suggests that the treatment assignment mechanism is driven by a small number of strong predictors that effectively partition the population into groups with very different treatment probabilities. While this pattern does not necessarily indicate a problem with the propensity score model (the model may be correctly capturing the true treatment mechanism), it signals a challenging estimation environment where the causal estimate will depend heavily on a few observations with moderate propensity scores.

```r
## Detailed propensity score distribution diagnostics
cat("== Propensity Score Distribution ==\n")
ps <- dat$ps

## Boundary proximity
cat("\nBoundary proximity:\n")
for (threshold in c(0.01, 0.025, 0.05, 0.10)) {
  n_low  <- sum(ps < threshold)
  n_high <- sum(ps > (1 - threshold))
  cat(sprintf("  PS < %.3f: %d obs (%.1f%%)  |  PS > %.3f: %d obs (%.1f%%)\n",
              threshold, n_low, 100 * n_low / length(ps),
              1 - threshold, n_high, 100 * n_high / length(ps)))
}

## Weight diagnostics
wt <- ifelse(dat$A == 1, 1/ps, 1/(1-ps))
cat("\nInverse probability weight summary:\n")
cat("  Mean:", round(mean(wt), 2), "\n")
cat("  SD:", round(sd(wt), 2), "\n")
cat("  Max:", round(max(wt), 1), "\n")
cat("  Coefficient of variation:", round(sd(wt)/mean(wt), 2), "\n")
cat("  Top 1% of weights account for",
    round(100 * sum(sort(wt, decreasing = TRUE)[1:ceiling(0.01*length(wt))]) /
            sum(wt), 1),
    "% of total weight\n")
```

#### 3.3.2 Covariate Balance Diagnostics

Covariate balance is the most practically relevant diagnostic for the treatment model, because it directly measures whether the weighting or adjustment procedure has succeeded in making the treatment groups comparable with respect to measured confounders. The standardized mean difference (SMD) for each covariate $W_j$ is defined as:

$$
\text{SMD}_j = \frac{\bar{W}_{j,1} - \bar{W}_{j,0}}{\sqrt{(s_{j,1}^2 + s_{j,0}^2)/2}}
$$

where $\bar{W}_{j,a}$ and $s_{j,a}^2$ are the mean and variance of $W_j$ in treatment group $a$. After weighting by the inverse probability weights, the weighted SMD should be computed using the weighted means and variances. A common threshold for adequate balance is $|\text{SMD}| < 0.1$, though this is a convention rather than a principled cutoff, and the appropriate threshold depends on the context.

```r
## Covariate balance diagnostics (Love plot)
compute_smd <- function(W, A, weights = NULL) {
  if (is.null(weights)) weights <- rep(1, length(A))

  smds <- sapply(1:ncol(W), function(j) {
    w1 <- weights[A == 1]
    w0 <- weights[A == 0]
    x1 <- W[A == 1, j]
    x0 <- W[A == 0, j]

    mean1 <- weighted.mean(x1, w1)
    mean0 <- weighted.mean(x0, w0)
    var1  <- sum(w1 * (x1 - mean1)^2) / sum(w1)
    var0  <- sum(w0 * (x0 - mean0)^2) / sum(w0)

    (mean1 - mean0) / sqrt((var1 + var0) / 2)
  })
  names(smds) <- colnames(W)
  return(smds)
}

W <- dat[, c("W1", "W2", "W3")]
ipw <- ifelse(dat$A == 1, 1/dat$ps, 1/(1 - dat$ps))

smd_unadj <- compute_smd(W, dat$A)
smd_weighted <- compute_smd(W, dat$A, weights = ipw)

balance_df <- data.frame(
  Covariate = rep(colnames(W), 2),
  SMD = c(smd_unadj, smd_weighted),
  Method = rep(c("Unadjusted", "IPW-Weighted"), each = ncol(W))
)

library(ggplot2)
ggplot(balance_df, aes(x = abs(SMD), y = Covariate,
                        color = Method, shape = Method)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0.1, linetype = "dashed", color = "gray50") +
  labs(x = "Absolute Standardized Mean Difference",
       title = "Covariate Balance (Love Plot)") +
  theme_minimal()

cat("== Covariate Balance Summary ==\n")
cat("Unadjusted max |SMD|:", round(max(abs(smd_unadj)), 3), "\n")
cat("Weighted max |SMD|:", round(max(abs(smd_weighted)), 3), "\n")
cat("Unadjusted mean |SMD|:", round(mean(abs(smd_unadj)), 3), "\n")
cat("Weighted mean |SMD|:", round(mean(abs(smd_weighted)), 3), "\n")
```

Covariate balance should be assessed not only for the first moments (means) of each covariate but also for higher-order terms---squares, interactions, and quantiles---when these are relevant to the outcome. A treatment model may achieve balance on the marginal means of each covariate while leaving substantial imbalance in their joint distribution or in their nonlinear transformations. For high-dimensional covariate sets, examining balance for all pairwise interactions may be impractical, but the analyst should at a minimum assess balance for squares of continuous covariates and for interactions between the strongest confounders.

#### 3.3.3 Balance in High Dimensions

When the covariate set $W$ is high-dimensional, individual SMDs for each covariate become unwieldy to examine, and it becomes necessary to use summary measures of overall balance. Several such measures are available.

The maximum absolute SMD across all covariates provides a worst-case measure. If the maximum SMD is below the threshold, all individual SMDs are also below the threshold. The average absolute SMD provides a central tendency measure. The post-weighting C-statistic (the area under the ROC curve for predicting treatment from covariates, using the weighted sample) provides a global measure of residual imbalance: a C-statistic near 0.5 indicates that the covariates no longer discriminate between treatment groups after weighting, suggesting adequate balance. A C-statistic substantially above 0.5 indicates residual imbalance.

```r
## Post-weighting C-statistic
## Fit a logistic regression predicting A from W using weighted data
post_wt_model <- glm(A ~ W1 + W2 + W3, data = dat,
                      family = binomial, weights = ipw)
post_wt_pred <- predict(post_wt_model, type = "response")

## C-statistic (AUC)
if (requireNamespace("pROC", quietly = TRUE)) {
  auc_post <- pROC::auc(dat$A, post_wt_pred)
  cat("Post-weighting C-statistic:", round(as.numeric(auc_post), 3),
      "(ideal = 0.5)\n")
}
```

#### 3.3.4 Balance Versus Predictive Accuracy

A critical conceptual distinction in treatment model diagnostics is between predictive accuracy and confounding control. These are related but not identical objectives, and optimizing one does not necessarily optimize the other.

Predictive accuracy---measured by the AUC, Brier score, or cross-validated log-likelihood---quantifies how well the propensity score model discriminates between treated and untreated individuals. High predictive accuracy is desirable insofar as it indicates that the model has captured the factors that drive treatment assignment. However, a highly predictive propensity score model can produce extreme weights (because it assigns very high or very low probabilities of treatment to many individuals), which increases the variance of the causal estimator and can produce practical positivity violations.

Confounding control---measured by covariate balance after weighting---quantifies whether the treatment model has succeeded in making the treatment groups comparable. A simpler propensity score model that sacrifices some predictive accuracy may produce more stable weights and better balance, particularly when the true treatment mechanism involves interactions or nonlinearities that the simpler model does not capture. In such cases, the simpler model "smooths" the propensity scores toward the center of the distribution, reducing weight variability at the cost of potentially incomplete confounding control.

The analyst should report both predictive accuracy and covariate balance, and should recognize that the two may conflict. When they do conflict, covariate balance is the more directly relevant diagnostic for the causal analysis, because the goal of the treatment model is not to predict treatment but to remove confounding. A treatment model that achieves perfect covariate balance with moderate predictive accuracy is preferable, for causal inference purposes, to a treatment model that achieves high predictive accuracy with poor balance and extreme weights.

::: {.callout-important}
## The Propensity Score Paradox

An important subtlety in propensity score estimation has been called the "propensity score paradox": including more covariates in the propensity score model can worsen the performance of the causal estimator, even when all included covariates are genuine confounders. This occurs because each additional covariate increases the dimensionality of the propensity score model, which can produce more variable propensity score estimates and more extreme weights, particularly in small samples. The paradox is most acute when the additional covariates are strong predictors of treatment but weak predictors of the outcome (instruments or near-instruments), because these variables increase weight variability without substantially reducing confounding bias.

The practical implication is that the propensity score model should not be constructed by including every available covariate. Variables that are strong predictors of the outcome but weak predictors of treatment should generally be included (they reduce variance without increasing weight instability). Variables that are strong predictors of treatment but weak predictors of the outcome should be included cautiously, as they may increase variance more than they reduce bias. Variables that predict neither treatment nor outcome should be excluded. The DAG provides the theoretical framework for these decisions; the diagnostics described in this section provide the empirical evidence needed to evaluate their consequences.
:::

---

## 4. Double Robust and EIF-Based Diagnostics {#sec-eif}

The efficient influence function (EIF) occupies a central position in the theory of semiparametric causal inference. It defines the efficient estimator, characterizes the double robustness property, determines the semiparametric efficiency bound, and provides the basis for inference. Diagnostics derived from the EIF therefore provide a uniquely powerful lens for evaluating the performance of doubly robust estimators such as TMLE and AIPW. This section develops a comprehensive set of EIF-based diagnostics, beginning with the mathematical structure of the EIF and proceeding to its diagnostic implications.

### 4.1 Efficient Influence Function Structure

For the average treatment effect $\psi_0 = E[E[Y \mid A=1, W] - E[Y \mid A=0, W]]$ under the nonparametric model, the efficient influence function at the true data-generating distribution $P_0$ is:

$$
D(O; \bar{Q}_0, g_0, \psi_0) = \frac{I(A=1)}{g_0(1 \mid W)}\left(Y - \bar{Q}_0(1, W)\right) - \frac{I(A=0)}{1 - g_0(1 \mid W)}\left(Y - \bar{Q}_0(0, W)\right) + \bar{Q}_0(1, W) - \bar{Q}_0(0, W) - \psi_0
$$

where $\bar{Q}_0(a, W) = E_0[Y \mid A=a, W]$ is the true outcome regression and $g_0(1 \mid W) = P_0(A=1 \mid W)$ is the true treatment mechanism. This function has mean zero under the true distribution: $E_{P_0}[D(O; \bar{Q}_0, g_0, \psi_0)] = 0$.

The EIF can be decomposed into three functionally distinct components that have different diagnostic implications:

**The $g$-dependent weighting component:** $\frac{I(A=a)}{g_0(a \mid W)}(Y - \bar{Q}_0(a, W))$ captures the contribution of the inverse probability weighting. When $g$ is misspecified, this component introduces bias. When $g$ is near zero, this component becomes large, introducing variance. The magnitude of this component is directly related to the clever covariate in TMLE and the weighting term in IPTW.

**The outcome regression component:** $\bar{Q}_0(1, W) - \bar{Q}_0(0, W)$ captures the contribution of the outcome model to the causal contrast. When $\bar{Q}$ is well-specified, this component provides a consistent estimate of the conditional average treatment effect (CATE), and the weighting component serves only to improve efficiency. When $\bar{Q}$ is misspecified, the weighting component corrects for the bias, provided $g$ is correctly specified.

**The centering constant:** $-\psi_0$ ensures that the EIF has mean zero. In practice, $\psi_0$ is replaced by the estimated causal parameter $\hat{\psi}$.

The double robustness property is encoded in the structure of the EIF. The key identity is:

$$
E_{P_0}[D(O; \bar{Q}, g, \psi_0)] = E_{P_0}\left[\left(\frac{g_0(A \mid W)}{g(A \mid W)} - 1\right)\left(\bar{Q}_0(A, W) - \bar{Q}(A, W)\right)\right] \cdot (\text{terms})
$$

This expression equals zero whenever either $g = g_0$ (the treatment model is correct) or $\bar{Q} = \bar{Q}_0$ (the outcome model is correct). The bias of the estimator is therefore a product of the errors in both nuisance functions---a second-order term that converges to zero faster than either individual error, provided both are estimated at rates faster than $n^{-1/4}$. This rate condition is the formal requirement for the "doubly robust" estimator to achieve $\sqrt{n}$-consistency when both nuisance functions are estimated with machine learning.

### 4.2 Empirical Mean of the EIF as a Diagnostic

The empirical mean of the estimated EIF, $P_n \hat{D} = n^{-1}\sum_{i=1}^{n} \hat{D}(O_i)$, is one of the most informative single diagnostic quantities for a doubly robust estimator.

For TMLE, the empirical mean of the EIF is exactly zero by construction. TMLE works by updating the initial outcome regression estimate $\hat{\bar{Q}}^0$ through a fluctuation that solves the empirical EIF equation $P_n \hat{D} = 0$. If the TMLE implementation reports a nonzero empirical EIF mean, this indicates a numerical problem in the targeting step---typically a convergence failure in the logistic fluctuation regression, or a boundary condition that prevents the update from achieving the desired solution.

For AIPW (the one-step estimator), the empirical mean of the EIF is not exactly zero but converges to zero at rate $o_P(n^{-1/2})$ under correct specification of at least one nuisance function. A nonzero empirical EIF mean for AIPW indicates either finite-sample bias or nuisance function misspecification. The magnitude of this mean, relative to the standard error of the estimate, provides a rough diagnostic: if $|P_n \hat{D}| / \text{SE}(\hat{\psi})$ is large (say, greater than 0.5), the finite-sample bias may be substantial relative to the sampling variability.

```r
## Empirical EIF mean diagnostic
## For TMLE
IC_tmle <- tmle_fit$estimates$ATE$IC
cat("== Empirical EIF Mean ==\n")
cat("TMLE:  mean(IC) =", format(mean(IC_tmle), digits = 6),
    "(should be ~0 by construction)\n")
cat("       |mean(IC)| / SE =",
    round(abs(mean(IC_tmle)) / (sd(IC_tmle) / sqrt(length(IC_tmle))), 4), "\n")

## For AIPW (one-step estimator)
## Manually compute the one-step EIF
Q1 <- predict(Q_model, newdata = transform(dat, A = 1))
Q0 <- predict(Q_model, newdata = transform(dat, A = 0))
ps <- dat$ps
A  <- dat$A
Y  <- dat$Y

EIF_onestep <- A/ps * (Y - Q1) - (1-A)/(1-ps) * (Y - Q0) + Q1 - Q0
psi_onestep <- mean(EIF_onestep)  ## one-step estimate

## Center the EIF
EIF_centered <- EIF_onestep - psi_onestep
cat("\nAIPW:  mean(EIF) =", format(mean(EIF_centered), digits = 6), "\n")
cat("       |mean(EIF)| / SE =",
    round(abs(mean(EIF_centered)) / (sd(EIF_centered) / sqrt(length(EIF_centered))), 4), "\n")
```

Comparing the empirical EIF mean between TMLE and AIPW provides a useful cross-check. When both are near zero, this provides reassurance that the nuisance functions are adequate. When the AIPW empirical EIF mean is substantially nonzero while the TMLE empirical EIF mean is zero (by construction), the difference highlights the bias-correction role of the TMLE targeting step. If the TMLE targeting step is performing a large correction (moving the estimate substantially from the initial plug-in value to the targeted value), this is a signal that the initial outcome regression estimate was substantially biased along the EIF direction, and the analyst should investigate why.

### 4.3 Variance Decomposition and Outlier Influence

The variance of the estimated EIF, $\hat{\sigma}^2 = n^{-1}\sum_{i=1}^{n} \hat{D}(O_i)^2$, is the basis for the Wald-type confidence interval $\hat{\psi} \pm z_{1-\alpha/2} \hat{\sigma} / \sqrt{n}$. Examining the contributions of individual observations to this variance reveals whether the estimate is driven by a broad base of observations or dominated by a few outliers.

The contribution of observation $i$ to the total EIF variance is proportional to $\hat{D}(O_i)^2$. The analyst should examine the distribution of $|\hat{D}(O_i)|$ and identify observations with disproportionately large contributions. A useful summary is the fraction of the total EIF variance attributable to the top 1\%, 5\%, and 10\% of observations by $|\hat{D}(O_i)|$. If a small fraction of observations accounts for a large fraction of the variance, the estimate is fragile: removing or perturbing these observations would substantially change the result.

```r
## Variance decomposition and outlier influence
IC <- tmle_fit$estimates$ATE$IC
IC_sq <- IC^2
total_var <- sum(IC_sq)

## Contribution of top observations
IC_ranked <- sort(IC_sq, decreasing = TRUE)
n <- length(IC)

cat("== EIF Variance Decomposition ==\n")
for (pct in c(0.01, 0.05, 0.10)) {
  top_n <- ceiling(pct * n)
  top_var <- sum(IC_ranked[1:top_n])
  cat(sprintf("  Top %.0f%% of observations (n=%d): %.1f%% of total variance\n",
              pct * 100, top_n, 100 * top_var / total_var))
}

## Leave-one-out influence on the point estimate
## The change in psi_hat from removing observation i is approximately IC_i / n
loo_influence <- IC / n
cat("\n== Leave-One-Out Influence ==\n")
cat("Max influence magnitude:", round(max(abs(loo_influence)), 6), "\n")
cat("Point estimate:", round(mean(IC) + mean(Q1 - Q0), 4), "\n")
cat("Max possible shift from removing one obs:",
    round(max(abs(loo_influence)), 6), "\n")

## Flag high-influence observations
threshold <- 3 * sd(IC)
high_influence_idx <- which(abs(IC) > threshold)
cat("\nObservations with |IC| > 3*SD:", length(high_influence_idx), "\n")
if (length(high_influence_idx) > 0 & length(high_influence_idx) <= 20) {
  cat("\nHigh-influence observations:\n")
  high_inf_df <- data.frame(
    Index = high_influence_idx,
    IC = round(IC[high_influence_idx], 3),
    A = dat$A[high_influence_idx],
    PS = round(dat$ps[high_influence_idx], 4),
    Y = round(dat$Y[high_influence_idx], 3)
  )
  print(high_inf_df, row.names = FALSE)
}
```

The source of large EIF values is diagnostically important. An observation can have a large $|\hat{D}(O_i)|$ for two distinct reasons. First, the inverse probability weighting component may be large because $g(A_i \mid W_i)$ is near zero, which indicates a positivity problem. Second, the residual $(Y_i - \hat{\bar{Q}}(A_i, W_i))$ may be large because the outcome model is poorly calibrated for observation $i$, which indicates a model fit problem. The analyst can decompose each large EIF value into its weighting and residual components to determine the source.

```r
## Decompose large EIF values into weighting and residual components
## For treated observations
weight_component <- ifelse(dat$A == 1, 1/dat$ps, 1/(1 - dat$ps))
residual_component <- ifelse(dat$A == 1,
                              abs(dat$Y - Q1),
                              abs(dat$Y - Q0))

## For high-influence observations
if (length(high_influence_idx) > 0) {
  decomp_df <- data.frame(
    Index = high_influence_idx,
    IC_magnitude = round(abs(IC[high_influence_idx]), 3),
    Weight = round(weight_component[high_influence_idx], 2),
    Residual = round(residual_component[high_influence_idx], 3),
    Primary_source = ifelse(
      weight_component[high_influence_idx] > 5, "Positivity",
      ifelse(residual_component[high_influence_idx] > 2 * sd(dat$Y),
             "Model fit", "Mixed"))
  )
  cat("\n== Source Decomposition of High-Influence Observations ==\n")
  print(decomp_df, row.names = FALSE)
}
```

This decomposition directly informs the appropriate remedial action. If high-influence observations are driven primarily by extreme weights (positivity problems), the analyst should consider truncation, estimand redefinition, or modified treatment policies. If they are driven primarily by large residuals (model fit problems), the analyst should consider expanding the Super Learner library, adding relevant covariates to the outcome model, or investigating whether the large residuals represent genuine outliers in the data.

### 4.4 Targeting Step Convergence in TMLE

The TMLE targeting step updates the initial outcome regression estimate $\hat{\bar{Q}}^0$ by fitting a univariate logistic regression (for bounded outcomes) or linear regression (for unbounded outcomes) of the outcome $Y$ on the clever covariate $H(A, W)$, using $\hat{\bar{Q}}^0$ as an offset. The coefficient $\epsilon$ from this regression determines the magnitude of the update: the targeted estimate is $\hat{\bar{Q}}^*(A, W) = \text{expit}(\text{logit}(\hat{\bar{Q}}^0(A, W)) + \epsilon H(A, W))$ for bounded outcomes. Several aspects of this targeting step admit diagnostic examination.

**The magnitude of $\epsilon$.** The fluctuation parameter $\epsilon$ quantifies the correction applied by the targeting step. A small $\epsilon$ (close to zero) indicates that the initial outcome regression estimate was already nearly unbiased along the EIF direction---the plug-in estimate and the targeted estimate are nearly identical. A large $\epsilon$ indicates substantial bias correction, which may signal poor initial estimation of $\bar{Q}$ or $g$, or it may reflect a genuine difference between the plug-in and efficient estimators. As a rough heuristic, $|\epsilon| > 0.5$ warrants investigation.

**Convergence of the targeting step.** The logistic fluctuation regression should converge without warnings. Convergence failures can occur when the clever covariate has extreme values (due to positivity violations) or when the initial outcome regression produces predicted values near the boundary of the parameter space (0 or 1 for probabilities). Convergence warnings from the `glm` call in the targeting step should be reported and investigated.

**The plug-in versus targeted estimate.** The difference between the initial plug-in estimate $\hat{\psi}^0 = n^{-1}\sum_i (\hat{\bar{Q}}^0(1, W_i) - \hat{\bar{Q}}^0(0, W_i))$ and the targeted estimate $\hat{\psi}^* = n^{-1}\sum_i (\hat{\bar{Q}}^*(1, W_i) - \hat{\bar{Q}}^*(0, W_i))$ measures the impact of the targeting step. A large difference indicates that the initial estimate was substantially biased, and the targeting step is performing an important correction. While this correction is theoretically valid, a large correction in finite samples may indicate that the estimator is operating in a regime where the asymptotic theory is unreliable.

```r
## Targeting step diagnostics for TMLE
## Access epsilon from TMLE fit
cat("== TMLE Targeting Step Diagnostics ==\n")

## Plug-in estimate (before targeting)
Q1_init <- tmle_fit$Qinit$Q[, "Q1W"]
Q0_init <- tmle_fit$Qinit$Q[, "Q0W"]
psi_plugin <- mean(Q1_init - Q0_init)

## Targeted estimate
Q1_star <- tmle_fit$Qstar[, "Q1W"]
Q0_star <- tmle_fit$Qstar[, "Q0W"]
psi_targeted <- tmle_fit$estimates$ATE$psi

cat("Plug-in estimate:", round(psi_plugin, 4), "\n")
cat("Targeted estimate:", round(psi_targeted, 4), "\n")
cat("Difference (targeting correction):",
    round(psi_targeted - psi_plugin, 4), "\n")
cat("Relative correction:",
    round(abs(psi_targeted - psi_plugin) / abs(psi_targeted) * 100, 1), "%\n")

## Fluctuation parameter
cat("\nFluctuation epsilon:",
    round(tmle_fit$epsilon, 6), "\n")
cat("|epsilon| > 0.5:", abs(tmle_fit$epsilon) > 0.5,
    "(large correction if TRUE)\n")
```

**Iterative TMLE.** Standard TMLE performs a single targeting step. Iterative TMLE repeats the targeting step---refitting the fluctuation regression using the updated outcome regression from the previous iteration---until the empirical EIF equation $P_n \hat{D} = 0$ is solved to within a specified tolerance. Iterative TMLE can improve finite-sample performance when a single targeting step is insufficient to solve the EIF equation, which can occur when the clever covariate distribution is highly skewed or when the initial estimate is far from the truth.

**Collaborative TMLE (C-TMLE).** C-TMLE adaptively selects which components of the nuisance function to target, using a data-adaptive procedure that balances bias reduction and variance control. C-TMLE is particularly relevant when positivity violations are present, because it can choose to target the outcome regression more aggressively in regions where the propensity score is bounded away from zero and less aggressively in regions where positivity violations would introduce excessive variance. Diagnostics for C-TMLE include the sequence of selected targeting steps, the cross-validated risk at each step, and the comparison with standard TMLE. When C-TMLE and standard TMLE produce substantially different estimates, this is a signal that the standard TMLE is being influenced by observations in low-support regions, and C-TMLE's adaptive approach may be more appropriate.

### 4.5 Failure Modes of AIPW Versus TMLE

AIPW and TMLE are both doubly robust estimators that target the same statistical estimand, but they differ in important ways that affect their diagnostic profiles and failure modes.

**AIPW (one-step estimator).** The AIPW estimator is computed as:

$$
\hat{\psi}_{AIPW} = \frac{1}{n}\sum_{i=1}^{n}\left[\frac{A_i}{\hat{g}(1 \mid W_i)}(Y_i - \hat{\bar{Q}}(1, W_i)) - \frac{1-A_i}{1-\hat{g}(1 \mid W_i)}(Y_i - \hat{\bar{Q}}(0, W_i)) + \hat{\bar{Q}}(1, W_i) - \hat{\bar{Q}}(0, W_i)\right]
$$

AIPW is a closed-form estimator that does not iterate. It adds a single correction term to the plug-in estimate. For bounded outcomes (e.g., probabilities), AIPW can produce estimates outside the parameter space $[0, 1]$, because the correction term is not constrained. This boundary violation is a diagnostic red flag: it indicates that the correction term is so large that it overwhelms the plug-in estimate, which typically occurs when extreme weights coincide with large residuals.

**TMLE (substitution estimator).** TMLE produces an updated estimate of the outcome regression that is then used to compute the plug-in causal parameter. Because the updated outcome regression is constrained to respect the parameter space (through the logistic submodel for bounded outcomes), TMLE cannot produce estimates outside $[0, 1]$ for probability-scale parameters. This substitution property provides a built-in safeguard against extreme estimates.

The practical implications of these differences for diagnostics are as follows.

First, if the AIPW estimate is outside the parameter space, this is a definitive signal of instability, and the analyst should prefer TMLE (which respects the bounds) or investigate the source of the instability. Second, if AIPW and TMLE produce similar estimates, this provides reassurance that both estimators are functioning adequately. Third, if they produce different estimates, the difference can be attributed to the targeting procedure (TMLE iterates to solve the EIF equation, while AIPW applies a single linear correction) and to the bounding behavior (TMLE constrains the outcome regression to the parameter space).

```r
## AIPW vs TMLE comparison
## Compute AIPW estimate manually
Q1_hat <- predict(Q_model, newdata = transform(dat, A = 1))
Q0_hat <- predict(Q_model, newdata = transform(dat, A = 0))
ps_hat <- dat$ps

aipw_obs <- dat$A/ps_hat * (dat$Y - Q1_hat) -
            (1-dat$A)/(1-ps_hat) * (dat$Y - Q0_hat) +
            Q1_hat - Q0_hat
psi_aipw <- mean(aipw_obs)
se_aipw  <- sd(aipw_obs) / sqrt(nrow(dat))

cat("== AIPW vs TMLE Comparison ==\n")
cat("AIPW estimate:", round(psi_aipw, 4),
    " SE:", round(se_aipw, 4), "\n")
cat("TMLE estimate:", round(tmle_fit$estimates$ATE$psi, 4),
    " SE:", round(sqrt(tmle_fit$estimates$ATE$var.psi), 4), "\n")
cat("Difference:", round(psi_aipw - tmle_fit$estimates$ATE$psi, 4), "\n")

## Check for parameter space violations (binary outcome)
if (all(dat$Y %in% c(0, 1))) {
  Q1_aipw <- mean(dat$A/ps_hat * dat$Y + (1 - dat$A/ps_hat) * Q1_hat)
  Q0_aipw <- mean((1-dat$A)/(1-ps_hat) * dat$Y +
                    (1 - (1-dat$A)/(1-ps_hat)) * Q0_hat)
  cat("\nAIPW E[Y(1)]:", round(Q1_aipw, 4))
  cat("  E[Y(0)]:", round(Q0_aipw, 4), "\n")
  if (Q1_aipw < 0 | Q1_aipw > 1 | Q0_aipw < 0 | Q0_aipw > 1) {
    cat("WARNING: AIPW counterfactual means outside [0,1].\n")
    cat("This indicates instability. Prefer TMLE.\n")
  }
}
```

**When double robustness masks instability.** A subtle failure mode of both AIPW and TMLE occurs when both nuisance functions are moderately misspecified but their errors partially cancel, producing a point estimate that appears reasonable but confidence intervals with incorrect coverage. This failure mode is difficult to detect from the data alone, because the usual diagnostics (EIF mean near zero, reasonable variance) may all look acceptable. The most effective diagnostic for this failure mode is a simulation study using a data-generating process calibrated to the observed data: if the simulated coverage is substantially below nominal, the double robustness is not providing adequate protection in the finite sample at hand.

Another manifestation of masked instability occurs when the EIF variance is dominated by a small number of observations. The point estimate may be stable (because the mean of the EIF is constrained to be near zero), but the variance estimate may be unreliable, leading to confidence intervals that are either too narrow (if the high-influence observations happen to have small EIF values in the particular sample) or too wide (if they happen to have large values). The variance decomposition diagnostics described in Section 4.3 are designed to detect this pattern.

---

## 5. Estimator Performance Diagnostics {#sec-performance}

The preceding sections addressed diagnostics for the identification assumptions (Section 2), the nuisance function estimates (Section 3), and the influence function structure (Section 4). This section turns to the performance of the causal estimator itself: given the fitted nuisance functions and the chosen estimation procedure, how reliable is the resulting causal estimate? Estimator performance diagnostics assess whether the finite-sample behavior of the estimator is consistent with its asymptotic theory, whether the variance estimates and confidence intervals are trustworthy, and whether alternative estimation strategies would produce substantially different conclusions.

### 5.1 Finite-Sample Bias and Monte Carlo Diagnostics

The asymptotic theory for TMLE and AIPW guarantees $\sqrt{n}$-consistency and asymptotic normality under conditions on the nuisance function estimation rates. In finite samples, however, these guarantees may not hold. The most direct way to assess finite-sample performance is through Monte Carlo simulation: the analyst defines a data-generating process (DGP), draws many samples, applies the estimation procedure to each sample, and computes summary statistics that characterize the estimator's behavior.

The key metrics for Monte Carlo diagnostics are:

**Bias**: $\text{Bias} = E[\hat{\psi}] - \psi_0$, estimated by the mean of the estimates across simulations minus the true parameter value. Nonzero bias indicates that the estimator is not centered on the truth in finite samples, either because the nuisance functions are insufficiently well-estimated or because the sample size is too small for the asymptotic bias correction to be effective.

**Root mean squared error (RMSE)**: $\text{RMSE} = \sqrt{E[(\hat{\psi} - \psi_0)^2]}$, which combines bias and variance into a single measure of estimation accuracy.

**Empirical coverage**: The fraction of simulated confidence intervals that contain the true parameter value. Nominal 95\% coverage should be near 0.95. Coverage below nominal indicates that the confidence intervals are too narrow---the variance is underestimated or the estimator's sampling distribution departs from normality.

**Average confidence interval width**: The mean width of the confidence intervals across simulations. Comparing the average width to the empirical standard deviation of the estimates provides a diagnostic for variance estimation: if the average width is substantially narrower than $2 \times 1.96 \times \text{SD}(\hat{\psi})$, the variance is underestimated.

```r
## Monte Carlo simulation for TMLE performance diagnostics
run_simulation <- function(n, B = 500, true_ATE = 0.5) {
  results <- data.frame(
    ATE = numeric(B),
    SE = numeric(B),
    CI_lower = numeric(B),
    CI_upper = numeric(B),
    covers = logical(B)
  )

  for (b in 1:B) {
    ## Generate data from known DGP
    W1 <- rnorm(n)
    W2 <- rbinom(n, 1, 0.3)
    A  <- rbinom(n, 1, plogis(-0.5 + 0.4 * W1 + 1.5 * W2))
    Y  <- true_ATE * A + 0.8 * W1 - 0.3 * W2 + rnorm(n, sd = 0.5)
    W  <- data.frame(W1 = W1, W2 = W2)

    fit <- tmle::tmle(Y = Y, A = A, W = W,
                      Q.SL.library = "SL.glm",
                      g.SL.library = "SL.glm")

    results$ATE[b]      <- fit$estimates$ATE$psi
    results$SE[b]       <- sqrt(fit$estimates$ATE$var.psi)
    results$CI_lower[b] <- fit$estimates$ATE$CI[1]
    results$CI_upper[b] <- fit$estimates$ATE$CI[2]
    results$covers[b]   <- (fit$estimates$ATE$CI[1] <= true_ATE) &
                           (fit$estimates$ATE$CI[2] >= true_ATE)
  }

  cat("== Monte Carlo Simulation Results ==\n")
  cat("Sample size:", n, " | Simulations:", B, "\n")
  cat("True ATE:", true_ATE, "\n\n")
  cat("Bias:", round(mean(results$ATE) - true_ATE, 4), "\n")
  cat("Empirical SD:", round(sd(results$ATE), 4), "\n")
  cat("Mean estimated SE:", round(mean(results$SE), 4), "\n")
  cat("SE ratio (mean SE / empirical SD):",
      round(mean(results$SE) / sd(results$ATE), 3), "\n")
  cat("RMSE:", round(sqrt(mean((results$ATE - true_ATE)^2)), 4), "\n")
  cat("Coverage:", round(mean(results$covers), 3), "\n")
  cat("Mean CI width:", round(mean(results$CI_upper - results$CI_lower), 4), "\n")

  invisible(results)
}

## Run at different sample sizes to assess finite-sample behavior
## sim_200 <- run_simulation(n = 200)
## sim_500 <- run_simulation(n = 500)
## sim_1000 <- run_simulation(n = 1000)
```

The design of the simulation DGP is as important as the summary statistics. The analyst should construct DGPs that stress-test the specific features of the problem at hand, rather than using generic simulations. Relevant stress tests include:

**Near-positivity violations.** Setting the treatment mechanism to produce propensity scores near the boundaries (e.g., $P(A=1 \mid W) \in [0.02, 0.98]$) and examining how much the bias, RMSE, and coverage degrade relative to a well-separated DGP.

**Outcome model misspecification.** Fitting the outcome regression with a simpler model (e.g., main-terms logistic regression) when the true outcome surface includes interactions or nonlinearities, and assessing whether the double robustness of TMLE/AIPW compensates for the misspecification.

**Rare outcomes.** For binary outcomes with low event rates, the effective information about the treatment effect is limited, and the normal approximation may be poor. Simulations with event rates of 1\%--5\% can reveal whether the confidence intervals have adequate coverage.

**Small samples.** Sample sizes below 200 are particularly challenging for semiparametric estimators that rely on cross-fitting, because the training folds may be too small for the nuisance function estimators to perform well. Simulations at $n = 50, 100, 200$ can identify the minimum sample size at which the estimator is reliable.

### 5.2 Bootstrap Diagnostics

The nonparametric bootstrap provides an independent assessment of the estimator's variability that can be compared to the influence-function-based variance estimate. For each bootstrap replicate $b = 1, \ldots, B$, the analyst draws a sample of size $n$ with replacement from the original data, applies the full estimation procedure (including nuisance function estimation), and records the resulting causal estimate $\hat{\psi}^{(b)}$. The bootstrap standard error is the standard deviation of $\hat{\psi}^{(b)}$ across replicates.

The bootstrap serves several diagnostic purposes in the context of causal inference.

**Variance estimation validation.** The bootstrap standard error can be compared to the influence-function-based standard error $\hat{\sigma}/\sqrt{n}$. Agreement between the two provides reassurance that the influence-function-based inference is reliable. Disagreement---particularly when the bootstrap standard error is substantially larger than the influence-function-based standard error---indicates that the asymptotic variance is underestimated, which may occur when the influence curve distribution is heavy-tailed, when the sample size is small, or when the nuisance functions are estimated with high variability.

**Normality assessment.** The distribution of bootstrap estimates $\{\hat{\psi}^{(b)}\}_{b=1}^{B}$ should be approximately normal if the asymptotic theory is a good guide to finite-sample behavior. A histogram or Q-Q plot of the bootstrap estimates can reveal skewness, bimodality, or heavy tails that would invalidate Wald-type confidence intervals.

**Confidence interval comparison.** Bootstrap percentile intervals (using the 2.5th and 97.5th percentiles of the bootstrap distribution) and BCa intervals (bias-corrected and accelerated) can be compared to Wald-type intervals. When these interval types disagree, the Wald-type interval is likely unreliable, and the analyst should consider using the bootstrap intervals instead or investigating the source of the disagreement.

```r
## Bootstrap diagnostics for TMLE
B_boot <- 500
boot_estimates <- numeric(B_boot)

for (b in 1:B_boot) {
  idx <- sample(1:nrow(dat), nrow(dat), replace = TRUE)
  dat_b <- dat[idx, ]

  fit_b <- tmle::tmle(
    Y = dat_b$Y, A = dat_b$A,
    W = dat_b[, c("W1", "W2", "W3")],
    Q.SL.library = "SL.glm",
    g.SL.library = "SL.glm"
  )
  boot_estimates[b] <- fit_b$estimates$ATE$psi
}

## Compare bootstrap and influence-function-based inference
IC_se <- sqrt(tmle_fit$estimates$ATE$var.psi)
boot_se <- sd(boot_estimates)

cat("== Bootstrap vs Influence Function Inference ==\n")
cat("IF-based SE:", round(IC_se, 4), "\n")
cat("Bootstrap SE:", round(boot_se, 4), "\n")
cat("Ratio (boot/IF):", round(boot_se / IC_se, 3), "\n\n")

## Wald interval
psi_hat <- tmle_fit$estimates$ATE$psi
wald_ci <- psi_hat + c(-1, 1) * 1.96 * IC_se

## Bootstrap percentile interval
boot_pctile_ci <- quantile(boot_estimates, c(0.025, 0.975))

cat("Wald CI:", round(wald_ci, 4), "\n")
cat("Boot percentile CI:", round(boot_pctile_ci, 4), "\n")
cat("Width ratio (boot/Wald):",
    round(diff(boot_pctile_ci) / diff(wald_ci), 3), "\n\n")

## Normality of bootstrap distribution
sw_test <- shapiro.test(boot_estimates)
cat("Shapiro-Wilk test for normality of bootstrap dist:\n")
cat("  W =", round(sw_test$statistic, 4),
    "  p =", round(sw_test$p.value, 4), "\n")
cat("  Skewness:", round(moments::skewness(boot_estimates), 3), "\n")
cat("  Kurtosis:", round(moments::kurtosis(boot_estimates), 3),
    "(normal = 3)\n")
```

An important caveat about bootstrap diagnostics for TMLE and AIPW is that the bootstrap may not be asymptotically valid for these estimators when highly adaptive machine learning methods are used for nuisance estimation. The bootstrap validity requires that the nuisance estimator behaves sufficiently smoothly as a function of the data, which may not hold for tree-based methods or other discontinuous learners. For this reason, the bootstrap is best used as a diagnostic tool (to check the plausibility of influence-function-based inference) rather than as the primary inference method. When the bootstrap and influence function approaches agree, the influence-function-based inference is likely reliable. When they disagree, the analyst should investigate the source of the disagreement rather than automatically defaulting to the bootstrap.

### 5.3 Influence Curve Stability and Normality

The Wald-type confidence interval for TMLE and AIPW relies on the central limit theorem applied to the estimated influence curve: $\sqrt{n}(\hat{\psi} - \psi_0) \approx n^{-1/2}\sum_{i=1}^{n} \hat{D}(O_i) \xrightarrow{d} N(0, \sigma^2)$. This approximation is valid when the influence curve distribution is sufficiently well-behaved---specifically, when it has finite variance and satisfies a Lindeberg-type condition. In practice, the approximation can fail when the influence curve distribution is heavy-tailed, highly skewed, or contains outliers that dominate the sum.

The analyst should assess the distribution of the estimated influence curve using the following diagnostics.

**Q-Q plots.** A normal Q-Q plot of the estimated influence curve values $\{\hat{D}(O_i)\}_{i=1}^{n}$ provides a visual assessment of normality. Systematic departures from the diagonal line indicate non-normality. Heavy tails (S-shaped departures) are the most common and most concerning pattern, as they indicate that a few observations exert disproportionate influence on the estimator.

**Skewness and kurtosis.** The skewness and excess kurtosis of the influence curve distribution provide numerical summaries. For a normal distribution, the skewness is 0 and the kurtosis is 3 (excess kurtosis is 0). Positive skewness indicates that the right tail is heavier than the left; negative skewness indicates the opposite. Excess kurtosis greater than 3 indicates heavy tails relative to the normal distribution. As a rough guideline, $|\text{skewness}| > 1$ or excess kurtosis $> 5$ warrant concern about the normality approximation.

**Formal normality tests.** The Shapiro-Wilk test provides a formal test of normality. However, its interpretation should be tempered by the fact that, in large samples, even trivial departures from normality will produce significant $p$-values. The test is most useful in moderate samples ($n = 100$--$500$) where the power is sufficient to detect meaningful departures but not so great as to reject normality for negligible reasons.

```r
## Influence curve stability diagnostics
IC <- tmle_fit$estimates$ATE$IC
n <- length(IC)

cat("== Influence Curve Distribution ==\n")
cat("Mean:", round(mean(IC), 6), "\n")
cat("SD:", round(sd(IC), 4), "\n")
cat("Skewness:", round(moments::skewness(IC), 3), "\n")
cat("Kurtosis:", round(moments::kurtosis(IC), 3), "(normal = 3)\n")
cat("Min:", round(min(IC), 3), "  Max:", round(max(IC), 3), "\n")

## Q-Q plot
qqnorm(IC, main = "Normal Q-Q Plot of Influence Curve")
qqline(IC, col = "red")

## Shapiro-Wilk test (on subsample if n > 5000)
if (n <= 5000) {
  sw <- shapiro.test(IC)
  cat("\nShapiro-Wilk: W =", round(sw$statistic, 4),
      " p =", round(sw$p.value, 4), "\n")
} else {
  sw <- shapiro.test(sample(IC, 5000))
  cat("\nShapiro-Wilk (subsample of 5000): W =",
      round(sw$statistic, 4),
      " p =", round(sw$p.value, 4), "\n")
}

## Tail contribution
cat("\nTail weight:\n")
for (k in c(2, 3, 4, 5)) {
  n_extreme <- sum(abs(IC) > k * sd(IC))
  expected_normal <- round(n * 2 * pnorm(-k), 1)
  cat(sprintf("  |IC| > %d*SD: %d observed vs %.1f expected under normality\n",
              k, n_extreme, expected_normal))
}
```

When the influence curve distribution is substantially non-normal, the analyst has several options. The bootstrap (Section 5.2) provides an alternative basis for confidence intervals that does not rely on the normal approximation. Small-sample corrections (Section 5.4) can improve the accuracy of Wald-type intervals. In some cases, the non-normality is driven by a small number of extreme observations associated with near-positivity violations, and truncation or estimand redefinition (Section 8) can address the root cause.

### 5.4 Small-Sample Corrections

When the sample size is small relative to the complexity of the estimation problem, several corrections can improve the finite-sample performance of TMLE and AIPW.

**Cross-validated TMLE (CV-TMLE).** Standard TMLE estimates the nuisance functions on the full sample and then applies the targeting step on the same sample. This can produce a subtle form of overfitting: the targeting step may exploit information about the relationship between the nuisance function estimates and the outcome that is specific to the sample at hand, producing an estimate that is overly optimistic. CV-TMLE addresses this by using cross-fitting: the data are split into $V$ folds, nuisance functions are estimated on each training fold, and the targeting step is applied using the validation fold predictions. The final estimate averages across folds.

CV-TMLE is recommended as the default approach when machine learning methods are used for nuisance estimation, because it ensures that the estimator's asymptotic linearity holds under weaker conditions on the nuisance function estimation rates. The diagnostic question is whether CV-TMLE and standard TMLE produce substantially different results. If they do, this indicates that the standard TMLE estimate is affected by overfitting, and the CV-TMLE estimate should be preferred.

```r
## CV-TMLE vs standard TMLE comparison
## Standard TMLE
fit_standard <- tmle::tmle(
  Y = dat$Y, A = dat$A,
  W = dat[, c("W1", "W2", "W3")],
  Q.SL.library = c("SL.glm", "SL.glmnet", "SL.ranger"),
  g.SL.library = c("SL.glm", "SL.glmnet", "SL.ranger")
)

## CV-TMLE (if available via drtmle or tmle3)
if (requireNamespace("drtmle", quietly = TRUE)) {
  fit_cvtmle <- drtmle::drtmle(
    Y = dat$Y, A = dat$A,
    W = dat[, c("W1", "W2", "W3")],
    SL_Q = c("SL.glm", "SL.glmnet", "SL.ranger"),
    SL_g = c("SL.glm", "SL.glmnet", "SL.ranger"),
    cvFolds = 10
  )

  cat("== CV-TMLE vs Standard TMLE ==\n")
  cat("Standard TMLE:", round(fit_standard$estimates$ATE$psi, 4), "\n")
  cat("CV-TMLE:", round(fit_cvtmle$drtmle$ATE$est, 4), "\n")
  cat("Difference:", round(fit_standard$estimates$ATE$psi -
                            fit_cvtmle$drtmle$ATE$est, 4), "\n")
}
```

**Finite-sample variance corrections.** The standard influence-function-based variance estimator $\hat{\sigma}^2/n$ can be anticonservative in small samples because it does not account for the variability in the nuisance function estimates. Several corrections have been proposed. The most straightforward is the degrees-of-freedom correction $\hat{\sigma}^2 / (n - p)$, where $p$ is the number of parameters estimated in the nuisance functions, but this is difficult to define precisely for nonparametric estimators. A more principled approach is the sandwich variance estimator that accounts for the estimation of $g$ and $\bar{Q}$, as implemented in the `drtmle` package.

**Higher-order TMLE.** Higher-order TMLE extends the standard TMLE by solving not only the first-order EIF equation but also higher-order estimating equations that provide additional bias correction. This can improve coverage in moderate samples where the first-order bias correction is insufficient. The practical implementation of higher-order TMLE is more complex than standard TMLE and is currently available only in specialized software.

The diagnostic question for all small-sample corrections is whether they make a meaningful difference. If the corrected and uncorrected estimates agree, the sample size is likely adequate for the standard approach. If they disagree, the corrected estimate should be preferred, and the analyst should report the disagreement as evidence that the sample size is marginal for the intended analysis.

### 5.5 Comparative Diagnostics Across Estimators

Running multiple estimators on the same data provides an informal but valuable robustness diagnostic. The logic is straightforward: if TMLE, AIPW, IPTW, and g-computation all produce similar estimates, the analyst can be more confident that the result is not an artifact of a particular estimation strategy. If the estimates diverge, the pattern of divergence can reveal which aspects of the estimation are driving the differences.

```r
## Comparative estimator diagnostics
cat("== Multi-Estimator Comparison ==\n\n")

## 1. TMLE
fit_tmle <- tmle::tmle(
  Y = dat$Y, A = dat$A, W = dat[, c("W1", "W2", "W3")],
  Q.SL.library = c("SL.glm", "SL.ranger"),
  g.SL.library = c("SL.glm", "SL.ranger")
)
cat("TMLE:       ATE =", round(fit_tmle$estimates$ATE$psi, 4),
    " SE =", round(sqrt(fit_tmle$estimates$ATE$var.psi), 4),
    " 95% CI:", round(fit_tmle$estimates$ATE$CI, 4), "\n")

## 2. AIPW (one-step)
Q1 <- predict(Q_model, newdata = transform(dat, A = 1))
Q0 <- predict(Q_model, newdata = transform(dat, A = 0))
ps <- dat$ps
eif_vals <- dat$A/ps * (dat$Y - Q1) - (1-dat$A)/(1-ps) * (dat$Y - Q0) +
            Q1 - Q0
psi_aipw <- mean(eif_vals)
se_aipw  <- sd(eif_vals) / sqrt(nrow(dat))
cat("AIPW:       ATE =", round(psi_aipw, 4),
    " SE =", round(se_aipw, 4),
    " 95% CI:", round(psi_aipw + c(-1,1)*1.96*se_aipw, 4), "\n")

## 3. IPTW (stabilized)
sw <- ifelse(dat$A == 1,
             mean(dat$A) / ps,
             (1 - mean(dat$A)) / (1 - ps))
psi_iptw <- weighted.mean(dat$Y[dat$A==1], sw[dat$A==1]) -
            weighted.mean(dat$Y[dat$A==0], sw[dat$A==0])
cat("IPTW (stab): ATE =", round(psi_iptw, 4), "\n")

## 4. G-computation (parametric)
Q_param <- glm(Y ~ A * (W1 + W2 + W3), data = dat)
Q1_param <- predict(Q_param, newdata = transform(dat, A = 1))
Q0_param <- predict(Q_param, newdata = transform(dat, A = 0))
psi_gcomp <- mean(Q1_param - Q0_param)
cat("G-comp:     ATE =", round(psi_gcomp, 4), "\n")

## Summary
estimates <- c(fit_tmle$estimates$ATE$psi, psi_aipw, psi_iptw, psi_gcomp)
cat("\nRange of estimates:", round(diff(range(estimates)), 4), "\n")
cat("Max pairwise difference:", round(max(dist(estimates)), 4), "\n")
```

The pattern of agreement and disagreement among estimators is diagnostically informative in specific ways.

**TMLE and AIPW agree but IPTW disagrees.** This pattern suggests that the outcome model is contributing meaningful information. TMLE and AIPW both use the outcome model (through the augmentation term), which partially compensates for inaccuracies in the propensity score model. IPTW relies entirely on the propensity score model and is therefore more sensitive to propensity score misspecification and positivity violations.

**All doubly robust estimators agree but differ from g-computation.** This pattern suggests that the outcome model is misspecified and the propensity score model is providing a useful correction. G-computation relies entirely on the outcome model, while the doubly robust estimators correct for outcome model bias using the propensity score.

**All estimators agree.** This provides the strongest evidence (among the diagnostics available from a single dataset) that the result is robust to the choice of estimation strategy. However, agreement does not prove correctness: if all estimators are biased in the same direction (due to unmeasured confounding, for example), they will agree on a biased answer.

**All estimators disagree.** This is the most concerning pattern. It indicates that the data do not strongly identify the causal parameter---small changes in the estimation approach lead to large changes in the estimate. The analyst should investigate the source of the disagreement (positivity violations, outcome model instability, small sample size) and consider whether the intended estimand is credibly supported by the data.

::: {.callout-important}
## When Asymptotic Normality Is Unreliable

The Wald-type confidence interval $\hat{\psi} \pm z_{1-\alpha/2} \hat{\sigma}/\sqrt{n}$ relies on the approximation $\sqrt{n}(\hat{\psi} - \psi_0) \approx N(0, \sigma^2)$. This approximation can fail in several identifiable circumstances, each of which has diagnostic indicators:

**Heavy-tailed influence curves.** When the kurtosis of the estimated influence curve exceeds 5--6, the normal approximation converges slowly, and the Wald interval may have poor coverage. Diagnostic: kurtosis of the influence curve, Q-Q plots, tail weight analysis.

**Near-positivity violations.** When propensity scores are near the boundaries, the influence curve has a heavy right tail (for observations with large $1/g$ terms), and the sum $n^{-1/2}\sum_i D(O_i)$ may not be well-approximated by a normal distribution even for moderate $n$. Diagnostic: maximum $|H|$ or $|D|$, effective sample size, proportion of observations with extreme weights.

**Small effective sample size.** When the effective sample size (Section 2.2.2) is much smaller than the nominal sample size, the central limit theorem is effectively applied to a smaller sample, requiring more observations for the approximation to be accurate. Diagnostic: ESS ratio, comparison of IF-based and bootstrap standard errors.

**Rare outcomes.** For binary outcomes with event rates below 5\%, the influence curve distribution is highly skewed (because most observations contribute small positive values and a few contribute large negative values, or vice versa). Diagnostic: event rate, skewness of the influence curve.

**Multiple testing.** When the analyst estimates effects for many subgroups or time points, the normal approximation errors accumulate. Diagnostic: number of comparisons, agreement between nominal and empirical coverage in simulation.

When any of these indicators suggest that asymptotic normality is unreliable, the analyst should consider alternative inference approaches: bootstrap confidence intervals, permutation tests, small-sample TMLE corrections, or---most importantly---honest acknowledgment that the confidence intervals may have incorrect coverage.
:::

---

## 6. Longitudinal and Dynamic Regime Diagnostics {#sec-longitudinal-outline}

This section will extend the diagnostic framework to longitudinal settings with time-varying treatment.

### 6.1 Time-Specific Positivity Diagnostics

- Assessing positivity at each time point separately
- Identifying time points where treatment probabilities are most extreme
- Time-specific propensity score distributions and clever covariate magnitudes

### 6.2 Cumulative Product Diagnostics

- The cumulative treatment mechanism $\prod_{t=0}^{T} g_t(A_t \mid \bar{A}_{t-1}, \bar{L}_t)$ for IPTW in longitudinal settings
- Diagnostics for the distribution of cumulative weights: mean, variance, maximum, effective sample size
- The multiplicative nature of cumulative weights means that moderate per-period violations produce extreme cumulative weights

### 6.3 Stabilized Weight Diagnostics

- Stabilized weights $\text{sw}_i = \prod_t P(A_t \mid \bar{A}_{t-1}) / \prod_t P(A_t \mid \bar{A}_{t-1}, \bar{L}_t)$ and their diagnostic properties
- The mean of stabilized weights should be approximately 1; departures indicate model misspecification
- Comparing stabilized and unstabilized weight distributions

### 6.4 Sequential Double Robustness in LTMLE

- How LTMLE achieves sequential double robustness and what diagnostics are relevant at each time step
- Time-specific targeting step diagnostics
- Cumulative EIF diagnostics for LTMLE

### 6.5 Diagnostics for Modified Treatment Policies and Stochastic Interventions

- Support diagnostics for MTPs: verifying that the shifted treatment $d(A_t, W_t)$ lies within the support of the natural treatment mechanism
- Density ratio estimation diagnostics for continuous treatments under the `lmtp` framework
- Positivity under continuous exposures and the role of density ratios rather than inverse probabilities

---

## 7. Simulation-Based Design Diagnostics {#sec-simulation-outline}

This section will develop a formal framework for diagnostics conducted before outcome data are accessed.

### 7.1 Outcome-Blind Simulation

- Using only the treatment and covariate data (no outcomes) to simulate plausible outcome-generating processes
- Assessing estimator performance under these simulated DGPs
- This approach permits design diagnostics without risking outcome-driven analysis decisions

### 7.2 Bias Under Plausible Data-Generating Processes

- Defining a range of plausible DGPs that span the analyst's uncertainty about the true data-generating process
- Computing bias, variance, and coverage of the intended estimator under each DGP
- Identifying conditions under which the estimator is likely to fail

### 7.3 Support Stress Testing

- Systematically varying the strength of positivity violations in simulated data to identify the threshold at which estimator performance degrades
- Relating these thresholds to the observed propensity score distribution in the real data

### 7.4 Sample Size Diagnostics Under TMLE

- Power and sample size calculations that account for the specific properties of TMLE (e.g., the efficiency bound, the impact of nuisance estimation on variance)
- Comparison with naive sample size calculations that assume parametric models

### 7.5 Regulatory Relevance of Design Diagnostics

- How simulation-based design diagnostics align with regulatory expectations for real-world evidence studies
- Pre-registration of design diagnostics as part of a statistical analysis plan

---

## 8. Estimand Adaptation When Diagnostics Fail {#sec-adaptation-outline}

This section will provide a structured decision framework for modifying the analysis when diagnostics reveal that the intended estimand is not credibly supported by the data.

### 8.1 Changing the Estimand Versus Changing the Estimator

- A fundamental distinction: if positivity fails, switching from IPTW to TMLE does not solve the problem---the estimand is the same, and both estimators target the same (unsupported) quantity
- Genuine resolution requires changing the estimand to one that the data can support

### 8.2 Decision Tree for Estimand Adaptation

- **Restrict the target population**: limit to the overlap region, defining a new target population for which positivity holds
- **Redefine the exposure**: coarsen the treatment variable, combine treatment levels, or restrict to a subset of treatment contrasts
- **Switch to stochastic interventions**: replace "set $A = 1$" with "shift the probability of $A = 1$ by $\delta$"
- **Target conditional effects**: instead of the marginal ATE, target the ATE conditional on a subset of covariates (CATE) for which support is adequate
- **Shift from static to modified treatment policies**: define the intervention as a shift from the natural value of treatment, which respects the observed support by construction

### 8.3 How Diagnostics Guide Adaptation

- Mapping specific diagnostic findings to specific adaptation strategies
- Documenting the rationale for estimand changes in the analysis report
- Maintaining the pre-specified nature of the analysis through planned adaptation rules

---

## 9. Sensitivity Analysis Frameworks {#sec-sensitivity-outline}

This section will provide a comprehensive treatment of sensitivity analyses beyond those covered in the exchangeability subsection.

### 9.1 Bias Function Sensitivity Analysis

- Formalizing the bias as a function of the departure from identification assumptions
- Computing bias-corrected estimates under a range of assumed departures

### 9.2 Partial Identification and Bounding Approaches

- When point identification fails, bounding the causal effect under minimal assumptions
- Manski bounds and their extensions
- The tradeoff between bound width and assumption strength

### 9.3 Selection Bias Modeling

- Sensitivity to selection into the study (as distinct from confounding)
- Inverse probability of selection weighting and its diagnostics

### 9.4 Sensitivity to Weight Truncation

- Systematic analysis of how truncation level affects point estimates and inference
- Reporting truncation sensitivity as a standard component of the analysis

### 9.5 Sensitivity to Learner Library Choice

- How the causal estimate depends on the composition of the Super Learner library
- Reporting estimates under multiple library specifications

### 9.6 Separating Statistical and Causal Uncertainty

- Statistical uncertainty (captured by confidence intervals) reflects sampling variability
- Causal uncertainty (captured by sensitivity analyses) reflects uncertainty about the identification assumptions
- Both should be reported and distinguished in the analysis

---

## 10. Reporting Recommendations {#sec-reporting-outline}

This section will provide a structured set of reporting guidelines for diagnostics in causal inference publications and regulatory submissions.

### 10.1 Required Diagnostics for Publication

- Minimum diagnostic reporting requirements: propensity score distributions, covariate balance tables, weight summaries, influence curve summaries, truncation sensitivity, and at least one sensitivity analysis for unmeasured confounding

### 10.2 Essential Plots and Tables

- Propensity score density plots by treatment group
- Covariate balance (Love) plots showing SMDs before and after weighting
- Weight distribution histograms
- Truncation sensitivity plots (estimate and CI versus truncation level)
- Calibration plots for outcome model
- Super Learner coefficient tables
- Influence curve distribution summaries

### 10.3 Documenting Estimator Tuning

- Reporting the Super Learner library, cross-validation scheme, and truncation parameters
- Documenting any deviations from the pre-specified analysis plan and their justification

### 10.4 Presenting Sensitivity Analyses

- E-values with interpretation
- Quantitative bias analysis results with parameter ranges
- Negative control results
- Truncation sensitivity
- Library sensitivity

### 10.5 Communicating Limitations

- Distinguishing between statistical precision and causal credibility
- Acknowledging specific threats to identification assumptions identified through diagnostics
- Avoiding both overstatement ("we proved causation") and understatement ("this is merely observational") of causal claims
- Framing conclusions in terms of the strength of evidence, given the diagnostic findings

---

## Sources and Further Reading

- VanderWeele TJ, Ding P (2017). Sensitivity analysis in observational research: introducing the E-value. *Annals of Internal Medicine* 167(4):268--274.
- Lipsitch M, Tchetgen Tchetgen E, Cohen T (2010). Negative controls: a tool for detecting confounding and bias in observational studies. *Epidemiology* 21(3):383--388.
- Petersen ML, Porter KE, Gruber S, et al. (2012). Diagnosing and responding to violations in the positivity assumption. *Statistical Methods in Medical Research* 21(1):31--54.
- Gruber S, van der Laan MJ (2010). An application of collaborative targeted maximum likelihood estimation in causal inference and genomics. *International Journal of Biostatistics* 6(1):Article 18.
- van der Laan MJ, Rose S (2011). *Targeted Learning: Causal Inference for Observational and Experimental Data*. Springer.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapters 11--12. [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- Diaz I, Williams N, Hoffman KL, et al. (2023). Nonparametric causal effects based on longitudinal modified treatment policies. *Journal of the American Statistical Association* 118(542):846--857.
- van der Laan MJ, Gruber S (2012). Targeted minimum loss based estimation of causal effects of multiple time point interventions. *International Journal of Biostatistics* 8(1):Article 9.
- Tchetgen Tchetgen E (2014). The control outcome calibration approach for causal inference with unobserved confounding. *American Journal of Epidemiology* 179(5):633--640.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data: integrating causal modeling and statistical estimation. *Epidemiology* 25(3):418--426.
- Ding P, VanderWeele TJ (2016). Sensitivity analysis without assumptions. *Epidemiology* 27(3):368--377.
- `EValue` R package: [CRAN](https://cran.r-project.org/package=EValue)
- `drtmle` R package: [CRAN](https://cran.r-project.org/package=drtmle)
- `lmtp` R package: [CRAN](https://cran.r-project.org/package=lmtp)
- `episensr` R package: [CRAN](https://cran.r-project.org/package=episensr)
- `sl3` R package: [GitHub](https://github.com/tlverse/sl3)

---

## Software Implementation (R)

This example demonstrates key diagnostic checks for a TMLE analysis: positivity assessment via propensity score overlap and clever covariate summaries, effective sample size computation, truncation sensitivity, influence curve inspection, and E-value computation.

```{r}
#| eval: false
set.seed(42)
n <- 1000

## ── Simulate data with a near-positivity violation ──
W1 <- rnorm(n)
W2 <- rbinom(n, 1, 0.3)
W3 <- rnorm(n, mean = 0.5, sd = 1.2)
## W2 = 1 strongly predicts treatment, creating near-positivity violation
A  <- rbinom(n, 1, plogis(-0.5 + 0.4 * W1 + 2.5 * W2 + 0.3 * W3))
Y  <- 0.5 * A + 0.8 * W1 - 0.3 * W2 + 0.2 * W3 + rnorm(n, sd = 0.5)

W <- data.frame(W1 = W1, W2 = W2, W3 = W3)
dat <- data.frame(W, A = A, Y = Y)

## ── 1. Propensity score overlap ──
ps_mod <- glm(A ~ W1 + W2 + W3, data = dat, family = binomial)
dat$ps <- predict(ps_mod, type = "response")

cat("== Propensity Score Summary ==\n")
print(summary(dat$ps))
cat("\nObservations with ps < 0.05:", sum(dat$ps < 0.05), "\n")
cat("Observations with ps > 0.95:", sum(dat$ps > 0.95), "\n")

## ── 2. Effective sample size ──
w_treated <- 1 / dat$ps[dat$A == 1]
w_control <- 1 / (1 - dat$ps[dat$A == 0])
ess_t <- sum(w_treated)^2 / sum(w_treated^2)
ess_c <- sum(w_control)^2 / sum(w_control^2)
cat("\n== Effective Sample Size ==\n")
cat("Treated:  nominal =", sum(dat$A == 1),
    " effective =", round(ess_t, 1),
    " ratio =", round(ess_t / sum(dat$A == 1), 3), "\n")
cat("Control:  nominal =", sum(dat$A == 0),
    " effective =", round(ess_c, 1),
    " ratio =", round(ess_c / sum(dat$A == 0), 3), "\n")

## ── 3. Clever covariate diagnostics ──
H <- dat$A / dat$ps - (1 - dat$A) / (1 - dat$ps)
cat("\n== Clever Covariate ==\n")
cat("Range:", round(range(H), 2), "\n")
cat("SD:", round(sd(H), 2), "\n")
cat("|H| > 20:", sum(abs(H) > 20), "obs\n")

## ── 4. TMLE with diagnostics ──
if (requireNamespace("tmle", quietly = TRUE)) {
  library(tmle)
  fit <- tmle(Y = Y, A = A, W = W,
              Q.SL.library = "SL.glm",
              g.SL.library = "SL.glm")

  cat("\n== TMLE Results ==\n")
  cat("ATE:", round(fit$estimates$ATE$psi, 4), "\n")
  cat("SE:", round(sqrt(fit$estimates$ATE$var.psi), 4), "\n")
  cat("95% CI:", round(fit$estimates$ATE$CI, 4), "\n")

  ## Influence curve diagnostics
  IC <- fit$estimates$ATE$IC
  cat("\n== Influence Curve ==\n")
  cat("Mean:", round(mean(IC), 6), "(should be near 0)\n")
  cat("SD:", round(sd(IC), 4), "\n")
  cat("Max |IC|:", round(max(abs(IC)), 2), "\n")
  cat("|IC| > 3*SD:", sum(abs(IC) > 3 * sd(IC)), "obs\n")

  ## ── 5. Truncation sensitivity ──
  cat("\n== Truncation Sensitivity ==\n")
  for (gb in c(0.005, 0.01, 0.025, 0.05, 0.10)) {
    fit_tr <- tmle(Y = Y, A = A, W = W,
                   Q.SL.library = "SL.glm",
                   g.SL.library = "SL.glm",
                   gbound = gb)
    cat(sprintf("  gbound = %.3f  ATE = %.4f  SE = %.4f\n",
                gb, fit_tr$estimates$ATE$psi,
                sqrt(fit_tr$estimates$ATE$var.psi)))
  }
} else {
  message("Install the 'tmle' package:  install.packages('tmle')")
}

## ── 6. E-value ──
RR <- 1.8
E_val <- RR + sqrt(RR * (RR - 1))
cat("\n== E-value ==\n")
cat("For RR =", RR, " E-value =", round(E_val, 2), "\n")
cat("An unmeasured confounder would need RR >=", round(E_val, 2),
    "\nwith both A and Y to explain away the observed effect.\n")
```

