---
title: "Chapter 3.2: Positivity, Overlap, and Data Support"
subtitle: "When your data cannot answer the causal question you are asking"
format: html
---

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4
)
```

# Chapter 3.2: Positivity, Overlap, and Data Support

## A cautionary tale from pharmacoepidemiology

A research team set out to compare two anticoagulants for stroke prevention in older
adults with atrial fibrillation. They carefully assembled a claims database, adjusted
for dozens of confounders, and ran an IPTW analysis. The result: a risk ratio of 0.12,
implying the newer drug cut stroke risk by 88%. Clinicians were skeptical, and rightly
so. When the analysts inspected the propensity score distribution they discovered the
problem. Patients over age 85, those with prior falls, and those on multiple
interacting medications were essentially *never* prescribed the newer drug. Yet the
IPTW estimator was trying to estimate what would happen if *everyone* (including
those patients) received it. A handful of patients in the treated group with near-zero
propensity scores received enormous weights, and those weights drove the entire
estimate.

The culprit was a **positivity violation**: the data contained patient profiles for
which one treatment was never (or almost never) observed. No amount of clever
modeling can conjure information that the data simply do not contain.

We touched on positivity briefly in @sec-identification (Chapter 1.2,
`identification-estimands.qmd`) and saw weight instability in @sec-iptw (Chapter 2.2,
`iptw.qmd`). But positivity is too important, and too commonly the root cause of
failed causal analyses, to scatter across chapters. This chapter gives it the dedicated
treatment it deserves.

::: {.callout-note}
## Learning objectives

After completing this chapter you will be able to:

1. State the positivity assumption formally and intuitively.
2. Distinguish structural from practical positivity violations.
3. Diagnose positivity problems using propensity score plots, overlap statistics, and effective sample size.
4. Explain how positivity violations distort IPTW, G-computation, and TMLE.
5. Apply practical remedies including population restriction, weight truncation, and modified treatment policies.
:::

---

# 1. What Positivity Means {#sec-positivity-defn}

## Intuition first

Imagine you want to estimate the causal effect of a drug versus no drug across all
patients in your study. Positivity says: **for every type of patient defined by the
confounders you adjust for, both treatment options must be possible.** If some patient
profiles always receive the drug (or never receive it), you have no observed
counterfactual for that stratum, and the causal contrast there is unidentifiable from
data alone.

## Formal definition

For every level of confounders $w$ that occurs in the population (that is,
$P(W = w) > 0$), we require:

$$
0 < P(A = a \mid W = w) < 1 \quad \text{for all treatment levels } a
$$
{#eq-positivity}

In words: the probability of receiving each treatment value, conditional on covariates,
is bounded away from zero and one.

## Connection to identification

Recall the G-computation identification formula from Chapter 1.2:

$$
\psi = E_W\!\Big[\,E[Y \mid A=1, W] - E[Y \mid A=0, W]\,\Big]
$$

This formula requires us to estimate $E[Y \mid A=a, W=w]$ for *both* values of $a$
at every $w$ in the population. If $P(A=1 \mid W=w) = 0$ for some $w$, we have no
treated patients at that covariate value and must extrapolate from other regions. That
extrapolation is unverifiable; we are guessing.

::: {.callout-tip}
## A useful analogy

Positivity is like having a restaurant review for every item on the menu. If nobody
has ever ordered the fish, you cannot credibly claim it is better or worse than the
chicken. You can speculate, but the data are silent on that comparison.
:::

---

# 2. Structural vs. Practical Positivity Violations

Not all positivity problems are alike. The distinction matters because the appropriate
response differs.

## Structural violations

A **structural violation** occurs when certain patients *cannot* receive a treatment by
definition or by design.

| Example | Why structural |
|---------|---------------|
| Comparing hysterectomy vs. watchful waiting: men cannot receive hysterectomy | Biological impossibility |
| Pediatric drug not approved for adults | Regulatory prohibition |
| Dialysis only available at certain facilities | Geographic/logistic constraint |

Because the treatment is truly impossible, you should **redefine the target population**
to exclude those strata. Trying to estimate a causal effect for patients who could
never be treated is asking a question the world cannot answer, not just a question the
data cannot answer.

## Practical violations

A **practical (or random) violation** occurs when treatment is *theoretically* possible
but so rare that the data contain few or no treated (or untreated) patients in certain
strata.

| Example | Why practical |
|---------|--------------|
| Very elderly patients rarely prescribed a newer drug | Clinical preference, not prohibition |
| Patients with severe liver disease rarely enrolled in trials | Selection bias, not impossibility |
| Patients at small rural sites rarely receive complex interventions | Resource constraints |

Practical violations are more insidious because they can be hard to detect. The
treatment is not forbidden; it just almost never happens. Analysts may not realize they
are forcing the estimator to extrapolate.

::: {.callout-warning}
## Adjusting for more confounders can make positivity worse

Adding covariates reduces confounding bias (good) but splits the data into finer
strata, each with fewer patients (bad for positivity). There is a genuine tension
between exchangeability and positivity. Targeted Learning methods help navigate this
trade-off, but they cannot eliminate it.
:::

---

# 3. How to Diagnose Positivity Problems {#sec-diagnose}

Let us build a simulated dataset with a deliberate positivity problem, then walk
through the diagnostic toolkit.

## 3.1 Simulate data with a positivity gap

```{r}
library(tidyverse)

set.seed(42)
n <- 2000

# Generate confounders
age <- rnorm(n, mean = 60, sd = 12)
severity <- rbinom(n, 1, prob = 0.4)

# Treatment assignment: strongly depends on age and severity
# Patients with age > 75 AND severity == 1 almost never get treated
logit_ps <- 0.5 - 0.06 * age - 1.8 * severity - 0.04 * (age > 75) * severity * 10
ps_true <- plogis(logit_ps)
A <- rbinom(n, 1, prob = ps_true)

# Outcome
Y <- 2 + 0.03 * age + 0.5 * severity - 1.5 * A + 0.02 * A * age + rnorm(n)

dat <- tibble(age, severity, A, Y, ps_true)

# Examine the positivity gap
dat |>
  mutate(age_group = case_when(
    age <= 55 ~ "<=55",
    age <= 65 ~ "56-65",
    age <= 75 ~ "66-75",
    TRUE       ~ ">75"
  )) |>
  group_by(age_group, severity) |>
  summarise(
    n = n(),
    n_treated = sum(A),
    prop_treated = mean(A),
    .groups = "drop"
  ) |>
  arrange(age_group, severity) |>
  knitr::kable(digits = 3, caption = "Treatment prevalence by age group and severity")
```

Look at the older, high-severity stratum. Treatment is rare there, sometimes
vanishingly so. That is our positivity problem.

## 3.2 Propensity score density plots

The single most useful positivity diagnostic is a density plot of the estimated
propensity score, separated by treatment group.

```{r}
# Fit a propensity score model
ps_model <- glm(A ~ age * severity + I(age^2), data = dat, family = binomial)
dat$ps_hat <- predict(ps_model, type = "response")

ggplot(dat, aes(x = ps_hat, fill = factor(A))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(
    values = c("0" = "#E69F00", "1" = "#56B4E9"),
    labels = c("Untreated", "Treated")
  ) +
  labs(
    x = "Estimated propensity score",
    y = "Density",
    fill = "Group",
    title = "Propensity score overlap plot",
    subtitle = "Poor overlap visible in the left tail (treated patients with very low PS)"
  ) +
  theme_minimal()
```

::: {.callout-tip}
## What to look for

Good overlap: the two densities sit on top of each other across most of the
propensity score range. Bad overlap: one group has density where the other does not.
The non-overlapping region is where positivity is violated.
:::

## 3.3 Numerical overlap summaries

```{r}
dat |>
  group_by(A) |>
  summarise(
    min_ps  = min(ps_hat),
    q05_ps  = quantile(ps_hat, 0.05),
    median  = median(ps_hat),
    q95_ps  = quantile(ps_hat, 0.95),
    max_ps  = max(ps_hat),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 4, caption = "Propensity score distribution by treatment group")
```

## 3.4 Identifying problematic covariate strata

Beyond the propensity score, inspect the raw covariate-treatment cross-tabulation.
Look for cells with zero or near-zero counts.

```{r}
dat |>
  mutate(
    age_cat = cut(age, breaks = c(-Inf, 55, 65, 75, Inf),
                  labels = c("<=55", "56-65", "66-75", ">75"))
  ) |>
  count(age_cat, severity, A) |>
  pivot_wider(names_from = A, values_from = n, values_fill = 0) |>
  rename(untreated = `0`, treated = `1`) |>
  mutate(pct_treated = treated / (treated + untreated)) |>
  knitr::kable(digits = 3,
               caption = "Covariate strata by treatment: look for near-empty cells")
```

---

# 4. How Positivity Problems Affect Different Estimators {#sec-estimator-effects}

Using our simulated data, let us see how each estimator behaves when positivity is
violated. The true ATE in the data-generating process depends on the covariate
distribution, so we will focus on *stability* and *plausibility* of estimates rather
than a single truth.

## 4.1 IPTW: extreme weights

IPTW inverts the propensity score, so patients with $P(A=1 \mid W) \approx 0$ who
nonetheless received treatment get enormous weights.

```{r}
# Compute IPTW weights
dat <- dat |>
  mutate(
    ipw = if_else(A == 1, 1 / ps_hat, 1 / (1 - ps_hat))
  )

# Inspect weight distribution
dat |>
  summarise(
    mean_wt   = mean(ipw),
    max_wt    = max(ipw),
    pct_gt_10 = mean(ipw > 10) * 100,
    pct_gt_20 = mean(ipw > 20) * 100
  ) |>
  knitr::kable(digits = 2, caption = "IPW weight summary: large weights signal trouble")

# IPTW estimate (Horvitz-Thompson style)
ate_iptw <- dat |>
  summarise(
    ate = weighted.mean(Y[A == 1], ipw[A == 1]) -
          weighted.mean(Y[A == 0], ipw[A == 0])
  ) |>
  pull(ate)

cat("IPTW ATE estimate:", round(ate_iptw, 3), "\n")
```

::: {.callout-important}
## Why extreme weights are dangerous

A single patient with a weight of 50 counts as much as 50 patients. If that patient
has an unusual outcome (by chance), it can shift the entire ATE estimate. This is not
bias in the usual sense; it is **finite-sample instability** driven by near-violations
of positivity.
:::

## 4.2 G-computation: silent extrapolation

G-computation fits an outcome model $E[Y \mid A, W]$ and then predicts under each
treatment value for every patient. When positivity is violated, those predictions
require the outcome model to extrapolate into covariate regions with no (or very few)
treated patients.

```{r}
out_model <- lm(Y ~ A * age * severity, data = dat)

dat <- dat |>
  mutate(
    Y1_hat = predict(out_model, newdata = mutate(dat, A = 1)),
    Y0_hat = predict(out_model, newdata = mutate(dat, A = 0))
  )

ate_gcomp <- mean(dat$Y1_hat - dat$Y0_hat)
cat("G-computation ATE estimate:", round(ate_gcomp, 3), "\n")
```

G-computation may look stable, but that stability is an illusion. It relies entirely on
the parametric model being correct in the extrapolation region. There is no data to
check it.

## 4.3 TMLE: clever covariate instability

TMLE (Chapter 2.3, `doubly-robust-tmle.qmd`) updates the initial outcome model
using a "clever covariate" $H(A, W) = A / g(W) - (1-A) / (1 - g(W))$. When $g(W)$
is near zero, this clever covariate explodes, just like the IPW weight. So TMLE
inherits the instability of IPTW through its targeting step, even though it also uses
an outcome model.

```{r}
# Demonstrate clever covariate magnitude
dat <- dat |>
  mutate(
    clever_cov = A / ps_hat - (1 - A) / (1 - ps_hat)
  )

dat |>
  summarise(
    max_abs_H  = max(abs(clever_cov)),
    q99_abs_H  = quantile(abs(clever_cov), 0.99),
    mean_abs_H = mean(abs(clever_cov))
  ) |>
  knitr::kable(digits = 2,
               caption = "Clever covariate magnitude: large values indicate positivity trouble")
```

::: {.callout-note}
## The bottom line

No estimator is immune to positivity violations. IPTW shows the problem as weight
instability. G-computation hides it behind model extrapolation. TMLE sits in between,
inheriting instability through the clever covariate while gaining some protection from
the outcome model. The real solution is not to choose a "better" estimator; it is to
address the positivity problem directly.
:::

---

# 5. Practical Responses to Positivity Violations {#sec-responses}

## 5.1 Redefine the target population

The most principled response is to restrict the analysis to the population where
overlap exists. You are changing the *question*, not patching the estimator.

```{r}
# Restrict to the overlap region (e.g., PS between 0.05 and 0.95)
dat_trimmed <- dat |>
  filter(ps_hat > 0.05, ps_hat < 0.95)

cat("Original sample:", nrow(dat), "\n")
cat("After trimming: ", nrow(dat_trimmed),
    sprintf("(%.0f%% retained)\n", 100 * nrow(dat_trimmed) / nrow(dat)))
```

This is honest. You are now estimating the ATE for the sub-population with adequate
support, not the full population. Report this clearly.

## 5.2 Weight truncation with sensitivity analysis

If trimming the population is not desirable, you can truncate (cap) the IPW weights at
a threshold. This introduces some bias but reduces variance. The key is to try several
thresholds and see how the estimate changes.

```{r}
truncation_sensitivity <- function(data, thresholds) {
  map_dfr(thresholds, function(thr) {
    d <- data |>
      mutate(ipw_trunc = pmin(ipw, thr))
    ate <- d |>
      summarise(
        ate = weighted.mean(Y[A == 1], ipw_trunc[A == 1]) -
              weighted.mean(Y[A == 0], ipw_trunc[A == 0])
      ) |>
      pull(ate)
    tibble(threshold = thr, ate_estimate = ate, max_weight = thr)
  })
}

thresholds <- c(5, 10, 20, 50, 100, Inf)
results <- truncation_sensitivity(dat, thresholds)

results |>
  mutate(threshold = if_else(is.infinite(threshold), "None", as.character(threshold))) |>
  knitr::kable(digits = 3, caption = "Sensitivity of IPTW to weight truncation")
```

```{r}
# Visualize truncation sensitivity
results_plot <- results |>
  filter(is.finite(threshold))

ggplot(results_plot, aes(x = threshold, y = ate_estimate)) +
  geom_point(size = 3) +
  geom_line() +
  geom_hline(
    yintercept = results$ate_estimate[is.infinite(results$threshold)],
    linetype = "dashed", color = "red"
  ) +
  annotate("text", x = max(results_plot$threshold), y = ate_iptw,
           label = "Untruncated", hjust = 1, vjust = -0.5, color = "red") +
  labs(
    x = "Weight truncation threshold",
    y = "ATE estimate",
    title = "Truncation sensitivity analysis",
    subtitle = "Stable estimates across thresholds suggest robust findings"
  ) +
  theme_minimal()
```

::: {.callout-tip}
## Interpreting the sensitivity plot

If the ATE estimate barely changes as you relax the truncation threshold, the result
is not driven by extreme weights and positivity violations are not dominating. If the
estimate swings wildly, the causal conclusion depends heavily on a few patients with
near-zero propensity scores, and you should be cautious.
:::

## 5.3 Modified treatment policies (shift interventions)

Instead of asking "What if everyone received treatment?", you can ask "What if
everyone's probability of treatment were shifted slightly upward?" This avoids
counterfactual scenarios that violate positivity. We will encounter these in detail in
later chapters; for now, just note the idea: if the static intervention
$\text{do}(A=1)$ creates positivity problems, a *stochastic* or *shift* intervention
may not.

## 5.4 Honest reporting

When positivity problems remain after reasonable adjustments, the most important
response is transparency. Report:

- The proportion of patients in low-overlap regions.
- The maximum and 99th-percentile IPW weights.
- The effective sample size (next section).
- Sensitivity analyses under different truncation levels.
- The sub-populations where the estimate is well supported versus extrapolated.

---

# 6. Effective Sample Size {#sec-ess}

Extreme weights do not just inflate variance; they reduce the amount of information in
your sample. The **effective sample size (ESS)** quantifies how many equally weighted
observations your weighted sample is equivalent to.

## Formula

For a set of weights $w_1, \ldots, w_n$:

$$
\text{ESS} = \frac{\left(\sum_{i=1}^n w_i\right)^2}{\sum_{i=1}^n w_i^2}
$$
{#eq-ess}

If all weights are equal, $\text{ESS} = n$. If one weight dominates,
$\text{ESS} \approx 1$.

## Code

```{r}
compute_ess <- function(weights) {
  sum(weights)^2 / sum(weights^2)
}

# ESS for each treatment group
ess_treated   <- compute_ess(dat$ipw[dat$A == 1])
ess_untreated <- compute_ess(dat$ipw[dat$A == 0])

tibble(
  group        = c("Treated", "Untreated", "Total"),
  n_actual     = c(sum(dat$A == 1), sum(dat$A == 0), nrow(dat)),
  ess          = c(ess_treated, ess_untreated, compute_ess(dat$ipw)),
  pct_retained = round(100 * ess / n_actual, 1)
) |>
  knitr::kable(digits = 1, caption = "Effective sample size by treatment group")
```

## Threshold guidance

There is no universal cutoff, but common recommendations include:

- **ESS < 50% of actual sample size**: warning flag; inspect weights.
- **ESS < 20% of actual sample size**: serious concern; consider trimming or population restriction.
- **ESS < n for one group but not the other**: asymmetric positivity, often indicating that one treatment direction has poor support.

::: {.callout-important}
## ESS is a summary, not a diagnosis

A healthy ESS does not guarantee that positivity holds everywhere. A few problematic
strata can be masked by many well-behaved ones. Always pair ESS with the overlap plots
and covariate-specific tables shown in @sec-diagnose.
:::

---

# 7. Summary {#sec-summary}

::: {.callout-tip}
## Chapter summary

1. **Positivity** requires that every confounder stratum has a non-zero probability
   of each treatment value. Without it, the causal effect is not identified from
   data alone.

2. **Structural violations** (treatment impossible) call for redefining the target
   population. **Practical violations** (treatment very rare) call for diagnostic
   checks and possibly modified estimands.

3. **Diagnosis toolkit**: propensity score overlap plots, covariate-by-treatment
   tables, effective sample size.

4. **All estimators suffer**: IPTW through extreme weights, G-computation through
   hidden extrapolation, TMLE through clever covariate instability.

5. **Practical responses**: restrict to the overlap population, truncate weights with
   sensitivity analysis, consider shift interventions, and report limitations honestly.

6. **Effective sample size** quantifies information loss from extreme weights and
   serves as a quick summary diagnostic.
:::

---

# 8. Check Your Understanding

**Question 1.** A colleague argues: "I use G-computation, so I don't need to worry
about positivity because I never compute weights." Why is this wrong?

<details>
<summary>Show answer</summary>

G-computation still requires estimating $E[Y \mid A=a, W=w]$ for both treatment
values at every covariate pattern. When no treated patients exist in a stratum,
G-computation extrapolates the outcome model into that region. The estimate there
depends entirely on the parametric model, not on data. The analyst may not even
realize this is happening because there are no extreme weights to flag the problem.
Positivity is an *identification* assumption, not an estimation assumption, so it
applies regardless of the estimator used.

</details>

**Question 2.** You fit a propensity score model and find that 95% of your sample has
propensity scores between 0.15 and 0.85, but 5% of treated patients have scores below
0.02. What should you do?

<details>
<summary>Show answer</summary>

First, investigate *who* those 5% are. Check which covariate patterns produce
near-zero propensity scores and ask whether those patients should be in the target
population at all (structural violation) or whether treatment is merely rare there
(practical violation). Then consider: (a) restricting to the overlap region
(PS > 0.02 or stricter), which changes the estimand to the sub-population with
adequate support; (b) truncating weights and running a sensitivity analysis across
several thresholds; (c) reporting the effective sample size with and without those
patients. Whatever you choose, be transparent that the ATE for the full population
is poorly supported by the data.

</details>

**Question 3.** Your effective sample size is 900 out of 1,000 actual patients. Does
this mean positivity holds?

<details>
<summary>Show answer</summary>

Not necessarily. ESS is a global summary. It is possible for a small subgroup to
have severe positivity problems that are diluted by the well-behaved majority. For
example, if 950 patients have moderate weights (around 1) and 50 patients in a
specific stratum have weights of 3 to 5, the overall ESS could still look healthy
while that stratum is poorly supported. Always supplement ESS with overlap plots and
covariate-specific tables to identify localized problems.

</details>

---

# 9. What Comes Next

Now that you understand when your data can and cannot support a causal question, you
are ready to see these ideas in action. Chapter 3.3 (`03-03-longitudinal-case-study.qmd`)
walks through a complete applied example where positivity diagnostics, weight
inspection, and population-restriction decisions arise naturally in a longitudinal
pharmacoepidemiology study. You will see how the toolkit from this chapter integrates
with the estimation methods from Chapters 2.2 and 2.3 in a realistic analysis
workflow.

---

```{r}
sessionInfo()
```
