---
subtitle: "When your data cannot answer the causal question you are asking"
format: html
---

*Estimated reading time: 50 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 7, fig.height = 4
)
```

# Positivity, Overlap, and Data Support {#sec-positivity-overlap}

### A cautionary tale from pharmacoepidemiology

A research team set out to estimate the effect of highly active antiretroviral
therapy (HAART) initiation on mortality
in an HIV cohort. They assembled a claims-derived dataset, adjusted for
dozens of confounders, and ran an inverse probability of treatment weighting
(IPTW) analysis. The result was startling: a risk
ratio of 0.12 for mortality, implying HAART cut the death rate by 88%. Clinicians
welcomed the magnitude, but treatment epidemiologists were skeptical. When the
analysts inspected the propensity score distribution they found the problem.
Patients with very low CD4 counts and recent opportunistic infections were
essentially *always* started on HAART under the guidelines in force at the time.
A small number who never initiated despite meeting those criteria looked
nothing like anyone else in the cohort. Those untreated, very-low-CD4 patients had
near-zero propensity scores for "no HAART," so their inverse-probability weights
were enormous. A handful of them dominated the entire estimate.

The culprit was a **positivity violation**: the data contained patient profiles for
which one treatment was never (or almost never) observed. No amount of clever
modeling can conjure information that the data simply do not contain.

We touched on positivity briefly in @sec-roadmap and saw weight instability in @sec-iptw. But positivity is too often the root cause of a failed causal analysis to leave scattered across chapters. This chapter treats it on its own.

::: {.callout-note}
## Learning objectives

After completing this chapter you will be able to:

1. State the positivity assumption formally and intuitively.
2. Distinguish structural from practical positivity violations.
3. Diagnose positivity problems using propensity score plots, overlap statistics, and effective sample size.
4. Explain how positivity violations distort IPTW, G-computation, and TMLE.
5. Apply practical remedies including population restriction, weight truncation, and modified treatment policies.
:::

---

## What Positivity Means {#sec-positivity-defn}

### Intuition first

Imagine you want to estimate the causal effect of a drug versus no drug across all
patients in your study. Positivity says: **for every type of patient defined by the
confounders you adjust for, both treatment options must be possible.** If some patient
profiles always receive the drug (or never receive it), you have no observed
counterfactual for that stratum, and the causal contrast there is not identifiable from
data alone. You cannot learn the effect of a treatment in patients who would never have received it, any more than a phone poll can capture the views of people who never pick up.

### Formal definition

For every level of confounders $w$ that occurs in the population (that is,
$P(W = w) > 0$), we require [@hernan2025whatif; @petersen2014causal]:

$$
0 < P(A = a \mid W = w) < 1 \quad \text{for all treatment levels } a
$$
{#eq-positivity}

In words: the probability of receiving each treatment value, conditional on covariates,
is bounded away from zero and one.

### Connection to identification

Recall the G-computation identification formula from @sec-roadmap:

$$
\psi = E_W\!\Big[\,E[Y \mid A=1, W] - E[Y \mid A=0, W]\,\Big]
$$

This formula requires us to estimate $E[Y \mid A=a, W=w]$ for *both* values of $a$
at every $w$ in the population. If $P(A=1 \mid W=w) = 0$ for some $w$, we have no
treated patients at that covariate value and must extrapolate from other regions. Nothing
in the data lets us check that extrapolation, so we are guessing.

Positivity is a bit like having a restaurant review for every item on the menu. If nobody has ever ordered the fish, you cannot credibly claim it is better or worse than the chicken. You can speculate, but the data are silent on that comparison.

### Positivity with continuous or high-dimensional covariates

The formal definition above is stated for discrete $W$, where we require $P(A = a \mid W = w) > 0$ at every support point. When $W$ is continuous or high-dimensional, the condition generalizes to positivity **almost surely**: the propensity score $g(W) = P(A = 1 \mid W)$ must be bounded away from 0 and 1 with probability one in the analysis population.

**Practical restatement.** There exist constants $0 < \delta_L \leq \delta_U < 1$ such that $\delta_L \leq g(W) \leq \delta_U$ almost surely. In applied work, this means the estimated propensity scores should be bounded away from the boundaries, not just on average, but for every observation that contributes to the analysis.

When propensity scores approach 0 or 1 for a non-negligible fraction of the sample, the effective sample size shrinks, influence-function-based variance estimates grow, and the causal estimand becomes poorly identified in the tails of the covariate distribution. If elderly patients with renal impairment are almost never prescribed the new drug, the propensity score for that group approaches zero, and estimates for their subgroup rest entirely on model extrapolation.

This is more than a technical annoyance. When positivity fails for a subpopulation, the most principled response is often to **revise the estimand** rather than truncate weights or extrapolate. Modified treatment policies and stochastic interventions (see @sec-adherence-shift) define interventions that stay within the observed data support by construction, so there is no need to estimate treatment effects in regions where the data are sparse.

Alternatively, one can restrict the target population to the subset with adequate overlap, acknowledging that the resulting estimate answers a different (but well-defined) causal question. This often arises in claims-based studies when certain comorbidity profiles are strongly associated with one treatment, leaving little or no overlap for the alternative.

---

## Structural vs. Practical Positivity Violations

Not all positivity problems are alike. The distinction between structural and
practical violations [@westreichcole2010positivity] matters because the appropriate
response differs.

### Structural violations

A **structural violation** occurs when certain patients *cannot* receive a treatment by
definition or by design [@petersen2014causal].

| Example | Why structural |
|---------|---------------|
| Comparing hysterectomy vs. watchful waiting: men cannot receive hysterectomy | Biological impossibility |
| Pediatric drug not approved for adults | Regulatory prohibition |
| Dialysis only available at certain facilities | Geographic/logistic constraint |

Because the treatment is truly impossible, you should **redefine the target population**
to exclude those strata. Estimating a causal effect for patients who could
never be treated asks a question the world cannot answer, not just one the
data cannot answer.

### Practical violations

A **practical (or random) violation** occurs when treatment is *theoretically* possible
but so rare that the data contain few or no treated (or untreated) patients in certain
strata.

| Example | Why practical |
|---------|--------------|
| Very elderly patients rarely prescribed a newer drug | Clinical preference, not prohibition |
| Patients with severe liver disease rarely enrolled in trials | Selection bias, not impossibility |
| Patients at small rural sites rarely receive complex interventions | Resource constraints |

Practical violations are more insidious because they can be hard to detect. The
treatment is not forbidden; it just almost never happens. An analyst may not realize the
estimator is being forced to extrapolate.

::: {.callout-warning}
## Adjusting for more confounders can make positivity worse

Adding covariates reduces confounding bias (good) but splits the data into finer
strata, each with fewer patients (bad for positivity). There is a genuine tension
between exchangeability and positivity. Targeted Learning methods help manage this
trade-off, but they cannot eliminate it.
:::

---

## How to Diagnose Positivity Problems {#sec-diagnose}

Let us build a simulated dataset with a deliberate positivity problem, then work
through the diagnostic toolkit on it.

### Simulate data with a positivity gap

```{r}
#| code-fold: true
library(tidyverse)

set.seed(42)
n <- 2000

# Generate confounders
age <- rnorm(n, mean = 60, sd = 12)
severity <- rbinom(n, 1, prob = 0.4)

# Treatment assignment: strongly depends on age and severity
# Patients with age > 75 AND severity == 1 almost never get treated
logit_ps <- 0.5 - 0.06 * age - 1.8 * severity - 0.04 * (age > 75) * severity * 10
ps_true <- plogis(logit_ps)
A <- rbinom(n, 1, prob = ps_true)

# Outcome (binary: adverse event within 1 year)
lp_y <- -3 + 0.03 * age + 0.8 * severity - 0.5 * A + 0.01 * A * age
Y <- rbinom(n, 1, plogis(lp_y))

dat <- tibble(age, severity, A, Y, ps_true)

# Examine the positivity gap
dat |>
  mutate(age_group = case_when(
    age <= 55 ~ "<=55",
    age <= 65 ~ "56-65",
    age <= 75 ~ "66-75",
    TRUE       ~ ">75"
  )) |>
  group_by(age_group, severity) |>
  summarise(
    n = n(),
    n_treated = sum(A),
    prop_treated = mean(A),
    .groups = "drop"
  ) |>
  arrange(age_group, severity) |>
  knitr::kable(digits = 3, caption = "Treatment prevalence by age group and severity")
```

Look at the older, high-severity stratum. Treatment is rare there, sometimes
vanishingly so. That is our positivity problem.

### Propensity score density plots

The single most useful positivity diagnostic is a density plot of the estimated
propensity score, separated by treatment group.

```{r}
#| label: fig-ps-overlap-sim
#| fig-cap: "Propensity score distributions by treatment group in the simulated data. Good overlap (both groups spanning the full range) supports positivity; mass piling up near 0 or 1 in only one group signals near-violations."
#| code-fold: true
# Fit a propensity score model
ps_model <- glm(A ~ age * severity + I(age^2), data = dat, family = binomial)
dat$ps_hat <- predict(ps_model, type = "response")

ggplot(dat, aes(x = ps_hat, fill = factor(A))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(
    values = c("0" = "#E69F00", "1" = "#56B4E9"),
    labels = c("Untreated", "Treated")
  ) +
  labs(
    x = "Estimated propensity score",
    y = "Density",
    fill = "Group",
    title = "Propensity score overlap plot",
    subtitle = "Non-overlapping regions in the tails, where one group carries density the other lacks, signal near-violations"
  ) +
  theme_minimal()
```

Good overlap looks like the two densities sitting on top of each other across most of the propensity score range. Bad overlap looks like one group carrying density where the other has none. The non-overlapping region is where positivity is violated.

### Numerical overlap summaries

```{r}
#| code-fold: true
dat |>
  group_by(A) |>
  summarise(
    min_ps  = min(ps_hat),
    q05_ps  = quantile(ps_hat, 0.05),
    median  = median(ps_hat),
    q95_ps  = quantile(ps_hat, 0.95),
    max_ps  = max(ps_hat),
    .groups = "drop"
  ) |>
  knitr::kable(digits = 4, caption = "Propensity score distribution by treatment group")
```

### Identifying problematic covariate strata

Beyond the propensity score, inspect the raw covariate-treatment cross-tabulation.
Look for cells with zero or near-zero counts.

```{r}
#| code-fold: true
dat |>
  mutate(
    age_cat = cut(age, breaks = c(-Inf, 55, 65, 75, Inf),
                  labels = c("<=55", "56-65", "66-75", ">75"))
  ) |>
  count(age_cat, severity, A) |>
  pivot_wider(names_from = A, values_from = n, values_fill = 0) |>
  rename(untreated = `0`, treated = `1`) |>
  mutate(pct_treated = treated / (treated + untreated)) |>
  knitr::kable(digits = 3,
               caption = "Covariate strata by treatment: look for near-empty cells")
```

### A reusable data-support report

The individual checks above are easier to run as a single routine that returns the numbers most worth seeing before you trust any weighted or standardized estimate. Pass the treatment, the estimated propensity score, and a covariate frame; the report gives back the propensity mass in the tails, the largest stabilized weight, the effective sample size, and the covariate balance before and after weighting. Later chapters call back to this report rather than re-derive the pieces.

```{r}
#| code-fold: show
data_support_report <- function(A, ps, W) {
  w   <- ifelse(A == 1, mean(A) / ps, mean(1 - A) / (1 - ps))   # stabilized weights
  ess <- sum(w)^2 / sum(w^2)
  smd <- sapply(W, function(x) {
    s  <- sqrt((var(x[A == 1]) + var(x[A == 0])) / 2)           # pooled SD
    c(before = (mean(x[A == 1]) - mean(x[A == 0])) / s,
      after  = (weighted.mean(x[A == 1], w[A == 1]) -
                weighted.mean(x[A == 0], w[A == 0])) / s)
  })
  list(
    pct_ps_below_05  = round(100 * mean(ps < 0.05), 1),
    pct_ps_above_95  = round(100 * mean(ps > 0.95), 1),
    max_stab_weight  = round(max(w), 1),
    effective_n      = round(ess),
    nominal_n        = length(A),
    smd_before_after = round(t(smd), 3)
  )
}

data_support_report(dat$A, dat$ps_hat, dat[, c("age", "severity")])
```

Read it as a checklist. Any appreciable mass with a propensity below 0.05 or above 0.95 flags patients with little counterfactual support. A maximum stabilized weight in the tens or hundreds, or an effective sample size far below the nominal count, means a few patients dominate. The standardized mean differences should shrink toward zero after weighting; a covariate whose imbalance persists is one the weights failed to balance, which sends you back to the propensity model. The covariate-by-treatment cell table above complements this by exposing exact structural zeros. The per-stratum version of the same positivity check, applied inside each subgroup, appears in @sec-effect-modification.

---

## How Positivity Problems Affect Different Estimators {#sec-estimator-effects}

Using our simulated data, let us see how each estimator behaves when positivity is
violated. The true average treatment effect (ATE) in the data-generating process depends on the covariate
distribution, so we will judge the estimates by their *stability* and *plausibility*
rather than against a single truth.

### IPTW: extreme weights

IPTW inverts the propensity score, so patients with $P(A=1 \mid W) \approx 0$ who
nonetheless received treatment get enormous weights.

```{r}
#| code-fold: true
# Compute IPTW weights
dat <- dat |>
  mutate(
    ipw = if_else(A == 1, 1 / ps_hat, 1 / (1 - ps_hat))
  )

# Inspect weight distribution
dat |>
  summarise(
    mean_wt   = mean(ipw),
    max_wt    = max(ipw),
    pct_gt_10 = mean(ipw > 10) * 100,
    pct_gt_20 = mean(ipw > 20) * 100
  ) |>
  knitr::kable(digits = 2, caption = "IPTW weight summary: large weights signal trouble")

# IPTW estimate (Hajek / stabilized style)
ate_iptw <- dat |>
  summarise(
    ate = weighted.mean(Y[A == 1], ipw[A == 1]) -
          weighted.mean(Y[A == 0], ipw[A == 0])
  ) |>
  pull(ate)

cat("IPTW ATE estimate:", round(ate_iptw, 3), "\n")
```

::: {.callout-important}
## Why extreme weights are dangerous

A single patient with a weight of 50 counts as much as 50 patients. If that patient
has an unusual outcome (by chance), it can shift the entire ATE estimate. This is not
bias in the usual sense; it is **finite-sample instability** driven by near-violations
of positivity.
:::

### G-computation: silent extrapolation

G-computation fits an outcome model $E[Y \mid A, W]$ and then predicts under each
treatment value for every patient. When positivity is violated, those predictions
require the outcome model to extrapolate into covariate regions with no (or very few)
treated patients.

```{r}
out_model <- lm(Y ~ A * age * severity, data = dat)

dat <- dat |>
  mutate(
    Y1_hat = predict(out_model, newdata = mutate(dat, A = 1)),
    Y0_hat = predict(out_model, newdata = mutate(dat, A = 0))
  )

ate_gcomp <- mean(dat$Y1_hat - dat$Y0_hat)
cat("G-computation ATE estimate:", round(ate_gcomp, 3), "\n")
```

G-computation may look stable, but that stability is an illusion. It rests entirely on
the parametric model being correct in the extrapolation region, and there is no data to
check it there. The model might predict that a 90-year-old with severe comorbidities would benefit from treatment, but if no such patient was ever treated, that prediction is speculation dressed up as an estimate.

### TMLE: clever covariate instability

TMLE (@sec-doubly-robust-tmle, `doubly-robust-tmle.qmd`) updates the initial outcome model
using a "clever covariate" $H(A, W) = A / g(W) - (1-A) / (1 - g(W))$. When $g(W)$
is near zero, this clever covariate explodes, just like the IPTW weight. TMLE
inherits the instability of IPTW through its targeting step, even though it also uses
an outcome model. In practice, you might see one or two observations with clever-covariate values ten times larger than the rest, a sign that those few patients are driving the entire estimate.

```{r}
#| code-fold: true
# Demonstrate clever covariate magnitude
dat <- dat |>
  mutate(
    clever_cov = A / ps_hat - (1 - A) / (1 - ps_hat)
  )

dat |>
  summarise(
    max_abs_H  = max(abs(clever_cov)),
    q99_abs_H  = quantile(abs(clever_cov), 0.99),
    mean_abs_H = mean(abs(clever_cov))
  ) |>
  knitr::kable(digits = 2,
               caption = "Clever covariate magnitude: large values indicate positivity trouble")
```

The bottom line is that no estimator is immune to positivity violations. IPTW shows the problem as weight instability. G-computation hides it behind model extrapolation. TMLE sits in between: it inherits instability through the clever covariate while gaining some protection from the outcome model. The real fix is not to reach for a "better" estimator but to address the positivity problem directly.

---

## Practical Responses to Positivity Violations {#sec-responses}

### Redefine the target population

The most principled response is to restrict the analysis to the population where
overlap exists [@crump2009trimming]. You are changing the *question*, not patching the estimator.

```{r}
# Restrict to the overlap region (e.g., PS between 0.05 and 0.95)
dat_trimmed <- dat |>
  filter(ps_hat > 0.05, ps_hat < 0.95)

cat("Original sample:", nrow(dat), "\n")
cat("After trimming: ", nrow(dat_trimmed),
    sprintf("(%.0f%% retained)\n", 100 * nrow(dat_trimmed) / nrow(dat)))
```

This is the honest move. You are now estimating the ATE for the sub-population with adequate
support, not the full population. Say so clearly in the write-up.

### Weight truncation with sensitivity analysis

If trimming the population is not desirable, you can truncate (cap) the IPTW weights at
a threshold. This introduces some bias but reduces variance [@cole2008ipw; @hernan2025whatif]. The trick is to try several
thresholds and watch how the estimate moves.

```{r}
#| code-fold: true
truncation_sensitivity <- function(data, thresholds) {
  map_dfr(thresholds, function(thr) {
    d <- data |>
      mutate(ipw_trunc = pmin(ipw, thr))
    ate <- d |>
      summarise(
        ate = weighted.mean(Y[A == 1], ipw_trunc[A == 1]) -
              weighted.mean(Y[A == 0], ipw_trunc[A == 0])
      ) |>
      pull(ate)
    tibble(threshold = thr, ate_estimate = ate, max_weight = max(d$ipw_trunc))
  })
}

thresholds <- c(5, 10, 20, 50, 100, Inf)
results <- truncation_sensitivity(dat, thresholds)

results |>
  mutate(threshold = if_else(is.infinite(threshold), "None", as.character(threshold))) |>
  knitr::kable(digits = 3, caption = "Sensitivity of IPTW to weight truncation")
```

```{r}
#| label: fig-truncation-sensitivity
#| fig-cap: "ATE estimate as a function of the propensity-score truncation threshold. A stable line indicates the estimate does not hinge on a handful of extreme weights; a steep slope flags a result driven by near-positivity violations."
#| code-fold: true
# Visualize truncation sensitivity
results_plot <- results |>
  filter(is.finite(threshold))

ggplot(results_plot, aes(x = threshold, y = ate_estimate)) +
  geom_point(size = 3) +
  geom_line() +
  geom_hline(
    yintercept = results$ate_estimate[is.infinite(results$threshold)],
    linetype = "dashed", color = "red"
  ) +
  annotate("text", x = max(results_plot$threshold), y = ate_iptw,
           label = "Untruncated", hjust = 1, vjust = -0.5, color = "red") +
  labs(
    x = "Weight truncation threshold",
    y = "ATE estimate",
    title = "Truncation sensitivity analysis",
    subtitle = "Stable estimates across thresholds suggest robust findings"
  ) +
  theme_minimal()
```

If the ATE estimate barely moves as you relax the truncation threshold, the result is not driven by extreme weights and positivity violations are not dominating. If the estimate swings wildly, the causal conclusion hangs on a few patients with near-zero propensity scores, and you should be cautious.

### Modified treatment policies (shift interventions)

Instead of asking "What if everyone received treatment?", you can ask "What if
everyone's probability of treatment were shifted slightly upward?" This sidesteps
counterfactual scenarios that violate positivity. Later chapters work through these in detail; for now, hold on to the idea: if the static intervention
$\text{do}(A=1)$ creates positivity problems, a *stochastic* or *shift* intervention
may not.

### A remedies decision guide

The responses above are not a menu to pick from at random; the diagnosis points to the remedy. The table maps what the diagnostics show to what to do, and to how the estimand changes as a result.

| What the diagnostics show | Recommended response | Estimand consequence |
|---|---|---|
| A few extreme weights, estimate stable under truncation | Truncate and report a truncation sensitivity curve | Mild; the target is essentially unchanged |
| A region of covariate space with no overlap | Restrict to the overlap region | The target becomes the ATE in the overlap population |
| The question is about the treated, and controls are sparse | Target the average treatment effect among the treated (ATT) rather than the ATE | The question narrows to the treated |
| A static "treat everyone" regime has no support | Switch to a stochastic or modified treatment policy (@sec-adherence-shift, @sec-applied-lmtp) | The target becomes a feasible, support-respecting intervention |
| A longitudinal regime loses support across visits | Use a feasible shift rule rather than a fixed path | As above, defined visit by visit |

The unifying principle is that a positivity problem tells you the data cannot answer the original question. The honest move is to answer a nearby question the data can support, and to state plainly which question that is.

### Honest reporting

When positivity problems remain after reasonable adjustments, the most important
response is transparency. Report:

- The proportion of patients in low-overlap regions.
- The maximum and 99th-percentile IPTW weights.
- The effective sample size (next section).
- Sensitivity analyses under different truncation levels.
- The sub-populations where the estimate is well supported versus extrapolated.

---

## Effective Sample Size {#sec-ess}

Extreme weights do not just inflate variance; they reduce the amount of information in
your sample. The **effective sample size (ESS)** tells you how many equally weighted
observations your weighted sample is worth [@hernan2025whatif].

### Formula

For a set of weights $w_1, \ldots, w_n$:

$$
\text{ESS} = \frac{\left(\sum_{i=1}^n w_i\right)^2}{\sum_{i=1}^n w_i^2}
$$
{#eq-ess}

If all weights are equal, $\text{ESS} = n$. If one weight dominates,
$\text{ESS} \approx 1$.

### Code

```{r}
#| code-fold: true
compute_ess <- function(weights) {
  sum(weights)^2 / sum(weights^2)
}

# ESS for each treatment group
ess_treated   <- compute_ess(dat$ipw[dat$A == 1])
ess_untreated <- compute_ess(dat$ipw[dat$A == 0])

tibble(
  group        = c("Treated", "Untreated", "Total"),
  n_actual     = c(sum(dat$A == 1), sum(dat$A == 0), nrow(dat)),
  ess          = c(ess_treated, ess_untreated, compute_ess(dat$ipw)),
  pct_retained = round(100 * ess / n_actual, 1)
) |>
  knitr::kable(digits = 1, caption = "Effective sample size by treatment group")
```

### Threshold guidance

There is no universal cutoff, but common recommendations include:

- **ESS < 50% of actual sample size**: warning flag; inspect weights.
- **ESS < 20% of actual sample size**: serious concern; consider trimming or population restriction.
- **ESS far below n for one group but not the other**: asymmetric positivity, which often means one treatment direction has poor support.

::: {.callout-important}
## ESS is a summary, not a diagnosis

A healthy ESS does not guarantee that positivity holds everywhere. A few problematic
strata can be masked by many well-behaved ones. Always pair ESS with the overlap plots
and covariate-specific tables shown in @sec-diagnose.
:::

---

## Running Example: Antiretroviral Therapy Cohort {#sec-running-example}

See the [Running Case Study](00-case-study.qmd) for the full data description. The HAART cohort gives a clear illustration of practical positivity violations. Clinicians initiate HAART based on declining CD4 counts, so patients with very low baseline CD4 are nearly always treated. The untreated group then has little or no representation in the low-CD4 region, exactly the pattern we diagnosed in the simulated examples above.

```{r}
#| label: fig-ps-overlap-haart
#| fig-cap: "Propensity score overlap for HAART initiation in the running cohort. Substantial overlap across initiators and non-initiators is what makes the confounding-by-indication adjustment credible; regions where one group is absent are where positivity is strained."
#| code-fold: true
source("_data/load-haart-cohort.R")

# Fit propensity score model for HAART initiation
ps_haart <- glm(
  ever_haart ~ age + sex + cd4_sqrt_baseline,
  family = binomial,
  data = haart_baseline
)
haart_baseline <- haart_baseline |>
  mutate(ps_hat = predict(ps_haart, type = "response"))

# Propensity score overlap plot
ggplot(haart_baseline, aes(x = ps_hat, fill = factor(ever_haart))) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(
    values = c("0" = "#E69F00", "1" = "#56B4E9"),
    labels = c("No HAART", "HAART")
  ) +
  labs(
    x = "Estimated propensity score",
    y = "Density",
    fill = "Group",
    title = "Propensity score overlap: HAART initiation"
  ) +
  theme_minimal()
```

The overlap plot shows that patients with low `cd4_sqrt_baseline` cluster near propensity scores of 1, because clinical guidelines make HAART initiation near-certain for severely immunocompromised patients. This is a practical (not structural) positivity violation: HAART is not literally required, but it is rarely withheld.

Any causal estimate for those patients leans heavily on model extrapolation rather than observed counterfactual data. One option is to restrict the target population to the overlap region; another is to report effective sample sizes (@sec-ess), which quantify how much information these near-violations consume.

---

## Positivity in the longitudinal setting {#sec-longitudinal-positivity}

In a longitudinal analysis the treatment decision repeats at every visit, and positivity must hold at each one. The danger is multiplicative. A treatment probability that is merely uncomfortable at a single visit becomes untenable once it is chained across many visits, because the longitudinal estimand depends on the product of the per-visit probabilities. A regime that asks a patient to have followed an increasingly unlikely treatment path can end up with almost no support, even when each single step looked acceptable.

The diagnostic mirrors the point-treatment one, but you compute it visit by visit, and the cumulative product is the quantity to watch.

```{r}
#| code-fold: show
set.seed(5)
n  <- 1000
g1 <- plogis(rnorm(n, 1.0, 0.7))      # probability of following the regime at each visit,
g2 <- plogis(rnorm(n, 0.7, 0.9))      # each individually plausible
g3 <- plogis(rnorm(n, 0.4, 1.1))
cum_g <- g1 * g2 * g3                  # probability of following the regime through visit 3

data.frame(
  visit        = c("v1", "v2", "v3", "cumulative"),
  min_prob     = round(c(min(g1), min(g2), min(g3), min(cum_g)), 3),
  pct_below_10 = round(100 * c(mean(g1 < 0.1), mean(g2 < 0.1),
                               mean(g3 < 0.1), mean(cum_g < 0.1)), 1)
)
```

Each visit on its own looks tolerable, but the cumulative probability grows a long tail near zero, and the inverse of that cumulative probability is the weight a longitudinal IPTW analysis would carry. TMLE-based longitudinal estimators (@sec-illustrated-ltmle) avoid forming explicit weight products, but they inherit the same support limitation: the contrast still rests on person-histories that barely occur. The visit-by-visit treatment-probability and effective-sample-size diagnostic in @sec-applied-lmtp is the applied version of this check. When the cumulative support is thin, the static regime is the wrong target, and a modified treatment policy that shifts treatment by a feasible amount (@sec-adherence-shift) restores support.

---

## Summary {#sec-summary}

::: {.callout-tip}
## Chapter summary

1. **Positivity** requires that every confounder stratum has a non-zero probability
   of each treatment value. Without it, the causal effect is not identified from
   data alone.

2. **Structural violations** (treatment impossible) call for redefining the target
   population. **Practical violations** (treatment very rare) call for diagnostic
   checks and possibly modified estimands.

3. **Diagnosis toolkit**: propensity score overlap plots, covariate-by-treatment
   tables, effective sample size.

4. **All estimators suffer**: IPTW through extreme weights, G-computation through
   hidden extrapolation, TMLE through clever covariate instability.

5. **Practical responses**: restrict to the overlap population, truncate weights with
   sensitivity analysis, consider shift interventions, and report limitations honestly.

6. **Effective sample size** quantifies information loss from extreme weights and
   works as a quick summary diagnostic.
:::

---

## Check Your Understanding

**Question 1.** A colleague argues: "I use G-computation, so I don't need to worry
about positivity because I never compute weights." Why is this wrong?

<details>
<summary>Show answer</summary>

G-computation still requires estimating $E[Y \mid A=a, W=w]$ for both treatment
values at every covariate pattern. When no treated patients exist in a stratum,
G-computation extrapolates the outcome model into that region. The estimate there
depends entirely on the parametric model, not on data. The analyst may not even
realize this is happening because there are no extreme weights to flag the problem.
Positivity is an *identification* assumption, not an estimation assumption, so it
applies regardless of the estimator used.

</details>

**Question 2.** You fit a propensity score model and find that 95% of your sample has
propensity scores between 0.15 and 0.85, but 5% of treated patients have scores below
0.02. What should you do?

<details>
<summary>Show answer</summary>

First, investigate *who* those 5% are. Check which covariate patterns produce
near-zero propensity scores and ask whether those patients should be in the target
population at all (structural violation) or whether treatment is merely rare there
(practical violation). Then consider: (a) restricting to the overlap region
(PS > 0.02 or stricter), which changes the estimand to the sub-population with
adequate support; (b) truncating weights and running a sensitivity analysis across
several thresholds; (c) reporting the effective sample size with and without those
patients. Whatever you choose, be transparent that the ATE for the full population
is poorly supported by the data.

</details>

**Question 3.** Your effective sample size is 900 out of 1,000 actual patients. Does
this mean positivity holds?

<details>
<summary>Show answer</summary>

Not necessarily. ESS is a global summary. It is possible for a small subgroup to
have severe positivity problems that are diluted by the well-behaved majority. For
example, if 950 patients have moderate weights (around 1) and 50 patients in a
specific stratum have weights of 3 to 5, the overall ESS could still look healthy
while that stratum is poorly supported. Always supplement ESS with overlap plots and
covariate-specific tables to identify localized problems.

</details>

---

## What Comes Next

Now that you know when your data can and cannot support a causal question, you
are ready to see these ideas at work. The [adherence shift methods chapter](03-03b-adherence-shift-methods.qmd)
walks through a full applied example where positivity diagnostics, weight
inspection, and population-restriction decisions come up naturally in a longitudinal
pharmacoepidemiology study. You will see how the toolkit from this chapter fits together
with the estimation methods from @sec-iptw and @sec-doubly-robust-tmle in a realistic analysis
workflow.

---

```{r}
sessionInfo()
```
