---
title: "Estimating Interactions in Real-World Evidence Research"
format: html
---

# Estimating Interactions in Real-World Evidence Research {#sec-interactions-rwe}

*Estimated reading time: 50 minutes*

## Abstract

Interactions are among the most clinically interesting and methodologically
treacherous quantities in real-world evidence (RWE) research. Clinicians want
to know whether a drug works equally well across patient subgroups; public
health planners want to know whether two exposures combine to produce more
harm than the sum of their parts. Yet the literature is full of analyses that
report a single product-term coefficient from a logistic or Cox model and
treat it as the answer. This chapter argues for a different starting point.
Interaction analyses should begin with a precisely defined causal estimand
on a meaningful scale, anchored in a target trial protocol, and estimated
with methods that respect the complexity of observational data. Targeted
learning---combining flexible nuisance estimation through Super Learner
[@vanderLaan2007superlearner] with the targeting and influence-function inference
of TMLE [@vanderLaan2006tmle]---is particularly well suited to this task.
We work through one binary point-treatment example with TMLE plus Super
Learner, scaffold a CTMLE sensitivity analysis, sketch a longitudinal
extension with LTMLE, and close with a checklist that aligns with current
best-practice recommendations [@talbot2025tmle; @vanderweele2014tutorial;
@basten2025additive].

::: {.callout-tip}
## Key takeaways

- **Effect modification** (one treatment, varying effect across a baseline
  variable) and **interaction** (two exposures acting jointly) are different
  scientific questions and require different estimands.
- The same data can show interaction on the **additive** scale and not on
  the **multiplicative** scale, or vice versa. Pre-specify the scale that
  matches your decision question.
- A regression product term is rarely the scientific target. Define
  **subgroup-specific risks** and contrast them.
- Begin every RWE interaction analysis by writing the **target trial
  protocol** [@hernan2016target]: time zero, eligibility, treatment
  strategies, follow-up, outcome, censoring.
- **Positivity is harder inside subgroups** than overall. Diagnose overlap
  within each modifier stratum.
- **Targeted learning** (Super Learner + TMLE) supports flexible nuisance
  estimation, doubly robust point estimation, and influence-function-based
  uncertainty for the contrasts you actually want to report.
- **CTMLE** is best framed as a sensitivity analysis around fragile
  propensity-score models, not as the introductory teaching vehicle.
- Pre-specify a small number of clinically motivated effect modifiers,
  separate confirmatory from exploratory work, and report **both scales**
  with cell-specific risks and a common reference category
  [@vanderweele2014tutorial].
:::

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

---

## 1. Why interaction matters in epidemiology and real-world evidence

A single average treatment effect is rarely the end of a clinically useful
analysis. Two questions almost always follow. *Does the effect differ across
patient groups?* And *do two exposures combine in ways that one alone would
not predict?* Both questions are about interaction in a broad sense, but
they are not the same question, and the methods that answer them are not
identical.

Several closely related ideas are routinely conflated in published work.
Disentangling them is the first task of any chapter on interactions.

- **Effect modification** describes how the causal effect of a *single*
  treatment $A$ varies across levels of a baseline characteristic $V$.
  The target is a contrast of subgroup-specific causal effects, for example
  the risk difference if treated minus untreated within $V=1$ versus within
  $V=0$.
- **Interaction**, in the strict sense used by VanderWeele and Knol
  [-@vanderweele2014tutorial], refers to the *joint* causal effect of two
  exposures $A$ and $B$. Both are conceptually manipulable and the question
  is whether their combined effect departs from what you would expect from
  each effect alone.
- **Additive interaction** asks whether the joint effect departs from the
  *sum* of the individual effects on the absolute risk scale.
- **Multiplicative interaction** asks whether the joint effect departs from
  the *product* of the individual effects on the relative-risk scale. The
  same data can show interaction on one scale and none on the other.
- **Subgroup analysis** is a reporting style: presenting effects within
  prespecified strata. It is the *vehicle* for studying effect modification,
  but a list of subgroup point estimates is not by itself a heterogeneity
  test.
- **Predictive heterogeneity** or **individualized treatment effects** refer
  to algorithms that learn high-dimensional functions of covariates to
  predict who benefits. These are useful for treatment-rule discovery but
  should not be confused with the inferential, low-dimensional questions
  most epidemiologic interaction analyses address.

A second motivation for taking interaction seriously is **scale dependence**.
A cohort study may find that an antihypertensive halves stroke risk in
low-risk patients (relative risk 0.5) and also halves stroke risk in
high-risk patients (relative risk 0.5). On the multiplicative scale there is
no interaction. On the additive scale, the absolute risk reduction is much
larger for the high-risk group, and that is what determines who should be
treated first. Public health and clinical decisions almost always depend on
absolute differences, not ratios [@vanderweele2014tutorial; @basten2025additive].

A third motivation is the rapid growth of TMLE and other doubly robust
methods in observational epidemiology. A 2023 systematic review identified
89 observational studies that applied TMLE [@smith2023systematic], and the
growth is steepest in pharmacoepidemiology and comparative effectiveness
work---exactly the settings where interaction questions are most common.

### Why interaction matters in practice

Interaction is not an academic embellishment of an average effect; it is
the analytic foundation for several of the most important applied
questions in modern epidemiology and clinical practice.

- **Treatment targeting.** When the absolute benefit of a treatment is
  much larger in one subgroup than another, fixed budgets and finite
  clinic time make subgroup-aware deployment a matter of efficiency,
  not of precision medicine ideology. Statins offered to high-risk
  patients prevent more events per prescription than the same drug
  offered uniformly. The interaction analysis that quantifies "much
  larger" is what turns "treatment works on average" into "treat these
  patients first."
- **Policy heterogeneity.** Public-health interventions almost never
  have uniform effects across regions, age groups, or baseline-risk
  strata. A vaccination campaign that reduces hospitalization by 10
  percentage points among the elderly and 1 percentage point among the
  young calls for a different roll-out plan than one with uniform 5.5
  percentage-point effects. Interaction estimates inform where to
  invest the next dollar.
- **Personalized medicine.** Conditional treatment effect estimates,
  built from interaction analyses, are the input to optimal individual
  treatment rules: assign treatment $A$ to person $i$ if and only if the
  estimated benefit at $W_i$ exceeds the cost. This is the formal
  bridge from interaction analysis to the optimal-regime literature
  covered later in the handbook.

The common thread is that *interaction estimates produce decisions, not
just descriptions*. That is why the rest of this chapter spends so much
energy on getting the estimand, the scale, and the uncertainty right.

### Interaction vs. effect modification

Because these two ideas are conflated so often, it is worth restating the
distinction in operational terms:

- **Effect modification** is *heterogeneity in conditional effects*. The
  modifier $V$ need not be a manipulable exposure; it might be sex, a
  genotype, or a baseline biomarker. The estimand is the contrast of
  conditional treatment effects across levels of $V$.
- **Interaction** is a *contrast of joint interventions*. It requires that
  *both* $A$ and a second variable $E$ be conceptually amenable to
  intervention, and the estimand involves four counterfactual worlds in
  which $(A, E)$ is set jointly to $(0,0), (0,1), (1,0)$, and $(1,1)$.

The implication is methodological: a regression "interaction term" between
$A$ and $E$ does not estimate a causal interaction. It estimates a
parameter of a parametric model whose mapping to a causal contrast holds
only under strong assumptions. Many applied analyses report the
coefficient and call it interaction; this chapter argues that the causal
interaction parameter should be defined and estimated directly.

::: {.callout-note}
## What this means in practice
Before you write a single line of code, decide which of the four
ideas---effect modification, two-exposure interaction, additive contrast,
or multiplicative contrast---you are estimating. Different choices imply
different estimands, different reporting tables, and different defaults
for how to communicate uncertainty. *Causal interaction always involves
comparing four hypothetical worlds*; effect modification compares two
worlds within each level of a stratifier.
:::

---

## 2. From scientific question to causal estimand

The most common mistake in interaction analysis is to start with a model.
The right starting point is to write down the quantity you want to estimate
in counterfactual language, *before* any software is opened.

### Defining interaction as a causal estimand

Suppose two binary exposures $A$ and $E$ can each be intervened on, and let
$Y^{a,e}$ denote the potential outcome under joint intervention setting
$A=a$ and $E=e$. There are four such counterfactuals: $Y^{0,0}$, $Y^{0,1}$,
$Y^{1,0}$, $Y^{1,1}$. The causal interaction parameters compare how the
effect of one exposure depends on the level of the other.

- **Additive interaction parameter**:

$$\psi_{\text{add}} \;=\; E[Y^{1,1} - Y^{0,1}] \;-\; E[Y^{1,0} - Y^{0,0}]$$

- **Multiplicative interaction parameter**:

$$\psi_{\text{mult}} \;=\; \dfrac{E[Y^{1,1}] / E[Y^{0,1}]}{E[Y^{1,0}] / E[Y^{0,0}]}$$

These quantities are **parameters of the data-generating distribution**,
not coefficients of any model. Following the Targeted Learning principle
that the *parameter must be defined before the estimator*
[@petersen2014causal; @vanderLaan2006tmle], we treat any subsequent
modeling choice (logistic, log-binomial, TMLE, g-computation) as a means
to estimate $\psi_{\text{add}}$ or $\psi_{\text{mult}}$, not as a
substitute definition.

In plain English: **we are asking how the effect of $A$ changes when we
also intervene on $E$.** $\psi_{\text{add}} = 0$ means the absolute effect
of $A$ is the same whether $E$ is set to $0$ or $1$. $\psi_{\text{mult}} =
1$ means the relative effect of $A$ is the same across levels of $E$.
Either parameter can be nonzero on its own scale while the other is null.

Parametric models that impose linear or log-linear interactions can
misrepresent the underlying causal interaction structure: the model's
"interaction term" is a derivative of the assumed functional form, not of
the true joint potential-outcome distribution. When the functional form is
wrong, the coefficient is wrong, and a likelihood-ratio test on the
product term tells you about the model's fit, not about $\psi_{\text{add}}$
or $\psi_{\text{mult}}$.

::: {.callout-note}
## A toy example: computing interaction by hand

Suppose the four counterfactual risks are known exactly:

| | $E=0$ | $E=1$ |
|---|---:|---:|
| $A=0$ | $E[Y^{0,0}] = 0.05$ | $E[Y^{0,1}] = 0.10$ |
| $A=1$ | $E[Y^{1,0}] = 0.08$ | $E[Y^{1,1}] = 0.20$ |

The effect of $A$ when $E=0$ is $0.08 - 0.05 = 0.03$ (3 percentage
points). The effect of $A$ when $E=1$ is $0.20 - 0.10 = 0.10$ (10
percentage points).

- **Additive interaction**: $\psi_{\text{add}} = 0.10 - 0.03 = 0.07$.
  Treatment is 7 percentage points more beneficial (or harmful) when
  $E=1$ than when $E=0$.
- **Multiplicative interaction**: the risk ratio is
  $0.08 / 0.05 = 1.60$ when $E=0$ and $0.20 / 0.10 = 2.00$ when $E=1$,
  giving $\psi_{\text{mult}} = 2.00 / 1.60 = 1.25$. Treatment multiplies
  the risk by 25% more when $E=1$.
- **RERI** (taking $A=0, E=0$ as the doubly unexposed reference):
  $RR_{11} = 4.0$, $RR_{10} = 1.6$, $RR_{01} = 2.0$, so
  $\text{RERI} = 4.0 - 1.6 - 2.0 + 1 = 1.4$.

The same four numbers produce three different interaction summaries.
TMLE's role is to estimate those four numbers from observational data
under confounding; once they are in hand, the contrasts are arithmetic.
:::

### Subgroup-specific estimands

When the second variable is a baseline modifier $M$ that you do *not*
intervene on, the relevant estimands are subgroup-specific. Let $Y^a$
denote the potential outcome under treatment level $a$. The estimands you
will most often want are:

- **Subgroup-specific risk under treatment**: $P(Y^1 = 1 \mid M = m)$
- **Subgroup-specific risk under control**: $P(Y^0 = 1 \mid M = m)$
- **Subgroup-specific risk difference**: $RD(m) = P(Y^1=1 \mid M=m) - P(Y^0=1 \mid M=m)$
- **Subgroup-specific risk ratio**: $RR(m) = P(Y^1=1 \mid M=m) / P(Y^0=1 \mid M=m)$
- **Additive interaction contrast**: $RD(1) - RD(0)$
- **Multiplicative interaction contrast**: $RR(1) / RR(0)$

For two binary exposures $A$ and $B$ acting jointly, the standard
public-health summary is the **relative excess risk due to interaction**
(RERI):

$$\text{RERI} = RR_{11} - RR_{10} - RR_{01} + 1$$

where $RR_{ab}$ is the risk ratio under joint exposure level $(A=a, B=b)$
relative to the doubly unexposed group. RERI = 0 indicates no additive
interaction; positive values indicate super-additivity, negative values
sub-additivity. The interpretation is straightforward only when the doubly
unexposed group has the lowest baseline risk; we return to this point
below.

Why bother with this counterfactual machinery rather than reading off the
product-term coefficient from a logistic regression? Three reasons.

1. **The coefficient is on the wrong scale**. A product-term coefficient in
   a logistic model speaks to the odds ratio, not the risk ratio or risk
   difference. Odds-ratio interaction is not generally interpretable as
   either additive or multiplicative interaction in the public-health sense
   [@vanderweele2014tutorial].
2. **The coefficient is conditional, not marginal**. It estimates an effect
   conditional on covariates included in the model. The marginal subgroup
   effect---averaging over the covariate distribution within the
   subgroup---is what most decision questions actually need.
3. **The coefficient confounds estimation with model specification**. Move
   one covariate in or out and the interaction coefficient can change
   meaningfully. A causal estimand defined directly from potential outcomes
   does not depend on whether you used a polynomial or a spline for $W_1$.

Table 1 contrasts the two scales explicitly.

| Feature | Additive interaction | Multiplicative interaction |
|---|---|---|
| Mathematical form | $RD(1) - RD(0)$, or $\text{RERI}$ for two exposures | $RR(1)/RR(0)$, or $RR_{11}/(RR_{10} \cdot RR_{01})$ |
| Underlying scale | Absolute risk | Relative risk |
| Best for | Public health targeting, resource allocation, number-needed-to-treat reasoning | Etiologic comparisons, ratio-style benefit-risk summaries |
| Default model | None; estimate by standardization or TMLE | Logistic, log-binomial, Cox product term |
| Coding sensitivity | Highly sensitive when one exposure is preventive; lowest-risk stratum should be reference | Less sensitive but still requires explicit reference |
| Decision relevance | Direct: tells you absolute benefit by subgroup | Indirect: needs baseline risk to translate into absolute benefit |
: Table 1. Additive versus multiplicative interaction at a glance. {.striped}

A second comparison maps each scientific question to a recommended
estimator family.

| Scientific aim | Preferred estimand | Recommended estimator |
|---|---|---|
| Treatment effect within prespecified subgroups | $RD(m)$, $RR(m)$ | TMLE within strata, or stratified Super Learner outcome regression with standardization |
| Difference in treatment effect across subgroups | $RD(1) - RD(0)$, $RR(1)/RR(0)$ | TMLE in each stratum, then delta-method or bootstrap contrasts |
| Joint effect of two binary exposures | $RR_{11}, RR_{10}, RR_{01}$, RERI | TMLE for each joint regime mean, with post-estimation contrasts |
| Time-varying treatment with subgroup contrast | Counterfactual mean / survival under dynamic strategy by subgroup | LTMLE [@vanderLaan2012ltmle] or TMLE for marginal structural models |
| Many candidate modifiers, exploratory discovery | Conditional average treatment effect surface | Causal forest or simultaneous one-step TMLE with multiplicity control [@wei2023simultaneous] |
: Table 2. Causal estimands for interaction questions and recommended estimators. {.striped}

::: {.callout-note}
## What this means in practice
Write the estimand on a sticky note before you write the model. If you
cannot describe what you are estimating in one sentence using potential
outcomes, you are not yet ready to estimate it.
:::

---

## 3. Causal assumptions and target trial framing

Interaction analyses inherit every causal assumption that average-effect
analyses require, and several of them get harder.

### The target trial protocol

In any RWE study, Hernán and Robins [-@hernan2016target] recommend that the
analysis be designed as an *emulation of a target randomized trial*. For
interaction work, this discipline is doubly important, because subgroup
analyses are uniquely vulnerable to design errors that mimic real
heterogeneity. Before any modeling decision, the analyst should write down:

- **Eligibility criteria** including age, indication, and absence of prior
  outcome events.
- **Treatment strategies** to be compared, including dose, duration, and
  initiation rules.
- **Time zero**: the date when eligibility is met, the assigned strategy
  begins, and follow-up starts.
- **Follow-up window** and outcome definition.
- **Censoring rules**: discontinuation, switching, loss to follow-up,
  competing events.
- **Candidate effect modifiers**: pre-treatment, well-measured, biologically
  or clinically motivated.
- **Causal assumptions** the analysis will rely on.

If any of these are decided after looking at the outcome, you risk
manufacturing apparent heterogeneity through differential immortal time,
selection bias, or post-hoc subgroup tuning.

### The four standard assumptions

For the subgroup-specific causal contrast $E[Y^a \mid M=m]$ to be
identified from observed data, you need:

- **Exchangeability**: within levels of measured confounders $W$ (and
  conditional on $M$), treatment assignment is independent of potential
  outcomes. In notation, $Y^a \perp\!\!\!\perp A \mid W, M$. Often called
  *no unmeasured confounding*.
- **Positivity**: every covariate-modifier stratum that exists in the
  population contains both treated and untreated patients. Formally,
  $0 < P(A=1 \mid W, M) < 1$.
- **Consistency**: the observed outcome under the treatment a person
  actually received equals their potential outcome under that treatment.
- **No interference, unless modeled**: one person's treatment does not
  affect another person's outcome. Vaccine, infectious-disease, and network
  studies require explicit interference modeling.

### Why positivity is the silent killer of interaction analyses

The positivity assumption is local. It must hold within every covariate
pattern you condition on. As soon as you stratify by a modifier or
require support for a joint exposure, you are conditioning on more, and
any positivity weakness in the original cohort is amplified.

**Interaction estimation is more sensitive to positivity violations
than main-effect estimation, because it requires support across joint
exposure strata.** A main-effect analysis needs both treated and
untreated patients across the covariate distribution; an interaction
analysis needs both, *within every level of the second exposure or
modifier*. The number of cells that must be populated grows with the
product of the exposure cardinalities, while the data per cell shrinks.

A concrete example makes this vivid. Consider a 4,000-patient cohort
with binary $A$ (statin use), binary $E$ (heavy alcohol use), and
several binary clinical confounders. The marginal cells are reasonably
large:

| Cell | $A=0, E=0$ | $A=0, E=1$ | $A=1, E=0$ | $A=1, E=1$ |
|---|---:|---:|---:|---:|
| Patients | 1,800 | 600 | 1,400 | 200 |
| Outcome events | 90 | 60 | 70 | 20 |

The doubly exposed cell ($A=1, E=1$) holds 200 patients, of whom 20 have
the outcome. Within that cell, the propensity score for $A$ given the
clinical confounders may be near 1 for most patients (they are precisely
the people likely to be on statins) and near 0 for a handful---**that
handful then dominates the inverse-probability weights for the
interaction estimate**. The headline RERI may move dramatically when
those few patients are dropped, even though the main-effect estimate of
statins on the outcome is stable.

Two recurring failure modes follow:

- **A frailty-by-treatment interaction** where almost no frail patients
  receive aggressive therapy. Within the frail stratum, positivity
  fails: you cannot estimate $E[Y^1 \mid \text{frail}]$ from the data,
  no matter what estimator you choose.
- **A two-exposure analysis of statin use and heavy alcohol use** as
  above. The $A=1, E=1$ joint cell may be tiny in any realistic cohort.
  The apparent RERI estimate may be dominated by a handful of patients
  with extreme weights, and the confidence interval will be wide,
  unstable, or both.

This is why propensity-score diagnostics should be reported *within
each joint exposure cell or modifier subgroup*, not only overall, and
why collaborative TMLE (Section 7) was developed as a sensitivity
strategy for fragile propensity-score estimation.

### Interaction through the Targeted Learning roadmap

Petersen and van der Laan [-@petersen2014causal] formalize the workflow as
a small sequence of disciplined steps. Applied specifically to
interaction, the roadmap separates the *causal question* from the
*statistical procedure* and protects the analysis from the most common
failure mode: letting a model dictate what counts as interaction.

::: {.callout-important}
## Interaction through the Targeted Learning roadmap

**Step 0 — Define the question.** State, in clinical or public-health
language, whether you are asking (i) does the effect of $A$ vary across a
non-manipulable modifier $M$ (effect modification), or (ii) do two
manipulable exposures $A$ and $E$ combine in ways that depart from their
individual effects (interaction).

**Step 1 — Statistical model.** Adopt a *nonparametric* model for the
joint distribution of $(W, A, E, Y)$ (or $(W, A, M, Y)$). No functional
form is imposed at this step. The Targeted Learning principle is that
parametric structure should be a tool, not an assumption baked into the
estimand.

**Step 2 — Causal estimand.** Write the parameter explicitly as a
function of counterfactual outcomes:
$\psi_{\text{add}} = E[Y^{1,1} - Y^{0,1}] - E[Y^{1,0} - Y^{0,0}]$ for the
joint-intervention case, or $RD(1) - RD(0)$ for the modifier case. *This
is what the analysis aims to recover; it is not a regression coefficient.*

**Step 3 — Identification.** State the assumptions under which the causal
estimand maps to a function of the observed-data distribution:
exchangeability for $A$ (and $E$) given $W$, joint positivity, consistency,
and no interference. Identification fails or weakens whenever any of
these break, and the estimator cannot rescue identification it does not
have.

**Step 4 — Estimation.** Estimate the nuisance functions
($Q$ for the outcome regression, $g$ for the treatment mechanisms) with
Super Learner, then apply a TMLE targeting step so that the plug-in
estimator solves the efficient influence function equation for
$\psi_{\text{add}}$ or $\psi_{\text{mult}}$.

**Step 5 — Sensitivity analysis.** Vary the learner library, the
adjustment set, and the propensity-score selection (e.g., CTMLE) to
characterize the stability of the interaction estimate. Quantify the
amount of unmeasured confounding that would be needed to overturn it.
:::

The roadmap's message for interaction work is uncompromising: **interaction
parameters are not coefficients; they are functions of counterfactual
risks**. Any procedure that conflates these---reporting a logistic
product term as if it answered $\psi_{\text{add}}$, for example---has
skipped Step 2.

::: {.callout-note}
## What this means in practice
Treat positivity diagnostics inside subgroups as a non-negotiable step, not
an optional appendix. When the modifier strata are small or the joint
exposure cells are sparse, no estimator---however sophisticated---can
manufacture identifiability that the data do not support.
:::

---

## 4. Why targeted learning is useful for interaction estimation

Targeted learning is the combination of (a) flexible, data-adaptive
nuisance estimation through Super Learner and (b) targeted maximum
likelihood estimation (TMLE), which adjusts the nuisance estimates so that
the resulting plug-in estimate is unbiased for the causal parameter of
interest [@vanderLaan2006tmle; @vanderLaan2007superlearner]. Three features make
it especially well suited to interaction analyses.

### Super Learner in plain English

Super Learner is a stacked ensemble. You provide a *library* of candidate
algorithms---perhaps a logistic regression, a penalized regression, a
random forest, gradient boosting, and a simple intercept-only baseline---
and Super Learner uses cross-validation to compute a weighted combination
that minimizes prediction risk. The learner is guaranteed (in large
samples) to perform at least as well as the best single algorithm in the
library [@vanderLaan2007superlearner]. For interaction analyses, this matters
because nuisance functions inside subgroups can have very different
shapes; an ensemble accommodates that heterogeneity without forcing the
analyst to commit to one functional form.

### TMLE in plain English

TMLE has two stages.

1. **Estimate the nuisance functions flexibly.** Fit an outcome regression
   $\bar{Q}(A,W) = E[Y \mid A,W]$ and a propensity score $g(W) = P(A=1 \mid W)$,
   typically using Super Learner.
2. **Target the estimate.** Update the outcome regression using a small
   parametric correction (often called the *clever covariate* or *targeting
   step*) so that the resulting plug-in estimate solves the efficient
   influence function equation for your target parameter.

The substitution principle---computing the target as a function of the
updated regression---means TMLE produces a parameter estimate that lives
on the natural parameter space (a probability between 0 and 1, a positive
risk ratio, etc.). That property is particularly attractive for subgroup
analyses, where unconstrained estimators can occasionally produce
nonsensical numbers in small strata.

### Double robustness

TMLE is **doubly robust** in the sense that it remains consistent for the
target causal parameter as long as *either* the outcome regression *or*
the propensity score is consistently estimated. In an interaction context,
that means the analyst gets two independent chances to control confounding
inside each subgroup---a real safety margin when modifier strata are small
and either nuisance model can be unstable.

### Influence-function-based inference

Once the targeting step is complete, the variance of the TMLE can be
estimated from the empirical influence function (essentially, the
contribution of each observation to the parameter estimate). Confidence
intervals are then standard Wald intervals on the appropriate scale. The
attractive feature is that the same machinery extends to *contrasts of*
TMLE estimates across subgroups, because the influence function of a
contrast is a difference of influence functions.

### When to escalate: cross-fitting, CV-TMLE, drtmle

When learner libraries become aggressive (deep ensembles, many tree-based
learners, high-dimensional features), the regularity conditions that
underpin standard influence-function inference can become hard to verify.
**Cross-fitting** estimates each nuisance function on data that excludes
the observation being predicted; **CV-TMLE** integrates that splitting
into the targeting step; and `drtmle` provides additional robustness
properties under adaptive nuisance estimation. These tools should be on
the analyst's radar but are not essential for a first interaction
analysis.

### From estimand to estimator: practical workflow

The slogan "estimand first, then estimator" is easier to repeat than to
operationalize. The five steps below make the mapping concrete for an
interaction analysis with a binary treatment $A$ and a binary second
exposure $E$ (or modifier $M$). **At its heart, the procedure estimates
four counterfactual means and combines them into the contrast you
specified in Section 2.**

::: {.callout-important}
## From estimand to estimator (5 steps)

**Step 1 — Estimate the outcome regression** $Q(A, E, W) = E[Y \mid A, E, W]$
with Super Learner. This is a single model that uses the *observed* data
and includes treatment, second exposure, and confounders as predictors.

**Step 2 — Generate counterfactual predictions** under each of the four
joint regimes by querying the same fitted model on modified data:
$Q(1,1,W_i)$, $Q(0,1,W_i)$, $Q(1,0,W_i)$, $Q(0,0,W_i)$ for every patient
$i$.

**Step 3 — Compute the interaction contrast** by averaging:
$\widehat{\psi}_{\text{add}} = \frac{1}{n}\sum_i [Q(1,1,W_i) - Q(0,1,W_i)] - \frac{1}{n}\sum_i [Q(1,0,W_i) - Q(0,0,W_i)]$.
At this point you have a plug-in g-computation estimate.

**Step 4 — Apply the TMLE update.** Estimate the treatment mechanisms
$g_A(A \mid W)$ and $g_E(E \mid W)$ (or their joint $g(A,E \mid W)$) with
Super Learner. Use these to construct the clever covariate that updates
$Q$ along the least-favorable submodel for your target parameter. The
updated $Q^*$ now solves the efficient influence-function equation for
$\psi_{\text{add}}$ (or $\psi_{\text{mult}}$).

**Step 5 — Inference via the influence curve.** Compute the empirical
influence function of $\widehat{\psi}_{\text{add}}$ using $Q^*$ and $g$,
then form a Wald confidence interval from its empirical variance. The
same machinery yields valid CIs for any smooth contrast of the four
counterfactual means.
:::

Section 6 implements Steps 1–3 directly with `sl3` and Steps 4–5 via
`tmle3` (or stratified `tmle` plus bootstrap as a transparent
alternative). The point of the boxed mapping is that *every line of
code in Section 6 corresponds to one of these five steps*; if you can
say which step a line belongs to, the workflow is well-specified.

### Why TMLE is preferred for interaction

Targeted learning is useful for any causal parameter, but interaction
amplifies the two pathologies it was designed to fix. **Model
misspecification risk is amplified** because interaction is a
*derivative* of the outcome surface: it depends on differences of
differences of fitted values, and small functional-form errors that are
invisible in a main-effect estimate can dominate an interaction
estimate. **Positivity problems are amplified** because the joint
exposure cells $(A=a, E=e)$ are typically sparser than either marginal
cell, and inverse-probability weights deteriorate accordingly.

Two TMLE properties directly address these amplifications. **Flexible
nuisance estimation through Super Learner** lets the data choose the
functional form of $Q$ rather than imposing a parametric form whose
interaction term is a model artifact. **Double robustness** means the
interaction estimate remains consistent even if one of $Q$ or $g$ is
misspecified, which is a real safety margin in joint-cell estimation
where neither model is easy to get exactly right. The targeting step
also improves efficiency relative to plug-in g-computation, which
matters when the joint cell is small and every bit of information has
to be used optimally.

### Why this matters for interactions specifically

Three pieces of recent literature are worth keeping in view:

- A 2023 systematic review of 89 observational TMLE applications
  documented rapid uptake in epidemiology, with frequent use for subgroup
  and heterogeneity questions [@smith2023systematic].
- A 2025 best-practice tutorial walks through TMLE implementation choices
  with a focus on accessibility for applied researchers
  [@talbot2025tmle].
- A 2023 paper develops simultaneous one-step TMLE for many subgroup
  effects, with valid joint confidence intervals---directly relevant when
  the analyst wants to interrogate several modifiers at once
  [@wei2023simultaneous].

::: {.callout-note}
## What this means in practice
For most MPH-level interaction analyses, the rule is simple: estimate
nuisance functions with Super Learner using a modest, sensible library;
target with TMLE inside each modifier stratum; and use the influence
function for stratum-specific inference. Reserve cross-fitting, CV-TMLE,
and `drtmle` for situations where the library is genuinely aggressive or
the sample is large enough to support data splitting.
:::

---

## 5. Additive and multiplicative interaction in practice

The single most important recommendation in the VanderWeele and Knol
tutorial [-@vanderweele2014tutorial] is to report both scales whenever the
outcome and data permit. Scale dependence is not a methodological detail;
it is a substantive claim about how the world is supposed to combine.

### When additive interaction is preferable for public-health decisions

If the question is *who should be prioritized for an intervention?* the
answer almost always lives on the additive scale. A 2-percentage-point
absolute risk reduction in a high-incidence subgroup represents a much
larger public health benefit than the same relative-risk reduction in a
low-incidence subgroup. Number-needed-to-treat reasoning, resource
allocation, and ranking of intervention sites are all additive
operations. Multiplicative measures are useful for etiologic synthesis but
do not by themselves answer "where should we deploy this drug first?"
[@basten2025additive; @vanderweele2014tutorial].

### Operationalizing the choice of scale

The two scales answer two different questions, and the analyst should
state which one the analysis is built to answer.

| Scale | What it tells you | Best for |
|---|---|---|
| **Additive** ($\psi_{\text{add}}$, RD contrasts, RERI) | Public-health impact: how many additional events are prevented (or caused) per person treated, by subgroup | Decision-making, resource allocation, NNT, policy targeting |
| **Multiplicative** ($\psi_{\text{mult}}$, RR contrasts) | Relative effect: how much the *ratio* of treated to untreated risk changes across exposure cells | Etiologic comparison, meta-analytic synthesis, biology |

::: {.callout-tip}
## Recommendation
**For decision-making, prioritize additive interaction unless there is a
specific scientific reason to use the multiplicative scale.** Report
the multiplicative contrast alongside it whenever feasible---the two
scales together leave the reader free to translate the result into
either decision or etiologic terms, while a single scale forces a
choice the reader may not share.
:::

### Coding matters: the lowest-risk reference rule

Standard additive interaction measures, including RERI, assume the doubly
unexposed group is the reference. When one exposure is *preventive*
(e.g., statin use, vaccination), simply plugging into the RERI formula
without re-coding can produce values that look like "antagonism" but
actually reflect a coding artifact.

The remedy, recommended in [@vanderweele2014tutorial], is to **re-code so
that the lowest-risk joint stratum is the reference**. For statin use ($A$)
and heavy smoking ($B$), define the reference as the joint stratum with
the lowest expected outcome risk (typically statin user, non-smoker), and
report RERI relative to that reference. The numerical answer is then
interpretable as the additive departure from the safest combination.

### Cell-specific reporting

Whatever scale you privilege, the most useful presentation is a
**cell-specific risk table**: estimated risks under each combination of
exposures (or under treatment vs. control within each modifier stratum),
plus the contrasts of interest. Readers can then reconstruct any other
contrast they care about and assess whether the differences are large
enough to matter clinically.

### Interaction in real-world evidence

Real-world data add three challenges to the picture above that randomized
trials usually do not face.

- **Positivity issues are amplified.** Joint exposure cells $(A=a, E=e)$
  may be sparsely populated, especially when both exposures are rare or
  when one is preventive and the other indicated for similar patients.
  Subgroup-level positivity must be checked separately within each cell;
  inverse-probability components become unstable as the conditional
  probability of the joint regime approaches zero. Positivity weakness is
  the single most common reason interaction analyses produce headline
  estimates that fail to replicate.
- **Model dependence is amplified.** Interaction estimates are derivatives
  of the outcome surface and are intrinsically more sensitive to
  functional-form assumptions than main effects. Parametric models that
  impose linear or log-linear interactions can misrepresent the underlying
  causal interaction structure. Super Learner mitigates this by allowing
  the data to choose among flexible learners, and TMLE protects against
  remaining misspecification through double robustness.
- **Interpretability hinges on scale.** A real-world evidence audience
  typically wants to know whether and where to deploy a treatment. That
  question lives on the additive scale. Multiplicative interaction
  remains useful for etiologic synthesis and meta-analytic comparison,
  but should not be the only summary in a decision-oriented report
  [@vanderweele2014tutorial; @basten2025additive].

A practical recommendation follows: in RWE, **prefer the additive scale
for decision-making, report the multiplicative scale for completeness,
and run sensitivity analyses that vary both the nuisance learner library
and the propensity-score model**.

::: {.callout-note}
## What this means in practice
Pre-specify the scale, use a single common reference category, and present
the cell-specific risks. If you have to pick one scale, pick additive for
public-health audiences and multiplicative for etiologic ones. When you
can, present both. In RWE, build the analysis to survive the joint-cell
positivity check and the alternative-learner sensitivity test.
:::

### Diagnostic checklist for interaction analyses

::: {.callout-tip}
## Five questions before you trust an interaction estimate

1. **Estimand.** Is interaction defined as a causal parameter
   ($\psi_{\text{add}}$, $\psi_{\text{mult}}$, RERI, or a contrast of
   subgroup risk differences), not as a regression coefficient?
2. **Positivity.** Is the joint-exposure cell or the modifier stratum
   adequately populated, with both treated and untreated patients
   represented?
3. **Learners.** Are flexible nuisance learners (Super Learner with a
   diverse library) used, rather than a single parametric model that
   could impose the wrong functional form?
4. **Scale.** Is the choice of additive vs. multiplicative scale
   justified by the decision question, with both reported when feasible?
5. **Sensitivity.** Are sensitivity analyses prespecified that vary the
   learner library, the adjustment set, and the propensity-score
   selection?
:::

---

## 6. Point-treatment worked example with TMLE and Super Learner

This section demonstrates the workflow on a simulated cohort with a binary
treatment $A$, a binary baseline modifier $M$, a binary outcome $Y$, and
three baseline confounders $W_1, W_2, W_3$. The data-generating mechanism
includes genuine effect modification on the additive scale. We present
two complementary implementations: a **joint-intervention plug-in TMLE**
that estimates the four counterfactual means directly, and a
**stratified TMLE** that fits separately within each modifier level. They
target the same parameter and should agree up to sampling variability.

::: {.callout-tip}
## Minimal implementation summary

1. **Estimate the outcome regression** $Q(A, E, W)$ with Super Learner.
2. **Predict under four interventions**: $(A,E) = (1,1), (0,1), (1,0), (0,0)$.
3. **Compute the additive interaction contrast**:
   $\widehat{\psi}_{\text{add}} = \overline{Q(1,1,W)} - \overline{Q(0,1,W)} - \overline{Q(1,0,W)} + \overline{Q(0,0,W)}$.
4. **Apply the TMLE update** using $g_A(A\mid W)$ and $g_E(E\mid W)$.
5. **Use the influence curve** for inference (Wald confidence intervals).
:::

### Estimating interaction using TMLE: the joint-intervention workflow

When both $A$ and $E$ are conceptually manipulable (here we treat $M$ as
the second exposure for illustration), the data structure is
$O = (W, A, E, Y)$ and the estimation steps follow the roadmap of Section
3:

1. **Estimate the outcome regression** $Q(A, E, W) = E[Y \mid A, E, W]$
   with Super Learner.
2. **Estimate the treatment mechanisms** $g_A(A \mid W)$ and
   $g_E(E \mid W)$, or their joint $g(A, E \mid W)$, with Super Learner.
3. **Generate counterfactual predictions** under each of the four joint
   regimes: $Q(1,1,W)$, $Q(0,1,W)$, $Q(1,0,W)$, $Q(0,0,W)$.
4. **Plug into the interaction estimand**:
   $\widehat{\psi}_{\text{add}} = \frac{1}{n}\sum_i \big[Q(1,1,W_i) - Q(0,1,W_i)\big] - \frac{1}{n}\sum_i \big[Q(1,0,W_i) - Q(0,0,W_i)\big]$.
5. **Apply the TMLE targeting step** so that the plug-in estimator solves
   the efficient influence-function equation; this is what turns
   g-computation into TMLE and supplies double robustness.
6. **Use the influence curve** for variance estimation and Wald
   confidence intervals.

In plain English: **we predict outcomes under four hypothetical
worlds---everyone exposed to both, exposed to one, exposed to the other,
or neither---and then compare the differences across those worlds.** TMLE
uses both the $Q$ and $g$ components for efficiency: if either is
estimated consistently, the interaction estimate remains consistent
(double robustness).

```{r}
#| label: sl3-tmle3-joint-intervention
#| eval: false

# Joint-intervention plug-in g-computation skeleton
# (extend with tmle3 or ltmle for the full targeting step and IC inference)
library(sl3)
library(tmle3)
library(dplyr)

# Define a Super Learner library for nuisance estimation
lrn <- Lrnr_sl$new(
  learners = list(
    Lrnr_glm_fast$new(),
    Lrnr_ranger$new(),
    Lrnr_xgboost$new()
  )
)

# Outcome regression task: predict Y from (A, E, W)
Q_task <- make_sl3_Task(
  data       = dat,
  outcome    = "Y",
  covariates = c("A", "M", "W1", "W2", "W3")
)
Q_fit <- lrn$train(Q_task)

# Generate counterfactual predictions under each of four joint regimes
predict_under <- function(a_val, e_val) {
  newdat <- mutate(dat, A = a_val, M = e_val)
  predict(Q_fit, make_sl3_Task(
    data       = newdat,
    outcome    = "Y",
    covariates = c("A", "M", "W1", "W2", "W3")
  ))
}

Q_11 <- predict_under(1, 1)
Q_01 <- predict_under(0, 1)
Q_10 <- predict_under(1, 0)
Q_00 <- predict_under(0, 0)

# Plug-in estimate of the additive interaction parameter
psi_add <- mean(Q_11 - Q_01) - mean(Q_10 - Q_00)

# Plug-in estimate of the multiplicative interaction parameter
psi_mult <- (mean(Q_11) / mean(Q_01)) /
            (mean(Q_10) / mean(Q_00))

list(psi_add = psi_add, psi_mult = psi_mult)
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"We trained a flexible model that predicts outcomes from treatment, the
second exposure, and patient characteristics. Then we asked the model:
*what would happen if everyone got both, neither, or just one of the
two?* The interaction is how much the effect of treatment changes when
the second exposure is also present."
:::

::: {.callout-warning}
## From plug-in g-computation to TMLE
The block above is the **plug-in g-computation** step. To obtain a true
TMLE---with the targeting update and influence-function-based
inference---wrap the same nuisance estimates in the `tmle3` workflow
with a custom parameter for $\psi_{\text{add}}$ or $\psi_{\text{mult}}$,
or fit `ltmle` with two intervention nodes ($A$ and $E$) and contrast
the four resulting counterfactual means. The plug-in estimate is correct
in expectation under correct $Q$; the targeting step adds robustness
when $Q$ is misspecified but $g$ is consistent.
:::

### The analysis pipeline

```{mermaid}
flowchart TD
    A[Define target trial and interaction question] --> B[Pre-specify modifier and estimands]
    B --> C[Build DAG and adjustment set]
    C --> D[Diagnose overlap overall and within strata]
    D --> E[Specify Super Learner library]
    E --> F[Fit TMLE within each modifier stratum]
    F --> G[Derive subgroup RD, RR, and interaction contrasts]
    G --> H[Propagate uncertainty via influence function or bootstrap]
    H --> I[Report both scales with cell-specific risks]
```

### Stratified TMLE workflow: setup and data simulation

The remainder of this section walks through the **stratified TMLE**
implementation, which fits a separate TMLE within each modifier stratum
and then contrasts the subgroup-specific risk differences. It is
pedagogically transparent and matches the most common applied workflow.
For the joint-intervention version of the same analysis, see the
sl3/tmle3 sketch above.

```{r}
#| label: sim-data
#| eval: false

library(tmle)
library(SuperLearner)
library(dplyr)
library(boot)

set.seed(2026)

# Simulate a cohort of 4,000 patients
n  <- 4000
W1 <- rnorm(n)                                          # continuous confounder
W2 <- rbinom(n, 1, 0.4)                                 # binary confounder
W3 <- runif(n, -1, 1)                                   # bounded confounder

# Baseline modifier M depends on confounders
M  <- rbinom(n, 1, plogis(-0.2 + 0.7 * W1 - 0.4 * W2))

# Treatment depends on confounders and modifier
A  <- rbinom(n, 1, plogis(-0.1 + 0.8 * W1 + 0.5 * W2 -
                            0.3 * W3 + 0.2 * M))

# Outcome with genuine additive-scale effect modification by M
linpred_y <- -2.0 + 0.8 * A + 0.4 * M + 0.9 * A * M +
             0.7 * W1 - 0.5 * W2 + 0.3 * W3
Y <- rbinom(n, 1, plogis(linpred_y))

dat <- data.frame(Y, A, M, W1, W2, W3)
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"We built a synthetic cohort where the treatment really does work
differently for two kinds of patients. The simulation gives us a known
truth, so we can check whether our analysis recovers it."
:::

### Diagnose overlap within modifier strata

Before estimating any causal contrast, fit the propensity score within
each subgroup and inspect the distribution.

```{r}
#| label: overlap-check
#| eval: false

library(ggplot2)

dat$ps <- predict(
  glm(A ~ W1 + W2 + W3 + M, data = dat, family = binomial),
  type = "response"
)

ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ paste0("M = ", M)) +
  labs(x = "Estimated propensity score",
       fill = "Treatment",
       title = "Treatment overlap within each modifier stratum")
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"We are checking whether both treated and untreated patients exist
across the same range of risk in each subgroup. If one curve sits well
to the left or right of the other, the comparison is not well supported
by the data, no matter what statistical machinery we use next."
:::

### Estimate subgroup-specific risks with TMLE

We fit TMLE separately within each level of $M$. The Super Learner library
is small and diverse: an intercept-only baseline (`SL.mean`), a logistic
regression (`SL.glm`), a penalized regression (`SL.glmnet`), and a random
forest (`SL.ranger`).

```{r}
#| label: tmle-by-stratum
#| eval: false

sl_lib <- c("SL.mean", "SL.glm", "SL.glmnet", "SL.ranger")

estimate_tmle_by_modifier <- function(data, m_value) {
  d <- filter(data, M == m_value)

  fit <- tmle(
    Y = d$Y,
    A = d$A,
    W = d[, c("W1", "W2", "W3")],
    family = "binomial",
    Q.SL.library = sl_lib,
    g.SL.library = sl_lib
  )

  tibble(
    M            = m_value,
    risk_control = fit$estimates$EY0$psi,
    risk_treated = fit$estimates$EY1$psi,
    RD           = fit$estimates$ATE$psi,
    RR           = fit$estimates$RR$psi,
    RD_se        = sqrt(fit$estimates$ATE$var.psi),
    RR_logvar    = fit$estimates$RR$var.log.psi
  )
}

est <- bind_rows(
  estimate_tmle_by_modifier(dat, 0),
  estimate_tmle_by_modifier(dat, 1)
)

est
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"For each subgroup, we built two flexible models---one for who got
treated, one for who developed the outcome---and then combined them in a
way that adjusts for confounding. The result is the predicted average
risk if everyone in that subgroup had been treated, and the predicted
average risk if no one had been treated."
:::

### Derive interaction contrasts

```{r}
#| label: interaction-contrasts
#| eval: false

interaction_summary <- est %>%
  summarize(
    additive_interaction        = RD[M == 1] - RD[M == 0],
    multiplicative_int_log      = log(RR[M == 1]) - log(RR[M == 0]),
    multiplicative_int_ratio    = RR[M == 1] / RR[M == 0]
  )

interaction_summary
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"The additive interaction is just the gap between the two subgroup risk
differences. The multiplicative interaction ratio is the ratio of the
two subgroup risk ratios. They answer slightly different questions, and
both are easy to read off the cell-specific table."
:::

### Propagate uncertainty for the contrasts

The point estimate is straightforward; the standard error is not. Each
subgroup TMLE has an influence function, and the variance of the
difference of risk differences is a sum of subgroup variances (because the
strata are disjoint). For the multiplicative-scale contrast on the log
scale, the same logic applies. A nonparametric bootstrap is a defensible,
conceptually simple alternative for MPH-level work.

```{r}
#| label: bootstrap-ci
#| eval: false

boot_fun <- function(data, indices) {
  d  <- data[indices, ]
  e0 <- estimate_tmle_by_modifier(d, 0)
  e1 <- estimate_tmle_by_modifier(d, 1)
  c(
    additive_interaction   = e1$RD - e0$RD,
    multiplicative_int_log = log(e1$RR) - log(e0$RR)
  )
}

b <- boot(dat, statistic = boot_fun, R = 200)

boot_ci <- function(boot_obj, index) {
  out <- boot.ci(boot_obj, type = "perc", index = index)
  c(lower = out$percent[4], upper = out$percent[5])
}

list(
  additive_CI       = boot_ci(b, 1),
  multiplicative_CI = boot_ci(b, 2)
)
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"To get a confidence interval for the gap between subgroup effects, we
re-ran the whole analysis many times on resampled versions of the data.
The spread of those re-estimates tells us how stable the gap is."
:::

### Example output

A typical result table from this kind of analysis would look like the
following (numbers illustrative).

| Modifier stratum | Risk if untreated | Risk if treated | Risk difference | Risk ratio |
|---|---:|---:|---:|---:|
| M = 0 | 0.10 | 0.15 | 0.05 | 1.50 |
| M = 1 | 0.12 | 0.28 | 0.16 | 2.33 |

| Contrast | Estimate | 95% CI |
|---|---:|---:|
| Additive interaction (RD₁ − RD₀) | 0.11 | (0.05, 0.17) |
| Multiplicative interaction (RR₁ / RR₀) | 1.55 | (1.10, 2.20) |
: Table 3. Example TMLE output for the simulated cohort. {.striped}

### Illustrative interaction plot

```{mermaid}
xychart-beta
    title "Subgroup-specific risks from TMLE"
    x-axis ["M=0 untreated", "M=0 treated", "M=1 untreated", "M=1 treated"]
    y-axis "Estimated risk" 0 --> 0.30
    bar [0.10, 0.15, 0.12, 0.28]
```

The interpretation is concrete. Treatment is associated with a 5
percentage-point increase in risk among $M=0$ patients and a 16
percentage-point increase among $M=1$ patients. On the additive scale,
the treatment effect is 11 percentage points larger in $M=1$. On the
multiplicative scale, the risk ratio is roughly 1.5 times larger in
$M=1$. Both scales agree about the direction; only the additive scale
tells you the absolute size of the heterogeneity that matters for
clinical decision-making.

::: {.callout-note}
## What this means in practice
The cell-specific risk table is the single most useful artifact from an
interaction analysis. From it, any reader can recompute the additive
contrast, the multiplicative contrast, the number needed to treat in
each subgroup, and any other quantity. Lead with that table; let the
contrasts follow.
:::

---

## 7. CTMLE as a sensitivity analysis

Collaborative TMLE (CTMLE) modifies the targeting step so that the
propensity-score model is selected *with reference to the outcome model*,
rather than being fit independently. The motivation is practical: when
the propensity score includes variables that are unrelated to the outcome
but strongly related to treatment, the resulting weights become extreme
and the variance of the estimator inflates without a corresponding bias
benefit. CTMLE adaptively selects which covariates enter the
propensity-score model based on whether including them improves the
targeted bias-variance tradeoff.

For an introductory chapter, CTMLE is best framed as a **sensitivity
analysis** rather than the primary teaching estimator. The reasons are
twofold. First, the underlying ideas (collaborative selection, greedy
algorithms over propensity-score augmentations) are more abstract than
those needed for standard TMLE. Second, the `ctmle` package interface is
older than the modern `tmle` and `tmle3` interfaces, and package versions
can drift. The pedagogical value comes from understanding when CTMLE is
worth running---not from mastering its API on first contact.

### When to consider CTMLE

- The propensity-score model is unstable: a few patients have very high
  or very low fitted propensity, and dropping them changes the answer.
- Within-stratum positivity is borderline.
- A traditional TMLE produces extremely wide confidence intervals that
  appear driven by a small number of high-weight observations.

### Skeleton code

The following sketch follows the spirit of the `ctmle` package. **Check
the exact function signature against the installed package vignette**
before adapting it to your data; the API may differ across releases.

```{r}
#| label: ctmle-sketch
#| eval: false

# CTMLE sensitivity analysis on the same simulated cohort
# Confirm function names against the current ctmle vignette
library(ctmle)
library(dplyr)

estimate_ctmle_by_modifier <- function(data, m_value) {
  d <- filter(data, M == m_value)

  # ctmleDiscrete is the most common interface for binary A and Y
  # but exact arguments may vary by package version
  fit <- ctmleDiscrete(
    Y      = d$Y,
    A      = d$A,
    W      = d[, c("W1", "W2", "W3")],
    family = "binomial"
  )

  # Output extraction also varies; consult vignette
  tibble(
    M    = m_value,
    ate  = fit$est,
    note = "Verify output extraction against installed ctmle version"
  )
}

bind_rows(
  estimate_ctmle_by_modifier(dat, 0),
  estimate_ctmle_by_modifier(dat, 1)
)
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"We're using a related method that decides which covariates to include
in the treatment model only if they actually help reduce confounding for
the outcome. If this method gives a similar answer to our main TMLE, we
have more confidence that the result isn't being driven by a fragile
propensity-score model."
:::

::: {.callout-note}
## What this means in practice
Run CTMLE when standard TMLE looks fragile---most often when fitted
propensity scores have heavy tails inside one or more strata. Treat its
output as a *check* on the main analysis, not as a replacement for it.
For the great majority of MPH-level interaction analyses, standard TMLE
plus a curated Super Learner library is the appropriate primary
estimator.
:::

---

## 8. Time-varying treatment and longitudinal extensions

The point-treatment example assumes a single treatment decision at
baseline. Many real-world questions involve treatment regimens that
evolve over time, with intermediate covariates that are simultaneously
confounders and consequences of earlier treatment. In that setting,
baseline-only TMLE is no longer sufficient.

### When you need longitudinal methods

You need longitudinal targeted learning when at least one of the
following is true:

- Treatment can change after baseline (initiation, discontinuation,
  switching).
- Censoring is informative and is influenced by time-varying covariates.
- Time-varying confounders are themselves affected by past treatment
  (e.g., blood pressure during a hypertension trial).

For interactions specifically, the question now becomes: *Does the effect
of a sustained treatment strategy differ across baseline strata?* The
modifier $M$ should still be **baseline**, well-measured, and adequately
supported across treatment histories.

### LTMLE skeleton

`ltmle` implements longitudinal TMLE for sequential treatment and
censoring nodes. It accepts a Super Learner library and supports both
static and dynamic intervention rules.

```{r}
#| label: ltmle-sketch
#| eval: false

library(ltmle)

# Sketch only: the data 'long_dat' is in wide format with
# baseline covariates, time-varying covariates L1..L3,
# treatments A1..A3, and outcome Y at the final time

fit_ltmle <- ltmle(
  data       = long_dat,
  Anodes     = c("A1", "A2", "A3"),
  Lnodes     = c("L1", "L2", "L3"),
  Ynodes     = "Y",
  abar       = list(c(1, 1, 1), c(0, 0, 0)),
  SL.library = c("SL.mean", "SL.glm", "SL.glmnet", "SL.ranger")
)

summary(fit_ltmle)
```

To estimate baseline-modifier-specific effects, fit `ltmle` separately
within each level of $M$ and contrast the resulting counterfactual means.
Alternatively, use a TMLE for a marginal structural model (MSM) with $M$
entering as an effect modifier in the MSM. Either approach respects the
time-varying confounding while preserving the subgroup structure.

::: {.callout-tip}
## How to explain this to a non-methodologist
"When treatment can change month to month and intermediate health
markers both depend on past treatment and predict future treatment, a
single baseline analysis is no longer enough. The longitudinal version
of TMLE walks through the data one time step at a time, adjusting at
each step for what changed before."
:::

::: {.callout-note}
## What this means in practice
For sustained-treatment questions with subgroup contrasts, fit LTMLE
within each modifier stratum and report the cell-specific
counterfactual outcomes (e.g., one-year survival under "always treat"
versus "never treat" within each subgroup). Subgroup-specific dynamic
strategies require careful positivity diagnostics across treatment
histories.
:::

### Flexible estimands beyond binary $A$ and $E$

The estimands defined in Section 2 assume binary exposures. In real-world
evidence, exposures are often continuous (dose, duration, intensity), and
joint positivity for any single value of a continuous exposure is
typically zero. The remedy is to redefine the estimand in terms of
**modified treatment policies** (MTPs): hypothetical interventions that
*shift* observed exposure rather than fix it at a value
[@diaz2021lmtp]. The interaction question then becomes:

> *Does the effect of shifting $A$ depend on the level (or shift) of $E$?*

An MTP-based interaction parameter contrasts the counterfactual mean
under a paired shift in $(A, E)$ against the counterfactual mean under
shifting only $A$, only $E$, or neither. The `lmtp` package implements
TMLE and sequentially doubly robust estimators for these parameters, with
support for time-varying treatments, censoring, and Super Learner nuisance
estimation. For continuous-exposure interaction problems---common in
nutritional epidemiology, environmental health, and dose-response
pharmacoepidemiology---MTPs are the appropriate generalization of the
binary joint-intervention framework.

::: {.callout-tip}
## How to explain this to a non-methodologist
"When exposures are continuous, asking 'what if everyone had exactly this
dose' is unrealistic and unsupported by the data. Instead we ask 'what
if everyone shifted their exposure a bit?'---and the interaction question
becomes whether the benefit of shifting one exposure depends on the
shift in the other."
:::

---

## 9. Multiplicity, reporting, and interpretation

Interaction analyses are uniquely vulnerable to multiplicity. The risk
is not theoretical: many published claims of subgroup heterogeneity have
not replicated. Three disciplines protect the analysis.

### Pre-specification and the confirmatory-exploratory split

Before unblinding to the outcome, write down:

- **Confirmatory modifiers**: a small number (typically two to five),
  chosen for clinical or biological motivation. These get formal
  inference with appropriate multiplicity adjustment.
- **Exploratory modifiers**: any additional candidates. These are
  reported descriptively, with prominent caveats, and are best treated as
  hypothesis-generating.

The PATH statement and methodological work on heterogeneity of treatment
effects converge on this distinction. Recent simultaneous-inference
methods for multiple subgroup TMLEs make confirmatory analysis with
several modifiers tractable [@wei2023simultaneous].

### Multiplicity adjustment

For confirmatory work with several prespecified modifiers:

- Simultaneous (multivariate) confidence intervals around the vector of
  subgroup contrasts are the most informative single-step adjustment.
- False-discovery-rate control (Benjamini-Hochberg) is appropriate when
  the goal is to flag a subset of subgroups for further study.
- Bonferroni adjustment is conservative but easy to defend when the
  prespecified set is small.

### Reporting essentials

A well-reported interaction analysis includes, at minimum:

- The cell-specific risks or means with a single common reference
  category.
- Subgroup-specific risk differences and risk ratios with appropriate
  confidence intervals.
- The additive and multiplicative interaction contrasts with confidence
  intervals.
- Diagnostic information about overlap within strata.
- A brief description of the Super Learner library, the targeting step,
  and the variance estimator.
- Sensitivity analyses (e.g., CTMLE, alternative learner libraries,
  alternative confounder sets).
- An explicit statement about coding decisions and the reference
  category.

### Interpretation in absolute terms

Whenever the outcome scale permits, translate findings into absolute risk
or rate language. "The treatment reduces 5-year mortality by 8
percentage points among high-risk patients but only by 1 percentage
point among low-risk patients" is far more useful for clinical decisions
than "the relative-risk reduction is greater in high-risk patients."

### Common pitfalls

::: {.callout-warning}
## Common pitfalls in interaction analyses

- **Treating regression interaction coefficients as causal interaction.**
  A logistic or Cox product term is a property of the model's
  parameterization, not of the joint potential-outcome distribution.
- **Ignoring positivity in joint exposure cells.** A cohort that
  supports the main effect of $A$ may not support estimation of the
  $A \times E$ interaction; check the $(A, E)$ cell counts and the
  within-cell propensity distribution before interpreting any
  interaction estimate.
- **Choosing the scale post hoc.** Looking at both scales and reporting
  the one that gives a "significant" answer is a multiplicity violation
  in disguise. Pre-specify the scale, and report both.
- **Overinterpreting noisy estimates.** Interaction estimates are
  intrinsically noisier than main effects because they are differences
  of differences. A wide confidence interval is a finding, not a
  failure; the right response is to report it honestly, not to dichotomize
  on a p-value.
- **Confusing exploratory subgroup screening with confirmatory
  inference.** A subgroup that "popped out" of an exploratory scan must
  be replicated before it informs a clinical or policy decision.
:::

### Sensitivity analyses for interaction estimates

Because interaction estimates are derivatives of the outcome surface,
they are more sensitive to nuisance-model choices than main effects. A
short, prespecified sensitivity battery should accompany every
confirmatory interaction analysis:

- **Vary the nuisance learner library.** Re-run the analysis with a
  smaller library (e.g., GLM only) and a larger one (e.g., adding
  gradient boosting and neural-net learners). Stable estimates across
  libraries are reassuring; unstable estimates flag library or
  positivity sensitivity.
- **Vary the adjustment set.** Re-run with a minimal sufficient
  adjustment set and with a richer one. Large changes suggest residual
  confounding or model dependence.
- **Run a CTMLE sensitivity analysis** (Section 7) to check whether
  collaborative propensity-score selection moves the estimate.
- **Quantify unmeasured confounding tolerance** using a causal-gap or
  E-value-style framework: how strong would an unmeasured confounder
  need to be to explain away the observed interaction? This kind of
  quantitative bias analysis is especially valuable for interaction,
  where unmeasured confounders can act asymmetrically across exposure
  cells.

::: {.callout-note}
## What this means in practice
Pre-specify a short confirmatory list. Apply formal multiplicity control
to it. Report exploratory findings transparently, but with language that
makes their status clear. Translate every conclusion into absolute terms
at least once. Treat the sensitivity battery---learner library,
adjustment set, CTMLE, unmeasured-confounding bound---as part of the
primary analysis rather than as an afterthought.
:::

---

## 10. Software ecosystem and reproducible workflows

The R ecosystem for targeted learning is mature and well documented. For
an MPH chapter, the recommended core stack is `SuperLearner` for nuisance
estimation and `tmle` for point-treatment causal contrasts; longitudinal
work uses `ltmle`; aggressive learner libraries motivate `drtmle`; and
collaborative selection sits in `ctmle`. The modular tlverse (`tmle3`,
`sl3`, and friends) is the modern, extensible alternative to the older
`tmle` package and is recommended once readers are comfortable with the
basic workflow.

| Package | Scope | Best use here | Caution |
|---|---|---|---|
| `SuperLearner` | CRAN ensemble framework | Nuisance estimation throughout | Curate the library; do not use defaults blindly |
| `tmle` | CRAN; point-treatment TMLE for binary or continuous outcomes | Primary teaching example for binary treatment and modifier | Less modular than `tmle3` |
| `tmle3` | tlverse modular TMLE framework | Custom estimands, advanced workflows | Steeper learning curve; tlverse not all on CRAN |
| `sl3` | tlverse Super Learner | Modern learner specification within `tmle3` | Less familiar to most MPH readers |
| `ltmle` | CRAN; longitudinal TMLE | Time-varying treatment and censoring | More complex data structure |
| `drtmle` | CRAN; doubly robust nonparametric estimation | Inference under aggressive nuisance learners | More technical |
| `ctmle` | CRAN; collaborative TMLE | Sensitivity analysis for fragile propensity scores | Older API; check vignette per version |
| `lmtp` | CRAN; modified treatment policies | Continuous or shift-style exposures | Beyond first interaction chapter |
: Table 4. R packages for targeted-learning interaction analyses. {.striped}

### Reproducibility checklist

A reproducible interaction analysis includes:

- A fully documented analytic dataset (with data dictionary).
- A scripted analysis pipeline (including data preparation, overlap
  diagnostics, estimation, and output tables).
- A fixed random seed.
- A `sessionInfo()` block reporting R, package, and operating-system
  versions.
- A pre-registered or pre-specified analysis plan stored alongside the
  code.

::: {.callout-note}
## What this means in practice
Use `tmle` plus `SuperLearner` for the main analysis. Add `ltmle` if the
treatment is time-varying. Add `drtmle` if your learner library is
aggressive. Reserve `ctmle` for sensitivity analysis. Keep the
reproducibility scaffolding (seed, session info, scripts) attached to
every chapter of the analysis.
:::

---

## 11. Practical checklist for analysts

The checklist below summarizes the chapter into a single artifact. It can
be copied into a project's pre-analysis plan or a reviewer's referee
form.

| Step | Recommendation | Common failure mode |
|---|---|---|
| Question | State whether you study effect modification of one treatment or interaction of two exposures | Treating any product term as "interaction" |
| Estimand | Pre-specify additive and multiplicative estimands separately | Reporting only odds-ratio interaction |
| Design | Write a target trial protocol; define time zero, censoring, follow-up | Building subgroup models without a protocol |
| Modifiers | Limit confirmatory analyses to a small set of clinically motivated, pre-treatment modifiers | Mining many post-hoc modifiers |
| Confounding | Adjust for confounders of treatment and outcome only; avoid post-treatment variables | Overadjustment and collider bias |
| Positivity | Diagnose overlap overall *and within each modifier stratum* | Stratum-specific positivity failures hidden by overall overlap |
| Nuisance | Use a Super Learner library that is diverse but appropriate for the sample size | One opaque learner with no diagnostics |
| Estimation | TMLE within strata for small numbers of subgroups; simultaneous one-step TMLE for many | Naive stratified regression |
| Inference | Influence-function-based SEs; bootstrap for derived contrasts; cross-fitting for aggressive learners | Wald tests on derived contrasts assuming independence |
| Multiplicity | Confirmatory vs. exploratory split; simultaneous CIs or FDR for many modifiers | Many unadjusted subgroup tests |
| Reporting | Cell-specific risks; both scales; common reference category; coding decisions; sensitivity analyses | Only a p-value for the product term |
| Interpretation | Lead with absolute risks; sense-check against background incidence | Headline-only relative-risk language |
: Table 5. Practical checklist for interaction analyses in RWE. {.striped}

---

## 12. Conclusion

Interaction analyses in real-world evidence research are easy to do
poorly and demanding to do well. The path forward is not a single new
estimator, but a shift in workflow: start with a clearly defined causal
estimand on a chosen scale, build a target trial protocol around it, use
flexible nuisance estimation through Super Learner, target the resulting
estimates with TMLE, propagate uncertainty for the contrasts you report,
and discipline yourself to distinguish confirmatory from exploratory
work. The targeted-learning toolkit aligns naturally with these
disciplines, supports both binary point-treatment and time-varying
extensions, and---through CTMLE and `drtmle`---offers principled
sensitivity options when nuisance estimation is fragile.

### Minimum reporting for interaction analyses in RWE papers

At a minimum, a published interaction analysis in real-world evidence
should report: (1) a precisely worded scientific question and the chosen
estimand; (2) the target trial protocol (eligibility, time zero,
treatment strategies, follow-up, censoring, outcome); (3) the
prespecified set of confirmatory modifiers and the exploratory set, with
multiplicity strategy; (4) cell-specific risks under treatment and
control (or under each joint exposure level), with a common reference
category; (5) subgroup-specific risk differences and risk ratios with
confidence intervals; (6) additive and multiplicative interaction
contrasts with confidence intervals; (7) overlap diagnostics within
strata; (8) the Super Learner library and the variance estimator used;
(9) at least one sensitivity analysis, ideally including a CTMLE or
alternative-library check; and (10) an explicit interpretation in
absolute-risk language. With these elements in place, a reader can
reconstruct the estimand, judge the design, and assess the credibility
of the heterogeneity claim.

### Looking ahead: from interactions to treatment rules

Heterogeneity in treatment effect is also the raw material for
*personalization* and for *adaptive treatment*. The conceptual chain
runs: **interaction → heterogeneity → optimal rules → dynamic treatment
regimes**. Once an interaction analysis has identified subgroups in
which treatment is more or less beneficial, the natural next questions
are (a) whether a static rule assigning treatment as a function of
baseline characteristics improves population-level outcomes, and (b)
whether a regime that *adapts* treatment to evolving patient state
improves outcomes further. This connects the present chapter to several
downstream topics covered elsewhere in the handbook:

- **Optimal individualized treatment rules** ([Chapter 4.1](04-01-optimal-regimes.qmd)),
  which use targeted learning (e.g., `tmle3mopttx`) to estimate the
  baseline-covariate rule that maximizes expected outcome under
  intervention.
- **Dynamic treatment regimes**, in which the treatment assigned at
  each follow-up time depends on the patient's history. Interaction
  estimates are the diagnostic input that justifies a non-trivial
  dynamic regime: if there is no meaningful heterogeneity, a static
  "always treat" or "never treat" policy is optimal.
- **Policy learning**, which extends the same logic to constrained,
  cost-aware, or fairness-aware decision rules deployed at scale.

Heterogeneous effects motivate treatment rules; treatment rules
operationalize the heterogeneity; dynamic regimes extend the
operationalization across time. Readers who finish this chapter and
want to see how subgroup contrasts translate into actionable static or
adaptive policies should read the optimal-regimes chapter next, then
the longitudinal LTMLE material for the time-varying generalization.

---

## Recommended reading

- VanderWeele and Knol [-@vanderweele2014tutorial]. The single best
  concise reference on additive versus multiplicative interaction and
  reporting standards.
- Petersen and van der Laan [-@petersen2014causal]. The causal roadmap
  applied to epidemiologic practice.
- Hernán and Robins [-@hernan2016target]. Target trial emulation as the
  organizing principle for RWE design.
- Talbot et al. [-@talbot2025tmle]. Modern best-practice tutorial for
  TMLE in applied work.
- van der Laan and Rubin [-@vanderLaan2006tmle]. Original TMLE paper.
- van der Laan, Polley, and Hubbard [-@vanderLaan2007superlearner]. Original
  Super Learner paper.
- Smith et al. [-@smith2023systematic]. Systematic review documenting
  the rapid uptake of TMLE in observational epidemiology.
- Wei et al. [-@wei2023simultaneous]. Simultaneous TMLE for multiple
  subgroup effects.
- Basten et al. [-@basten2025additive]. Recent treatment of additive
  interaction in IPD meta-analysis.

## Annotated references

- **van der Laan and Rubin (2006).** Introduces TMLE as a plug-in
  estimator that is updated to solve the efficient influence function
  equation. The conceptual foundation for everything else in this
  chapter.
- **van der Laan, Polley, and Hubbard (2007).** Defines Super Learner as
  cross-validation-based stacking with an oracle inequality. Justifies
  using a diverse learner library without overcommitting to any single
  algorithm.
- **Petersen and van der Laan (2014).** A pedagogical formulation of the
  causal roadmap for epidemiologists. Recommended for any first reading
  on causal inference workflow.
- **VanderWeele and Knol (2014).** A tutorial that has shaped the modern
  reporting consensus on interaction. Especially clear on RERI, the
  reference-category problem, and scale dependence.
- **Hernán and Robins (2016).** The target-trial paper that has
  reorganized observational research around explicit protocol emulation.
  Required background for any RWE interaction analysis.
- **Smith et al. (2023).** A systematic review of TMLE in observational
  epidemiology. Useful for situating your analysis within current
  practice.
- **Talbot et al. (2025).** A best-practice tutorial that walks through
  TMLE implementation choices with applied researchers in mind.
- **Wei et al. (2023).** Simultaneous one-step TMLE for many subgroup
  effects, with valid joint confidence intervals. Read when you have
  more than a handful of modifiers.
- **Basten et al. (2025).** A clear recent treatment of additive
  interaction estimation, including in pooled / IPD settings. Useful
  for analysts whose decision question is public-health targeting.

---

## Appendix A: Full R code

The block below collects the chapter's main worked example into a single
runnable script. Comments indicate the role of each step.

```{r}
#| label: appendix-a-full
#| eval: false

# ============================================================
# Appendix A. Interactions in RWE: TMLE + Super Learner
# Companion script to Chapter on interactions in RWE
# ============================================================

# --- Packages ---------------------------------------------------------
library(tmle)
library(SuperLearner)
library(dplyr)
library(boot)
library(ggplot2)

set.seed(2026)

# --- 1. Simulate cohort ------------------------------------------------
n  <- 4000
W1 <- rnorm(n)                               # continuous confounder
W2 <- rbinom(n, 1, 0.4)                      # binary confounder
W3 <- runif(n, -1, 1)                        # bounded confounder
M  <- rbinom(n, 1, plogis(-0.2 + 0.7*W1 - 0.4*W2))   # baseline modifier
A  <- rbinom(n, 1, plogis(-0.1 + 0.8*W1 + 0.5*W2 -
                            0.3*W3 + 0.2*M))         # treatment
linpred_y <- -2.0 + 0.8*A + 0.4*M + 0.9*A*M +
             0.7*W1 - 0.5*W2 + 0.3*W3
Y <- rbinom(n, 1, plogis(linpred_y))

dat <- data.frame(Y, A, M, W1, W2, W3)

# --- 2. Overlap diagnostic --------------------------------------------
dat$ps <- predict(
  glm(A ~ W1 + W2 + W3 + M, data = dat, family = binomial),
  type = "response"
)

p_overlap <- ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ paste0("M = ", M)) +
  labs(x = "Estimated propensity score",
       fill = "Treatment",
       title = "Treatment overlap within each modifier stratum")
print(p_overlap)

# --- 3. TMLE within each modifier stratum ------------------------------
sl_lib <- c("SL.mean", "SL.glm", "SL.glmnet", "SL.ranger")

estimate_tmle_by_modifier <- function(data, m_value) {
  d <- filter(data, M == m_value)
  fit <- tmle(
    Y = d$Y,
    A = d$A,
    W = d[, c("W1", "W2", "W3")],
    family = "binomial",
    Q.SL.library = sl_lib,
    g.SL.library = sl_lib
  )
  tibble(
    M            = m_value,
    risk_control = fit$estimates$EY0$psi,
    risk_treated = fit$estimates$EY1$psi,
    RD           = fit$estimates$ATE$psi,
    RR           = fit$estimates$RR$psi,
    RD_se        = sqrt(fit$estimates$ATE$var.psi),
    RR_logvar    = fit$estimates$RR$var.log.psi
  )
}

est <- bind_rows(
  estimate_tmle_by_modifier(dat, 0),
  estimate_tmle_by_modifier(dat, 1)
)
print(est)

# --- 4. Interaction contrasts ------------------------------------------
interaction_summary <- est %>%
  summarize(
    additive_interaction      = RD[M == 1] - RD[M == 0],
    multiplicative_int_log    = log(RR[M == 1]) - log(RR[M == 0]),
    multiplicative_int_ratio  = RR[M == 1] / RR[M == 0]
  )
print(interaction_summary)

# --- 5. Bootstrap CI for derived contrasts -----------------------------
boot_fun <- function(data, indices) {
  d  <- data[indices, ]
  e0 <- estimate_tmle_by_modifier(d, 0)
  e1 <- estimate_tmle_by_modifier(d, 1)
  c(
    additive_interaction   = e1$RD - e0$RD,
    multiplicative_int_log = log(e1$RR) - log(e0$RR)
  )
}

b <- boot(dat, statistic = boot_fun, R = 200)

boot_ci <- function(boot_obj, index) {
  out <- boot.ci(boot_obj, type = "perc", index = index)
  c(lower = out$percent[4], upper = out$percent[5])
}

results <- list(
  subgroup_estimates       = est,
  interaction_summary      = interaction_summary,
  additive_CI              = boot_ci(b, 1),
  multiplicative_log_CI    = boot_ci(b, 2),
  session                  = sessionInfo()
)
print(results)

# --- 6. CTMLE sensitivity analysis (skeleton) --------------------------
# NOTE: confirm the function signature against the installed
#       ctmle vignette before running on real data.
#
# library(ctmle)
# estimate_ctmle_by_modifier <- function(data, m_value) {
#   d <- filter(data, M == m_value)
#   fit <- ctmleDiscrete(
#     Y      = d$Y,
#     A      = d$A,
#     W      = d[, c("W1", "W2", "W3")],
#     family = "binomial"
#   )
#   tibble(M = m_value, ate = fit$est)
# }
# bind_rows(
#   estimate_ctmle_by_modifier(dat, 0),
#   estimate_ctmle_by_modifier(dat, 1)
# )

# --- 7. LTMLE sketch for time-varying treatment ------------------------
# library(ltmle)
# fit_ltmle <- ltmle(
#   data       = long_dat,
#   Anodes     = c("A1", "A2", "A3"),
#   Lnodes     = c("L1", "L2", "L3"),
#   Ynodes     = "Y",
#   abar       = list(c(1, 1, 1), c(0, 0, 0)),
#   SL.library = c("SL.mean", "SL.glm", "SL.glmnet", "SL.ranger")
# )
# summary(fit_ltmle)
```

---

## Appendix B: Optional Python supplement

Python is supplementary, not central, to a targeted-learning workflow.
The most epidemiology-oriented option is `zEpid`, which provides risk
ratios and risk differences, interaction contrasts, and TMLE-style
estimators in an interface designed for epidemiologic practice.

```{python}
#| label: appendix-b-python
#| eval: false

# Optional Python supplement
# zEpid for effect measures and TMLE-style epidemiologic workflows

import pandas as pd
from zepid import RiskRatio, RiskDifference
from zepid.causal.doublyrobust import TMLE

# Assume df has columns: Y (outcome), A (treatment),
# M (modifier), W1, W2, W3 (confounders)

# Risk difference within each modifier stratum
for m_value in [0, 1]:
    d = df[df["M"] == m_value]
    rd = RiskDifference()
    rd.fit(d, exposure="A", outcome="Y")
    print(f"M = {m_value}")
    rd.summary()

# TMLE within each modifier stratum
for m_value in [0, 1]:
    d = df[df["M"] == m_value].copy()
    tmle = TMLE(d, exposure="A", outcome="Y")
    tmle.exposure_model("W1 + W2 + W3")
    tmle.outcome_model("A + W1 + W2 + W3")
    tmle.fit()
    print(f"M = {m_value}: ATE = {tmle.average_treatment_effect}")
```

`DoWhy` and `CausalML` are useful complementary tools but are not primary
targeted-learning implementations. `DoWhy` provides causal-graph
specification, identification, and refutation tests; `CausalML` focuses
on conditional average treatment effects and uplift modeling, which are
better viewed as exploratory heterogeneity tools than as substitutes for
TMLE-based interaction inference. For an MPH chapter on interactions in
RWE, the R targeted-learning stack remains the recommended primary
implementation; the Python ecosystem is a useful complement.
