---
title: "L-TMLE By Hand: A Technical Companion"
subtitle: "Full step-by-step implementation for two time points"
format: html
---

*Estimated reading time: 30 minutes*

::: {.callout-note}
## Optional / advanced chapter
This chapter provides the full by-hand L-TMLE implementation for readers who want to see exactly how the algorithm works at the code level. It uses the same two-time-point HIV simulation from the preceding conceptual chapter. **Most applied readers can skip this chapter** and proceed to the LMTP chapters, which provide the recommended practical workflow for realistic longitudinal interventions. Return here if you need to understand the internals of the algorithm.
:::

```{r}
#| include: false
library(tidyverse)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
logit <- function(p) log(p / (1 - p))
```


# Setup: simulated data

We use the same two-time-point HIV cohort from the [conceptual L-TMLE chapter](03-09-illustrated-ltmle.qmd). The data encodes treatment-confounder feedback: ART adherence at visit 1 ($A_1$) raises CD4 ($L_1$), which affects visit-2 adherence ($A_2$) and virologic failure ($Y$).

```{r}
set.seed(2026)
n <- 2000

W_age    <- rnorm(n, mean = 40, sd = 10)
W_male   <- rbinom(n, 1, 0.6)
W_cd4_bl <- pmax(50, rnorm(n, mean = 350, sd = 120))
W_vl_bl  <- rnorm(n, mean = 4.2, sd = 0.7)
W <- 0.01 * (W_age - 40) - 0.001 * (W_cd4_bl - 350) + 0.3 * W_vl_bl - 0.1 * W_male

g1_true <- plogis(0.5 + 0.8 * W)
A1 <- rbinom(n, 1, g1_true)
L1 <- rnorm(n, mean = 0.5 * W + 0.7 * A1, sd = 1)
g2_true <- plogis(-0.3 + 0.6 * L1 + 0.4 * A1)
A2 <- rbinom(n, 1, g2_true)
Y <- rbinom(n, 1, plogis(-0.5 - 0.5 * A1 - 0.6 * A2 - 0.3 * L1 + 0.4 * W))

dat <- tibble(W, A1, L1, A2, Y)
```

```{r}
# Oracle truth (separate seed, large MC sample)
set.seed(9999)
n_mc <- 100000
W_mc <- rep(W, length.out = n_mc)
L1_cf11 <- rnorm(n_mc, mean = 0.5 * W_mc + 0.7, sd = 1)
Y_cf11  <- plogis(-0.5 - 0.5 - 0.6 - 0.3 * L1_cf11 + 0.4 * W_mc)
true_EY_11 <- mean(Y_cf11)
cat("True E[Y(1,1)]:", round(true_EY_11, 4), "\n")
```


# Phase A: Estimate the treatment mechanisms

At each time point, estimate the probability of receiving treatment conditional on the observed history.

```{r}
# Time 1: P(A1 = 1 | W)
g1_mod <- glm(A1 ~ W, family = binomial, data = dat)
g1_hat <- pmax(0.01, pmin(0.99, predict(g1_mod, type = "response")))

# Time 2: P(A2 = 1 | A1, L1, W)
g2_mod <- glm(A2 ~ L1 + A1 + W, family = binomial, data = dat)
g2_hat <- pmax(0.01, pmin(0.99, predict(g2_mod, type = "response")))

# Evaluate g2 under the intervention value A1=1
g2_under_a1 <- pmax(0.01, pmin(0.99,
  predict(g2_mod, newdata = dat %>% mutate(A1 = 1), type = "response")))

# Cumulative probability of following "always adhere"
g_cumulative <- g1_hat * g2_under_a1
cat("Cumulative g range:", round(range(g_cumulative), 3), "\n")
```


# Phase B: Sequential outcome regression (backwards)

## Step 2a: Model the final outcome

Fit $E[Y \mid A_2, L_1, A_1, W]$, conditioning on $L_1$ to deconfound $A_2$. Then predict under $A_2 = 1$.

```{r}
Q2_mod <- glm(Y ~ A2 + L1 + A1 + W + A1:A2, family = binomial, data = dat)
Q2_A2is1 <- predict(Q2_mod, newdata = dat %>% mutate(A2 = 1), type = "response")
cat("Q2(A2=1) range:", round(range(Q2_A2is1), 4), "\n")
```

## Step 2b: Move backwards (the critical step)

Regress the time-2 predictions on $(A_1, W)$ only, **without** conditioning on $L_1$. This integrates $L_1$ out, preserving the mediated effect of $A_1$ through $L_1$.

```{r}
Q1_mod <- glm(Q2_A2is1 ~ A1 + W, family = quasibinomial, data = dat)
Q1_A1is1 <- predict(Q1_mod, newdata = dat %>% mutate(A1 = 1), type = "response")
cat("G-computation estimate E[Y(1,1)]:", round(mean(Q1_A1is1), 4), "\n")
cat("True E[Y(1,1)]:                  ", round(true_EY_11, 4), "\n")
```

This is longitudinal G-computation. It depends on all models being correctly specified. The targeting steps below add robustness.


# Phase C: Targeting for double robustness

## Target at time 2

The clever covariate at time 2 is the inverse cumulative probability of following the full regime:

$$
H_2 = \frac{I(A_1 = 1, A_2 = 1)}{\hat{g}_1(W) \times \hat{g}_2(1, L_1, W)}
$$

```{r}
H2 <- (dat$A1 == 1 & dat$A2 == 1) / (g1_hat * g2_under_a1)
QA_init_t2 <- predict(Q2_mod, type = "response")

# Fluctuation model
fluc2 <- glm(dat$Y ~ -1 + offset(logit(QA_init_t2)) + H2, family = binomial)
eps2 <- coef(fluc2)

# Update Q2 predictions under A2=1
Q2_star <- plogis(logit(Q2_A2is1) + eps2 / (g1_hat * g2_under_a1))
cat("Epsilon at time 2:", round(eps2, 5), "\n")
```

The coefficient $\hat{\epsilon}_2$ measures how much the initial outcome model needs correcting. A small value means the outcome model was already well-calibrated for the causal estimand.

## Target at time 1

```{r}
Q1_mod_updated <- glm(Q2_star ~ A1 + W, family = quasibinomial, data = dat)
Q1_updated <- predict(Q1_mod_updated, type = "response")
Q1_under_a1 <- predict(Q1_mod_updated, newdata = dat %>% mutate(A1 = 1), type = "response")

H1 <- (dat$A1 == 1) / g1_hat

fluc1 <- glm(Q2_star ~ -1 + offset(logit(Q1_updated)) + H1, family = quasibinomial)
eps1 <- coef(fluc1)

Q1_star <- plogis(logit(Q1_under_a1) + eps1 / g1_hat)
cat("Epsilon at time 1:", round(eps1, 5), "\n")
```


# The final L-TMLE estimate

```{r}
ltmle_est_11 <- mean(Q1_star)
cat("L-TMLE estimate E[Y(1,1)]:", round(ltmle_est_11, 4), "\n")
cat("True E[Y(1,1)]:           ", round(true_EY_11, 4), "\n")
```


# Comparing estimators

```{r}
naive_mod <- glm(Y ~ A1 + A2 + L1 + W, family = binomial, data = dat)
naive_pred_11 <- mean(predict(naive_mod, newdata = dat %>% mutate(A1 = 1, A2 = 1), type = "response"))

iptw_w <- ifelse(dat$A1 == 1 & dat$A2 == 1, 1 / (g1_hat * g2_hat), 0)
iptw_est <- sum(iptw_w * dat$Y) / sum(iptw_w)

comparison <- tibble(
  Method = c("True E[Y(1,1)]", "Naive regression", "G-computation", "IPTW", "L-TMLE"),
  Estimate = round(c(true_EY_11, naive_pred_11, mean(Q1_A1is1), iptw_est, ltmle_est_11), 4)
)
comparison |> as.data.frame()
```

```{r}
#| label: fig-estimator-comparison
#| fig-cap: "Comparison of estimators for the counterfactual failure rate under sustained high adherence."
ggplot(comparison, aes(x = Estimate, y = factor(Method, levels = rev(Method)))) +
  geom_point(size = 3) +
  geom_vline(xintercept = true_EY_11, linetype = "dashed", color = "red") +
  annotate("text", x = true_EY_11, y = 5.4, label = "Truth", color = "red", size = 3) +
  labs(x = "Estimated E[Y(1,1)]", y = "") +
  theme_minimal()
```


# Sequential double robustness

At each time point, L-TMLE is consistent if either the outcome model or the treatment model is correctly specified:

| Correct at time 2 | Correct at time 1 | L-TMLE consistent? |
|---|---|---|
| $Q_2$ only | $Q_1$ only | Yes |
| $g_2$ only | $g_1$ only | Yes |
| $Q_2$ only | $g_1$ only | Yes |
| $g_2$ only | $Q_1$ only | Yes |
| Neither | Anything | Not guaranteed |


# Scaling to T time points

With $T$ time points, the same pattern applies:

1. Estimate $\hat{g}_t$ for $t = 1, \ldots, T$.
2. Start at time $T$: fit $\hat{Q}_T$, target with $H_T$.
3. Move to $T-1$: fit $\hat{Q}_{T-1}$ using targeted $\hat{Q}_T^*$ as outcome, target with $H_{T-1}$.
4. Repeat until $\hat{Q}_1^*$.
5. Final estimate: $\hat{\psi} = \frac{1}{n}\sum_i \hat{Q}_1^*(a_1, W_i)$.

Each backwards step conditions on $L_t$ where it confounds future treatment, and integrates it out where it mediates past treatment.


## Sources

- Petersen ML, Schwab J, Gruber S, et al. (2014). Targeted maximum likelihood estimation for dynamic and static longitudinal marginal structural working models. *J Causal Inference* 2(2):147-185.
- van der Laan MJ, Gruber S (2012). Targeted minimum loss based estimation of causal effects of multiple time point interventions. *Int J Biostat* 8(1):Article 9.
- Lendle SD, Schwab J, Petersen ML, van der Laan MJ (2017). ltmle: An R package implementing targeted minimum loss-based estimation for longitudinal data. *J Stat Softw* 81(1):1-21.
