---
title: "Longitudinal TMLE: Concepts and Intuition"
subtitle: "How backwards sequential regression solves treatment-confounder feedback"
format: html
---

*Estimated reading time: 30 minutes*

::: {.callout-note}
## Learning objectives
- Understand the backwards sequential regression logic that underpins L-TMLE
- Explain why conditioning on the time-varying confounder is needed at one time step but harmful at another
- Describe the role of targeting (fluctuation steps) in the longitudinal setting
- Recognize when static regimes violate positivity and when LMTP offers a better framework
:::

```{r}
#| include: false
#| code-fold: true
library(dagitty)
library(ggdag)
library(ggplot2)
library(tidyverse)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
logit <- function(p) log(p / (1 - p))
```


# Where this chapter fits

The previous chapter explained why treatment-confounder feedback breaks standard regression. This chapter explains how L-TMLE solves the problem conceptually. The focus is on the logic of the algorithm, not a full code walkthrough. Readers who want the complete by-hand implementation can consult the [technical companion chapter](03-09b-ltmle-by-hand.qmd). For practical applied analysis of realistic longitudinal interventions, the LMTP chapters that follow are the recommended pathway.


# The causal question

We use a two-time-point HIV cohort to make the mechanics visible. The question: what is the probability of virologic failure if every patient maintained high ART adherence at each visit, compared to never maintaining high adherence?

This is a comparison of two static regimes, $\bar{a} = (1,1)$ versus $\bar{a} = (0,0)$, with treatment-confounder feedback through CD4 count. We use two time points for clarity; the same logic extends to any number of visits.

```{r}
#| label: fig-hiv-node-order
#| fig-cap: "Node ordering for the two-time-point HIV cohort."
#| fig-width: 9
#| fig-height: 3.5
#| code-fold: true
dag_hiv <- dagify(
  A1 ~ W, L1 ~ A1 + W, A2 ~ L1 + A1 + W, Y ~ A1 + A2 + L1 + W,
  exposure = "A1", outcome = "Y",
  labels = c(W = "Baseline\n(age, sex, CD4_0, VL_0)",
             A1 = "ART adherence\nvisit 1",
             L1 = "CD4 count\nvisit 1",
             A2 = "ART adherence\nvisit 2",
             Y = "Virologic\nfailure"),
  coords = list(
    x = c(W = 0, A1 = 1, L1 = 2, A2 = 3, Y = 4),
    y = c(W = 1, A1 = 0, L1 = 0.8, A2 = 0, Y = 0)
  )
)
ggdag(dag_hiv, text = FALSE, use_labels = "label", text_size = 2.8) +
  theme_dag()
```


# Simulated data

```{r}
set.seed(2026)
n <- 2000

W_age    <- rnorm(n, mean = 40, sd = 10)
W_male   <- rbinom(n, 1, 0.6)
W_cd4_bl <- pmax(50, rnorm(n, mean = 350, sd = 120))
W_vl_bl  <- rnorm(n, mean = 4.2, sd = 0.7)
W <- 0.01 * (W_age - 40) - 0.001 * (W_cd4_bl - 350) + 0.2 * W_vl_bl - 0.1 * W_male

A1 <- rbinom(n, 1, plogis(0.3 + 0.5 * W))
L1 <- rnorm(n, mean = 0.3 * W + 1.2 * A1, sd = 1)  # strong treatment -> confounder
A2 <- rbinom(n, 1, plogis(-0.5 + 0.8 * L1 + 0.3 * A1))  # strong confounder -> treatment
Y  <- rbinom(n, 1, plogis(-0.8 - 0.4 * A1 - 0.5 * A2 - 0.6 * L1 + 0.2 * W))  # strong L1 effect

dat <- tibble(W, A1, L1, A2, Y)
cat("Failure rate:", round(mean(Y), 3), "\n")
```


# The backwards regression idea

L-TMLE estimates the causal effect through three ideas working together.

**Sequential regressions, working backwards.** Start at the final outcome and model $E[Y \mid A_2, L_1, A_1, W]$. This conditions on $L_1$ to deconfound $A_2$. Then move backwards: take the predicted outcomes under $A_2 = 1$ and regress them on $(A_1, W)$ only, without conditioning on $L_1$. This integrates $L_1$ out, preserving the mediated effect of $A_1$ through $L_1$.

**Treatment mechanism models.** Estimate the probability of receiving treatment at each visit conditional on history: $\hat{g}_1(W)$ and $\hat{g}_2(A_1, L_1, W)$.

**Targeting steps.** Use the treatment mechanism to construct clever covariates that correct the sequential regressions for any remaining bias. At each time point, the final estimate is consistent if either the outcome regression or the treatment model is correctly specified (sequential double robustness).

The key insight: conditioning on $L_1$ at time 2 deconfounds $A_2$, while NOT conditioning on $L_1$ at time 1 preserves the indirect effect $A_1 \to L_1 \to Y$. Standard regression cannot do both in a single model.


# How the recipe works in plain language

**Step 1.** At each visit, model how likely each patient was to actually adhere. Sicker patients are less likely to adhere; patients who already missed doses are less likely to catch up.

**Step 2.** Starting at visit 2, predict each patient's failure risk as if they had high adherence. Adjust using the treatment model so that the prediction is calibrated for the causal estimand, not just for prediction accuracy. This is the targeting step.

**Step 3.** Take the adjusted visit-2 predictions and use them as the "outcome" for visit 1. Predict under high adherence at visit 1, and target again. The visit-1 model does not condition on CD4, because CD4 is a consequence of visit-1 adherence.

**Step 4.** Average the final targeted predictions across all patients. This is the L-TMLE estimate of $E[Y(1,1)]$.


# Naive regression vs L-TMLE

A quick demonstration of why the backwards structure matters:

```{r}
# Oracle truth
set.seed(9999)
n_mc <- 100000; W_mc <- rep(W, length.out = n_mc)
L1_cf11 <- rnorm(n_mc, mean = 0.3 * W_mc + 1.2, sd = 1)
true_EY_11 <- mean(plogis(-0.8 - 0.4 - 0.5 - 0.6 * L1_cf11 + 0.2 * W_mc))

# Naive: condition on L1 (blocks A1 -> L1 -> Y)
naive_mod <- glm(Y ~ A1 + A2 + L1 + W, family = binomial, data = dat)
naive_pred <- mean(predict(naive_mod, newdata = dat %>% mutate(A1 = 1, A2 = 1), type = "response"))

cat("True E[Y(1,1)]:", round(true_EY_11, 4), "\n")
cat("Naive (conditions on L1):", round(naive_pred, 4), "\n")
```

The naive estimate is biased because conditioning on $L_1$ blocks part of the causal pathway. L-TMLE avoids this by conditioning on $L_1$ only where it confounds ($A_2$) and integrating it out where it mediates ($A_1$).

::: {.callout-note}
## Full by-hand implementation
The complete L-TMLE algorithm, with code for treatment mechanism estimation, sequential outcome regression, clever covariates, fluctuation models, and estimator comparison, is in the [technical companion chapter](03-09b-ltmle-by-hand.qmd). That chapter walks through every line of the two-time-point implementation.
:::


# When static regimes break down

The "always adhere" regime is pedagogically useful, but it has a practical limitation: no real intervention achieves 100% adherence. When the static regime assigns treatment values that are rare in the observed data, positivity breaks down: propensity scores approach zero, clever covariates explode, and the estimate becomes unstable.

This is not a failure of L-TMLE as a method. It is a limitation of the estimand. Asking "what if everyone achieved perfect adherence?" requires extrapolating far outside the data for patients who are very unlikely to adhere. The result is high variance and potential bias.

The solution is to define more realistic interventions that stay within the observed data support. The LMTP framework provides this: instead of forcing a specific treatment value, it shifts the treatment process by a feasible amount. The [LMTP chapters](03-03-longitudinal-case-study.qmd) develop this approach.


# Package implementation

The `ltmle` R package automates the full L-TMLE procedure. For static regimes, the interface is straightforward:

```{r}
#| eval: false
#| code-fold: true
library(ltmle)

dat_ltmle <- data.frame(W = dat$W, A1 = dat$A1, L1 = dat$L1, A2 = dat$A2, Y = dat$Y)

result <- ltmle(
  data = dat_ltmle,
  Anodes = c("A1", "A2"),
  Lnodes = c("L1"),
  Ynodes = "Y",
  abar = list(treatment = c(1, 1), control = c(0, 0)),
  SL.library = c("SL.glm", "SL.mean")
)

summary(result)
```

For realistic interventions estimated with `lmtp`, see the [LMTP case study](03-03-longitudinal-case-study.qmd) and the [applied workflow chapter](05-01-applied-lmtp-workflow.qmd).


## Check your understanding

<details>
<summary><strong>Q1. Why does L-TMLE work backwards through time?</strong></summary>

Working backwards allows L-TMLE to condition on $L_t$ where it confounds future treatment and integrate it out where it mediates past treatment. A forward approach cannot do both simultaneously.

</details>

<details>
<summary><strong>Q2. What is sequential double robustness?</strong></summary>

At each time point, L-TMLE is consistent if either the outcome model or the treatment model is correctly specified. Different models can fail at different time points, as long as at each point at least one is correct.

</details>

<details>
<summary><strong>Q3. When should you use LMTP instead of static L-TMLE?</strong></summary>

When the static regime (e.g., "always adhere") assigns treatment values that are rare in the observed data. LMTP defines feasible shifts that stay within the data support, producing more stable estimates and answering more clinically relevant questions.

</details>

---

## Sources and further reading

- Petersen ML, Schwab J, Gruber S, et al. (2014). Targeted maximum likelihood estimation for dynamic and static longitudinal marginal structural working models. *J Causal Inference* 2(2):147-185.
- Lendle SD, Schwab J, Petersen ML, van der Laan MJ (2017). ltmle: An R package implementing targeted minimum loss-based estimation for longitudinal data. *J Stat Softw* 81(1):1-21.
- `ltmle` R package: [CRAN](https://cran.r-project.org/package=ltmle)
