---
title: "Chapter 3.3: Case Study – Real-World Application in Longitudinal Analysis"
format: html
---

# Case Study: Real-World Longitudinal Causal Inference Using Targeted Learning

Parts I through III of this handbook focused on point-treatment settings: a single treatment decision at one moment in time. But real-world treatment trajectories are rarely that simple. Patients start, stop, and switch medications. Their health evolves, and those changes influence both future treatment decisions and outcomes. This part of the handbook addresses longitudinal causal inference, and we begin with a complete applied case study.

This chapter walks through a **complete applied example** of longitudinal causal inference using real-world data concepts. The goal is to help you translate the earlier theoretical chapters into a practical analysis workflow.

We use the motivating real-world question:

> **What is the 3-year cardiovascular risk difference if all eligible patients initiating osteoporosis therapy remained on denosumab vs. zoledronic acid under full adherence?**

Although we cannot use the proprietary data from the actual studies, this chapter recreates a *plausible longitudinal structure* and walks you through the entire pipeline:

- Constructing the longitudinal dataset  
- Defining the causal question, intervention, and estimand  
- Identifying longitudinal confounders and censoring  
- Using SuperLearner for nuisance function estimation  
- Applying TMLE or LMTP for longitudinal causal effects  
- Interpreting the results in a regulatory and clinical context  

This chapter integrates lessons from Chapter 1 (causal roadmap) and Chapter 2 (estimation).

::: {.callout-note}
## Learning objectives
- Construct a longitudinal dataset suitable for causal analysis from wide and long formats
- Define a causal estimand for a longitudinal treatment strategy with time-varying confounders
- Identify and handle censoring, treatment switching, and loss to follow-up in the data structure
- Apply L-TMLE or LMTP to estimate the effect of a sustained treatment strategy
- Interpret longitudinal causal effect estimates in a clinical and regulatory context
:::

*This chapter is educational. Causal conclusions depend on identification assumptions (e.g., consistency, exchangeability, positivity) and on diagnostic evidence that the data support the target estimand. When flexible machine learning is used for nuisance estimation, valid inference typically requires cross-fitting or a cross-validated TMLE variant, plus appropriate rate conditions.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)
```

---

# 1. The Clinical and Real-World Context

Osteoporosis treatments such as **denosumab** and **zoledronic acid** are widely used in postmenopausal women. But real-world treatment trajectories are messy. Observational data registries show:

- Frequent **treatment interruptions**
- **Switching** between agents
- **Loss to follow-up**
- Time-varying covariates (e.g., frailty, kidney function, fall risk)

These complexities mean that a single baseline snapshot cannot capture the full causal story. Patients change treatments over time, their risk factors evolve, and those evolving risk factors influence both future treatment decisions and the outcome. **Key point.** Standard regression on baseline data alone will produce biased estimates when treatment and confounders co-evolve over time. This is why longitudinal causal inference is necessary.

:::{.callout-warning}
## Why a Naive Cox Model Fails Here

**Pitfall.** If you adjust for time-varying confounders in a Cox model, you **block causal pathways** (overadjustment bias). If you do not adjust, you have **uncontrolled confounding**. Longitudinal TMLE and LMTP are designed to solve this dilemma.
:::

We instead use **Longitudinal TMLE** or **LMTP** to estimate causal effects under hypothetical longitudinal interventions.

---

# 2. The Longitudinal Data Structure

:::{.callout-note}
## The Canonical Longitudinal Data Structure

**Definition.** Each patient's observed data is an ordered sequence:

$$
O_i = (W_i, A_{i0}, L_{i1}, A_{i1}, C_{i1}, L_{i2}, A_{i2}, C_{i2}, \dots, Y_i)
$$

- **$W$**: baseline covariates (age, CVD history, frailty)
- **$A_t$**: treatment status at time $t$ (1 = denosumab, 0 = zoledronic acid)
- **$L_t$**: time-varying covariates (kidney function, fracture risk, comorbidities)
- **$C_t$**: censoring/loss to follow-up (1 = censored)
- **$Y$**: cardiovascular event by 36 months

At each time point, covariates ($L_t$) are observed first, then treatment ($A_t$), then censoring ($C_t$). This ordering encodes our causal assumptions.
:::

We simulate a small example below.

---

# 3. Simulation of a Plausible Longitudinal Dataset

```{r}
library(tidyverse)

set.seed(2027)
n <- 2000
Tmax <- 6   # six timepoints for illustration

# Baseline
W_age  <- rnorm(n, 75, 6)
W_cvd  <- rbinom(n, 1, plogis(0.12 * (W_age - 70)))

# Initial treatment
A0 <- rbinom(n, 1, plogis(-1 + 0.1*(W_age - 70) + 1.4*W_cvd))

# Containers
L <- matrix(NA, n, Tmax)
A <- matrix(NA, n, Tmax)
C <- matrix(0, n, Tmax)

A[,1] <- A0

# Generate longitudinal covariates & treatment
for (t in 1:Tmax) {
  # time-varying comorbidity
  L[,t] <- rnorm(n, mean = 0.3*W_age + 1.2*W_cvd + 0.5*A[,max(1,t-1)], sd = 1)

  # treatment switching
  A[,t] <- rbinom(n, 1, plogis(-2 + 0.3*L[,t] + 0.8*A[,max(1,t-1)]))

  # censoring
  C[,t] <- rbinom(n, 1, plogis(-4 + 0.2*L[,t] + 0.4*A[,t]))
}

# Outcome depends on full history
Y <- rbinom(n, 1, plogis(-3 + 0.5*rowMeans(A) + 0.1*rowMeans(L)))

dat_long <- tibble(
  id = 1:n,
  age = W_age,
  cvd = W_cvd,
  Y = Y
)

# Expand wide data into long format (illustrative)
for (t in 1:Tmax) {
  dat_long[[paste0("L",t)]] <- L[,t]
  dat_long[[paste0("A",t)]] <- A[,t]
  dat_long[[paste0("C",t)]] <- C[,t]
}

glimpse(dat_long)
```

The code above creates **2,000 hypothetical patients** each observed over **6 time points**:

1. **Baseline**: Each patient starts with an age (centered around 75) and a baseline CVD history that depends on age.
2. **Initial treatment ($A_0$)**: Whether a patient starts on denosumab depends on age and CVD history (confounding by indication).
3. **At each time point**: A time-varying comorbidity score ($L_t$) evolves based on age, CVD history, *and the prior treatment*, the hallmark of **time-varying confounding affected by prior treatment**. Treatment can then switch based on the current comorbidity score.
4. **Censoring**: Dropout probability depends on current comorbidity and treatment, making censoring **informative**.
5. **Outcome**: The final cardiovascular event depends on the *entire treatment history* and the *entire covariate history*.

**Key point.** This is exactly the kind of data structure where naive methods break down: treatment affects future confounders, which affect future treatment, which affects the outcome.

---

# 4. Defining the Longitudinal Causal Question

:::{.callout-note}
## Static Treatment Regimes

**Definition.** A **static treatment regime** assigns the same treatment at every visit regardless of patient history. We compare two regimes:

- $d_1$: $A_t = 1$ for all $t$ (always denosumab)
- $d_0$: $A_t = 0$ for all $t$ (always zoledronic acid)

The target estimand is the 36-month cardiovascular event risk difference under perfect adherence to each regime.
:::

This is the longitudinal analogue of the per-protocol effect from RCTs, estimated from observational data using methods that handle time-varying confounding. **Key point.** We can also define **dynamic regimes** (treatment depends on evolving patient characteristics) or **stochastic regimes** (shift treatment probabilities), but static interventions keep the exposition clear.

---

# 5. Identification: Assumptions for Longitudinal Effects

:::{.callout-warning}
## Identification Assumptions: Sequential Exchangeability

As discussed in [Chapter 1.2](01-02-causal-roadmap.qmd), identification requires consistency, exchangeability, and positivity. In the longitudinal setting, these assumptions must hold at every time point (sequential versions):

- **Sequential exchangeability:** $Y^{\bar{a}} \perp A_t \mid \bar{L}_t, \bar{A}_{t-1}, W$ for all $t$. Conditional on the full history, treatment is "as good as random."
- **Sequential positivity:** Each treatment must remain possible at every time point for every covariate history.
- **Consistency:** The observed outcome under regime $\bar{a}$ equals the potential outcome under intervention $\bar{a}$.

**Practical guidance:** In your DAG and protocol, explicitly list which time-varying confounders you condition on and which might be unmeasured. Sensitivity analyses (Section 9) can help assess robustness.
:::

---

# 6. Using LMTP for Longitudinal Interventions

:::{.callout-tip}
## What LMTP Does in Plain Language

**Key idea.** The **lmtp** R package estimates counterfactual outcomes under hypothetical treatment strategies that unfold over time. You specify a treatment rule (static, dynamic, or stochastic), and LMTP uses doubly robust estimation with SuperLearner to adjust for time-varying confounding, treatment switching, and informative censoring, all in a single automated pipeline.
:::

Here we implement:

```{r, eval=F}
# Example skeleton 
 library(lmtp)

result <- lmtp_tmle(
  data = dat_long,
  trt = paste0("A", 1:Tmax),
  outcome = "Y",
  time = 1:Tmax,
  shift = function(data, trt) { rep(1, length(trt)) },  # always-denosumab
  learners_outcome = list(SL.glm, SL.ranger),
  learners_trt = list(SL.glm, SL.ranger),
  folds = 5
)

result
```

This gives a TMLE estimate of:

- Risk under always-denosumab  
- Risk under always-ZA  
- Risk difference  
- Confidence intervals  

---

# 7. Using Longitudinal TMLE (Manual Concept)

:::{.callout-tip}
## The Backward Regression Intuition

**Key idea.** Longitudinal TMLE works **backward from the outcome**: at each time step it (1) estimates the expected outcome given the current history, (2) estimates the treatment/censoring probability, and (3) applies a targeting update that makes the final estimate doubly robust. This backward iteration is automated by LMTP, but understanding the intuition helps you diagnose problems (e.g., propensity scores near 0 or 1 at a given step signal a positivity violation).
:::

---

# 8. Interpretation of Results

:::{.callout-tip}
## Interpreting Longitudinal Causal Estimates

Assume hypothetical results:

| Regime | Estimated 36-mo risk |
|--------|----------------------|
| Always denosumab | 0.051 |
| Always ZA        | 0.055 |

**Key idea.** The 0.4 percentage-point risk difference is clinically small. These estimates reflect a **per-protocol effect** (what would happen under perfect adherence), not what is observed in practice. Subgroup and sensitivity analyses should explore heterogeneity and robustness to assumption violations.
:::

---

# 9. Sensitivity & Diagnostic Considerations

:::{.callout-warning}
## Positivity Violations in Longitudinal Settings

**Pitfall.** Positivity must hold *at every time point* for *every observed covariate history*, a much stronger requirement than in single-timepoint studies. Plot estimated propensity scores at each visit; if scores cluster near 0 or 1 at later time points, you likely have a structural positivity violation. LMTP and L-TMLE handle extreme weights better than IPTW (substitution vs. inverse weighting), but you should still examine influence-function values for outliers and consider density-ratio truncation for sparse treatment histories.
:::

Beyond positivity, you should also assess whether unmeasured confounding might bias your primary analysis.

:::{.callout-important}
## Negative Control Outcomes

**Practical implementation.** Run your entire LMTP pipeline substituting a **negative control outcome**, one that *should not* be affected by the treatment (e.g., accidental injury rates). If the estimated effect is close to the null, that is reassuring. A substantial effect suggests residual confounding that may also bias your primary analysis.
:::

---

# 10. Summary

:::{.callout-tip}
## Chapter Takeaways

- **Build longitudinal data** using the canonical ordering $O = (W, A_0, L_1, A_1, C_1, \dots, Y)$
- **Define static interventions** as hypothetical "always treat" regimes to formalize the causal question
- **Apply LMTP/TMLE** for doubly robust estimation that handles time-varying confounding, informative censoring, and treatment switching
- **Interpret results** as per-protocol causal effects, distinct from observed associational quantities

**The bottom line:** When treatment, confounders, and censoring co-evolve over time, standard regression and Cox models produce biased estimates. Longitudinal TMLE and LMTP provide a principled framework, automated by the `lmtp` R package.
:::

## Check your understanding

<details>
<summary><strong>Q1. Why can't you simply add time-varying confounders as covariates in a Cox regression when those confounders are affected by prior treatment?</strong></summary>

Because time-varying confounders that are affected by prior treatment sit on the causal pathway between earlier treatment and the outcome. Adjusting for them in a standard regression **blocks part of the treatment effect** (overadjustment bias), while omitting them leaves **uncontrolled confounding**. This is the fundamental "treatment-confounder feedback" problem. Longitudinal TMLE and LMTP solve it by using sequential modeling that properly separates confounding adjustment from causal-pathway preservation at each time point.

</details>

<details>
<summary><strong>Q2. What does "sequential exchangeability" require that ordinary exchangeability does not?</strong></summary>

Ordinary (point-treatment) exchangeability requires that treatment is unconfounded conditional on baseline covariates measured at a single time point. **Sequential exchangeability** requires this to hold **at every time point**, conditional on the entire covariate and treatment history up to that point. In practice, this means you must measure every confounder that influences both treatment decisions and the outcome *at every visit*, a much harder bar to clear, especially when physician judgment, symptom severity, or patient preferences evolve over time and may not be captured in the data.

</details>

<details>
<summary><strong>Q3. The LMTP estimate compares "always denosumab" vs "always ZA": is this an intention-to-treat or per-protocol effect?</strong></summary>

It is a **per-protocol effect** (more precisely, a "treatment-policy" or "sustained-strategy" effect under full adherence). Unlike the intention-to-treat (ITT) effect, which estimates the impact of initial treatment *assignment* regardless of subsequent adherence, the LMTP estimate asks what *would* happen if every patient remained on the assigned treatment at every visit. This is the clinically relevant question when you want to know the causal effect of the treatment itself, but it requires the additional assumption of sequential exchangeability, which is not needed for ITT analyses in randomised trials.

</details>

## What comes next

This case study applied longitudinal methods to a single treatment comparison. The next part of the handbook broadens the scope to advanced applications: combining RCT and real-world data, pre-specified regulatory analyses, mediation decomposition, and time-to-event estimands.

```{r}
sessionInfo()
```

---

## Sources and further reading

- Petersen ML, Schwab J, Gruber S, Blaser N, Schomaker M, van der Laan MJ (2014). Targeted maximum likelihood estimation for dynamic and static longitudinal marginal structural working models. *J Causal Inference* 2(2):147-185.
- Lendle SD, Schwab J, Petersen ML, van der Laan MJ (2017). ltmle: An R package implementing targeted minimum loss-based estimation for longitudinal data. *J Stat Softw* 81(1):1-21.
- Dang LE, Tager I, Petersen ML (2023). A causal roadmap for generating high-quality real-world evidence. *J Clin Transl Sci* 7(1):e127.
- Diaz I, Williams N, Hoffman KL, et al. (2023). lmtp: An R package for estimating the causal effects of modified treatment policies. *Observational Studies* 9:103-122.
- van der Laan MJ, Gruber S (2012). Targeted minimum loss based estimation of causal effects of multiple time point interventions. *Int J Biostat* 8(1):Article 9.
- `ltmle` R package: [CRAN](https://cran.r-project.org/package=ltmle)
- `lmtp` R package: [CRAN](https://cran.r-project.org/package=lmtp)

---

## Software Implementation (R)

This example uses the `ltmle` package to estimate the effect of a **time-varying treatment** on a binary outcome, mirroring the longitudinal case study structure (denosumab vs. zoledronic acid).

- Simulate 2-time-point data: baseline $W$, time-varying confounder $L_1$, treatment at two times $A_0, A_1$, and outcome $Y$
- $L_1$ is affected by $A_0$ (treatment-confounder feedback)
- Use `ltmle()` to estimate the ATE of "always treat" vs. "never treat"
- Falls back to a manual sequential regression if `ltmle` is unavailable

```{r}
#| eval: false
set.seed(1)
n <- 500
W     <- rnorm(n)
A_0   <- rbinom(n, 1, plogis(0.3 * W))
L_1   <- rnorm(n, mean = 0.5 * A_0 + 0.3 * W)  # affected by A_0
A_1   <- rbinom(n, 1, plogis(0.2 * L_1 + 0.3 * W))
Y     <- rbinom(n, 1, plogis(-1.5 + 0.5 * A_0 + 0.5 * A_1 + 0.4 * L_1 + 0.3 * W))

dat <- data.frame(W = W, A_0 = A_0, L_1 = L_1, A_1 = A_1, Y = Y)

if (requireNamespace("ltmle", quietly = TRUE)) {
  library(ltmle)

  result <- ltmle(
    data = dat,
    Anodes  = c("A_0", "A_1"),
    Lnodes  = "L_1",
    Ynodes  = "Y",
    abar = list(treatment = c(1, 1), control = c(0, 0)),
    SL.library = "glm"
  )
  cat("LTMLE estimate (always treat vs never treat):\n")
  print(summary(result))
} else {
  message("Install the 'ltmle' package:  install.packages('ltmle')")
  message("Falling back to manual sequential regression...\n")

  ## Manual sequential g-computation (no targeting)
  Q_mod <- glm(Y ~ A_0 + A_1 + L_1 + W, family = binomial, data = dat)
  dat1 <- dat; dat1$A_0 <- 1; dat1$A_1 <- 1
  dat0 <- dat; dat0$A_0 <- 0; dat0$A_1 <- 0
  gcomp <- mean(predict(Q_mod, dat1, type = "response")) -
           mean(predict(Q_mod, dat0, type = "response"))
  cat("G-computation ATE (no targeting):", round(gcomp, 3), "\n")
}
```
