---
subtitle: "From adherence thresholds to intervention-defined forgiveness curves"
format: html
---

# Adherence, Forgiveness, and Distribution-Shift Interventions {#sec-adherence-shift}

*Estimated reading time: 55 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
library(tidyverse)
```


# What this chapter is about

The previous chapters introduced longitudinal TMLE for static interventions ("always maintain high adherence") and LMTP for modified treatment policies. This chapter brings those tools to bear on a clinically important question: how do two antiretroviral drugs compare when we account for realistic adherence patterns? It focuses on the adherence- and forgiveness-specific shift methods; for the full end-to-end LMTP workflow on a dynamic treatment rule, from question to estimate, see @sec-applied-lmtp.

The key concept is **regimen forgiveness**: a drug is forgiving if patients can miss some doses without a large increase in treatment failure. A drug with a flat adherence-response curve is forgiving; one with a steep curve is not. Forgiveness is a causal quantity, not an observable property of the data, because the adherence-response relationship is confounded by time-varying health status.

We use a simulated two-drug HIV cohort designed to capture the essential features of the adherence problem: treatment-confounder feedback between adherence and health, differential forgiveness across drugs, and the positivity challenges that arise when targeting extreme adherence levels. The simulation is calibrated to real pharmacoepidemiologic settings and uses continuous proportion of days covered (PDC) as the adherence measure.


The figure below previews the quantity we are after. A **forgiveness curve** plots the counterfactual risk of virologic failure against the adherence level a patient sustains, one curve per drug. The vertical distance between the two curves is the **forgiveness gap**: how much more a missed-dose pattern costs on the less forgiving regimen. The curve is a causal object, the risk we *would* see if adherence were set to each level, and it is not the same as the risk we observe among patients who happen to adhere at that level. Those two can diverge sharply, and most of this chapter is about why, and how to estimate the causal version.

```{r}
#| label: fig-forgiveness-motivation
#| fig-cap: "An illustrative forgiveness curve. Counterfactual failure risk rises as sustained adherence falls, and it rises faster for the less forgiving regimen. The vertical distance is the forgiveness gap. This is the causal quantity the chapter estimates; the curve here is a stylized illustration, not yet an estimate from data."
#| code-fold: true
library(ggplot2)
adh <- seq(0.30, 1, length.out = 150)
data.frame(
  adherence = rep(adh, 2),
  risk = c(plogis(-3 + 4.5 * (1 - adh)),
           plogis(-3 + 4.5 * (1 - adh) - 0.8 * (1 - adh))),
  Drug = rep(c("Drug B (less forgiving)", "Drug A (more forgiving)"), each = 150)
) |>
  ggplot(aes(adherence, risk, color = Drug)) +
  geom_line(linewidth = 1.2) +
  annotate("segment", x = 0.5, xend = 0.5, y = 0.16, yend = 0.55,
           linetype = "dashed", color = "grey40") +
  annotate("text", x = 0.52, y = 0.36, label = "forgiveness\ngap",
           hjust = 0, size = 3.4, color = "grey40") +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(values = c("Drug B (less forgiving)" = "#EE6677",
                                "Drug A (more forgiving)" = "#228833")) +
  labs(x = "Sustained adherence (PDC)", y = "Virologic failure risk", color = NULL) +
  theme_minimal() + theme(legend.position = "bottom")
```


# The clinical question

Consider two antiretroviral regimens for HIV treatment:

- **Drug A** (e.g., a long-acting formulation): pharmacologically forgiving, meaning that occasional missed doses have a smaller impact on viral suppression.
- **Drug B** (e.g., a short half-life oral regimen): less forgiving, where missed doses lead to faster viral rebound.

In practice, both drugs are prescribed and adherence varies. The adherence distributions may differ across drugs (one may be easier to take consistently), and the clinical consequences of low adherence differ (one may be more forgiving). The causal question is: **how much of the observed outcome difference between drugs is due to adherence patterns versus pharmacological forgiveness?**

This question requires separating three components:

1. **Adherence distribution**: Do Drug A and Drug B patients adhere at different rates?
2. **Forgiveness**: At any given adherence level, does one drug produce better outcomes than the other?
3. **Adherence-mediated effect**: How much of the total drug difference operates through the adherence pathway?

Each component requires longitudinal causal methods because adherence and health co-evolve over time [@petersen2006msm].


# Why this needs a causal method

Adherence is measured after treatment starts and changes over time, and it sits in a feedback loop with health [@rosenblum2009virologic]. Worse health lowers adherence through side effects and complications; lower adherence worsens virologic control; and worse control feeds back into health, which shapes the next period's adherence. The variable that confounds the adherence-to-failure relationship, current health status $L_t$, is therefore also a *consequence* of past adherence.

```{mermaid}
%%| label: fig-adherence-dag
%%| fig-cap: "Three blocks of the adherence process (K = 3). Baseline severity W0 channels the drug choice A0 (confounding by indication, the arrow W0 to A0). Health status L at each block confounds the adherence-to-failure relationship and is itself moved by prior adherence (the feedback arrows PDC1 to L2 and PDC2 to L3); this is treatment-confounder feedback. The drug A0 modifies the adherence-to-failure relationship at every block (forgiveness). A regression that conditions on L removes the confounding but also blocks the pathway through which past adherence acts; a regression that omits L leaves the confounding in place."
flowchart LR
  W0["W0<br/>baseline severity"] --> A0["A0<br/>drug"]
  W0 --> L1["L1<br/>health"]
  A0 --> Y["Y<br/>failure"]
  A0 --> P1["PDC1<br/>adherence"]
  L1 --> P1
  P1 --> Y
  L1 --> Y
  P1 --> L2["L2<br/>health"]
  W0 --> L2
  L2 --> P2["PDC2<br/>adherence"]
  P2 --> Y
  L2 --> Y
  P2 --> L3["L3<br/>health"]
  W0 --> L3
  L3 --> P3["PDC3<br/>adherence"]
  P3 --> Y
  L3 --> Y
```

This is the dilemma that defeats ordinary regression. Conditioning on observed adherence adjusts for the health-status confounding and, at the same time, blocks the adherence effect we want to measure, because that effect runs through health. Leaving health out of the model leaves the confounding in place. No single regression can do both. This is exactly the treatment-confounder-feedback setting of @sec-longitudinal-confounding, and it is why estimating a forgiveness curve needs the g-methods developed there.


# The data-generating process

We use a simulation that generates longitudinal data in 90-day blocks (K = 3 blocks). The DGP produces treatment-confounder feedback: poor adherence worsens health (higher L), and worse health reduces future adherence. Drug assignment is confounded by baseline severity (confounding by indication: sicker patients are channeled toward Drug A), which is the study's primary scenario; setting `kappaW = 0` recovers a randomized trial. Drug A is more forgiving at low adherence levels, encoded through a drug-by-adherence interaction in the failure hazard.

The code below reproduces the study's `functions/dgp.R` at its default parameters, simplified for teaching: the latent adherence residual is i.i.d. (the study's default `rho = 0`), and the optional censoring, AR(1) persistence, differential-tolerability, and extra-confounding levers are omitted. Confounding by indication is on.

```{r}
#| label: sim-dgp
# Longitudinal DGP for the adherence-forgiveness simulation (K = 3 blocks).
# Causal structure each block t:  W0, A0 -> L_t -> PDC_t -> Y_t,
# with feedback  PDC_{t-1} -> L_t -> PDC_t  (time-dependent confounding).
# Parameters and equations follow functions/dgp.R at its defaults.
simulate_adherence_data <- function(n = 2000, K = 3, seed = 123, kappaW = 1.0) {
  set.seed(seed)

  # --- Parameters (study defaults) -------------------------------------------
  # Adherence (PDC) on the logit scale
  alpha0 <- 1.30    # intercept; plogis(1.30) is about 0.79
  alphaA <- -0.35   # Drug A shifts adherence slightly lower
  alphaW <- -0.20   # worse baseline severity -> lower adherence
  alphaL <- -0.30   # worse interim health -> lower next-block adherence
  sigma_u <- 0.30   # between-person adherence-trait SD
  sigma_e <- 0.65   # within-person per-block noise SD (i.i.d.; study default rho = 0)

  # Time-varying health confounder L_t
  beta0   <- 0.00
  betaW   <- 0.30   # baseline severity -> worse health
  betaLag <- 0.35   # health persistence
  betaP   <- 0.80   # prior non-adherence -> worse health (the feedback arrow)
  sigma_L <- 0.50

  # Virologic-failure hazard (logit)
  gamma0  <- -3.80  # low baseline hazard
  gammaL  <- 0.50   # sicker -> higher failure hazard
  gammaW  <- 0.30   # baseline severity -> higher hazard
  gammaP  <- 2.50   # poor adherence -> higher hazard
  gammaA  <- 0.00   # no main drug effect (drugs equal at perfect adherence)
  gammaAP <- -1.20  # Drug A x poor-adherence: protective => Drug A is MORE FORGIVING

  cut_low  <- 0.60; cut_high <- 0.80   # adherence-band cutpoints (low/mid/high)

  # --- Baseline --------------------------------------------------------------
  W0  <- rnorm(n)                              # baseline severity
  # Confounding by indication: sicker patients channeled to Drug A.
  # kappaW = 0 reproduces a randomized trial (A0 ~ Bernoulli(0.5)).
  A0  <- rbinom(n, 1, plogis(kappaW * W0))     # 1 = Drug A, 0 = Drug B
  u_i <- rnorm(n, 0, sigma_u)                  # persistent adherence trait (unmeasured)

  # --- Storage (rows = patients, cols = blocks) ------------------------------
  L   <- PDC <- matrix(NA_real_, n, K)
  Y   <- matrix(0L, n, K)
  failed <- rep(FALSE, n)                      # absorbing failure flag

  for (t in 1:K) {
    # Time-varying health L_t
    if (t == 1) {
      L[, t] <- beta0 + betaW * W0 + rnorm(n, 0, sigma_L)
    } else {
      L[, t] <- beta0 + betaW * W0 +
                betaLag * L[, t - 1] +
                betaP   * (1 - PDC[, t - 1]) +     # prior non-adherence worsens health
                rnorm(n, 0, sigma_L)
    }

    # Adherence PDC_t in (0, 1)
    eta <- alpha0 + alphaA * A0 + alphaW * W0 + u_i + rnorm(n, 0, sigma_e)
    if (t > 1) eta <- eta + alphaL * L[, t - 1]    # interim health feeds back into adherence
    PDC[, t] <- plogis(eta)

    # Virologic failure (absorbing): poor adherence raises hazard, less so on Drug A
    at_risk <- !failed
    lp <- gamma0 +
          gammaL  * L[at_risk, t] +
          gammaW  * W0[at_risk] +
          gammaP  * (1 - PDC[at_risk, t]) +
          gammaA  * A0[at_risk] +
          gammaAP * A0[at_risk] * (1 - PDC[at_risk, t])
    new_fail <- rbinom(sum(at_risk), 1, plogis(lp)) == 1L
    idx <- which(at_risk)
    Y[idx[new_fail], t] <- 1L
    failed[idx[new_fail]] <- TRUE
    if (t > 1) Y[Y[, t - 1] == 1L, t] <- 1L        # carry failure forward (absorbing)
  }

  # --- Assemble wide data frame ---------------------------------------------
  dat <- data.frame(id = 1:n, W0 = W0, A0 = A0)
  for (t in 1:K) {
    dat[[paste0("L_",   t)]] <- L[, t]
    dat[[paste0("PDC_", t)]] <- PDC[, t]
    dat[[paste0("Y_",   t)]] <- Y[, t]
    dat[[paste0("PDC_cat_", t)]] <- cut(PDC[, t],
                                        breaks = c(-Inf, cut_low, cut_high, Inf),
                                        labels = c("low", "mid", "high"))
  }
  dat$Y_final <- as.integer(rowSums(Y) > 0)        # failed by end of follow-up
  attr(dat, "K") <- K
  dat
}

dat <- simulate_adherence_data(n = 2000, K = 3, seed = 123, kappaW = 1.0)
```

The key structural features of this DGP:

**Confounding by indication.** Drug assignment depends on baseline severity through `A0 ~ Bernoulli(plogis(kappaW * W0))`: sicker patients (higher `W0`) are more likely to start Drug A. Setting `kappaW = 0` reproduces a randomized trial in which `A0` is a coin flip independent of severity.

**Treatment-confounder feedback.** Poor adherence (low PDC) at block $t-1$ worsens health ($L_t$), which then reduces adherence at block $t$ (through `alphaL`) and independently increases failure risk (through `gammaL`). This is the same feedback structure discussed in @sec-longitudinal-confounding.

**Drug-by-adherence interaction (the forgiveness mechanism).** The parameter `gammaAP = -1.20` means that Drug A's penalty for poor adherence is smaller than Drug B's. At high adherence, the two drugs perform similarly. At low adherence, Drug A is more forgiving. This interaction is what makes the forgiveness curve differ between drugs.

For `lmtp`'s survival mode the data need one further preparation step: once a patient fails, the post-failure treatment and confounder nodes are set to `NA` and the failure indicator is carried forward as 1 (the absorbing state). In the study this is handled by the `make_lmtp_data()` helper in `functions/prepare_lmtp.R`; we reference it here rather than re-deriving it inline.

Before turning to methods, here is the cohort the simulation produced, summarized by drug.

```{r}
#| label: cohort-summary
dat |>
  dplyr::mutate(Drug = ifelse(A0 == 1, "Drug A", "Drug B")) |>
  dplyr::group_by(Drug) |>
  dplyr::summarise(
    n = dplyr::n(),
    mean_PDC_1   = round(mean(PDC_1), 3),
    failure_pct  = round(100 * mean(Y_final), 1),
    .groups = "drop"
  ) |>
  as.data.frame()
```

Read this table as a description of the observed cohort, not as a forgiveness estimate. Taken at face value, Drug B looks like the better regimen here: its patients have higher mean adherence and a lower crude failure percentage than Drug A's. That crude ranking is misleading, and in fact it inverts the underlying pharmacology, because Drug A is the more forgiving drug by construction (the `gammaAP = -1.20` interaction). The reason is confounding by indication: sicker patients (higher `W0`) are channeled toward Drug A, so the Drug A column is enriched for patients who were always at higher risk and who adhere less well, for reasons that have nothing to do with the drug itself. The two drug groups are not exchangeable, so any crude difference in adherence or failure between the columns mixes the drugs' pharmacology with the difference in who received each drug. The crude cross-drug comparison is therefore confounded and is not the forgiveness estimand; recovering forgiveness, and seeing Drug A's advantage emerge, requires the causal machinery developed below.

# Why naive adherence analysis is biased

Forgiveness is an **intervention-based estimand**, not a within-stratum association. The question is what the failure risk *would be* if adherence were set to a given level, with health allowed to respond, not what the failure rate happens to be among patients who adhered at that level.

That distinction is what defeats ordinary regression here, because of treatment-confounder feedback. Conditioning on observed adherence does remove the health confounding, but it also blocks the *mediated* part of the adherence effect, the part that runs through interim health $L_t$, so the within-stratum comparison answers a different question than the one we asked. Omitting interim health from the model leaves the confounding in place instead. No single regression escapes the bind: adjust for $L_t$ and you block the pathway; drop it and you reintroduce the confounding.

The longitudinal g-formula, realized here as a modified treatment policy (MTP), is built precisely for this setting. It intervenes on the adherence sequence and propagates the consequences forward through the health process, so the mediated effect is preserved and the confounding is still controlled. The practical lesson is to name the estimand before choosing the adjustment.

We can make the bias visible because the data come from a known simulation. The causal (oracle) forgiveness curve is obtained by re-running the failure process with adherence held fixed at each level and health allowed to respond; the crude curve is the observed failure rate among patients binned by their actual average adherence. Overlaying the two shows the gap.

```{r}
#| label: fig-oracle-vs-crude
#| fig-cap: "Causal (oracle) forgiveness curves, as solid lines, versus crude adherence-stratified failure rates, as open points. The crude points sit above the causal curve at low adherence and below it at high adherence, because low-adherence patients are sicker for reasons the crude comparison wrongly attributes to adherence. The crude curve is therefore too steep."
#| code-fold: true
library(ggplot2)

# Oracle forgiveness curve: re-run the DGP failure process with PDC held fixed
# at a sustained level, health L allowed to respond, separately by drug.
# Parameter values must match simulate_adherence_data() exactly.
oracle_forgiveness_curve <- function(a_grid, drug, n = 50000, K = 3, seed = 2027) {
  set.seed(seed)
  beta0  <- 0.00; betaW <- 0.30; betaLag <- 0.35; betaP <- 0.80; sigma_L <- 0.50
  gamma0 <- -3.80; gammaL <- 0.50; gammaW <- 0.30
  gammaP <- 2.50; gammaA <- 0.00; gammaAP <- -1.20
  risk <- numeric(length(a_grid))
  for (j in seq_along(a_grid)) {
    a  <- a_grid[j]
    W0 <- rnorm(n)
    A0 <- rep(drug, n)
    L  <- matrix(NA_real_, n, K)
    failed <- rep(FALSE, n)
    for (t in 1:K) {
      if (t == 1) {
        L[, t] <- beta0 + betaW * W0 + rnorm(n, 0, sigma_L)
      } else {
        L[, t] <- beta0 + betaW * W0 + betaLag * L[, t - 1] +
                  betaP * (1 - a) + rnorm(n, 0, sigma_L)   # PDC fixed at a
      }
      at_risk <- !failed
      lp <- gamma0 + gammaL * L[at_risk, t] + gammaW * W0[at_risk] +
            gammaP * (1 - a) + gammaA * A0[at_risk] +
            gammaAP * A0[at_risk] * (1 - a)
      new_fail <- rbinom(sum(at_risk), 1, plogis(lp)) == 1L
      failed[which(at_risk)[new_fail]] <- TRUE
    }
    risk[j] <- mean(failed)
  }
  data.frame(adherence = a_grid, risk = risk,
             Drug = ifelse(drug == 1, "Drug A", "Drug B"))
}

a_grid <- seq(0.40, 0.95, by = 0.01)
oracle <- rbind(
  oracle_forgiveness_curve(a_grid, drug = 1),
  oracle_forgiveness_curve(a_grid, drug = 0)
)

# Crude points: bin patients by observed average PDC, per drug.
crude <- dat |>
  dplyr::mutate(
    avg_pdc = rowMeans(cbind(PDC_1, PDC_2, PDC_3)),
    Drug    = ifelse(A0 == 1, "Drug A", "Drug B"),
    bin     = cut(avg_pdc, breaks = seq(0.30, 1.00, by = 0.10))
  ) |>
  dplyr::group_by(Drug, bin) |>
  dplyr::summarise(adherence = mean(avg_pdc),
                   risk = mean(Y_final),
                   n_bin = dplyr::n(), .groups = "drop") |>
  dplyr::filter(n_bin >= 10)

ggplot(oracle, aes(adherence, risk, color = Drug)) +
  geom_line(linewidth = 1.1) +
  geom_point(data = crude, shape = 21, fill = "white", size = 2.4,
             stroke = 1.0, aes(color = Drug)) +
  scale_x_continuous(labels = scales::percent) +
  scale_y_continuous(labels = scales::percent) +
  scale_color_manual(values = c("Drug A" = "#228833", "Drug B" = "#EE6677")) +
  labs(x = "Sustained adherence (PDC)", y = "12-month failure risk",
       color = NULL,
       title = "Oracle forgiveness curves (lines) vs crude risk by adherence (points)") +
  theme_minimal() + theme(legend.position = "bottom")
```

The solid lines are the causal forgiveness curves: the failure risk that would obtain if adherence were set to each level, with health allowed to respond. The open points are the crude failure rates among patients whose observed average adherence fell in each bin. Within each drug, the crude points tend to lie above the causal curve at low adherence and below it at high adherence, because low-adherence patients are sicker for reasons the crude comparison attributes to adherence. The crude comparison therefore overstates the cost of poor adherence and understates the benefit of good adherence, exactly the bias the feedback loop predicts. Note that this within-drug gap is treatment-confounder feedback, a separate issue from the confounding by indication that distorts the cross-drug comparison; reading each drug's curve against its own crude points keeps the two from being conflated.


# Two estimands, and the estimator each one needs

The forgiveness question can be made precise in two ways, and they call for different estimators.

The first is a **static adherence regime**: what would the failure risk be if every patient *sustained* a fixed adherence level, say low, medium, or high, across all K = 3 blocks? This is the estimand that traces the forgiveness curve at fixed points, and the longitudinal TMLE of @sec-illustrated-ltmle estimates it. Its weakness is positivity, and the weakness sharpens with the horizon: requiring a patient to sit in an extreme adherence band at *every* block is a conjunction over blocks, so the support for "low every block" or "near-perfect every block" thins as K grows. The ends of the curve are exactly where the data are thinnest.

The second is a **feasible shift**: what would happen if every patient's adherence were nudged up by a realistic amount, say 5 percentage points, from wherever it currently sits? This is a modified treatment policy [@haneuse2013mtp; @diaz2021lmtp], and LMTP estimates it. Because the shifted adherence stays close to what each patient already does, it respects the support of the data and answers a question an adherence-support program could actually act on.

The two are complementary summaries of the same counterfactual risk surface. The static curve shows the whole shape but needs support at every level, and that support is fragile at the extremes; the feasible shift gives a stable, policy-relevant slice near the observed distribution. Reading them together is the point: the static curve for shape, the shift for a number you can act on.

The positivity strain on the static regime is something you can diagnose directly. The table below tabulates per-block adherence-band membership; tracking how the count in an extreme band shifts across blocks shows where the support for a sustained-extreme regime lives.

```{r}
#| label: positivity-band-diagnostic
#| eval: true
# Tabulate per-block adherence-band membership. Tracking how the count in an
# extreme band (e.g. "low") shifts from block 1 to block K shows the support
# for a sustained-extreme static regime thinning as the horizon grows.
table(dat$PDC_cat_1)
table(dat$PDC_cat_2)
table(dat$PDC_cat_3)
```

The per-block counts only hint at the problem; what a sustained static regime actually requires is the same patient sitting in the extreme band at *every* block. Counting those patients makes the positivity strain concrete.

```{r}
#| label: positivity-sustained
pdc <- as.matrix(dat[, paste0("PDC_", 1:3)])
data.frame(
  regime = c("sustained low (< 0.50 every block)",
             "sustained high (> 0.90 every block)"),
  n_patients = c(sum(rowMeans(pdc < 0.50) == 1),
                 sum(rowMeans(pdc > 0.90) == 1)),
  pct = round(100 * c(mean(rowMeans(pdc < 0.50) == 1),
                      mean(rowMeans(pdc > 0.90) == 1)), 1)
)
```

Only a handful of patients sustain an extreme adherence band across all three blocks, so the static forgiveness-curve estimate rests on almost no data at the ends and is fragile there; this is exactly why the feasible shift, which stays near the observed distribution, is preferred. The longitudinal positivity diagnostic in @sec-positivity-overlap is the general version of this check. A feasible shift sidesteps the problem, which is the motivation for the LMTP approach below.


# LMTP estimation

We now code the estimators themselves. One estimand, the feasible shift, is fit live below so the reader sees a concrete estimate; the heavier study-grade fits (the static regime, the cross-regimen risk difference, and the quantile shift) are shown as `#| eval: false` reference code. Throughout, `lmtp`'s survival mode returns $P(\text{survival})$, so the failure risk is $1 - \text{estimate}$, and a contrast on that scale is a difference in survival probability.

## Shared setup

A single small chunk fixes the node names and the learner library shared by every fit. The study fits with the cross-fitted `lmtp_tmle` (or the sequentially doubly robust `lmtp_sdr`), both with the same learner-agnostic library; `SL.earth` (MARS) is what lets the fit discover the drug-by-adherence interaction without being told it is there.

```{r}
#| label: lmtp-setup
#| message: false
#| warning: false
#| eval: true
library(lmtp)
K        <- 3
trt      <- paste0("PDC_", 1:K)            # continuous adherence nodes
outcome  <- paste0("Y_", 1:K)              # absorbing failure nodes
baseline <- c("W0", "A0")
tv       <- lapply(1:K, function(t) paste0("L_", t))   # time-varying confounder per block
# Learner-agnostic library used by the study (earth = MARS, finds the drug x adherence
# interaction without being told it is there). The study fits with lmtp_tmle or the
# sequentially doubly robust lmtp_sdr; both use this library with cross-fitting (folds = 5).
learners <- c("SL.mean", "SL.glm", "SL.earth")
learners_fast <- c("SL.mean", "SL.glm")   # lighter library for the rendered fit below
```

## Static regime: a point on the forgiveness curve

A static intervention sets PDC to a fixed level at every block. Fitting it on one drug's patients gives one point on that drug's forgiveness curve.

```{r}
#| label: lmtp-static
#| eval: false
# Static intervention: set adherence to a fixed value at every block.
make_static_pdc_shift <- function(pdc_value) {
  force(pdc_value)
  function(data, trt) rep(pdc_value, nrow(data))   # ignore observed value; set to target
}

# Forgiveness-curve point for one drug at sustained adherence = 0.70:
fit_static_A <- lmtp_tmle(
  data = subset(dat, A0 == 1), trt = trt, outcome = outcome,
  baseline = "W0", time_vary = tv,
  shift = make_static_pdc_shift(0.70),
  mtp = TRUE, outcome_type = "survival",        # mtp=TRUE: lmtp treats a fixed set as a degenerate MTP
  learners_trt = learners, learners_outcome = learners, folds = 5)
# risk = 1 - tidy(fit_static_A)$estimate[1]
```

Sweeping the fixed value over a grid traces the whole forgiveness curve, which is the static regime's strength. Its weakness is positivity, and it bites hardest at sustained extreme adherence: the support for "low every block" (or "near-perfect every block") thins as the horizon K grows, so the ends of the curve rest on little data. This is what motivates the feasible shift.

## Feasible capped additive shift

The study's canonical bounded MTP raises each patient's PDC by a small `delta`, but only when the result stays under a `cap`, so no probability mass piles up at PDC = 1.

The rendered fit below uses the lighter learner library (`learners_fast`) and only two cross-fitting folds so the chapter knits quickly; the study uses the richer library with `SL.earth` and `folds = 5` exactly as shown in the reference chunks above and below.

```{r}
#| label: lmtp-feasible
#| eval: true
#| cache: true
#| message: false
#| warning: false
set.seed(2027)   # reproducible cross-fitting splits for the rendered fit

# Capped additive rule: raise PDC by delta, but only for patients not already near the cap,
# so no probability mass piles up at PDC = 1 (which would break positivity / the IF variance).
feasible_add_rule <- function(a, delta, cap) {
  sh <- a + delta
  ifelse(!is.na(a) & sh <= cap, sh, a)            # shift if it stays <= cap, else leave as-is
}
make_feasible_add_shift <- function(delta, cap) {
  force(delta); force(cap)
  function(data, trt) {
    a   <- data[[trt]]
    out <- feasible_add_rule(a, delta, cap)
    out[is.na(a)] <- NA_real_
    out
  }
}
natural_shift <- function(data, trt) data[[trt]]   # no intervention (observed adherence)

fit_shift   <- lmtp_tmle(dat, trt, outcome, baseline = baseline, time_vary = tv,
                         shift = make_feasible_add_shift(0.05, 0.95),
                         mtp = TRUE, outcome_type = "survival",
                         learners_trt = learners_fast, learners_outcome = learners_fast, folds = 2)
fit_natural <- lmtp_tmle(dat, trt, outcome, baseline = baseline, time_vary = tv,
                         shift = natural_shift,
                         mtp = TRUE, outcome_type = "survival",
                         learners_trt = learners_fast, learners_outcome = learners_fast, folds = 2)
```

The cap is what keeps the estimator well-behaved. A naive "+delta capped at 1.0" shift would push every already-high patient to exactly 1.0, piling mass on the boundary and sending the density ratio (and with it the influence-function variance) toward infinity. The capped rule shifts only patients who stay strictly below the cap and leaves the rest at their observed value, so the density ratio stays bounded by construction.

With both fits in hand, the contrast compares the capped +5pp regime against the natural course.

```{r}
#| label: lmtp-feasible-contrast
lmtp::lmtp_contrast(fit_shift, ref = fit_natural, type = "additive")
```

Because `lmtp`'s survival mode returns the survival probability, this additive contrast reports the difference in survival probability between the capped +5pp regime and the natural course. A positive value means the shift raised survival, equivalently it lowered failure risk, so the risk difference is the negative of the reported contrast. Concretely, the magnitude is the change in 12-month virologic-failure risk an adherence-support program would buy if it nudged every patient's PDC up by 5 percentage points (capped at 0.95) at each block, holding everything else as observed.

You can also confirm that the cap kept the density ratios bounded rather than assume it.

```{r}
#| label: density-ratio-diagnostic
#| eval: true
# The capped rule should keep the per-block density ratios bounded; inspecting
# their summary is the check behind the cap (a heavy right tail would flag the
# boundary pile-up the cap is designed to prevent).
summary(fit_shift$density_ratios)
```

## Cross-regimen risk difference: the forgiveness estimand

The manuscript's headline quantity is a cross-regimen risk difference: a *joint* intervention that sets the drug and applies the adherence regime at once. The pattern prepends `A0` as the first treatment node, uses the single absorbing binomial outcome `Y_final` (so the estimate is the failure risk directly), and fits both drug arms on the *same* patients so their influence functions align before contrasting.

```{r}
#| label: lmtp-rd
#| eval: false
# One arm: set A0 = target_drug for everyone AND apply the adherence regime.
fit_joint_arm <- function(data, target_drug, delta = 0.05, cap = 0.95, K = 3) {
  pdc_nodes <- paste0("PDC_", 1:K)
  shift_fn <- function(d, trt) {
    if (trt == "A0") return(rep(as.integer(target_drug), nrow(d)))  # set the drug
    a <- d[[trt]]; o <- feasible_add_rule(a, delta, cap); o[is.na(a)] <- NA_real_; o
  }
  lmtp_tmle(
    data = data, trt = c("A0", pdc_nodes), outcome = "Y_final",
    baseline = NULL,
    time_vary = c(list("W0"), lapply(1:K, function(t) paste0("L_", t))),
    shift = shift_fn, mtp = TRUE, outcome_type = "binomial",
    learners_trt = learners, learners_outcome = learners, folds = 5)
}
fit_A <- fit_joint_arm(dat, target_drug = 1)   # Drug A under the regime
fit_B <- fit_joint_arm(dat, target_drug = 0)   # Drug B under the regime

# Cross-regimen forgiveness RD (Drug A - Drug B under the same adherence regime).
# type = "additive" gives the RD; type = "rr" gives the risk ratio.
lmtp_contrast(fit_A, ref = fit_B, type = "additive")
```

This cross-regimen RD is the forgiveness estimand: it asks how much the failure risk differs between drugs when both are held to the same adherence regime. It is a different question from a shift-versus-natural contrast within a single arm, `lmtp_contrast(fit_shift, ref = fit_natural, type = "additive")`, which answers the policy question "what would a universal +5pp adherence program achieve?". For a static-regime RD, swap the arm's adherence rule for `make_static_pdc_shift(value)` on the PDC nodes (keeping `mtp = TRUE`).

## Quantile shift: what if Drug B adhered like Drug A?

The quantile shift maps each from-drug patient's PDC to the matching quantile of the to-drug's PDC, block by block, so a patient at the 30th percentile of one drug's adherence is moved to the 30th percentile of the other's.

```{r}
#| label: lmtp-quantile
#| eval: false
# Map each from_drug patient's PDC to the matching quantile of to_drug's PDC, per block.
make_quantile_match_shift <- function(df, from_drug, to_drug, K = 3) {
  ecdf_from <- vector("list", K); vals_to <- vector("list", K)
  for (t in 1:K) {
    col <- paste0("PDC_", t)
    ecdf_from[[t]] <- ecdf(df[[col]][df$A0 == from_drug])
    vals_to[[t]]   <- sort(df[[col]][df$A0 == to_drug])
  }
  trt_nodes <- paste0("PDC_", 1:K)
  function(data, trt) {
    t_idx <- match(trt, trt_nodes); if (is.na(t_idx)) return(data[[trt]])
    a <- data[[trt]]
    out <- vapply(seq_along(a), function(i) {
      if (is.na(a[i])) return(NA_real_)
      q <- ecdf_from[[t_idx]](a[i])                  # patient's rank in from_drug
      quantile(vals_to[[t_idx]], probs = q, type = 1)  # same rank in to_drug
    }, numeric(1))
    out
  }
}
```

Framed informally, the quantile shift separates a real-world-effectiveness comparison (drugs as actually adhered to) from an adherence-controlled forgiveness comparison (drugs held to a common adherence profile). This is an ICH-E9-style distinction between effectiveness and a more mechanism-focused estimand, not a formal mediation decomposition.


# The forgiveness curve

The forgiveness curve maps sustained adherence level to counterfactual failure risk for each drug:

$$
r_d(a) = P\big(Y = 1 \mid \text{do}(\bar{A} = a), D = d\big)
$$

A flat curve means the drug is forgiving (patients can miss doses without large increases in failure risk). A steep curve means small adherence drops produce large outcome differences.

Static-regime L-TMLE traces this curve by estimating the failure risk at a grid of fixed PDC thresholds. The LMTP shift approach provides a complementary view: it estimates the marginal benefit of a small adherence improvement, integrated over the observed adherence distribution. The two approaches are different summaries of the same counterfactual risk surface.

The curve is well-estimated only where there is adequate positivity, and positivity is weakest at the extremes. Sustaining a low adherence level across all K = 3 blocks is a conjunction over blocks, so the supporting data thin as the horizon grows, and the ends of the curve become unstable. LMTP shift interventions avoid this problem because they stay close to the observed distribution.


# Interpretation

**Feasible, not hypothetical.** The capped +5pp shift corresponds to something a real adherence support program might achieve. The estimate answers a concrete policy question.

**Causal, not descriptive.** The LMTP framework accounts for the treatment-confounder feedback between adherence and health. When adherence changes, health status changes too, and those changes are propagated through the sequential regression structure.

**Drug-specific effects.** Running the analysis separately by drug reveals whether the marginal return to improved adherence differs by regimen [@nugent2023mtp]. If Drug A is more forgiving, the benefit of an adherence intervention is smaller for Drug A patients (they are already doing relatively well at moderate adherence) and larger for Drug B patients.

**Read the shift estimate alongside the static curve, not alone.** Under positivity strain the variance of the LMTP shift estimator can be underestimated and its interval can under-cover. The feasible-shift number is most trustworthy when it is reported next to the static-regime estimates and the support diagnostics above, rather than as a single standalone figure.


# Connection to mediation

The quantile-matching analysis provides an informal decomposition of the total drug effect into an adherence-mediated component and a pharmacological (forgiveness) component. This is not formal mediation analysis; it is an approximation based on distribution shifts. For the formal TMLE-based mediation framework with proper identification conditions, see @sec-mediation.


## Sources and further reading

- [@diaz2021lmtp; @williams2025lmtp] — lmtp R package and the LMTP framework for modified treatment policies.
- [@haneuse2013mtp] — original modified treatment policy framework.
- [@rosenblum2009virologic] — adherence and virologic failure in HIV.
- [@petersen2006msm] — marginal structural models for adherence effects.
- [@nugent2023mtp] — applied modified treatment policy analysis.
