---
title: "Estimating Interactions in Real-World Evidence Research"
format: html
---

# Estimating Interactions in Real-World Evidence Research {#sec-interactions-rwe}

*Estimated reading time: 50 minutes*

## Abstract

Interactions are among the most clinically interesting and the most
methodologically demanding quantities in real-world evidence research.
Clinicians want to know whether a drug works equally well across patient
subgroups. Public-health planners want to know whether two exposures
combine to produce more harm than the sum of their parts. Many published
analyses answer these questions by reporting a single product-term
coefficient from a logistic or Cox model. This chapter argues for a
different starting point. Interaction analyses should begin with a
clearly defined causal estimand on a meaningful scale, anchored in a
target trial protocol, and should be estimated with methods that respect
the complexity of observational data. Targeted learning — flexible
nuisance estimation through Super Learner [@vanderLaan2007superlearner]
combined with the targeting step and influence-function inference of
TMLE [@vanderLaan2006tmle] — is particularly well suited to this task.
The chapter walks through a binary point-treatment example with TMLE and
Super Learner, scaffolds a collaborative TMLE sensitivity analysis,
sketches a longitudinal extension using LTMLE and TMLE for marginal
structural models, and closes with a reporting checklist that aligns
with current best-practice recommendations
[@talbot2025tmle; @vanderweele2014tutorial; @basten2025additive].

::: {.callout-tip}
## Key takeaways

- **Effect modification** (one treatment, varying effect across a baseline
  variable) and **interaction** (two exposures acting jointly) are different
  scientific questions and require different estimands.
- The same data can show interaction on the **additive** scale and not on
  the **multiplicative** scale, or vice versa. Pre-specify the scale that
  matches your decision question.
- A regression product term is rarely the scientific target. Define
  **subgroup-specific risks** and contrast them.
- Begin every RWE interaction analysis by writing the **target trial
  protocol** [@hernan2016target]: time zero, eligibility, treatment
  strategies, follow-up, outcome, censoring.
- **Positivity is harder inside subgroups** than overall. Diagnose overlap
  within each modifier stratum.
- **Targeted learning** (Super Learner + TMLE) supports flexible nuisance
  estimation, doubly robust point estimation, and influence-function-based
  uncertainty for the contrasts you actually want to report.
- **CTMLE** is best framed as a sensitivity analysis around fragile
  propensity-score models, not as the introductory teaching vehicle.
- Pre-specify a small number of clinically motivated effect modifiers,
  separate confirmatory from exploratory work, and report **both scales**
  with cell-specific risks and a common reference category
  [@vanderweele2014tutorial].
:::

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

---

## 1. Why interaction matters in epidemiology and real-world evidence

A single average treatment effect is rarely the end of a clinically useful
analysis. Two follow-up questions almost always arise. *Does the effect
differ across patient groups?* And *do two exposures combine in ways that
one alone would not predict?* Both questions are about interaction in a
broad sense, but they are not the same question, and the methods that
answer them are not identical.

::: {.callout-note}
## Scope of this chapter
This chapter focuses on **causal interaction**: the joint effect of two
manipulable exposures, and the contrast of subgroup-specific effects when
that contrast is itself the scientific target. The companion chapter,
[Effect Modification and Heterogeneous Treatment Effects](03-04-effect-modification.qmd),
covers the closely related but distinct question of how the effect of a
single treatment varies across baseline covariates (the CATE), with tools
including TMLE for prespecified subgroups, causal forests, and treatment-effect
variable importance.

The two chapters are designed to be read together. Effect modification
asks *who benefits more from one treatment*; interaction asks *how two
exposures combine*.

**Single time-point analyses** occupy Sections 1 through 10. **Time-varying
treatment, longitudinal interactions, and marginal structural models** are
concentrated in Section 11, and should be read after the single time-point
material is comfortable. Section 12 closes the chapter.
:::

Several closely related ideas are routinely conflated in published work.
Disentangling them is the first task of any chapter on interactions.

- **Effect modification** describes how the causal effect of a *single*
  treatment $A$ varies across levels of a baseline characteristic $V$.
  The target is a contrast of subgroup-specific causal effects, for example
  the risk difference if treated minus untreated within $V=1$ versus within
  $V=0$.
- **Interaction**, in the strict sense used by VanderWeele and Knol
  [-@vanderweele2014tutorial], refers to the *joint* causal effect of two
  exposures $A$ and $B$. Both are conceptually manipulable and the question
  is whether their combined effect departs from what you would expect from
  each effect alone.
- **Additive interaction** asks whether the joint effect departs from the
  *sum* of the individual effects on the absolute risk scale.
- **Multiplicative interaction** asks whether the joint effect departs from
  the *product* of the individual effects on the relative-risk scale. The
  same data can show interaction on one scale and none on the other.
- **Subgroup analysis** is a reporting style: presenting effects within
  prespecified strata. It is the *vehicle* for studying effect modification,
  but a list of subgroup point estimates is not by itself a heterogeneity
  test.
- **Predictive heterogeneity** refers to algorithms that learn high-dimensional
  functions of covariates to predict who benefits — for example, the conditional
  average treatment effect surface produced by a causal forest. These tools are
  covered in the companion [Effect Modification](03-04-effect-modification.qmd)
  chapter and should not be confused with the inferential, low-dimensional
  interaction parameters that are the focus of this chapter.

A second motivation for taking interaction seriously is **scale dependence**.
A cohort study may find that an antihypertensive halves stroke risk in
low-risk patients (relative risk 0.5) and also halves stroke risk in
high-risk patients (relative risk 0.5). On the multiplicative scale there is
no interaction. On the additive scale, the absolute risk reduction is much
larger for the high-risk group, and that is what determines who should be
treated first. Public health and clinical decisions almost always depend on
absolute differences, not ratios [@vanderweele2014tutorial; @basten2025additive].

A third motivation is the rapid growth of TMLE and other doubly robust
methods in observational epidemiology. A 2023 systematic review identified
89 observational studies that applied TMLE [@smith2023systematic], and the
growth is steepest in pharmacoepidemiology and comparative effectiveness
work---exactly the settings where interaction questions are most common.

### Why interaction matters in practice

Interaction is not an academic embellishment of an average effect. It
is the analytic foundation for several of the most important applied
questions in modern epidemiology and clinical practice.

- **Treatment targeting.** When the absolute benefit of a treatment is
  much larger in one subgroup than another, fixed budgets and finite
  clinic time turn subgroup-aware deployment into a matter of
  efficiency. Statins offered to high-risk patients prevent more events
  per prescription than the same drug offered uniformly. Interaction
  estimates are what turn "treatment works on average" into "treat
  these patients first."
- **Policy heterogeneity.** Public-health interventions rarely have
  uniform effects across regions, age groups, or baseline-risk strata.
  A vaccination campaign that reduces hospitalisation by 10 percentage
  points among the elderly and 1 percentage point among the young
  calls for a different roll-out plan than one with uniform 5.5
  percentage-point effects. Interaction estimates inform where to
  invest the next public-health dollar.
- **Joint-exposure aetiology.** When two exposures may share a
  biological pathway — environmental tobacco smoke and occupational
  asbestos, for example — interaction analyses on the additive scale
  carry information about whether the combined exposure produces
  excess events beyond what either alone predicts. That information
  matters for screening, regulation, and aetiologic theory.
- **Comorbidity-by-treatment interaction in pharmacoepidemiology.**
  Whether a drug's safety profile depends on a baseline comorbidity —
  for example, whether the cardiovascular risk associated with a
  glucose-lowering agent differs in patients with versus without prior
  heart failure — is a recurrent question in comparative-safety
  studies. The same machinery applies here, with the comorbidity
  playing the role of the second exposure $E$.
- **Biomarker-by-biomarker interaction.** When two continuous clinical
  measurements (blood pressure and HbA1c, for example) jointly index
  cardiometabolic risk, the interaction question is whether their
  effects on the outcome compound beyond what either alone predicts.
  These analyses typically need the modified-treatment-policy
  framework introduced in Section 11 rather than binary-cell
  contrasts.

The common thread is that interaction estimates inform decisions and
explanations, not just summaries. That is why the rest of this chapter
spends so much energy on getting the estimand, the scale, and the
uncertainty right.

### Interaction vs. effect modification

Because these two ideas are conflated so often, it is worth restating the
distinction in operational terms:

- **Effect modification** is *heterogeneity in conditional effects*. The
  modifier $V$ need not be a manipulable exposure; it might be sex, a
  genotype, or a baseline biomarker. The estimand is the contrast of
  conditional treatment effects across levels of $V$.
- **Interaction** is a *contrast of joint interventions*. It requires that
  *both* $A$ and a second variable $E$ be conceptually amenable to
  intervention, and the estimand involves four counterfactual worlds in
  which $(A, E)$ is set jointly to $(0,0), (0,1), (1,0)$, and $(1,1)$.

The implication is methodological: a regression "interaction term" between
$A$ and $E$ does not estimate a causal interaction. It estimates a
parameter of a parametric model whose mapping to a causal contrast holds
only under strong assumptions. Many applied analyses report the
coefficient and call it interaction; this chapter argues that the causal
interaction parameter should be defined and estimated directly.

::: {.callout-note}
## What this means in practice
Before you write a single line of code, decide which of the four
ideas---effect modification, two-exposure interaction, additive contrast,
or multiplicative contrast---you are estimating. Different choices imply
different estimands, different reporting tables, and different defaults
for how to communicate uncertainty. *Causal interaction always involves
comparing four hypothetical worlds*; effect modification compares two
worlds within each level of a stratifier.
:::

---

## 2. From scientific question to causal estimand

The most common mistake in interaction analysis is to start with a model.
The right starting point is to write down the quantity you want to estimate
in counterfactual language, *before* any software is opened.

### Defining interaction as a causal estimand

Suppose two binary exposures $A$ and $E$ can each be intervened on, and let
$Y^{a,e}$ denote the potential outcome under joint intervention setting
$A=a$ and $E=e$. There are four such counterfactuals: $Y^{0,0}$, $Y^{0,1}$,
$Y^{1,0}$, $Y^{1,1}$. The causal interaction parameters compare how the
effect of one exposure depends on the level of the other.

- **Additive interaction parameter**:

$$\psi_{\text{add}} \;=\; E[Y^{1,1} - Y^{0,1}] \;-\; E[Y^{1,0} - Y^{0,0}]$$

- **Multiplicative interaction parameter**:

$$\psi_{\text{mult}} \;=\; \dfrac{E[Y^{1,1}] / E[Y^{0,1}]}{E[Y^{1,0}] / E[Y^{0,0}]}$$

These quantities are **parameters of the data-generating distribution**,
not coefficients of any model. Following the Targeted Learning principle
that the *parameter must be defined before the estimator*
[@petersen2014causal; @vanderLaan2006tmle], we treat any subsequent
modeling choice (logistic, log-binomial, TMLE, g-computation) as a means
to estimate $\psi_{\text{add}}$ or $\psi_{\text{mult}}$, not as a
substitute definition.

In plain English: **we are asking how the effect of $A$ changes when we
also intervene on $E$.** $\psi_{\text{add}} = 0$ means the absolute effect
of $A$ is the same whether $E$ is set to $0$ or $1$. $\psi_{\text{mult}} =
1$ means the relative effect of $A$ is the same across levels of $E$.
Either parameter can be nonzero on its own scale while the other is null.

Parametric models that impose linear or log-linear interactions can
misrepresent the underlying causal interaction structure: the model's
"interaction term" is a derivative of the assumed functional form, not of
the true joint potential-outcome distribution. When the functional form is
wrong, the coefficient is wrong, and a likelihood-ratio test on the
product term tells you about the model's fit, not about $\psi_{\text{add}}$
or $\psi_{\text{mult}}$.

::: {.callout-note}
## A toy example: computing interaction by hand

Suppose the four counterfactual risks are known exactly:

| | $E=0$ | $E=1$ |
|---|---:|---:|
| $A=0$ | $E[Y^{0,0}] = 0.05$ | $E[Y^{0,1}] = 0.10$ |
| $A=1$ | $E[Y^{1,0}] = 0.08$ | $E[Y^{1,1}] = 0.20$ |

The effect of $A$ when $E=0$ is $0.08 - 0.05 = 0.03$ (3 percentage
points). The effect of $A$ when $E=1$ is $0.20 - 0.10 = 0.10$ (10
percentage points).

- **Additive interaction**: $\psi_{\text{add}} = 0.10 - 0.03 = 0.07$.
  Treatment is 7 percentage points more beneficial (or harmful) when
  $E=1$ than when $E=0$.
- **Multiplicative interaction**: the risk ratio is
  $0.08 / 0.05 = 1.60$ when $E=0$ and $0.20 / 0.10 = 2.00$ when $E=1$,
  giving $\psi_{\text{mult}} = 2.00 / 1.60 = 1.25$. Treatment multiplies
  the risk by 25% more when $E=1$.
- **RERI** (taking $A=0, E=0$ as the doubly unexposed reference):
  $RR_{11} = 4.0$, $RR_{10} = 1.6$, $RR_{01} = 2.0$, so
  $\text{RERI} = 4.0 - 1.6 - 2.0 + 1 = 1.4$.
- **Attributable proportion due to interaction**: $\text{AP} = \text{RERI}/RR_{11} = 1.4/4.0 = 0.35$.
  About 35% of the risk in the doubly exposed group is attributable to
  the interaction beyond the sum of the individual effects.
- **Synergy index**: $S = (RR_{11} - 1)\big/[(RR_{10} - 1) + (RR_{01} - 1)] = 3.0 / 1.6 = 1.88$.
  The joint excess risk is 1.88 times the sum of the separate excess
  risks; values $> 1$ indicate super-additivity.

The same four numbers produce five different interaction summaries.
TMLE's role is to estimate those four counterfactual risks from
observational data under confounding; once they are in hand, all of
RERI, AP, $S$, and $\psi_{\text{add}}$ are arithmetic. RERI, AP, and
$S$ are the standard additive-interaction trio in epidemiology
[@vanderweele2014tutorial]; reporting all three is uncommon but
inexpensive once the joint counterfactual means are estimated.
:::

### Subgroup-specific estimands

When the second variable is a baseline modifier $M$ that you do *not*
intervene on, the relevant estimands are subgroup-specific. Let $Y^a$
denote the potential outcome under treatment level $a$. The estimands you
will most often want are:

- **Subgroup-specific risk under treatment**: $P(Y^1 = 1 \mid M = m)$
- **Subgroup-specific risk under control**: $P(Y^0 = 1 \mid M = m)$
- **Subgroup-specific risk difference**: $RD(m) = P(Y^1=1 \mid M=m) - P(Y^0=1 \mid M=m)$
- **Subgroup-specific risk ratio**: $RR(m) = P(Y^1=1 \mid M=m) / P(Y^0=1 \mid M=m)$
- **Additive interaction contrast**: $RD(1) - RD(0)$
- **Multiplicative interaction contrast**: $RR(1) / RR(0)$

For two binary exposures $A$ and $B$ acting jointly, the standard
public-health summary is the **relative excess risk due to interaction**
(RERI):

$$\text{RERI} = RR_{11} - RR_{10} - RR_{01} + 1$$

where $RR_{ab}$ is the risk ratio under joint exposure level $(A=a, B=b)$
relative to the doubly unexposed group. RERI = 0 indicates no additive
interaction; positive values indicate super-additivity, negative values
sub-additivity. The interpretation is straightforward only when the doubly
unexposed group has the lowest baseline risk; we return to this point
below.

Why bother with this counterfactual machinery rather than reading off the
product-term coefficient from a logistic regression? Three reasons.

1. **The coefficient is on the wrong scale**. A product-term coefficient in
   a logistic model speaks to the odds ratio, not the risk ratio or risk
   difference. Odds-ratio interaction is not generally interpretable as
   either additive or multiplicative interaction in the public-health sense
   [@vanderweele2014tutorial].
2. **The coefficient is conditional, not marginal**. It estimates an effect
   conditional on covariates included in the model. The marginal subgroup
   effect---averaging over the covariate distribution within the
   subgroup---is what most decision questions actually need.
3. **The coefficient confounds estimation with model specification**. Move
   one covariate in or out and the interaction coefficient can change
   meaningfully. A causal estimand defined directly from potential outcomes
   does not depend on whether you used a polynomial or a spline for $W_1$.

Table 1 contrasts the two scales explicitly.

| Feature | Additive interaction | Multiplicative interaction |
|---|---|---|
| Mathematical form | $RD(1) - RD(0)$, or $\text{RERI}$ for two exposures | $RR(1)/RR(0)$, or $RR_{11}/(RR_{10} \cdot RR_{01})$ |
| Underlying scale | Absolute risk | Relative risk |
| Best for | Public health targeting, resource allocation, number-needed-to-treat reasoning | Etiologic comparisons, ratio-style benefit-risk summaries |
| Default model | None; estimate by standardization or TMLE | Logistic, log-binomial, Cox product term |
| Coding sensitivity | Highly sensitive when one exposure is preventive; lowest-risk stratum should be reference | Less sensitive but still requires explicit reference |
| Decision relevance | Direct: tells you absolute benefit by subgroup | Indirect: needs baseline risk to translate into absolute benefit |
: Table 1. Additive versus multiplicative interaction at a glance. {.striped}

A second comparison maps each scientific question to a recommended
estimator family.

| Scientific aim | Preferred estimand | Recommended estimator |
|---|---|---|
| Treatment effect within prespecified subgroups | $RD(m)$, $RR(m)$ | TMLE within strata, or stratified Super Learner outcome regression with standardization |
| Difference in treatment effect across subgroups | $RD(1) - RD(0)$, $RR(1)/RR(0)$ | TMLE in each stratum, then delta-method or bootstrap contrasts |
| Joint effect of two binary exposures | $RR_{11}, RR_{10}, RR_{01}$, RERI | TMLE for each joint regime mean, with post-estimation contrasts |
| Time-varying treatment with subgroup contrast | Counterfactual mean / survival under dynamic strategy by subgroup | LTMLE [@vanderLaan2012ltmle] or TMLE for marginal structural models |
| Many candidate modifiers, exploratory discovery | Conditional average treatment effect surface | Causal forest or simultaneous one-step TMLE with multiplicity control [@wei2023simultaneous] |
: Table 2. Causal estimands for interaction questions and recommended estimators. {.striped}

::: {.callout-note}
## What this means in practice
Write the estimand on a sticky note before you write the model. If you
cannot describe what you are estimating in one sentence using potential
outcomes, you are not yet ready to estimate it.
:::

---

## 3. Interaction-specific design and identification

Interaction analyses inherit the standard causal assumptions covered in
detail in [Chapter 1.2](01-02-causal-roadmap.qmd) — consistency,
exchangeability given measured confounders, positivity, and no
interference unless explicitly modelled. This section focuses only on
the issues that get *harder* when interaction is the target.

### The target trial protocol

In any RWE study, Hernán and Robins [-@hernan2016target] recommend that the
analysis be designed as an *emulation of a target randomized trial*. For
interaction work, this discipline is doubly important, because subgroup
analyses are uniquely vulnerable to design errors that mimic real
heterogeneity. Before any modeling decision, the analyst should write down:

- **Eligibility criteria** including age, indication, and absence of prior
  outcome events.
- **Treatment strategies** to be compared, including dose, duration, and
  initiation rules.
- **Time zero**: the date when eligibility is met, the assigned strategy
  begins, and follow-up starts.
- **Follow-up window** and outcome definition.
- **Censoring rules**: discontinuation, switching, loss to follow-up,
  competing events.
- **Candidate effect modifiers**: pre-treatment, well-measured, biologically
  or clinically motivated.
- **Causal assumptions** the analysis will rely on.

If any of these are decided after looking at the outcome, you risk
manufacturing apparent heterogeneity through differential immortal time,
selection bias, or post-hoc subgroup tuning.

### Interaction must be estimated on a common target population

A recurring failure mode in applied work is to compute an interaction
contrast from *separately matched or stratified subgroup analyses*. A
typical version of this looks like the following: the analyst matches
$E=1$ patients to $E=0$ patients on baseline covariates, estimates the
treatment effect within each matched set, and then computes an
"interaction" as the difference of those two subgroup-specific
effects. The result is not generally an interpretable causal interaction
parameter, for three reasons.

- **The matched samples differ from each other.** Each match is built
  to balance covariates *within its own subgroup*. The two resulting
  populations do not overlap on the same target, so the contrast is
  not estimating a single counterfactual quantity defined on one
  population.
- **The contrast is not anchored to a counterfactual estimand.** $\psi_{\text{add}} = E[Y^{1,1}] - E[Y^{0,1}] - E[Y^{1,0}] + E[Y^{0,0}]$
  averages each potential outcome over the *same* covariate distribution
  — typically the full study cohort. A difference of two separately
  matched effects does not.
- **Confounding control is uneven across exposure cells.** Patterns of
  confounding can differ by $E$, so subgroup-specific matching may
  succeed in one stratum and fail in another. The interaction estimate
  inherits the worst of those failures.

The remedy is to estimate the interaction parameter directly on the
full cohort, with a single adjustment specification (regression,
Super Learner, or a TMLE plug-in) that targets all four counterfactual
means jointly. Section 6 implements exactly this workflow. When
inverse-probability weights or matching are used as part of the
adjustment, they should be constructed on the full cohort with treatment
defined as the *joint* exposure $(A, E)$, not on each subgroup
separately. If the only data available are pre-aggregated
subgroup-specific estimates, the most defensible approach is to report
each subgroup's effect with its own confidence interval and explicitly
flag that no formal interaction parameter has been identified.

### Why positivity is the silent killer of interaction analyses

The positivity assumption is local. It must hold within every covariate
pattern you condition on. As soon as you stratify by a modifier or
require support for a joint exposure, you are conditioning on more, and
any positivity weakness in the original cohort is amplified.

**Interaction estimation is more sensitive to positivity violations
than main-effect estimation, because it requires support across joint
exposure strata.** A main-effect analysis needs both treated and
untreated patients across the covariate distribution; an interaction
analysis needs both, *within every level of the second exposure or
modifier*. The number of cells that must be populated grows with the
product of the exposure cardinalities, while the data per cell shrinks.

A concrete example makes this vivid. Consider a 4,000-patient cohort
with binary $A$ (statin use), binary $E$ (heavy alcohol use), and
several binary clinical confounders. The marginal cells are reasonably
large:

| Cell | $A=0, E=0$ | $A=0, E=1$ | $A=1, E=0$ | $A=1, E=1$ |
|---|---:|---:|---:|---:|
| Patients | 1,800 | 600 | 1,400 | 200 |
| Outcome events | 90 | 60 | 70 | 20 |

The doubly exposed cell ($A=1, E=1$) holds 200 patients, of whom 20 have
the outcome. Within that cell, the propensity score for $A$ given the
clinical confounders may be near 1 for most patients (they are precisely
the people likely to be on statins) and near 0 for a handful---**that
handful then dominates the inverse-probability weights for the
interaction estimate**. The headline RERI may move dramatically when
those few patients are dropped, even though the main-effect estimate of
statins on the outcome is stable.

Two recurring failure modes follow:

- **A frailty-by-treatment interaction** where almost no frail patients
  receive aggressive therapy. Within the frail stratum, positivity
  fails: you cannot estimate $E[Y^1 \mid \text{frail}]$ from the data,
  no matter what estimator you choose.
- **A two-exposure analysis of statin use and heavy alcohol use** as
  above. The $A=1, E=1$ joint cell may be tiny in any realistic cohort.
  The apparent RERI estimate may be dominated by a handful of patients
  with extreme weights, and the confidence interval will be wide,
  unstable, or both.

This is why propensity-score diagnostics should be reported *within
each joint exposure cell or modifier subgroup*, not only overall, and
why collaborative TMLE (Section 7) was developed as a sensitivity
strategy for fragile propensity-score estimation.

### Interaction through the Targeted Learning roadmap

Petersen and van der Laan [-@petersen2014causal] formalize the workflow as
a small sequence of disciplined steps. Applied specifically to
interaction, the roadmap separates the *causal question* from the
*statistical procedure* and protects the analysis from the most common
failure mode: letting a model dictate what counts as interaction.

::: {.callout-important}
## Interaction through the Targeted Learning roadmap

**Step 0 — Define the question.** State, in clinical or public-health
language, whether you are asking (i) does the effect of $A$ vary across a
non-manipulable modifier $M$ (effect modification), or (ii) do two
manipulable exposures $A$ and $E$ combine in ways that depart from their
individual effects (interaction).

**Step 1 — Statistical model.** Adopt a *nonparametric* model for the
joint distribution of $(W, A, E, Y)$ (or $(W, A, M, Y)$). No functional
form is imposed at this step. The Targeted Learning principle is that
parametric structure should be a tool, not an assumption baked into the
estimand.

**Step 2 — Causal estimand.** Write the parameter explicitly as a
function of counterfactual outcomes:
$\psi_{\text{add}} = E[Y^{1,1} - Y^{0,1}] - E[Y^{1,0} - Y^{0,0}]$ for the
joint-intervention case, or $RD(1) - RD(0)$ for the modifier case. *This
is what the analysis aims to recover; it is not a regression coefficient.*

**Step 3 — Identification.** State the assumptions under which the causal
estimand maps to a function of the observed-data distribution:
exchangeability for $A$ (and $E$) given $W$, joint positivity, consistency,
and no interference. Identification fails or weakens whenever any of
these break, and the estimator cannot rescue identification it does not
have.

**Step 4 — Estimation.** Estimate the nuisance functions
($Q$ for the outcome regression, $g$ for the treatment mechanisms) with
Super Learner, then apply a TMLE targeting step so that the plug-in
estimator solves the efficient influence function equation for
$\psi_{\text{add}}$ or $\psi_{\text{mult}}$.

**Step 5 — Sensitivity analysis.** Vary the learner library, the
adjustment set, and the propensity-score selection (e.g., CTMLE) to
characterize the stability of the interaction estimate. Quantify the
amount of unmeasured confounding that would be needed to overturn it.
:::

The roadmap's message for interaction work is uncompromising: **interaction
parameters are not coefficients; they are functions of counterfactual
risks**. Any procedure that conflates these---reporting a logistic
product term as if it answered $\psi_{\text{add}}$, for example---has
skipped Step 2.

::: {.callout-note}
## What this means in practice
Treat positivity diagnostics inside subgroups as a non-negotiable step, not
an optional appendix. When the modifier strata are small or the joint
exposure cells are sparse, no estimator---however sophisticated---can
manufacture identifiability that the data do not support.
:::

---

## 4. Why targeted learning is useful for interaction estimation

Super Learner [@vanderLaan2007superlearner] and TMLE
[@vanderLaan2006tmle] are introduced in [Chapter 2.3](02-03-doubly-robust-tmle.qmd)
and [Chapter 2.4](02-04-superlearner.qmd); the cross-validated, collaborative,
and adaptive variants (CV-TMLE, CTMLE, A-TMLE) are covered in
[Chapter 2.5](02-05-advanced-tmle.qmd). This section assumes that
material as background and focuses on what changes when the target
parameter is an interaction.

### Why TMLE is preferred for interaction

Targeted learning is useful for any causal parameter, but interaction
amplifies the two pathologies it was designed to fix. **Model
misspecification risk is amplified** because an interaction is a
*derivative* of the outcome surface: it depends on differences of
differences of fitted values, so small functional-form errors that are
invisible in a main-effect estimate can dominate the interaction estimate.
**Positivity problems are amplified** because the joint exposure cells
$(A=a, E=e)$ are typically sparser than either marginal cell, and
inverse-probability weights deteriorate accordingly.

Two TMLE properties directly address these amplifications. **Flexible
nuisance estimation through Super Learner** lets the data choose the
shape of $Q$ rather than imposing a parametric form whose interaction
term is a model artefact. **Double robustness** means the interaction
estimate remains consistent if either $Q$ or $g$ is misspecified — a real
safety margin in joint-cell estimation, where neither model is easy to
get exactly right. The targeting step also improves efficiency relative
to plug-in g-computation, which matters when the joint cell is small and
every bit of information has to be used optimally.

### From estimand to estimator: practical workflow

The slogan "estimand first, then estimator" is easier to repeat than to
operationalize. The five steps below make the mapping concrete for an
interaction analysis with a binary treatment $A$ and a binary second
exposure $E$ (or modifier $M$). **At its heart, the procedure estimates
four counterfactual means and combines them into the contrast you
specified in Section 2.**

::: {.callout-important}
## From estimand to estimator (5 steps)

**Step 1 — Estimate the outcome regression** $Q(A, E, W) = E[Y \mid A, E, W]$
with Super Learner. This is a single model that uses the *observed* data
and includes treatment, second exposure, and confounders as predictors.

**Step 2 — Generate counterfactual predictions** under each of the four
joint regimes by querying the same fitted model on modified data:
$Q(1,1,W_i)$, $Q(0,1,W_i)$, $Q(1,0,W_i)$, $Q(0,0,W_i)$ for every patient
$i$.

**Step 3 — Compute the interaction contrast** by averaging:
$\widehat{\psi}_{\text{add}} = \frac{1}{n}\sum_i [Q(1,1,W_i) - Q(0,1,W_i)] - \frac{1}{n}\sum_i [Q(1,0,W_i) - Q(0,0,W_i)]$.
At this point you have a plug-in g-computation estimate.

**Step 4 — Apply the TMLE update.** Estimate the treatment mechanisms
$g_A(A \mid W)$ and $g_E(E \mid W)$ (or their joint $g(A,E \mid W)$) with
Super Learner. Use these to construct the clever covariate that updates
$Q$ along the least-favorable submodel for your target parameter. The
updated $Q^*$ now solves the efficient influence-function equation for
$\psi_{\text{add}}$ (or $\psi_{\text{mult}}$).

**Step 5 — Inference via the influence curve.** Compute the empirical
influence function of $\widehat{\psi}_{\text{add}}$ using $Q^*$ and $g$,
then form a Wald confidence interval from its empirical variance. The
same machinery yields valid CIs for any smooth contrast of the four
counterfactual means.
:::

Section 6 implements Steps 1–3 directly with `sl3` and Steps 4–5 via
`tmle3` (or stratified `tmle` plus bootstrap as a transparent
alternative). The point of the boxed mapping is that *every line of
code in Section 6 corresponds to one of these five steps*; if you can
say which step a line belongs to, the workflow is well-specified.

### Recent literature

Three recent papers are worth keeping in view: a 2023 systematic review
of 89 observational TMLE applications that documented rapid uptake in
epidemiology and frequent use for heterogeneity questions
[@smith2023systematic]; a 2025 best-practice tutorial that walks through
TMLE implementation choices for applied researchers [@talbot2025tmle];
and a 2023 paper on simultaneous one-step TMLE for many subgroup effects,
with valid joint confidence intervals [@wei2023simultaneous].

::: {.callout-note}
## What this means in practice
For most MPH-level interaction analyses, the rule is simple. Estimate
the nuisance functions with Super Learner using a modest, sensible
library, target with TMLE for the joint-intervention parameter, and use
the influence function for inference. Escalate to CV-TMLE or `drtmle`
only when the learner library is genuinely aggressive or the joint
exposure cells are large enough to support data splitting (see
[Chapter 2.5](02-05-advanced-tmle.qmd)).
:::

---

## 5. Additive and multiplicative interaction in practice

The most important recommendation in the VanderWeele and Knol tutorial
[-@vanderweele2014tutorial] is to report both scales whenever the
outcome and data permit. Scale dependence is not a methodological
detail. It is a substantive claim about how two exposures combine to
produce risk.

### When additive interaction is preferable for public-health decisions

If the question is *who should be prioritized for an intervention?* the
answer almost always lives on the additive scale. A 2-percentage-point
absolute risk reduction in a high-incidence subgroup represents a much
larger public health benefit than the same relative-risk reduction in a
low-incidence subgroup. Number-needed-to-treat reasoning, resource
allocation, and ranking of intervention sites are all additive
operations. Multiplicative measures are useful for etiologic synthesis but
do not by themselves answer "where should we deploy this drug first?"
[@basten2025additive; @vanderweele2014tutorial].

### Operationalizing the choice of scale

The two scales answer two different questions, and the analyst should
state which one the analysis is built to answer.

| Scale | What it tells you | Best for |
|---|---|---|
| **Additive** ($\psi_{\text{add}}$, RD contrasts, RERI) | Public-health impact: how many additional events are prevented (or caused) per person treated, by subgroup | Decision-making, resource allocation, NNT, policy targeting |
| **Multiplicative** ($\psi_{\text{mult}}$, RR contrasts) | Relative effect: how much the *ratio* of treated to untreated risk changes across exposure cells | Etiologic comparison, meta-analytic synthesis, biology |

::: {.callout-tip}
## Recommendation
**For decision-making, prioritize additive interaction unless there is a
specific scientific reason to use the multiplicative scale.** Report
the multiplicative contrast alongside it whenever feasible---the two
scales together leave the reader free to translate the result into
either decision or etiologic terms, while a single scale forces a
choice the reader may not share.
:::

### Coding matters: the lowest-risk reference rule

Standard additive interaction measures, including RERI, assume the doubly
unexposed group is the reference. When one exposure is *preventive*
(e.g., statin use, vaccination), simply plugging into the RERI formula
without re-coding can produce values that look like "antagonism" but
actually reflect a coding artifact.

The remedy, recommended in [@vanderweele2014tutorial], is to **re-code so
that the lowest-risk joint stratum is the reference**. For statin use ($A$)
and heavy smoking ($B$), define the reference as the joint stratum with
the lowest expected outcome risk (typically statin user, non-smoker), and
report RERI relative to that reference. The numerical answer is then
interpretable as the additive departure from the safest combination.

### Cell-specific reporting

Whatever scale you privilege, the most useful presentation is a
**cell-specific risk table**: estimated risks under each combination of
exposures (or under treatment vs. control within each modifier stratum),
plus the contrasts of interest. Readers can then reconstruct any other
contrast they care about and assess whether the differences are large
enough to matter clinically.

### Time-to-event outcomes: choose a risk-scale estimand

Many real-world evidence analyses have time-to-event outcomes, and the
default model in that setting is the Cox proportional-hazards
regression. The default temptation, then, is to introduce an interaction
term in the linear predictor and report a hazard-ratio interaction. For
interaction analyses specifically, **this is the wrong default**.
Hazard-ratio interactions are problematic for three connected reasons.

- **Hazards are non-collapsible.** The marginal hazard ratio is not in
  general a weighted average of conditional hazard ratios over baseline
  covariates, so even an unconfounded marginal hazard-ratio interaction
  is not a clean function of the underlying counterfactual hazards
  [@martinussen2018hazardratios; @aalen2015cox].
- **Hazards condition on survival.** $\lambda(t \mid A, E)$ is the
  instantaneous risk *given* survival to time $t$, and the survivors at
  $t$ are themselves a selected subset of the original cohort. Even if
  treatment is randomly assigned at baseline, the survivors under
  $A=1, E=1$ and the survivors under $A=0, E=0$ at time $t$ are
  comparable populations only at $t = 0$. A hazard-ratio interaction
  therefore mixes a true treatment-by-covariate effect with a
  time-varying selection effect.
- **Additive interaction on hazards is not additive interaction on
  cumulative incidence.** A model that is multiplicative on hazards is
  highly nonlinear on the cumulative incidence scale, and the
  interaction parameter on the hazard scale does not generally map
  onto $\psi_{\text{add}}$, RERI, AP, or $S$.

The recommendation that follows is straightforward. **Define and target
an interaction parameter on the cumulative-risk scale at a clinically
meaningful time horizon $t^*$**, not on the hazard scale. Useful
risk-scale estimands include:

- **Cumulative incidence at $t^*$**: $F^{a,e}(t^*) = P(T^{a,e} \le t^*)$. The four counterfactual cumulative incidences populate exactly the same 2$\times$2 table as the binary-outcome example in Section 2, and RERI / AP / $S$ are computed directly from them.
- **Restricted mean survival time (RMST) at $t^*$**: $\mu^{a,e}(t^*) = E[\min(T^{a,e}, t^*)]$. RMST contrasts have a natural decision-theoretic interpretation (expected event-free time gained) and an additive scale that is straightforward to interpret.
- **Risk difference at $t^*$**: $F^{1,1}(t^*) - F^{0,0}(t^*)$ and the corresponding interaction contrast $F^{1,1}(t^*) - F^{0,1}(t^*) - F^{1,0}(t^*) + F^{0,0}(t^*)$.

Each of these can be estimated with a TMLE that targets the
counterfactual survival or cumulative-incidence function directly,
using `survtmle`, `concrete`, `ltmle`, or `lmtp` (with survival outcome
nodes). The point-treatment chapter on time-to-event outcomes
([Chapter 3.11](03-11-time-to-event-competing-risks.qmd)) walks through
the survival-TMLE machinery; combining it with the joint-intervention
workflow of Section 6 — by replacing $Q(A, E, W) = E[Y \mid A, E, W]$
with $\hat F^{a,e}(t^*; W)$ and contrasting four counterfactual
cumulative incidences — yields an interaction estimate that is free of
hazard-scale interpretive problems.

::: {.callout-warning}
## Avoid: hazard-ratio interaction terms as the headline
A Cox model with a $A \times E$ product term is fine as a sensitivity
analysis or as an exploratory description of how hazards differ by
subgroup, but the headline interaction estimate in a real-world
evidence paper should be defined and reported on the cumulative-risk
scale, with an explicit time horizon. The same data analysed at
$t^* = 1$ year, $t^* = 2$ years, and $t^* = 5$ years can give
materially different interaction conclusions, and that is information
the reader needs.
:::

### Interaction in real-world evidence

Real-world data add three challenges to the picture above that randomized
trials usually do not face.

- **Positivity issues are amplified.** Joint exposure cells $(A=a, E=e)$
  may be sparsely populated, especially when both exposures are rare or
  when one is preventive and the other indicated for similar patients.
  Subgroup-level positivity must be checked separately within each cell;
  inverse-probability components become unstable as the conditional
  probability of the joint regime approaches zero. Positivity weakness is
  the single most common reason interaction analyses produce headline
  estimates that fail to replicate.
- **Model dependence is amplified.** Interaction estimates are derivatives
  of the outcome surface and are intrinsically more sensitive to
  functional-form assumptions than main effects. Parametric models that
  impose linear or log-linear interactions can misrepresent the underlying
  causal interaction structure. Super Learner mitigates this by allowing
  the data to choose among flexible learners, and TMLE protects against
  remaining misspecification through double robustness.
- **Interpretability hinges on scale.** A real-world evidence audience
  typically wants to know whether and where to deploy a treatment. That
  question lives on the additive scale. Multiplicative interaction
  remains useful for etiologic synthesis and meta-analytic comparison,
  but should not be the only summary in a decision-oriented report
  [@vanderweele2014tutorial; @basten2025additive].

A practical recommendation follows: in RWE, **prefer the additive scale
for decision-making, report the multiplicative scale for completeness,
and run sensitivity analyses that vary both the nuisance learner library
and the propensity-score model**.

::: {.callout-note}
## What this means in practice
Pre-specify the scale, use a single common reference category, and present
the cell-specific risks. If you have to pick one scale, pick additive for
public-health audiences and multiplicative for etiologic ones. When you
can, present both. In RWE, build the analysis to survive the joint-cell
positivity check and the alternative-learner sensitivity test.
:::

### Diagnostic checklist for interaction analyses

::: {.callout-tip}
## Five questions before you trust an interaction estimate

1. **Estimand.** Is interaction defined as a causal parameter
   ($\psi_{\text{add}}$, $\psi_{\text{mult}}$, RERI, or a contrast of
   subgroup risk differences), not as a regression coefficient?
2. **Positivity.** Is the joint-exposure cell or the modifier stratum
   adequately populated, with both treated and untreated patients
   represented?
3. **Learners.** Are flexible nuisance learners (Super Learner with a
   diverse library) used, rather than a single parametric model that
   could impose the wrong functional form?
4. **Scale.** Is the choice of additive vs. multiplicative scale
   justified by the decision question, with both reported when feasible?
5. **Sensitivity.** Are sensitivity analyses prespecified that vary the
   learner library, the adjustment set, and the propensity-score
   selection?
:::

---

## 6. Point-treatment worked example with TMLE and Super Learner

This section demonstrates the workflow on a simulated cohort with a binary
treatment $A$, a binary baseline modifier $M$, a binary outcome $Y$, and
three baseline confounders $W_1, W_2, W_3$. The data-generating mechanism
includes genuine effect modification on the additive scale. We present
two complementary implementations: a **joint-intervention plug-in TMLE**
that estimates the four counterfactual means directly, and a
**stratified TMLE** that fits separately within each modifier level. They
target the same parameter and should agree up to sampling variability.

::: {.callout-tip}
## Minimal implementation summary

1. **Estimate the outcome regression** $Q(A, E, W)$ with Super Learner.
2. **Predict under four interventions**: $(A,E) = (1,1), (0,1), (1,0), (0,0)$.
3. **Compute the additive interaction contrast**:
   $\widehat{\psi}_{\text{add}} = \overline{Q(1,1,W)} - \overline{Q(0,1,W)} - \overline{Q(1,0,W)} + \overline{Q(0,0,W)}$.
4. **Apply the TMLE update** using $g_A(A\mid W)$ and $g_E(E\mid W)$.
5. **Use the influence curve** for inference (Wald confidence intervals).
:::

### Simulated data

```{r}
#| label: sim-data
#| eval: false

library(dplyr)
set.seed(2026)

# Simulate a cohort of 2,000 patients. n = 2000 keeps the Super Learner
# + four-cell counterfactual prediction step fast in both the static
# and the WebR (interactive) builds. Production analyses can use any n.
n  <- 2000
W1 <- rnorm(n)                                       # continuous confounder
W2 <- rbinom(n, 1, 0.4)                              # binary confounder
W3 <- runif(n, -1, 1)                                # bounded confounder

# Second exposure E (here playing the role of a baseline modifier M)
E  <- rbinom(n, 1, plogis(-0.2 + 0.7 * W1 - 0.4 * W2))

# Treatment depends on confounders and the second exposure
A  <- rbinom(n, 1, plogis(-0.1 + 0.8 * W1 + 0.5 * W2 -
                            0.3 * W3 + 0.2 * E))

# Outcome with genuine additive-scale interaction between A and E
linpred_y <- -2.0 + 0.8 * A + 0.4 * E + 0.9 * A * E +
             0.7 * W1 - 0.5 * W2 + 0.3 * W3
Y <- rbinom(n, 1, plogis(linpred_y))

dat <- data.frame(Y, A, E, W1, W2, W3)
```

### Estimating interaction using TMLE: the joint-intervention workflow

The data structure is $O = (W, A, E, Y)$ and the estimation steps follow
the five-step mapping from Section 4:

1. **Estimate the outcome regression** $Q(A, E, W) = E[Y \mid A, E, W]$
   with Super Learner.
2. **Estimate the treatment mechanisms** $g_A(A \mid W)$ and
   $g_E(E \mid W)$ (or their joint $g(A, E \mid W)$) with Super Learner.
3. **Generate counterfactual predictions** under each of the four joint
   regimes: $Q(1,1,W)$, $Q(0,1,W)$, $Q(1,0,W)$, $Q(0,0,W)$.
4. **Plug into the interaction estimand**:
   $\widehat{\psi}_{\text{add}} = \frac{1}{n}\sum_i \big[Q(1,1,W_i) - Q(0,1,W_i)\big] - \frac{1}{n}\sum_i \big[Q(1,0,W_i) - Q(0,0,W_i)\big]$.
5. **Apply the TMLE targeting step** so that the plug-in solves the
   efficient influence-function equation, then **use the influence curve**
   for Wald confidence intervals.

In plain English: we predict outcomes under four hypothetical worlds —
everyone exposed to both, exposed to one, exposed to the other, or
neither — and compare the differences across those worlds.

```{r}
#| label: sl3-tmle3-joint-intervention
#| eval: false

# Joint-intervention plug-in g-computation skeleton
# (extend with tmle3 or ltmle for the full targeting step and IC inference)
library(sl3)
library(tmle3)
library(dplyr)

# Define a Super Learner library for nuisance estimation
lrn <- Lrnr_sl$new(
  learners = list(
    Lrnr_glm_fast$new(),
    Lrnr_ranger$new(),
    Lrnr_xgboost$new()
  )
)

# Outcome regression task: predict Y from (A, E, W)
Q_task <- make_sl3_Task(
  data       = dat,
  outcome    = "Y",
  covariates = c("A", "E", "W1", "W2", "W3")
)
Q_fit <- lrn$train(Q_task)

# Generate counterfactual predictions under each of four joint regimes.
# sl3 uses R6 classes, so call $predict() on the fitted learner rather
# than the generic predict() function.
predict_under <- function(a_val, e_val) {
  newdat <- mutate(dat, A = a_val, E = e_val)
  Q_fit$predict(make_sl3_Task(
    data       = newdat,
    outcome    = "Y",
    covariates = c("A", "E", "W1", "W2", "W3")
  ))
}

Q_11 <- predict_under(1, 1)
Q_01 <- predict_under(0, 1)
Q_10 <- predict_under(1, 0)
Q_00 <- predict_under(0, 0)

# Plug-in estimate of the additive interaction parameter
psi_add <- mean(Q_11 - Q_01) - mean(Q_10 - Q_00)

# Plug-in estimate of the multiplicative interaction parameter
psi_mult <- (mean(Q_11) / mean(Q_01)) /
            (mean(Q_10) / mean(Q_00))

list(psi_add = psi_add, psi_mult = psi_mult)
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"We trained a flexible model that predicts outcomes from treatment, the
second exposure, and patient characteristics. Then we asked the model:
*what would happen if everyone got both, neither, or just one of the
two?* The interaction is how much the effect of treatment changes when
the second exposure is also present."
:::

::: {.callout-warning}
## From plug-in g-computation to TMLE
The block above is the **plug-in g-computation** step. To obtain a true
TMLE---with the targeting update and influence-function-based
inference---wrap the same nuisance estimates in the `tmle3` workflow
with a custom parameter for $\psi_{\text{add}}$ or $\psi_{\text{mult}}$,
or fit `ltmle` with two intervention nodes ($A$ and $E$) and contrast
the four resulting counterfactual means. The plug-in estimate is correct
in expectation under correct $Q$; the targeting step adds robustness
when $Q$ is misspecified but $g$ is consistent.
:::

### Diagnose overlap within joint exposure cells

Before estimating any causal contrast, fit the propensity score and
inspect the distribution within each level of $E$. This is the
joint-cell version of the standard positivity diagnostic introduced in
[Chapter 3.2](03-02-positivity-overlap.qmd).

```{r}
#| label: overlap-check
#| eval: false

library(ggplot2)

dat$ps <- predict(
  glm(A ~ W1 + W2 + W3 + E, data = dat, family = binomial),
  type = "response"
)

ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ paste0("E = ", E)) +
  labs(x = "Estimated propensity score",
       fill = "Treatment",
       title = "Treatment overlap within each level of the second exposure")
```

### Example output and interpretation

The four counterfactual means produced by the joint-intervention TMLE
populate a 2×2 cell-specific risk table. Numbers are illustrative.

| | $E = 0$ | $E = 1$ |
|---|---:|---:|
| **$A = 0$** (risk if untreated) | 0.10 | 0.12 |
| **$A = 1$** (risk if treated)   | 0.15 | 0.28 |

From these four numbers the analyst reads off both interaction
parameters directly:

- $\widehat{\psi}_{\text{add}} = (0.28 - 0.12) - (0.15 - 0.10) = 0.11$.
  Treatment is 11 percentage points more harmful (or beneficial) when
  $E = 1$ than when $E = 0$.
- $\widehat{\psi}_{\text{mult}} = (0.28 / 0.12) \div (0.15 / 0.10) =
  1.55$. The treated-versus-untreated risk ratio is about 1.5 times
  larger when $E = 1$.

Both scales agree about the direction; only the additive scale tells
you the absolute size of the heterogeneity that matters for clinical
decision-making.

```{mermaid}
xychart-beta
    title "Joint-intervention counterfactual risks from TMLE"
    x-axis ["A=0,E=0", "A=1,E=0", "A=0,E=1", "A=1,E=1"]
    y-axis "Estimated risk" 0 --> 0.30
    bar [0.10, 0.15, 0.12, 0.28]
```

### Alternative: stratified TMLE when $E$ is a non-manipulable modifier

When the second variable is a non-manipulable baseline modifier rather
than an intervenable exposure, the natural analysis is stratified TMLE:
fit a separate TMLE within each level of the modifier, then contrast
the subgroup-specific risk differences. That is the standard
effect-modification workflow and is covered in detail in
[Chapter 3.4](03-04-effect-modification.qmd) §5–6, including code,
positivity diagnostics within strata, and uncertainty propagation. For
interaction analyses, the joint-intervention approach above is preferred
whenever both variables are conceptually manipulable, because it
estimates the four counterfactual means directly and avoids the
small-stratum instability of a separate TMLE per cell.

::: {.callout-note}
## What this means in practice
The cell-specific risk table is the single most useful artifact from an
interaction analysis. From it, any reader can recompute the additive
contrast, the multiplicative contrast, the number needed to treat in
each subgroup, and any other quantity. Lead with that table; let the
contrasts follow.
:::

---

## 7. CTMLE as a sensitivity analysis

Collaborative TMLE (CTMLE) is covered in detail in
[Chapter 2.5](02-05-advanced-tmle.qmd). This section explains only when
it is worth running for an interaction analysis.

CTMLE adaptively selects which covariates enter the propensity-score
model, guided by whether including them improves the bias-variance
trade-off for the target parameter. That makes it most useful exactly
where interaction analyses tend to suffer: in joint exposure cells with
a small number of patients and extreme propensity-score weights driven
by covariates that are strongly associated with treatment but only
weakly with the outcome.

Use CTMLE as a sensitivity check when standard TMLE looks fragile —
typically when fitted propensity scores have heavy tails inside one or
more joint cells, when the confidence interval is wide and appears
driven by a handful of high-weight observations, or when the headline
estimate moves substantially after dropping those observations. If the
CTMLE estimate agrees with the standard TMLE, the result is more
credible. If it disagrees, the standard TMLE was probably being pulled
by a fragile propensity model. See [Chapter 2.5](02-05-advanced-tmle.qmd)
§C-TMLE for the implementation details and code; the same workflow
applies to interaction analyses, with the propensity-score model now
including both $A$ and $E$ as treatments and the joint cell of interest
specified as the conditioning set.

---

## 8. Multiplicity, reporting, and interpretation

Interaction analyses are uniquely vulnerable to multiplicity. The risk
is not theoretical: many published claims of subgroup heterogeneity have
not replicated. Three disciplines protect the analysis.

### Pre-specification and the confirmatory-exploratory split

Before unblinding to the outcome, write down:

- **Confirmatory modifiers**: a small number (typically two to five),
  chosen for clinical or biological motivation. These get formal
  inference with appropriate multiplicity adjustment.
- **Exploratory modifiers**: any additional candidates. These are
  reported descriptively, with prominent caveats, and are best treated as
  hypothesis-generating.

The PATH statement and methodological work on heterogeneity of treatment
effects converge on this distinction. Recent simultaneous-inference
methods for multiple subgroup TMLEs make confirmatory analysis with
several modifiers tractable [@wei2023simultaneous].

### Multiplicity adjustment

For confirmatory work with several prespecified modifiers:

- Simultaneous (multivariate) confidence intervals around the vector of
  subgroup contrasts are the most informative single-step adjustment.
- False-discovery-rate control (Benjamini-Hochberg) is appropriate when
  the goal is to flag a subset of subgroups for further study.
- Bonferroni adjustment is conservative but easy to defend when the
  prespecified set is small.

### Reporting essentials

A well-reported interaction analysis includes, at minimum:

- The cell-specific risks or means with a single common reference
  category.
- Subgroup-specific risk differences and risk ratios with appropriate
  confidence intervals.
- The additive and multiplicative interaction contrasts with confidence
  intervals.
- Diagnostic information about overlap within strata.
- A brief description of the Super Learner library, the targeting step,
  and the variance estimator.
- Sensitivity analyses (e.g., CTMLE, alternative learner libraries,
  alternative confounder sets).
- An explicit statement about coding decisions and the reference
  category.

### Interpretation in absolute terms

Whenever the outcome scale permits, translate findings into absolute risk
or rate language. "The treatment reduces 5-year mortality by 8
percentage points among high-risk patients but only by 1 percentage
point among low-risk patients" is far more useful for clinical decisions
than "the relative-risk reduction is greater in high-risk patients."

### Common pitfalls

::: {.callout-warning}
## Common pitfalls in interaction analyses

- **Treating regression interaction coefficients as causal interaction.**
  A logistic or Cox product term is a property of the model's
  parameterization, not of the joint potential-outcome distribution.
- **Ignoring positivity in joint exposure cells.** A cohort that
  supports the main effect of $A$ may not support estimation of the
  $A \times E$ interaction; check the $(A, E)$ cell counts and the
  within-cell propensity distribution before interpreting any
  interaction estimate.
- **Choosing the scale post hoc.** Looking at both scales and reporting
  the one that gives a "significant" answer is a multiplicity violation
  in disguise. Pre-specify the scale, and report both.
- **Overinterpreting noisy estimates.** Interaction estimates are
  intrinsically noisier than main effects because they are differences
  of differences. A wide confidence interval is a finding, not a
  failure; the right response is to report it honestly, not to dichotomize
  on a p-value.
- **Confusing exploratory subgroup screening with confirmatory
  inference.** A subgroup that "popped out" of an exploratory scan must
  be replicated before it informs a clinical or policy decision.
:::

### Sensitivity analyses for interaction estimates

Because interaction estimates are derivatives of the outcome surface,
they are more sensitive to nuisance-model choices than main effects. A
short, prespecified sensitivity battery should accompany every
confirmatory interaction analysis:

- **Vary the nuisance learner library.** Re-run the analysis with a
  smaller library (e.g., GLM only) and a larger one (e.g., adding
  gradient boosting and neural-net learners). Stable estimates across
  libraries are reassuring; unstable estimates flag library or
  positivity sensitivity.
- **Vary the adjustment set.** Re-run with a minimal sufficient
  adjustment set and with a richer one. Large changes suggest residual
  confounding or model dependence.
- **Run a CTMLE sensitivity analysis** (Section 7) to check whether
  collaborative propensity-score selection moves the estimate.
- **Quantify unmeasured confounding tolerance** using a causal-gap or
  E-value-style framework: how strong would an unmeasured confounder
  need to be to explain away the observed interaction? This kind of
  quantitative bias analysis is especially valuable for interaction,
  where unmeasured confounders can act asymmetrically across exposure
  cells.

::: {.callout-note}
## What this means in practice
Pre-specify a short confirmatory list. Apply formal multiplicity control
to it. Report exploratory findings transparently, but with language that
makes their status clear. Translate every conclusion into absolute terms
at least once. Treat the sensitivity battery---learner library,
adjustment set, CTMLE, unmeasured-confounding bound---as part of the
primary analysis rather than as an afterthought.
:::

---

## 9. Software for interaction analyses

The general R targeted-learning stack — `SuperLearner`, `tmle`, `tmle3`,
`sl3`, `ltmle`, `drtmle`, `ctmle` — is introduced in
[Chapter 2.3](02-03-doubly-robust-tmle.qmd),
[Chapter 2.4](02-04-superlearner.qmd), and
[Chapter 2.5](02-05-advanced-tmle.qmd). The table below highlights only
the packages that earn special mention for interaction work.

| Package | Why it matters for interaction |
|---|---|
| `tmle3` + `sl3` | The joint-intervention plug-in workflow in Section 6 maps cleanly onto the modular tlverse grammar; defining a custom interaction parameter is more natural here than in `tmle`. |
| `ltmle` | Sustained-treatment interactions and TMLE for marginal structural models with $M$ as an interaction term (Section 11). |
| `lmtp` | Continuous or shift-style exposures, where the interaction question is reframed as a modified treatment policy (Section 11). |
| `InterXshift` / `JointXshift` | Pairwise additive-scale interaction in mixed exposures via stochastic shifts, with CV-TMLE inference (Section 11). |
| `IsoXshift` | Isobologram-style synergy: smallest joint shift achieving a target outcome level (Section 11). |
| `CVtreeMLE` | Cross-validated decision-tree search for threshold combinations across multiple exposures, with TMLE inference (Section 11). |
| `SuperNOVA` | Automatic discovery of variables, exposure pairs, and exposure-covariate combinations driving an outcome under stochastic intervention (Section 11). |
| `drtmle` | Aggressive Super Learner libraries combined with sparse joint cells push beyond the regularity conditions of standard TMLE; `drtmle` provides additional robustness. |
| `ctmle` | Sensitivity analysis when joint-cell propensity scores are fragile (Section 7). |

The standard reproducibility scaffolding — fixed seed, scripted
pipeline, documented analytic dataset, `sessionInfo()` — applies as in
every other chapter of this handbook.

---

## 10. Practical checklist for analysts

The checklist below summarizes the chapter into a single artifact. It can
be copied into a project's pre-analysis plan or a reviewer's referee
form.

| Step | Recommendation | Common failure mode |
|---|---|---|
| Question | State whether you study effect modification of one treatment or interaction of two exposures | Treating any product term as "interaction" |
| Estimand | Pre-specify additive and multiplicative estimands separately | Reporting only odds-ratio interaction |
| Design | Write a target trial protocol; define time zero, censoring, follow-up | Building subgroup models without a protocol |
| Modifiers | Limit confirmatory analyses to a small set of clinically motivated, pre-treatment modifiers | Mining many post-hoc modifiers |
| Confounding | Adjust for confounders of treatment and outcome only; avoid post-treatment variables | Overadjustment and collider bias |
| Positivity | Diagnose overlap overall *and within each modifier stratum* | Stratum-specific positivity failures hidden by overall overlap |
| Nuisance | Use a Super Learner library that is diverse but appropriate for the sample size | One opaque learner with no diagnostics |
| Estimation | TMLE within strata for small numbers of subgroups; simultaneous one-step TMLE for many | Naive stratified regression |
| Inference | Influence-function-based SEs; bootstrap for derived contrasts; cross-fitting for aggressive learners | Wald tests on derived contrasts assuming independence |
| Multiplicity | Confirmatory vs. exploratory split; simultaneous CIs or FDR for many modifiers | Many unadjusted subgroup tests |
| Reporting | Cell-specific risks; both scales; common reference category; coding decisions; sensitivity analyses | Only a p-value for the product term |
| Interpretation | Lead with absolute risks; sense-check against background incidence | Headline-only relative-risk language |
: Table 5. Practical checklist for interaction analyses in RWE. {.striped}

---

## 11. Time-varying interactions and marginal structural models

Sections 1 through 10 dealt with single time-point analyses: a baseline
treatment, a baseline modifier or second exposure, a single outcome
measurement. Many real-world questions involve treatment regimens that
evolve over time, with intermediate covariates that are simultaneously
confounders of later treatment and consequences of earlier treatment.
In that setting, baseline-only TMLE is no longer sufficient, and the
interaction question must be reframed in terms of sustained or dynamic
treatment strategies.

### When you need longitudinal methods

Longitudinal targeted learning becomes necessary when at least one of
the following is true:

- Treatment can change after baseline through initiation,
  discontinuation, or switching.
- Censoring is informative and is itself influenced by time-varying
  covariates.
- Time-varying confounders are themselves affected by past treatment,
  for example blood pressure during a hypertension trial.

For interaction analyses specifically, the question becomes: *does the
effect of a sustained treatment strategy differ across baseline
strata?* The modifier $M$ should still be a baseline variable, well
measured, and adequately supported across the relevant treatment
histories. Time-varying modifiers can also be of interest, but they
introduce additional positivity and identification challenges and are
typically handled inside a marginal structural model rather than by
stratification.

### LTMLE for sustained-treatment interactions

The `ltmle` package implements longitudinal TMLE for sequential
treatment and censoring nodes. It accepts a Super Learner library and
supports both static and dynamic intervention rules.

```{r}
#| label: ltmle-sketch
#| eval: false

library(ltmle)

# Sketch only: the data 'long_dat' is in wide format with
# baseline covariates, time-varying covariates L1..L3,
# treatments A1..A3, and outcome Y at the final time
fit_ltmle <- ltmle(
  data       = long_dat,
  Anodes     = c("A1", "A2", "A3"),
  Lnodes     = c("L1", "L2", "L3"),
  Ynodes     = "Y",
  abar       = list(c(1, 1, 1), c(0, 0, 0)),
  SL.library = c("SL.mean", "SL.glm", "SL.glmnet", "SL.ranger")
)

summary(fit_ltmle)
```

To estimate baseline-modifier-specific effects, fit `ltmle` separately
within each level of $M$ and contrast the resulting counterfactual means.
The output is an estimate of, for example, the one-year survival
probability under "always treat" versus "never treat" within each
subgroup, along with the contrast across subgroups.

::: {.callout-tip}
## How to explain this to a non-methodologist
When treatment can change month to month and intermediate health
markers both depend on past treatment and predict future treatment, a
single baseline analysis is no longer enough. The longitudinal version
of TMLE walks through the data one time step at a time, adjusting at
each step for what has already changed.
:::

### Marginal structural models for interaction

An alternative is to fit a marginal structural model (MSM) with the
modifier $M$ entering as an interaction term. The MSM specifies how the
counterfactual mean depends on the treatment regime and on $M$, and the
interaction coefficient in the MSM is a parameter of the marginal
counterfactual distribution rather than of any conditional outcome
model. TMLE for MSMs estimates this coefficient by fitting nuisance
regressions, computing counterfactual outcomes under the regimes
indexed by the MSM, and then targeting the MSM coefficient directly.
The `tmleMSM` framework and the corresponding `ltmle` MSM facilities
support this workflow.

In an MSM-based interaction analysis the analyst should be explicit
about: which regimes the MSM indexes, whether $M$ is baseline or
time-varying, and how the MSM is parameterised on the additive or
multiplicative scale. The same scale considerations from Section 5
apply: pre-specify the scale, use a common reference, and report the
cell-specific counterfactual means whenever possible.

::: {.callout-note}
## What this means in practice
For sustained-treatment interaction questions, fit LTMLE within each
modifier stratum or fit a TMLE-MSM with the modifier as an interaction
term. Report the cell-specific counterfactual outcomes — for example,
one-year survival under "always treat" versus "never treat" within
each subgroup — and the across-subgroup contrast. Diagnose positivity
across treatment histories, not only at baseline.
:::

### Flexible estimands beyond binary exposures

The estimands defined in Section 2 assume binary exposures. In
real-world evidence, exposures are often continuous (dose, duration,
intensity), and joint positivity for any single value of a continuous
exposure is typically zero. The remedy is to redefine the estimand in
terms of **modified treatment policies** (MTPs): hypothetical
interventions that shift observed exposure rather than fix it at a
value [@diaz2021lmtp]. The interaction question then becomes: *does
the effect of shifting $A$ depend on the level or shift of $E$?*

An MTP-based interaction parameter contrasts the counterfactual mean
under a paired shift in $(A, E)$ against the counterfactual mean under
shifting only $A$, only $E$, or neither. The general-purpose `lmtp`
package implements TMLE and sequentially doubly robust estimators for
arbitrary MTPs, with support for time-varying treatments, censoring,
and Super Learner nuisance estimation [@diaz2021lmtp]. For
continuous-exposure interaction problems — common in nutritional
epidemiology, environmental health, and dose-response
pharmacoepidemiology — MTPs are the appropriate generalisation of the
binary joint-intervention framework.

::: {.callout-tip}
## How to explain this to a non-methodologist
When exposures are continuous, asking "what if everyone had exactly
this dose?" is unrealistic and unsupported by the data. Instead, ask
"what if everyone shifted their exposure a bit?" The interaction
question becomes whether the benefit of shifting one exposure depends
on the shift in the other.
:::

### The `Xshift` family: TMLE for interactions in mixed exposures

A focused toolkit for stochastic-intervention interaction analysis
under mixed exposures — typical of environmental epidemiology — is the
`Xshift` family of R packages developed by McCoy and colleagues
[@mccoy2023interaction]. The packages share a common philosophy: define
the interaction parameter directly in terms of *contrasts of
counterfactual means under stochastic shifts*, and estimate it with
cross-validated TMLE (CV-TMLE) so that the data-adaptive selection of
which exposures to interrogate does not invalidate the confidence
intervals.

| Package | Estimand | Best use |
|---|---|---|
| [`InterXshift`](https://github.com/blind-contours/InterXshift) / [`JointXshift`](https://github.com/blind-contours/JointXshift) | Counterfactual mean under a *joint* shift of two exposures, contrasted with the *sum* of the two individual-shift means. Positive values indicate synergy; negative values indicate antagonism. | Pairwise interaction in a mixture, with valid CIs on the joint-vs-additive contrast |
| [`EffectXshift`](https://github.com/blind-contours/EffectXshift) | Counterfactual mean under a stochastic shift of one exposure, contrasted across data-adaptively learned subgroups | Effect modification of a continuous-exposure shift across covariate strata |
| [`IsoXshift`](https://github.com/blind-contours/IsoXshift) | Smallest joint shift in $(A, E)$ that achieves a target counterfactual outcome level, in the spirit of an isobologram | Policy-oriented synergy: which combinations of exposure reductions are most efficient? |

A typical `InterXshift` call looks like:

```{r}
#| label: interxshift-sketch
#| eval: false

library(InterXshift)

results <- InterXshift(
  w       = covariates,         # confounders (data frame)
  a       = exposures,          # mixed exposure matrix
  y       = outcome,            # outcome vector
  delta   = list(X1 = 1, X2 = 1),  # candidate shifts for each exposure
  n_folds = 5,                  # CV-TMLE folds
  top_n   = 3                   # how many exposure pairs to rank
)
```

The output is a ranked list of exposure pairs with the joint-shift
counterfactual mean, the sum-of-individual-shifts counterfactual mean,
their CV-TMLE-based contrast, and an associated 95% confidence interval
and $p$-value for each pair.

::: {.callout-note}
## When to reach for the `Xshift` family
- The exposures are continuous or many-valued (e.g., metals
  concentrations, pesticide loads, social-determinant indices), and
  fixing them at single values is biologically or pragmatically
  unrealistic.
- The analyst wants the additive-scale interaction parameter without
  collapsing exposures to binary categories.
- Several candidate exposure pairs are of interest, and the analysis
  needs valid confidence intervals despite data-adaptive ranking.
:::

### Discovery: which exposures interact?

The single-time-point worked example in Section 6 assumes the analyst
already knows which two exposures to interrogate. When the candidate
pool is larger — a typical mixture has 5 to 20 exposures — the
question shifts to *which pairs interact at all*, with valid inference
on the answer. Two TL-based discovery tools complement the
hypothesis-driven framework:

- [`CVtreeMLE`](https://github.com/blind-contours/CVtreeMLE) combines a
  cross-validated decision-tree search with TMLE to identify *threshold
  combinations* across multiple exposures that minimise or maximise
  outcome risk [@mccoy2023cvtreemle]. It produces $k$-fold-specific
  estimates with confidence intervals and a pooled oracle estimate
  with multiplicity-adjusted $p$-values.
- [`SuperNOVA`](https://github.com/blind-contours/SuperNOVA) uses
  basis-function estimators (MARS via `earth`, the highly adaptive
  lasso via `hal9001`) and an ANOVA-style variance decomposition to
  rank variables, exposure pairs, and exposure-covariate combinations
  by their contribution to the counterfactual outcome under stochastic
  intervention. It then estimates the relevant target parameters with
  TMLE on a held-out estimation sample.

These tools play, in the interaction setting, the same role that
treatment-effect variable importance ([Chapter 3.4](03-04-effect-modification.qmd) §9)
plays for effect modification: they rank candidate parameters with
valid inference, so that follow-up confirmatory work can be focused on
the interactions that actually appear to matter. As with all discovery
methods, findings should be replicated or prespecified in a
follow-up study before being treated as confirmatory.

---

## 12. Conclusion

Interaction analyses in real-world evidence research are easy to do
poorly and demanding to do well. The path forward is not a single new
estimator, but a shift in workflow: start with a clearly defined causal
estimand on a chosen scale, build a target trial protocol around it, use
flexible nuisance estimation through Super Learner, target the resulting
estimates with TMLE, propagate uncertainty for the contrasts you report,
and discipline yourself to distinguish confirmatory from exploratory
work. The targeted-learning toolkit aligns naturally with these
disciplines, supports both binary point-treatment and time-varying
extensions, and---through CTMLE and `drtmle`---offers principled
sensitivity options when nuisance estimation is fragile.

### Minimum reporting for interaction analyses in RWE papers

At a minimum, a published interaction analysis in real-world evidence
should report: (1) a precisely worded scientific question and the chosen
estimand; (2) the target trial protocol (eligibility, time zero,
treatment strategies, follow-up, censoring, outcome); (3) the
prespecified set of confirmatory modifiers and the exploratory set, with
multiplicity strategy; (4) cell-specific risks under treatment and
control (or under each joint exposure level), with a common reference
category; (5) subgroup-specific risk differences and risk ratios with
confidence intervals; (6) additive and multiplicative interaction
contrasts with confidence intervals; (7) overlap diagnostics within
strata; (8) the Super Learner library and the variance estimator used;
(9) at least one sensitivity analysis, ideally including a CTMLE or
alternative-library check; and (10) an explicit interpretation in
absolute-risk language. With these elements in place, a reader can
reconstruct the estimand, judge the design, and assess the credibility
of the heterogeneity claim.

### What comes next

Two natural follow-up questions arise after a careful interaction
analysis. The first is *how do we know the analysis is trustworthy?*
That question is the focus of the chapter on diagnostics and
sensitivity analysis ([Chapter 3.12](03-12-sensitivity-analysis.qmd)),
which covers identifiability checks, nuisance-model assessment,
positivity diagnostics, and quantitative bias analysis for unmeasured
confounding. The second is *which covariates contribute most to the
heterogeneity we found?* That question is the focus of the companion
chapter on effect modification
([Chapter 3.4](03-04-effect-modification.qmd)), which introduces
treatment-effect variable importance measures and causal forests as
tools for ranking and visualising the modifiers behind a contrast.

---

## Recommended reading

- VanderWeele and Knol [-@vanderweele2014tutorial]. The single best
  concise reference on additive versus multiplicative interaction and
  reporting standards.
- Petersen and van der Laan [-@petersen2014causal]. The causal roadmap
  applied to epidemiologic practice.
- Hernán and Robins [-@hernan2016target]. Target trial emulation as the
  organizing principle for RWE design.
- Talbot et al. [-@talbot2025tmle]. Modern best-practice tutorial for
  TMLE in applied work.
- van der Laan and Rubin [-@vanderLaan2006tmle]. Original TMLE paper.
- van der Laan, Polley, and Hubbard [-@vanderLaan2007superlearner]. Original
  Super Learner paper.
- Smith et al. [-@smith2023systematic]. Systematic review documenting
  the rapid uptake of TMLE in observational epidemiology.
- Wei et al. [-@wei2023simultaneous]. Simultaneous TMLE for multiple
  subgroup effects.
- Basten et al. [-@basten2025additive]. Recent treatment of additive
  interaction in IPD meta-analysis.
- McCoy et al. [-@mccoy2023interaction]. Stochastic-intervention TL for
  interaction and effect modification in mixed exposures; companion to
  the `Xshift` family of R packages.
- McCoy et al. [-@mccoy2023cvtreemle]. CVtreeMLE: cross-validated
  decision trees with TMLE for mixed-exposure threshold discovery.

## Annotated references

- **van der Laan and Rubin (2006).** Introduces TMLE as a plug-in
  estimator that is updated to solve the efficient influence function
  equation. The conceptual foundation for everything else in this
  chapter.
- **van der Laan, Polley, and Hubbard (2007).** Defines Super Learner as
  cross-validation-based stacking with an oracle inequality. Justifies
  using a diverse learner library without overcommitting to any single
  algorithm.
- **Petersen and van der Laan (2014).** A pedagogical formulation of the
  causal roadmap for epidemiologists. Recommended for any first reading
  on causal inference workflow.
- **VanderWeele and Knol (2014).** A tutorial that has shaped the modern
  reporting consensus on interaction. Especially clear on RERI, the
  reference-category problem, and scale dependence.
- **Hernán and Robins (2016).** The target-trial paper that has
  reorganized observational research around explicit protocol emulation.
  Required background for any RWE interaction analysis.
- **Smith et al. (2023).** A systematic review of TMLE in observational
  epidemiology. Useful for situating your analysis within current
  practice.
- **Talbot et al. (2025).** A best-practice tutorial that walks through
  TMLE implementation choices with applied researchers in mind.
- **Wei et al. (2023).** Simultaneous one-step TMLE for many subgroup
  effects, with valid joint confidence intervals. Read when you have
  more than a handful of modifiers.
- **Basten et al. (2025).** A clear recent treatment of additive
  interaction estimation, including in pooled / IPD settings. Useful
  for analysts whose decision question is public-health targeting.
- **McCoy et al. (2023, arXiv:2305.01849).** Develops the
  semi-parametric TL framework for interaction and effect modification
  via stochastic interventions on mixed exposures, and motivates the
  `InterXshift` / `EffectXshift` / `JointXshift` / `IsoXshift`
  packages.
- **McCoy et al. (2023, *JOSS*).** Companion paper to the `CVtreeMLE`
  package: cross-validated tree-based threshold discovery with TMLE
  inference for mixed-exposure policy questions.

---

## Appendix A: Full R code

The block below collects the joint-intervention worked example from
Section 6 into a single runnable script.

```{r}
#| label: appendix-a-full
#| eval: false

# ============================================================
# Appendix A. Joint-intervention TMLE for interaction
# Companion script to the chapter on interactions in RWE
# ============================================================

# --- Packages ---------------------------------------------------------
library(sl3)
library(tmle3)
library(dplyr)
library(ggplot2)

set.seed(2026)

# --- 1. Simulate cohort ------------------------------------------------
# n = 2000 keeps Appendix A fast in the WebR (interactive) build.
n  <- 2000
W1 <- rnorm(n)
W2 <- rbinom(n, 1, 0.4)
W3 <- runif(n, -1, 1)
E  <- rbinom(n, 1, plogis(-0.2 + 0.7*W1 - 0.4*W2))
A  <- rbinom(n, 1, plogis(-0.1 + 0.8*W1 + 0.5*W2 -
                            0.3*W3 + 0.2*E))
linpred_y <- -2.0 + 0.8*A + 0.4*E + 0.9*A*E +
             0.7*W1 - 0.5*W2 + 0.3*W3
Y <- rbinom(n, 1, plogis(linpred_y))

dat <- data.frame(Y, A, E, W1, W2, W3)

# --- 2. Joint-cell overlap diagnostic ---------------------------------
dat$ps <- predict(
  glm(A ~ W1 + W2 + W3 + E, data = dat, family = binomial),
  type = "response"
)
ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ paste0("E = ", E)) +
  labs(x = "Estimated propensity score",
       fill = "Treatment")

# --- 3. Outcome regression with Super Learner --------------------------
lrn <- Lrnr_sl$new(
  learners = list(
    Lrnr_glm_fast$new(),
    Lrnr_ranger$new(),
    Lrnr_xgboost$new()
  )
)

Q_task <- make_sl3_Task(
  data       = dat,
  outcome    = "Y",
  covariates = c("A", "E", "W1", "W2", "W3")
)
Q_fit <- lrn$train(Q_task)

# --- 4. Counterfactual predictions under four joint regimes -----------
# sl3 uses R6 classes, so call $predict() on the fitted learner rather
# than the generic predict() function.
predict_under <- function(a_val, e_val) {
  newdat <- mutate(dat, A = a_val, E = e_val)
  Q_fit$predict(make_sl3_Task(
    data       = newdat,
    outcome    = "Y",
    covariates = c("A", "E", "W1", "W2", "W3")
  ))
}
Q_11 <- predict_under(1, 1)
Q_01 <- predict_under(0, 1)
Q_10 <- predict_under(1, 0)
Q_00 <- predict_under(0, 0)

# --- 5. Plug-in interaction parameters --------------------------------
psi_add  <- mean(Q_11 - Q_01) - mean(Q_10 - Q_00)
psi_mult <- (mean(Q_11) / mean(Q_01)) /
            (mean(Q_10) / mean(Q_00))

list(psi_add = psi_add, psi_mult = psi_mult)

# To obtain a fully targeted estimate with influence-function-based
# inference, wrap the same nuisance estimates in a tmle3 workflow with
# a custom parameter for psi_add, or fit ltmle with two intervention
# nodes (A and E) and contrast the four counterfactual means.
# See Section 11 for time-varying / MSM extensions.
```

---

## Appendix B: Optional Python supplement

Python is supplementary, not central, to a targeted-learning workflow.
The most epidemiology-oriented option is `zEpid`, which provides risk
ratios and risk differences, interaction contrasts, and TMLE-style
estimators in an interface designed for epidemiologic practice.

```{python}
#| label: appendix-b-python
#| eval: false

# Optional Python supplement
# zEpid for effect measures and TMLE-style epidemiologic workflows
import pandas as pd
from zepid import RiskRatio, RiskDifference
from zepid.causal.doublyrobust import TMLE

# Assume df has columns: Y (outcome), A (treatment),
# E (second exposure), W1, W2, W3 (confounders)

# Risk difference within each level of the second exposure
for e_value in [0, 1]:
    d = df[df["E"] == e_value]
    rd = RiskDifference()
    rd.fit(d, exposure="A", outcome="Y")
    print(f"E = {e_value}")
    rd.summary()

# TMLE within each level of the second exposure
for e_value in [0, 1]:
    d = df[df["E"] == e_value].copy()
    tmle = TMLE(d, exposure="A", outcome="Y")
    tmle.exposure_model("W1 + W2 + W3")
    tmle.outcome_model("A + W1 + W2 + W3")
    tmle.fit()
    print(f"E = {e_value}: ATE = {tmle.average_treatment_effect}")
```

`DoWhy` and `CausalML` are useful complementary tools but are not primary
targeted-learning implementations. `DoWhy` provides causal-graph
specification, identification, and refutation tests; `CausalML` focuses
on conditional average treatment effects and uplift modeling, which are
better viewed as exploratory heterogeneity tools than as substitutes for
TMLE-based interaction inference. For an MPH chapter on interactions in
RWE, the R targeted-learning stack remains the recommended primary
implementation; the Python ecosystem is a useful complement.
