---
title: "Adherence, Forgiveness, and Distribution-Shift Interventions"
subtitle: "Comparing drugs under realistic adherence using LMTP"
format: html
---

*Estimated reading time: 55 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
library(tidyverse)
```


# What this chapter is about

The previous chapters introduced longitudinal TMLE for static interventions ("always maintain high adherence") and LMTP for modified treatment policies. This chapter brings those tools to bear on a clinically important question: how do two antiretroviral drugs compare when we account for realistic adherence patterns?

The key concept is **regimen forgiveness**: a drug is forgiving if patients can miss some doses without a large increase in treatment failure. A drug with a flat adherence-response curve is forgiving; one with a steep curve is not. Forgiveness is a causal quantity, not an observable property of the data, because the adherence-response relationship is confounded by time-varying health status.

We use a simulated two-drug HIV cohort designed to capture the essential features of the adherence problem: treatment-confounder feedback between adherence and health, differential forgiveness across drugs, and the positivity challenges that arise when targeting extreme adherence levels. The simulation is calibrated to real pharmacoepidemiologic settings and uses continuous proportion of days covered (PDC) as the adherence measure.


# The clinical question

Consider two antiretroviral regimens for HIV treatment:

- **Drug A** (e.g., a long-acting formulation): pharmacologically forgiving, meaning that occasional missed doses have a smaller impact on viral suppression.
- **Drug B** (e.g., a short half-life oral regimen): less forgiving, where missed doses lead to faster viral rebound.

In practice, both drugs are prescribed and adherence varies. The adherence distributions may differ across drugs (one may be easier to take consistently), and the clinical consequences of low adherence differ (one may be more forgiving). The causal question is: **how much of the observed outcome difference between drugs is due to adherence patterns versus pharmacological forgiveness?**

This question requires separating three components:

1. **Adherence distribution**: Do Drug A and Drug B patients adhere at different rates?
2. **Forgiveness**: At any given adherence level, does one drug produce better outcomes than the other?
3. **Adherence-mediated effect**: How much of the total drug difference operates through the adherence pathway?

Each component requires longitudinal causal methods because adherence and health co-evolve over time.


# The data-generating process

We use a simulation that generates longitudinal data with 90-day blocks over one year (K = 4 blocks). The DGP produces treatment-confounder feedback: poor adherence worsens health (higher L), and worse health reduces future adherence. Drug A is more forgiving at low adherence levels, encoded through a drug-by-adherence interaction in the failure hazard.

```{r}
#| label: sim-dgp
# Simplified DGP for the adherence forgiveness simulation
# Based on sim_v4_dgp.R from the adherence simulation project

simulate_adherence_data <- function(n = 500, K = 4, seed = 2026) {
  set.seed(seed)

  # Parameters
  alpha0 <- 1.30; alphaA <- -0.35; alphaW <- -0.20; alphaL <- -0.30
  rho <- 0.45; sigma_u <- 0.50; sigma_e <- 0.65
  beta0 <- 0.0; betaW <- 0.30; betaA <- 0.0; betaLag <- 0.35; betaP <- 0.80
  sigma_L <- 0.50
  gamma0 <- -3.80; gammaL <- 0.50; gammaW <- 0.30
  gammaP <- 2.50; gammaA <- 0.0; gammaAP <- -1.20

  # Baseline
  W0 <- rnorm(n)
  A0 <- rbinom(n, 1, 0.5)  # 1 = Drug A, 0 = Drug B
  u_i <- rnorm(n, 0, sigma_u)
  mu_i <- alpha0 + alphaA * A0 + alphaW * W0 + u_i

  # Storage
  L_mat <- PDC_mat <- matrix(NA_real_, n, K)
  Y_mat <- matrix(0L, n, K)
  eta_prev <- rep(NA_real_, n)
  failed <- rep(FALSE, n)

  for (t in 1:K) {
    # Time-varying health confounder L_t
    noise_L <- rnorm(n, 0, sigma_L)
    if (t == 1) {
      L_mat[, t] <- beta0 + betaW * W0 + betaA * A0 + noise_L
    } else {
      L_mat[, t] <- beta0 + betaW * W0 + betaA * A0 +
        betaLag * L_mat[, t-1] + betaP * (1 - PDC_mat[, t-1]) + noise_L
    }

    # Adherence (PDC) on logit scale with AR(1) persistence
    noise_e <- rnorm(n, 0, sigma_e)
    if (t == 1) {
      eta_t <- mu_i + noise_e
    } else {
      eta_t <- (1 - rho) * mu_i + rho * eta_prev + alphaL * L_mat[, t-1] + noise_e
    }
    PDC_mat[, t] <- plogis(eta_t)
    eta_prev <- eta_t

    # Virologic failure (absorbing)
    at_risk <- !failed
    if (any(at_risk)) {
      lp <- gamma0 + gammaL * L_mat[at_risk, t] + gammaW * W0[at_risk] +
        gammaP * (1 - PDC_mat[at_risk, t]) + gammaA * A0[at_risk] +
        gammaAP * A0[at_risk] * (1 - PDC_mat[at_risk, t])
      new_fail <- rbinom(sum(at_risk), 1, plogis(lp)) == 1L
      idx <- which(at_risk)
      Y_mat[idx[new_fail], t] <- 1L
      failed[idx[new_fail]] <- TRUE
    }
    if (t > 1) Y_mat[Y_mat[, t-1] == 1L, t] <- 1L
  }

  # Assemble wide data
  df <- data.frame(id = 1:n, W0 = W0, A0 = A0)
  for (t in 1:K) {
    df[[paste0("L_", t)]]   <- L_mat[, t]
    df[[paste0("PDC_", t)]] <- PDC_mat[, t]
    df[[paste0("Y_", t)]]   <- Y_mat[, t]
  }
  df$Y_final <- as.integer(rowSums(Y_mat) > 0)
  attr(df, "K") <- K
  df
}

dat <- simulate_adherence_data(n = 500, K = 4, seed = 2026)
```

The key structural features of this DGP:

**Treatment-confounder feedback.** Poor adherence (low PDC) at block $t-1$ worsens health ($L_t$), which then reduces adherence at block $t$ (through `alphaL`) and independently increases failure risk (through `gammaL`). This is the same feedback structure discussed in [Chapter 3.7](03-07-longitudinal-td-confounding.qmd).

**Drug-by-adherence interaction (the forgiveness mechanism).** The parameter `gammaAP = -1.20` means that Drug A's penalty for poor adherence is smaller than Drug B's. At high adherence, the two drugs perform similarly. At low adherence, Drug A is more forgiving. This interaction is what makes the forgiveness curve differ between drugs.

```{r}
#| label: fig-adherence-dist
#| fig-cap: "PDC distributions at block 1 by drug. Drug A patients have slightly lower mean adherence (the drug assignment shifts the adherence intercept), but Drug A is more forgiving of that lower adherence."
dat |>
  mutate(Drug = ifelse(A0 == 1, "Drug A", "Drug B")) |>
  ggplot(aes(x = PDC_1, fill = Drug)) +
  geom_density(alpha = 0.4) +
  labs(x = "PDC (block 1)", y = "Density") +
  scale_fill_manual(values = c("Drug A" = "#228833", "Drug B" = "#EE6677")) +
  theme_minimal()
```

```{r}
# Cohort summary
dat |>
  mutate(Drug = ifelse(A0 == 1, "Drug A", "Drug B")) |>
  group_by(Drug) |>
  summarise(
    n = n(),
    mean_PDC_1 = round(mean(PDC_1), 3),
    failure_pct = round(100 * mean(Y_final), 1),
    .groups = "drop"
  ) |>
  as.data.frame()
```


# Why naive adherence analysis is biased

A naive analysis would stratify patients by observed adherence level and compare outcomes within strata. This is systematically biased because of treatment-confounder feedback: patients with low adherence tend to be sicker (higher $L_t$), and sicker patients have higher failure risk regardless of adherence. The naive analysis overestimates the effect of adherence because the low-adherence stratum is enriched for patients who were going to fail anyway.

```{r}
#| label: fig-naive-forgiveness
#| fig-cap: "Naive (crude) adherence-outcome relationship. Each point shows the failure rate among patients with PDC in that bin during block 1. The steep relationship is partly causal and partly confounded by the treatment-confounder feedback."
dat |>
  mutate(
    Drug = ifelse(A0 == 1, "Drug A", "Drug B"),
    PDC_bin = cut(PDC_1, breaks = seq(0, 1, 0.1), include.lowest = TRUE)
  ) |>
  group_by(Drug, PDC_bin) |>
  summarise(failure_rate = mean(Y_final), n = n(), .groups = "drop") |>
  filter(n >= 5) |>
  ggplot(aes(x = PDC_bin, y = failure_rate, color = Drug, group = Drug)) +
  geom_point(size = 2) +
  geom_line() +
  labs(x = "PDC bin (block 1)", y = "Failure rate (crude)") +
  scale_color_manual(values = c("Drug A" = "#228833", "Drug B" = "#EE6677")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

The crude curves overestimate the adherence effect because they compare patients who are not exchangeable. The causal adherence-response curve (the forgiveness curve) requires g-methods that account for the feedback between adherence and health. The L-TMLE chapter ([Chapter 3.9](03-09-illustrated-ltmle.qmd)) showed how to estimate static adherence regimes; here we focus on the feasible shift approach using LMTP.


# LMTP estimation: feasible adherence shifts

Rather than targeting extreme static interventions ("force everyone to perfect adherence"), which violate positivity for patients who cannot realistically achieve high adherence, we use LMTP to define feasible adherence shifts [@diaz2021lmtp].

The shift function increases each patient's PDC by a fixed amount (e.g., 10 percentage points), capped at 1.0. This stays within the support of the observed data because no patient is assigned an adherence level that is impossible given their covariate history.

```{r}
#| label: lmtp-estimation
#| eval: false
library(lmtp)

# Prepare variable names
K <- 4
trt_vars     <- paste0("PDC_", 1:K)
outcome_vars <- paste0("Y_", 1:K)
baseline     <- c("W0", "A0")
time_vary    <- lapply(1:K, function(t) paste0("L_", t))

# Additive shift: increase PDC by 10 percentage points
shift_up <- function(data, trt) {
  pmin(data[[trt]] + 0.10, 1.0)
}

# Natural course (no intervention)
shift_natural <- function(data, trt) data[[trt]]

# Estimate outcomes under each regime
fit_shifted <- lmtp_tmle(
  data = dat, trt = trt_vars, outcome = outcome_vars,
  baseline = baseline, time_vary = time_vary,
  shift = shift_up, mtp = TRUE, outcome_type = "survival",
  learners_outcome = c("SL.glm", "SL.mean"),
  learners_trt = c("SL.glm", "SL.mean"),
  folds = 2
)

fit_natural <- lmtp_tmle(
  data = dat, trt = trt_vars, outcome = outcome_vars,
  baseline = baseline, time_vary = time_vary,
  shift = shift_natural, mtp = TRUE, outcome_type = "survival",
  learners_outcome = c("SL.glm", "SL.mean"),
  learners_trt = c("SL.glm", "SL.mean"),
  folds = 2
)

# Contrast: effect of shifting adherence up by 10pp
contrast <- lmtp_contrast(fit_shifted, ref = fit_natural, type = "additive")
contrast
```

The contrast estimates how much the failure rate would change if every patient's adherence increased by 10 percentage points at each visit. Because `lmtp` uses `outcome_type = "survival"`, it returns $P(\text{survive})$, so the risk is $1 - \theta$. A negative risk difference means the shift reduces failure.


# Alternative shift approaches

The additive shift is one way to operationalize "what if adherence were better?" Other approaches answer subtly different questions.

## Quantile matching: what if Drug B adhered like Drug A?

This maps each Drug B patient's adherence to the corresponding quantile of Drug A's distribution. A patient at the 30th percentile of Drug B adherence gets mapped to the 30th percentile of Drug A adherence. This answers: what would happen to Drug B's outcomes if its patients had Drug A's adherence profile?

```{r}
#| label: quantile-shift
#| eval: false
# Build quantile-matching shift function
ecdf_B <- list()
sorted_A <- list()
for (t in 1:K) {
  col <- paste0("PDC_", t)
  ecdf_B[[t]] <- ecdf(dat[[col]][dat$A0 == 0])
  sorted_A[[t]] <- sort(dat[[col]][dat$A0 == 1])
}

shift_match_A <- function(data, trt) {
  t_idx <- as.integer(gsub("PDC_", "", trt))
  a_obs <- data[[trt]]
  a_shifted <- a_obs
  is_b <- data$A0 == 0
  if (any(is_b)) {
    p_rank <- ecdf_B[[t_idx]](a_obs[is_b])
    p_rank <- pmax(0.01, pmin(0.99, p_rank))
    a_shifted[is_b] <- sorted_A[[t_idx]][
      pmin(length(sorted_A[[t_idx]]),
           ceiling(p_rank * length(sorted_A[[t_idx]])))
    ]
  }
  a_shifted
}
```

## Mean-calibrated additive shift

Add a constant to every Drug B patient's adherence so that Drug B's mean matches Drug A's mean. This is the simplest cross-drug comparison: it shifts the location of the distribution but preserves the shape.

## Fixed additive shift (policy evaluation)

The +10pp shift applied to everyone, regardless of drug. This answers a purely policy question: what would a universal adherence support program achieve? It does not reference the other drug at all and has the strongest positivity properties because the shift is small and uniform.


# The forgiveness curve

The forgiveness curve maps sustained adherence level to counterfactual failure risk for each drug:

$$
r_d(a) = P\big(Y = 1 \mid \text{do}(\bar{A} = a), D = d\big)
$$

A flat curve means the drug is forgiving (patients can miss doses without large increases in failure risk). A steep curve means small adherence drops produce large outcome differences.

Static-regime L-TMLE traces this curve by estimating the failure risk at a grid of fixed PDC thresholds. The LMTP shift approach provides a complementary view: it estimates the marginal benefit of a small adherence improvement, integrated over the observed adherence distribution. The two approaches are different summaries of the same counterfactual risk surface.

The curve is well-estimated only where there is adequate positivity. Below approximately PDC 0.70, few patients sustain such low adherence for four consecutive blocks, so the estimates become unstable. LMTP shift interventions avoid this problem because they stay close to the observed distribution.


# Interpretation

**Feasible, not hypothetical.** The +10pp shift corresponds to something a real adherence support program might achieve. The estimate answers a concrete policy question.

**Causal, not descriptive.** The LMTP framework accounts for the treatment-confounder feedback between adherence and health. When adherence changes, health status changes too, and those changes are propagated through the sequential regression structure.

**Drug-specific effects.** Running the analysis separately by drug reveals whether the marginal return to improved adherence differs by regimen. If Drug A is more forgiving, the benefit of an adherence intervention is smaller for Drug A patients (they are already doing relatively well at moderate adherence) and larger for Drug B patients.


# Connection to mediation

The quantile-matching analysis provides an informal decomposition of the total drug effect into an adherence-mediated component and a pharmacological (forgiveness) component. This is not formal mediation analysis; it is an approximation based on distribution shifts. For the formal TMLE-based mediation framework with proper identification conditions, see [Chapter 3.10](03-10-tmle-mediation.qmd).


## Sources and further reading

- Diaz I, Williams N, Hoffman KL, Schenck EJ (2023). Nonparametric causal effects based on longitudinal modified treatment policies. *Journal of the American Statistical Association* 118(542):846-856.
- Bangsberg DR (2006). Less than 95% adherence to nonnucleoside reverse-transcriptase inhibitor therapy can lead to viral suppression. *Clinical Infectious Diseases* 43(7):939-941.
- Petersen ML, Schwab J, Gruber S, et al. (2014). Targeted maximum likelihood estimation for dynamic and static longitudinal marginal structural working models. *J Causal Inference* 2(2):147-185.
- `lmtp` R package: [CRAN](https://cran.r-project.org/package=lmtp)
