---
format: html
---

# Estimating Interactions in Real-World Evidence Research {#sec-interactions-rwe}

*Estimated reading time: 50 minutes*

::: {.callout-tip}
## If you only read one thing
A causal interaction is about a joint intervention on two exposures, summarized by the additive interaction contrast $\psi_{\text{add}} = p_{11} - p_{10} - p_{01} + p_{00}$, the extra events per patient beyond what the two effects would produce separately. Estimate it with TMLE across the four joint-exposure cells, and check joint positivity first: those four cells are smaller than the margins, so overlap fails for an interaction long before it fails for a single exposure.
:::

## Abstract

Interactions are among the most clinically interesting and the most
methodologically demanding quantities in real-world evidence research.
Clinicians want to know whether a drug works equally well across patient
subgroups. Public-health planners want to know whether two exposures
combine to produce more harm than the sum of their parts. Many published
analyses answer these questions by reporting a single product-term
coefficient from a logistic or Cox model. This chapter argues for a
different starting point. Interaction analyses should begin with a
clearly defined causal estimand on a meaningful scale, anchored in a
target trial protocol, and should be estimated with methods that respect
the complexity of observational data. Targeted learning, which combines
flexible nuisance estimation through Super Learner
[@vanderLaan2007superlearner] with the targeting step and
influence-function inference of TMLE [@vanderLaan2006tmle], is
particularly well suited to this task.
The chapter walks through a binary point-treatment example with TMLE and
Super Learner, scaffolds a collaborative TMLE sensitivity analysis,
sketches a longitudinal extension using LTMLE and TMLE for marginal
structural models, and closes with a reporting checklist that aligns
with current best-practice recommendations
[@talbot2025tmle; @vanderweele2014tutorial; @basten2025additive].

::: {.callout-tip}
## Key takeaways

- **Effect modification** (one treatment, varying effect across a baseline
  variable) and **interaction** (two exposures acting jointly) are different
  scientific questions and require different estimands.
- The same data can show interaction on the **additive** scale and not on
  the **multiplicative** scale, or vice versa. Pre-specify the scale that
  matches your decision question.
- A regression product term is rarely the scientific target. Define
  **subgroup-specific risks** and contrast them.
- Begin every RWE interaction analysis by writing the **target trial
  protocol** [@hernan2016target]: time zero, eligibility, treatment
  strategies, follow-up, outcome, censoring.
- **Positivity is harder inside subgroups** than overall. Diagnose overlap
  within each modifier stratum.
- **Targeted learning** (Super Learner + TMLE) supports flexible nuisance
  estimation, doubly robust point estimation, and influence-function-based
  uncertainty for the contrasts you actually want to report.
- **CTMLE** is best framed as a sensitivity analysis around fragile
  propensity-score models, not as the introductory teaching vehicle.
- Pre-specify a small number of clinically motivated effect modifiers,
  separate confirmatory from exploratory work, and report **both scales**
  with cell-specific risks and a common reference category
  [@vanderweele2014tutorial].
:::

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

---

## 1. Why interaction matters in epidemiology and real-world evidence

A single average treatment effect is rarely the end of a clinically useful
analysis. Two follow-up questions almost always arise. *Does the effect
differ across patient groups?* And *do two exposures combine in ways that
one alone would not predict?* Both questions are about interaction in a
broad sense, but they are not the same question, and the methods that
answer them are not identical.

In real-world evidence studies, interaction questions arise because
exposures rarely occur one at a time. A patient with chronic hepatitis B
infection may also have diabetes, obesity, kidney disease, or a lipid
disorder. Some conditions may appear harmful when considered alone,
while others may appear neutral or even protective after adjustment for
the rest. The scientific question is then not simply whether each
condition is associated with advanced liver disease, but whether
combinations of conditions create more excess risk than would be
expected from considering each condition separately.

The same problem appears in co-medication studies. A drug may have one
safety profile when used alone and a different profile when taken with
another medication. Interaction analysis gives us a structured way to
ask whether the joint exposure pattern matters beyond the marginal
effects of each component. Under the target trial interpretation
described in @sec-roadmap, the joint
exposure pattern corresponds to a hypothetical intervention that sets
both exposures simultaneously, and the interaction parameter is the
contrast between this joint intervention and the sum of the two
single-exposure interventions.

::: {.callout-note}
## Scope of this chapter
This chapter focuses on **causal interaction**: the joint effect of two
manipulable exposures, and the contrast of subgroup-specific effects when
that contrast is itself the scientific target. The companion chapter,
[Effect Modification and Heterogeneous Treatment Effects](03-04-effect-modification.qmd),
covers the closely related but distinct question of how the effect of a
single treatment varies across baseline covariates (the CATE), with tools
including TMLE for prespecified subgroups, causal forests, and treatment-effect
variable importance.

The two chapters are designed to be read together. Effect modification
asks *who benefits more from one treatment*; interaction asks *how two
exposures combine*.

**Single time-point analyses** occupy Sections 1 through 10. **Time-varying
treatment, longitudinal interactions, and marginal structural models** are
concentrated in Section 11, and should be read after the single time-point
material is comfortable. Section 12 closes the chapter.
:::

Several closely related ideas are routinely conflated in published work.
Disentangling them is the first task of any chapter on interactions.

- **Effect modification** describes how the causal effect of a *single*
  treatment $A$ varies across levels of a baseline characteristic $V$.
  The target is a contrast of subgroup-specific causal effects, for example
  the risk difference if treated minus untreated within $V=1$ versus within
  $V=0$.
- **Interaction**, in the strict sense used by VanderWeele and Knol
  [-@vanderweele2014tutorial], refers to the *joint* causal effect of two
  exposures $A$ and $B$. Both are conceptually manipulable and the question
  is whether their combined effect departs from what you would expect from
  each effect alone.
- **Additive interaction** asks whether the joint effect departs from the
  *sum* of the individual effects on the absolute risk scale.
- **Multiplicative interaction** asks whether the joint effect departs from
  the *product* of the individual effects on the relative-risk scale. The
  same data can show interaction on one scale and none on the other.
- **Subgroup analysis** is a reporting style: presenting effects within
  prespecified strata. It is the *vehicle* for studying effect modification,
  but a list of subgroup point estimates is not by itself a heterogeneity
  test.
- **Predictive heterogeneity** refers to algorithms that learn high-dimensional
  functions of covariates to predict who benefits, for example the conditional
  average treatment effect surface produced by a causal forest. These tools are
  covered in the companion [Effect Modification](03-04-effect-modification.qmd)
  chapter and should not be confused with the inferential, low-dimensional
  interaction parameters that are the focus of this chapter.

A second motivation for taking interaction seriously is **scale dependence**.
A cohort study may find that an antihypertensive halves stroke risk in
low-risk patients (relative risk 0.5) and also halves stroke risk in
high-risk patients (relative risk 0.5). On the multiplicative scale there is
no interaction. On the additive scale, the absolute risk reduction is much
larger for the high-risk group, and that is what determines who should be
treated first. Public health and clinical decisions almost always depend on
absolute differences, not ratios [@vanderweele2014tutorial; @basten2025additive].

A third motivation is the rapid growth of TMLE and other doubly robust
methods in observational epidemiology. A 2023 systematic review identified
89 observational studies that applied TMLE [@smith2023systematic], and the
growth is steepest in pharmacoepidemiology and comparative effectiveness
work---exactly the settings where interaction questions are most common.

### Why interaction matters in practice

Interaction is not an academic embellishment of an average effect. It
is the analytic foundation for several of the most important applied
questions in modern epidemiology and clinical practice.

- **Treatment targeting.** When the absolute benefit of a treatment is
  much larger in one subgroup than another, fixed budgets and finite
  clinic time turn subgroup-aware deployment into a matter of
  efficiency. HAART offered preferentially to patients with the lowest
  baseline CD4 prevents more deaths per regimen-month than the same
  drug offered uniformly. Interaction estimates are what turn "treatment
  works on average" into "treat these patients first."
- **Policy heterogeneity.** Public-health interventions rarely have
  uniform effects across regions, age groups, or baseline-risk strata.
  A vaccination campaign that reduces hospitalisation by 10 percentage
  points among the elderly and 1 percentage point among the young
  calls for a different roll-out plan than one with uniform 5.5
  percentage-point effects. Interaction estimates inform where to
  invest the next public-health dollar.
- **Joint-exposure aetiology.** When two exposures may share a
  biological pathway, for example environmental tobacco smoke and
  occupational asbestos, interaction analyses on the additive scale
  carry information about whether the combined exposure produces
  excess events beyond what either alone predicts. That information
  matters for screening, regulation, and aetiologic theory.
- **Comorbidity-by-treatment interaction in pharmacoepidemiology.**
  Whether a drug's safety profile depends on a baseline comorbidity,
  for example whether the cardiovascular risk associated with a
  glucose-lowering agent differs in patients with versus without prior
  heart failure, is a recurrent question in comparative-safety
  studies. The same machinery applies here, with the comorbidity
  playing the role of the second exposure $E$.
- **Biomarker-by-biomarker interaction.** When two continuous clinical
  measurements (blood pressure and HbA1c, for example) jointly index
  cardiometabolic risk, the interaction question is whether their
  effects on the outcome compound beyond what either alone predicts.
  These analyses typically need the modified-treatment-policy
  framework introduced in Section 11 rather than binary-cell
  contrasts.

The common thread is that interaction estimates inform decisions and
explanations, not just summaries. That is why the rest of this chapter
spends so much energy on getting the estimand, the scale, and the
uncertainty right.

### Interaction vs. effect modification

Because these two ideas are conflated so often, it is worth restating the
distinction in operational terms:

- **Effect modification** is *heterogeneity in conditional effects*. The
  modifier $V$ need not be a manipulable exposure; it might be sex, a
  genotype, or a baseline biomarker. The estimand is the contrast of
  conditional treatment effects across levels of $V$.
- **Interaction** is a *contrast of joint interventions*. It requires that
  *both* $A$ and a second variable $E$ be conceptually amenable to
  intervention, and the estimand involves four counterfactual worlds in
  which $(A, E)$ is set jointly to $(0,0), (0,1), (1,0)$, and $(1,1)$.

The implication is methodological: a regression "interaction term" between
$A$ and $E$ does not estimate a causal interaction. It estimates a
parameter of a parametric model whose mapping to a causal contrast holds
only under strong assumptions. Many applied analyses report the
coefficient and call it interaction; this chapter argues that the causal
interaction parameter should be defined and estimated directly.

### Effect modification is not the same as causal interaction

Throughout this chapter, treat these as different scientific questions.
If only one exposure is interpreted as a manipulable treatment and the
second variable is a baseline characteristic such as age, sex, frailty,
or chronic kidney disease, the target is usually effect modification:
how much the treatment effect differs across levels of that
characteristic. If both variables are treated as interventions or
exposures whose joint effect is of interest, the target is causal
interaction. In that second case, the estimand is built from four
counterfactual mean outcomes under the joint exposure conditions
$(0,0)$, $(1,0)$, $(0,1)$, and $(1,1)$. This distinction matters
because identification assumptions, interpretation, and recommended
software all differ [@hernan2020whatif; @knol2012recommendations].

If the second variable is not itself assigned or conceptualised as an
intervention, avoid calling the estimand a causal interaction unless
that stronger interpretation is justified.

Before you write a single line of code, decide which of the four
ideas (effect modification, two-exposure interaction, additive
contrast, or multiplicative contrast) you are estimating. Different
choices imply different estimands, different reporting tables, and
different defaults for how to communicate uncertainty. Causal
interaction always involves comparing four hypothetical worlds;
effect modification compares two worlds within each level of a
stratifier.

---

## 2. From scientific question to causal estimand

The most common mistake in interaction analysis is to start with a model.
The right starting point is to write down the quantity you want to
estimate in counterfactual language, *before* any software is opened.
Product-term coefficients from logistic, log-binomial, or Cox
regressions are model parameters. They are not automatically the
causal estimand of interest.

### A. Joint interventions and four counterfactual means

Suppose two binary exposures $A$ and $E$ can each be intervened on, and
let $Y^{a,e}$ denote the potential outcome under joint intervention
setting $A=a$ and $E=e$. There are four such counterfactuals:
$Y^{0,0}$, $Y^{0,1}$, $Y^{1,0}$, $Y^{1,1}$. Write $p_{ae} = E[Y^{a,e}]$
for the marginal mean under the joint intervention. Every additive and
multiplicative interaction summary below is a function of those four
numbers.

These quantities are **parameters of the data-generating distribution**,
not coefficients of any model. Following the targeted-learning
principle that the *parameter must be defined before the estimator*
[@petersen2014causal; @vanderLaan2006tmle], any subsequent modelling
choice (logistic, log-binomial, TMLE, g-computation) is a means to
estimate functions of $p_{00}, p_{10}, p_{01}, p_{11}$, not a
substitute for their definition.

### B. Additive interaction contrast on the risk scale

The simplest additive-scale interaction parameter is the **interaction
contrast** on absolute risks:

Writing $p_{ae}$ for the event risk when $A=a$ and $E=e$, so that $p_{11}$ is the risk when both exposures are present and $p_{00}$ when neither is, the contrast (in events per patient) is:

$$\psi_{\text{add}} \;=\; p_{11} - p_{10} - p_{01} + p_{00}.$$

Equivalently, $\psi_{\text{add}}$ is the difference between the effect
of $A$ when $E=1$ and the effect of $A$ when $E=0$:

$$\psi_{\text{add}} \;=\; \big(p_{11} - p_{01}\big) \;-\; \big(p_{10} - p_{00}\big).$$

$\psi_{\text{add}} = 0$ means the absolute effect of $A$ is the same
whether $E$ is set to $0$ or $1$. Positive values indicate
super-additivity: the joint exposure produces more events per person
than the two single-exposure effects would predict on their own
[@vanderweele2014tutorial; @knol2012recommendations].

### C. RERI on the relative-risk scale

When risk ratios are the natural summary, the **relative excess risk
due to interaction** generalises the additive-contrast idea to ratio
form. Taking $(A=0, E=0)$ as the joint reference category and writing
$RR_{ae} = p_{ae}/p_{00}$,

$$\text{RERI} \;=\; RR_{11} - RR_{10} - RR_{01} + 1.$$

$\text{RERI} = 0$ corresponds to no additive interaction; positive
values indicate that the joint exposure produces more risk than the
sum of the two individual excess risks
[@vanderweele2014tutorial; @knol2012recommendations]. Two derived
summaries that appear regularly in the epidemiologic literature are
the **attributable proportion due to interaction**, $\text{AP} = \text{RERI}/RR_{11}$,
and the **synergy index**, $S = (RR_{11} - 1)/[(RR_{10} - 1) + (RR_{01} - 1)]$.

### D. Multiplicative interaction as a secondary scale

Multiplicative interaction asks whether the joint risk ratio departs
from the product of the two marginal risk ratios. Writing
$\psi_{\text{mult}}$ for the parameter,

$$\psi_{\text{mult}} \;=\; \dfrac{RR_{11}}{RR_{10} \cdot RR_{01}}.$$

$\psi_{\text{mult}} = 1$ means the joint risk ratio is exactly the
product of the marginal risk ratios. Multiplicative interaction is a
useful complement to the additive summaries when the scientific
question is itself ratio-based or when the result has to sit alongside
other ratio-scale literature. It is rarely the right starting point
for decision-making, where the additive scale answers "how many
additional events?" directly.

### E. Subgroup-specific treatment effects when the second variable is a non-manipulable modifier

When the second variable is a baseline modifier $M$ that you do *not*
intervene on, the relevant estimands are subgroup-specific causal
contrasts of one treatment. The full conditional-effect machinery is
covered in the companion [Effect Modification chapter](03-04-effect-modification.qmd).
The summaries most often paired with this chapter's interaction
language are:

- **Subgroup-specific risk under treatment**: $P(Y^1 = 1 \mid M = m)$.
- **Subgroup-specific risk under control**: $P(Y^0 = 1 \mid M = m)$.
- **Subgroup-specific risk difference**: $RD(m) = P(Y^1=1 \mid M=m) - P(Y^0=1 \mid M=m)$.
- **Subgroup-specific risk ratio**: $RR(m) = P(Y^1=1 \mid M=m) / P(Y^0=1 \mid M=m)$.
- **Additive heterogeneity contrast**: $RD(1) - RD(0)$.
- **Multiplicative heterogeneity contrast**: $RR(1) / RR(0)$.

These look numerically similar to the joint-intervention contrasts
above, but their causal interpretation is different: $M$ is a
stratifier, not an intervened-on exposure. Calling the result a "causal
interaction" is appropriate only when $M$ itself is conceptually
manipulable.

### Estimands at a glance

The compact table below summarises the four estimands this chapter
foregrounds.

| Question | Estimand | Formula | Plain-English interpretation | Preferred use |
|---|---|---|---|---|
| Does treatment benefit differ across subgroups defined by $V$? | Subgroup-specific risk difference / CATE contrast | $RD(v) = E[Y^1 - Y^0 \mid V=v]$, contrasted across $v$ | "Treatment prevents $RD(v)$ more events per 100 patients in subgroup $V=v$" | Treatment allocation, subgroup targeting |
| Do two joint interventions combine super-additively on the risk scale? | Additive interaction contrast | $\psi_{\text{add}} = p_{11} - p_{10} - p_{01} + p_{00}$ | "Joint exposure produces $\psi_{\text{add}}$ additional events per person beyond what the two effects would predict separately" | Public-health decision-making, RWE targeting |
| Do two joint interventions combine super-additively on the relative-risk scale? | RERI | $\text{RERI} = RR_{11} - RR_{10} - RR_{01} + 1$ | "Excess risk above 1.0 once the two single-exposure RRs are subtracted" | Epidemiologic synthesis when risks are not directly modelled |
| Do two joint interventions combine multiplicatively? | Multiplicative interaction | $\psi_{\text{mult}} = RR_{11} / (RR_{10} \cdot RR_{01})$ | "Joint RR is $\psi_{\text{mult}}$ times the product of the marginals" | Secondary summary when a ratio-scale question is explicit |
: Table 1. Core interaction estimands and their plain-English readings. {.striped}

::: {.callout-note}
## A toy example: computing interaction by hand

Suppose the four counterfactual risks are known exactly:

| | $E=0$ | $E=1$ |
|---|---:|---:|
| $A=0$ | $p_{00} = 0.05$ | $p_{01} = 0.11$ |
| $A=1$ | $p_{10} = 0.09$ | $p_{11} = 0.22$ |

The effect of $A$ when $E=0$ is $0.09 - 0.05 = 0.04$, that is 4
additional events per 100 people. The effect of $A$ when $E=1$ is
$0.22 - 0.11 = 0.11$, that is 11 additional events per 100 people.

- **Additive interaction contrast**: $\psi_{\text{add}} = 0.22 - 0.09 - 0.11 + 0.05 = 0.07$.
  The combined exposure produces **7 additional events per 100 people**
  beyond what would be expected if the two component effects simply
  added.
- **RERI** (taking $A=0, E=0$ as the doubly unexposed reference):
  $RR_{10} = 1.8$, $RR_{01} = 2.2$, $RR_{11} = 4.4$, so
  $\text{RERI} = 4.4 - 1.8 - 2.2 + 1 = 1.4$.
- **Attributable proportion due to interaction**: $\text{AP} = \text{RERI}/RR_{11} = 1.4/4.4 \approx 0.32$.
  About 32% of the risk in the doubly exposed group is attributable to
  the interaction beyond the sum of the individual effects.
- **Synergy index**: $S = (RR_{11} - 1)/[(RR_{10} - 1) + (RR_{01} - 1)] = 3.4/2.0 = 1.70$.
- **Multiplicative interaction**: $\psi_{\text{mult}} = 4.4/(1.8 \cdot 2.2) \approx 1.11$.
  The joint risk ratio is about 11% larger than the product of the two
  marginal risk ratios.

The same four numbers produce every standard interaction summary.
TMLE's role is to estimate those four counterfactual risks from
observational data under confounding; once they are in hand, all of
$\psi_{\text{add}}$, RERI, AP, $S$, and $\psi_{\text{mult}}$ are
arithmetic [@vanderweele2014tutorial; @knol2012recommendations].
:::

Why bother with this counterfactual machinery rather than reading off the
product-term coefficient from a logistic regression? Three reasons.

1. **The coefficient is on the wrong scale**. A product-term coefficient in
   a logistic model speaks to the odds ratio, not the risk ratio or risk
   difference. Odds-ratio interaction is not generally interpretable as
   either additive or multiplicative interaction in the public-health sense
   [@vanderweele2014tutorial].
2. **The coefficient is conditional, not marginal**. It estimates an effect
   conditional on covariates included in the model. The marginal subgroup
   effect---averaging over the covariate distribution within the
   subgroup---is what most decision questions actually need.
3. **The coefficient confounds estimation with model specification**. Move
   one covariate in or out and the interaction coefficient can change
   meaningfully. A causal estimand defined directly from potential outcomes
   does not depend on whether you used a polynomial or a spline for $W_1$.

Table 2 contrasts the two scales explicitly, and Table 3 maps each
question type to a recommended estimator family.

| Feature | Additive interaction | Multiplicative interaction |
|---|---|---|
| Mathematical form | $RD(1) - RD(0)$, or $\text{RERI}$ for two exposures | $RR(1)/RR(0)$, or $RR_{11}/(RR_{10} \cdot RR_{01})$ |
| Underlying scale | Absolute risk | Relative risk |
| Best for | Public health targeting, resource allocation, number-needed-to-treat reasoning | Etiologic comparisons, ratio-style benefit-risk summaries |
| Default model | None; estimate by standardisation or TMLE | Logistic, log-binomial, Cox product term |
| Coding sensitivity | Highly sensitive when one exposure is preventive; lowest-risk stratum should be reference | Less sensitive but still requires explicit reference |
| Decision relevance | Direct: tells you absolute benefit by subgroup | Indirect: needs baseline risk to translate into absolute benefit |
: Table 2. Additive versus multiplicative interaction at a glance. {.striped}

| Scientific aim | Preferred estimand | Recommended estimator |
|---|---|---|
| Treatment effect within prespecified subgroups | $RD(m)$, $RR(m)$ | TMLE within strata, or stratified Super Learner outcome regression with standardisation |
| Difference in treatment effect across subgroups | $RD(1) - RD(0)$, $RR(1)/RR(0)$ | TMLE in each stratum, then delta-method or bootstrap contrasts |
| Joint effect of two binary exposures | $p_{ae}$, $\psi_{\text{add}}$, RERI | TMLE for each joint regime mean, with post-estimation contrasts |
| Time-varying treatment with subgroup contrast | Counterfactual mean / survival under dynamic strategy by subgroup | LTMLE [@vanderLaan2012ltmle] or TMLE for marginal structural models |
| Many candidate modifiers, exploratory discovery | Conditional average treatment effect surface | Causal forest or simultaneous one-step TMLE with multiplicity control [@wei2023simultaneous] |
: Table 3. Causal estimands for interaction questions and recommended estimators. {.striped}

Write the estimand on a sticky note before you write the model. If you
cannot describe what you are estimating in one sentence using potential
outcomes, you are not yet ready to estimate it.

---

## 3. Interaction-specific design and identification

Interaction analyses inherit the standard causal assumptions covered in
detail in @sec-roadmap: consistency,
exchangeability given measured confounders, positivity, and no
interference unless explicitly modelled. This section focuses only on
the issues that get *harder* when interaction is the target.

### The target trial protocol

In any RWE study, Hernán and Robins [-@hernan2016target] recommend that the
analysis be designed as an *emulation of a target randomized trial*. For
interaction work, this discipline is doubly important, because subgroup
analyses are uniquely vulnerable to design errors that mimic real
heterogeneity. Before any modeling decision, the analyst should write down:

- **Eligibility criteria** including age, indication, and absence of prior
  outcome events.
- **Treatment strategies** to be compared, including dose, duration, and
  initiation rules.
- **Time zero**: the date when eligibility is met, the assigned strategy
  begins, and follow-up starts.
- **Follow-up window** and outcome definition.
- **Censoring rules**: discontinuation, switching, loss to follow-up,
  competing events.
- **Candidate effect modifiers**: pre-treatment, well-measured, biologically
  or clinically motivated.
- **Causal assumptions** the analysis will rely on.

If any of these are decided after looking at the outcome, you risk
manufacturing apparent heterogeneity through differential immortal time,
selection bias, or post-hoc subgroup tuning.

### Interaction must be estimated on a common target population

A recurring failure mode in applied work is to compute an interaction
contrast from *separately matched or stratified subgroup analyses*. A
typical version of this looks like the following: the analyst matches
$E=1$ patients to $E=0$ patients on baseline covariates, estimates the
treatment effect within each matched set, and then computes an
"interaction" as the difference of those two subgroup-specific
effects. The result is not generally an interpretable causal interaction
parameter, for three reasons.

- **The matched samples differ from each other.** Each match is built
  to balance covariates *within its own subgroup*. The two resulting
  populations do not overlap on the same target, so the contrast is
  not estimating a single counterfactual quantity defined on one
  population.
- **The contrast is not anchored to a counterfactual estimand.** $\psi_{\text{add}} = E[Y^{1,1}] - E[Y^{0,1}] - E[Y^{1,0}] + E[Y^{0,0}]$
  averages each potential outcome over the *same* covariate distribution,
  typically the full study cohort. A difference of two separately
  matched effects does not.
- **Confounding control is uneven across exposure cells.** Patterns of
  confounding can differ by $E$, so subgroup-specific matching may
  succeed in one stratum and fail in another. The interaction estimate
  inherits the worst of those failures.

The remedy is to estimate the interaction parameter directly on the
full cohort, with a single adjustment specification (regression,
Super Learner, or a TMLE plug-in) that targets all four counterfactual
means jointly. Section 6 implements exactly this workflow. When
inverse-probability weights or matching are used as part of the
adjustment, they should be constructed on the full cohort with treatment
defined as the *joint* exposure $(A, E)$, not on each subgroup
separately. If the only data available are pre-aggregated
subgroup-specific estimates, the most defensible approach is to report
each subgroup's effect with its own confidence interval and explicitly
flag that no formal interaction parameter has been identified.

### Identification assumptions for joint-intervention interaction

Identification of a causal interaction between two exposures requires
the standard assumptions of @sec-roadmap,
restated here in joint-treatment form
[@hernan2020whatif; @petersen2014causal]:

- **Consistency.** For each patient, the observed outcome under the
  observed exposure pair $(A, E)$ equals the corresponding
  counterfactual outcome $Y^{A, E}$.
- **Exchangeability for the joint treatment.** Given the measured
  confounders $W$, the joint exposure $(A, E)$ is independent of all
  four potential outcomes $Y^{a, e}$. Single-treatment exchangeability
  for $A$ alone is not sufficient.
- **Positivity for the joint treatment.** Every covariate pattern $W$
  observed in the target population has positive probability of every
  joint exposure level: $P(A=a, E=e \mid W) > 0$ for all
  $(a, e) \in \{0,1\}^2$.
- **A common target population.** All four counterfactual means are
  averaged over the same distribution of $W$, typically the full
  study cohort. Without a common target, the four estimated means do
  not combine into a single interpretable contrast.

In interaction analyses, positivity failures often appear only after
crossing the two exposures with key confounders or modifier strata, so
it is not enough to inspect overall treatment overlap. A propensity
score for $A$ that is well-behaved marginally can still collapse to
zero or one inside specific $(W, E)$ cells.

::: {.callout-important}
## Joint positivity checklist
Before reporting an interaction estimate, verify all five:

1. **Cell counts overall.** Inspect the number of patients in each
   $A \times E$ cell across the full cohort.
2. **Cell counts within key strata.** Repeat the cell-count inspection
   inside important covariate or modifier strata. A cell that is
   adequate marginally can collapse inside the strata that matter
   for the scientific question.
3. **Joint propensity scores.** Summarise the estimated joint
   propensity $\widehat g(A, E \mid W)$ (or its two factor pieces)
   across the cohort and within strata.
4. **Extreme weights and near-zero probabilities.** Flag observations
   with very small $\widehat g$ in any cell, very large
   inverse-probability weights, or fitted joint probabilities that
   compress against 0 or 1.
5. **Simplify when cells are too sparse.** When one or more joint
   cells has essentially no data, simplify the estimand: collapse
   modifier categories, restrict the time horizon, or report
   single-exposure effects with an explicit positivity disclaimer
   instead of a poorly identified interaction parameter.
:::

### Why positivity is the central threat to interaction analyses

The positivity assumption is local. It must hold within every covariate
pattern you condition on. As soon as you stratify by a modifier or
require support for a joint exposure, you are conditioning on more, and
any positivity weakness in the original cohort is amplified.

**Interaction estimation is more sensitive to positivity violations
than main-effect estimation, because it requires support across joint
exposure strata.** For a binary $A$ and a binary $E$, the analysis
must learn about four joint exposure states: $(A=0, E=0)$,
$(A=1, E=0)$, $(A=0, E=1)$, and $(A=1, E=1)$. If very few patients
are observed in one of those cells, or if co-exposed patients are
clinically very different from everyone else in the cohort, the
estimate for that joint state will rely heavily on extrapolation from
the outcome regression. The number of cells that must be populated
grows with the product of the exposure cardinalities, while the data
per cell shrinks.

A concrete example makes this vivid. Consider a 4,000-patient HIV
cohort with binary $A$ (HAART initiation), binary $E$ (heavy alcohol
use), and several binary clinical confounders. The marginal cells are
reasonably large:

| Cell | $A=0, E=0$ | $A=0, E=1$ | $A=1, E=0$ | $A=1, E=1$ |
|---|---:|---:|---:|---:|
| Patients | 1,800 | 600 | 1,400 | 200 |
| Outcome events | 90 | 60 | 70 | 20 |

The doubly exposed cell ($A=1, E=1$) holds 200 patients, of whom 20 have
the outcome. Within that cell, the propensity score for $A$ given the
clinical confounders may be near 0 for most patients (heavy drinkers are
counselled against initiating HAART without adherence support and many
defer) and near 1 for a handful with severe immunosuppression who were
initiated despite the alcohol use---**that handful then dominates the
inverse-probability weights for the interaction estimate**. The headline
RERI may move dramatically when those few patients are dropped, even
though the main-effect estimate of HAART on the outcome is stable.

Two recurring failure modes follow:

- **A frailty-by-treatment interaction** where almost no frail patients
  receive aggressive therapy. Within the frail stratum, positivity
  fails: you cannot estimate $E[Y^1 \mid \text{frail}]$ from the data,
  no matter what estimator you choose.
- **A two-exposure analysis of HAART initiation and heavy alcohol use**
  as above. The $A=1, E=1$ joint cell may be tiny in any realistic
  cohort. The apparent RERI estimate may be dominated by a handful of
  patients with extreme weights, and the confidence interval will be
  wide, unstable, or both.

This is why propensity-score diagnostics should be reported *within
each joint exposure cell or modifier subgroup*, not only overall, and
why collaborative TMLE (Section 7) was developed as a sensitivity
strategy for fragile propensity-score estimation.

### Interaction through the Targeted Learning roadmap

Petersen and van der Laan [-@petersen2014causal] formalize the workflow as
a small sequence of disciplined steps. Applied specifically to
interaction, the roadmap separates the *causal question* from the
*statistical procedure* and protects the analysis from the most common
failure mode: letting a model dictate what counts as interaction.

**Step 0. Define the question.** State, in clinical or public-health
language, whether you are asking (i) does the effect of $A$ vary across a
non-manipulable modifier $M$ (effect modification), or (ii) do two
manipulable exposures $A$ and $E$ combine in ways that depart from their
individual effects (interaction).

**Step 1. Statistical model.** Adopt a *nonparametric* model for the
joint distribution of $(W, A, E, Y)$ (or $(W, A, M, Y)$). No functional
form is imposed at this step. The Targeted Learning principle is that
parametric structure should be a tool, not an assumption baked into the
estimand.

**Step 2. Causal estimand.** Write the parameter explicitly as a
function of counterfactual outcomes:
$\psi_{\text{add}} = E[Y^{1,1} - Y^{0,1}] - E[Y^{1,0} - Y^{0,0}]$ for the
joint-intervention case, or $RD(1) - RD(0)$ for the modifier case. *This
is what the analysis aims to recover; it is not a regression coefficient.*

**Step 3. Identification.** State the assumptions under which the causal
estimand maps to a function of the observed-data distribution:
exchangeability for $A$ (and $E$) given $W$, joint positivity, consistency,
and no interference. Identification fails or weakens whenever any of
these break, and the estimator cannot rescue identification it does not
have.

**Step 4. Estimation.** Estimate the nuisance functions
($Q$ for the outcome regression, $g$ for the treatment mechanisms) with
Super Learner, then apply a TMLE targeting step so that the plug-in
estimator solves the efficient influence function equation for
$\psi_{\text{add}}$ or $\psi_{\text{mult}}$.

**Step 5. Sensitivity analysis.** Vary the learner library, the
adjustment set, and the propensity-score selection (e.g., CTMLE) to
characterize the stability of the interaction estimate. Quantify the
amount of unmeasured confounding that would be needed to overturn it.

The roadmap's message for interaction work is uncompromising: **interaction
parameters are not coefficients; they are functions of counterfactual
risks**. Any procedure that conflates these---reporting a logistic
product term as if it answered $\psi_{\text{add}}$, for example---has
skipped Step 2.

Treat positivity diagnostics inside subgroups as a non-negotiable step, not
an optional appendix. When the modifier strata are small or the joint
exposure cells are sparse, no estimator---however sophisticated---can
manufacture identifiability that the data do not support.

---

## 4. Estimand-first targeted learning for interaction

Super Learner [@vanderLaan2007superlearner] and TMLE
[@vanderLaan2006tmle] are introduced in @sec-doubly-robust-tmle
and @sec-superlearner; the cross-validated, collaborative,
and adaptive variants (CV-TMLE, CTMLE, A-TMLE) are covered in
@sec-advanced-tmle. This section assumes that
material as background and focuses on what changes when the target
parameter is an interaction.

### What TMLE is targeting here

For interaction estimation, TMLE does not begin with a product term.
It begins with the four mean counterfactual outcomes
$p_{00}, p_{10}, p_{01}, p_{11}$ under the joint exposure conditions
defined in Section 2. The additive interaction estimand
$\psi_{\text{add}} = p_{11} - p_{10} - p_{01} + p_{00}$ is then a
contrast of those four means. TMLE improves a plug-in estimator of
those means by incorporating the treatment mechanism and updating the
initial outcome fit so that the final estimate is optimised for the
target parameter [@vanderLaan2006tmle; @schuler2017tmle].

In plain English: estimate the average outcome we would see if everyone
were in each of four hypothetical worlds: nobody exposed, only $A$, only
$E$, both. Then ask whether the "both exposed" world is larger than what
the two single-exposure worlds would predict. The interaction parameter
is exactly that comparison.

### Workflow at a glance

The full estimand-to-report workflow for an interaction analysis is
sketched below. The single decision at the top (one or two
manipulable interventions?) determines whether the analyst is in this
chapter or in the [Effect Modification](03-04-effect-modification.qmd)
chapter.

```{mermaid}
flowchart TD
    A[Scientific question] --> B[Choose estimand]
    B --> C{One or two manipulable interventions?}
    C -->|One| D[Effect modification chapter]
    C -->|Two| E[Interaction chapter]
    E --> F[Specify four joint counterfactual means]
    F --> G[Check identification and joint positivity]
    G --> H[Estimate Q and g with Super Learner]
    H --> I[Target the four means with TMLE]
    I --> J[Compute additive and multiplicative contrasts]
    J --> K[Report cell-specific risks, contrasts, diagnostics]
```

### Why TMLE is preferred for interaction

Targeted learning is useful for any causal parameter, but interaction
amplifies the two pathologies it was designed to fix. **Model
misspecification risk is amplified** because an interaction is a
*derivative* of the outcome surface: it depends on differences of
differences of fitted values, so small functional-form errors that are
invisible in a main-effect estimate can dominate the interaction estimate.
**Positivity problems are amplified** because the joint exposure cells
$(A=a, E=e)$ are typically sparser than either marginal cell, and
inverse-probability weights deteriorate accordingly.

Two TMLE properties directly address these amplifications. **Flexible
nuisance estimation through Super Learner** lets the data choose the
shape of $Q$ rather than imposing a parametric form whose interaction
term is a model artefact. **Double robustness** means the interaction
estimate remains consistent if either $Q$ or $g$ is misspecified, a real
safety margin in joint-cell estimation, where neither model is easy to
get exactly right. The targeting step also improves efficiency relative
to plug-in g-computation, which matters when the joint cell is small and
every bit of information has to be used optimally.

### From estimand to estimator: practical workflow

The slogan "estimand first, then estimator" is easier to repeat than to
operationalize. The five steps below make the mapping concrete for an
interaction analysis with a binary treatment $A$ and a binary second
exposure $E$ (or modifier $M$). **At its heart, the procedure estimates
four counterfactual means and combines them into the contrast you
specified in Section 2.**

**Step 1. Estimate the outcome regression** $Q(A, E, W) = E[Y \mid A, E, W]$
with Super Learner. This is a single model that uses the *observed* data
and includes treatment, second exposure, and confounders as predictors.

**Step 2. Generate counterfactual predictions** under each of the four
joint regimes by querying the same fitted model on modified data:
$Q(1,1,W_i)$, $Q(0,1,W_i)$, $Q(1,0,W_i)$, $Q(0,0,W_i)$ for every patient
$i$.

**Step 3. Compute the interaction contrast** by averaging:
$\widehat{\psi}_{\text{add}} = \frac{1}{n}\sum_i [Q(1,1,W_i) - Q(0,1,W_i)] - \frac{1}{n}\sum_i [Q(1,0,W_i) - Q(0,0,W_i)]$.
At this point you have a plug-in g-computation estimate.

**Step 4. Apply the TMLE update.** Estimate the treatment mechanisms
$g_A(A \mid W)$ and $g_E(E \mid W)$ (or their joint $g(A,E \mid W)$) with
Super Learner. Use these to construct the clever covariate that updates
$Q$ along the least-favorable submodel for your target parameter. The
updated $Q^*$ now solves the efficient influence-function equation for
$\psi_{\text{add}}$ (or $\psi_{\text{mult}}$).

**Step 5. Inference via the influence curve.** Compute the empirical
influence function of $\widehat{\psi}_{\text{add}}$ using $Q^*$ and $g$,
then form a Wald confidence interval from its empirical variance. The
same machinery yields valid CIs for any smooth contrast of the four
counterfactual means.

Section 6 implements Steps 1–3 directly with `sl3` and Steps 4–5 via
`tmle3` (or stratified `tmle` plus bootstrap as a transparent
alternative). The point of the boxed mapping is that *every line of
code in Section 6 corresponds to one of these five steps*; if you can
say which step a line belongs to, the workflow is well-specified.

### Recent literature

Three recent papers are worth keeping in view: a 2023 systematic review
of 89 observational TMLE applications that documented rapid uptake in
epidemiology and frequent use for heterogeneity questions
[@smith2023systematic]; a 2025 best-practice tutorial that walks through
TMLE implementation choices for applied researchers [@talbot2025tmle];
and a 2023 paper on simultaneous one-step TMLE for many subgroup effects,
with valid joint confidence intervals [@wei2023simultaneous].

For most MPH-level interaction analyses, the rule is simple. Estimate
the nuisance functions with Super Learner using a modest, sensible
library, target with TMLE for the joint-intervention parameter, and use
the influence function for inference. Escalate to CV-TMLE or `drtmle`
only when the learner library is genuinely aggressive or the joint
exposure cells are large enough to support data splitting (see
@sec-advanced-tmle).

---

## 5. Additive and multiplicative interaction in practice

The most important recommendation in the VanderWeele and Knol tutorial
[-@vanderweele2014tutorial] is to report both scales whenever the
outcome and data permit. Scale dependence is not a methodological
detail. It is a substantive claim about how two exposures combine to
produce risk.

### Why additive interaction is often the default in RWE

Additive interaction compares **excess risks**, not ratios. It answers
the question of whether the combined exposure produces more cases than
would be expected from adding the separate risk differences for each
exposure. That question is usually the most actionable for public
health, safety monitoring, and clinical prioritization. It maps
directly onto the number of cases per population unit, which is what
policy and resource decisions are made in.

Multiplicative interaction, in contrast, compares ratios of ratios.
That summary remains useful for etiologic and comparative-effect
questions, and for putting a finding alongside other ratio-based
literature. It is rarely the right starting point when the goal is to
identify where the absolute burden of disease is greatest, or where
intervention would prevent the most cases.

For most RWE decision problems, the additive scale should be the
default starting point. If the goal is to identify where the greatest
excess number of cases may occur, or which co-exposed group may benefit
most from intervention, risk differences and additive interaction
contrasts are usually more interpretable than ratios. A
2-percentage-point absolute risk reduction in a high-incidence
subgroup represents a much larger public-health benefit than the same
relative-risk reduction in a low-incidence subgroup
[@basten2025additive; @vanderweele2014tutorial].

### Operationalizing the choice of scale

The two scales answer two different questions, and the analyst should
state which one the analysis is built to answer.

| Scale | What it tells you | Best for |
|---|---|---|
| **Additive** ($\psi_{\text{add}}$, RD contrasts, RERI) | Public-health impact: how many additional events are prevented (or caused) per person treated, by subgroup | Decision-making, resource allocation, NNT, policy targeting |
| **Multiplicative** ($\psi_{\text{mult}}$, RR contrasts) | Relative effect: how much the *ratio* of treated to untreated risk changes across exposure cells | Etiologic comparison, meta-analytic synthesis, biology |

**For decision-making, prioritize additive interaction unless there is
a specific scientific reason to use the multiplicative scale.** Report
the multiplicative contrast alongside it whenever feasible. Reporting
both scales together leaves the reader free to translate the result
into either decision or etiologic terms. A single scale forces a
choice the reader may not share.

### Do not compare interaction estimates across exposure pairs too quickly

When analysts estimate several pairwise interactions in the same study,
it is tempting to rank them. For example, one might report that the
diabetes-by-obesity interaction is larger than the kidney-disease-by-
lipid-disorder interaction, and conclude that the first pair is more
biologically synergistic, more clinically important, or more
policy-relevant than the second. That comparison can be misleading,
for several practical reasons.

Each pairwise estimate is a function of the *target population* over
which the four counterfactual means are averaged. Different exposure
pairs can define different observed support, can rest on different
parts of the data, and can be affected by different positivity and
sparsity limitations. Baseline risk varies across the strata that one
pair defines and the strata another pair defines. A larger interaction
contrast for one pair does not automatically mean that pair is more
synergistic, more important, or more actionable than another.

Before ranking interaction estimates across pairs, an analyst should
confirm:

- the same target population was used to standardize each estimate,
- the same outcome scale was used (additive vs. multiplicative, risk vs. RMST vs. cumulative incidence at the same horizon),
- the same follow-up period and intervention definitions apply, and
- all four joint exposure cells are adequately supported for every pair under consideration.

If those conditions do not hold, the pair-by-pair estimates can each
be reported with their own confidence intervals, but they should not
be ranked as if they answered the same question.

### Protective exposures require recoding

Additive interaction summaries such as RERI and AP are *not* invariant
to coding when one or both exposures are preventive. The standard
formulas assume that $(A=0, E=0)$ is the doubly unexposed reference and
that the joint reference cell carries the lowest expected risk. When
one exposure is preventive (HAART initiation, vaccination, adherence to
guideline-recommended therapy) the doubly exposed cell may not be the
highest-risk cell, and applying the unmodified RERI formula can produce
values that look like "antagonism" but actually reflect a coding
artefact rather than a substantive finding.

The remedy is mechanical and is implemented automatically by the
`interactionR` package: **before computing additive interaction
summaries, recode each exposure so that the lowest-risk joint stratum
is the reference category** [@knol2011preventive;
@vanderweele2014tutorial; @alli2021interactionr]. For HAART initiation
($A$) and heavy smoking ($B$), the reference is typically HAART
initiator without smoking; RERI and AP are then interpretable as the
additive departure from the safest combination. This is one of the most
common applied mistakes in published RERI estimates and should be
checked explicitly in any analysis with a preventive exposure.

### Cell-specific reporting

Whatever scale you privilege, the most useful presentation is a
**cell-specific risk table**: estimated risks under each combination of
exposures (or under treatment vs. control within each modifier stratum),
plus the contrasts of interest. Readers can then reconstruct any other
contrast they care about and assess whether the differences are large
enough to matter clinically.

### Time-to-event outcomes: choose a risk-scale estimand

Many real-world evidence analyses have time-to-event outcomes, and the
default model in that setting is the Cox proportional-hazards
regression. The default temptation, then, is to introduce an interaction
term in the linear predictor and report a hazard-ratio interaction. For
interaction analyses specifically, **this is the wrong default**.
Hazard-ratio interactions are problematic for three connected reasons.

- **Hazards are non-collapsible.** The marginal hazard ratio is not in
  general a weighted average of conditional hazard ratios over baseline
  covariates, so even an unconfounded marginal hazard-ratio interaction
  is not a clean function of the underlying counterfactual hazards
  [@martinussen2018hazardratios; @aalen2015cox].
- **Hazards condition on survival.** $\lambda(t \mid A, E)$ is the
  instantaneous risk *given* survival to time $t$, and the survivors at
  $t$ are themselves a selected subset of the original cohort. Even if
  treatment is randomly assigned at baseline, the survivors under
  $A=1, E=1$ and the survivors under $A=0, E=0$ at time $t$ are
  comparable populations only at $t = 0$. A hazard-ratio interaction
  therefore mixes a true treatment-by-covariate effect with a
  time-varying selection effect.
- **Additive interaction on hazards is not additive interaction on
  cumulative incidence.** A model that is multiplicative on hazards is
  highly nonlinear on the cumulative incidence scale, and the
  interaction parameter on the hazard scale does not generally map
  onto $\psi_{\text{add}}$, RERI, AP, or $S$.

The recommendation that follows is straightforward. **Define and target
an interaction parameter on the cumulative-risk scale at a clinically
meaningful time horizon $t^*$**, not on the hazard scale. Useful
risk-scale estimands include:

- **Cumulative incidence at $t^*$**: $F^{a,e}(t^*) = P(T^{a,e} \le t^*)$. The four counterfactual cumulative incidences populate exactly the same 2$\times$2 table as the binary-outcome example in Section 2, and RERI / AP / $S$ are computed directly from them.
- **Restricted mean survival time (RMST) at $t^*$**: $\mu^{a,e}(t^*) = E[\min(T^{a,e}, t^*)]$. RMST contrasts have a natural decision-theoretic interpretation (expected event-free time gained) and an additive scale that is straightforward to interpret.
- **Risk difference at $t^*$**: $F^{1,1}(t^*) - F^{0,0}(t^*)$ and the corresponding interaction contrast $F^{1,1}(t^*) - F^{0,1}(t^*) - F^{1,0}(t^*) + F^{0,0}(t^*)$.

Each of these can be estimated with a TMLE that targets the
counterfactual survival or cumulative-incidence function directly,
using `survtmle`, `concrete`, `ltmle`, or `lmtp` (with survival outcome
nodes). The point-treatment chapter on time-to-event outcomes
(@sec-tte-competing) walks through
the survival-TMLE machinery. Combining it with the joint-intervention
workflow of Section 6 (by replacing $Q(A, E, W) = E[Y \mid A, E, W]$
with $\hat F^{a,e}(t^*; W)$ and contrasting four counterfactual
cumulative incidences) yields an interaction estimate that is free of
hazard-scale interpretive problems.

::: {.callout-warning}
## Avoid: hazard-ratio interaction terms as the headline
A Cox model with a $A \times E$ product term is fine as a sensitivity
analysis or as an exploratory description of how hazards differ by
subgroup, but the headline interaction estimate in a real-world
evidence paper should be defined and reported on the cumulative-risk
scale, with an explicit time horizon. The same data analysed at
$t^* = 1$ year, $t^* = 2$ years, and $t^* = 5$ years can give
materially different interaction conclusions, and that is information
the reader needs.
:::

### Interaction in real-world evidence

Real-world data add three challenges to the picture above that randomized
trials usually do not face.

- **Positivity issues are amplified.** Joint exposure cells $(A=a, E=e)$
  may be sparsely populated, especially when both exposures are rare or
  when one is preventive and the other indicated for similar patients.
  Subgroup-level positivity must be checked separately within each cell;
  inverse-probability components become unstable as the conditional
  probability of the joint regime approaches zero. Positivity weakness is
  the single most common reason interaction analyses produce headline
  estimates that fail to replicate.
- **Model dependence is amplified.** Interaction estimates are derivatives
  of the outcome surface and are intrinsically more sensitive to
  functional-form assumptions than main effects. Parametric models that
  impose linear or log-linear interactions can misrepresent the underlying
  causal interaction structure. Super Learner mitigates this by allowing
  the data to choose among flexible learners, and TMLE protects against
  remaining misspecification through double robustness.
- **Interpretability hinges on scale.** A real-world evidence audience
  typically wants to know whether and where to deploy a treatment. That
  question lives on the additive scale. Multiplicative interaction
  remains useful for etiologic synthesis and meta-analytic comparison,
  but should not be the only summary in a decision-oriented report
  [@vanderweele2014tutorial; @basten2025additive].

A practical recommendation follows: in RWE, **prefer the additive scale
for decision-making, report the multiplicative scale for completeness,
and run sensitivity analyses that vary both the nuisance learner library
and the propensity-score model**.

### Operational RWE recommendations

Four operational rules concentrate the chapter's recommendations for
real-world evidence work:

- **Treatment targeting and public-health decisions live on the additive
  scale.** Risk differences and additive interaction contrasts answer
  "how many additional events?" directly. Multiplicative summaries are
  useful complements but should not be the basis of a deployment
  decision.
- **Pre-specification matters more for interaction work than for
  main-effect work.** The analyst can easily search across many
  candidate modifiers, exposure pairs, scales, and adjustment sets.
  Write down the confirmatory list and the scale before unblinding,
  and label everything else as exploratory in the manuscript
  [@kent2020path].
- **Use outcome-blind simulations to choose nuisance-learning
  strategies.** Before unblinding the comparative treatment-outcome
  results, run candidate Super Learner libraries, propensity-score
  specifications, and TMLE variants on outcome-blind versions of the
  data and select the configuration that performs best on the
  outcome-blind metrics. This is the targeted-learning analogue of
  the design-stage discipline emphasised in the propensity-score
  literature [@rubin2007design; @sofrygin2017simcausal].
- **Pre-specify the sensitivity battery.** Vary the learner library
  (a small library and a richer one), vary the adjustment set
  (minimal sufficient and richer), and vary the positivity-handling
  choice (no truncation, modest truncation, CTMLE). Stable estimates
  across these variations are reassuring; large movements flag
  fragility worth disclosing.

Pre-specify the scale, use a single common reference category, and
present the cell-specific risks. If you have to pick one scale, pick
additive for public-health audiences and multiplicative for etiologic
ones. When you can, present both. In RWE, build the analysis to
survive the joint-cell positivity check and the alternative-learner
sensitivity test.

### Diagnostic checklist for interaction analyses

::: {.callout-tip}
## Five questions before you trust an interaction estimate

1. **Estimand.** Is interaction defined as a causal parameter
   ($\psi_{\text{add}}$, $\psi_{\text{mult}}$, RERI, or a contrast of
   subgroup risk differences), not as a regression coefficient?
2. **Positivity.** Is the joint-exposure cell or the modifier stratum
   adequately populated, with both treated and untreated patients
   represented?
3. **Learners.** Are flexible nuisance learners (Super Learner with a
   diverse library) used, rather than a single parametric model that
   could impose the wrong functional form?
4. **Scale.** Is the choice of additive vs. multiplicative scale
   justified by the decision question, with both reported when feasible?
5. **Sensitivity.** Are sensitivity analyses prespecified that vary the
   learner library, the adjustment set, and the propensity-score
   selection?
:::

---

## 6. Point-treatment worked example with TMLE and Super Learner

This section demonstrates the workflow on a simulated cohort with a
binary treatment $A$, a binary second exposure $E$ (here playing the
role of a baseline modifier), a binary outcome $Y$, and three baseline
confounders $W_1, W_2, W_3$. The data-generating mechanism includes a
genuine additive-scale interaction. The example proceeds in two layers.

**Layer 1 (this section)** is a transparent **plug-in g-computation**
workflow that makes the four-counterfactual-means logic visible: fit
$Q(A, E, W)$ with Super Learner, predict outcomes under each of the
four joint regimes, average to get $\widehat p_{ae}$, and combine the
four marginal risks into $\widehat\psi_{\text{add}}$ and RERI.

**Layer 2 (later in this section)** explains what a full TMLE
implementation adds on top of Layer 1: estimate the joint treatment
mechanism $g(A, E \mid W)$, run a targeting update on the four
counterfactual-mean estimates, and use the efficient influence curve
for inference (or bootstrap the derived contrast when working with a
two-step implementation). Layers 1 and 2 target the same estimand;
Layer 2 supplies the targeting and inference machinery that the
plug-in version skips.

A minimal implementation has five steps:

1. **Estimate the outcome regression** $Q(A, E, W) = E[Y \mid A, E, W]$ with Super Learner.
2. **Predict outcomes under the four joint exposure states** $(A, E) \in \{(0,0), (1,0), (0,1), (1,1)\}$.
3. **Average each set of predictions over the target population** to get four marginal counterfactual risks.
4. **Combine the four marginal risks into the additive or multiplicative interaction parameter** ($\psi_{\text{add}}$ or $\psi_{\text{mult}}$, plus RERI, AP, and $S$ if useful).
5. **Use a TMLE update and the efficient influence curve for inference** when using a targeted estimator. The plug-in version is a g-computation step; the TMLE update adds robustness when the outcome regression is misspecified but the treatment mechanism is consistent.

### Simulated data

```{r}
#| label: sim-data
#| eval: false

library(dplyr)
set.seed(2026)

# Simulate a cohort of 2,000 patients. n = 2000 keeps the Super Learner
# + four-cell counterfactual prediction step fast in both the static
# and the WebR (interactive) builds. Production analyses can use any n.
n  <- 2000
W1 <- rnorm(n)                                       # continuous confounder
W2 <- rbinom(n, 1, 0.4)                              # binary confounder
W3 <- runif(n, -1, 1)                                # bounded confounder

# Second exposure E (here playing the role of a baseline modifier M)
E  <- rbinom(n, 1, plogis(-0.2 + 0.7 * W1 - 0.4 * W2))

# Treatment depends on confounders and the second exposure
A  <- rbinom(n, 1, plogis(-0.1 + 0.8 * W1 + 0.5 * W2 -
                            0.3 * W3 + 0.2 * E))

# Outcome with genuine additive-scale interaction between A and E
linpred_y <- -2.0 + 0.8 * A + 0.4 * E + 0.9 * A * E +
             0.7 * W1 - 0.5 * W2 + 0.3 * W3
Y <- rbinom(n, 1, plogis(linpred_y))

dat <- data.frame(Y, A, E, W1, W2, W3)
```

### Layer 1: transparent plug-in g-computation

The data structure is $O = (W, A, E, Y)$. The plug-in workflow is the
first three steps of the estimand-to-estimator mapping from Section 4:

1. **Estimate the outcome regression** $Q(A, E, W) = E[Y \mid A, E, W]$
   with Super Learner.
2. **Generate counterfactual predictions** under each of the four joint
   regimes: $Q(1,1,W)$, $Q(0,1,W)$, $Q(1,0,W)$, $Q(0,0,W)$.
3. **Average each set of predictions over the target population** to
   get $\widehat p_{ae} = \frac{1}{n}\sum_i Q(a, e, W_i)$, and combine
   the four marginal risks into $\widehat\psi_{\text{add}}$, RERI, AP,
   and $\widehat\psi_{\text{mult}}$.

In plain English: we predict outcomes under four hypothetical worlds
(everyone exposed to both, exposed to one, exposed to the other, or
neither) and compare the differences across those worlds.

This is the transparent plug-in version. A full TMLE implementation
uses the same estimand but adds the treatment model and the targeting
step (Layer 2 below).

```{r}
#| label: sl3-tmle3-joint-intervention
#| eval: false

# Joint-intervention plug-in g-computation skeleton
# (extend with tmle3 or ltmle for the full targeting step and IC inference)
library(sl3)
library(tmle3)
library(dplyr)

# Define a Super Learner library for nuisance estimation
lrn <- Lrnr_sl$new(
  learners = list(
    Lrnr_glm_fast$new(),
    Lrnr_ranger$new(),
    Lrnr_xgboost$new()
  )
)

# Outcome regression task: predict Y from (A, E, W)
Q_task <- make_sl3_Task(
  data       = dat,
  outcome    = "Y",
  covariates = c("A", "E", "W1", "W2", "W3")
)
Q_fit <- lrn$train(Q_task)

# Generate counterfactual predictions under each of four joint regimes.
# sl3 uses R6 classes, so call $predict() on the fitted learner rather
# than the generic predict() function.
predict_under <- function(a_val, e_val) {
  newdat <- mutate(dat, A = a_val, E = e_val)
  Q_fit$predict(make_sl3_Task(
    data       = newdat,
    outcome    = "Y",
    covariates = c("A", "E", "W1", "W2", "W3")
  ))
}

Q_11 <- predict_under(1, 1)
Q_01 <- predict_under(0, 1)
Q_10 <- predict_under(1, 0)
Q_00 <- predict_under(0, 0)

# Plug-in estimate of the additive interaction parameter
psi_add <- mean(Q_11 - Q_01) - mean(Q_10 - Q_00)

# Plug-in estimate of the multiplicative interaction parameter
psi_mult <- (mean(Q_11) / mean(Q_01)) /
            (mean(Q_10) / mean(Q_00))

list(psi_add = psi_add, psi_mult = psi_mult)
```

::: {.callout-tip}
## How to explain this to a non-methodologist
"We trained a flexible model that predicts outcomes from treatment, the
second exposure, and patient characteristics. Then we asked the model:
*what would happen if everyone got both, neither, or just one of the
two?* The interaction is how much the effect of treatment changes when
the second exposure is also present."
:::

### Layer 2: what full TMLE adds

The plug-in workflow above is honest about its assumption: it requires
$Q(A, E, W)$ to be consistently estimated. A full TMLE implementation
adds three pieces.

1. **Estimate the joint treatment mechanism** $g(A, E \mid W)$ (or the
   two factor mechanisms $g_A(A \mid W)$ and $g_E(E \mid W)$) with
   Super Learner.
2. **Target the four counterfactual-mean estimates** by updating the
   initial $Q$ along a least-favourable submodel that uses $g$ as the
   clever covariate, so that the targeted means $\widehat p_{ae}^*$
   solve the efficient influence-function equation for the joint
   regime mean [@vanderLaan2006tmle; @gruber2012tmlepackage].
3. **Use the influence curve** of the resulting interaction parameter
   (a difference of differences of influence functions) for valid
   Wald confidence intervals. When the contrast is computed in two
   steps from independently fitted TMLEs, the nonparametric bootstrap
   is a defensible alternative.

The result is doubly robust: the targeted estimate is consistent if
either $Q$ or $g$ is consistent, not both. The chapter does not assume
that any single R package exposes a turnkey "interaction parameter"
button. The two practical paths are: (i) define a custom parameter
object inside `tmle3` for the additive interaction contrast; or
(ii) use `ltmle` with two intervention nodes ($A$ and $E$), extract
the four targeted regime means, and compute $\widehat\psi_{\text{add}}$
and the bootstrapped CI from those.

### Diagnose overlap within joint exposure cells

Before estimating any causal contrast, fit the propensity score and
inspect the distribution within each level of $E$. This is the
joint-cell version of the standard positivity diagnostic introduced in
@sec-positivity-overlap.

```{r}
#| label: overlap-check
#| eval: false

library(ggplot2)

dat$ps <- predict(
  glm(A ~ W1 + W2 + W3 + E, data = dat, family = binomial),
  type = "response"
)

ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ paste0("E = ", E)) +
  labs(x = "Estimated propensity score",
       fill = "Treatment",
       title = "Treatment overlap within each level of the second exposure")
```

### Example output and interpretation

The four counterfactual means produced by the joint-intervention TMLE
populate a 2×2 cell-specific risk table. Numbers are illustrative.

| | $E = 0$ | $E = 1$ |
|---|---:|---:|
| **$A = 0$** (risk if untreated) | 0.10 | 0.12 |
| **$A = 1$** (risk if treated)   | 0.15 | 0.28 |

From these four numbers the analyst reads off both interaction
parameters directly:

- $\widehat{\psi}_{\text{add}} = (0.28 - 0.12) - (0.15 - 0.10) = 0.11$.
  Treatment is 11 percentage points more harmful (or beneficial) when
  $E = 1$ than when $E = 0$.
- $\widehat{\psi}_{\text{mult}} = (0.28 / 0.12) \div (0.15 / 0.10) =
  1.55$. The treated-versus-untreated risk ratio is about 1.5 times
  larger when $E = 1$.

Both scales agree about the direction; only the additive scale tells
you the absolute size of the heterogeneity that matters for clinical
decision-making.

```{mermaid}
xychart-beta
    title "Joint-intervention counterfactual risks from TMLE"
    x-axis ["A=0,E=0", "A=1,E=0", "A=0,E=1", "A=1,E=1"]
    y-axis "Estimated risk" 0 --> 0.30
    bar [0.10, 0.15, 0.12, 0.28]
```

### Alternative: stratified TMLE when $E$ is a non-manipulable modifier

When the second variable is a non-manipulable baseline modifier rather
than an intervenable exposure, the natural analysis is stratified TMLE:
fit a separate TMLE within each level of the modifier, then contrast
the subgroup-specific risk differences. That is the standard
effect-modification workflow and is covered in detail in
@sec-effect-modification §5–6, including code,
positivity diagnostics within strata, and uncertainty propagation. For
interaction analyses, the joint-intervention approach above is preferred
whenever both variables are conceptually manipulable, because it
estimates the four counterfactual means directly and avoids the
small-stratum instability of a separate TMLE per cell.

The cell-specific risk table is the single most useful artifact from an
interaction analysis. From it, any reader can recompute the additive
contrast, the multiplicative contrast, the number needed to treat in
each subgroup, and any other quantity. Lead with that table; let the
contrasts follow.

---

## 7. CTMLE as a sensitivity analysis

Collaborative TMLE (CTMLE) is covered in detail in
@sec-advanced-tmle. This section explains only when
it is worth running for an interaction analysis. CTMLE is **not** the
default first-line method for most readers of this chapter; standard
TMLE plus a curated Super Learner library is. CTMLE earns its place in
three specific situations.

- **As a sensitivity analysis when the propensity model is fragile.**
  CTMLE adaptively selects which covariates enter the propensity-score
  model, guided by whether including them improves the bias-variance
  trade-off for the target parameter. This is most useful exactly
  where interaction analyses tend to suffer: in joint exposure cells
  with a small number of patients and extreme propensity-score weights
  driven by covariates that are strongly associated with treatment but
  only weakly with the outcome.
- **As a check on whether collaborative variable selection changes
  the interaction estimate materially.** Run CTMLE alongside the main
  TMLE and report both. Agreement is reassuring. Material disagreement
  flags that the main TMLE was being pulled by a fragile propensity
  model and that the propensity-score adjustment set is doing more
  work than the data really support.
- **Not as the default for every interaction analysis.** CTMLE adds
  algorithmic complexity, can be sensitive to its own tuning choices,
  and is harder to communicate to non-methodologist reviewers than a
  standard TMLE with a transparent Super Learner library. Use it
  because the standard analysis raised a specific concern, not as a
  routine substitute.

See @sec-advanced-tmle §C-TMLE for the
implementation details and code. For interaction work the same code
applies, with the propensity-score model defined over the joint
exposure $(A, E)$ rather than over $A$ alone.

---

## 8. Multiplicity, reporting, and interpretation

Interaction analyses are uniquely vulnerable to multiplicity. The risk
is not theoretical: many published claims of subgroup heterogeneity have
not replicated. Three disciplines protect the analysis.

### Pre-specification and the confirmatory-exploratory split

Before unblinding to the outcome, write down:

- **Confirmatory modifiers**: a small number (typically two to five),
  chosen for clinical or biological motivation. These get formal
  inference with appropriate multiplicity adjustment.
- **Exploratory modifiers**: any additional candidates. These are
  reported descriptively, with prominent caveats, and are best treated as
  hypothesis-generating.

The PATH statement and methodological work on heterogeneity of treatment
effects converge on this distinction. Recent simultaneous-inference
methods for multiple subgroup TMLEs make confirmatory analysis with
several modifiers tractable [@wei2023simultaneous].

### Practical issues when screening many interactions

A growing number of RWE projects involve a high-dimensional screen of
many candidate pairwise interactions, often as part of a comorbidity
review or a co-medication safety analysis. The discipline below keeps
that kind of screen interpretable.

- **Pre-specify biologically or clinically plausible interactions**
  whenever possible. Treat the prespecified set as confirmatory and
  the rest as exploratory.
- **Avoid treating all pairwise combinations as confirmatory.** A 20
  by 20 comorbidity matrix produces 190 pairs; an unreflective screen
  guarantees several "significant" findings by chance.
- **Label high-dimensional screens as exploratory** in the manuscript.
  Report the number of interaction contrasts examined.
- **Report sparsity diagnostics** (counts and proportions in each
  joint exposure cell, across all pairs screened) and any practical
  positivity violations.
- **Use Super Learner or other flexible nuisance estimation** to
  reduce the bias contribution of misspecified outcome and treatment
  models, but do not imply that machine learning resolves
  non-positivity or unmeasured confounding. It does not.
- **Consider outcome-blind simulations or design-stage diagnostics**
  for major interaction analyses, so that estimator behavior is
  understood before the outcome is unblinded.

::: {.callout-tip}
## Checklist for a multi-interaction screen
- Is each interaction hypothesis pre-specified, or is the screen
  explicitly exploratory?
- Are all four joint exposure groups represented for every pair?
- Is the additive or multiplicative scale chosen *before* looking at
  the results, with a clear scientific justification?
- Are estimates standardized to the same target population?
- Are sparse cells, extreme propensity-score weights, or unstable
  fold-specific estimates reported?
- Are exploratory findings clearly labeled as hypothesis-generating
  rather than confirmatory?
:::

### Multiplicity adjustment

For confirmatory work with several prespecified modifiers:

- Simultaneous (multivariate) confidence intervals around the vector of
  subgroup contrasts are the most informative single-step adjustment.
- False-discovery-rate control (Benjamini-Hochberg) is appropriate when
  the goal is to flag a subset of subgroups for further study.
- Bonferroni adjustment is conservative but easy to defend when the
  prespecified set is small.

### Reporting essentials

The Knol and VanderWeele guidelines for interaction reporting are the
practical template [@knol2012recommendations; @vanderweele2014tutorial].
A well-reported interaction analysis includes, at minimum:

- **The four cell-specific effect estimates or risks**
  $\widehat p_{00}, \widehat p_{10}, \widehat p_{01}, \widehat p_{11}$,
  using one common reference category (the lowest-risk cell when any
  exposure is preventive).
- **The within-stratum effect of $A$ across levels of $E$**: the
  effect of $A$ when $E=0$ and the effect of $A$ when $E=1$, each with
  a confidence interval.
- **The within-stratum effect of $E$ across levels of $A$** when both
  exposures are part of the scientific question, in symmetric form to
  the above.
- **The additive interaction summary**: $\widehat\psi_{\text{add}}$ or
  RERI (and AP and $S$ when the team chooses to report them) with
  confidence intervals.
- **The multiplicative interaction summary**: $\widehat\psi_{\text{mult}}$
  with a confidence interval.
- **The confounders adjusted for**, including the rationale for the
  adjustment set and a note on whether matching, IPW, regression
  adjustment, or TMLE was used.
- **The confirmatory versus exploratory status** of the analysis,
  including any multiplicity adjustment and how the modifier or
  exposure-pair list was chosen.

A brief description of the Super Learner library, the targeting step,
the variance estimator, and the sensitivity analyses run (alternative
learner libraries, alternative confounder sets, CTMLE) belongs in
either the methods section or an appendix.

### Interpretation in absolute terms

Whenever the outcome scale permits, translate findings into absolute risk
or rate language. "The treatment reduces 5-year mortality by 8
percentage points among high-risk patients but only by 1 percentage
point among low-risk patients" is far more useful for clinical decisions
than "the relative-risk reduction is greater in high-risk patients."

### Common pitfalls

::: {.callout-warning}
## Common pitfalls in interaction analyses

- **Treating regression interaction coefficients as causal interaction.**
  A logistic or Cox product term is a property of the model's
  parameterization, not of the joint potential-outcome distribution.
- **Ignoring positivity in joint exposure cells.** A cohort that
  supports the main effect of $A$ may not support estimation of the
  $A \times E$ interaction; check the $(A, E)$ cell counts and the
  within-cell propensity distribution before interpreting any
  interaction estimate.
- **Choosing the scale post hoc.** Looking at both scales and reporting
  the one that gives a "significant" answer is a multiplicity violation
  in disguise. Pre-specify the scale, and report both.
- **Overinterpreting noisy estimates.** Interaction estimates are
  intrinsically noisier than main effects because they are differences
  of differences. A wide confidence interval is a finding, not a
  failure; the right response is to report it honestly, not to dichotomize
  on a p-value.
- **Confusing exploratory subgroup screening with confirmatory
  inference.** A subgroup that "popped out" of an exploratory scan must
  be replicated before it informs a clinical or policy decision.
:::

### Sensitivity analyses for interaction estimates

Because interaction estimates are derivatives of the outcome surface,
they are more sensitive to nuisance-model choices than main effects. A
short, prespecified sensitivity battery should accompany every
confirmatory interaction analysis:

- **Vary the nuisance learner library.** Re-run the analysis with a
  smaller library (e.g., GLM only) and a larger one (e.g., adding
  gradient boosting and neural-net learners). Stable estimates across
  libraries are reassuring; unstable estimates flag library or
  positivity sensitivity.
- **Vary the adjustment set.** Re-run with a minimal sufficient
  adjustment set and with a richer one. Large changes suggest residual
  confounding or model dependence.
- **Run a CTMLE sensitivity analysis** (Section 7) to check whether
  collaborative propensity-score selection moves the estimate.
- **Quantify unmeasured confounding tolerance** using a causal-gap or
  E-value-style framework: how strong would an unmeasured confounder
  need to be to explain away the observed interaction? This kind of
  quantitative bias analysis is especially valuable for interaction,
  where unmeasured confounders can act asymmetrically across exposure
  cells.

Pre-specify a short confirmatory list. Apply formal multiplicity control
to it. Report exploratory findings transparently, but with language that
makes their status clear. Translate every conclusion into absolute terms
at least once. Treat the sensitivity battery---learner library,
adjustment set, CTMLE, unmeasured-confounding bound---as part of the
primary analysis rather than as an afterthought.

---

## 9. Software for interaction analyses

The core R targeted-learning stack is introduced in
@sec-doubly-robust-tmle,
@sec-superlearner, and
@sec-advanced-tmle. The table below is the
short-list a beginner needs for interaction work, plus two
epidemiology-oriented options outside the tlverse and one Python entry.

| Package | Language | Best use in this chapter | Strength | Limitation |
|---|---|---|---|---|
| `tmle` | R | Single-treatment TMLE for $E[Y^a]$ inside each modifier stratum, then post-estimation contrast for $\psi_{\text{add}}$ | Stable CRAN release, easy first pass | No first-class joint-intervention parameter |
| `tmle3` + `sl3` | R | Joint-intervention TMLE for the four counterfactual means, with a user-defined interaction parameter | Modular and extensible; integrates with HAL, ranger, xgboost | Steeper learning curve than `tmle` |
| `ltmle` | R | Time-varying interactions; can also handle two intervention nodes ($A$ and $E$) at the same time point and return the four targeted regime means | Mature; supports censoring and dynamic regimes | Wide-format data layout takes practice |
| `SuperLearner` | R | Nuisance estimation for $Q$ and $g$ in any of the above | Curated learner library, well documented | Requires user choice of learners |
| `drtmle` | R | Robustness check when learners are aggressive or joint cells are sparse; returns DR, TMLE, and AIPW for comparison | Doubly robust asymptotic normality | Less interaction-specific tooling |
| `ctmle` | R | Sensitivity analysis when the propensity model is fragile (Section 7) | Collaborative variable selection for $g$ | Older API; less actively developed |
| `interactionR` | R | Compute and recode RERI / AP / synergy index from estimated risks or model output, with automatic preventive-exposure handling | Implements the Knol and VanderWeele recoding rule; useful even when nuisance estimation is done elsewhere | Reporting-focused, not an estimator on its own |
| `grf` | R | Exploratory CATE / heterogeneity surface; orienting tool before a confirmatory interaction analysis | Honest sample splitting, individual-level CATE estimates | Not doubly robust; should not be the primary interaction estimator |
| `zEpid` | Python | Epidemiology-oriented TMLE / AIPW / IPTW with explicit cross-fitting; useful when the team is Python-first | Cross-fitting available; epi-friendly API | Smaller learner ecosystem than R |
| `PyTMLE` | Python | Targeted estimation outside the R ecosystem | Active development | Less interaction-specific tooling than the R stack |
: Table 4. Core R and Python packages for interaction analyses. {.striped}

### Emerging methods for mixed-exposure interaction discovery

A small but rapidly developing set of R packages targets *discovery*
of interactions across many candidate exposures, rather than
estimation of a prespecified joint contrast. These are advanced or
emerging tools, useful when the candidate list of exposure pairs is
large (environmental mixtures, polypharmacy screens, comorbidity
matrices) and the analyst wants targeted-learning inference on the
discovery process itself.

- [`InterXshift`](https://github.com/blind-contours/InterXshift):
  pairwise additive-scale interaction under stochastic shifts of
  continuous exposures, estimated with CV-TMLE.
- [`CVtreeMLE`](https://github.com/blind-contours/CVtreeMLE):
  cross-validated decision-tree search for threshold combinations
  across multiple exposures, with TMLE inference on the resulting
  policy effect.
- [`SuperNOVA`](https://github.com/blind-contours/SuperNOVA):
  basis-function discovery (MARS, HAL) of variables, exposure pairs,
  and exposure-covariate combinations driving outcomes under
  stochastic intervention.

Use these as advanced complements to the workflow in Section 6, not as
the main starting point for a beginner. Their additional flexibility
also means more decisions about library composition, cross-fitting,
and exposure-shift definitions.

The standard reproducibility scaffolding (fixed seed, scripted
pipeline, documented analytic dataset, `sessionInfo()`) applies as in
every other chapter of this handbook.

---

## 10. Practical checklist for analysts

The checklist below summarizes the chapter into a single artifact. It can
be copied into a project's pre-analysis plan or a reviewer's referee
form.

| Step | Recommendation | Common failure mode |
|---|---|---|
| Question | State whether you study effect modification of one treatment or interaction of two exposures | Treating any product term as "interaction" |
| Estimand | Pre-specify additive and multiplicative estimands separately | Reporting only odds-ratio interaction |
| Design | Write a target trial protocol; define time zero, censoring, follow-up | Building subgroup models without a protocol |
| Modifiers | Limit confirmatory analyses to a small set of clinically motivated, pre-treatment modifiers | Mining many post-hoc modifiers |
| Confounding | Adjust for confounders of treatment and outcome only; avoid post-treatment variables | Overadjustment and collider bias |
| Positivity | Diagnose overlap overall *and within each modifier stratum* | Stratum-specific positivity failures hidden by overall overlap |
| Nuisance | Use a Super Learner library that is diverse but appropriate for the sample size | One opaque learner with no diagnostics |
| Estimation | TMLE within strata for small numbers of subgroups; simultaneous one-step TMLE for many | Naive stratified regression |
| Inference | Influence-function-based SEs; bootstrap for derived contrasts; cross-fitting for aggressive learners | Wald tests on derived contrasts assuming independence |
| Multiplicity | Confirmatory vs. exploratory split; simultaneous CIs or FDR for many modifiers | Many unadjusted subgroup tests |
| Reporting | Cell-specific risks; both scales; common reference category; coding decisions; sensitivity analyses | Only a p-value for the product term |
| Interpretation | Lead with absolute risks; sense-check against background incidence | Headline-only relative-risk language |
: Table 5. Practical checklist for interaction analyses in RWE. {.striped}

---

## 11. Time-varying interactions and marginal structural models

Sections 1 through 10 dealt with single time-point analyses: a baseline
treatment, a baseline modifier or second exposure, a single outcome
measurement. Many real-world questions involve treatment regimens that
evolve over time, with intermediate covariates that are simultaneously
confounders of later treatment and consequences of earlier treatment.
In that setting, baseline-only TMLE is no longer sufficient, and the
interaction question must be reframed in terms of sustained or dynamic
treatment strategies.

### When you need longitudinal methods

Longitudinal targeted learning becomes necessary when at least one of
the following is true:

- Treatment can change after baseline through initiation,
  discontinuation, or switching.
- Censoring is informative and is itself influenced by time-varying
  covariates.
- Time-varying confounders are themselves affected by past treatment,
  for example blood pressure during a hypertension trial.

For interaction analyses specifically, the question becomes: *does the
effect of a sustained treatment strategy differ across baseline
strata?* The modifier $M$ should still be a baseline variable, well
measured, and adequately supported across the relevant treatment
histories. Time-varying modifiers can also be of interest, but they
introduce additional positivity and identification challenges and are
typically handled inside a marginal structural model rather than by
stratification.

### LTMLE for sustained-treatment interactions

The `ltmle` package implements longitudinal TMLE for sequential
treatment and censoring nodes. It accepts a Super Learner library and
supports both static and dynamic intervention rules.

```{r}
#| label: ltmle-sketch
#| eval: false

library(ltmle)

# Sketch only: the data 'long_dat' is in wide format with
# baseline covariates, time-varying covariates L1..L3,
# treatments A1..A3, and outcome Y at the final time
fit_ltmle <- ltmle(
  data       = long_dat,
  Anodes     = c("A1", "A2", "A3"),
  Lnodes     = c("L1", "L2", "L3"),
  Ynodes     = "Y",
  abar       = list(c(1, 1, 1), c(0, 0, 0)),
  SL.library = c("SL.mean", "SL.glm", "SL.glmnet", "SL.ranger")
)

summary(fit_ltmle)
```

To estimate baseline-modifier-specific effects, fit `ltmle` separately
within each level of $M$ and contrast the resulting counterfactual means.
The output is an estimate of, for example, the one-year survival
probability under "always treat" versus "never treat" within each
subgroup, along with the contrast across subgroups.

When treatment can change month to month and intermediate health
markers both depend on past treatment and predict future treatment, a
single baseline analysis is no longer enough. The longitudinal version
of TMLE walks through the data one time step at a time, adjusting at
each step for what has already changed.

### Marginal structural models for interaction

An alternative is to fit a marginal structural model (MSM) with the
modifier $M$ entering as an interaction term. The MSM specifies how the
counterfactual mean depends on the treatment regime and on $M$, and the
interaction coefficient in the MSM is a parameter of the marginal
counterfactual distribution rather than of any conditional outcome
model. TMLE for MSMs estimates this coefficient by fitting nuisance
regressions, computing counterfactual outcomes under the regimes
indexed by the MSM, and then targeting the MSM coefficient directly.
The `tmleMSM` framework and the corresponding `ltmle` MSM facilities
support this workflow.

In an MSM-based interaction analysis the analyst should be explicit
about: which regimes the MSM indexes, whether $M$ is baseline or
time-varying, and how the MSM is parameterised on the additive or
multiplicative scale. The same scale considerations from Section 5
apply: pre-specify the scale, use a common reference, and report the
cell-specific counterfactual means whenever possible.

For sustained-treatment interaction questions, fit LTMLE within each
modifier stratum or fit a TMLE-MSM with the modifier as an interaction
term. Report the cell-specific counterfactual outcomes, for example
one-year survival under "always treat" versus "never treat" within
each subgroup, and the across-subgroup contrast. Diagnose positivity
across treatment histories, not only at baseline.

### Flexible estimands beyond binary exposures

The estimands defined in Section 2 assume binary exposures. In
real-world evidence, exposures are often continuous (dose, duration,
intensity), and joint positivity for any single value of a continuous
exposure is typically zero. The remedy is to redefine the estimand in
terms of **modified treatment policies** (MTPs): hypothetical
interventions that shift observed exposure rather than fix it at a
value [@diaz2021lmtp]. The interaction question then becomes: *does
the effect of shifting $A$ depend on the level or shift of $E$?*

An MTP-based interaction parameter contrasts the counterfactual mean
under a paired shift in $(A, E)$ against the counterfactual mean under
shifting only $A$, only $E$, or neither. The general-purpose `lmtp`
package implements TMLE and sequentially doubly robust estimators for
arbitrary MTPs, with support for time-varying treatments, censoring,
and Super Learner nuisance estimation [@diaz2021lmtp]. For
continuous-exposure interaction problems, common in nutritional
epidemiology, environmental health, and dose-response
pharmacoepidemiology, MTPs are the appropriate generalisation of the
binary joint-intervention framework.

When exposures are continuous, asking "what if everyone had exactly
this dose?" is unrealistic and unsupported by the data. Instead, ask
"what if everyone shifted their exposure a bit?" The interaction
question becomes whether the benefit of shifting one exposure depends
on the shift in the other.

### The `Xshift` family: TMLE for interactions in mixed exposures

A focused toolkit for stochastic-intervention interaction analysis
under mixed exposures, typical of environmental epidemiology, is the
`Xshift` family of R packages developed by McCoy and colleagues
[@mccoy2023interaction]. The packages share a common philosophy: define
the interaction parameter directly in terms of *contrasts of
counterfactual means under stochastic shifts*, and estimate it with
cross-validated TMLE (CV-TMLE) so that the data-adaptive selection of
which exposures to interrogate does not invalidate the confidence
intervals.

| Package | Estimand | Best use |
|---|---|---|
| [`InterXshift`](https://github.com/blind-contours/InterXshift) / [`JointXshift`](https://github.com/blind-contours/JointXshift) | Counterfactual mean under a *joint* shift of two exposures, contrasted with the *sum* of the two individual-shift means. Positive values indicate synergy; negative values indicate antagonism. | Pairwise interaction in a mixture, with valid CIs on the joint-vs-additive contrast |
| [`EffectXshift`](https://github.com/blind-contours/EffectXshift) | Counterfactual mean under a stochastic shift of one exposure, contrasted across data-adaptively learned subgroups | Effect modification of a continuous-exposure shift across covariate strata |
| [`IsoXshift`](https://github.com/blind-contours/IsoXshift) | Smallest joint shift in $(A, E)$ that achieves a target counterfactual outcome level, in the spirit of an isobologram | Policy-oriented synergy: which combinations of exposure reductions are most efficient? |

A typical `InterXshift` call looks like:

```{r}
#| label: interxshift-sketch
#| eval: false

library(InterXshift)

results <- InterXshift(
  w       = covariates,         # confounders (data frame)
  a       = exposures,          # mixed exposure matrix
  y       = outcome,            # outcome vector
  delta   = list(X1 = 1, X2 = 1),  # candidate shifts for each exposure
  n_folds = 5,                  # CV-TMLE folds
  top_n   = 3                   # how many exposure pairs to rank
)
```

The output is a ranked list of exposure pairs with the joint-shift
counterfactual mean, the sum-of-individual-shifts counterfactual mean,
their CV-TMLE-based contrast, and an associated 95% confidence interval
and $p$-value for each pair.

::: {.callout-note}
## When to reach for the `Xshift` family
- The exposures are continuous or many-valued (e.g., metals
  concentrations, pesticide loads, social-determinant indices), and
  fixing them at single values is biologically or pragmatically
  unrealistic.
- The analyst wants the additive-scale interaction parameter without
  collapsing exposures to binary categories.
- Several candidate exposure pairs are of interest, and the analysis
  needs valid confidence intervals despite data-adaptive ranking.
:::

### Discovery: which exposures interact?

The single-time-point worked example in Section 6 assumes the analyst
already knows which two exposures to interrogate. When the candidate
pool is larger (a typical mixture has 5 to 20 exposures), the
question shifts to *which pairs interact at all*, with valid inference
on the answer. Two TL-based discovery tools complement the
hypothesis-driven framework:

- [`CVtreeMLE`](https://github.com/blind-contours/CVtreeMLE) combines a
  cross-validated decision-tree search with TMLE to identify *threshold
  combinations* across multiple exposures that minimise or maximise
  outcome risk [@mccoy2023cvtreemle]. It produces $k$-fold-specific
  estimates with confidence intervals and a pooled oracle estimate
  with multiplicity-adjusted $p$-values.
- [`SuperNOVA`](https://github.com/blind-contours/SuperNOVA) uses
  basis-function estimators (MARS via `earth`, the highly adaptive
  lasso via `hal9001`) and an ANOVA-style variance decomposition to
  rank variables, exposure pairs, and exposure-covariate combinations
  by their contribution to the counterfactual outcome under stochastic
  intervention. It then estimates the relevant target parameters with
  TMLE on a held-out estimation sample.

These tools play, in the interaction setting, the same role that
treatment-effect variable importance (@sec-effect-modification §9)
plays for effect modification: they rank candidate parameters with
valid inference, so that follow-up confirmatory work can be focused on
the interactions that actually appear to matter. As with all discovery
methods, findings should be replicated or prespecified in a
follow-up study before being treated as confirmatory.

---

## 12. Conclusion

Interaction analyses in real-world evidence research are easy to do
poorly and demanding to do well. The path forward is not a single new
estimator, but a shift in workflow: start with a clearly defined causal
estimand on a chosen scale, build a target trial protocol around it, use
flexible nuisance estimation through Super Learner, target the resulting
estimates with TMLE, propagate uncertainty for the contrasts you report,
and discipline yourself to distinguish confirmatory from exploratory
work. The targeted-learning toolkit aligns naturally with these
disciplines, supports both binary point-treatment and time-varying
extensions, and---through CTMLE and `drtmle`---offers principled
sensitivity options when nuisance estimation is fragile.

### Minimum reporting for interaction analyses in RWE papers

At a minimum, a published interaction analysis in real-world evidence
should report: (1) a precisely worded scientific question and the chosen
estimand; (2) the target trial protocol (eligibility, time zero,
treatment strategies, follow-up, censoring, outcome); (3) the
prespecified set of confirmatory modifiers and the exploratory set, with
multiplicity strategy; (4) cell-specific risks under treatment and
control (or under each joint exposure level), with a common reference
category; (5) subgroup-specific risk differences and risk ratios with
confidence intervals; (6) additive and multiplicative interaction
contrasts with confidence intervals; (7) overlap diagnostics within
strata; (8) the Super Learner library and the variance estimator used;
(9) at least one sensitivity analysis, ideally including a CTMLE or
alternative-library check; and (10) an explicit interpretation in
absolute-risk language. With these elements in place, a reader can
reconstruct the estimand, judge the design, and assess the credibility
of the heterogeneity claim.

### What comes next

Three natural follow-up questions arise after a careful interaction
analysis.

The first is *how do we know the analysis is trustworthy?* That
question is the focus of the chapter on diagnostics and sensitivity
analysis (@sec-sensitivity-analysis), which
covers identifiability checks, nuisance-model assessment, positivity
diagnostics, and quantitative bias analysis for unmeasured
confounding.

The second is *which covariates contribute most to the heterogeneity
we found?* That question is the focus of the companion chapter on
effect modification (@sec-effect-modification),
which introduces treatment-effect variable importance measures and
causal forests as tools for ranking and visualising the modifiers
behind a contrast.

The third is *what should we do with the heterogeneity?* Interaction
analysis is often a first step toward more individualized
decision-making. If treatment benefit or harm differs across
clinically meaningful combinations of covariates, the natural next
step is to construct a treatment rule that assigns therapy as a
function of those covariates. The chapters on optimal treatment rules
and policy learning provide a more formal framework for translating
the heterogeneity that interaction analysis surfaces into a decision
rule with quantifiable expected value.

---

## Recommended reading

- VanderWeele and Knol [-@vanderweele2014tutorial]. The single best
  concise reference on additive versus multiplicative interaction and
  reporting standards.
- Petersen and van der Laan [-@petersen2014causal]. The causal roadmap
  applied to epidemiologic practice.
- Hernán and Robins [-@hernan2016target]. Target trial emulation as the
  organizing principle for RWE design.
- Talbot et al. [-@talbot2025tmle]. Modern best-practice tutorial for
  TMLE in applied work.
- van der Laan and Rubin [-@vanderLaan2006tmle]. Original TMLE paper.
- van der Laan, Polley, and Hubbard [-@vanderLaan2007superlearner]. Original
  Super Learner paper.
- Smith et al. [-@smith2023systematic]. Systematic review documenting
  the rapid uptake of TMLE in observational epidemiology.
- Wei et al. [-@wei2023simultaneous]. Simultaneous TMLE for multiple
  subgroup effects.
- Basten et al. [-@basten2025additive]. Recent treatment of additive
  interaction in IPD meta-analysis.
- McCoy et al. [-@mccoy2023interaction]. Stochastic-intervention TL for
  interaction and effect modification in mixed exposures; companion to
  the `Xshift` family of R packages.
- McCoy et al. [-@mccoy2023cvtreemle]. CVtreeMLE: cross-validated
  decision trees with TMLE for mixed-exposure threshold discovery.

## Annotated references

- **van der Laan and Rubin (2006).** Introduces TMLE as a plug-in
  estimator that is updated to solve the efficient influence function
  equation. The conceptual foundation for everything else in this
  chapter.
- **van der Laan, Polley, and Hubbard (2007).** Defines Super Learner as
  cross-validation-based stacking with an oracle inequality. Justifies
  using a diverse learner library without overcommitting to any single
  algorithm.
- **Petersen and van der Laan (2014).** A pedagogical formulation of the
  causal roadmap for epidemiologists. Recommended for any first reading
  on causal inference workflow.
- **VanderWeele and Knol (2014).** A tutorial that has shaped the modern
  reporting consensus on interaction. Especially clear on RERI, the
  reference-category problem, and scale dependence.
- **Hernán and Robins (2016).** The target-trial paper that has
  reorganized observational research around explicit protocol emulation.
  Required background for any RWE interaction analysis.
- **Smith et al. (2023).** A systematic review of TMLE in observational
  epidemiology. Useful for situating your analysis within current
  practice.
- **Talbot et al. (2025).** A best-practice tutorial that walks through
  TMLE implementation choices with applied researchers in mind.
- **Wei et al. (2023).** Simultaneous one-step TMLE for many subgroup
  effects, with valid joint confidence intervals. Read when you have
  more than a handful of modifiers.
- **Basten et al. (2025).** A clear recent treatment of additive
  interaction estimation, including in pooled / IPD settings. Useful
  for analysts whose decision question is public-health targeting.
- **McCoy et al. (2023, arXiv:2305.01849).** Develops the
  semi-parametric TL framework for interaction and effect modification
  via stochastic interventions on mixed exposures, and motivates the
  `InterXshift` / `EffectXshift` / `JointXshift` / `IsoXshift`
  packages.
- **McCoy et al. (2023, *JOSS*).** Companion paper to the `CVtreeMLE`
  package: cross-validated tree-based threshold discovery with TMLE
  inference for mixed-exposure policy questions.

---

## Appendix A: Full R code

The block below collects the joint-intervention worked example from
Section 6 into a single runnable script.

```{r}
#| label: appendix-a-full
#| eval: false

# ============================================================
# Appendix A. Joint-intervention TMLE for interaction
# Companion script to the chapter on interactions in RWE
# ============================================================

# --- Packages ---------------------------------------------------------
library(sl3)
library(tmle3)
library(dplyr)
library(ggplot2)

set.seed(2026)

# --- 1. Simulate cohort ------------------------------------------------
# n = 2000 keeps Appendix A fast in the WebR (interactive) build.
n  <- 2000
W1 <- rnorm(n)
W2 <- rbinom(n, 1, 0.4)
W3 <- runif(n, -1, 1)
E  <- rbinom(n, 1, plogis(-0.2 + 0.7*W1 - 0.4*W2))
A  <- rbinom(n, 1, plogis(-0.1 + 0.8*W1 + 0.5*W2 -
                            0.3*W3 + 0.2*E))
linpred_y <- -2.0 + 0.8*A + 0.4*E + 0.9*A*E +
             0.7*W1 - 0.5*W2 + 0.3*W3
Y <- rbinom(n, 1, plogis(linpred_y))

dat <- data.frame(Y, A, E, W1, W2, W3)

# --- 2. Joint-cell overlap diagnostic ---------------------------------
dat$ps <- predict(
  glm(A ~ W1 + W2 + W3 + E, data = dat, family = binomial),
  type = "response"
)
ggplot(dat, aes(x = ps, fill = factor(A))) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ paste0("E = ", E)) +
  labs(x = "Estimated propensity score",
       fill = "Treatment")

# --- 3. Outcome regression with Super Learner --------------------------
lrn <- Lrnr_sl$new(
  learners = list(
    Lrnr_glm_fast$new(),
    Lrnr_ranger$new(),
    Lrnr_xgboost$new()
  )
)

Q_task <- make_sl3_Task(
  data       = dat,
  outcome    = "Y",
  covariates = c("A", "E", "W1", "W2", "W3")
)
Q_fit <- lrn$train(Q_task)

# --- 4. Counterfactual predictions under four joint regimes -----------
# sl3 uses R6 classes, so call $predict() on the fitted learner rather
# than the generic predict() function.
predict_under <- function(a_val, e_val) {
  newdat <- mutate(dat, A = a_val, E = e_val)
  Q_fit$predict(make_sl3_Task(
    data       = newdat,
    outcome    = "Y",
    covariates = c("A", "E", "W1", "W2", "W3")
  ))
}
Q_11 <- predict_under(1, 1)
Q_01 <- predict_under(0, 1)
Q_10 <- predict_under(1, 0)
Q_00 <- predict_under(0, 0)

# --- 5. Plug-in interaction parameters --------------------------------
psi_add  <- mean(Q_11 - Q_01) - mean(Q_10 - Q_00)
psi_mult <- (mean(Q_11) / mean(Q_01)) /
            (mean(Q_10) / mean(Q_00))

list(psi_add = psi_add, psi_mult = psi_mult)

# To obtain a fully targeted estimate with influence-function-based
# inference, wrap the same nuisance estimates in a tmle3 workflow with
# a custom parameter for psi_add, or fit ltmle with two intervention
# nodes (A and E) and contrast the four counterfactual means.
# See Section 11 for time-varying / MSM extensions.
```

---

## Appendix B: Optional Python supplement

Python is supplementary, not central, to a targeted-learning workflow.
The most epidemiology-oriented option is `zEpid`, which provides risk
ratios and risk differences, interaction contrasts, and TMLE-style
estimators in an interface designed for epidemiologic practice.

```{python}
#| label: appendix-b-python
#| eval: false

# Optional Python supplement
# zEpid for effect measures and TMLE-style epidemiologic workflows
import pandas as pd
from zepid import RiskRatio, RiskDifference
from zepid.causal.doublyrobust import TMLE

# Assume df has columns: Y (outcome), A (treatment),
# E (second exposure), W1, W2, W3 (confounders)

# Risk difference within each level of the second exposure
for e_value in [0, 1]:
    d = df[df["E"] == e_value]
    rd = RiskDifference()
    rd.fit(d, exposure="A", outcome="Y")
    print(f"E = {e_value}")
    rd.summary()

# TMLE within each level of the second exposure
for e_value in [0, 1]:
    d = df[df["E"] == e_value].copy()
    tmle = TMLE(d, exposure="A", outcome="Y")
    tmle.exposure_model("W1 + W2 + W3")
    tmle.outcome_model("A + W1 + W2 + W3")
    tmle.fit()
    print(f"E = {e_value}: ATE = {tmle.average_treatment_effect}")
```

`DoWhy` and `CausalML` are useful complementary tools but are not primary
targeted-learning implementations. `DoWhy` provides causal-graph
specification, identification, and refutation tests; `CausalML` focuses
on conditional average treatment effects and uplift modeling, which are
better viewed as exploratory heterogeneity tools than as substitutes for
TMLE-based interaction inference. For an MPH chapter on interactions in
RWE, the R targeted-learning stack remains the recommended primary
implementation; the Python ecosystem is a useful complement.
