---
title: "Chapter 3.4: Effect Modification and Heterogeneous Treatment Effects"
format: html
---

*Estimated reading time: 35 minutes*

# Chapter 3.4

## Effect Modification and Heterogeneous Treatment Effects
*Identifying who benefits most (or least) from treatment*

A large observational study estimates that a new antihypertensive reduces cardiovascular events by 4 percentage points overall. A clinician reviewing the result asks: "Does this drug work equally well for my 82-year-old patient with CKD stage 4 and my 55-year-old patient with no comorbidities?" The overall average treatment effect cannot answer that question, because it averages over patients who may respond very differently to the drug.

A natural first instinct is to split the data into subgroups and run separate analyses. But this naive approach fails for several reasons. Unadjusted within-subgroup comparisons do not control for confounding that may vary across strata. Testing many subgroups inflates the false positive rate. And the "winner’s curse" means that whichever subgroup looks most impressive is also the most likely to be a false lead. A principled framework is needed: one that defines subgroup-specific causal estimands, controls confounding within each subgroup, and honestly distinguishes confirmatory from exploratory findings.

This chapter provides that framework. You will learn to:

1. Define subgroup-specific causal estimands (the CATE)
2. Use TMLE to estimate prespecified subgroup effects with double robustness
3. Use causal forests as an exploratory, data-adaptive tool for discovering heterogeneity
4. Recognize the key pitfalls: positivity fragility in small subgroups, multiple comparisons, and the distinction between confirmatory and exploratory findings

The chapter proceeds from estimands (Section 2) through identification (Section 3), to estimation with TMLE (Sections 5--6) and causal forests (Sections 7--8), and closes with interpretation guidance (Sections 10--11).

::: {.callout-note}
## Learning objectives
- Define effect modification, heterogeneous treatment effects (HTE), and the conditional average treatment effect (CATE)
- Distinguish interaction from effect modification in the causal inference framework
- Use TMLE to estimate subgroup-specific causal effects
- Understand causal forests as a data-adaptive approach to discovering treatment effect heterogeneity
- Recognize the risks of subgroup analyses: multiple testing, post-hoc selection, and the winner's curse
:::

*Causal conclusions require that the identification assumptions hold; see [Chapter 1.2](01-02-causal-roadmap.qmd) for details. When flexible machine learning is used for nuisance estimation, valid inference typically requires cross-fitting or a cross-validated TMLE variant.*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

---

## 1. Why Study Effect Modification?

**Key point.** The ATE tells you the mean treatment effect in the population, but patients differ in age, comorbidities, biomarkers, baseline risk, and genetics. Clinically relevant questions (*Does the drug work better in high-risk patients? Are cardiovascular risks higher with CKD? Does benefit differ by frailty?*) require studying **who benefits most** (or least). Effect modification is essential for translating population-level evidence into individualized clinical and policy recommendations.

---

## 2. Causal Estimands for Effect Modification

To study heterogeneity, we define **conditional average treatment effects (CATE)**.

:::{.callout-note}
## What Is a CATE? (Plain-Language Version)

The **Conditional Average Treatment Effect (CATE)** answers a simple question: *"What is the average treatment effect for people who share a specific characteristic?"*

For example, the CATE among women tells you the average effect of the treatment *specifically among women*. The CATE at age 80 tells you the expected effect *for an 80-year-old*.

Formally, let \(X\) be a baseline covariate or set of covariates. The CATE is:

\[
CATE(x) = E[Y(1) - Y(0) \mid X = x].
\]

For categorical variables (e.g., age group):

\[
ATE_{	ext{age<75}} = E[Y(1) - Y(0) \mid 	ext{age} < 75],
\]

and similarly for other strata.

Think of it this way: the ATE gives you one number for *everyone*. The CATE gives you a *different number for each subgroup*, letting you see where the treatment works best (or worst).
:::

Your choice of estimand must match the scientific question:

- **Prespecified subgroups** (sex, race, CKD stage)
- **Post hoc continuous moderators** (eGFR, age)
- **Machine-learned subgroups** (via trees / forests)

---

## 3. Identification Assumptions

:::{.callout-warning}
## Assumptions Still Apply and Get Harder in Subgroups

Effect modification estimands inherit the same identification assumptions as the overall ATE: exchangeability, positivity, and consistency (see [Chapter 1.2](01-02-causal-roadmap.qmd)). Additionally, moderator variables must be **measured before treatment**, and you should avoid conditioning on post-treatment variables.

**Pitfall.** Positivity is especially fragile in subgroup analyses. When you slice the data into smaller groups, some subgroups may have very few treated or untreated individuals. Always check that both treatment arms are adequately represented in every subgroup you analyze.
:::

---

## 4. A Workflow for Effect Modification

A practical workflow for investigating effect modification:

1. **Define causal estimands** (subgroup-specific ATE, CATE(x))
2. **Choose moderators** (prespecified vs exploratory)
3. **Estimate nuisance functions** with SuperLearner
4. **Estimate heterogeneous effects** using TMLE or causal forests
5. **Quantify uncertainty**
6. **Document all steps and prespecifications**

**Key point.** Decide your subgroups and moderators before looking at results. Prespecification protects against cherry-picking. When you explore post hoc, label those findings clearly as hypothesis-generating.

We now walk through TMLE and causal forest approaches.

---

## 5. TMLE for Subgroup-Specific Effects

:::{.callout-important}
## TMLE for Subgroup Effects

**Practical implementation.** Subset your data to a prespecified subgroup and run TMLE within that subset to estimate $ATE_g = E[Y(1) - Y(0) \mid X_g = 1]$. Because TMLE is doubly robust and uses SuperLearner, you get reliable subgroup-specific estimates, as long as you have enough data and positivity holds.
:::

Example: ATE among patients younger than 75.

```{r, eval=FALSE}
library(tmle)
library(SuperLearner)
library(tidyverse)

# Assume dat has columns: Y (outcome), A (treatment), age, cvd, eGFR, etc.
set.seed(123)

dat <- tibble(
  Y = rbinom(2000, 1, 0.2),
  A = rbinom(2000, 1, 0.5),
  age = rnorm(2000, 75, 6),
  cvd = rbinom(2000, 1, 0.4),
  eGFR = rnorm(2000, 60, 15)
)

dat <- dat %>%
  mutate(age_under75 = if_else(age < 75, 1, 0))

sl_lib <- c("SL.glm", "SL.ranger", "SL.mean")

tmle_sub <- tmle(
  Y = dat$Y[dat$age_under75 == 1],             # <1>
  A = dat$A[dat$age_under75 == 1],             # <1>
  W = dat %>%
    filter(age_under75 == 1) %>%
    select(cvd, eGFR),                         # <2>
  family = "binomial",
  Q.SL.library = sl_lib,                       # <3>
  g.SL.library = sl_lib                        # <3>
)

tmle_sub$estimates$ATE
```
1. Subset Y and A to the target subgroup before passing to `tmle()`; this estimates the **subgroup-specific ATE** $E[Y(1) - Y(0) \mid \text{age} < 75]$ rather than the marginal ATE; the estimand is conditional on being in the subgroup
2. W contains the **confounders** for the subgroup analysis. Notably, the subgrouping variable `age_under75` itself is **excluded** from W; including it would condition the propensity score model on the stratification variable, collapsing variation and potentially destabilizing the estimate
3. The same SL library is used for both the outcome model (Q) and propensity score (g); in small subgroups you may need to simplify the library or reduce cross-validation folds to prevent empty folds

Interpretation:

> Among patients younger than 75, the TMLE-estimated risk difference is …

This approach:

- Gives valid subgroup-specific estimates  
- Leverages SuperLearner for nuisance models  

Limitations:

- Does not provide a *continuous* CATE curve  
- Requires prespecified groups and adequate sample size in each  

---

## 6. TMLE Extensions for Continuous Effect Modification

For continuous effect modifiers, you may want:

\[
CATE(x) = E[Y(1) - Y(0) \mid X = x]
\]

for all x (e.g., all ages). TMLE-based approaches often use:

- Projection of the CATE onto basis functions of X (splines, polynomials)
- Estimation of projection coefficients via TMLE
- Reconstruction of \( CATE(x) \) from the projected function

:::{.callout-tip}
## CATE-TMLE: A Smooth, Doubly Robust Curve

**Key idea.** Project the CATE onto basis functions of $X$, estimate each projection coefficient via TMLE, and reconstruct a smooth $\hat{CATE}(x)$. This gives you a continuous picture of how the treatment effect varies across a modifier, with the doubly robust guarantees of TMLE.
:::  

---

## 7. Modern ML Approach: Causal Forests

:::{.callout-note}
## Causal Forests in Plain Language

**Definition.** Causal forests are tree-based methods designed to estimate CATEs. Think of them as “smart subgroup finders” that search across covariates for splits maximizing treatment-effect heterogeneity. They use honest sample splitting and averaging across many trees to produce smooth, individual-level CATE estimates $\hat\tau(x)$ with approximate pointwise confidence intervals.
:::

Causal forests are powerful exploratory tools, but they come with important limitations.

:::{.callout-warning}
## Caveats About Causal Forests

**Pitfall.** Causal forests are not doubly robust, can overfit to noise in small samples, and their output is associational unless embedded in a rigorous causal framework. Treat causal forest results as **hypothesis-generating**, not confirmatory. Validate findings in independent data or prespecify them in future studies.
:::

---

## 8. Implementing Causal Forests in R

We use the `grf` package.

```{r, eval=FALSE}
# install.packages("grf")
library(grf)

set.seed(2028)

# Simulate data with treatment effect heterogeneity
n <- 3000
X <- tibble(
  age = rnorm(n, 75, 6),
  cvd = rbinom(n, 1, 0.4),
  risk_score = rnorm(n, 0, 1)
)

A <- rbinom(n, 1, plogis(-0.5 + 0.2 * X$age - 0.8 * X$risk_score))
# Treatment effect depends on risk_score
tau_true <- -0.05 + 0.1 * (X$risk_score > 0)
Y <- rbinom(n, 1, plogis(-2 + tau_true * A + 0.05 * (X$age - 75) + 0.5 * X$cvd))

X_mat <- as.matrix(X)

cf <- causal_forest(
  X = X_mat,   # <1>
  Y = Y,       # <2>
  W = A        # <3>
)

# Individualized CATE estimates
tau_hat <- predict(cf)$predictions

summary(tau_hat)
```
1. `X`: covariate matrix that includes all potential effect modifiers; the forest will search for splits across these covariates that maximize treatment effect heterogeneity; unlike TMLE's W, this should include variables you suspect are **moderators** even if they are not confounders
2. `Y`: outcome vector; `causal_forest` internally uses a Robinson (1988) decomposition, which separately estimates $E[Y \mid X]$ and $E[W \mid X]$ before fitting the heterogeneous treatment effect
3. `W = A`: the binary treatment indicator (not to be confused with W the confounder matrix); `causal_forest` uses an **honest splitting** approach: each tree splits on X for heterogeneity but estimates treatment effects using held-out data from that split, preventing overfitting

### Visualizing CATE vs. risk score

```{r, eval=FALSE}
plot(X$risk_score, tau_hat,
     xlab = "Baseline risk score",
     ylab = "Estimated CATE",
     pch = 16, cex = 0.3)
abline(h = 0, col = "red")
```

You should see different CATE patterns for low vs high risk scores.

---

## 9. Subgrouping Using CATE Estimates

We can create subgroups such as:

- “High predicted benefit” vs “low predicted benefit”

```{r, eval=FALSE}
threshold <- median(tau_hat)

res <- X %>%
  mutate(
    CATE_hat = tau_hat,
    group = if_else(CATE_hat > threshold, "high benefit", "low benefit")
  )

table(res$group)

res %>%
  group_by(group) %>%
  summarize(
    mean_CATE = mean(CATE_hat),
    .groups = "drop"
  )
```

This is **exploratory**, not confirmatory.  
It can inform:

- Hypothesis generation  
- Prespecified subgroup definitions in future trials  

---

## 10. Causal Forest vs. TMLE: Which Should You Use?

:::{.callout-tip}
## Choosing the Right Tool for the Job

**Key idea.** Use **TMLE** when you know *which* subgroups to test (prespecified, doubly robust, parameter-focused). Use **causal forests** when you want the data to *show* you where heterogeneity exists (exploratory, data-adaptive). In practice, the strongest analyses do both: TMLE for confirmatory subgroup effects, then causal forests to explore residual heterogeneity and generate new hypotheses.
:::

---

## 11. Interpretation and Reporting: Avoiding Pitfalls

:::{.callout-warning}
## The Multiple Comparisons Trap

**Pitfall.** Testing 20 subgroups at alpha = 0.05 yields one “significant” finding by chance alone. Prespecify main effect modifiers, correct for multiplicity (FDR or family-wise error), emphasize uncertainty, treat causal forest findings as exploratory, and always present the overall ATE alongside any heterogeneity results.
:::  

---

## 12. Example: Combining TMLE and Causal Forests

```{r, eval=FALSE}
# Step 1: ATE with TMLE
sl_lib <- c("SL.glm", "SL.ranger", "SL.mean")

tmle_ate <- tmle(
  Y = Y,
  A = A,
  W = X,                        # <1>
  family = "binomial",
  Q.SL.library = sl_lib,        # <2>
  g.SL.library = sl_lib         # <2>
)

tmle_ate$estimates$ATE

# Step 2: CATEs with causal forest (already computed as tau_hat)
summary(tau_hat)
```
1. `W = X` passes the full covariate matrix including potential effect modifiers like `age`, `cvd`, `risk_score`; the TMLE ATE averages over all of these, while the subsequent causal forest step decomposes heterogeneity across them
2. Using the same `sl_lib` for Q and g is standard when the covariate space is the same; consider richer libraries for g if treatment assignment is more complex (e.g., multiple prescribers, non-random clinic assignment)

- TMLE gives a **population-level effect** (with solid identifiability and efficiency).
- Causal forest gives **individual-level estimated CATEs** to explore how effects vary.

---

## Running Example: Antiretroviral Therapy Cohort

See the [Running Case Study](00-case-study.qmd) for the full data description. Effect modification is clinically relevant in the HAART cohort because the benefit of treatment may vary across patient characteristics. Patients with very low baseline CD4 counts face the highest mortality risk and may experience the largest absolute risk reduction from HAART, while patients with higher CD4 at baseline have less room to benefit. Age is another plausible modifier, since immune reconstitution dynamics differ for older patients.

```{r}
#| eval: false
source("_data/load-haart-cohort.R")

# Define clinically meaningful subgroups by baseline CD4
haart_baseline <- haart_baseline %>%
  mutate(cd4_group = if_else(cd4_baseline < median(cd4_baseline),
                             "low_cd4", "high_cd4"))

# Prespecified subgroup analysis: estimate the CATE within each CD4 group
# using TMLE with confounders age and sex.
# A causal forest could also explore heterogeneity across cd4_baseline
# and age simultaneously as an exploratory complement.
table(haart_baseline$cd4_group, haart_baseline$ever_haart)
```

Before estimating subgroup-specific effects, verify that positivity holds within each stratum (Chapter 3.2). Both the HAART-treated (`ever_haart = 1`) and untreated groups should be well represented in each CD4 subgroup. With only 1,200 patients and 31 deaths, very fine stratification will produce sparse cells and unstable CATE estimates, so analyses should focus on a small number of prespecified subgroups rather than exhaustive slicing.

---

## 13. Summary

:::{.callout-tip}
## Key Takeaways

- Define causal estimands (CATE) before looking at data; use TMLE for confirmatory subgroup effects and causal forests for exploratory heterogeneity
- Positivity and multiple comparisons are the two biggest threats in subgroup analyses
- Always report the overall ATE first, prespecify subgroups, correct for multiplicity, and label exploratory findings honestly
:::

---

## Sources and further reading

- Athey S, Imbens GW (2016). Recursive partitioning for heterogeneous causal effects. *PNAS* 113(27):7353-7360.
- Wager S, Athey S (2018). Estimation and inference of heterogeneous treatment effects using random forests. *JASA* 113(523):1228-1242.
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.
- Luedtke AR, van der Laan MJ (2016). Super learning of an optimal dynamic treatment rule. *Int J Biostat* 12(1):305-332.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapter 4 (Effect modification). [Free online](https://www.hsph.harvard.edu/miguel-hernan/causal-inference-book/)
- VanderWeele TJ (2009). On the distinction between interaction and effect modification. *Epidemiology* 20(6):863-871.
- `grf` R package (causal forests): [CRAN](https://cran.r-project.org/package=grf)
- `tmle` R package: [CRAN](https://cran.r-project.org/package=tmle)

## Check your understanding

<details>
<summary><strong>Q1. You run a causal forest and find that patients over 80 have a larger treatment effect than younger patients. Is this a confirmatory or exploratory finding?</strong></summary>

This is an **exploratory** finding. Causal forests are data-adaptive methods that search across covariates to discover heterogeneity. Because the "patients over 80" subgroup was not prespecified before the analysis, the finding is hypothesis-generating. To treat it as confirmatory, you would need to prespecify this subgroup in a new study or validation dataset, with appropriate multiplicity correction. Presenting it as a confirmed subgroup effect risks the winner's curse and inflated type I error.

</details>

<details>
<summary><strong>Q2. Why is positivity especially fragile in subgroup analyses compared to the overall ATE analysis?</strong></summary>

Positivity requires that both treated and untreated individuals exist within every stratum of confounders. When you slice the data into subgroups, the sample size in each stratum shrinks, making it increasingly likely that some confounder-by-treatment combinations have zero or near-zero representation. For example, if you estimate the ATE among patients over 80 with CKD stage 4, there may be very few (or no) untreated individuals in that subgroup, making the causal effect unidentifiable there. The overall ATE analysis pools across subgroups and is therefore less vulnerable to these sparse-data positivity violations.

</details>

<details>
<summary><strong>Q3. What is the key difference between using TMLE vs causal forests for studying effect modification?</strong></summary>

**TMLE** is designed for **confirmatory** inference on prespecified subgroups or estimands. It is doubly robust (consistent if either the outcome model or propensity model is correct), provides valid confidence intervals grounded in semiparametric efficiency theory, and targets a clearly defined causal parameter. **Causal forests** are designed for **exploratory** heterogeneity discovery: they search data-adaptively across covariates to find where treatment effects vary, but they are not doubly robust, are more susceptible to overfitting, and their output should be treated as hypothesis-generating rather than confirmatory. The strongest analyses use both: TMLE for pre-planned subgroup effects, and causal forests to explore where additional heterogeneity might exist.

</details>

## What comes next

Effect modification analysis raises an important follow-up question: how do you know if your analysis is trustworthy? The next chapter provides a systematic framework for diagnostics, covering identifiability checks, nuisance model assessment, estimator behavior, and sensitivity analyses.

```{r}
sessionInfo()
```
