# Time-to-Event, Censoring, and Competing Risks with Targeted Learning {#sec-tte-competing}

::: {.callout-tip}
## Learning Objectives

After completing this chapter, you will be able to:

1. Distinguish risk at time *t*, restricted mean survival time (RMST), and hazard ratios as causal estimands, and articulate why hazard ratios have problematic causal interpretations
2. Map ICH E9(R1) intercurrent event strategies to censoring and handling approaches in TMLE
3. Implement IPCW for informative censoring and diagnose weight instability at long follow-up
4. Compare Fine-Gray and cause-specific approaches to competing risks, and understand how TMLE handles competing events
5. Apply survival TMLE to estimate a marginal risk difference at time *t* with competing risks
:::

## Time-to-Event Estimands {#sec-tte-estimands}

### Risk at Time *t*

The most interpretable causal estimand in survival analysis is the counterfactual risk (cumulative incidence) at a pre-specified time point *t*:

$$\Psi = P(T^{a=1} \leq t) - P(T^{a=0} \leq t)$$

This is a marginal risk difference: the difference in the probability of the event by time *t* under treatment vs. control, averaged over the target population. It has a clear causal interpretation under standard identifiability assumptions.

### Restricted Mean Survival Time (RMST)

The RMST up to time $\tau$ is the area under the survival curve from 0 to $\tau$:

$$\text{RMST}(\tau) = \int_0^\tau S(t) \, dt$$

The causal contrast is the difference in RMST between treatment arms. This estimand summarizes the entire survival curve up to $\tau$ without requiring proportional hazards, making it robust and interpretable.

### The Hazard Ratio Problem

::: {.callout-warning}
## Hazard Ratios Are Not Straightforward Causal Parameters

The hazard ratio has two fundamental problems as a causal quantity:

1. **Non-collapsibility**: Even in the absence of confounding, the conditional hazard ratio (within strata) does not equal the marginal hazard ratio. This means that a perfectly adjusted hazard ratio from a Cox model estimates a different quantity than the marginal hazard ratio in the population---even when adjustment is correct.

2. **Built-in selection bias**: The hazard at time *t* is defined only among those who have survived to *t*. Under treatment, the survivors at time *t* are a *selected* subset of the original population. Even in a randomized trial, comparing hazards at later time points compares groups that differ in composition because the treatment has already differentially removed individuals from the risk set.

For causal inference, prefer risk differences, risk ratios at fixed time points, or RMST differences. The target trial framework [@hernan2016target] emphasizes that the causal question should drive the estimand, and hazard ratios rarely correspond to a well-defined causal contrast.
:::

## ICH E9(R1) Intercurrent Events and Censoring Strategies {#sec-ich-e9r1}

The ICH E9(R1) addendum defines five strategies for handling intercurrent events (events that affect interpretation of the treatment effect). Each maps to a specific censoring or outcome-handling approach in TMLE.

| ICH E9(R1) Strategy | Description | TMLE Implementation |
|---------------------|-------------|---------------------|
| **Treatment policy** | Outcome observed regardless of intercurrent event | No censoring at intercurrent event; analyze all observed outcomes |
| **Composite** | Intercurrent event is part of the outcome | Redefine outcome to include the intercurrent event (e.g., death or discontinuation) |
| **Hypothetical** | Outcome in a world where the intercurrent event could not occur | Censor at intercurrent event and use IPCW to adjust for informative censoring |
| **Principal stratum** | Effect among those who would not experience the intercurrent event under either treatment | Requires principal stratification; typically not handled by standard TMLE |
| **While on treatment** | Outcome while on assigned treatment | Censor at treatment discontinuation; use IPCW for informative censoring |

: ICH E9(R1) strategies mapped to TMLE approaches {#tbl-ich-e9r1}

The **hypothetical** and **while-on-treatment** strategies both require IPCW. The key difference is what triggers censoring: under the hypothetical strategy, a biological event (e.g., a competing risk) triggers censoring, while under the while-on-treatment strategy, treatment discontinuation triggers censoring.

## Informative Censoring and IPCW {#sec-ipcw}

### The Problem

Standard Kaplan-Meier and Cox regression assume that censoring is non-informative---that censored individuals have the same future risk as those who remain under observation. When censoring is related to prognosis (e.g., sicker patients are more likely to switch treatments or drop out), this assumption is violated and naive estimates are biased.

### Inverse Probability of Censoring Weights

IPCW corrects for informative censoring by upweighting individuals who remain uncensored but are similar to those who were censored. At each time point $t_k$, the censoring weight for an individual still under observation is:

$$W_C(t_k) = \prod_{j=1}^{k} \frac{1}{P(C > t_j \mid C > t_{j-1}, \bar{A}_{j-1}, \bar{L}_{j-1})}$$

### Cumulative Weight Instability

::: {.callout-important}
## The Cumulative Product Problem

Because IPCW weights are a *product* over time points, instability compounds. Even if the probability of remaining uncensored at each time point is 0.95, after 20 time points the weight becomes $1/0.95^{20} \approx 2.65$. If some individuals have per-period censoring probabilities of 0.80, their weights explode: $1/0.80^{20} \approx 87$.

Practical consequences:

- **Positivity violations accumulate**: At later time points, fewer covariate strata have uncensored individuals, violating positivity
- **Effective sample size shrinks**: Even with a large nominal sample, the effective sample size can become very small at later time points
- **Variance inflation**: A few highly weighted individuals dominate the estimate

**Remedies**: Truncate weights (with sensitivity analysis), restrict the analysis to a time horizon where weights remain stable, or use TMLE which can stabilize estimation through its targeting step.
:::

## Competing Risks {#sec-competing-risks}

A competing risk is an event that prevents the outcome of interest from occurring. For example, when studying time to cardiovascular death, non-cardiovascular death is a competing event.

### Two Approaches

**Cause-specific approach**: Treat competing events as censoring. This estimates the cause-specific hazard, which answers: "Among those currently alive, what is the instantaneous rate of the event of interest?" However, the resulting survival curve does not have a probability interpretation when competing risks exist.

**Fine-Gray (subdistribution) approach**: Keep individuals in the risk set after experiencing a competing event. This directly models the cumulative incidence function (CIF) and answers: "What is the probability of the event of interest by time *t*, accounting for the fact that competing events prevent it?"

### TMLE for Competing Risks

TMLE can handle competing risks in two ways:

1. **Competing event as censoring**: Apply IPCW for the competing event, treating it like any other form of informative censoring. This targets the cause-specific risk in a hypothetical world without the competing event.

2. **Composite outcome**: Combine the event of interest and the competing event into a single outcome. This targets the risk of "any event" by time *t*.

3. **Direct CIF estimation**: Use a TMLE that directly targets the cumulative incidence function. This is the most principled approach and corresponds to the subdistribution quantity.

The choice depends on the scientific question:

- "What would the risk be if we could eliminate the competing event?" $\rightarrow$ Cause-specific with IPCW
- "What is the actual probability of the event in the real world where competing events exist?" $\rightarrow$ CIF approach

## Worked Example: Survival TMLE with Competing Risks {#sec-tte-example}

### Data Simulation

We simulate a two-arm study with 500 patients, administrative censoring, and one competing event (non-cardiovascular death competing with cardiovascular death).

```{r}
#| label: sim-tte-data
#| code-fold: false

set.seed(2026)
n <- 500

# Baseline covariates
age <- rnorm(n, mean = 65, sd = 10)
female <- rbinom(n, 1, 0.45)
comorbidity <- rbinom(n, 1, 0.3)

# Treatment assignment (observational, depends on covariates)
ps <- plogis(-1 + 0.02 * age + 0.3 * comorbidity)
A <- rbinom(n, 1, ps)

# Event times (cardiovascular death)
lambda_cv <- exp(-5 + 0.03 * age + 0.5 * comorbidity - 0.4 * A)
T_cv <- rexp(n, rate = lambda_cv)

# Competing event times (non-cardiovascular death)
lambda_noncv <- exp(-5.5 + 0.04 * age + 0.2 * comorbidity - 0.1 * A)
T_noncv <- rexp(n, rate = lambda_noncv)

# Administrative censoring at 5 years
C_admin <- 5

# Observed data
T_obs <- pmin(T_cv, T_noncv, C_admin)
event_type <- ifelse(T_obs == C_admin, 0,       # censored
               ifelse(T_obs == T_cv, 1,          # CV death
                      2))                         # non-CV death

dat <- data.frame(
  age = age,
  female = female,
  comorbidity = comorbidity,
  A = A,
  time = T_obs,
  event = event_type
)

# Summary
table(dat$event)
summary(dat$time)
```

### Naive Kaplan-Meier Estimate

The standard Kaplan-Meier estimator ignores confounding and treats competing events as censoring.

```{r}
#| label: km-naive
#| code-fold: false

library(survival)

# Naive KM treating competing events as censored
km_fit <- survfit(Surv(time, event == 1) ~ A, data = dat)

plot(km_fit, col = c("red", "blue"), lwd = 2,
     xlab = "Time (years)", ylab = "Survival probability",
     main = "Naive Kaplan-Meier (ignoring confounding & competing risks)")
legend("bottomleft", c("Control (A=0)", "Treated (A=1)"),
       col = c("red", "blue"), lwd = 2)
```

::: {.callout-note}
This Kaplan-Meier estimate is biased for two reasons: (1) it does not adjust for confounding (age and comorbidity affect both treatment and outcome), and (2) it treats competing events as non-informative censoring.
:::

### Cumulative Incidence (Aalen-Johansen)

A better descriptive analysis uses the Aalen-Johansen estimator, which accounts for competing risks but still does not adjust for confounding.

```{r}
#| label: aalen-johansen
#| code-fold: false

# Cumulative incidence with competing risks
library(survival)

# Multi-state event indicator
dat$event_ms <- factor(dat$event, levels = 0:2,
                       labels = c("censor", "CV death", "Non-CV death"))

cuminc_fit <- survfit(Surv(time, event_ms) ~ A, data = dat)
plot(cuminc_fit, col = c("red", "blue"), lwd = 2,
     xlab = "Time (years)", ylab = "Cumulative incidence",
     main = "Aalen-Johansen Cumulative Incidence by Treatment")
```

### TMLE for Risk Difference at Time *t*

We estimate the causal risk difference for cardiovascular death at $t = 3$ years using a TMLE approach. This adjusts for confounding and can properly handle competing events.

```{r}
#| label: tmle-survival
#| eval: false
#| code-fold: false

# --- TMLE for survival: risk difference at t = 3 ---
# This uses a discretized hazard TMLE approach
# Requires: survtmle package (install from GitHub if needed)
# devtools::install_github("benkeser/survtmle")

library(survtmle)

# Discretize time for TMLE (quarterly intervals)
t_target <- 3  # 3-year risk

# Prepare data for survtmle
# Treating competing events as censoring (cause-specific approach)
result <- survtmle(
  ftime   = dat$time,
  ftype   = dat$event,
  trt     = dat$A,
  adjustVars = dat[, c("age", "female", "comorbidity")],
  t0      = t_target,
  SL.ftime = c("SL.glm", "SL.mean"),
  SL.ctime = c("SL.glm", "SL.mean"),
  SL.trt   = c("SL.glm", "SL.mean"),
  method   = "hazard",
  ftypeOfInterest = 1  # CV death
)

# Extract risk difference
print(result)

# Risk under treatment vs control at t = 3
risk_A1 <- result$est[2]  # P(T <= 3 | do(A=1))
risk_A0 <- result$est[1]  # P(T <= 3 | do(A=0))
RD <- risk_A1 - risk_A0

cat(sprintf("Risk under A=1: %.3f\n", risk_A1))
cat(sprintf("Risk under A=0: %.3f\n", risk_A0))
cat(sprintf("Risk difference: %.3f\n", RD))
```

### Diagnostic: Weight Trajectories and Effective Sample Size {#sec-tte-diagnostics}

Monitoring how IPCW weights behave over time is essential for any survival analysis using inverse weighting.

```{r}
#| label: fig-weight-diagnostics
#| fig-cap: "IPCW weight trajectories and effective sample size over time"
#| code-fold: false

# Demonstrate weight instability with a simple IPCW example
# Fit a censoring model
library(survival)

# Create indicators for censoring (event == 0 means administratively censored,
# but we also treat competing events as a form of censoring for this exercise)
dat$censored <- as.integer(dat$event == 0 | dat$event == 2)

# Time grid for diagnostics
time_grid <- seq(0.5, 4.5, by = 0.5)

# At each time point, compute weights for those still at risk
weight_summary <- data.frame(
  time = time_grid,
  n_at_risk = NA,
  mean_weight = NA,
  max_weight = NA,
  ess = NA
)

# Simple time-varying censoring model for illustration
cens_model <- glm(censored ~ age + comorbidity + A,
                  data = dat, family = binomial)
p_uncens <- 1 - predict(cens_model, type = "response")

for (i in seq_along(time_grid)) {
  t_k <- time_grid[i]
  at_risk <- dat$time >= t_k
  n_risk <- sum(at_risk)

  # Simplified cumulative weight (illustrative)
  # In practice, these would be computed from a discrete-time censoring model
  w <- (1 / p_uncens[at_risk])^(i)  # cumulative product approximation
  w <- w / mean(w)  # stabilize

  weight_summary$n_at_risk[i] <- n_risk
  weight_summary$mean_weight[i] <- mean(w)
  weight_summary$max_weight[i] <- max(w)
  weight_summary$ess[i] <- sum(w)^2 / sum(w^2)
}

par(mfrow = c(1, 2))

# Plot 1: Max weight over time
plot(weight_summary$time, weight_summary$max_weight, type = "b",
     pch = 19, col = "darkred", lwd = 2,
     xlab = "Time (years)", ylab = "Maximum weight",
     main = "Weight instability over time")
abline(h = 10, lty = 2, col = "gray50")
text(1, 10.5, "Concern threshold", col = "gray50", cex = 0.8)

# Plot 2: Effective sample size over time
plot(weight_summary$time, weight_summary$ess, type = "b",
     pch = 19, col = "steelblue", lwd = 2,
     xlab = "Time (years)", ylab = "Effective sample size",
     main = "ESS decreases with follow-up")
abline(h = 50, lty = 2, col = "gray50")
text(3, 55, "Minimum ESS threshold", col = "gray50", cex = 0.8)

par(mfrow = c(1, 1))
```

## Practical Diagnostics for Survival TMLE {#sec-surv-diagnostics}

### Censoring Model Calibration

Assess whether your censoring model is well-specified by comparing predicted vs. observed censoring rates across risk strata:
```{r}
#| label: cens-calibration
#| code-fold: false

# Calibration of the censoring model
pred_cens <- predict(cens_model, type = "response")
risk_groups <- cut(pred_cens, breaks = quantile(pred_cens, probs = c(0, 0.25, 0.5, 0.75, 1)),
                   include.lowest = TRUE)

cal_table <- aggregate(
  cbind(predicted = pred_cens, observed = dat$censored),
  by = list(risk_group = risk_groups),
  FUN = mean
)
print(cal_table)
```

### Weight Trajectory Plots

For individual-level diagnostics, plot the weight trajectory for a random sample of individuals:

```{r}
#| label: weight-trajectories-individual
#| eval: false
#| code-fold: false

# In a full implementation, you would extract individual-level weights
# from the survtmle or ltmle output and plot them:
#
# plot(time_points, weights_individual[i, ], type = "l",
#      xlab = "Time", ylab = "IPCW weight",
#      main = "Individual weight trajectories")
#
# Key things to look for:
# - Extreme weights (> 10-20x) suggest positivity violations
# - Rapidly diverging weights suggest model misspecification
# - Weights that "explode" at a specific time point suggest
#   a covariate stratum with very high censoring probability
```

### Effective Sample Size Over Time

The effective sample size (ESS) is the most important summary diagnostic for weighted survival analyses:

$$\text{ESS}(t) = \frac{\left(\sum_{i: T_i \geq t} w_i(t)\right)^2}{\sum_{i: T_i \geq t} w_i(t)^2}$$

Rules of thumb:

- ESS < 50 at your target time point: results are unreliable
- ESS drops below 10% of nominal sample size: consider shortening follow-up
- ESS plot has a sudden drop: investigate what changed at that time point (e.g., a policy change causing differential censoring)

## Check Your Understanding

::: {.callout-caution}
## Questions

1. **Estimand choice**: A colleague proposes using a Cox proportional hazards model to estimate the causal effect of a new drug on overall survival in an observational study. They argue that adjusting for confounders in the Cox model produces a valid causal estimate. What are two problems with this approach, and what estimand would you recommend instead?

2. **Competing risks strategy**: In a study of time to kidney transplant, death before transplant is a competing event. A reviewer asks you to justify your choice of estimand. Under what scientific question would you (a) treat death as censoring with IPCW, and (b) use a cumulative incidence approach?

3. **Weight diagnostics**: Your IPCW-based survival TMLE shows that the effective sample size drops from 400 to 25 between year 1 and year 3 of follow-up. What does this indicate, and what are two strategies to address it?
:::

::: {.callout-tip collapse="true"}
## Answers

1. Two problems with the causal HR from a Cox model: (a) Non-collapsibility means the adjusted HR estimates a conditional (within-stratum) quantity that differs from the marginal causal effect, even under correct adjustment. (b) Built-in selection bias at later time points means the hazard comparison is among differentially selected survivors, not the original population. Recommended estimand: the marginal risk difference or RMST difference at a pre-specified time point, estimated using TMLE or IPTW.

2. (a) Treating death as censoring with IPCW answers: "What would the transplant rate be in a hypothetical world where patients could not die before transplant?" This is useful for understanding the biological mechanism of the treatment on transplant. (b) The cumulative incidence approach answers: "What is the actual probability of receiving a transplant by time *t*, given that some patients will die first?" This is useful for clinical planning and resource allocation.

3. The ESS drop indicates severe positivity violations or model misspecification in the censoring model at later time points. The cumulative product of censoring weights is concentrating mass on a few individuals. Strategies: (a) Shorten the analysis time horizon to a point where ESS remains adequate (e.g., report 2-year rather than 3-year results). (b) Truncate extreme weights at a threshold (e.g., the 99th percentile) and report this as a sensitivity analysis, acknowledging the bias-variance tradeoff.
:::

## Sources and Further Reading

- @hernan2016target --- Target trial emulation framework and why estimand choice matters
- @vanderLaan2011targeted --- Targeted learning for causal inference including survival outcomes
- @zheng2011crossvalidated --- Cross-validated TMLE for optimal bias-variance tradeoff
- @petersen2014causal --- Integrating causal modeling and statistical estimation
- @dang2023causal --- Causal roadmap for high-quality real-world evidence
