# Time-to-Event, Censoring, and Competing Risks with Targeted Learning {#sec-tte-competing}

*Estimated reading time: 35 minutes*

::: {.callout-tip}
## Learning Objectives

After completing this chapter, you will be able to:

1. Distinguish risk at time *t*, restricted mean survival time (RMST), and hazard ratios as causal estimands, and articulate why hazard ratios have problematic causal interpretations
2. Apply the Causal Roadmap to move from a clinical question to a survival estimand, identification assumptions, and estimator choice
3. Map ICH E9(R1) intercurrent event strategies (treatment policy, composite, hypothetical, while-on-treatment, principal stratum) to censoring and handling approaches in survival TMLE
4. Implement IPCW for informative censoring and diagnose weight instability at long follow-up
5. Compare cause-specific and subdistribution approaches to competing risks, and understand how continuous-time TMLE handles competing events
6. Apply survival TMLE to estimate a marginal risk difference at time *t* with competing risks, using the ART cohort data
:::

The longitudinal methods from the preceding chapters extend naturally to time-to-event outcomes. The main new considerations are: choosing an estimand that maps cleanly to a causal question, handling informative censoring over time, and accounting for competing risks. This chapter uses the Causal Roadmap framework [@petersen2014causal; @dang2023causal] to connect each analytic decision back to the scientific question being asked.


## Why Estimand Choice Matters for Survival Data {#sec-tte-estimands}

In survival analysis, the default approach has been to fit a Cox proportional hazards model and report a hazard ratio. The Causal Roadmap asks us to slow down and separate three steps: (1) define the causal question and estimand, (2) state the identification assumptions, and (3) choose an estimator suited to the data structure [@chen2023beyond]. This separation prevents the common mistake of letting the statistical model dictate the causal question.

### Risk at Time *t* (Cumulative Incidence)

The counterfactual risk (cumulative incidence) at a pre-specified time point *t*:

$$\Psi = P(T^{a=1} \leq t) - P(T^{a=0} \leq t)$$

This is a marginal risk difference with a clear causal interpretation: "If everyone in the population received treatment versus control, how much would the probability of the event by time *t* differ?" It is compatible with regulator recommendations for trial estimands and does not require proportional hazards.

### Restricted Mean Survival Time (RMST)

The RMST up to time $\tau$ is the area under the survival curve from 0 to $\tau$:

$$\text{RMST}(\tau) = \int_0^\tau S(t) \, dt$$

The causal contrast $\text{RMST}^{a=1}(\tau) - \text{RMST}^{a=0}(\tau)$ summarizes the entire survival curve up to $\tau$. It has a natural clinical interpretation as the difference in expected event-free time over a defined horizon, and does not require proportional hazards [@stensrud2019limitations].

### The Hazard Ratio Problem

::: {.callout-warning}
## Hazard Ratios Are Not Straightforward Causal Parameters

The hazard ratio has two fundamental problems as a causal quantity [@stensrud2019limitations; @aalen2015cox; @martinussen2018hazardratios]:

1. **Built-in selection bias**: The hazard at time *t* is defined only among survivors to *t*. Even in a perfectly executed randomized trial, if treatment is effective, high-risk individuals are depleted from the placebo arm faster than from the treatment arm. Comparing hazards at later times therefore compares groups with systematically different baseline risk profiles. This means the population-level hazard ratio conflates the treatment effect with differential selection of survivors.

2. **Non-collapsibility**: The conditional HR (within strata) does not equal the marginal HR, even without confounding. This means that covariate adjustment in a Cox model not only affects precision and bias, but actually changes the estimand being targeted. Researchers who adjust for covariates to handle informative censoring may inadvertently redefine their target quantity.

3. **Dependence on censoring distribution**: Under non-proportional hazards, the estimand targeted by Cox regression is defined implicitly through an estimating equation that depends on the censoring distribution. Two trials in identical populations with different accrual patterns or follow-up durations can produce different "true" hazard ratios, even if the treatment effect is the same.

For causal inference, prefer risk differences, risk ratios at fixed time points, or RMST differences [@hernan2016target; @chen2023beyond; @rufibach2019estimands].
:::


## The Causal Roadmap for Survival Data {#sec-causal-roadmap}

The Causal Roadmap [@petersen2014causal; @dang2023causal] provides a systematic framework for integrating causal inference and statistical estimation. Applied to a time-to-event RCT or observational study, the steps are:

1. **Causal model and question**: Define the counterfactual data structure, including potential outcomes under each treatment. What is the ideal experiment?
2. **Causal estimand**: Express the treatment effect as a function of counterfactual outcomes (e.g., risk ratio at time *t*).
3. **Observed data and identification**: Specify how the observed data relates to the ideal experiment, including what gets censored and why.
4. **Identification assumptions**: State consistency, positivity (for both treatment and censoring), and conditional exchangeability. Different analyses require different assumptions of different strengths.
5. **Statistical estimand**: The observed-data parameter that equals the causal estimand under the identification assumptions.
6. **Estimator**: Choose an estimator with desirable properties (efficiency, robustness, doubly robust inference).
7. **Sensitivity analysis**: Assess the impact of potential assumption violations.

::: {.callout-important}
## Different Questions, Different Assumptions

A single dataset can support multiple causal analyses, each targeting a different estimand and requiring different identification assumptions. For example, using the SUSTAIN-6 cardiovascular outcome trial as an illustration [@chen2023concrete; @marso2016sustain6]:

- **Intent-to-treat**: What is the effect of being *assigned* to semaglutide vs. placebo? Only requires that loss-to-follow-up is non-informative.
- **On-treatment (right-censoring)**: What would the effect be if all patients *adhered* to their assigned treatment? Requires that treatment discontinuation for any reason is conditionally independent of MACE given baseline covariates---a much stronger assumption.
- **On-treatment (competing risks)**: What is the effect on MACE occurring *before* treatment discontinuation due to adverse events? Requires only that discontinuation for reasons *other* than adverse events is conditionally independent of MACE---arguably more plausible than the right-censoring assumption.

These three analyses use the same data but answer different clinical questions, have different estimands, and require different assumptions. The Causal Roadmap forces explicit articulation of each.
:::


## ICH E9(R1) Intercurrent Events and Censoring Strategies {#sec-ich-e9r1}

The ICH E9(R1) addendum defines five strategies for intercurrent events. Each maps to a specific approach for how censoring or outcome handling is implemented in survival TMLE [@rufibach2019estimands].

| ICH E9(R1) Strategy | Description | Survival TMLE Implementation |
|-------------------:|-------------|---------------------|
| **Treatment policy** | Outcome observed regardless of intercurrent event | No censoring at intercurrent event; analyze all observed outcomes. This is the survival analogue of "intent-to-treat" |
| **Composite** | Intercurrent event becomes part of the outcome | Redefine the event indicator to include the intercurrent event (e.g., PFS = min(progression, death)) |
| **Hypothetical** | Outcome in a world where the intercurrent event could not occur | Censor at intercurrent event and use IPCW to adjust for informative censoring. Alternatively, estimate via continuous-time TMLE with the censoring mechanism modeled |
| **While on treatment** | Outcome while receiving assigned treatment | Censor at treatment discontinuation; model the censoring mechanism. Estimand differs from hypothetical: targets risk *while on treatment* rather than risk *in absence of intercurrent event* |
| **Principal stratum** | Effect among those who would not experience the intercurrent event under either treatment | Requires principal stratification methods; not handled by standard survival TMLE |

: ICH E9(R1) strategies mapped to survival TMLE approaches {#tbl-ich-e9r1}

::: {.callout-tip}
## Censoring Is Part of the Estimator, Not the Estimand

A key distinction from the estimand framework: the decision to censor at an intercurrent event is an *estimation strategy*, not a definition of the quantity of interest. The estimand comes first (what would we like to know?), and the censoring approach follows (how do we estimate it from data where the intercurrent event occurs?). Both the hypothetical and while-on-treatment strategies may censor at the same event, but the *estimand* they target differs, and this difference manifests in the choice of summary measure [@rufibach2019estimands].
:::


## Informative Censoring and IPCW {#sec-ipcw}

When censoring depends on covariates that also predict the outcome, naive Kaplan-Meier or Cox estimates are biased. Inverse probability of censoring weighting (IPCW) applies the same logic from the IPTW chapter to the censoring mechanism: upweight individuals who remain uncensored but have covariate profiles similar to those who were censored.

At each time point $t_k$, the cumulative IPCW weight is:

$$W_C(t_k) = \prod_{j=1}^{k} \frac{1}{P(C > t_j \mid C > t_{j-1}, \bar{A}_{j-1}, \bar{L}_{j-1})}$$

::: {.callout-important}
## Cumulative Weight Instability

Because IPCW weights are a *product* over time points, instability compounds exponentially. With per-period uncensored probability of 0.95, after 20 time points the weight is $1/0.95^{20} \approx 2.65$; at 0.80 per period, weights explode to $1/0.80^{20} \approx 87$. Positivity violations accumulate, effective sample size shrinks, and a few individuals can dominate the estimate.

**Remedies**:

- Truncate weights (with sensitivity analysis on the truncation threshold)
- Restrict follow-up to a horizon with stable weights
- Use stabilized weights: $SW_C(t_k) = \prod_{j=1}^k \frac{P(C > t_j \mid C > t_{j-1})}{P(C > t_j \mid C > t_{j-1}, \bar{A}_{j-1}, \bar{L}_{j-1})}$
- Use TMLE, whose targeting step provides additional stabilization relative to pure IPCW
:::


## Competing Risks {#sec-competing-risks}

A competing risk is an event that prevents the outcome of interest from being observed. When studying time to cardiovascular death, non-cardiovascular death is a competing event. Competing risks have been consistently under-addressed in time-to-event RCTs [@austin2017competing].

### Two Classical Approaches

**Cause-specific approach**: Censor at competing events. Estimates the cause-specific hazard---the instantaneous rate of the event of interest among those who have not yet experienced *any* event. However, 1 minus the cause-specific Kaplan-Meier estimate does *not* have a probability interpretation when competing events exist, because it ignores that some censored individuals experienced the competing event rather than remaining at risk.

**Subdistribution (Fine-Gray) approach**: Keep individuals in the risk set after competing events. Directly models the cumulative incidence function (CIF): the actual probability of the event by time *t*, accounting for the fact that competing events remove individuals from eligibility. The CIF is the quantity patients and clinicians typically want for planning.

### Connecting Competing Risks to the Causal Roadmap

The choice between approaches is an *estimand* choice, not a statistical one:

- **"What would the risk of MACE be if patients could not die from other causes?"** --- This is a *hypothetical* estimand. Estimate via cause-specific IPCW, treating non-CV death as informative censoring. Requires the (untestable) assumption that the competing event mechanism is conditionally independent of the outcome.

- **"What is the actual probability of MACE by time *t*, given that patients can also die from other causes?"** --- This is the *cumulative incidence* (CIF) estimand. Estimate via Aalen-Johansen or subdistribution methods. This is the clinically relevant quantity for prognostic planning.

- **"What is the effect on MACE occurring before treatment discontinuation due to adverse events?"** --- This targets a competing-risk cumulative incidence where treatment discontinuation is the competing event. This is the estimand used in the SUSTAIN-6 competing risk analysis [@chen2023concrete].

::: {.callout-tip}
## When the Competing Event Tells the Story

In the SUSTAIN-6 trial analysis, semaglutide reduced MACE risk but *increased* treatment discontinuation due to adverse events. The competing risk analysis revealed that the increased discontinuation risk was larger in magnitude than the decreased MACE risk. This trade-off is fundamentally an estimand choice issue: the competing risk analysis answers a different clinical question than the right-censoring analysis, and the difference in results reflects this, rather than a modeling artifact [@chen2023concrete].
:::


## Continuous-Time vs. Discrete-Time Survival TMLE {#sec-ct-vs-dt}

Standard survival TMLE implementations (e.g., `survtmle`) discretize continuous event times into bins and apply longitudinal TMLE within each bin. This works well when the time grid is coarse relative to the event rate, but can introduce problems when applied to continuous-time survival data [@chen2023concrete; @rytgaard2021continuous]:

- **Bias from discretization**: Binning changes the causal question (from continuous to discrete time) and can shift point estimates toward the null.
- **Variance instability**: Fine bins with few events per bin lead to unstable influence-curve-based variance estimates.
- **Grid sensitivity**: Results can depend on the choice of bin width.

A continuous-time TMLE, implemented in the `concrete` R package [@chen2023concrete; @rytgaard2023estimation; @rytgaard2022tmle], avoids these issues by:

1. Representing the observed data through counting processes and intensities
2. Estimating conditional cause-specific hazards using Cox regression or penalized Cox (coxnet) via Super Learner
3. Applying a targeting step that solves the efficient influence curve equation in continuous time

In the SUSTAIN-6 application, continuous-time TMLE provided 7--29% efficiency gains over unadjusted Kaplan-Meier and Aalen-Johansen estimators, while discrete-time TMLE with non-optimized bin widths sometimes performed worse than the unadjusted estimators [@chen2023concrete].


## Worked Example: Survival TMLE with the ART Cohort {#sec-tte-example}

We use the HIV antiretroviral therapy (ART) cohort from the `ipw` package---the running case study for this handbook. Here we frame a survival analysis: does initiating HAART reduce the risk of death by 2,000 days (approximately 5.5 years), accounting for confounding by indication and the competing event of dropout?

### Setup and Data Preparation

```{r}
#| label: setup-tte
#| code-fold: true
#| message: false
#| warning: false

library(survival)
library(ipw)
library(dplyr)
library(ggplot2)

# Load the ART cohort
data(haartdat)

# Create a point-treatment dataset: first record per patient
haart_surv <- haartdat |>
  group_by(patient) |>
  summarise(
    sex = first(sex),
    age = first(age),
    cd4_sqrt_bl = first(cd4.sqrt),
    A = max(haartind),           # ever initiated HAART
    time = max(fuptime),         # total follow-up (days)
    died = max(event),           # death indicator
    dropout = max(dropout),      # dropout indicator
    .groups = "drop"
  ) |>
  mutate(
    # Event type: 0 = censored (administrative), 1 = death, 2 = dropout
    event_type = case_when(
      died == 1 ~ 1L,
      dropout == 1 ~ 2L,
      TRUE ~ 0L
    ),
    # For survival analysis, use death as primary event
    time_years = time / 365.25
  )

cat("Sample size:", nrow(haart_surv), "\n")
cat("Event distribution:\n")
table(haart_surv$event_type,
      dnn = "Event") |> print()
cat("\nHAART initiation:\n")
table(haart_surv$A,
      dnn = "HAART") |> print()
```

### Naive Kaplan-Meier

The unadjusted Kaplan-Meier ignores confounding by indication (sicker patients, with lower CD4, are more likely to receive HAART and more likely to die) and treats dropout as non-informative censoring.

```{r}
#| label: fig-km-naive
#| fig-cap: "Naive Kaplan-Meier survival curves by HAART initiation. The crossing or convergence of curves reflects confounding by indication, not the causal effect of treatment."
#| code-fold: show

km_fit <- survfit(Surv(time, died) ~ A, data = haart_surv)

plot(km_fit, col = c("firebrick", "steelblue"), lwd = 2,
     xlab = "Time (days)", ylab = "Survival probability",
     main = "Naive Kaplan-Meier by HAART Initiation")
legend("bottomleft", c("No HAART (A=0)", "HAART (A=1)"),
       col = c("firebrick", "steelblue"), lwd = 2, bty = "n")
```

This naive comparison is misleading: confounding by indication (low CD4 drives both treatment initiation and mortality) biases the comparison.

### Cumulative Incidence with Competing Risks (Aalen-Johansen)

Dropout can be treated as a competing event. The Aalen-Johansen estimator provides cause-specific cumulative incidence functions, but does not adjust for confounding.

```{r}
#| label: fig-aalen-johansen
#| fig-cap: "Aalen-Johansen cumulative incidence for death and dropout by HAART status. Accounts for competing risks but not confounding."
#| code-fold: show

# Multi-state event factor
haart_surv$event_ms <- factor(haart_surv$event_type, levels = 0:2,
                              labels = c("censored", "death", "dropout"))

aj_fit <- survfit(Surv(time, event_ms) ~ A, data = haart_surv)
plot(aj_fit, col = c("firebrick", "steelblue"), lwd = 2,
     xlab = "Time (days)", ylab = "Cumulative incidence",
     main = "Aalen-Johansen Cumulative Incidence by HAART Status")
```

### Adjusted Survival via IPTW

To address confounding, we compute stabilized IPTW weights using baseline covariates and apply them to the Kaplan-Meier estimator.

```{r}
#| label: fig-iptw-survival
#| fig-cap: "IPTW-weighted Kaplan-Meier curves adjusting for confounding by age, sex, and baseline CD4. Compare to the naive KM above."
#| code-fold: show

# Propensity score model
ps_model <- glm(A ~ age + sex + cd4_sqrt_bl,
                data = haart_surv, family = binomial)
haart_surv$ps <- predict(ps_model, type = "response")

# Stabilized weights
p_A <- mean(haart_surv$A)
haart_surv$sw <- ifelse(haart_surv$A == 1,
                        p_A / haart_surv$ps,
                        (1 - p_A) / (1 - haart_surv$ps))

# Truncate extreme weights at 1st and 99th percentiles
q_bounds <- quantile(haart_surv$sw, c(0.01, 0.99))
haart_surv$sw_trunc <- pmin(pmax(haart_surv$sw, q_bounds[1]), q_bounds[2])

cat("Weight summary (truncated):\n")
summary(haart_surv$sw_trunc)
cat("\nEffective sample size:",
    round(sum(haart_surv$sw_trunc)^2 / sum(haart_surv$sw_trunc^2)), "\n")

# Weighted KM
km_iptw <- survfit(Surv(time, died) ~ A,
                   data = haart_surv,
                   weights = sw_trunc)

plot(km_iptw, col = c("firebrick", "steelblue"), lwd = 2,
     xlab = "Time (days)", ylab = "Survival probability",
     main = "IPTW-Weighted Kaplan-Meier (Confounding Adjusted)")
legend("bottomleft", c("No HAART (A=0)", "HAART (A=1)"),
       col = c("firebrick", "steelblue"), lwd = 2, bty = "n")
```

### TMLE for Risk Difference at Time *t*

For a doubly robust estimate, we use `survtmle` to target the risk difference at a specific time point. This discretizes time but provides both outcome-model and censoring-model robustness.

```{r}
#| label: tmle-surv
#| code-fold: show
#| message: false
#| warning: false

library(survtmle)

# survtmle requires positive event times; filter patients with time = 0
haart_pos <- haart_surv |> filter(time > 0)

# Target: risk of death by day 2000 (~5.5 years)
t_target <- 2000

fit_tmle <- survtmle(
  ftime      = haart_pos$time,
  ftype      = ifelse(haart_pos$died == 1, 1, 0),
  trt        = haart_pos$A,
  adjustVars = haart_pos[, c("age", "sex", "cd4_sqrt_bl")],
  t0         = t_target,
  SL.ftime   = c("SL.glm", "SL.mean"),
  SL.ctime   = c("SL.glm", "SL.mean"),
  SL.trt     = c("SL.glm", "SL.mean"),
  method     = "hazard"
)

# Extract marginal risks and risk difference
risk_A0 <- fit_tmle$est[1]
risk_A1 <- fit_tmle$est[2]
RD <- risk_A1 - risk_A0

cat(sprintf("Risk under A=0 (no HAART): %.4f\n", risk_A0))
cat(sprintf("Risk under A=1 (HAART):    %.4f\n", risk_A1))
cat(sprintf("Risk difference (A=1 - A=0): %.4f\n", RD))
cat(sprintf("Risk ratio: %.3f\n", risk_A1 / risk_A0))
```

### Continuous-Time TMLE with `concrete`

For continuous-time survival data, the `concrete` R package avoids discretization entirely. It estimates cause-specific hazards using Cox/penalized Cox models via Super Learner, then applies the continuous-time TMLE targeting step. This approach provided consistent precision gains of 7--29% over unadjusted estimators in the SUSTAIN-6 application, while discrete-time TMLE with naive bin choices sometimes performed worse than unadjusted methods [@chen2023concrete].

```{r}
#| label: concrete-tmle
#| code-fold: show
#| message: false
#| warning: false

library(concrete)

# concrete requires positive event times; filter patients with time = 0
haart_pos <- haart_surv |> filter(time > 0)

# Format data for concrete
concrete_data <- data.frame(
  T = haart_pos$time,
  Delta = haart_pos$event_type,  # 0 = censored, 1 = death, 2 = dropout
  A = haart_pos$A,
  age = haart_pos$age,
  sex = haart_pos$sex,
  cd4_sqrt_bl = haart_pos$cd4_sqrt_bl
)

# makeITT() creates the intent-to-treat intervention for a binary treatment
intervention <- makeITT()

# Format arguments and fit nuisance parameters via Super Learner
concrete_est <- formatArguments(
  DataTable = concrete_data,
  EventTime = "T",
  EventType = "Delta",
  Treatment = "A",
  Covariates = c("age", "sex", "cd4_sqrt_bl"),
  TargetTime = 2000,
  TargetEvent = 1,         # death (primary event of interest)
  Intervention = intervention,
  MaxUpdateIter = 250,
  CVArg = list(V = 5),
  Verbose = FALSE
)

# Run continuous-time TMLE
concrete_est <- doConcrete(concrete_est)

# Extract results: risk estimates, differences, and CIs
concrete_results <- getOutput(concrete_est)
print(concrete_results)
```

The `concrete` output provides risk estimates under each treatment level at the target time, along with risk differences and risk ratios with influence-curve-based confidence intervals. Compare these to the discrete-time `survtmle` estimates above---both target the same estimand, but `concrete` operates in continuous time.


## Diagnostics for Survival TMLE {#sec-surv-diagnostics}

### Weight Trajectories and Effective Sample Size

The effective sample size (ESS) is the key diagnostic for weighted survival analyses:

$$\text{ESS}(t) = \frac{\left(\sum_{i: T_i \geq t} w_i(t)\right)^2}{\sum_{i: T_i \geq t} w_i(t)^2}$$

```{r}
#| label: fig-weight-diagnostics
#| fig-cap: "Diagnostic plots for IPTW weights: maximum weight and effective sample size over follow-up. Instability at later time points signals positivity concerns."
#| code-fold: show
#| fig-width: 9
#| fig-height: 4

# Compute ESS at multiple time horizons
time_grid <- seq(200, 3000, by = 200)
diag_df <- data.frame(
  time = time_grid,
  n_at_risk = NA_real_,
  max_weight = NA_real_,
  ess = NA_real_
)

for (i in seq_along(time_grid)) {
  tk <- time_grid[i]
  at_risk <- haart_surv$time >= tk
  w <- haart_surv$sw_trunc[at_risk]

  diag_df$n_at_risk[i] <- sum(at_risk)
  diag_df$max_weight[i] <- max(w)
  diag_df$ess[i] <- sum(w)^2 / sum(w^2)
}

par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

plot(diag_df$time, diag_df$max_weight, type = "b",
     pch = 19, col = "darkred", lwd = 2,
     xlab = "Time (days)", ylab = "Maximum weight",
     main = "Weight instability over time")
abline(h = 10, lty = 2, col = "gray50")

plot(diag_df$time, diag_df$ess, type = "b",
     pch = 19, col = "steelblue", lwd = 2,
     xlab = "Time (days)", ylab = "Effective sample size",
     main = "ESS decreases with follow-up")
abline(h = 50, lty = 2, col = "gray50")

par(mfrow = c(1, 1))
```

### Rules of Thumb for ESS

- **ESS < 50** at your target time point: results are unreliable
- **ESS drops below 10%** of nominal sample size: consider shortening follow-up
- **Sudden ESS drop**: investigate what changed at that time point (e.g., a protocol amendment causing differential censoring)

### Censoring Model Calibration

```{r}
#| label: tbl-cens-calibration
#| code-fold: show

# Calibration of a simple censoring model
cens_model <- glm(
  (event_type == 0 | event_type == 2) ~ age + cd4_sqrt_bl + A,
  data = haart_surv, family = binomial
)
pred_cens <- predict(cens_model, type = "response")

risk_groups <- cut(pred_cens,
                   breaks = quantile(pred_cens, probs = seq(0, 1, 0.25)),
                   include.lowest = TRUE)

cal_table <- aggregate(
  cbind(predicted = pred_cens,
        observed = as.integer(haart_surv$event_type == 0 |
                              haart_surv$event_type == 2)),
  by = list(risk_group = risk_groups),
  FUN = mean
)
print(cal_table)
```


## Practical Guidance {#sec-tte-guidance}

::: {.callout-important}
## Choosing Your Survival Estimand

| If your question is... | Target this estimand | Handle intercurrent events by... |
|:---|:---|:---|
| Effect of *assignment* to treatment | Risk difference at time *t* (ITT) | Treatment policy: observe outcomes regardless of discontinuation |
| Effect of *receiving* treatment | Risk difference at time *t* (per-protocol) | Hypothetical: censor at discontinuation, use IPCW |
| Effect on outcome *before* a competing event | Cumulative incidence difference | Competing risk: model both event types jointly |
| Summary effect over the survival curve | RMST difference | Integrate the survival curve to $\tau$ |
:::

::: {.callout-tip}
## Recommendation: When to Use What

- **Small trials, rare events, long follow-up**: Be cautious with IPCW; weight instability may dominate. Consider restricting the time horizon.
- **Continuous-time data from RCTs**: Consider `concrete` for continuous-time TMLE---avoids discretization artifacts and provides reliable efficiency gains over Kaplan-Meier.
- **Observational studies with many confounders**: Use TMLE with Super Learner for both the outcome and censoring models. The doubly robust property protects against misspecification of either model.
- **Competing risks present**: Decide the estimand *first*. If you want the "hypothetical world without competing events," use cause-specific IPCW. If you want the actual cumulative incidence, use Aalen-Johansen or TMLE targeting the CIF.
:::


## Check Your Understanding

::: {.callout-caution}
## Questions

1. **Estimand choice**: A colleague proposes a confounding-adjusted Cox model for a causal effect on survival. Identify two problems with this approach, and recommend an alternative estimand.

2. **Competing risks**: In a study of time to kidney transplant where death is a competing event, when would you (a) treat death as censoring with IPCW, and (b) use a cumulative incidence approach? What identification assumption does (a) require that (b) does not?

3. **Weight diagnostics**: ESS drops from 400 to 25 between year 1 and year 3. What does this indicate, and what are two strategies to address it?

4. **Continuous vs. discrete time**: An analysis discretizes daily event times into 6-month bins for a discrete-time TMLE. What are two potential consequences of this discretization?
:::

::: {.callout-tip collapse="true"}
## Answers

1. (a) Non-collapsibility: the adjusted HR estimates a conditional quantity that differs from the marginal causal effect; adjustment changes the estimand, not just bias and precision. (b) Built-in selection bias: hazard comparisons at later times compare differentially selected survivors, even in an RCT. **Recommendation**: target a marginal risk difference or RMST difference at a pre-specified time point, estimated via TMLE or IPCW.

2. (a) Cause-specific IPCW targets: "What would transplant rates be if patients could not die?" This requires that death is conditionally independent of transplant given measured covariates (a hypothetical estimand). (b) CIF targets: "What is the actual probability of transplant by time *t*, given that death can occur?" This requires only that administrative censoring is non-informative. Strategy (a) requires the much stronger assumption that the competing event mechanism is independent of the outcome conditional on covariates.

3. The ESS drop indicates near-positivity violations in the censoring or treatment model---a few heavily weighted individuals dominate the estimate at later time points. Strategies: (a) Shorten the analysis horizon to where ESS remains adequate. (b) Truncate extreme weights and report truncation thresholds as a sensitivity analysis.

4. (a) Discretization changes the causal question from continuous to discrete time, and wide bins can bias point estimates toward the null by pooling heterogeneous event times. (b) Bins with few events lead to unstable influence-curve-based variance estimates, potentially yielding confidence intervals that are too wide or too narrow. Consider continuous-time TMLE via `concrete` to avoid these artifacts.
:::


## Sources and Further Reading

- @petersen2014causal and @dang2023causal: The Causal Roadmap framework for integrating causal modeling and statistical estimation
- @chen2023beyond: Application of the Causal Roadmap to the LEADER cardiovascular outcome trial, demonstrating TMLE's efficiency and robustness advantages over Cox regression
- @chen2023concrete: Continuous-time survival TMLE applied to SUSTAIN-6, with the `concrete` R package; includes competing risks
- @rytgaard2023estimation and @rytgaard2022tmle: Theory of continuous-time TMLE for survival and competing risks
- @stensrud2019limitations: Accessible overview of why hazard ratios are problematic for causal inference in clinical trials
- @rufibach2019estimands: ICH E9(R1) estimand framework applied to time-to-event endpoints
- @austin2017competing: Review of competing risks handling in RCTs
- @hernan2016target: Target trial emulation framework and why estimand choice matters
