---
title: "TMLE for Mediation Analysis"
subtitle: "Decomposing total effects into direct and indirect pathways"
format: html
params:
  fast: true
---

*Estimated reading time: 45 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
library(dagitty)
library(ggdag)
library(ggplot2)
```

# Why This Matters

In pharmacoepidemiology, we often want to know not just **whether** a drug works, but **how** it works. Does dapagliflozin reduce cardiovascular events because it lowers blood pressure, because it improves glucose control, or through some other pathway? Mediation analysis decomposes a total causal effect into the part that operates through a specified intermediate variable (the indirect effect) and the part that does not (the direct effect). Getting this decomposition right requires careful attention to the estimand, the causal model, and the assumptions that connect the two. This chapter walks through the logic and implementation of TMLE-based mediation using the `crumble` R package.

**What you will learn:**

- How to specify a mediation question as a formal causal estimand and why the choice of estimand matters.
- The assumptions required for identification, and what breaks when they fail.
- How TMLE-based estimation with cross-fitting and flexible learners works for mediation.
- How to use the `crumble` package to estimate natural, randomized interventional, and recanting-twins effects.
- How to interpret and report mediation results responsibly.

::: {.callout-important}
## Sources and scope

This chapter is educational. Causal conclusions depend on identification assumptions (consistency, exchangeability, positivity) and on diagnostic evidence that the data support the target estimand. The worked examples use simulated data with known ground truth; real analyses require subject-matter justification for every assumption.

Key references: Zheng & van der Laan (2012), VanderWeele (2015), Díaz & Hejazi (2020), Robins & Richardson (2010), Williams & Díaz (2024, `crumble`). Full citations appear at the end.
:::


# Clinical Motivation

## The Pharmacoepidemiologic Question

SGLT2 inhibitors (like dapagliflozin) have shown surprising cardiovascular benefits in type 2 diabetes trials. A natural follow-up question:

> **How much of dapagliflozin's effect on reducing MACE operates through HbA1c reduction (the glycemic pathway), and how much operates through other mechanisms (blood pressure, weight loss, direct cardiac effects)?**

This matters for labeling, for understanding whether other glucose-lowering drugs should have similar benefits, and for guiding future drug development.

::: {.callout-note}
## The Setup

- **Treatment (A):** Dapagliflozin vs. sulfonylurea
- **Mediator (M):** 6-month HbA1c change (glycemic improvement)
- **Outcome (Y):** 1-year MACE (binary)
- **Baseline confounders (W):** Age, BMI, baseline HbA1c, prior CVD, eGFR, statin use
- **Question:** How much of the $A \to Y$ effect goes through $A \to M \to Y$?
:::


# Conceptual Primer: Mediator, Confounder, Moderator, Collider

Before diving into estimation, it helps to be precise about what a mediator is, and what it is not.

```{r}
#| label: fig-primer-dags
#| fig-cap: "Four fundamental variable roles in causal diagrams"
#| fig-width: 10
#| fig-height: 3

dag_confounder <- dagify(A ~ W, Y ~ W,
  coords = list(x = c(W = 1, A = 0, Y = 2), y = c(W = 1, A = 0, Y = 0))
)
dag_moderator <- dagify(Y ~ A + V,
  coords = list(x = c(A = 0, V = 1, Y = 2), y = c(A = 0, V = 1, Y = 0))
)
dag_mediator <- dagify(M ~ A, Y ~ A + M,
  coords = list(x = c(A = 0, M = 1, Y = 2), y = c(A = 0, M = 0.5, Y = 0))
)
dag_collider <- dagify(C ~ A + Y,
  coords = list(x = c(A = 0, C = 1, Y = 2), y = c(A = 0, C = 0.5, Y = 0))
)

p1 <- ggdag(dag_confounder) + theme_dag() + labs(title = "Confounder (W)")
p2 <- ggdag(dag_moderator) + theme_dag() + labs(title = "Moderator (V)")
p3 <- ggdag(dag_mediator) + theme_dag() + labs(title = "Mediator (M)")
p4 <- ggdag(dag_collider) + theme_dag() + labs(title = "Collider (C)")

library(patchwork)
p1 + p2 + p3 + p4 + plot_layout(nrow = 1)
```

- A **confounder** is a common cause of treatment and outcome. Adjusting for it removes bias.
- A **moderator** modifies the size of the treatment effect (interaction) but is not on the causal pathway.
- A **mediator** is on the causal pathway from treatment to outcome. It is caused by treatment and in turn causes the outcome.
- A **collider** is caused by two variables. Conditioning on a collider opens a spurious path.

**Temporality is essential.** A variable can only mediate the treatment effect if it occurs *after* treatment and *before* the outcome. Adjusting for a post-outcome variable or a cross-sectionally measured variable as if it were a mediator is a common error that can introduce bias rather than remove it.

::: {.callout-warning}
## The intermediate confounder problem

A variable $Z$ that is affected by treatment and confounds the mediator-outcome relationship creates the hardest problem in mediation analysis. If $Z$ exists and you ignore it, your decomposition is biased. We return to this below when choosing the estimand.

```{r}
#| label: fig-intermediate-confounder
#| fig-cap: "Intermediate confounder Z: affected by A, confounds M → Y"
#| fig-width: 5
#| fig-height: 3
dag_ic <- dagify(
  Z ~ A, M ~ A + Z, Y ~ A + M + Z,
  coords = list(
    x = c(A = 0, Z = 1, M = 2, Y = 3),
    y = c(A = 0, Z = 0.5, M = 0, Y = 0)
  )
)
ggdag(dag_ic) + theme_dag()
```
:::


# Mediation Analysis: The Causal Roadmap

## Step 1: Define the Estimand

Mediation decomposes a total effect into pieces. But there are several competing definitions, and the choice matters because they require different assumptions.

**Natural direct and indirect effects (NDE/NIE).** The most common definitions. In plain language, the NDE captures the part of the treatment effect that bypasses the mediator entirely, while the NIE captures the part that works through the mediator. Together they add up to the total effect.

$$\text{NDE} = E[Y(1, M(0))] - E[Y(0, M(0))]$$
$$\text{NIE} = E[Y(1, M(1))] - E[Y(1, M(0))]$$

The NDE asks: what is the effect of treatment if we could hold the mediator at the value it would have taken under control? The NIE asks: what is the effect of shifting the mediator as treatment would, while keeping treatment on? These decompose the total effect: $\text{TE} = \text{NDE} + \text{NIE}$.

The trouble is that $Y(1, M(0))$ is a **cross-world** quantity that combines treatment from one "world" ($A=1$) with a mediator value from another ($A=0$). This requires a strong, untestable assumption (no treatment-induced mediator-outcome confounding).

**Randomized interventional effects (RIE).** Instead of setting $M$ to a specific counterfactual value, we draw $M$ from its distribution under control. Think of this as asking: "What would the outcome look like if we kept the treatment on, but randomly reassigned each patient's mediator value from the pool of mediator values seen under the control condition?" This avoids the cross-world problem by using distributions rather than fixed individual values.

$$\text{IDE} = E[Y(1, G_0)] - E[Y(0, G_0)]$$
$$\text{IIE} = E[Y(1, G_1)] - E[Y(1, G_0)]$$

where $G_a$ denotes a random draw from $P(M \mid A = a, W)$. These do not require cross-world quantities and are identified under weaker conditions.

**Recanting twins (RT).** When there is an intermediate confounder $Z$ (affected by treatment, confounding the mediator-outcome path), even randomized interventional effects need further care. The recanting-twins approach (Robins & Richardson, 2010) handles this by decomposing the effect into path-specific components using two independent "copies" of the treatment.

::: {.callout-tip}
## Which estimand should you choose?

- If you are confident there is **no intermediate confounder** (no $Z$), natural effects (`"N"`) are the most interpretable and standard choice.
- If intermediate confounding is plausible or you want **weaker assumptions**, randomized interventional effects (`"RI"`) are a good default.
- If you have a **measured intermediate confounder** $Z$ and want to decompose the effect further, recanting twins (`"RT"`) handles this explicitly.

For this tutorial, we will estimate all three to compare.
:::


## Step 2: Causal Model and Time Ordering

The mediation DAG for our example:

```{r}
#| label: fig-mediation-dag-simple
#| fig-cap: "Simple mediation DAG: W confounds all relationships, A affects Y directly and through M"
#| fig-width: 6
#| fig-height: 4
dag_simple <- dagify(
  A ~ W, M ~ A + W, Y ~ A + M + W,
  exposure = "A", outcome = "Y",
  coords = list(
    x = c(W = 1, A = 0, M = 1, Y = 2),
    y = c(W = 1, A = 0, M = 0, Y = 0)
  )
)
ggdag(dag_simple) + theme_dag()
```

The time ordering must be: $W$ (baseline) $\to$ $A$ (treatment assignment) $\to$ $M$ (6-month HbA1c) $\to$ $Y$ (1-year MACE). This ordering is not negotiable. If $M$ is measured after $Y$, or if $A$ and $M$ are measured simultaneously, the decomposition is not credible.

When we add an intermediate confounder $Z$ (e.g., blood pressure change caused by treatment, affecting both mediator trajectory and outcome):

```{r}
#| label: fig-mediation-dag-z
#| fig-cap: "Mediation DAG with intermediate confounder Z (affected by A, confounds M → Y)"
#| fig-width: 7
#| fig-height: 4
dag_with_z <- dagify(
  A ~ W, Z ~ A + W, M ~ A + Z + W, Y ~ A + M + Z + W,
  exposure = "A", outcome = "Y",
  coords = list(
    x = c(W = 1.5, A = 0, Z = 1, M = 2, Y = 3),
    y = c(W = 1, A = 0, Z = 0, M = 0, Y = 0)
  )
)
ggdag(dag_with_z) + theme_dag()
```

Now $Z$ is post-treatment and confounds $M \to Y$. Natural effects are not identified without the cross-world assumption. Recanting twins and randomized interventional effects can handle this.


## Step 3: Identification Assumptions

To identify mediation effects, we need the standard causal assumptions from Chapter 1.2 (consistency, exchangeability, positivity) **plus** additional ones specific to the mediator pathway:

1. **Consistency:** as in Chapter 1.2.
2. **No unmeasured treatment-outcome confounding:** $Y(a, m) \perp\!\!\!\perp A \mid W$
3. **No unmeasured treatment-mediator confounding:** $M(a) \perp\!\!\!\perp A \mid W$
4. **No unmeasured mediator-outcome confounding:** $Y(a, m) \perp\!\!\!\perp M \mid A, W$ (or $A, W, Z$ if intermediate confounders exist)
5. **Positivity:** For all strata of $W$, both treatment values have positive probability, and all mediator values observed under one treatment could also occur under the other.

For **natural effects** only, add:

6. **Cross-world independence:** No confounder of $M \to Y$ is affected by $A$. This is the assumption that fails when an intermediate confounder $Z$ exists.

For **randomized interventional** and **recanting twins** effects, assumption 6 is replaced by the weaker requirement that intermediate confounders are measured and adjusted for.

::: {.callout-note collapse="true"}
## Assumption checklist (expand)

Before running a mediation analysis, verify:

- [ ] Treatment precedes mediator precedes outcome in time.
- [ ] Baseline confounders $W$ are measured before treatment.
- [ ] No variable in $W$ is itself affected by treatment (otherwise it belongs in $Z$ or $M$).
- [ ] Confounders of $A \to Y$, $A \to M$, and $M \to Y$ are identified and measured.
- [ ] If using natural effects: no treatment-induced mediator-outcome confounder exists (or if one does, switch to RI/RT).
- [ ] If using RI or RT effects: intermediate confounders are measured and specified in the model.
- [ ] Positivity: treatment groups overlap in confounder space; mediator distributions overlap across treatment arms.
- [ ] The mediator is not a proxy for treatment (if $M$ is nearly deterministic given $A$, the indirect effect is poorly identified).

None of these can be fully verified from data. Subject-matter knowledge is required.
:::


## Step 4: Estimation Strategy

TMLE-based mediation estimation (as implemented in `crumble`) works in several stages:

1. **Nuisance estimation with cross-fitting.** Fit flexible models for the outcome ($E[Y \mid A, M, W]$), the mediator distribution ($P(M \mid A, W)$), and the treatment mechanism ($P(A \mid W)$) using an ensemble of learners. Cross-fitting (sample splitting) avoids overfitting bias.

2. **Riesz representer estimation.** The targeting step in TMLE requires a "clever covariate" that depends on the estimand. For mediation, this involves density ratios of the mediator distribution under treatment vs. control. The `crumble` package estimates these using a neural network (the `nn_module` argument) rather than requiring explicit density estimation.

3. **Targeting (fluctuation).** Update the initial outcome model to solve the efficient influence function equation, making the estimator doubly robust: consistent if either the outcome model or the combination of treatment/mediator models is correct.

4. **Plug-in estimation and inference.** Compute the effect estimate from the targeted predictions and obtain standard errors from the influence function.

::: {.callout-note collapse="true"}
## Technical detail: the mediation g-computation formulas (expand)

Under the identification assumptions, the NDE and NIE can be expressed as weighted averages over the outcome model and the mediator distribution. The NDE formula compares the expected outcome under treatment versus control, holding the mediator distribution at its control-arm values. The NIE formula holds treatment on and compares how the outcome changes when the mediator distribution shifts from its control-arm shape to its treatment-arm shape.

$$\text{NDE} = \sum_m \sum_w \big[E[Y \mid A=1, M=m, W=w] - E[Y \mid A=0, M=m, W=w]\big] \cdot P(M=m \mid A=0, W=w) \cdot P(W=w)$$

$$\text{NIE} = \sum_m \sum_w E[Y \mid A=1, M=m, W=w] \cdot \big[P(M=m \mid A=1, W=w) - P(M=m \mid A=0, W=w)\big] \cdot P(W=w)$$

These require an outcome model $E[Y \mid A, M, W]$, a mediator model $P(M \mid A, W)$, and marginalization over $W$. TMLE targets these functionals to achieve double robustness.
:::


## Step 5: Diagnostics, Sensitivity, and Reporting

After estimation, check:

- **Propensity score overlap:** extreme treatment probabilities signal positivity problems.
- **Mediator distribution overlap:** if the mediator distributions under treatment and control barely overlap, the density ratios will be extreme and estimates unstable.
- **Influence function diagnostics:** large or highly variable influence function values indicate fragile estimates.
- **Sensitivity analysis:** consider how much unmeasured confounding would be needed to explain away the result (e.g., E-values for mediation, or simulation-based sensitivity).

We demonstrate diagnostics in the worked example below.


# Data Simulation

We simulate a dataset where dapagliflozin reduces MACE both directly and through HbA1c improvement. The data-generating process is simple enough to compute ground-truth effects for comparison.

```{r}
library(tidyverse)
set.seed(2026)
n <- 2000
```

**Baseline confounders**

```{r}
age      <- rnorm(n, mean = 62, sd = 10)
bmi      <- rnorm(n, mean = 31, sd = 5)
hba1c_bl <- rnorm(n, mean = 8.2, sd = 1.3)
cvd      <- rbinom(n, 1, plogis(-3.5 + 0.04 * age + 0.03 * bmi))
egfr     <- pmax(15, rnorm(n, mean = 95 - 0.5 * (age - 60), sd = 15))
statin   <- rbinom(n, 1, plogis(-1.5 + 1.8 * cvd + 0.02 * age))
```

**Treatment assignment (A)**

```{r}
lp_trt <- -0.5 +
  -0.02 * (age - 62) +
   0.03 * (bmi - 31) +
  -0.10 * (hba1c_bl - 8.2) +
  -0.50 * cvd +
   0.015 * (egfr - 80)

A <- rbinom(n, 1, plogis(lp_trt))
```

**Intermediate confounder (Z): blood pressure change**

This is a variable affected by treatment that confounds the mediator-outcome relationship. Including $Z$ lets us demonstrate recanting-twins estimation.

```{r}
Z <- rnorm(n,
  mean = -2 * A + 0.3 * (bmi - 31) + 0.02 * (age - 62),
  sd = 4
)
```

**Mediator (M): 6-month HbA1c change**

```{r}
M <- rnorm(n,
  mean = -0.3 - 0.5 * A + 0.1 * (hba1c_bl - 8.2) -
         0.005 * egfr + 0.01 * (age - 62) + 0.02 * Z,
  sd = 0.6
)
```

**Outcome (Y): 1-year MACE**

```{r}
lp_out <- -3.5 +
  -0.25 * A +
   0.40 * M +
   0.05 * Z +
   0.04 * (age - 62) +
   0.02 * (bmi - 31) +
   0.90 * cvd +
  -0.01 * (egfr - 80) +
   0.25 * statin +
   0.15 * A * cvd

Y <- rbinom(n, 1, plogis(lp_out))
```

**Assemble dataset**

```{r}
dat <- tibble(age, bmi, hba1c_bl, cvd, egfr, statin, A, Z, M, Y)
```

Quick sanity check:

```{r}
dat |>
  group_by(A) |>
  summarise(
    n = n(),
    mean_M = mean(M),
    mean_Z = mean(Z),
    mean_Y = mean(Y),
    .groups = "drop"
  )
```

The DAG for this simulation:

```{r}
#| label: fig-simulation-dag
#| fig-cap: "DAG for the simulated data-generating process"
#| fig-width: 7
#| fig-height: 4
dag_sim <- dagify(
  A ~ W, Z ~ A + W, M ~ A + Z + W, Y ~ A + M + Z + W,
  exposure = "A", outcome = "Y",
  labels = c(W = "W\n(age, bmi, hba1c_bl,\ncvd, egfr, statin)",
             A = "A", Z = "Z", M = "M", Y = "Y"),
  coords = list(
    x = c(W = 1.5, A = 0, Z = 1, M = 2, Y = 3),
    y = c(W = 1, A = 0, Z = 0, M = 0, Y = 0)
  )
)
ggdag(dag_sim, text = TRUE, use_labels = "label", text_size = 2.8) +
  theme_dag()
```

Because $Z$ is affected by $A$ and confounds $M \to Y$, natural effects require the cross-world assumption. Recanting twins and randomized interventional effects handle $Z$ explicitly.


## True Causal Effects

We compute ground-truth effects from the structural model via Monte Carlo integration.

```{r}
#| cache: true
set.seed(999)
n_mc <- 50

# Mediator and Z means under each treatment
Z_mean_A0 <- -2 * 0 + 0.3 * (bmi - 31) + 0.02 * (age - 62)
Z_mean_A1 <- -2 * 1 + 0.3 * (bmi - 31) + 0.02 * (age - 62)

compute_EY <- function(a_val, m_vals, z_vals, idx) {
  lp <- -3.5 +
    -0.25 * a_val +
     0.40 * m_vals +
     0.05 * z_vals +
     0.04 * (age[idx] - 62) +
     0.02 * (bmi[idx] - 31) +
     0.90 * cvd[idx] +
    -0.01 * (egfr[idx] - 80) +
     0.25 * statin[idx] +
     0.15 * a_val * cvd[idx]
  mean(plogis(lp))
}

true_nde_vals <- numeric(n)
true_nie_vals <- numeric(n)

for (i in 1:n) {
  # Draw Z from P(Z | A=0, W) and P(Z | A=1, W)
  z0 <- rnorm(n_mc, mean = Z_mean_A0[i], sd = 4)
  z1 <- rnorm(n_mc, mean = Z_mean_A1[i], sd = 4)

  # M means given Z draws
  M_mean_fn <- function(a, z) {
    -0.3 - 0.5 * a + 0.1 * (hba1c_bl[i] - 8.2) -
      0.005 * egfr[i] + 0.01 * (age[i] - 62) + 0.02 * z
  }

  # Draw M from P(M | A=0, Z(0), W)
  m_A0_Z0 <- rnorm(n_mc, mean = M_mean_fn(0, z0), sd = 0.6)
  # Draw M from P(M | A=1, Z(1), W)
  m_A1_Z1 <- rnorm(n_mc, mean = M_mean_fn(1, z1), sd = 0.6)

  # NDE: E[Y(1, M(0), Z(0))] - E[Y(0, M(0), Z(0))]
  # (holding mediator and Z at control values)
  EY_1_M0 <- compute_EY(1, m_A0_Z0, z0, i)
  EY_0_M0 <- compute_EY(0, m_A0_Z0, z0, i)
  EY_1_M1 <- compute_EY(1, m_A1_Z1, z1, i)

  true_nde_vals[i] <- EY_1_M0 - EY_0_M0
  true_nie_vals[i] <- EY_1_M1 - EY_1_M0
}

true_NDE <- mean(true_nde_vals)
true_NIE <- mean(true_nie_vals)
true_TE  <- true_NDE + true_NIE

cat("True Total Effect:", round(true_TE, 4), "\n")
cat("True NDE (direct):", round(true_NDE, 4), "\n")
cat("True NIE (via M): ", round(true_NIE, 4), "\n")
cat("NDE + NIE:        ", round(true_NDE + true_NIE, 4), "\n")
```


# Why Naive Approaches Fail

## The Baron-Kenny Trap

The classic Baron-Kenny approach fits regressions for $Y \sim A$ and $Y \sim A + M$, then reads off coefficients. For binary outcomes, this fails for several reasons:

```{r}
total_mod <- glm(Y ~ A + age + bmi + hba1c_bl + cvd + egfr + statin,
                 family = binomial, data = dat)

full_mod  <- glm(Y ~ A + M + age + bmi + hba1c_bl + cvd + egfr + statin,
                 family = binomial, data = dat)

med_mod   <- lm(M ~ A + age + bmi + hba1c_bl + cvd + egfr + statin,
                data = dat)

cat("Total effect (A without M):", round(coef(total_mod)["A"], 3), "(log-OR)\n")
cat("Direct effect (A with M):  ", round(coef(full_mod)["A"], 3), "(log-OR)\n")
cat("Treatment on mediator:     ", round(coef(med_mod)["A"], 3), "(mean change)\n")
```

**Why this is wrong:**

1. **Non-collapsibility.** The log-odds ratio changes when you add $M$ even if there is no mediation, simply because the logistic link is non-collapsible. The "change in coefficient" is not the indirect effect.

2. **Scale dependence.** The difference-in-coefficients method requires additivity on the scale of the link function, which logistic regression does not provide for risk differences.

3. **Post-treatment confounding.** Adding $M$ (a post-treatment variable) to the regression without proper identification opens backdoor paths through any $Z$ that affects both $M$ and $Y$.

4. **No valid inference.** Standard errors for the "indirect effect" (product of coefficients) rely on normality assumptions that are wrong for binary outcomes.

These are not minor quibbles; they produce qualitatively misleading answers. A proper mediation analysis requires specifying the estimand, verifying identification, and using an estimator matched to both.


# Worked Example: Mediation with `crumble`

The `crumble` package (Williams & Díaz, 2024) implements TMLE-based mediation analysis with cross-fitting, flexible machine-learning nuisance estimation, and neural-network-based Riesz representer estimation. It supports natural effects, randomized interventional effects, organic effects, and recanting twins.

```{r}
#| eval: false
# Install crumble (if not already installed):
# remotes::install_github("nt-williams/crumble")
# Also needs: torch, mlr3superlearner, earth, ranger
```

## Example 1: `weight_behavior` Dataset

We start with the example dataset from `mma::weight_behavior`, which is used in the `crumble` README and is known to work.

```{r}
#| eval: false
library(crumble)
data(weight_behavior, package = "mma")

# Treatment: sports participation (factor with levels 1, 2)
# Outcome: bmi (continuous)
# Mediators: exercises, overweigh
# Intermediate confounder: snack (affected by sports, confounds mediator→outcome)
# Covariates: age, sex, tvhours

# --- Natural effects (assumes no intermediate confounding) ---
fit_natural <- crumble(
  data      = weight_behavior,
  trt       = "sports",
  outcome   = "bmi",
  covar     = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  d0 = \(data, trt) factor(rep(1, nrow(data)), levels = c("1", "2")),
  d1 = \(data, trt) factor(rep(2, nrow(data)), levels = c("1", "2")),
  effect    = "N",
  learners  = c("mean", "glm", "earth", "ranger"),
  nn_module = sequential_module(),
  control   = crumble_control(crossfit_folds = 2L, epochs = 20L)
)

print(fit_natural)
```

```{r}
#| eval: false
# --- Randomized interventional effects ---
fit_ri <- crumble(
  data      = weight_behavior,
  trt       = "sports",
  outcome   = "bmi",
  covar     = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  moc       = "snack",
  d0 = \(data, trt) factor(rep(1, nrow(data)), levels = c("1", "2")),
  d1 = \(data, trt) factor(rep(2, nrow(data)), levels = c("1", "2")),
  effect    = "RI",
  learners  = c("mean", "glm", "earth", "ranger"),
  nn_module = sequential_module(),
  control   = crumble_control(crossfit_folds = 2L, epochs = 20L)
)

print(fit_ri)
```

```{r}
#| eval: false
# --- Recanting twins (handles intermediate confounding explicitly) ---
fit_rt <- crumble(
  data      = weight_behavior,
  trt       = "sports",
  outcome   = "bmi",
  covar     = c("age", "sex", "tvhours"),
  mediators = c("exercises", "overweigh"),
  moc       = "snack",
  d0 = \(data, trt) factor(rep(1, nrow(data)), levels = c("1", "2")),
  d1 = \(data, trt) factor(rep(2, nrow(data)), levels = c("1", "2")),
  effect    = "RT",
  learners  = c("mean", "glm", "earth", "ranger"),
  nn_module = sequential_module(),
  control   = crumble_control(crossfit_folds = 2L, epochs = 20L)
)

print(fit_rt)
```


## Example 2: Simulated Pharmacoepi Data

Now we use the simulated dataset from this chapter. We need to convert treatment to a factor so that `d0` and `d1` return the correct type.

```{r}
#| eval: false
library(crumble)

# Prepare data: treatment must be a factor
sim_dat <- dat |>
  mutate(A = factor(A, levels = c("0", "1")))

# Define intervention functions
d0_fn <- \(data, trt) factor(rep("0", nrow(data)), levels = c("0", "1"))
d1_fn <- \(data, trt) factor(rep("1", nrow(data)), levels = c("0", "1"))

# Covariate and mediator names
covar_names <- c("age", "bmi", "hba1c_bl", "cvd", "egfr", "statin")
```

### Natural effects (no intermediate confounder adjustment)

```{r}
#| eval: false
fit_sim_N <- crumble(
  data      = sim_dat,
  trt       = "A",
  outcome   = "Y",
  covar     = covar_names,
  mediators = "M",
  d0        = d0_fn,
  d1        = d1_fn,
  effect    = "N",
  learners  = c("mean", "glm", "earth", "ranger"),
  nn_module = sequential_module(),
  control   = crumble_control(crossfit_folds = 2L, epochs = 5L)
)

print(fit_sim_N)
```

### Recanting twins (with intermediate confounder Z)

Because our DGP includes $Z$ (blood pressure change) as a treatment-induced confounder of $M \to Y$, recanting twins is the more appropriate estimand.

```{r}
#| eval: false
fit_sim_RT <- crumble(
  data      = sim_dat,
  trt       = "A",
  outcome   = "Y",
  covar     = covar_names,
  mediators = "M",
  moc       = "Z",
  d0        = d0_fn,
  d1        = d1_fn,
  effect    = "RT",
  learners  = c("mean", "glm", "earth", "ranger"),
  nn_module = sequential_module(),
  control   = crumble_control(crossfit_folds = 2L, epochs = 5L)
)

print(fit_sim_RT)
```

### Randomized interventional effects (with Z)

```{r}
#| eval: false
fit_sim_RI <- crumble(
  data      = sim_dat,
  trt       = "A",
  outcome   = "Y",
  covar     = covar_names,
  mediators = "M",
  moc       = "Z",
  d0        = d0_fn,
  d1        = d1_fn,
  effect    = "RI",
  learners  = c("mean", "glm", "earth", "ranger"),
  nn_module = sequential_module(),
  control   = crumble_control(crossfit_folds = 2L, epochs = 5L)
)

print(fit_sim_RI)
```

::: {.callout-tip}
## Faster vs. more accurate settings

The settings above (`crossfit_folds = 2L, epochs = 5L`) are deliberately fast for tutorials. For real analyses, increase to `crossfit_folds = 10L, epochs = 100L` (the defaults) and consider adding more flexible learners (e.g., `"xgboost"`, `"glmnet"`).
:::


# Interpretation

## What do the effects mean in context?

In the dapagliflozin example:

- The **direct effect** captures the cardiovascular benefit of dapagliflozin that does *not* operate through HbA1c reduction. This includes blood pressure lowering, weight loss, ketone body effects, and direct cardiac remodeling. A large direct effect suggests SGLT2 inhibitors have a unique cardioprotective mechanism beyond glucose lowering.

- The **indirect effect** (via M) captures how much of the cardiovascular benefit is attributable to the drug's effect on HbA1c. If this is large, other glucose-lowering drugs might produce similar cardiovascular benefits.

- The **total effect** is the sum (for additive decompositions). It should be close to what you would estimate from a standard TMLE or IPTW analysis of $A$ on $Y$.

The choice of effect definition matters for interpretation. Natural effects answer "what if we could intervene on the mediator to hold it at its control value?", a hypothetical intervention that may not be physically realizable. Randomized interventional effects answer "what if we randomly assigned the mediator value from its distribution under control?", a stochastic intervention that is closer to policy-relevant questions. Recanting twins decompose the total effect into path-specific components that remain identifiable even with intermediate confounding.

None of these claims establish "mechanism" in a biological sense without strong assumptions. The decomposition tells you how much of the effect flows through the measured mediator *given the causal model and identification assumptions*. Unmeasured mediator-outcome confounders, measurement error in $M$, or misspecification of the time ordering can all distort the decomposition.


## Caution: proportion mediated

The "proportion mediated" (NIE / TE) is commonly reported but has several problems:

- It is **undefined** when the total effect is zero or near zero.
- It can be **greater than 1 or negative** when the direct and indirect effects have opposite signs.
- It is **highly variable** because it is a ratio of two estimated quantities, each with uncertainty.
- It is **scale-dependent**: the proportion mediated on the risk-difference scale differs from that on the odds-ratio scale.

Report the direct and indirect effects on the absolute scale (risk difference) with confidence intervals. If you report proportion mediated, note these caveats.


# Diagnostics

## Propensity Score Overlap

```{r}
ps <- predict(
  glm(A ~ age + bmi + hba1c_bl + cvd + egfr + statin,
      family = binomial, data = dat),
  type = "response"
)

ggplot(dat, aes(x = ps, fill = factor(A, labels = c("Sulfonylurea", "Dapagliflozin")))) +
  geom_density(alpha = 0.45) +
  labs(x = "P(A=1 | W)", fill = "Treatment",
       title = "Propensity Score Overlap") +
  theme_minimal() +
  scale_fill_manual(values = c("#4477AA", "#EE6677"))
```

## Mediator Distribution by Treatment

The indirect effect depends on treatment actually shifting the mediator distribution. If the distributions overlap completely, there is no indirect effect to detect.

```{r}
ggplot(dat, aes(x = M, fill = factor(A, labels = c("Sulfonylurea", "Dapagliflozin")))) +
  geom_density(alpha = 0.45) +
  labs(x = "6-month HbA1c change (M)", fill = "Treatment",
       title = "Mediator Distribution by Treatment Group") +
  theme_minimal() +
  scale_fill_manual(values = c("#4477AA", "#EE6677"))
```

## Mediator Density Ratio

For TMLE-based mediation, the density ratio $f(M \mid A=0, W) / f(M \mid A=1, W)$ is a key intermediate quantity. Extreme values indicate instability.

```{r}
m_mod <- lm(M ~ A + age + bmi + hba1c_bl + egfr + Z, data = dat)
sigma_m <- sigma(m_mod)

M_mean_a0 <- predict(m_mod, newdata = dat |> mutate(A = 0))
M_mean_a1 <- predict(m_mod, newdata = dat |> mutate(A = 1))

density_ratio <- dnorm(dat$M, M_mean_a0, sigma_m) /
  pmax(dnorm(dat$M, M_mean_a1, sigma_m), 1e-10)

ggplot(tibble(r = density_ratio), aes(x = r)) +
  geom_histogram(bins = 50, fill = "#CC6677", alpha = 0.7) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  labs(x = "f(M|A=0,W) / f(M|A=1,W)", y = "Count",
       title = "Mediator Density Ratio") +
  theme_minimal()

cat("Density ratio summary:\n")
summary(density_ratio)
cat("Proportion > 3:", mean(density_ratio > 3), "\n")
```

## Mediator-Outcome Relationship

```{r}
ggplot(dat, aes(x = M, y = Y)) +
  geom_smooth(method = "loess", se = TRUE, color = "#228833") +
  facet_wrap(~factor(A, labels = c("Sulfonylurea", "Dapagliflozin"))) +
  labs(x = "6-month HbA1c change (M)", y = "P(MACE)",
       title = "Mediator-Outcome Relationship by Treatment") +
  theme_minimal()
```


# What to Report

When publishing a mediation analysis, include:

1. **Effect definition:** Which estimand (natural, interventional, recanting twins) and why.
2. **Time ordering:** When each variable was measured and why the ordering supports mediation claims.
3. **Confounder sets:** Which variables are treated as baseline confounders ($W$), mediators ($M$), and intermediate confounders ($Z$), with justification.
4. **Identification assumptions:** State them explicitly. Note which are testable and which require subject-matter judgment.
5. **Estimator details:** TMLE with cross-fitting, number of folds, learner library.
6. **Point estimates and uncertainty:** Effect estimates on the risk-difference scale with 95% confidence intervals.
7. **Diagnostics:** Propensity score overlap, mediator distribution overlap, density ratio stability.
8. **Sensitivity analysis:** How robust are conclusions to unmeasured confounding or model misspecification?
9. **Limitations:** Measurement error in the mediator, unmeasured pathways, temporal resolution of the data.


# Key Takeaways

::: {.callout-note}
## Summary

1. **Mediation analysis decomposes a total effect** into direct and indirect components. This tells us *how much* of a drug's benefit flows through a specific mechanism, but only under strong assumptions.

2. **The choice of estimand matters.** Natural effects require the cross-world assumption (no treatment-induced mediator-outcome confounders). Randomized interventional effects and recanting twins relax this at the cost of a slightly different interpretation.

3. **Baron-Kenny regression fails** for binary outcomes because of non-collapsibility and post-treatment confounding. Even for continuous outcomes, it assumes linearity and no interactions.

4. **TMLE-based mediation** (via `crumble`) combines flexible nuisance estimation, cross-fitting, and targeting to provide doubly robust estimates with valid inference.

5. **Diagnostics are essential:** propensity score overlap, mediator distribution shift, density ratio stability, and the mediator-outcome relationship should all be examined.

6. **Report responsibly.** State the estimand, assumptions, time ordering, and limitations. Avoid over-interpreting "proportion mediated" or claiming biological mechanism without acknowledging the assumptions required.
:::


# Method Comparison

| Method | Handles nonlinearity | Doubly robust | Works with ML | Cross-world needed | Package |
|--------|:---:|:---:|:---:|:---:|---------|
| Baron-Kenny | No | No | No | Implicitly | Base R |
| Parametric G-comp | Partially | No | No | Yes | `mediation` |
| Weighting | Yes | No | Partially | Yes | `CMAverse` |
| **TMLE (natural)** | **Yes** | **Yes** | **Yes** | **Yes** | `crumble` |
| **TMLE (interv.)** | **Yes** | **Yes** | **Yes** | **No** | `crumble` |
| **TMLE (recanting)** | **Yes** | **Yes** | **Yes** | **No** | `crumble` |


```{r}
sessionInfo()
```


## Sources and Further Reading

- Zheng Z, van der Laan MJ (2012). Targeted maximum likelihood estimation of natural direct effects. *Int J Biostat* 8(1):Article 3.
- VanderWeele TJ (2015). *Explanation in Causal Inference: Methods for Mediation and Interaction*. Oxford University Press.
- Díaz I, Hejazi NS (2020). Causal mediation analysis for stochastic interventions. *JRSS-B* 82(3):661--683.
- Pearl J (2001). Direct and indirect effects. *Proceedings of the 17th Conference on Uncertainty in Artificial Intelligence*, 411--420.
- Robins JM, Richardson TS (2010). Alternative graphical causal models and the identification of direct effects. In *Causality and Psychopathology*, Oxford University Press.
- Williams NT, Díaz I (2024). `crumble`: An R package for TMLE-based mediation analysis. [GitHub](https://github.com/nt-williams/crumble).
- van der Laan MJ, Rose S (2011). *Targeted Learning*. Springer.

## What comes next

Mediation analysis asks how a treatment works. The final chapter asks a complementary question for safety studies: what does your analysis actually estimate when patients switch treatments, discontinue, or experience competing events? We close with the ICH E9(R1) estimand framework for time-to-event data.
