---
format: html
---

# Lessons from the Antiretroviral Therapy Case Study {#sec-case-study-synthesis}

*Estimated reading time: 10 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

Throughout this handbook, we have used the longitudinal HIV antiretroviral therapy cohort to illustrate every major concept, from defining causal questions to diagnosing estimator behavior [@petersen2014causal; @hernan2025whatif]. See the [Running Case Study](00-case-study.qmd) for the full data description, including the variable dictionary and target trial protocol. This final section draws the lessons together and traces the progression from naive methods to principled causal estimation. See the [Running Case Study](00-case-study.qmd) for the data description and variable dictionary.

## What a naive analysis might have done

Without the roadmap, a researcher would likely fit a Cox or pooled logistic regression, adjust for time-varying CD4 because it is a confounder, and report the hazard ratio as the effect of HAART. Two of those moves are the traps developed earlier in the book. Adjusting for CD4 blocks part of the treatment effect, because CD4 is also affected by prior treatment (@sec-longitudinal-confounding), and the hazard ratio answers a different question from the one a clinician asks (@sec-regression-vs-causal). As the running case study shows, the result can make an effective treatment look harmful.

## How the Causal Roadmap clarified the question

Running the seven steps of @sec-roadmap on this cohort fixed each of those problems before any model was fit. A target trial protocol set the question (the effect on mortality of initiating HAART at seroconversion versus not); a DAG made the dual role of CD4, a confounder of future treatment and a mediator of past treatment, explicit; the estimand was the marginal risk difference rather than a hazard ratio; and the identification assumptions (exchangeability, consistency, positivity) were stated together with where they might fail, namely unmeasured adherence and clinical staging. Only then were g-computation, IPTW, and TMLE applied and compared.

## Why time-dependent confounding required longitudinal methods

The central lesson of this case study is that **standard regression cannot handle treatment-confounder feedback**. When a time-varying confounder (`cd4.sqrt`) is both a consequence of prior treatment (`haartind`) and a cause of future treatment, the analyst faces an impossible choice in a regression framework:

- **Do not adjust for time-varying CD4:** leaves confounding of future treatment unaddressed.
- **Adjust for time-varying CD4:** blocks part of the treatment effect and introduces collider bias.

Neither option produces a consistent estimate of the causal effect, as demonstrated concretely in @sec-longitudinal-confounding. The g-methods (g-computation, IPTW with marginal structural models, longitudinal TMLE) resolve this dilemma by modeling the treatment mechanism separately from the outcome and using that model to reweight or standardize the analysis. @sec-illustrated-ltmle walked through the mechanics of longitudinal TMLE step by step, showing how backwards sequential regression conditions on `cd4.sqrt` where it confounds future treatment and integrates it out where it mediates past treatment. This is not a minor technical improvement; it is a qualitative change in what can be estimated.

## What the example illustrates about real-world evidence analyses

The HIV/HAART example is not unique. Treatment-confounder feedback arises whenever:

- Treatment decisions respond to time-varying patient status (lab values, symptoms, disease progression)
- Those same patient characteristics are also affected by prior treatment
- The outcome of interest occurs at the end of a follow-up period during which both treatment and confounders evolve

This pattern is ubiquitous in chronic disease epidemiology, pharmacoepidemiology, and comparative effectiveness research. Examples include:

- Insulin dose adjustments based on HbA1c in diabetes
- Immunosuppressant dosing based on kidney function in transplant recipients
- Chemotherapy regimen changes based on tumor markers in oncology
- Blood pressure medication titration based on BP readings

In all of these settings, the methods covered in this handbook, not standard regression, are needed to estimate causal effects correctly. The Causal Roadmap provides the structure, and the estimation methods (g-computation, IPTW, TMLE, longitudinal TMLE) provide the tools.

## Summary

```{r}
#| code-fold: true
source("_data/load-haart-cohort.R")

# Final summary table of approaches and their limitations
tibble::tribble(
  ~Approach, ~`Handles confounding`, ~`Handles feedback`, ~`Marginal effect`, ~`Valid inference`,
  "Crude comparison",          "No",  "No",  "Yes", "No",
  "Baseline regression",       "Partially", "No", "No (conditional)", "Yes (if correct)",
  "Time-varying regression",   "Partially", "No (biased)", "No (conditional)", "Yes (if correct)",
  "G-computation",             "Yes", "Yes (longitudinal)", "Yes", "Bootstrap only",
  "IPTW / MSM",                "Yes", "Yes", "Yes", "Sandwich SE",
  "TMLE",                      "Yes", "Yes (longitudinal)", "Yes", "Influence curve"
) |>
  knitr::kable(caption = "Comparison of approaches for the HAART case study")
```

Even after applying longitudinal methods, the results depend on untestable identification assumptions. @sec-sensitivity-analysis showed how to probe this dependence using E-values, quantitative bias analysis, and truncation sensitivity. For the HAART cohort, unmeasured adherence is the most plausible threat: patients who adhere to treatment tend to be healthier, potentially inflating the apparent benefit. Sensitivity analysis quantifies how strong this confounding would need to be to explain away the estimated effect.

The progression from naive regression to longitudinal TMLE is not just methodological sophistication for its own sake. Each step solves a real problem that the previous approach could not handle. The antiretroviral therapy case study makes those problems, and their solutions, concrete.

## Sources

- Robins JM, Hernan MA, Brumback B (2000). Marginal structural models and causal inference in epidemiology. *Epidemiology* 11(5):550-560.
- Hernan MA, Brumback B, Robins JM (2000). Marginal structural models to estimate the causal effect of zidovudine on the survival of HIV-positive men. *Epidemiology* 11(5):561-570.
- Van der Wal WM, Geskus RB (2011). ipw: An R Package for Inverse Probability Weighting. *Journal of Statistical Software* 43(13):1-23.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- Hernan MA, Robins JM (2020). *Causal Inference: What If*. Chapman & Hall/CRC.
