---
subtitle: "End-to-end workflow for a dynamic treatment rule in the HIV cohort"
format: html
---

# From Question to Estimate: An Applied Longitudinal Analysis with LMTP {#sec-applied-lmtp}

*Estimated reading time: 55 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

By this point the moving parts of a longitudinal targeted-learning analysis are already in hand. We have the backwards sequential-regression logic of longitudinal TMLE (@sec-illustrated-ltmle), the shift-function and `lmtp_tmle` workflow for modified treatment policies (@sec-adherence-shift), and the positivity diagnostics that decide whether any of it can be trusted (@sec-positivity-overlap). This chapter does not re-teach that machinery. It spends its length on the one thing the method chapters leave out: a single analysis carried from a clinical question to a defended estimate, with every decision in view.

The thread is a dynamic treatment rule, initiate highly active antiretroviral therapy (HAART) when CD4 falls below a clinically meaningful threshold, estimated with the `lmtp` package for longitudinal modified treatment policies (LMTP) [@diaz2021lmtp; @williams2025lmtp]. A dynamic rule is a sharper question than the adherence-distribution shift of @sec-adherence-shift or the static "treat everyone" contrast of the earlier chapters, and it is the kind of question clinical guidelines actually pose.


## The causal question

Clinical guidelines for HIV treatment have evolved over time, but a recurring question is: **what would mortality look like if HAART were initiated whenever a patient's CD4 count fell below 350 cells/µL, compared to the observed treatment pattern where initiation depends on clinical judgment?**

This is a dynamic treatment rule. The intervention depends on the patient's evolving health status, not just baseline characteristics, and it corresponds to a clinically interpretable policy: start treatment when the immune system shows meaningful decline. The comparison is between this threshold-based rule and the natural course (whatever clinicians actually did).

The estimand is the risk difference in mortality under the two regimes, marginalized over the study population. In the language of modified treatment policies [@haneuse2013mtp; @diaz2021lmtp], we are shifting the treatment process so that HAART is initiated at any visit where CD4 falls below the threshold, for patients who were not already on treatment.

A precise word on the estimand: the CD4 < 350 rule is a deterministic dynamic treatment regime, an assignment that is a function of the patient's covariate history rather than of the treatment a clinician would otherwise have chosen [@young2020failure]. A modified treatment policy is the broader class of interventions whose assigned value depends on the natural (observed) treatment value [@haneuse2013mtp]. The LMTP framework of [@diaz2021lmtp] estimates both, which is why the same `lmtp_tmle` workflow applies to this dynamic rule.


## Why a simpler analysis would be inadequate

A static point-treatment comparison (ever-treated vs. never-treated) ignores the timing and clinical context of treatment initiation. Two patients who both "ever started HAART" may have initiated at quite different times and for quite different reasons, and their prognoses at initiation may have been nothing alike. The static comparison conflates the effect of early initiation in healthier patients with late initiation in sicker patients.

A dynamic-rule analysis asks instead what would happen if initiation followed one clinical rule applied consistently to everyone, which is closer to what guideline developers need. Because CD4 both responds to past treatment and drives future treatment, this is the treatment-confounder-feedback setting of @sec-longitudinal-confounding, and it needs the longitudinal g-methods developed there and in @sec-illustrated-ltmle. The identification assumptions are their longitudinal versions: sequential exchangeability, positivity at each visit, and consistency of the dynamic rule. Informative censoring, when sicker patients drop out, is handled by the same machinery: `lmtp_tmle` models the censoring process alongside treatment and covariates whenever a censoring node is supplied. The simulated cohort used below has complete follow-up, so no censoring node is passed here; with dropout one would add per-visit censoring indicators through the `cens` argument, as in `lmtp_tmle(..., cens = c("C1", "C2"))`.


## Data setup

The real `haartdat` cohort [@vanderwal2011ipw] (introduced in the [Running Case Study](00-case-study.qmd)) follows 1,200 patients over irregularly spaced visits, but it contains only 31 deaths spread across up to 38 visits. That outcome sparsity makes a short-horizon longitudinal estimator unstable, so for a fully reproducible end-to-end run we use a **simulated extension of the cohort**: a balanced three-visit panel calibrated to the same clinical story, where CD4 drives both treatment initiation and mortality and HAART raises subsequent CD4. The simulation keeps the feature that matters pedagogically here, treatment-confounder feedback, while guaranteeing the estimator has enough events and enough low-CD4-untreated person-time to estimate the contrast.

```{r}
#| label: lmtp-simulate
#| code-fold: show
#| message: false
#| warning: false
library(dplyr)
library(lmtp)

set.seed(2024)
n <- 700
cd4_threshold <- 350  # cells/µL: the initiation rule we will evaluate

# Baseline covariates
age <- round(rnorm(n, 40, 9))
sex <- rbinom(n, 1, 0.45)

# Visit 1: CD4, then treatment (more likely to start if CD4 is low)
L_1 <- pmax(50, rnorm(n, 420, 140))
A_1 <- rbinom(n, 1, plogis(-0.5 + 1.2 * (L_1 < cd4_threshold) + 0.01 * (40 - age)))

# Visit 2: CD4 responds to treatment; HAART is absorbing (once on, stays on)
L_2 <- pmax(50, L_1 + ifelse(A_1 == 1, 60, -30) + rnorm(n, 0, 80))
A_2 <- ifelse(A_1 == 1, 1, rbinom(n, 1, plogis(-0.5 + 1.2 * (L_2 < cd4_threshold))))

# Visit 3
L_3 <- pmax(50, L_2 + ifelse(A_2 == 1, 60, -30) + rnorm(n, 0, 80))
A_3 <- ifelse(A_2 == 1, 1, rbinom(n, 1, plogis(-0.5 + 1.2 * (L_3 < cd4_threshold))))

# Death by end of follow-up: driven by low CD4 and lack of treatment
lp <- -1.6 + 1.4 * (L_3 < 250) + 0.8 * (L_2 < 250) - 0.7 * (A_1 + A_2 + A_3) / 3
died <- rbinom(n, 1, plogis(lp))

dat_wide <- data.frame(
  age, sex,
  L_1, L_2, L_3,   # time-varying CD4 (confounder), measured before treatment each visit
  A_1, A_2, A_3,   # HAART indicator at each visit (absorbing)
  Y = died
)

dat_wide |>
  summarise(
    patients = n(),
    deaths = sum(Y),
    pct_died = round(100 * mean(Y), 1),
    pct_treated_by_v3 = round(100 * mean(A_3), 1)
  ) |>
  as.data.frame()
```


### Defining the dynamic rule

The modified treatment policy we evaluate is: **initiate HAART at the first visit where CD4 count falls below 350 cells/µL, for patients not already on treatment.** Once initiated, HAART continues for all subsequent visits (the absorbing property).

In `lmtp`, the shift function receives the full data frame and the name of the *current* treatment node. We parse the visit index from that name, look at all CD4 measurements up to and including that visit, and set the treatment to 1 if CD4 has ever fallen below the threshold, or if the patient was already treated. This makes the rule absorbing by construction.

```{r}
#| label: lmtp-shift
#| code-fold: show
shift_cd4_rule <- function(data, trt) {
  t <- as.integer(sub("A_", "", trt))         # visit index for this node
  L_cols <- paste0("L_", seq_len(t))           # CD4 at this and all prior visits
  ever_low <- Reduce(`|`, lapply(L_cols, function(cc) data[[cc]] < 350))
  as.numeric(data[[trt]] == 1 | ever_low)
}
```

This function has the `(data, trt)` signature that `lmtp` expects. Because the shifted value depends on the patient's evolving CD4 and on the natural treatment value (through the absorbing `data[[trt]] == 1` term) rather than on a fixed assignment, we pass `mtp = TRUE` below so that `lmtp` applies the density-ratio machinery used for interventions that reference the observed treatment.


### Specifying the LMTP node structure

`lmtp` needs the treatment nodes in temporal order, a list of the time-varying covariates measured before each treatment node, and the baseline covariates. Our simulated frame is already in the wide, one-row-per-patient format `lmtp` expects.

```{r}
#| label: lmtp-nodes
#| code-fold: show
trt_vars  <- c("A_1", "A_2", "A_3")        # treatment in temporal order
time_vary <- list("L_1", "L_2", "L_3")     # time-varying confounder at each visit
baseline  <- c("age", "sex")               # fixed baseline covariates
```


## Estimation with LMTP

With the data prepared and the shift function defined, estimation follows the same `lmtp` workflow used in @sec-adherence-shift: estimate the counterfactual mortality risk under the dynamic rule, estimate it under the natural course (`shift = NULL`), and contrast the two.

```{r}
#| label: lmtp-estimate
#| code-fold: show
#| message: false
#| warning: false
#| cache: true

set.seed(42)
fit_rule <- lmtp_tmle(
  data = dat_wide,
  trt = trt_vars,
  outcome = "Y",
  baseline = baseline,
  time_vary = time_vary,
  shift = shift_cd4_rule,
  mtp = TRUE,
  outcome_type = "binomial",
  learners_outcome = c("SL.glm", "SL.mean"),
  learners_trt = c("SL.glm", "SL.mean"),
  folds = 5
)

fit_natural <- lmtp_tmle(
  data = dat_wide,
  trt = trt_vars,
  outcome = "Y",
  baseline = baseline,
  time_vary = time_vary,
  shift = NULL,                 # natural course: no intervention
  # mtp not needed here: shift = NULL is the observed (natural) course
  outcome_type = "binomial",
  learners_outcome = c("SL.glm", "SL.mean"),
  learners_trt = c("SL.glm", "SL.mean"),
  folds = 5
)

fit_rule
fit_natural
```

The call pattern is the one from @sec-adherence-shift. `lmtp_tmle()` fits a sequentially doubly robust estimator with cross-fitting [@diaz2021lmtp], and it calls the shift function at each visit to set the counterfactual treatment, with `mtp = TRUE` because the shifted value depends on the natural one. What is new here is only the rule that shift function encodes.

The contrast turns the two intervention-specific risks into the policy-relevant risk difference:

```{r}
#| label: lmtp-contrast
#| code-fold: show
ct <- lmtp_contrast(fit_rule, ref = fit_natural, type = "additive")
ct
```

The `estimate` column is the marginal risk difference: how much the CD4 < 350 initiation rule would change mortality relative to observed practice, with a negative value indicating fewer deaths under the rule. Here the rule changes mortality by an estimated `r round(ct$estimates$estimate, 4)` (95% CI `r round(ct$estimates$conf.low, 4)` to `r round(ct$estimates$conf.high, 4)`).


## Diagnostics and empirical support

Before interpreting the results, several diagnostic checks are important.

### Is the dynamic rule well-supported by the data?

The rule initiates HAART when CD4 drops below 350. For this to be estimable, we need person-visits where CD4 was below 350 but the patient did *not* start HAART (the counterfactual contrast). If every patient with low CD4 was immediately treated, there is no empirical support for the comparison, and the estimator relies entirely on model extrapolation.

```{r}
#| label: lmtp-support
#| code-fold: show
# Count person-visits where CD4 < 350 but the patient was not (yet) treated:
# these are the observations the rule would shift, and where support lives.
support <- with(dat_wide,
  (L_1 < 350 & A_1 == 0) +
  (L_2 < 350 & A_2 == 0) +
  (L_3 < 350 & A_3 == 0)
)
data.frame(
  shiftable_person_visits = sum(support),
  patients_with_any_shift = sum(support > 0)
)
```

If the rule would shift very few observations (most patients with low CD4 are already treated), the analysis has limited power. If the rule would shift many observations but the natural treatment probability is near 1 in that stratum, positivity is thin and the estimate is fragile.

### Propensity score overlap

The treatment mechanism at each visit should show adequate overlap. Estimated probabilities of treatment near 0 or 1 at any visit signal positivity concerns that compound across time points. The block below fits the per-visit initiation model among patients still at risk of initiating (those not already on HAART) and reports the range of fitted probabilities at each visit.

```{r}
#| label: lmtp-overlap
#| code-fold: show
# Treatment-initiation probability at each visit, among patients still at risk
# of initiating. Values near 0 or 1 flag positivity strain; because the
# longitudinal estimand chains across visits, thin overlap at any single visit
# propagates into the contrast.
g1 <- predict(glm(A_1 ~ age + sex + L_1, binomial, dat_wide), type = "response")

at_risk2 <- dat_wide$A_1 == 0
g2 <- predict(glm(A_2 ~ age + sex + L_2, binomial, dat_wide[at_risk2, ]),
              type = "response")

at_risk3 <- dat_wide$A_2 == 0
g3 <- predict(glm(A_3 ~ age + sex + L_3, binomial, dat_wide[at_risk3, ]),
              type = "response")

data.frame(
  visit     = c("A_1", "A_2", "A_3"),
  n_at_risk = c(nrow(dat_wide), sum(at_risk2), sum(at_risk3)),
  g_min     = round(c(min(g1), min(g2), min(g3)), 3),
  g_max     = round(c(max(g1), max(g2), max(g3)), 3)
)
```

### Effective sample size

In longitudinal settings, the effective sample size can shrink rapidly because the cumulative product of inverse-probability weights grows with the number of visits. Even with TMLE-based estimation, which avoids explicit weight products, near-positivity violations at individual time points can inflate the influence of a few observations. The visit-1 weights below illustrate the idea. The effective sample size is the squared sum of weights over the sum of squared weights, and it falls further once weights are multiplied across visits.

```{r}
#| label: lmtp-ess
#| code-fold: show
# Inverse-probability weights for the observed (natural-course) treatment at
# visit 1, and the resulting effective sample size. An ESS well below the
# nominal n means a few person-histories dominate the estimate.
w1   <- ifelse(dat_wide$A_1 == 1, 1 / g1, 1 / (1 - g1))
ess1 <- sum(w1)^2 / sum(w1^2)

data.frame(
  nominal_n     = nrow(dat_wide),
  max_weight_v1 = round(max(w1), 1),
  ess_visit1    = round(ess1, 1)
)
```

A large gap between `nominal_n` and `ess_visit1`, or a large `max_weight_v1`, at any visit is the longitudinal analogue of the point-treatment positivity problem: the contrast is being carried by a small number of person-histories, and the confidence interval understates the true fragility.


## Interpretation

The contrast from `lmtp_contrast()` estimates the risk difference in mortality under the CD4 < 350 initiation rule versus the natural course. A negative value indicates that the rule would reduce mortality.

Several points matter for interpretation:

**The estimand is policy-relevant.** Unlike a static "always treat" intervention, which may be clinically impossible, this rule corresponds to something a clinician could actually follow. The estimate answers a concrete question: how many deaths per 100 patients would be prevented if HAART were initiated at CD4 < 350 rather than at the discretion of individual clinicians?

**The estimate depends on identification assumptions.** Sequential exchangeability (no unmeasured time-varying confounders), positivity at each visit, and consistency of the dynamic rule must all hold. If unmeasured factors such as substance use, housing instability, or clinic-specific prescribing norms affect both treatment timing and mortality, the estimate is biased regardless of the estimator.

**Support limitations constrain the estimate.** If the rule is very different from observed practice, say most patients with CD4 < 350 were already treated, the estimate relies on extrapolation. Varying the rule threshold (comparing CD4 < 250, CD4 < 350, and CD4 < 500) can help assess whether the result is robust or driven by sparse data in a particular region.


## Sensitivity and robustness checks

### Alternative rule thresholds

Running the analysis at multiple CD4 thresholds gives a dose-response view of early versus late initiation. The helper below rebuilds the shift function for a given threshold and returns the contrast against the natural course.

```{r}
#| label: lmtp-threshold-sweep
#| code-fold: show
#| message: false
#| warning: false
#| cache: true

contrast_at <- function(thresh) {
  shift_t <- function(data, trt) {
    t <- as.integer(sub("A_", "", trt))
    L_cols <- paste0("L_", seq_len(t))
    ever_low <- Reduce(`|`, lapply(L_cols, function(cc) data[[cc]] < thresh))
    as.numeric(data[[trt]] == 1 | ever_low)
  }
  set.seed(42)
  fit_t <- lmtp_tmle(
    data = dat_wide, trt = trt_vars, outcome = "Y",
    baseline = baseline, time_vary = time_vary,
    shift = shift_t, mtp = TRUE, outcome_type = "binomial",
    learners_outcome = c("SL.glm", "SL.mean"),
    learners_trt = c("SL.glm", "SL.mean"), folds = 2
  )
  ct <- lmtp_contrast(fit_t, ref = fit_natural, type = "additive")
  data.frame(threshold = thresh,
             risk_diff = round(ct$estimates$estimate, 4),
             conf_low = round(ct$estimates$conf.low, 4),
             conf_high = round(ct$estimates$conf.high, 4))
}

threshold_results <- do.call(rbind, lapply(c(250, 350, 500), contrast_at))
threshold_results
```

```{r}
#| label: fig-threshold-doseresponse
#| fig-cap: "Risk difference versus the natural course for CD4-based initiation rules at three thresholds, with 95% confidence intervals. Reading left to right traces a dose-response curve for earlier initiation; the dashed line marks no benefit over the observed treatment pattern."
#| code-fold: true
library(ggplot2)
ggplot(threshold_results, aes(factor(threshold), risk_diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray60") +
  geom_pointrange(aes(ymin = conf_low, ymax = conf_high)) +
  labs(x = "CD4 initiation threshold (cells/µL)",
       y = "Risk difference vs. natural course",
       title = "Dose-response across CD4 initiation thresholds") +
  theme_minimal()
```

@fig-threshold-doseresponse plots the risk difference at each threshold against the natural course. If the estimated benefit increases smoothly with earlier initiation, that supports the clinical intuition. If the estimate becomes unstable or reverses at very early thresholds, that likely reflects positivity violations rather than a genuine clinical effect reversal.

### Nuisance model sensitivity

The `lmtp` estimator is sequentially doubly robust, but in finite samples the choice of learner library matters. Rerunning the analysis with different Super Learner libraries, say GLM-only versus a richer library that includes random forests and LASSO, assesses model sensitivity. Large changes in the estimate across libraries suggest that the nuisance functions are poorly estimated and the result should be read cautiously.


## Key lessons

This analysis pulls together the major threads of the handbook in a single applied example:

1. **The causal question comes first.** The dynamic rule was defined before any model was fit. The estimand, a marginal risk difference under a threshold-based initiation policy, was chosen because it answers a question that clinicians and guideline developers care about.

2. **Longitudinal methods are necessary.** The treatment-confounder feedback between HAART and CD4 means no cross-sectional analysis can correctly estimate the effect of a dynamic rule. The `lmtp` framework handles this by modeling the treatment, covariate, and censoring processes sequentially at each visit.

3. **Diagnostics matter as much as estimation.** Whether the dynamic rule is well-supported by the data, whether positivity holds at each time point, and whether the influence of individual observations is bounded all determine whether the estimate is credible. A precisely estimated quantity from a poorly supported intervention is not useful.

4. **Interpretation requires restraint.** The estimate is a causal claim only if the identification assumptions hold. In observational HIV cohort data, unmeasured confounding from clinical judgment, adherence patterns, and social determinants of health is a real concern. Sensitivity analysis and negative controls are part of the analysis, not optional additions to it.


## Sources and further reading

- @diaz2021lmtp introduces the LMTP framework for realistic longitudinal interventions that depend on the natural treatment value.
- @williams2025lmtp, the `lmtp` R package on CRAN with a comprehensive vignette.
- @haneuse2013mtp and @diaz2021lmtp define modified treatment policies and their identification.
- @kennedy2019ipsi, the incremental propensity score intervention (IPSI) as an alternative stochastic shift.
- @vanderLaan2007superlearner, the Super Learner / stacking estimator used for nuisance functions.
- @vanderLaan2012ltmle, longitudinal TMLE (LTMLE) for time-varying treatments with treatment-confounder feedback.
- @vanderwal2011ipw, the `ipw` R package and the `haartdat` HIV cohort used as the running example.
- @williams2024beyondatesite, the Beyond the ATE course site with additional applied lmtp examples.
- @nugent2023mtp, an applied MTP example illustrating the workflow in a real-world pharmacoepidemiology setting.
- @diaz2024survival and @wen2023ipsi for survival outcomes and the IPSI approach with time-to-event data.
- @petersen2014causal and @dang2023causal develop the Causal Roadmap and its extension to real-world evidence.
- `lmtp` R package: [CRAN](https://cran.r-project.org/package=lmtp)
