---
title: "From Question to Estimate: An Applied Longitudinal Analysis with LMTP"
subtitle: "End-to-end workflow for a dynamic treatment rule in the HIV cohort"
format: html
---

*Estimated reading time: 55 minutes*

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4,
  fig.align = "center"
)
```

The preceding chapters taught each piece of the causal inference toolkit separately: the roadmap, g-computation, IPTW, TMLE, Super Learner, and longitudinal methods. This chapter puts them together. We walk through a complete applied analysis, from the causal question through estimation, diagnostics, and interpretation, using a realistic longitudinal intervention estimated with the `lmtp` package [@diaz2021lmtp].

The example is deliberately more advanced than the earlier teaching examples. Instead of a static "everyone treated vs. no one treated" contrast, we define a dynamic treatment rule: initiate HAART when CD4 count drops below a clinically meaningful threshold. This is the kind of question that clinical guidelines actually address, and it requires the longitudinal methods from Part IV.


# The causal question

Clinical guidelines for HIV treatment have evolved over time, but a recurring question is: **what would mortality look like if HAART were initiated whenever a patient's CD4 count fell below 350 cells/µL, compared to the observed treatment pattern where initiation depends on clinical judgment?**

This is a dynamic treatment rule. The intervention depends on the patient's evolving health status, not just baseline characteristics. It corresponds to a clinically interpretable policy: "start treatment when the immune system shows meaningful decline." The comparison is between this threshold-based rule and the natural course (whatever clinicians actually did).

The estimand is the risk difference in mortality under the two regimes, marginalized over the study population. In the language of modified treatment policies, we are shifting the treatment process so that HAART is initiated at any visit where CD4 falls below the threshold, for patients who were not already on treatment.


# Why a simpler analysis would be inadequate

A static point-treatment comparison (ever-treated vs. never-treated) ignores the timing and clinical context of treatment initiation. Two patients who both "ever started HAART" may have initiated at very different times, for very different reasons, with very different prognoses at initiation. The static comparison conflates the effect of early initiation in healthier patients with late initiation in sicker patients.

A dynamic rule analysis addresses this directly. It asks: what would happen if treatment initiation followed a specific clinical rule, applied consistently to everyone? This is closer to the question that guideline developers actually need answered, and it requires the longitudinal g-methods from Part IV because:

- CD4 evolves over time and is affected by prior treatment (treatment-confounder feedback)
- The intervention is defined at each visit conditional on current health status
- Censoring is potentially informative (sicker patients may drop out)

The identification assumptions are the longitudinal versions introduced in [Chapter 3.7](03-07-longitudinal-td-confounding.qmd): sequential exchangeability, positivity at each time point, and consistency of the dynamic rule.


# Data setup

We use a simulated extension of the HAART cohort that includes the variables needed for a longitudinal LMTP analysis. The key addition beyond the baseline `haartdat` is a treatment variable at each visit that reflects whether HAART would be initiated under the dynamic rule.

```{r}
#| code-fold: true
library(tidyverse)
source("_data/load-haart-cohort.R")

# Work with the longitudinal panel
# Each row = one 100-day interval for one patient
# Key variables: haartind (treatment), cd4.sqrt (time-varying confounder),
#   event (death), dropout (censoring), age, sex

# Create visit-level CD4 in original units for interpretability
haartdat <- haartdat |>
  mutate(cd4 = cd4.sqrt^2)

# Summary of the longitudinal structure
haartdat |>
  group_by(patient) |>
  summarise(
    n_visits = n(),
    ever_haart = max(haartind),
    died = max(event),
    dropped = max(dropout),
    .groups = "drop"
  ) |>
  summarise(
    patients = n(),
    median_visits = median(n_visits),
    pct_haart = round(100 * mean(ever_haart), 1),
    pct_died = round(100 * mean(died), 1),
    pct_dropped = round(100 * mean(dropped), 1)
  ) |>
  as.data.frame()
```


## Defining the dynamic rule

The modified treatment policy we evaluate is: **initiate HAART at the first visit where CD4 count falls below 350 cells/µL, for patients not already on treatment.** Once initiated, HAART continues for all subsequent visits (the absorbing property).

In `lmtp` terms, the shift function modifies the treatment variable at each visit based on the patient's current CD4:

```{r}
#| eval: false
# Shift function for the dynamic rule:
# If not yet on HAART and CD4 < 350, initiate (set haartind = 1)
# If already on HAART, stay on (haartind remains 1)
# Otherwise, leave unchanged
shift_cd4_rule <- function(data, trt) {
  a <- data[[trt]]
  # Patients already on HAART stay on
  already_on <- a == 1
  # For those not yet on HAART, check CD4
  cd4_below <- data[["cd4"]] < 350
  # Apply the rule: initiate if CD4 < 350 and not already on
  a[!already_on & cd4_below] <- 1
  return(a)
}
```

This function has the `(data, trt)` signature that `lmtp` expects. It modifies the treatment value based on the patient's current state, which is what makes it a modified treatment policy rather than a static intervention.


## Preparing the data for LMTP

The `lmtp` package expects wide-format data with one row per patient and separate columns for each time point. We reshape the longitudinal panel accordingly.

```{r}
#| eval: false
library(lmtp)

# Determine max number of visits
max_visits <- haartdat |>
  count(patient) |>
  pull(n) |>
  max()

# For this analysis, use the first 10 visits (1000 days)
T_max <- 10

# Reshape to wide format
dat_wide <- haartdat |>
  group_by(patient) |>
  mutate(visit = row_number()) |>
  filter(visit <= T_max) |>
  ungroup()

# Create wide-format columns for treatment, covariates, censoring
# (This is a simplified version; a full analysis would handle
# the reshaping more carefully)
trt_vars  <- paste0("A_", 1:T_max)
cens_vars <- paste0("C_", 1:T_max)
baseline  <- c("age", "sex")
time_vary <- lapply(1:T_max, function(t) paste0("cd4_", t))
```

The details of the reshaping depend on the specific data structure. The key requirement is that `lmtp` receives: treatment columns in temporal order, time-varying covariate columns, censoring indicators, baseline covariates, and the outcome.


# Estimation with LMTP

With the data prepared and the shift function defined, estimation follows the standard `lmtp` workflow.

```{r}
#| eval: false
# Estimate the counterfactual outcome under the dynamic rule
fit_rule <- lmtp_tmle(
  data = dat_wide,
  trt = trt_vars,
  outcome = "Y",
  baseline = baseline,
  time_vary = time_vary,
  cens = cens_vars,
  shift = shift_cd4_rule,
  mtp = TRUE,
  outcome_type = "binomial",
  learners_outcome = c("SL.glm", "SL.mean"),
  learners_trt = c("SL.glm", "SL.mean"),
  folds = 5
)

# Estimate the natural course (no intervention)
shift_identity <- function(data, trt) data[[trt]]

fit_natural <- lmtp_tmle(
  data = dat_wide,
  trt = trt_vars,
  outcome = "Y",
  baseline = baseline,
  time_vary = time_vary,
  cens = cens_vars,
  shift = shift_identity,
  mtp = TRUE,
  outcome_type = "binomial",
  learners_outcome = c("SL.glm", "SL.mean"),
  learners_trt = c("SL.glm", "SL.mean"),
  folds = 5
)

# Contrast
contrast <- lmtp_contrast(fit_rule, ref = fit_natural, type = "additive")
contrast
```

The `lmtp_tmle()` function implements a sequentially doubly robust estimator with cross-fitting. At each time point, it estimates the outcome regression and the treatment/censoring mechanisms, then applies a TMLE-style targeting step. The shift function is called at each time point to determine the counterfactual treatment value, and the `mtp = TRUE` flag tells `lmtp` that the shifted value depends on the natural treatment value.


# Diagnostics and empirical support

Before interpreting the results, several diagnostic checks are important.

## Is the dynamic rule well-supported by the data?

The rule initiates HAART when CD4 drops below 350. For this to be estimable, we need patients in the data who had CD4 below 350 but did *not* start HAART (the counterfactual contrast). If every patient with low CD4 was immediately treated, there is no empirical support for the comparison, and the estimator relies entirely on model extrapolation.

```{r}
#| code-fold: true
# Check: among visits where CD4 < 350 and patient not yet on HAART,
# what fraction actually started HAART?
haartdat |>
  filter(cd4 < 350, haartind == 0) |>
  summarise(
    n_visits = n(),
    n_patients = n_distinct(patient),
    pct_would_shift = 100
  ) |>
  as.data.frame()
```

If the rule would shift very few observations (most patients with low CD4 are already treated), the analysis has limited power. If the rule would shift many observations but the natural treatment probability is near 1 in that stratum, positivity is thin and the estimate is fragile.

## Propensity score overlap

The treatment mechanism at each visit should show adequate overlap. Estimated probabilities of treatment near 0 or 1 at any visit indicate positivity concerns that compound across time points.

## Effective sample size

In longitudinal settings, the effective sample size can shrink rapidly because the cumulative product of inverse-probability weights grows with the number of visits. Even with TMLE-based estimation (which avoids explicit weight products), near-positivity violations at individual time points can inflate the influence of a few observations. Monitoring the distribution of influence-curve values across patients helps detect this.


# Interpretation

The contrast from `lmtp_contrast()` estimates the risk difference in mortality under the CD4 < 350 initiation rule versus the natural course. A negative value indicates that the rule would reduce mortality.

Several points matter for interpretation:

**The estimand is policy-relevant.** Unlike a static "always treat" intervention, which may be clinically impossible, this rule corresponds to something a clinician could actually follow. The estimate answers: how many deaths per 100 patients would be prevented if HAART were initiated at CD4 < 350 rather than at the discretion of individual clinicians?

**The estimate depends on identification assumptions.** Sequential exchangeability (no unmeasured time-varying confounders), positivity at each visit, and consistency of the dynamic rule must all hold. If unmeasured factors (substance use, housing instability, clinic-specific prescribing norms) affect both treatment timing and mortality, the estimate is biased regardless of the estimator.

**Support limitations constrain the estimate.** If the rule is very different from observed practice (for example, if most patients with CD4 < 350 were already treated), the estimate relies on extrapolation. Sensitivity to the rule threshold (e.g., comparing CD4 < 250 vs CD4 < 350 vs CD4 < 500) can help assess whether the result is robust or driven by sparse data in a particular region.


# Sensitivity and robustness checks

## Alternative rule thresholds

Running the analysis at multiple CD4 thresholds provides a dose-response perspective on early vs. late initiation:

```{r}
#| eval: false
# Compare rules at different thresholds
thresholds <- c(200, 250, 350, 500)

# For each threshold, define the shift function, run lmtp_tmle,
# and extract the contrast vs. natural course.
# Results can be plotted as risk difference vs threshold.
```

If the estimated benefit increases smoothly with earlier initiation, this supports the clinical intuition. If the estimate becomes unstable or reverses at very early thresholds, this likely reflects positivity violations rather than a genuine clinical effect reversal.

## Nuisance model sensitivity

The `lmtp` estimator is sequentially doubly robust, but in finite samples, the choice of learner library matters. Running the analysis with different Super Learner libraries (e.g., GLM-only vs. a richer library including random forests and LASSO) assesses model sensitivity. Large changes in the estimate across libraries suggest that the nuisance functions are poorly estimated and the result should be interpreted cautiously.


# Key lessons

This analysis integrates every major concept from the handbook into a single applied example:

1. **The causal question comes first.** The dynamic rule was defined before any model was fit. The estimand, a marginal risk difference under a threshold-based initiation policy, was chosen because it answers a question that clinicians and guideline developers care about.

2. **Longitudinal methods are necessary.** The treatment-confounder feedback between HAART and CD4 means that no cross-sectional analysis can correctly estimate the effect of a dynamic rule. The `lmtp` framework handles this by modeling the treatment, covariate, and censoring processes sequentially at each visit.

3. **Diagnostics matter as much as estimation.** Whether the dynamic rule is well-supported by the data, whether positivity holds at each time point, and whether the influence of individual observations is bounded all determine whether the estimate is credible. A precisely estimated quantity from a poorly supported intervention is not useful.

4. **Interpretation requires restraint.** The estimate is a causal claim only if the identification assumptions hold. In observational HIV cohort data, unmeasured confounding from clinical judgment, adherence patterns, and social determinants of health is a real concern. Sensitivity analysis and negative controls are not optional additions. They are part of the analysis.


## Sources and further reading

- Diaz I, Williams N, Hoffman KL, Schenck EJ (2023). Nonparametric causal effects based on longitudinal modified treatment policies. *Journal of the American Statistical Association* 118(542):846-856.
- Petersen ML, van der Laan MJ (2014). Causal models and learning from data. *Epidemiology* 25(3):418-426.
- Dang LE, Tager I, Petersen ML (2023). A causal roadmap for generating high-quality real-world evidence. *Journal of Clinical and Translational Science* 7(1):e127.
- `lmtp` R package: [CRAN](https://cran.r-project.org/package=lmtp)
