---
title: "Causal Learning Handbook"
---

# Welcome! {.unnumbered}

This handbook introduces the **Causal Roadmap**, **Targeted Maximum Likelihood Estimation (TMLE)**, and **Super Learner** for modern causal inference in epidemiology, with an emphasis on pharmacoepidemiology and clinical trial analysis. Each chapter builds on the previous ones, progressing from foundational concepts through estimation and diagnostics to longitudinal methods and advanced applications, with worked R code examples throughout.

## How to use this book

The chapters are organized into five parts plus appendices. Part I builds the conceptual foundations you need before running any estimator. Part II walks through each major estimation strategy with hands-on code. Part III covers diagnostics and treatment effect heterogeneity. Part IV tackles longitudinal settings with time-varying confounding. Part V addresses advanced applications including hybrid RCT-RWD designs, mediation, and time-to-event estimands. You can read the book front-to-back or jump to specific chapters as needed.

---

## Running Case Study

A single recurring example is used across the entire handbook: the longitudinal HIV antiretroviral therapy cohort example used in the marginal structural model literature (Van der Wal & Geskus, 2011). Start here to meet the dataset, understand why naive methods fail, and see how each chapter builds on the same example.

| Chapter | Description |
|---------|-------------|
| [Running Case Study](00-case-study.qmd) | Clinical motivation, observed data description, causal question, target trial protocol, DAGs with treatment-confounder feedback, why naive methods fail, and a roadmap of how the example appears in each chapter. |

---

## Part I: Foundations

These chapters establish the conceptual framework for causal inference: why standard regression is not enough, how the Causal Roadmap structures an analysis, and how we formally link causal questions to quantities we can estimate from data.

| Chapter | Description |
|---------|-------------|
| [1.1 Regression vs. Causal Inference](01-01-regression-vs-causal.qmd) | Why regression coefficients are not causal effects, and what the counterfactual framework offers instead. |
| [1.2 The Causal Roadmap](01-02-causal-roadmap.qmd) | The six-step Causal Roadmap: target trial emulation, causal models, estimands, identification (consistency, exchangeability, positivity), estimation, and interpretation. |

---

## Part II: Core Estimators

These chapters cover the core estimation approaches for causal inference, from outcome modeling to doubly robust methods, with progressive worked examples in R.

| Chapter | Description |
|---------|-------------|
| [2.1 G-Computation](02-01-gcomputation.qmd) | Outcome modeling and standardization: fit a model for $E[Y \mid A, W]$, predict under interventions, and average to get marginal causal effects. |
| [2.2 Inverse Probability of Treatment Weighting](02-02-iptw.qmd) | Propensity-score-based estimation: reweight the sample to create a pseudo-population where treatment is independent of confounders. Covers stabilized weights, truncation, and diagnostics. |
| [2.3 Doubly Robust Estimation and TMLE](02-03-doubly-robust-tmle.qmd) | AIPW and TMLE side by side: why combining outcome and treatment models gives you two chances to get it right, with efficient influence curve-based inference. |
| [2.4 From Question to Estimate: TMLE in Practice](02-05-tmle-teaching-examples.qmd) | A progressive, illustrated walkthrough comparing naive, g-computation, IPTW, AIPW, and TMLE estimators on a single cardiovascular safety dataset. |

---

## Part III: Diagnostics and Heterogeneity

These chapters address how to assess whether your analysis is trustworthy and how to explore treatment effect variation across patient subgroups.

| Chapter | Description |
|---------|-------------|
| [3.4 Effect Modification](03-04-effect-modification.qmd) | Conditional average treatment effects (CATE), TMLE-based subgroup analyses, and causal forests for heterogeneous treatment effect estimation. |
| [3.5 Advanced Diagnostics](03-05-advanced-diagnostics.qmd) | A systematic framework for identifiability, nuisance model, estimator, and sensitivity diagnostics, with executable R examples. |

---

## Part IV: Longitudinal Causal Inference

These chapters extend the core methods to longitudinal settings where treatment, confounders, and censoring co-evolve over time.

| Chapter | Description |
|---------|-------------|
| [3.7 Imperfect Adherence and Time-Varying Confounding](03-07-longitudinal-td-confounding.qmd) | Why conditioning on time-varying adherence biases causal effects, and how longitudinal TMLE and LMTP address treatment-confounder feedback. |
| [3.9 An Illustrated Guide to Longitudinal TMLE](03-09-illustrated-ltmle.qmd) | A visual, step-by-step guide to L-TMLE: backwards sequential regression, cumulative treatment probabilities, and time-step-specific targeting. |
| [3.3 Longitudinal Case Study](03-03-longitudinal-case-study.qmd) | A worked case study applying the Causal Roadmap to a longitudinal pharmacoepidemiology setting with time-varying confounding. |

---

## Part V: Advanced Applications

These chapters tackle specialized settings: hybrid RCT-RWD designs, pre-specified regulatory analyses, mediation decomposition, and time-to-event estimands.

| Chapter | Description |
|---------|-------------|
| [3.6 RWD Meets RCT: Hybrid Designs](03-06-rwd-meets-rct-hybrid-designs.qmd) | External control arms, transportability, ES-CV-TMLE, and A-TMLE for combining randomized and real-world data sources. |
| [3.8 TMLE in the Clean-Room Framework](03-08-tmle-clean-room.qmd) | Pre-specified TMLE analysis for regulatory pharmacoepidemiology: locked analysis plans, reproducible scripts, and audit trails. |
| [3.10 TMLE for Mediation Analysis](03-10-tmle-mediation.qmd) | Decomposing total effects into direct and indirect pathways using TMLE-based mediation with the `crumble` package. |
| [3.11 Time-to-Event, Censoring, and Competing Risks](03-11-time-to-event-competing-risks.qmd) | Time-to-event estimands, ICH E9(R1) intercurrent events, informative censoring with IPCW, and competing risks with targeted learning. |
| [3.13 Transportability and Generalizability](03-13-transportability-generalizability.qmd) | Trial-sample vs target-population effects, assumptions for transportability, IOSW/TMLE methods, and diagnostics for covariate shift. |

---

## Synthesis

| Chapter | Description |
|---------|-------------|
| [Lessons from the Case Study](04-03-case-study-synthesis.qmd) | What a naive analysis might have done, how the Causal Roadmap clarified the question, and why time-dependent confounding required longitudinal methods. |

---

## Appendices

| Chapter | Description |
|---------|-------------|
| [Common Pitfalls](common-pitfalls.Rmd) | A catalog of the most frequent mistakes in causal inference and how to avoid them. |
| [Annotated Bibliography](anotated_bibliography.Rmd) | Key readings in causal inference, targeted learning, and real-world evidence generation. |
| [Other Resources](other_resources.Rmd) | Links to external courses, tutorials, and reference sites for further learning. |

---

## Package installations

To run the code examples in this book, install the following R packages:

```{r}
#| eval: false

## Core CRAN packages
cran_pkgs <- c(
  "tidyverse", "here", "sandwich", "lmtest", "Matrix", "survival",
  "ipw", "dagitty", "ggdag"
)

## Super Learner ecosystem
sl_pkgs <- c(
  "SuperLearner", "glmnet", "gam", "ranger", "randomForest", "xgboost"
)

## Targeted Learning and causal inference
causal_pkgs <- c(
  "tmle", "ltmle", "lmtp", "hal9001", "origami"
)

## Diagnostics and sensitivity analysis
diagnostic_pkgs <- c("EValue", "episensr", "causalsens")

## Combine and install missing packages
all_pkgs <- unique(c(cran_pkgs, sl_pkgs, causal_pkgs, diagnostic_pkgs))
installed <- rownames(installed.packages())
to_install <- setdiff(all_pkgs, installed)

if (length(to_install) > 0) {
  install.packages(to_install, dependencies = TRUE)
}


```




```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'tidyverse', 'here'
), 'packages.bib')
```

```{r, eval=F, include=F}

#to create the book
quarto render

git add docs/
git commit -m "Rebuild book"
git push


#https://amertens.github.io/causal-learning-handbook/


```
