---
title: "Causal Learning Handbook"
---

# Welcome! {.unnumbered}

This handbook introduces the **Causal Roadmap**, **Targeted Maximum Likelihood Estimation (TMLE)**, and **Super Learner** for modern causal inference in epidemiology, with an emphasis on pharmacoepidemiology and clinical trial analysis. Each chapter builds on the previous ones, with worked R code examples throughout.

::: {.callout-tip icon=false}
## ▶ Run the code in your browser

Every R block in this book is also available as a live, editable code cell that runs entirely in the browser via [WebR](https://docs.r-wasm.org/webr/), with no R installation required. Edit the code, re-run it, and see the output update in place.

[**Open the interactive version →**](https://amertens.github.io/causal-learning-handbook/workshop/){.btn .btn-primary role="button"}
:::

This handbook was developed by [Andrew Mertens](mailto:amertens@berkeley.edu) as part of work for the [UC Berkeley Center for Targeted Machine Learning and Causal Inference](https://ctml.berkeley.edu/). For questions, suggestions, or bug reports, please contact Andrew at [amertens@berkeley.edu](mailto:amertens@berkeley.edu).

## Who is this book for?

This handbook is written for analysts who already know how to fit regression models in R and now need to do *causal* analysis on observational health data, primarily **pharmacoepidemiologists, biostatisticians, and epidemiologists** working on drug effectiveness, safety, and real-world evidence. The worked examples assume comfort with R and basic regression; they do not assume prior exposure to potential-outcomes notation, targeted learning, or g-methods.

Two secondary audiences are in scope:

- **Reviewers** (journal, HTA, or regulatory) who want a reference for what a well-specified causal analysis should look like and what diagnostics to expect. Parts I, IV, and the case-study synthesis are written so they can be read without working through the code.
- **Methods-oriented graduate students** in biostatistics, epidemiology, or health data science who want a code-first introduction to TMLE, Super Learner, and LMTP grounded in a single running example.

The book is *not* a substitute for a foundational causal inference text (for that, we recommend Hernán & Robins, *Causal Inference: What If*), but it is intended to be self-contained for someone working through a longitudinal real-world-data analysis from scratch. A [Glossary](glossary.qmd) is provided in the appendix for the abbreviations and short technical terms used throughout.

## How to use this book

**Short on time?** Start with the [Short Course](00-short-course.qmd). It covers the Causal Roadmap, TMLE (by hand and with software), longitudinal causal inference, and the `lmtp` package (**L**ongitudinal **M**odified **T**reatment **P**olicies) in a single chapter. You can work through it in about an hour and come away with a working understanding of the core framework.

**Want the full treatment?** Read the handbook front-to-back. The chapters are organized into six parts. Part I builds the conceptual foundations: why standard regression is not enough for causal questions, and how the Causal Roadmap structures an analysis. Part II walks through each major estimation strategy with hands-on code. Part III extends to longitudinal settings with time-varying confounding, longitudinal TMLE (L-TMLE), and Longitudinal Modified Treatment Policies (LMTP). Part IV covers diagnostics and sensitivity analysis. Part V is an end-to-end applied workflow built around the `lmtp` package. Part VI addresses special topics including effect heterogeneity, mediation, hybrid RCT/RWD designs, and time-to-event estimands. A short Synthesis part then revisits the running case study, and the appendices collect a glossary, an annotated bibliography, and external resources.

You can also jump to specific chapters as needed. Each chapter is self-contained enough to be read independently, though they build on each other.

## License, citation, and version

This handbook is released under a [Creative Commons Attribution 4.0 International (CC BY 4.0)](https://creativecommons.org/licenses/by/4.0/) license, so you are free to share and adapt the material, including for teaching and commercial use, as long as you give appropriate credit. If you prefer a more restrictive option, CC BY-NC 4.0 or CC BY-SA 4.0 are common alternatives. The source lives at the [GitHub repository](https://github.com/amertens/causal-learning-handbook).

**How to cite.** Mertens A. *Causal Learning Handbook*. Version 0.1, 2026. <https://amertens.github.io/causal-learning-handbook/>. DOI: *pending* (see note).

A versioned DOI can be minted by enabling the Zenodo and GitHub integration and tagging a release; Zenodo then archives each tagged version and returns a citable DOI to replace the placeholder above.

**Version 0.1, last updated 2026.** This is an actively maintained work and chapters may be revised; see the repository's release history for changes.

---

## Running Case Study

A single recurring example is used across the handbook: the longitudinal HIV antiretroviral therapy cohort from the marginal structural model literature [@vanderwal2011ipw].

| Chapter | Description |
|---------|-------------|
| [Running Case Study](00-case-study.qmd) | Clinical motivation, data structure, causal question, target trial protocol, and longitudinal DAGs. |

---

## Part I: Foundations

| Chapter | Description |
|---------|-------------|
| [Regression vs. Causal Inference](01-01-regression-vs-causal.qmd) | Why a regression coefficient for a treatment is not, on its own, a causal effect; the counterfactual ("what if everyone had been treated?") framework; and the "Table 2 fallacy", interpreting every adjusted coefficient in a regression output as a causal effect for its own variable. |
| [The Causal Roadmap](01-02-causal-roadmap.qmd) | Seven-step roadmap: target trial, causal model, estimand, identification, estimation, sensitivity analysis, and interpretation. |

---

## Part II: Core Estimators

| Chapter | Description |
|---------|-------------|
| [G-Computation](02-01-gcomputation.qmd) | Outcome modeling and standardization: fit a model for the outcome, predict each person's outcome under each treatment value, and average across the sample. |
| [IPTW](02-02-iptw.qmd) | Inverse-probability-of-treatment weighting: weight each individual by the inverse of their probability of receiving the treatment they actually got, so the weighted sample mimics one in which treatment was randomized. Covers stabilized weights, truncation, and diagnostics. |
| [Doubly Robust Estimation and TMLE](02-03-doubly-robust-tmle.qmd) | Augmented Inverse Probability Weighting (AIPW) and Targeted Maximum Likelihood Estimation (TMLE): combining outcome and treatment models for double robustness. |
| [Super Learner](02-04-superlearner.qmd) | Ensemble machine learning for nuisance estimation and integration with TMLE. |

---

## Part III: Longitudinal Causal Inference

| Chapter | Description |
|---------|-------------|
| [Time-Dependent Confounding](03-07-longitudinal-td-confounding.qmd) | Why treatment–confounder feedback is not addressed by standard regression and motivates g-methods. |
| [Longitudinal TMLE: Concepts](03-09-illustrated-ltmle.qmd) | Backwards sequential regression, targeting in the longitudinal setting, and when static regimes fail. |
| [L-TMLE By Hand](03-09b-ltmle-by-hand.qmd) | Technical companion: full by-hand implementation for two time points (optional). |
| [Adherence, Forgiveness, and Shift Methods](03-03b-adherence-shift-methods.qmd) | Comparing drugs under realistic adherence using interventions that shift the *distribution of adherence* (rather than setting adherence to a fixed value). Covers forgiveness curves, the adherence data-generating process (DGP), and four ways to shift an adherence distribution. |

---

## Part IV: Diagnostics and Sensitivity Analysis

| Chapter | Description |
|---------|-------------|
| [Positivity and Overlap](03-02-positivity-overlap.qmd) | Structural vs practical positivity violations, propensity score diagnostics, and effective sample size. |
| [Advanced Diagnostics](03-05-advanced-diagnostics.qmd) | Assumption, model, estimator, and design diagnostics for targeted learning analyses. |
| [Sensitivity Analysis](03-12-sensitivity-analysis.qmd) | E-values, quantitative bias analysis, truncation sensitivity, and negative controls. |

---

## Part V: Putting It All Together

| Chapter | Description |
|---------|-------------|
| [Applied Longitudinal Workflow with LMTP](05-01-applied-lmtp-workflow.qmd) | End-to-end analysis of a dynamic treatment rule using `lmtp`, from question to interpretation. |

---

## Part VI: Special Topics

::: {.callout-warning}
## In development
These chapters are under active development. Content may be incomplete or change substantially.
:::

| Chapter | Description |
|---------|-------------|
| [Effect Modification and Heterogeneous Treatment Effects (HTE)](03-04-effect-modification.qmd) | Estimating the conditional average treatment effect (CATE), TMLE-based subgroup analyses, and causal forests. |
| [TMLE for Mediation](03-10-tmle-mediation.qmd) | Decomposing effects into direct and indirect pathways with the `crumble` package. |
| [Time-to-Event and Competing Risks](03-11-time-to-event-competing-risks.qmd) | Survival estimands, ICH E9(R1) mapping, IPCW, and competing risks. |
| [RWD Meets RCT: Hybrid Designs](03-06-rwd-meets-rct-hybrid-designs.qmd) | External controls, ES-CV-TMLE, and A-TMLE for combining data sources. |
| [Data Fusion Tutorial](03-06b-data-fusion-tutorial.qmd) | Practical tutorial on ES-CV-TMLE and A-TMLE with the `EScvtmle` and `atmle` packages. |
| [Transportability](03-13-transportability-generalizability.qmd) | Trial-sample vs target-population effects, and transportability methods based on inverse odds of sampling weights (IOSW) and TMLE. |

---

## Synthesis

| Chapter | Description |
|---------|-------------|
| [Lessons from the Case Study](04-03-case-study-synthesis.qmd) | What naive analysis misses, how the Causal Roadmap clarifies the question, and why longitudinal g-methods are needed to recover an unbiased estimate of the HAART-on-mortality effect under treatment–confounder feedback. |

---

## Appendices

| Chapter | Description |
|---------|-------------|
| [Glossary](glossary.qmd) | One- to two-sentence definitions for the abbreviations and short technical terms used throughout the handbook. |
| [Annotated Bibliography](anotated_bibliography.Rmd) | Key readings in causal inference, targeted learning, and real-world evidence. |
| [Other Resources](other_resources.Rmd) | External courses, tutorials, and reference sites. |

---

## Package installations

```{r}
#| eval: false
#| code-fold: true
## Core CRAN packages
cran_pkgs <- c(
  "tidyverse", "here", "sandwich", "lmtest", "Matrix", "survival",
  "ipw", "dagitty", "ggdag"
)

## Super Learner ecosystem
sl_pkgs <- c(
  "SuperLearner", "glmnet", "gam", "ranger", "randomForest", "xgboost"
)

## Targeted Learning and causal inference
causal_pkgs <- c(
  "tmle", "ltmle", "lmtp", "hal9001", "origami"
)

## Diagnostics and sensitivity analysis
diagnostic_pkgs <- c("EValue", "episensr", "causalsens")

## Combine and install missing packages
all_pkgs <- unique(c(cran_pkgs, sl_pkgs, causal_pkgs, diagnostic_pkgs))
installed <- rownames(installed.packages())
to_install <- setdiff(all_pkgs, installed)
if (length(to_install) > 0) {
  install.packages(to_install, dependencies = TRUE)
}
```

```{r include=FALSE}
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'tidyverse', 'here'
), 'packages.bib')
```
