---
title: "Causal Learning Handbook"
---

# Welcome! {.unnumbered}

This handbook introduces the **Causal Roadmap**, **Targeted Maximum Likelihood Estimation (TMLE)**, and **Super Learner** for modern causal inference in epidemiology, with an emphasis on pharmacoepidemiology and clinical trial analysis. Each chapter builds on the previous ones, with worked R code examples throughout.

## How to use this book

The chapters are organized into six parts. Part I builds the conceptual foundations: why regression is not enough, and how the Causal Roadmap structures an analysis. Part II walks through each major estimation strategy with hands-on code. Part III extends to longitudinal settings with time-varying confounding, L-TMLE, and LMTP. Part IV covers diagnostics and sensitivity analysis. Part V is a synthesis chapter applying everything to a realistic longitudinal analysis. Part VI addresses special topics including effect heterogeneity, mediation, hybrid RCT-RWD designs, and time-to-event estimands.

You can read the book front-to-back or jump to specific chapters as needed.

---

## Running Case Study

A single recurring example is used across the handbook: the longitudinal HIV antiretroviral therapy cohort from the marginal structural model literature (Van der Wal & Geskus, 2011).

| Chapter | Description |
|---------|-------------|
| [Running Case Study](00-case-study.qmd) | Clinical motivation, data structure, causal question, target trial protocol, and longitudinal DAGs. |

---

## Part I: Foundations

| Chapter | Description |
|---------|-------------|
| [Regression vs. Causal Inference](01-01-regression-vs-causal.qmd) | Why regression coefficients are not causal effects, the counterfactual framework, and the table-2 fallacy. |
| [The Causal Roadmap](01-02-causal-roadmap.qmd) | Seven-step roadmap: target trial, causal model, estimand, identification, estimation, sensitivity analysis, and interpretation. |

---

## Part II: Core Estimators

| Chapter | Description |
|---------|-------------|
| [G-Computation](02-01-gcomputation.qmd) | Outcome modeling and standardization: predict under interventions and average. |
| [IPTW](02-02-iptw.qmd) | Propensity-score reweighting, stabilized weights, truncation, and diagnostics. |
| [Doubly Robust Estimation and TMLE](02-03-doubly-robust-tmle.qmd) | AIPW and TMLE: combining outcome and treatment models for double robustness. |
| [Super Learner](02-04-superlearner.qmd) | Ensemble machine learning for nuisance estimation and integration with TMLE. |

---

## Part III: Longitudinal Causal Inference

| Chapter | Description |
|---------|-------------|
| [Time-Dependent Confounding](03-07-longitudinal-td-confounding.qmd) | Why treatment-confounder feedback breaks standard regression and motivates g-methods. |
| [Longitudinal TMLE: Concepts](03-09-illustrated-ltmle.qmd) | Backwards sequential regression, targeting in the longitudinal setting, and when static regimes fail. |
| [L-TMLE By Hand](03-09b-ltmle-by-hand.qmd) | Technical companion: full by-hand implementation for two time points (optional). |
| [LMTP Case Study](03-03-longitudinal-case-study.qmd) | Comparing drugs under realistic adherence with distribution-shift interventions. |
| [Adherence, Forgiveness, and Shift Methods](03-03b-adherence-shift-methods.qmd) | Forgiveness curves, the adherence DGP, and four ways to shift an adherence distribution. |

---

## Part IV: Diagnostics and Sensitivity Analysis

| Chapter | Description |
|---------|-------------|
| [Positivity and Overlap](03-02-positivity-overlap.qmd) | Structural vs practical positivity violations, propensity score diagnostics, and effective sample size. |
| [Advanced Diagnostics](03-05-advanced-diagnostics.qmd) | Assumption, model, estimator, and design diagnostics for targeted learning analyses. |
| [Sensitivity Analysis](03-12-sensitivity-analysis.qmd) | E-values, quantitative bias analysis, truncation sensitivity, and negative controls. |

---

## Part V: Putting It All Together

| Chapter | Description |
|---------|-------------|
| [Applied Longitudinal Workflow with LMTP](05-01-applied-lmtp-workflow.qmd) | End-to-end analysis of a dynamic treatment rule using `lmtp`, from question to interpretation. |

---

## Part VI: Special Topics

::: {.callout-warning}
## In development
These chapters are under active development. Content may be incomplete or change substantially.
:::

| Chapter | Description |
|---------|-------------|
| [Effect Modification and HTE](03-04-effect-modification.qmd) | CATE estimation, TMLE-based subgroup analyses, and causal forests. |
| [TMLE for Mediation](03-10-tmle-mediation.qmd) | Decomposing effects into direct and indirect pathways with the `crumble` package. |
| [Time-to-Event and Competing Risks](03-11-time-to-event-competing-risks.qmd) | Survival estimands, ICH E9(R1) mapping, IPCW, and competing risks. |
| [RWD Meets RCT: Hybrid Designs](03-06-rwd-meets-rct-hybrid-designs.qmd) | External controls, ES-CV-TMLE, and A-TMLE for combining data sources. |
| [Data Fusion Tutorial](03-06b-data-fusion-tutorial.qmd) | Practical tutorial on ES-CV-TMLE and A-TMLE with the `EScvtmle` and `atmle` packages. |
| [Transportability](03-13-transportability-generalizability.qmd) | Trial-sample vs target-population effects and IOSW/TMLE transportability methods. |
| [TMLE in the Clean-Room Framework](03-08-tmle-clean-room.qmd) | Pre-specified analysis for regulatory pharmacoepidemiology. |

---

## Synthesis

| Chapter | Description |
|---------|-------------|
| [Lessons from the Case Study](04-03-case-study-synthesis.qmd) | What naive analysis misses, how the Causal Roadmap clarifies the question, and why longitudinal methods are needed. |

---

## Appendices

| Chapter | Description |
|---------|-------------|
| [Annotated Bibliography](anotated_bibliography.Rmd) | Key readings in causal inference, targeted learning, and real-world evidence. |
| [Other Resources](other_resources.Rmd) | External courses, tutorials, and reference sites. |

---

## Package installations

```{r}
#| eval: false
## Core CRAN packages
cran_pkgs <- c(
  "tidyverse", "here", "sandwich", "lmtest", "Matrix", "survival",
  "ipw", "dagitty", "ggdag"
)

## Super Learner ecosystem
sl_pkgs <- c(
  "SuperLearner", "glmnet", "gam", "ranger", "randomForest", "xgboost"
)

## Targeted Learning and causal inference
causal_pkgs <- c(
  "tmle", "ltmle", "lmtp", "hal9001", "origami"
)

## Diagnostics and sensitivity analysis
diagnostic_pkgs <- c("EValue", "episensr", "causalsens")

## Combine and install missing packages
all_pkgs <- unique(c(cran_pkgs, sl_pkgs, causal_pkgs, diagnostic_pkgs))
installed <- rownames(installed.packages())
to_install <- setdiff(all_pkgs, installed)
if (length(to_install) > 0) {
  install.packages(to_install, dependencies = TRUE)
}
```

```{r include=FALSE}
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'tidyverse', 'here'
), 'packages.bib')
```
