---
title: "Talk Results Companion V3: Strong TV Confounding DGP"
subtitle: "Figure and Table Bank — Strong Time-Varying Confounding Parameters"
author: "Sara Levintow & Andrew Mertens"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-summary: "Show code"
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    self-contained: true
execute:
  echo: false
  warning: false
  message: false
  cache: false
---

<!-- ============================================================ -->
<!-- V3: STRONG TIME-VARYING CONFOUNDING DGP                      -->
<!-- Uses modified parameters with strong TV feedback loops        -->
<!-- that make naive/baseline regression badly biased while        -->
<!-- L-TMLE remains approximately unbiased.                       -->
<!-- ============================================================ -->


# Setup {#sec-setup}

```{r setup}
#| cache: false
library(ggplot2)
library(knitr)
library(kableExtra)
library(ltmle)
library(lmtp)
library(grid)
library(gridExtra)
library(dagitty)
library(ggdag)

talk_theme <- theme_minimal(base_size = 14) +
  theme(legend.position = "bottom",
        plot.title    = element_text(face = "bold", size = 16),
        subtitle      = element_text(size = 13, colour = "grey40"),
        axis.title    = element_text(size = 13),
        legend.text   = element_text(size = 12))
theme_set(talk_theme)

# ── Tableau 10 colour-blind friendly palette ──
tab10 <- c(
  blue   = "#4E79A7",
  orange = "#F28E2B",
  red    = "#E15759",
  teal   = "#76B7B2",
  green  = "#59A14F",
  yellow = "#EDC948",
  purple = "#B07AA1",
  pink   = "#FF9DA7",
  brown  = "#9C755F",
  grey   = "#BAB0AC"
)

drug_pal   <- c("Drug A" = tab10[["blue"]], "Drug B" = tab10[["orange"]])
method_pal <- c("Naive"              = tab10[["red"]],
                "Reg (baseline)"     = tab10[["purple"]],
                "Reg (baseline + TV)" = tab10[["brown"]],
                "L-TMLE"             = tab10[["blue"]],
                "LTMLE (GLM)"        = tab10[["blue"]],
                "LMTP (+10%)"        = tab10[["green"]])

# ---- Figure collector: stores ggplot objects for later export ----
.fig_store <- new.env(parent = emptyenv())
save_fig <- function(name, plot = NULL) {
  if (is.null(plot)) plot <- ggplot2::last_plot()
  if (!is.null(plot)) {
    .fig_store[[name]] <- plot
    cat(sprintf("  [saved: %s]\n", name))
  } else {
    cat(sprintf("  [WARN: %s — no plot to save]\n", name))
  }
  invisible(plot)
}

# proj_root = "simulation V4/" directory
# repo_root = top-level repo (parent of "simulation V4/" and "simulation_v4_mc/")
proj_root <- normalizePath(file.path(getwd(), ".."), winslash = "/")
repo_root <- normalizePath(file.path(proj_root, ".."), winslash = "/")
cat("proj_root:", proj_root, "\n")
cat("repo_root:", repo_root, "\n")
source(file.path(proj_root, "R", "sim_v4_dgp.R"))
source(file.path(proj_root, "R", "oracle_v4.R"))
source(file.path(proj_root, "R", "prepare_ltmle_v4.R"))
source(file.path(proj_root, "R", "prepare_lmtp_v4.R"))
source(file.path(proj_root, "R", "diagnostics_v4.R"))
```

```{r strong-tv-params}
#| cache: false

# ── Strong time-varying confounding parameters ──
# Weak baseline confounding, strong TV feedback loops.
# This makes naive and baseline-only regression badly biased,
# while L-TMLE (which adjusts for TV confounders) stays unbiased.
strong_tv_params <- list(
  betaW   = 0.10,   # weak baseline severity -> L (was 0.30)
  alphaW  = -0.05,  # weak baseline severity -> adherence (was -0.20)
  gammaW  = 0.10,   # weak baseline severity -> failure (was 0.30)
  betaP   = 1.50,   # strong: poor adherence worsens future health (was 0.80)
  betaLag = 0.50,   # strong: persistence of L (was 0.35)
  alphaL  = -0.60,  # strong: sicker -> lower adherence (was -0.30)
  gammaL  = 1.00,   # strong: sicker -> higher hazard (was 0.50)
  alphaA  = -0.50   # Drug A has notably lower adherence (was -0.35)
)
```


# Simulate Data {#sec-simulate}

```{r simulate}
#| label: simulate
N_SIM <- 5000
K     <- 4
SEED  <- 2024

set.seed(SEED)
dat <- simulate_v4_data(n = N_SIM, K = K, seed = SEED, params = strong_tv_params)
p   <- attr(dat, "dgp_params")

dat$drug <- factor(ifelse(dat$A0 == 1, "Drug A", "Drug B"),
                   levels = c("Drug A", "Drug B"))
pdc_cols <- paste0("PDC_", 1:K)
```


# Oracle Forgiveness Curves {#sec-oracle}

```{r oracle-grid}
#| label: oracle-grid
pdc_grid <- seq(0.30, 1.00, by = 0.05)
oracle_curve <- data.frame(pdc = numeric(), drug = character(),
                           risk = numeric(), stringsAsFactors = FALSE)
for (d in 0:1) {
  dlabel <- ifelse(d == 0, "Drug B", "Drug A")
  for (pdc_val in pdc_grid) {
    r <- oracle_static_pdc_v4(pdc_val, drug = d, K = K,
                              n_oracle = 100000,
                              seed = SEED + d * 100 + round(pdc_val * 100),
                              params = strong_tv_params)
    oracle_curve <- rbind(oracle_curve,
      data.frame(pdc = pdc_val, drug = dlabel, risk = r, stringsAsFactors = FALSE))
  }
}
```

```{r fig-oracle}
#| label: fig-oracle
#| fig-cap: "True cumulative failure risk by adherence level and drug under strong TV confounding DGP. Drug A shows greater forgiveness at sub-optimal adherence."
#| fig-height: 5

g <- ggplot(oracle_curve, aes(x = pdc, y = risk, colour = drug)) +
  geom_line(linewidth = 1.4) +
  geom_point(size = 2.5) +
  scale_x_continuous(labels = scales::percent, breaks = seq(0.3, 1, 0.1)) +
  scale_y_continuous(labels = scales::percent) +
  scale_colour_manual(values = drug_pal) +
  labs(x = "Adherence (PDC)", y = paste0(K * 3, "-Month Failure Risk"),
       colour = "Regimen",
       title = "Oracle Forgiveness Curves (Strong TV Confounding DGP)",
       subtitle = "Strong feedback: PDC -> L -> PDC creates severe time-varying confounding")
print(g)
save_fig("fig-oracle", plot = g)
```


# DAG: Time-Dependent Confounding {#sec-dag}

```{r fig-dag}
#| label: fig-dag
#| fig-cap: "Two-period DAG with time-dependent confounding. Red dashed arrows show feedback from treatment/outcome to future confounders — the structure that breaks standard regression."
#| fig-height: 6
#| fig-width: 11

library(dagitty)
library(ggdag)

dag <- dagitty('dag {
  W   [pos="0,0.5"]
  A0  [pos="1,-1"]
  L1  [pos="2.5,1.5"]
  PDC1 [pos="4,-1"]
  Y1  [pos="5.5,0"]
  L2  [pos="7,1.5"]
  PDC2 [pos="8.5,-1"]
  Y2  [pos="10,0"]

  W -> A0
  W -> L1
  W -> PDC1
  A0 -> L1
  A0 -> PDC1
  L1 -> PDC1
  L1 -> Y1
  PDC1 -> Y1
  L1 -> L2
  PDC1 -> L2
  PDC1 -> PDC2
  Y1 -> L2
  Y1 -> PDC2
  Y1 -> Y2
  L2 -> PDC2
  L2 -> Y2
  PDC2 -> Y2
}')

dag_tidy <- tidy_dagitty(dag)
dag_df <- dag_tidy$data

# Classify edges
feedback_pairs <- list(
  c("PDC1", "L2"),
  c("Y1",   "L2"),
  c("Y1",   "PDC2")
)

dag_df$edge_type <- "causal"
for (fp in feedback_pairs) {
  idx <- !is.na(dag_df$name) & !is.na(dag_df$to) &
         dag_df$name == fp[1] & dag_df$to == fp[2]
  dag_df$edge_type[idx] <- "feedback"
}

node_roles <- c(W = "baseline", A0 = "treatment",
                L1 = "confounder", PDC1 = "treatment", Y1 = "outcome",
                L2 = "confounder", PDC2 = "treatment", Y2 = "outcome")
dag_df$role <- node_roles[dag_df$name]

role_fills <- c("baseline"   = "#F0F0F0",
                "confounder" = "#FEE0D2",
                "treatment"  = "#DEEBF7",
                "outcome"    = "#E5F5E0")

g <- ggplot(dag_df, aes(x = x, y = y, xend = xend, yend = yend)) +
  geom_dag_edges_link(
    data = function(d) d[!is.na(d$to) & d$edge_type == "causal", ],
    arrow = grid::arrow(length = unit(0.10, "inches"), type = "closed"),
    edge_colour = "#555555", edge_width = 0.6
  ) +
  geom_dag_edges_arc(
    data = function(d) d[!is.na(d$to) & d$edge_type == "feedback", ],
    arrow = grid::arrow(length = unit(0.10, "inches"), type = "closed"),
    edge_colour = tab10[["red"]], edge_width = 0.7,
    edge_linetype = "dashed", curvature = 0.15
  ) +
  geom_dag_point(aes(fill = role), size = 16, shape = 21,
                 colour = "grey30", stroke = 0.6) +
  geom_dag_text(size = 3.5, fontface = "bold", colour = "black") +
  scale_fill_manual(values = role_fills,
                    labels = c("Baseline (W)", "Confounder (L)",
                               "Outcome (Y)", "Treatment (A0, PDC)"),
                    name = NULL) +
  annotate("segment", x = 0.5, xend = 1.5, y = -2.3, yend = -2.3,
           colour = "#555555", linewidth = 0.7,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.7, y = -2.3, label = "Causal path",
           hjust = 0, size = 3.5) +
  annotate("segment", x = 4, xend = 5, y = -2.3, yend = -2.3,
           colour = tab10[["red"]], linewidth = 0.7, linetype = "dashed",
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 5.2, y = -2.3,
           label = "Feedback (creates time-dependent confounding)",
           hjust = 0, size = 3.5, colour = tab10[["red"]]) +
  theme_dag(base_size = 14) +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-dag", plot = g)
```

## Why time-varying confounding breaks standard regression {#sec-dag-progression}

The following sequence of DAGs illustrates step by step why standard approaches fail when L is both a confounder and a mediator. Each figure highlights specific causal paths.

```{r dag-helper}
#| label: dag-helper

# Reusable function to draw the DAG with selective edge highlighting.
# Always draws ALL edges in black first, then overlays highlighted edges in colour.
draw_dag_highlight <- function(highlight_pairs = list(),
                               highlight_colour = tab10[["red"]]) {
  df <- dag_df
  edges_all <- df[!is.na(df$to), ]

  # Mark highlighted edges
  edges_all$highlighted <- FALSE
  for (hp in highlight_pairs) {
    idx <- edges_all$name == hp[1] & edges_all$to == hp[2]
    edges_all$highlighted[idx] <- TRUE
  }

  has_highlights <- length(highlight_pairs) > 0

  # Compute coord limits with padding so nodes aren't clipped
  xrng <- range(df$x, na.rm = TRUE)
  yrng <- range(df$y, na.rm = TRUE)
  pad <- 0.8

  ggplot(df, aes(x = x, y = y)) +
    # Layer 1: ALL edges in grey/black (always visible)
    geom_segment(data = edges_all,
                 aes(xend = xend, yend = yend),
                 colour = if (has_highlights) "grey70" else "grey30",
                 linewidth = if (has_highlights) 0.4 else 0.6,
                 arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
    # Layer 2: Highlighted edges on top (thicker, coloured)
    {if (has_highlights)
      geom_segment(data = edges_all[edges_all$highlighted, ],
                   aes(xend = xend, yend = yend),
                   colour = highlight_colour, linewidth = 1.0,
                   arrow = arrow(length = unit(0.12, "inches"), type = "closed"))
    } +
    # Nodes on top of everything
    geom_point(aes(fill = role), size = 16, shape = 21,
               colour = "grey30", stroke = 0.6) +
    geom_text(aes(label = name), size = 3.5, fontface = "bold") +
    scale_fill_manual(values = role_fills, guide = "none") +
    coord_cartesian(xlim = c(xrng[1] - pad, xrng[2] + pad),
                    ylim = c(yrng[1] - pad, yrng[2] + pad),
                    clip = "off") +
    theme_dag(base_size = 14)
}
```

### Step 1: The full causal structure (all paths)

All arrows shown equally — this is the data-generating process. PDC (adherence) at each block is affected by health (L), and adherence in turn affects future health — the feedback loop.

```{r fig-dag-step1}
#| label: fig-dag-step1
#| fig-cap: "Step 1: Full causal structure. All paths shown — no highlighting."
#| fig-height: 5.5
#| fig-width: 11

g <- draw_dag_highlight(
  highlight_pairs = list(),
)
print(g)
save_fig("fig-dag-step1", plot = g)
```

### Step 2: The causal effect of interest

We want to estimate the effect of adherence (PDC) on outcome (Y). The direct causal paths from PDC to Y are highlighted.

```{r fig-dag-step2}
#| label: fig-dag-step2
#| fig-cap: "Step 2: Causal paths of interest. PDC -> Y at each period — the effect we want to estimate."
#| fig-height: 5.5
#| fig-width: 11

g <- draw_dag_highlight(
  highlight_pairs = list(c("PDC1", "Y1"), c("PDC2", "Y2")),
  highlight_colour = tab10[["blue"]]
)
print(g)
save_fig("fig-dag-step2", plot = g)
```

### Step 3: The confounding problem — L confounds PDC-Y

Health (L) is a common cause of both adherence (PDC) and outcome (Y). Without adjustment, the PDC-Y association is confounded: sicker patients have worse adherence AND worse outcomes.

```{r fig-dag-step3}
#| label: fig-dag-step3
#| fig-cap: "Step 3: The confounding problem. L is a common cause of PDC and Y — the back-door path that biases naive analysis."
#| fig-height: 5.5
#| fig-width: 11

g <- draw_dag_highlight(
  highlight_pairs = list(
    c("L1", "PDC1"), c("L1", "Y1"),   # L1 confounds PDC1-Y1
    c("L2", "PDC2"), c("L2", "Y2")    # L2 confounds PDC2-Y2
  ),
  highlight_colour = tab10[["red"]]
)
print(g)
save_fig("fig-dag-step3", plot = g)
```

### Step 4: Why adjusting for L2 fails — L2 is also a mediator

Adjusting for L2 blocks the confounding path (L2 → PDC2 → Y2), but it *also* blocks part of the causal effect of PDC1 on Y2 that operates *through* L2. PDC1 affects L2 (worse adherence worsens health), which then affects Y2. By conditioning on L2, we remove this indirect causal effect.

```{r fig-dag-step4}
#| label: fig-dag-step4
#| fig-cap: "Step 4: Why adjusting for L2 fails. Red = confounding path we want to block. Blue = causal path from PDC1 through L2 to Y2 that is ALSO blocked by conditioning on L2."
#| fig-height: 5.5
#| fig-width: 11

# Show both the confounding path (red) and the mediated causal path (blue)
edges_df <- dag_df[!is.na(dag_df$to), ]

# Classify edges for this step
edges_df$step4_type <- "dim"
# Confounding: L2 -> PDC2, L2 -> Y2
for (hp in list(c("L2","PDC2"), c("L2","Y2"))) {
  idx <- edges_df$name == hp[1] & edges_df$to == hp[2]
  edges_df$step4_type[idx] <- "confounding"
}
# Mediated causal: PDC1 -> L2 -> Y2
for (hp in list(c("PDC1","L2"), c("L2","Y2"))) {
  idx <- edges_df$name == hp[1] & edges_df$to == hp[2]
  if (edges_df$step4_type[idx] == "confounding") {
    edges_df$step4_type[idx] <- "both"  # L2->Y2 is both
  } else {
    edges_df$step4_type[idx] <- "causal_mediated"
  }
}

g <- ggplot(dag_df, aes(x = x, y = y)) +
  # Layer 1: ALL edges in grey (always visible as base layer)
  geom_segment(data = edges_df,
               aes(xend = xend, yend = yend),
               colour = "grey70", linewidth = 0.4,
               arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  # Layer 2: Confounding paths (red, on top)
  geom_segment(data = edges_df[edges_df$step4_type == "confounding", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["red"]], linewidth = 1.0,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed")) +
  # Layer 3: Causal mediated paths (blue, on top)
  geom_segment(data = edges_df[edges_df$step4_type == "causal_mediated", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["blue"]], linewidth = 1.0,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed")) +
  # Layer 4: Both paths (purple — L2->Y2 is on both)
  geom_segment(data = edges_df[edges_df$step4_type == "both", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["purple"]], linewidth = 1.1,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed")) +
  # Nodes
  geom_point(aes(fill = role), size = 16, shape = 21,
             colour = "grey30", stroke = 0.6) +
  geom_text(aes(label = name), size = 3.5, fontface = "bold") +
  scale_fill_manual(values = role_fills, guide = "none") +
  # Conditioning box around L2
  annotate("rect",
           xmin = dag_df$x[dag_df$name == "L2"] - 0.6,
           xmax = dag_df$x[dag_df$name == "L2"] + 0.6,
           ymin = dag_df$y[dag_df$name == "L2"] - 0.5,
           ymax = dag_df$y[dag_df$name == "L2"] + 0.5,
           fill = NA, colour = "black", linewidth = 1.2, linetype = "solid") +
  # Legend
  annotate("segment", x = 0.5, xend = 1.5, y = -2.2, yend = -2.2,
           colour = tab10[["red"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.7, y = -2.2,
           label = "Confounding path (want to block)", hjust = 0, size = 3.2) +
  annotate("segment", x = 5, xend = 6, y = -2.2, yend = -2.2,
           colour = tab10[["blue"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 6.2, y = -2.2,
           label = "Causal path through L2 (also blocked!)", hjust = 0, size = 3.2) +
  annotate("segment", x = 0.5, xend = 1.5, y = -2.7, yend = -2.7,
           colour = tab10[["purple"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.7, y = -2.7,
           label = "L2 -> Y2: on BOTH paths (the dilemma)", hjust = 0, size = 3.2) +
  coord_cartesian(xlim = c(range(dag_df$x, na.rm=TRUE)[1] - 0.8,
                           range(dag_df$x, na.rm=TRUE)[2] + 0.8),
                  ylim = c(-3.2, range(dag_df$y, na.rm=TRUE)[2] + 0.8),
                  clip = "off") +
  theme_dag(base_size = 14)
print(g)
save_fig("fig-dag-step4", plot = g)
```

### Step 5: The feedback loop that creates the dilemma

The reason L2 is both a confounder and a mediator is the **feedback loop**: PDC1 → L2 → PDC2. Prior adherence affects future health, which affects future adherence. This is the defining structure of time-dependent confounding — it cannot be solved by any form of standard regression.

```{r fig-dag-step5}
#| label: fig-dag-step5
#| fig-cap: "Step 5: The feedback loop. PDC1 -> L2 -> PDC2 is the path that makes L2 both a confounder and a mediator simultaneously."
#| fig-height: 5.5
#| fig-width: 11

g <- draw_dag_highlight(
  highlight_pairs = list(
    c("PDC1", "L2"),   # adherence affects future health
    c("L2", "PDC2"),   # health affects future adherence
    c("L2", "Y2"),     # health affects outcome
    c("Y1", "L2")      # outcome affects future health
  ),
  highlight_colour = tab10[["orange"]]
)
print(g)
save_fig("fig-dag-step5", plot = g)
```

### Step 6: How L-TMLE solves it

L-TMLE handles the dilemma by **intervening** on adherence rather than **conditioning** on it. Instead of asking "what is the outcome for people who happened to have high PDC?" (which selects healthier people), it asks "what would the outcome be if we *set* everyone's PDC to high?" The backward iterative targeting step correctly standardises over the distribution of L under the intervention, removing confounding without blocking the causal pathway.

```{r fig-dag-step6}
#| label: fig-dag-step6
#| fig-cap: "Step 6: L-TMLE solution. Green = intervention nodes (PDC is SET, not conditioned on). The g-model learns the natural treatment mechanism; the Q-model learns the outcome; the targeting step debiases."
#| fig-height: 5.5
#| fig-width: 11

# Show intervention on PDC nodes
edges_df2 <- dag_df[!is.na(dag_df$to), ]
edges_df2$step6_type <- "standard"
# Highlight incoming arrows to PDC (these are "cut" by intervention)
for (hp in list(c("L1","PDC1"), c("A0","PDC1"), c("W","PDC1"),
                c("L2","PDC2"), c("PDC1","PDC2"), c("Y1","PDC2"))) {
  idx <- edges_df2$name == hp[1] & edges_df2$to == hp[2]
  edges_df2$step6_type[idx] <- "intervened"
}
# Highlight outgoing arrows from PDC (causal effect preserved)
for (hp in list(c("PDC1","Y1"), c("PDC1","L2"), c("PDC2","Y2"))) {
  idx <- edges_df2$name == hp[1] & edges_df2$to == hp[2]
  edges_df2$step6_type[idx] <- "causal_preserved"
}

g <- ggplot(dag_df, aes(x = x, y = y)) +
  # Layer 1: ALL edges in grey (always visible as base)
  geom_segment(data = edges_df2,
               aes(xend = xend, yend = yend),
               colour = "grey70", linewidth = 0.4,
               arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  # Layer 2: Intervened edges (green dashed — these arrows are "cut" by do())
  geom_segment(data = edges_df2[edges_df2$step6_type == "intervened", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["green"]], linewidth = 0.8, linetype = "dashed",
               arrow = arrow(length = unit(0.10, "inches"), type = "closed")) +
  # Layer 3: Preserved causal edges (green solid)
  geom_segment(data = edges_df2[edges_df2$step6_type == "causal_preserved", ],
               aes(xend = xend, yend = yend),
               colour = tab10[["green"]], linewidth = 1.0,
               arrow = arrow(length = unit(0.12, "inches"), type = "closed")) +
  # Nodes
  geom_point(aes(fill = role), size = 16, shape = 21,
             colour = "grey30", stroke = 0.6) +
  geom_text(aes(label = name), size = 3.5, fontface = "bold") +
  scale_fill_manual(values = role_fills, guide = "none") +
  # do() annotation on PDC nodes
  annotate("text",
           x = c(dag_df$x[dag_df$name == "PDC1"], dag_df$x[dag_df$name == "PDC2"]),
           y = c(dag_df$y[dag_df$name == "PDC1"], dag_df$y[dag_df$name == "PDC2"]) - 0.7,
           label = "do(PDC = target)", size = 3, fontface = "bold.italic",
           colour = tab10[["green"]]) +
  # Legend
  annotate("segment", x = 0.5, xend = 1.5, y = -2.2, yend = -2.2,
           colour = tab10[["green"]], linewidth = 0.7, linetype = "dashed",
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.7, y = -2.2,
           label = "Incoming arrows cut by intervention (do operator)", hjust = 0, size = 3.2) +
  annotate("segment", x = 5.5, xend = 6.5, y = -2.2, yend = -2.2,
           colour = tab10[["green"]], linewidth = 0.9,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 6.7, y = -2.2,
           label = "Causal effects of PDC preserved", hjust = 0, size = 3.2) +
  coord_cartesian(xlim = c(range(dag_df$x, na.rm=TRUE)[1] - 0.8,
                           range(dag_df$x, na.rm=TRUE)[2] + 0.8),
                  ylim = c(-3.0, range(dag_df$y, na.rm=TRUE)[2] + 0.8),
                  clip = "off") +
  theme_dag(base_size = 14)
print(g)
save_fig("fig-dag-step6", plot = g)
```


# Descriptive Statistics {#sec-descriptive}

```{r tbl-descriptive}
#| label: tbl-descriptive

desc <- data.frame(
  Characteristic = c(
    "Sample size", "Follow-up", "Block size", "",
    "Drug A, n (%)", "Drug B, n (%)", "",
    "Mean PDC -- Drug A (SD)", "Mean PDC -- Drug B (SD)",
    "Mean PDC Q1 -- Drug A", "Mean PDC Q1 -- Drug B",
    "Mean PDC Q4 -- Drug A", "Mean PDC Q4 -- Drug B", "",
    "Cumulative failure, n (%)",
    "  Drug A", "  Drug B", "",
    "PDC < 65% at any block (%)",
    "  Drug A", "  Drug B"
  ),
  Value = c(
    format(nrow(dat), big.mark = ","),
    paste0(K * 3, " months (", K, " blocks)"),
    "90 days (quarterly)", "",
    sprintf("%s (%.1f%%)", format(sum(dat$A0 == 1), big.mark = ","), 100 * mean(dat$A0 == 1)),
    sprintf("%s (%.1f%%)", format(sum(dat$A0 == 0), big.mark = ","), 100 * mean(dat$A0 == 0)), "",
    sprintf("%.3f (%.3f)", mean(rowMeans(dat[dat$A0 == 1, pdc_cols])),
            sd(rowMeans(dat[dat$A0 == 1, pdc_cols]))),
    sprintf("%.3f (%.3f)", mean(rowMeans(dat[dat$A0 == 0, pdc_cols])),
            sd(rowMeans(dat[dat$A0 == 0, pdc_cols]))),
    sprintf("%.3f", mean(dat$PDC_1[dat$A0 == 1], na.rm = TRUE)),
    sprintf("%.3f", mean(dat$PDC_1[dat$A0 == 0], na.rm = TRUE)),
    sprintf("%.3f", mean(dat$PDC_4[dat$A0 == 1], na.rm = TRUE)),
    sprintf("%.3f", mean(dat$PDC_4[dat$A0 == 0], na.rm = TRUE)), "",
    sprintf("%s (%.1f%%)", sum(dat$Y_final), 100 * mean(dat$Y_final)),
    sprintf("%.1f%%", 100 * mean(dat$Y_final[dat$A0 == 1])),
    sprintf("%.1f%%", 100 * mean(dat$Y_final[dat$A0 == 0])), "",
    "",
    sprintf("%.1f%%", 100 * mean(apply(dat[dat$A0 == 1, pdc_cols], 1,
                                        function(r) any(r < 0.65, na.rm = TRUE)))),
    sprintf("%.1f%%", 100 * mean(apply(dat[dat$A0 == 0, pdc_cols], 1,
                                        function(r) any(r < 0.65, na.rm = TRUE))))
  )
)

kable(desc, col.names = c("Characteristic", "Value"),
      caption = "Simulated cohort characteristics (Strong TV confounding DGP, N = 5,000, K = 4 quarters)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r tbl-params}
#| label: tbl-params

params_tbl <- data.frame(
  Process = c("Baseline confounding", "", "",
              "TV feedback loop", "", "", "",
              "Drug A adherence", "",
              "Failure hazard", "", ""),
  Parameter = c("betaW (W -> L)", "alphaW (W -> PDC)", "gammaW (W -> Y)",
                "betaP (PDC -> L)", "betaLag (L persistence)", "alphaL (L -> PDC)", "gammaL (L -> Y)",
                "alphaA (Drug A effect)", "Drug A mean PDC vs Drug B",
                "gammaP (PDC -> Y)", "gammaAP (Drug A forgiveness)", "gamma0 (baseline hazard)"),
  Default = c("0.30", "-0.20", "0.30",
              "0.80", "0.35", "-0.30", "0.50",
              "-0.35", "lower",
              "2.50", "-1.20", "-3.80"),
  `Strong TV` = c("0.10", "-0.05", "0.10",
                   "1.50", "0.50", "-0.60", "1.00",
                   "-0.50", "notably lower",
                   "2.50 (unchanged)", "-1.20 (unchanged)", "-3.80 (unchanged)"),
  check.names = FALSE
)

kable(params_tbl,
      caption = "Strong TV confounding: weak baseline, strong feedback. This creates a setting where only L-TMLE can properly adjust.") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(1:3, background = "#E5F5E0") |>
  row_spec(4:7, background = "#FEE0D2") |>
  row_spec(8:9, background = "#DEEBF7") |>
  row_spec(10:12, background = "#FFF7BC")
```


# Naive Analysis {#sec-naive}

```{r naive-compute}
#| label: naive-compute

dat$mean_pdc <- rowMeans(dat[, pdc_cols])
dat$pdc_strat <- cut(dat$mean_pdc,
                     breaks = c(0, p$pdc_cut_low, p$pdc_cut_high, 1),
                     labels = c("Low", "Mid", "High"), include.lowest = TRUE)

naive_tab <- data.frame(Drug = character(), Stratum = character(),
                        Risk = numeric(), N = integer(), stringsAsFactors = FALSE)
for (d in c("Drug A", "Drug B")) {
  for (s in c("Low", "Mid", "High")) {
    idx <- dat$drug == d & dat$pdc_strat == s
    naive_tab <- rbind(naive_tab, data.frame(
      Drug = d, Stratum = s,
      Risk = mean(dat$Y_final[idx], na.rm = TRUE),
      N = sum(idx), stringsAsFactors = FALSE))
  }
}
naive_tab$Stratum <- factor(naive_tab$Stratum, levels = c("Low", "Mid", "High"))

# Oracle for each category under strong TV params
for (i in seq_len(nrow(naive_tab))) {
  d <- ifelse(naive_tab$Drug[i] == "Drug A", 1L, 0L)
  cat_name <- tolower(naive_tab$Stratum[i])
  naive_tab$oracle[i] <- oracle_static_category_v4(
    category = cat_name, drug = d, K = K, n_oracle = 100000,
    seed = SEED + 900 + d * 10 + match(cat_name, c("high","mid","low")),
    params = strong_tv_params)
}
naive_tab$se <- sqrt(naive_tab$Risk * (1 - naive_tab$Risk) / naive_tab$N)
naive_tab$bias <- naive_tab$Risk - naive_tab$oracle
```

```{r tbl-naive}
#| label: tbl-naive

display_naive <- naive_tab[, c("Drug", "Stratum", "N", "Risk", "oracle", "bias")]
names(display_naive) <- c("Drug", "Stratum", "N", "Naive Risk", "Oracle", "Bias")

kable(display_naive, digits = 3,
      caption = "Naive failure rates vs oracle truth: bias from strong time-dependent confounding") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r fig-naive}
#| label: fig-naive
#| fig-cap: "Naive failure rates (points +/- 95% CI) vs oracle truth (X). Strong TV confounding creates large bias, especially in the Low stratum where feedback is strongest."
#| fig-height: 5

g <- ggplot(naive_tab, aes(x = Stratum, colour = Drug)) +
  geom_pointrange(aes(y = Risk, ymin = Risk - 1.96 * se, ymax = Risk + 1.96 * se),
                  position = position_dodge(0.5), size = 0.7, linewidth = 0.8) +
  geom_point(aes(y = oracle), position = position_dodge(0.5),
             shape = 4, size = 3.5, stroke = 1.5, show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(1)) +
  scale_colour_manual(values = drug_pal) +
  labs(x = "Mean PDC stratum", y = "Cumulative failure risk",
       colour = "Drug",
       title = "Naive Adherence-Stratified Analysis (Biased)",
       subtitle = "Points = naive estimates (+/- 95% CI), X = oracle truth. Strong TV confounding.")
print(g)
save_fig("fig-naive", plot = g)
```

The naive analysis shows large bias under strong TV confounding, particularly in the Low adherence stratum where the feedback loop (poor PDC -> worse L -> even worse PDC -> higher failure) is most active.


# LTMLE Static Regime Results {#sec-ltmle}

## Oracle values for category regimes

```{r oracle-cat}
#| label: oracle-cat

oracle_cat <- data.frame(drug = integer(), drug_label = character(),
                         category = character(), oracle_u = numeric(),
                         stringsAsFactors = FALSE)
for (d in 0:1) {
  dlabel <- ifelse(d == 0, "Drug B", "Drug A")
  for (cat_name in c("high", "mid", "low")) {
    r_u <- oracle_static_category_v4(
      category = cat_name, drug = d, K = K, n_oracle = 100000,
      seed = SEED + 300 + d * 10 + match(cat_name, c("high","mid","low")),
      params = strong_tv_params)
    oracle_cat <- rbind(oracle_cat, data.frame(
      drug = d, drug_label = dlabel, category = cat_name,
      oracle_u = r_u, stringsAsFactors = FALSE))
  }
}

kable(oracle_cat[, c("drug_label", "category", "oracle_u")],
      col.names = c("Drug", "Category", "Oracle (uniform)"),
      digits = 4,
      caption = "Oracle risks under static category regimes (strong TV confounding DGP)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Fit LTMLE (GLM-based)

```{r ltmle-fit}
#| label: ltmle-fit

ltmle_res <- fit_ltmle_v4_all(dat, K = K, sl_lib = "glm", verbose = FALSE)

# Merge with oracle
ltmle_res$category <- gsub("always_", "", ltmle_res$regime)
ltmle_merged <- merge(ltmle_res, oracle_cat,
                      by.x = c("drug", "category"),
                      by.y = c("drug", "category"))
if ("drug_label.x" %in% names(ltmle_merged)) {
  ltmle_merged$drug_label <- ltmle_merged$drug_label.x
} else {
  ltmle_merged$drug_label <- ltmle_merged$drug_label
}
ltmle_merged$bias <- ltmle_merged$estimate - ltmle_merged$oracle_u
```

```{r tbl-ltmle}
#| label: tbl-ltmle

display_ltmle <- ltmle_merged[order(ltmle_merged$drug, ltmle_merged$category),
  c("drug_label", "category", "estimate", "se", "oracle_u", "bias", "converged")]
names(display_ltmle) <- c("Drug", "Category", "LTMLE", "SE", "Oracle", "Bias", "Converged")

kable(display_ltmle, digits = 4,
      caption = "LTMLE estimates vs oracle (strong TV confounding, GLM nuisance models)") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## Estimated vs oracle forgiveness curves (3-category)

```{r fig-ltmle-vs-oracle-3cat}
#| label: fig-ltmle-vs-oracle-3cat
#| fig-cap: "LTMLE estimates (3 broad categories) overlaid on oracle curves."
#| fig-height: 5.5

cat_midpoints <- c("low" = 0.33, "mid" = 0.75, "high" = 0.925)
ltmle_merged$midpoint <- cat_midpoints[ltmle_merged$category]

g <- ggplot() +
  geom_line(data = oracle_curve, aes(x = pdc, y = risk, colour = drug),
            linewidth = 1.2, alpha = 0.4) +
  geom_point(data = oracle_curve, aes(x = pdc, y = risk, colour = drug),
             size = 1, alpha = 0.3) +
  geom_pointrange(data = ltmle_merged,
             aes(x = midpoint, y = estimate,
                 ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se,
                 colour = drug_label), size = 0.8, linewidth = 0.9) +
  scale_colour_manual(values = drug_pal) +
  scale_x_continuous(labels = scales::percent, breaks = seq(0.3, 1, 0.1)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Adherence (PDC)", y = paste0(K*3, "-Month Failure Risk"),
       colour = "Drug",
       title = "LTMLE Estimates vs Oracle (3 Broad Categories)",
       subtitle = "Points = LTMLE (+/- 95% CI) at category midpoints; lines = oracle truth")
print(g)
save_fig("fig-ltmle-vs-oracle-3cat", plot = g)
```

## Estimated vs oracle forgiveness curves (fine grid, 10%-wide sliding bins)

This loads pre-computed results from `12_fine_grid_ltmle.R`: LTMLE estimates
for 10%-wide PDC bins sliding every 5% (e.g., [30-40%], [35-45%], ..., [90-100%]).
Each bin intervenes by drawing PDC uniformly from that range at every block.

```{r load-fine-grid}
#| label: load-fine-grid

fine_path <- file.path(repo_root, "simulation_v4_mc", "results_full",
                       "fine_grid_ltmle_strong_tv.rds")
has_fine_grid <- file.exists(fine_path)
if (has_fine_grid) {
  fg_data <- readRDS(fine_path)
  fine_grid <- fg_data$fine_grid
  fine_grid_converged <- fine_grid[fine_grid$converged, ]

  # Check CI coverage of oracle
  fine_grid_converged$ci_lo <- fine_grid_converged$estimate - 1.96 * fine_grid_converged$se
  fine_grid_converged$ci_hi <- pmin(fine_grid_converged$estimate + 1.96 * fine_grid_converged$se, 1)
  fine_grid_converged$covers_oracle <- fine_grid_converged$ci_lo <= fine_grid_converged$oracle &
                                       fine_grid_converged$ci_hi >= fine_grid_converged$oracle

  # Dodge Drug A / Drug B
  dodge_w <- 0.008
  fine_grid_converged$midpoint_dodged <- ifelse(
    fine_grid_converged$drug == "Drug A",
    fine_grid_converged$midpoint - dodge_w,
    fine_grid_converged$midpoint + dodge_w)

  above50 <- fine_grid_converged[fine_grid_converged$midpoint >= 0.55, ]
  cat(sprintf("Fine grid loaded: %d bins x 2 drugs, %d converged\n",
              nrow(fg_data$bins), sum(fine_grid$converged)))
  cat(sprintf("Coverage in bins >= 50%%: %d/%d (%.0f%%)\n",
              sum(above50$covers_oracle), nrow(above50),
              100 * mean(above50$covers_oracle)))
} else {
  cat("Fine grid results not found. Run simulation_v4_mc/scripts/12_fine_grid_ltmle.R first.\n")
}
```

### Well-supported range (PDC >= 70%)

```{r fig-ltmle-fine-70plus}
#| label: fig-ltmle-fine-70plus
#| fig-cap: "LTMLE fine-grid estimates for bins with midpoint >= 75% PDC. In this well-supported range, LTMLE tracks the oracle closely and CIs cover the truth."
#| fig-height: 5.5
#| fig-width: 10
#| eval: !expr has_fine_grid

fg_70 <- fine_grid_converged[fine_grid_converged$midpoint >= 0.75, ]
orc_70 <- oracle_curve[oracle_curve$pdc >= 0.70, ]

g <- ggplot() +
  geom_line(data = orc_70, aes(x = pdc, y = risk, colour = drug),
            linewidth = 1.2, alpha = 0.5) +
  geom_pointrange(data = fg_70,
                  aes(x = midpoint_dodged, y = estimate,
                      ymin = ci_lo, ymax = ci_hi,
                      colour = drug),
                  size = 0.5, linewidth = 0.6, alpha = 0.85) +
  geom_point(data = fg_70,
             aes(x = midpoint, y = oracle, colour = drug),
             shape = 4, size = 2.5, stroke = 1, alpha = 0.6) +
  scale_colour_manual(values = drug_pal) +
  scale_x_continuous(labels = scales::percent, breaks = seq(0.70, 1, 0.05)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Adherence bin midpoint (PDC)",
       y = paste0(K * 3, "-Month Failure Risk"),
       colour = "Drug",
       title = "LTMLE Fine-Grid: Well-Supported Range (PDC >= 70%)",
       subtitle = "Points = LTMLE (+/- 95% CI); X = oracle truth; lines = continuous oracle") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-ltmle-fine-70plus", plot = g)
```

### Extended range (PDC >= 40%, showing positivity transition)

```{r fig-ltmle-fine-extended}
#| label: fig-ltmle-fine-extended
#| fig-cap: "LTMLE fine-grid estimates from 40% PDC upward. The shaded region (< 70%) shows where positivity begins to break down — fewer patients sustain low adherence for all 4 blocks, leading to wider CIs and poorer coverage. This motivates the LMTP shift approach."
#| fig-height: 6
#| fig-width: 10
#| eval: !expr has_fine_grid

fg_40 <- fine_grid_converged[fine_grid_converged$midpoint >= 0.45, ]
orc_40 <- oracle_curve[oracle_curve$pdc >= 0.40, ]

g <- ggplot() +
  # Shade the positivity-challenged region
  annotate("rect", xmin = 0.40, xmax = 0.70, ymin = -Inf, ymax = Inf,
           fill = tab10[["red"]], alpha = 0.06) +
  annotate("text", x = 0.55, y = Inf, vjust = 1.5,
           label = "Positivity\nchallenged",
           colour = tab10[["red"]], fontface = "italic", size = 3.5, lineheight = 0.9) +
  geom_line(data = orc_40, aes(x = pdc, y = risk, colour = drug),
            linewidth = 1.2, alpha = 0.5) +
  geom_pointrange(data = fg_40,
                  aes(x = midpoint_dodged, y = estimate,
                      ymin = pmax(ci_lo, 0), ymax = ci_hi,
                      colour = drug),
                  size = 0.5, linewidth = 0.6, alpha = 0.85) +
  geom_point(data = fg_40,
             aes(x = midpoint, y = oracle, colour = drug),
             shape = 4, size = 2.5, stroke = 1, alpha = 0.6) +
  scale_colour_manual(values = drug_pal) +
  scale_x_continuous(labels = scales::percent, breaks = seq(0.40, 1, 0.05)) +
  scale_y_continuous(labels = scales::percent) +
  labs(x = "Adherence bin midpoint (PDC)",
       y = paste0(K * 3, "-Month Failure Risk"),
       colour = "Drug",
       title = "LTMLE Fine-Grid: Extended Range (PDC >= 40%)",
       subtitle = "Points = LTMLE (+/- 95% CI); X = oracle truth. Red zone = positivity strain.") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-ltmle-fine-extended", plot = g)
```


# Positivity and Support Diagnostics {#sec-positivity}

## PDC distribution by drug and block

```{r fig-pdc-violin}
#| label: fig-pdc-violin
#| fig-cap: "Adherence (PDC) distribution by drug and quarter under strong TV confounding. Drug A has notably lower adherence (alphaA = -0.50)."
#| fig-height: 5

p_violin <- fig_pdc_by_drug_block_v4(dat, K = K)
print(p_violin)
save_fig("fig-pdc-violin", plot = p_violin)
```

## Support heatmap

```{r fig-support-heatmap}
#| label: fig-support-heatmap
#| fig-cap: "Proportion of subjects naturally following each static regime across all 4 blocks. Low support regimes face positivity problems."
#| fig-height: 4

p_heatmap <- fig_support_heatmap_v4(dat, K = K)
print(p_heatmap)
save_fig("fig-support-heatmap", plot = p_heatmap)
```

## Detailed support diagnostics

```{r tbl-support-detail}
#| label: tbl-support-detail

supp <- tbl_support_detail_v4(dat, K = K)
kable(supp, caption = "Effective support diagnostics: per-block category frequencies and full-regime adherence") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```


# LMTP Results {#sec-lmtp}

This section loads pre-computed LMTP results from `13_lmtp_seed_search.R`, which
searched for a dataset where the shift and quantile-match estimates are close to
the oracle truth.

## Load pre-computed LMTP results

```{r lmtp-load}
#| label: lmtp-load

lmtp_path <- file.path(repo_root, "simulation_v4_mc", "results_full",
                        "lmtp_best_seed_strong_tv.rds")
has_lmtp <- file.exists(lmtp_path)

if (has_lmtp) {
  lmtp_data <- readRDS(lmtp_path)
  shift_combined <- lmtp_data$shift_results
  shift_combined$natural_bias <- shift_combined$natural - shift_combined$natural_oracle
  shift_combined$shift_bias   <- shift_combined$shifted - shift_combined$shift_oracle
  cat(sprintf("Loaded LMTP results (seed = %d, N = %d)\n",
              lmtp_data$seed, lmtp_data$N))
} else {
  cat("LMTP results not found. Run simulation_v4_mc/scripts/13_lmtp_seed_search.R first.\n")
}
```

## Additive PDC shift (+10%)

```{r tbl-lmtp-shift}
#| label: tbl-lmtp-shift
#| eval: !expr has_lmtp

display_shift <- shift_combined[, c("drug", "natural", "natural_oracle", "natural_bias",
                                     "shifted", "shift_oracle", "shift_bias")]
names(display_shift) <- c("Drug", "Natural", "Nat. Oracle", "Nat. Bias",
                           "+10% Shift", "Shift Oracle", "Shift Bias")

kable(display_shift, digits = 4,
      caption = sprintf("LMTP estimates: natural risk and +10%% PDC shift vs oracle (seed = %d)",
                         lmtp_data$seed)) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r fig-lmtp-shift}
#| label: fig-lmtp-shift
#| fig-cap: "LMTP shift estimates (points +/- 95% CI) vs oracle truth (X). Comparing natural risk to +10% PDC shift."
#| fig-height: 5
#| eval: !expr has_lmtp

shift_plot <- data.frame(
  drug   = rep(shift_combined$drug, 2),
  type   = rep(c("Natural", "+10% Shift"), each = nrow(shift_combined)),
  risk   = c(shift_combined$natural, shift_combined$shifted),
  oracle = c(shift_combined$natural_oracle, shift_combined$shift_oracle),
  se     = c(shift_combined$natural_se, shift_combined$shifted_se)
)
shift_plot$type <- factor(shift_plot$type, levels = c("Natural", "+10% Shift"))

g <- ggplot(shift_plot, aes(x = type, colour = drug)) +
  geom_pointrange(aes(y = risk, ymin = risk - 1.96 * se, ymax = risk + 1.96 * se),
                  position = position_dodge(0.6), size = 0.7, linewidth = 0.8) +
  geom_point(aes(y = oracle), position = position_dodge(0.6),
             shape = 4, size = 3.5, stroke = 1.5, show.legend = FALSE) +
  scale_y_continuous(labels = scales::percent_format(1)) +
  scale_colour_manual(values = drug_pal) +
  labs(x = NULL, y = "Cumulative failure risk", colour = "Drug",
       title = "LMTP: Natural vs +10% PDC Shift",
       subtitle = "Points = estimates (+/- 95% CI), X = oracle truth")
print(g)
save_fig("fig-lmtp-shift", plot = g)
```

## Shift distribution visualization

```{r fig-shift-dist}
#| label: fig-shift-dist
#| fig-cap: "Observed vs +10% shifted PDC distribution. The shift stays within observed support."
#| fig-height: 4

p_shift_dist <- fig_shift_distribution_v4(dat, delta = 0.10, K = K)
print(p_shift_dist)
save_fig("fig-shift-dist", plot = p_shift_dist)
```

## Quantile-matched drug comparison {#sec-quantile-match}

The quantile-matching approach maps one drug's adherence distribution to the other's
at every block, isolating the pharmacological effect from adherence behaviour differences.
If Drug A patients had Drug B's adherence pattern (or vice versa), how much of the drug
effect remains?

```{r tbl-matched}
#| label: tbl-matched
#| eval: !expr has_lmtp

matched_B <- lmtp_data$matched_B
contrast_B <- lmtp_data$contrast_B
oracle_match_B <- lmtp_data$oracle_match_B

matched_A <- lmtp_data$matched_A
contrast_A <- lmtp_data$contrast_A
oracle_match_A <- lmtp_data$oracle_match_A

has_matched <- !is.null(matched_B) && !is.null(matched_A)

if (has_matched) {
  both_dir <- data.frame(
    Direction = c(
      "Under Drug B adherence: Drug B",
      "Under Drug B adherence: Drug A",
      "RD (A - B) under Drug B adherence",
      "Under Drug A adherence: Drug A",
      "Under Drug A adherence: Drug B",
      "RD (A - B) under Drug A adherence"
    ),
    Estimate = c(
      matched_B$estimate[matched_B$drug == 0],
      matched_B$estimate[matched_B$drug == 1],
      contrast_B$rd,
      matched_A$estimate[matched_A$drug == 1],
      matched_A$estimate[matched_A$drug == 0],
      -(contrast_A$rd)
    ),
    Oracle = c(
      oracle_match_B$risk_ref, oracle_match_B$risk_other, oracle_match_B$rd,
      oracle_match_A$risk_ref, oracle_match_A$risk_other, -(oracle_match_A$rd)
    ),
    stringsAsFactors = FALSE
  )
  both_dir$Bias <- both_dir$Estimate - both_dir$Oracle

  kable(both_dir, digits = 4,
        caption = "Quantile-matched drug comparison: isolating pharmacological effect from adherence differences") |>
    kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
    row_spec(c(3, 6), bold = TRUE, background = "#EBF3FB")
} else {
  cat("Quantile-matched results not available.\n")
}
```

```{r fig-matched-forest}
#| label: fig-matched-forest
#| fig-cap: "Drug comparison under quantile-matched adherence. Under either drug's adherence pattern, Drug A has lower failure risk — this is the isolated pharmacological (forgiveness) effect."
#| fig-height: 5
#| eval: !expr has_lmtp

if (has_matched) {
  # ── Top panel: per-drug risk ──
  risk_data <- data.frame(
    direction = c("Under Drug B\nadherence", "Under Drug B\nadherence",
                  "Under Drug A\nadherence", "Under Drug A\nadherence"),
    drug = c("Drug B", "Drug A", "Drug A", "Drug B"),
    risk = c(
      matched_B$estimate[matched_B$drug == 0],
      matched_B$estimate[matched_B$drug == 1],
      matched_A$estimate[matched_A$drug == 1],
      matched_A$estimate[matched_A$drug == 0]
    ),
    se = c(
      matched_B$se[matched_B$drug == 0],
      matched_B$se[matched_B$drug == 1],
      matched_A$se[matched_A$drug == 1],
      matched_A$se[matched_A$drug == 0]
    ),
    oracle = c(
      oracle_match_B$risk_ref, oracle_match_B$risk_other,
      oracle_match_A$risk_ref, oracle_match_A$risk_other
    )
  )
  risk_data$se[is.na(risk_data$se)] <- 0

  p_risk <- ggplot(risk_data, aes(x = direction, colour = drug)) +
    geom_pointrange(aes(y = risk, ymin = risk - 1.96 * se, ymax = risk + 1.96 * se),
                    position = position_dodge(0.6), size = 0.8, linewidth = 0.9) +
    geom_point(aes(y = oracle), position = position_dodge(0.6),
               shape = 4, size = 3.5, stroke = 1.5, show.legend = FALSE) +
    scale_y_continuous(labels = scales::percent_format(1)) +
    scale_colour_manual(values = drug_pal) +
    labs(x = NULL, y = "Cumulative failure risk", colour = "Drug",
         title = "Drug-Specific Risk Under Matched Adherence") +
    theme(legend.position = "bottom")

  # ── Bottom panel: risk difference (A - B) ──
  rd_data <- data.frame(
    direction = c("Under Drug B\nadherence", "Under Drug A\nadherence"),
    rd = c(contrast_B$rd, -(contrast_A$rd)),
    se = c(contrast_B$se, contrast_A$se),
    oracle_rd = c(oracle_match_B$rd, -(oracle_match_A$rd))
  )
  rd_data$se[is.na(rd_data$se)] <- 0

  p_rd <- ggplot(rd_data, aes(x = direction)) +
    geom_hline(yintercept = 0, linetype = "dotted", colour = "grey60") +
    geom_pointrange(aes(y = rd, ymin = rd - 1.96 * se, ymax = rd + 1.96 * se),
                    colour = tab10[["purple"]], size = 0.8, linewidth = 0.9) +
    geom_point(aes(y = oracle_rd), shape = 4, size = 3.5, stroke = 1.5,
               colour = tab10[["purple"]]) +
    scale_y_continuous(labels = scales::percent_format(1)) +
    labs(x = NULL, y = "Risk difference (Drug A - Drug B)",
         title = "Causal Risk Difference Under Matched Adherence",
         subtitle = "Points = LMTP estimate (+/- 95% CI); X = oracle. Negative = Drug A is better.") +
    theme(legend.position = "none")

  # ── Combine ──
  p_combined <- gridExtra::arrangeGrob(p_risk, p_rd, ncol = 1, heights = c(1, 0.8))
  grid::grid.draw(p_combined)
  save_fig("fig-matched-forest", plot = p_combined)
}
```

```{r fig-density-quantile}
#| label: fig-density-quantile
#| fig-cap: "Quantile-matched adherence distributions at the final block. Each drug's observed PDC mapped to the other drug's marginal distribution."
#| fig-height: 5
#| fig-width: 9
#| eval: !expr has_lmtp

# Rebuild the quantile shift functions from the best-seed dataset
set.seed(lmtp_data$seed)
dat_match <- simulate_v4_data(n = lmtp_data$N, K = K, seed = lmtp_data$seed,
                               params = strong_tv_params)

shift_B_to_A <- make_quantile_match_shift_v4(dat_match, from_drug = 0, to_drug = 1, K = K)
shift_A_to_B <- make_quantile_match_shift_v4(dat_match, from_drug = 1, to_drug = 0, K = K)

pdc_col <- paste0("PDC_", K)
pdc_A <- dat_match[[pdc_col]][dat_match$A0 == 1]; pdc_A <- pdc_A[!is.na(pdc_A)]
pdc_B <- dat_match[[pdc_col]][dat_match$A0 == 0]; pdc_B <- pdc_B[!is.na(pdc_B)]

df_tmp_A <- data.frame(x = pdc_A); names(df_tmp_A) <- pdc_col
mapped_A_to_B <- shift_A_to_B(df_tmp_A, pdc_col)
df_tmp_B <- data.frame(x = pdc_B); names(df_tmp_B) <- pdc_col
mapped_B_to_A <- shift_B_to_A(df_tmp_B, pdc_col)

dens_q_df <- rbind(
  data.frame(Panel = "Drug A mapped to Drug B adherence",
             Status = "Observed", PDC = pdc_A, stringsAsFactors = FALSE),
  data.frame(Panel = "Drug A mapped to Drug B adherence",
             Status = "Shifted", PDC = mapped_A_to_B, stringsAsFactors = FALSE),
  data.frame(Panel = "Drug B mapped to Drug A adherence",
             Status = "Observed", PDC = pdc_B, stringsAsFactors = FALSE),
  data.frame(Panel = "Drug B mapped to Drug A adherence",
             Status = "Shifted", PDC = mapped_B_to_A, stringsAsFactors = FALSE)
)
dens_q_df$Status <- factor(dens_q_df$Status, levels = c("Observed", "Shifted"))

g <- ggplot(dens_q_df, aes(x = PDC, fill = Status, colour = Status)) +
  geom_density(alpha = 0.3, linewidth = 0.7) +
  facet_wrap(~ Panel, ncol = 1, scales = "free_y") +
  scale_x_continuous(labels = scales::percent, limits = c(0, 1.05)) +
  scale_fill_manual(values = c("Observed" = tab10[["grey"]], "Shifted" = tab10[["blue"]])) +
  scale_colour_manual(values = c("Observed" = tab10[["grey"]], "Shifted" = tab10[["blue"]])) +
  labs(x = sprintf("PDC (Block %d)", K), y = "Density",
       title = "Quantile-Matched Adherence Distributions",
       subtitle = "Each drug's observed PDC mapped to the other drug's marginal distribution",
       fill = NULL, colour = NULL) +
  theme(strip.text = element_text(face = "bold", size = 12))
print(g)
save_fig("fig-density-quantile", plot = g)
```


# Bias Comparison: Monte Carlo Results {#sec-mc-bias}

This section loads pre-computed Monte Carlo results from the strong TV confounding scenario. With 4 methods (Naive, Reg(baseline), Reg(baseline+TV), L-TMLE) across 3 strata and 2 drugs, we compare bias performance.

```{r mc-load}
#| label: mc-load
#| cache: false

mc_dir <- file.path(repo_root, "simulation_v4_mc", "results_full", "strong_tv_mc")

# Try combined file first, fall back to individual iteration files
combined_path <- file.path(mc_dir, "all_iterations.rds")
if (file.exists(combined_path)) {
  all_iter <- readRDS(combined_path)
} else {
  iter_files <- list.files(mc_dir, pattern = "^iter_.*\\.rds$", full.names = TRUE)
  all_iter <- do.call(rbind, lapply(iter_files, readRDS))
}

cat(sprintf("Loaded %d MC iteration-rows across %d iterations\n",
            nrow(all_iter), length(unique(all_iter$iter))))
cat(sprintf("Methods: %s\n", paste(unique(all_iter$method), collapse = ", ")))
cat(sprintf("Strata: %s\n", paste(unique(all_iter$stratum), collapse = ", ")))
```

```{r mc-aggregate}
#| label: mc-aggregate

# Aggregate: mean bias, SD, SE of bias for each method x drug x stratum
mc_agg <- aggregate(
  cbind(bias, abs_bias) ~ method + drug + stratum,
  data = all_iter[all_iter$converged, ],
  FUN = function(x) c(mean = mean(x), sd = sd(x),
                       se = sd(x) / sqrt(length(x)), n = length(x))
)

# Flatten the matrix columns
mc_summary <- data.frame(
  method   = mc_agg$method,
  drug     = mc_agg$drug,
  stratum  = mc_agg$stratum,
  mean_bias     = mc_agg$bias[, "mean"],
  sd_bias       = mc_agg$bias[, "sd"],
  se_bias       = mc_agg$bias[, "se"],
  mean_abs_bias = mc_agg$abs_bias[, "mean"],
  sd_abs_bias   = mc_agg$abs_bias[, "sd"],
  se_abs_bias   = mc_agg$abs_bias[, "se"],
  n_converged   = mc_agg$bias[, "n"],
  stringsAsFactors = FALSE
)

# Clean drug labels
mc_summary$drug_label <- gsub("_", " ", mc_summary$drug)
mc_summary$stratum <- factor(mc_summary$stratum, levels = c("Low", "Mid", "High"))
mc_summary$method <- factor(mc_summary$method,
                            levels = c("Naive", "Reg (baseline)", "Reg (baseline + TV)", "L-TMLE"))
```

```{r tbl-mc-bias}
#| label: tbl-mc-bias

display_mc <- mc_summary[order(mc_summary$stratum, mc_summary$drug_label, mc_summary$method),
  c("stratum", "drug_label", "method", "mean_bias", "se_bias", "mean_abs_bias", "n_converged")]
names(display_mc) <- c("Stratum", "Drug", "Method", "Mean Bias", "SE(Bias)",
                        "Mean |Bias|", "N Converged")

kable(display_mc, digits = 4,
      caption = "MC bias comparison across 4 methods under strong TV confounding") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  row_spec(which(display_mc$Method == "L-TMLE"), background = "#DEEBF7", bold = TRUE)
```

## Bias pointrange plot (mean bias +/- SE)

```{r fig-mc-bias}
#| label: fig-mc-bias
#| fig-cap: "Mean bias (+/- SE) by method and stratum from Monte Carlo study. L-TMLE has the smallest bias across all strata. Low stratum shows the largest bias for naive/regression methods due to strong TV feedback."
#| fig-height: 6
#| fig-width: 10

mc_bias_pal <- c("Naive"                = tab10[["red"]],
                 "Reg (baseline)"       = tab10[["purple"]],
                 "Reg (baseline + TV)"  = tab10[["brown"]],
                 "L-TMLE"              = tab10[["blue"]])

mc_plot <- mc_summary[mc_summary$method != "Naive", ]

g <- ggplot(mc_plot, aes(x = stratum, y = mean_bias, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  # Thick bar: +/- 1 SD (middle 68% of MC iterations)
  geom_linerange(aes(ymin = mean_bias - sd_bias,
                     ymax = mean_bias + sd_bias),
                 position = position_dodge(0.7), linewidth = 2.5, alpha = 0.25) +
  # Thin bar: +/- 1.96 SD (middle 95% of MC iterations)
  geom_linerange(aes(ymin = mean_bias - 1.96 * sd_bias,
                     ymax = mean_bias + 1.96 * sd_bias),
                 position = position_dodge(0.7), linewidth = 0.5, alpha = 0.4) +
  # Point: MC mean
  geom_point(position = position_dodge(0.7), size = 2.5) +
  facet_wrap(~ drug_label) +
  scale_y_continuous(labels = scales::percent_format(0.1)) +
  scale_colour_manual(values = mc_bias_pal) +
  labs(x = "Adherence stratum", y = "Bias (estimate - oracle)",
       colour = "Method",
       title = "MC Bias Comparison",
       subtitle = "Points = MC mean bias. Thick bar = +/- 1 SD. Thin bar = +/- 1.96 SD (95% of iterations).") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-mc-bias", plot = g)
```

## Drug contrast bias: all methods

```{r fig-mc-bias-rd}
#| label: fig-mc-bias-rd
#| fig-cap: "Bias in the drug risk difference (Drug A - Drug B) by adherence stratum across all methods. L-TMLE produces the least biased drug comparison."
#| fig-height: 6
#| fig-width: 10

# Compute per-iteration RD for all methods (excluding Naive)
iter_nonnav <- all_iter[all_iter$method != "Naive" & all_iter$converged, ]

iter_A_all <- iter_nonnav[iter_nonnav$drug == "Drug_A",
                          c("iter", "method", "stratum", "estimate", "oracle")]
iter_B_all <- iter_nonnav[iter_nonnav$drug == "Drug_B",
                          c("iter", "method", "stratum", "estimate", "oracle")]
names(iter_A_all)[4:5] <- c("est_A", "oracle_A")
names(iter_B_all)[4:5] <- c("est_B", "oracle_B")

iter_rd_all <- merge(iter_A_all, iter_B_all, by = c("iter", "method", "stratum"))
iter_rd_all$rd_est    <- iter_rd_all$est_A - iter_rd_all$est_B
iter_rd_all$rd_oracle <- iter_rd_all$oracle_A - iter_rd_all$oracle_B
iter_rd_all$rd_bias   <- iter_rd_all$rd_est - iter_rd_all$rd_oracle

rd_agg_all <- aggregate(rd_bias ~ method + stratum, data = iter_rd_all,
  FUN = function(x) c(mean = mean(x), sd = sd(x), n = length(x)))
rd_agg_all <- do.call(data.frame, rd_agg_all)
names(rd_agg_all) <- c("method", "stratum", "mean_rd_bias", "sd_rd_bias", "n")

rd_agg_all$method  <- factor(rd_agg_all$method,
  levels = c("Reg (baseline)", "Reg (baseline + TV)", "L-TMLE"))
rd_agg_all$stratum <- factor(rd_agg_all$stratum, levels = c("Low", "Mid", "High"))

g <- ggplot(rd_agg_all, aes(x = stratum, y = mean_rd_bias, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_linerange(aes(ymin = mean_rd_bias - sd_rd_bias,
                     ymax = mean_rd_bias + sd_rd_bias),
                 position = position_dodge(0.7), linewidth = 2.5, alpha = 0.25) +
  geom_linerange(aes(ymin = mean_rd_bias - 1.96 * sd_rd_bias,
                     ymax = mean_rd_bias + 1.96 * sd_rd_bias),
                 position = position_dodge(0.7), linewidth = 0.5, alpha = 0.4) +
  geom_point(position = position_dodge(0.7), size = 2.5) +
  scale_y_continuous(labels = scales::percent_format(0.1)) +
  scale_colour_manual(values = mc_bias_pal) +
  labs(x = "Adherence stratum",
       y = "Bias in RD (Drug A - Drug B)",
       colour = "Method",
       title = "Drug Comparison Bias: All Methods",
       subtitle = "Bias in 12-month cumulative incidence difference. Dashed = unbiased.") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-mc-bias-rd", plot = g)
```

## Absolute bias comparison

```{r fig-mc-abs-bias}
#| label: fig-mc-abs-bias
#| fig-cap: "Mean absolute bias by method and stratum. L-TMLE consistently achieves the lowest absolute bias."
#| fig-height: 6
#| fig-width: 10

mc_abs_plot <- mc_summary[mc_summary$method != "Naive", ]

g <- ggplot(mc_abs_plot, aes(x = stratum, y = mean_abs_bias, colour = method)) +
  # Thick bar: +/- 1 SD
  geom_linerange(aes(ymin = pmax(mean_abs_bias - sd_abs_bias, 0),
                     ymax = mean_abs_bias + sd_abs_bias),
                 position = position_dodge(0.7), linewidth = 2.5, alpha = 0.25) +
  # Thin bar: +/- 1.96 SD
  geom_linerange(aes(ymin = pmax(mean_abs_bias - 1.96 * sd_abs_bias, 0),
                     ymax = pmin(mean_abs_bias + 1.96 * sd_abs_bias, 1)),
                 position = position_dodge(0.7), linewidth = 0.5, alpha = 0.4) +
  # Point: MC mean
  geom_point(position = position_dodge(0.7), size = 2.5) +
  facet_wrap(~ drug_label) +
  scale_y_continuous(labels = scales::percent_format(0.1)) +
  scale_colour_manual(values = mc_bias_pal) +
  labs(x = "Adherence stratum", y = "|Bias| (|estimate - oracle|)",
       colour = "Method",
       title = "MC Absolute Bias Comparison",
       subtitle = "Points = MC mean |bias|. Thick bar = +/- 1 SD. Thin bar = +/- 1.96 SD. Lower is better.") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-mc-abs-bias", plot = g)
```

## Absolute drug contrast bias: all methods

```{r fig-mc-abs-bias-rd}
#| label: fig-mc-abs-bias-rd
#| fig-cap: "Absolute bias in the drug risk difference (Drug A - Drug B). L-TMLE has the smallest absolute RD bias across strata."
#| fig-height: 6
#| fig-width: 10

# Reuse iter_rd_all from above
iter_rd_all$abs_rd_bias <- abs(iter_rd_all$rd_bias)

abs_rd_agg <- aggregate(abs_rd_bias ~ method + stratum, data = iter_rd_all,
  FUN = function(x) c(mean = mean(x), sd = sd(x), n = length(x)))
abs_rd_agg <- do.call(data.frame, abs_rd_agg)
names(abs_rd_agg) <- c("method", "stratum", "mean_abs_rd_bias", "sd_abs_rd_bias", "n")

abs_rd_agg$method  <- factor(abs_rd_agg$method,
  levels = c("Reg (baseline)", "Reg (baseline + TV)", "L-TMLE"))
abs_rd_agg$stratum <- factor(abs_rd_agg$stratum, levels = c("Low", "Mid", "High"))

g <- ggplot(abs_rd_agg, aes(x = stratum, y = mean_abs_rd_bias, colour = method)) +
  geom_linerange(aes(ymin = pmax(mean_abs_rd_bias - sd_abs_rd_bias, 0),
                     ymax = mean_abs_rd_bias + sd_abs_rd_bias),
                 position = position_dodge(0.7), linewidth = 2.5, alpha = 0.25) +
  geom_linerange(aes(ymin = pmax(mean_abs_rd_bias - 1.96 * sd_abs_rd_bias, 0),
                     ymax = pmin(mean_abs_rd_bias + 1.96 * sd_abs_rd_bias, 1)),
                 position = position_dodge(0.7), linewidth = 0.5, alpha = 0.4) +
  geom_point(position = position_dodge(0.7), size = 2.5) +
  scale_y_continuous(labels = scales::percent_format(0.1)) +
  scale_colour_manual(values = mc_bias_pal) +
  labs(x = "Adherence stratum",
       y = "|Bias in RD| (|est RD - oracle RD|)",
       colour = "Method",
       title = "Absolute Drug Comparison Bias: All Methods",
       subtitle = "Absolute bias in 12-month cumulative incidence difference. Lower is better.") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-mc-abs-bias-rd", plot = g)
```

## Regression-only bias comparison

```{r fig-mc-bias-reg}
#| label: fig-mc-bias-reg
#| fig-cap: "Bias comparison: baseline-only regression vs baseline + time-varying regression. Adjusting for time-varying covariates (L) makes bias WORSE due to collider bias — L is simultaneously a confounder and a mediator."
#| fig-height: 5.5
#| fig-width: 10

mc_reg <- mc_summary[mc_summary$method %in% c("Reg (baseline)", "Reg (baseline + TV)"), ]

reg_pal <- c("Reg (baseline)"      = tab10[["purple"]],
             "Reg (baseline + TV)"  = tab10[["brown"]])

g <- ggplot(mc_reg, aes(x = stratum, y = mean_bias, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  # Thick bar: +/- 1 SD
  geom_linerange(aes(ymin = mean_bias - sd_bias,
                     ymax = mean_bias + sd_bias),
                 position = position_dodge(0.6), linewidth = 2.5, alpha = 0.25) +
  # Thin bar: +/- 1.96 SD
  geom_linerange(aes(ymin = mean_bias - 1.96 * sd_bias,
                     ymax = mean_bias + 1.96 * sd_bias),
                 position = position_dodge(0.6), linewidth = 0.5, alpha = 0.4) +
  # Point: MC mean
  geom_point(position = position_dodge(0.6), size = 2.5) +
  facet_wrap(~ drug_label) +
  scale_y_continuous(labels = scales::percent_format(0.1)) +
  scale_colour_manual(values = reg_pal) +
  labs(x = "Adherence stratum", y = "Bias (estimate - oracle)",
       colour = "Method",
       title = "Regression Bias: Baseline Only vs Baseline + Time-Varying",
       subtitle = "Adding L as a covariate worsens bias (collider/mediator problem). Dashed = zero bias.") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-mc-bias-reg", plot = g)
```

## Drug contrast bias (Risk Difference: Drug A - Drug B)

This figure shows the bias in the **12-month cumulative incidence difference between drugs** (RD = Drug A risk - Drug B risk) within each adherence stratum. The oracle RD is computed from the per-drug oracles. If a method is equally biased for both drugs, the RD bias cancels; if it's differentially biased (e.g., more confounded for one drug), the RD bias persists.

```{r fig-mc-bias-reg-rd}
#| label: fig-mc-bias-reg-rd
#| fig-cap: "Bias in the drug risk difference (Drug A - Drug B) by adherence stratum. Regression methods show differential bias across drugs, leading to biased drug comparisons — especially in the Low stratum."
#| fig-height: 5.5
#| fig-width: 10

# Compute per-iteration RD = Drug A estimate - Drug B estimate, for each method x stratum
reg_methods <- c("Reg (baseline)", "Reg (baseline + TV)")
iter_reg <- all_iter[all_iter$method %in% reg_methods & all_iter$converged, ]

# Pivot: one row per iter x method x stratum, columns for Drug_A and Drug_B estimates
iter_A <- iter_reg[iter_reg$drug == "Drug_A",
                   c("iter", "method", "stratum", "estimate", "oracle")]
iter_B <- iter_reg[iter_reg$drug == "Drug_B",
                   c("iter", "method", "stratum", "estimate", "oracle")]
names(iter_A)[4:5] <- c("est_A", "oracle_A")
names(iter_B)[4:5] <- c("est_B", "oracle_B")

iter_rd <- merge(iter_A, iter_B, by = c("iter", "method", "stratum"))
iter_rd$rd_est    <- iter_rd$est_A - iter_rd$est_B
iter_rd$rd_oracle <- iter_rd$oracle_A - iter_rd$oracle_B
iter_rd$rd_bias   <- iter_rd$rd_est - iter_rd$rd_oracle

# Aggregate
rd_agg <- aggregate(rd_bias ~ method + stratum, data = iter_rd,
  FUN = function(x) c(mean = mean(x), sd = sd(x), n = length(x)))
rd_agg <- do.call(data.frame, rd_agg)
names(rd_agg) <- c("method", "stratum", "mean_rd_bias", "sd_rd_bias", "n")

rd_agg$method  <- factor(rd_agg$method, levels = reg_methods)
rd_agg$stratum <- factor(rd_agg$stratum, levels = c("Low", "Mid", "High"))

g <- ggplot(rd_agg, aes(x = stratum, y = mean_rd_bias, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  # Thick bar: +/- 1 SD
  geom_linerange(aes(ymin = mean_rd_bias - sd_rd_bias,
                     ymax = mean_rd_bias + sd_rd_bias),
                 position = position_dodge(0.6), linewidth = 2.5, alpha = 0.25) +
  # Thin bar: +/- 1.96 SD
  geom_linerange(aes(ymin = mean_rd_bias - 1.96 * sd_rd_bias,
                     ymax = mean_rd_bias + 1.96 * sd_rd_bias),
                 position = position_dodge(0.6), linewidth = 0.5, alpha = 0.4) +
  # Point: MC mean
  geom_point(position = position_dodge(0.6), size = 2.5) +
  scale_y_continuous(labels = scales::percent_format(0.1)) +
  scale_colour_manual(values = reg_pal) +
  labs(x = "Adherence stratum",
       y = "Bias in RD (Drug A - Drug B)",
       colour = "Method",
       title = "Drug Comparison Bias: Risk Difference (Drug A - Drug B)",
       subtitle = "Bias in the 12-month cumulative incidence difference. Dashed = zero (unbiased).") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-mc-bias-reg-rd", plot = g)
```


# Estimand Diagnostic {#sec-estimand-diagnostic}

This section loads the estimand diagnostic results showing how each method's implicit oracle (what it is actually estimating) compares to the true L-TMLE oracle.

```{r estimand-load}
#| label: estimand-load
#| cache: false

estimand_path <- file.path(repo_root, "simulation_v4_mc", "results_full",
                           "strong_tv_confounding.rds")
estimand_data <- readRDS(estimand_path)
```

```{r tbl-estimand-diagnostic}
#| label: tbl-estimand-diagnostic

diag_tbl <- estimand_data$estimand_diagnostic
diag_tbl$drug_label <- gsub("_", " ", diag_tbl$drug)

display_diag <- diag_tbl[, c("drug_label", "stratum",
                               "naive_oracle", "reg_baseline_oracle", "reg_tv_oracle",
                               "ltmle_oracle", "gap_naive", "gap_reg_base")]
names(display_diag) <- c("Drug", "Stratum",
                          "Naive Oracle", "Reg(base) Oracle", "Reg(TV) Oracle",
                          "L-TMLE Oracle", "Gap: Naive", "Gap: Reg(base)")

kable(display_diag, digits = 4,
      caption = "Estimand diagnostic: each method targets a different oracle under strong TV confounding. The gap columns show how far naive/regression oracles are from the true causal estimand (L-TMLE oracle).") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r fig-estimand-diagnostic}
#| label: fig-estimand-diagnostic
#| fig-cap: "Estimand comparison: what each method is actually estimating vs the true causal oracle. Large gaps for naive/regression in the Low stratum reveal that these methods target fundamentally different (wrong) estimands."
#| fig-height: 6
#| fig-width: 10

# Reshape for plotting
diag_long <- rbind(
  data.frame(drug = diag_tbl$drug_label, stratum = diag_tbl$stratum,
             method = "Naive", oracle = diag_tbl$naive_oracle, stringsAsFactors = FALSE),
  data.frame(drug = diag_tbl$drug_label, stratum = diag_tbl$stratum,
             method = "Reg (baseline)", oracle = diag_tbl$reg_baseline_oracle, stringsAsFactors = FALSE),
  data.frame(drug = diag_tbl$drug_label, stratum = diag_tbl$stratum,
             method = "Reg (baseline + TV)", oracle = diag_tbl$reg_tv_oracle, stringsAsFactors = FALSE),
  data.frame(drug = diag_tbl$drug_label, stratum = diag_tbl$stratum,
             method = "L-TMLE", oracle = diag_tbl$ltmle_oracle, stringsAsFactors = FALSE)
)
diag_long$stratum <- factor(diag_long$stratum, levels = c("Low", "Mid", "High"))
diag_long$method <- factor(diag_long$method,
                           levels = c("Naive", "Reg (baseline)", "Reg (baseline + TV)", "L-TMLE"))

g <- ggplot(diag_long, aes(x = stratum, y = oracle, colour = method, shape = method)) +
  geom_point(position = position_dodge(0.6), size = 3.5) +
  facet_wrap(~ drug) +
  scale_y_continuous(labels = scales::percent_format(1)) +
  scale_colour_manual(values = mc_bias_pal) +
  scale_shape_manual(values = c(16, 17, 15, 18)) +
  labs(x = "Adherence stratum", y = "Oracle risk (what the method targets)",
       colour = "Method", shape = "Method",
       title = "Estimand Diagnostic: Different Methods Target Different Oracles",
       subtitle = "Under strong TV confounding, naive/regression target the wrong estimand in Low stratum") +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-estimand-diagnostic", plot = g)
```

Key finding: Under strong TV confounding, naive and baseline-regression methods target fundamentally different estimands from L-TMLE, particularly in the Low stratum. The gap is largest where the feedback loop is most active: patients with low PDC experience worsening health (via strong betaP), which further depresses their adherence (via strong alphaL), creating a vicious cycle that confounds the naive estimate.


# Summary Takeaways {#sec-summary}

```{r tbl-summary-takeaway}
#| label: tbl-summary-takeaway

takeaway <- data.frame(
  Finding = c(
    "1. Strong TV confounding creates large naive bias",
    "2. Baseline regression fails too",
    "3. Regression with TV covariates overshoots",
    "4. L-TMLE is approximately unbiased",
    "5. Positivity strain in Low stratum",
    "6. LMTP shift stays within support",
    "7. Different methods target different estimands"
  ),
  Detail = c(
    "Naive analysis systematically underestimates risk in Low stratum (negative bias ~20 pp for Drug B) due to the PDC -> L -> PDC feedback loop.",
    "Adjusting only for baseline W does not help because the confounding operates through time-varying L, not W (betaW = 0.10 is weak).",
    "Including L as a covariate in regression introduces collider bias: conditioning on the mediator (L is both confounder and mediator of prior PDC effects).",
    "L-TMLE properly handles the TV confounding by intervening rather than conditioning, producing bias < 2 pp across all strata.",
    "The Low stratum has limited positivity (few subjects sustained low PDC for all 4 blocks), increasing variance for all methods.",
    "The +10% additive shift from LMTP respects the natural support and avoids extrapolation problems.",
    "Estimand diagnostic shows naive/regression target different (wrong) oracles under strong TV confounding, especially in the Low stratum."
  ),
  stringsAsFactors = FALSE
)

kable(takeaway, col.names = c("Finding", "Detail"),
      caption = "Key takeaways: strong TV confounding amplifies the advantage of L-TMLE") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  column_spec(1, bold = TRUE, width = "30%") |>
  column_spec(2, width = "70%")
```


# Extension: Intercurrent Events {#sec-intercurrent}

In real-world EHR data, we also face informative censoring, treatment switching, and informative monitoring. These are additional time-varying nodes in the DAG that L-TMLE can handle through the same framework.

```{r fig-intercurrent-dag}
#| label: fig-intercurrent-dag
#| fig-cap: "Longitudinal DAG with intercurrent events. At each period: L -> A -> C -> S -> Y. Censoring (C) and switching (S) are driven by health and adherence. Red dashed arrows show feedback from treatment/outcome to future confounders."
#| fig-height: 6.5
#| fig-width: 12

# Nodes: 2 periods, each with L, A, C, S, Y plus baseline W, A0
ice_nodes <- data.frame(
  label = c("W",    "A0",
            "L1",   "A1",   "C1",   "S1",   "Y1",
            "L2",   "A2",   "C2",   "S2",   "Y2"),
  x     = c(0,      1.5,
            3.5,    5,      6.5,    8,      9.5,
            11.5,   13,     14.5,   16,     17.5),
  y     = c(1.5,    0,
            1.5,    0,      -1.5,   -1.5,   0,
            1.5,    0,      -1.5,   -1.5,   0),
  role  = c("baseline", "baseline",
            "confounder", "treatment", "ice", "ice", "outcome",
            "confounder", "treatment", "ice", "ice", "outcome"),
  stringsAsFactors = FALSE
)

ice_role_pal <- c("baseline"   = "#F0F0F0",
                  "confounder" = "#FEE0D2",
                  "treatment"  = "#DEEBF7",
                  "ice"        = "#FFF7BC",
                  "outcome"    = "#E5F5E0")

# Edges
ice_edges <- data.frame(
  from = c(
    # Baseline -> Period 1
    "W","W","W", "A0","A0",
    # Within Period 1
    "L1","L1","L1","L1",
    "A1","A1","A1",
    "C1", "S1",
    # Period 1 -> Period 2 (forward)
    "L1","A1","Y1",
    # Period 1 -> Period 2 (feedback)
    "A1","Y1",
    # Within Period 2
    "L2","L2","L2","L2",
    "A2","A2","A2",
    "C2", "S2"
  ),
  to = c(
    "L1","A1","C1", "L1","A1",
    "A1","C1","S1","Y1",
    "C1","S1","Y1",
    "Y1", "Y1",
    "L2","A2","Y2",
    "L2","L2",
    "A2","C2","S2","Y2",
    "C2","S2","Y2",
    "Y2", "Y2"
  ),
  stringsAsFactors = FALSE
)

# Look up coordinates
ice_edges$x    <- ice_nodes$x[match(ice_edges$from, ice_nodes$label)]
ice_edges$y    <- ice_nodes$y[match(ice_edges$from, ice_nodes$label)]
ice_edges$xend <- ice_nodes$x[match(ice_edges$to,   ice_nodes$label)]
ice_edges$yend <- ice_nodes$y[match(ice_edges$to,   ice_nodes$label)]

# Feedback arrows
ice_edges$feedback <- (ice_edges$from == "A1" & ice_edges$to == "L2") |
                      (ice_edges$from == "Y1" & ice_edges$to == "L2")

# Shorten arrows
shorten <- 0.45
for (i in seq_len(nrow(ice_edges))) {
  dx <- ice_edges$xend[i] - ice_edges$x[i]
  dy <- ice_edges$yend[i] - ice_edges$y[i]
  len <- sqrt(dx^2 + dy^2)
  if (len > 0) {
    ux <- dx/len; uy <- dy/len
    ice_edges$x[i]    <- ice_edges$x[i]    + shorten * ux
    ice_edges$y[i]    <- ice_edges$y[i]    + shorten * uy
    ice_edges$xend[i] <- ice_edges$xend[i] - shorten * ux
    ice_edges$yend[i] <- ice_edges$yend[i] - shorten * uy
  }
}

g <- ggplot() +
  # Period backgrounds
  annotate("rect", xmin = 2.5, xmax = 10.5, ymin = -2.5, ymax = 2.5,
           fill = "#F7F7F7", colour = "grey80", linetype = "dotted") +
  annotate("rect", xmin = 10.5, xmax = 18.5, ymin = -2.5, ymax = 2.5,
           fill = "#FAFAFA", colour = "grey80", linetype = "dotted") +
  annotate("text", x = 6.5, y = 2.3, label = "Period 1", fontface = "bold",
           size = 4, colour = "grey50") +
  annotate("text", x = 14.5, y = 2.3, label = "Period 2", fontface = "bold",
           size = 4, colour = "grey50") +
  # All edges in grey first
  geom_segment(data = ice_edges,
               aes(x = x, y = y, xend = xend, yend = yend),
               colour = "grey70", linewidth = 0.4,
               arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  # Feedback edges on top (red dashed)
  geom_segment(data = ice_edges[ice_edges$feedback, ],
               aes(x = x, y = y, xend = xend, yend = yend),
               colour = tab10[["red"]], linewidth = 0.7, linetype = "dashed",
               arrow = arrow(length = unit(0.10, "inches"), type = "closed")) +
  # Nodes
  geom_point(data = ice_nodes, aes(x = x, y = y, fill = role),
             shape = 21, size = 13, colour = "grey30", stroke = 0.6) +
  geom_text(data = ice_nodes, aes(x = x, y = y, label = label),
            size = 3.5, fontface = "bold") +
  scale_fill_manual(values = ice_role_pal,
                    labels = c("Baseline", "Confounder (L)",
                               "Intercurrent (C, S)",
                               "Outcome (Y)", "Treatment (A)"),
                    name = NULL) +
  # Role labels
  annotate("text", x = 19.5, y = 1.5,  label = "L: Health", hjust = 0,
           size = 3, colour = tab10[["orange"]]) +
  annotate("text", x = 19.5, y = 0,    label = "A: Adherence\nY: Outcome", hjust = 0,
           size = 3, colour = tab10[["blue"]], lineheight = 0.9) +
  annotate("text", x = 19.5, y = -1.5, label = "C: Censoring\nS: Switching", hjust = 0,
           size = 3, colour = tab10[["brown"]], lineheight = 0.9) +
  coord_cartesian(xlim = c(-0.8, 22), ylim = c(-3, 3), clip = "off") +
  theme_void(base_size = 14) +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-intercurrent-dag", plot = g)
```

### How L-TMLE handles intercurrent events

| Intercurrent Event | Challenge | L-TMLE Handling |
|---|---|---|
| **Informative censoring (LTFU)** | Patients who drop out are sicker / less adherent | IPCW: weight by inverse probability of remaining uncensored |
| **Informative monitoring** | Sicker patients tested more often — ascertainment bias | Include monitoring as an L-node; adjust in the g-model |
| **Treatment switching** | Patients switch based on failure / side effects | Include switching as treatment node or censor at switch |
| **Competing risks (death)** | Death precludes failure observation | Treat as censoring; estimate cause-specific risk |

The key point: each intercurrent event becomes an additional node in the DAG. L-TMLE handles them all through the same backward iterative targeting — intervene on treatment/adherence, adjust for confounders, weight for censoring. No new methodology is needed, just more nodes.


# Mediation Analysis Framework {#sec-mediation}

How much of the treatment effect operates *through* adherence vs through other pathways? The mediation DAG below shows the decomposition. The **total effect** of Drug (A) on Outcome (Y) can be decomposed into:

- **Natural indirect effect (NIE)**: the part mediated through adherence (A → PDC → Y)
- **Natural direct effect (NDE)**: the part operating through other pathways (pharmacokinetics, resistance barrier, etc.)

In the presence of time-varying confounding, standard mediation (Baron & Kenny, structural equation models) is biased. Longitudinal mediation requires interventional effects or the mediational g-formula.

```{r fig-mediation-dag}
#| label: fig-mediation-dag
#| fig-cap: "Mediation DAG for ART adherence. The total effect of treatment on outcome decomposes into: (1) the indirect effect through adherence (red path), (2) the direct effect through other pathways (blue path), and (3) the interaction between treatment and adherence on health (purple, dashed)."
#| fig-height: 6
#| fig-width: 11

med_dag <- dagitty('dag {
  A0        [pos="0,0"]
  L1        [pos="2,1.5"]
  PDC1      [pos="3,-1"]
  Y1        [pos="5,0"]
  L2        [pos="6,1.5"]
  PDC2      [pos="7,-1"]
  Y2        [pos="9,0"]

  A0 -> L1
  A0 -> PDC1
  A0 -> Y1
  L1 -> PDC1
  L1 -> Y1
  PDC1 -> Y1
  PDC1 -> L2
  Y1 -> L2
  Y1 -> PDC2
  Y1 -> Y2
  L1 -> L2
  L2 -> PDC2
  L2 -> Y2
  PDC2 -> Y2
  A0 -> PDC2
  A0 -> Y2
}')

med_tidy <- tidy_dagitty(med_dag)
med_df <- med_tidy$data

# Classify edges into mediation pathways
indirect_pairs <- list(
  c("A0", "PDC1"), c("PDC1", "Y1"), c("PDC1", "L2"),
  c("A0", "PDC2"), c("PDC2", "Y2")
)
direct_pairs <- list(
  c("A0", "Y1"), c("A0", "Y2"), c("A0", "L1")
)
interaction_pairs <- list(
  c("L1", "PDC1"), c("L2", "PDC2"),
  c("Y1", "PDC2")
)

med_df$path_type <- "confounder"
for (fp in indirect_pairs) {
  idx <- !is.na(med_df$name) & !is.na(med_df$to) &
         med_df$name == fp[1] & med_df$to == fp[2]
  med_df$path_type[idx] <- "indirect"
}
for (fp in direct_pairs) {
  idx <- !is.na(med_df$name) & !is.na(med_df$to) &
         med_df$name == fp[1] & med_df$to == fp[2]
  med_df$path_type[idx] <- "direct"
}
for (fp in interaction_pairs) {
  idx <- !is.na(med_df$name) & !is.na(med_df$to) &
         med_df$name == fp[1] & med_df$to == fp[2]
  med_df$path_type[idx] <- "interaction"
}

# Node roles
med_roles <- c(A0 = "treatment", L1 = "confounder", PDC1 = "mediator",
               Y1 = "outcome", L2 = "confounder", PDC2 = "mediator", Y2 = "outcome")
med_df$role <- med_roles[med_df$name]

role_fills <- c("treatment"  = "#DEEBF7",
                "confounder" = "#FEE0D2",
                "mediator"   = "#E8F4E4",
                "outcome"    = "#FFF7BC")

g <- ggplot(med_df, aes(x = x, y = y, xend = xend, yend = yend)) +
  # Confounder paths (grey)
  geom_dag_edges_link(
    data = function(d) d[!is.na(d$to) & d$path_type == "confounder", ],
    arrow = grid::arrow(length = unit(0.08, "inches"), type = "closed"),
    edge_colour = "grey60", edge_width = 0.4
  ) +
  # Direct effect paths (blue)
  geom_dag_edges_link(
    data = function(d) d[!is.na(d$to) & d$path_type == "direct", ],
    arrow = grid::arrow(length = unit(0.10, "inches"), type = "closed"),
    edge_colour = tab10[["blue"]], edge_width = 0.7
  ) +
  # Indirect effect paths through adherence (red/orange)
  geom_dag_edges_link(
    data = function(d) d[!is.na(d$to) & d$path_type == "indirect", ],
    arrow = grid::arrow(length = unit(0.10, "inches"), type = "closed"),
    edge_colour = tab10[["red"]], edge_width = 0.7
  ) +
  # Interaction paths (purple, dashed)
  geom_dag_edges_arc(
    data = function(d) d[!is.na(d$to) & d$path_type == "interaction", ],
    arrow = grid::arrow(length = unit(0.08, "inches"), type = "closed"),
    edge_colour = tab10[["purple"]], edge_width = 0.5,
    edge_linetype = "dashed", curvature = 0.15
  ) +
  geom_dag_point(aes(fill = role), size = 16, shape = 21,
                 colour = "grey30", stroke = 0.6) +
  geom_dag_text(size = 3.5, fontface = "bold", colour = "black") +
  scale_fill_manual(values = role_fills,
                    labels = c("Confounder (L)", "Mediator (PDC)",
                               "Outcome (Y)", "Treatment (Drug)"),
                    name = NULL) +
  # Legend for path types
  annotate("segment", x = 0.5, xend = 1.5, y = -2.2, yend = -2.2,
           colour = tab10[["blue"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.7, y = -2.2, label = "Direct effect (not through PDC)",
           hjust = 0, size = 3.2, colour = tab10[["blue"]]) +
  annotate("segment", x = 5, xend = 6, y = -2.2, yend = -2.2,
           colour = tab10[["red"]], linewidth = 0.8,
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 6.2, y = -2.2, label = "Indirect effect (through PDC)",
           hjust = 0, size = 3.2, colour = tab10[["red"]]) +
  annotate("segment", x = 0.5, xend = 1.5, y = -2.7, yend = -2.7,
           colour = tab10[["purple"]], linewidth = 0.6, linetype = "dashed",
           arrow = arrow(length = unit(0.08, "inches"), type = "closed")) +
  annotate("text", x = 1.7, y = -2.7, label = "Interaction (L affects PDC, creates TV confounding of mediation)",
           hjust = 0, size = 3.2, colour = tab10[["purple"]]) +
  theme_dag(base_size = 14) +
  theme(legend.position = "bottom")
print(g)
save_fig("fig-mediation-dag", plot = g)
```

### What the mediation analysis would estimate

| Estimand | Question | Method |
|----------|----------|--------|
| **Total effect (TE)** | What is the overall effect of Drug A vs B on failure? | LMTP natural risk comparison |
| **Natural direct effect (NDE)** | What is the drug effect if we *fix* adherence at its natural level under Drug B? | Interventional direct effect via longitudinal g-formula |
| **Natural indirect effect (NIE)** | How much of the drug effect operates by changing adherence? | TE - NDE, or interventional indirect effect |
| **Controlled direct effect (CDE)** | What is the drug effect at a *specific fixed* adherence level? | L-TMLE static regime, compare Drug A vs B at same PDC |

The quantile-matching analysis in the previous section provides an informal version of the CDE — it fixes the adherence distribution rather than a single level. A formal mediation decomposition (NDE/NIE) requires additional assumptions (cross-world independence) and is a planned extension.


# Export Figures {#sec-export}

All figures rendered during this document are saved to a single `.rds` file.
Load in R with `figs <- readRDS("talk_v3_figures.rds")` then access
individual plots by name (e.g., `figs$"fig-oracle"`).

```{r export-figures}
#| label: export-figures
#| cache: false

fig_list <- as.list(.fig_store)
out_path <- file.path(getwd(), "talk_v3_figures.rds")
saveRDS(fig_list, out_path)
cat(sprintf("Saved %d figure(s) to %s\n", length(fig_list), out_path))
cat("Available figures:\n")
cat(paste(" -", sort(names(fig_list))), sep = "\n")
```
