---
title: "Theme Preview: slate + breezedark"
format:
  html:
    theme: slate
    highlight-style: breezedark
    toc: true
    number-sections: true
    code-fold: show
    code-tools: true
    code-summary: "Show code"
    self-contained: true
---

# G-Computation Example

This page previews the **slate** theme with **breezedark** code highlighting.

## Data Setup

```{r}
#| code-fold: true
library(tidyverse)

set.seed(2026)
n <- 1000
age <- rnorm(n, 65, 10)
cvd <- rbinom(n, 1, plogis(-2 + 0.03 * age))
A <- rbinom(n, 1, plogis(-0.5 + 0.02 * age + 0.3 * cvd))
Y <- rbinom(n, 1, plogis(-2.5 + 0.03 * age + 0.5 * cvd - 0.4 * A))

dat <- tibble(age, cvd, A, Y)
```

## Outcome Model

```{r}
mod <- glm(Y ~ A + age + cvd, family = binomial, data = dat)
broom::tidy(mod, conf.int = TRUE) |>
  knitr::kable(digits = 3, caption = "Outcome model coefficients")
```

## Counterfactual Predictions

```{r}
p1 <- predict(mod, newdata = mutate(dat, A = 1), type = "response")
p0 <- predict(mod, newdata = mutate(dat, A = 0), type = "response")

risk1 <- mean(p1)
risk0 <- mean(p0)
ate <- risk1 - risk0

data.frame(
  Measure = c("Risk under treatment", "Risk under control", "ATE (risk difference)"),
  Estimate = round(c(risk1, risk0, ate), 4)
) |> knitr::kable()
```

## Propensity Score Overlap

```{r}
#| code-fold: true
#| fig-cap: "Propensity score overlap by treatment group"
ps <- predict(glm(A ~ age + cvd, family = binomial, data = dat), type = "response")

ggplot(tibble(ps = ps, A = factor(dat$A, labels = c("Control", "Treated"))),
       aes(x = ps, fill = A)) +
  geom_density(alpha = 0.4) +
  labs(x = "Propensity Score", y = "Density", fill = "Group") +
  scale_fill_manual(values = c("#EE6677", "#228833")) +
  theme_minimal()
```

## TMLE Targeting Step

```{r}
# Clever covariate
H <- dat$A / ps - (1 - dat$A) / (1 - ps)

# Fluctuation
logit <- function(p) log(p / (1 - p))
Qinit <- predict(mod, type = "response")
epsilon <- glm(dat$Y ~ -1 + offset(logit(Qinit)) + H, family = binomial)$coef

# Update predictions
H1 <- 1 / ps
H0 <- -1 / (1 - ps)
Q1_star <- plogis(logit(p1) + epsilon * H1)
Q0_star <- plogis(logit(p0) + epsilon * H0)

tmle_ate <- mean(Q1_star) - mean(Q0_star)
cat("G-computation ATE:", round(ate, 4), "\n")
cat("TMLE ATE:         ", round(tmle_ate, 4), "\n")
```

::: {.callout-note}
## Key Takeaway
The TMLE targeting step adjusts the initial outcome predictions to reduce bias for the ATE specifically. The fluctuation parameter $\hat{\epsilon}$ captures the correction needed.
:::

::: {.callout-warning}
## In development
This is a theme preview page, not part of the handbook.
:::

## Math Display

The ATE is identified by the g-computation formula:

$$
E[Y(1)] - E[Y(0)] = E_W\!\left[E[Y \mid A=1, W] - E[Y \mid A=0, W]\right]
$$

Under consistency, exchangeability conditional on $W$, and positivity.
