# Competing Risks in Estimands: Choosing What You Estimate Before How {#sec-competing-risks-estimands}

*Estimated reading time: 50 minutes*

::: {.callout-tip}
## Learning Objectives

After completing this chapter, you will be able to:

1. Articulate why the estimand must be defined before the estimator when competing events are present, using the ICH E9(R1) framework.
2. Distinguish **cause-specific hazards**, **subdistribution hazards**, and **cumulative incidence functions** in plain language and in formulas, and explain which scientific question each answers.
3. Recognize when "censoring at the competing event and computing $1-\hat S(t)$" silently changes the estimand and overestimates observed-world risk.
4. Choose between **total effect on cumulative incidence**, **cause-specific hazard estimands**, **hypothetical (controlled-direct-effect) estimands**, and **separable effects**, and state the identifiability assumptions each requires.
5. Implement **TMLE for competing risks** in R (and prototype it in Python or Julia), targeting an adjusted absolute risk at a clinically meaningful horizon with influence-function-based inference.
6. Diagnose positivity, weight stability, and curve agreement, and design sensitivity analyses around alternative estimands rather than around alternative software defaults.
:::

This chapter is the conceptual companion to [@sec-tte-competing]. Where that chapter walked through the Causal Roadmap and the mechanics of survival TMLE end-to-end, this one zooms in on the single most consequential decision in any competing-risks analysis: what quantity, exactly, are you trying to estimate? Once that is settled, the estimator usually picks itself.

## Executive Summary {#sec-cr-summary}

The order of work is: define the estimand first, then choose the estimator. Under ICH E9(R1), a competing event is not merely a statistical nuisance. It is an intercurrent event that can change the meaning of the clinical question or render the endpoint nonexistent (a patient who dies of a non-cancer cause cannot subsequently relapse). It must therefore be handled explicitly in the estimand.

Five practical claims follow:

1. **Censoring subjects at the competing event and reporting $1-\hat S_{\text{KM}}(t)$ does not estimate the real-world risk of the event of interest.** It estimates either a cause-specific hazard or, if turned into a survival probability, a hypothetical world in which the competing event had been prevented. That is why $1-\hat S_{\text{KM}}(t)$ typically *overestimates* observed-world cumulative incidence when competing events are non-trivial.
2. **If the question is the actual probability of the event by time $t$ in the world where competing events occur,** the competing event must be part of the estimand, and the natural targets are cumulative incidence functions (CIFs) estimated nonparametrically by the Aalen--Johansen (AJ) estimator or by adjusted absolute-risk models.
3. **Cause-specific hazards remain useful for etiologic and mechanistic questions,** but they do not, on their own, answer the probability question that patients and policy makers usually ask.
4. **Subdistribution-hazard (Fine--Gray) coefficients are mathematically tied to cumulative incidence** but are harder to interpret causally because they blend effects on the event of interest and on the competing event through an artificial risk set.
5. **TMLE is especially attractive when the target is an adjusted absolute risk or cumulative incidence at a clinically meaningful horizon.** It targets the estimand directly, remains a plug-in estimator constrained to $[0,1]$, supports flexible nuisance modeling, and yields influence-function-based inference. Competing-risks TMLE is most mature in R (`survtmle`, `concrete`), with newer support in Python and a generic TMLE infrastructure in Julia.

## 1. Why Competing Risks Belong in the Estimand {#sec-cr-why}

### Opening pharmacoepi scenario: heart failure hospitalization

A study compares Drug A versus Drug B on hospitalization for heart failure (HF) by 24 months. Death before HF hospitalization is common, especially in older and frailer patients. Death prevents future hospitalization. Before writing any model, the analyst has to choose which question is being asked:

- The risk of HF hospitalization before death, in the world where some patients die first.
- The risk of HF hospitalization if death could somehow be prevented.
- The risk of HF hospitalization or death as a composite endpoint.
- The effect on HF hospitalization among people who would survive under both strategies (a principal stratum).
- The effect while still on treatment, censoring at discontinuation.

Each of these is a different estimand. Each requires different identification assumptions. Each can give a different numerical answer with the same data, and at most one of them is the question the prescriber, regulator, or patient actually wants answered. The rest of this chapter is about choosing among them on purpose.

### A failure mode that motivates everything else

Now consider an adjuvant oncology trial. The event of interest is cancer recurrence; non-cancer death is a competing event. A reviewer reads in the methods section:

> "Patients were censored at non-cancer death. Recurrence-free survival was estimated by the Kaplan--Meier method."

This single sentence has, almost certainly, swapped the estimand silently. Kaplan--Meier treats non-cancer deaths as if they were administrative dropouts who *might still recur* in a world where they had been followed longer. They cannot. The Kaplan--Meier curve thus answers a counterfactual question that was never the trial's clinical question:

> "What would the recurrence probability be in a hypothetical world where non-cancer death had been prevented?"

If treatment also affects non-cancer mortality (and most cancer therapies do), this hypothetical question is not even close to the patient-facing question ("What is my chance of relapse by 5 years?"), and the two can disagree in direction.

::: {.callout-warning}
## Common mistake #1: $1-\hat S_{\text{KM}}(t)$ as "risk" with competing events

Reporting $1$ minus the cause-specific Kaplan--Meier estimate as a probability of the event of interest in the presence of competing events:

- typically overestimates observed-world cumulative incidence;
- silently switches the estimand to a hypothetical "competing-event-free world";
- and breaks the comparability of arms when treatment affects the competing event.

Use the Aalen--Johansen estimator or an adjusted CIF method instead [@austin2017competing; @stensrud2019limitations].
:::

### Three families of scientific questions

Once we agree that competing events change *what we mean*, the next step is to recognize that there are at least three legitimate families of scientific questions, each pointing to a different estimand class.

| Question family | Plain-English form | Natural estimand | Default estimator |
|---|---|---|---|
| **Observed-world risk** | "By 5 years, what proportion of patients will have relapsed under treatment $a$, in the world where some die of other causes first?" | Treatment-specific cumulative incidence $F_1^a(t) = P(T^a \le t, J^a = 1)$ | AJ, adjusted CIF, FG, TMLE for CIF |
| **Etiologic / mechanistic** | "Among patients still alive and event-free, how does treatment change the *rate* at which relapse occurs?" | Cause-specific hazard ratio | Cause-specific Cox |
| **Pathway / hypothetical** | "What would the relapse risk be if we could prevent non-cancer death?" or "What part of the treatment effect on relapse is *not* mediated through mortality?" | Controlled direct effect on CIF; separable effects | IPCW, g-formula, AIPW, TMLE with hypothetical censoring |

: Three families of questions and their natural estimands {#tbl-cr-families}

The first family (observed-world risk) is what patients, clinicians, regulators, and payers usually want. The second family is appropriate for biology-oriented questions where one truly wants to characterize an instantaneous rate among survivors. The third family is rarely the right *primary* estimand but is sometimes scientifically necessary, and when it is, it should be defended explicitly, not reached for as a default.

## 2. Core Concepts {#sec-cr-concepts}

### Plain-English vocabulary

> A competing risk is an event that prevents the event of interest from occurring or being observed as originally defined. Censoring is different. Censoring means we stop observing someone before we know what would have happened. A competing event means something clinically meaningful happened that changes what can happen next.

It helps to keep five distinct ideas separate. They often get lumped under "censoring" but should not be.

| Term | What it is | How it should usually be handled |
|---|---|---|
| **Administrative censoring** | The study's follow-up window ends before the event happens. The patient is still alive and event-free at last contact. | Treat as independent right censoring. |
| **Loss to follow-up** | The patient stops being observed for reasons unrelated (or weakly related) to the outcome process. | Treat as right censoring; check whether it is plausibly conditionally independent given measured covariates. |
| **Treatment discontinuation** | The patient stops the assigned treatment. Outcome ascertainment may continue. | An *intercurrent event* under ICH E9(R1). The strategy depends on the estimand: treatment-policy keeps following the patient, while-on-treatment censors at discontinuation. |
| **Death (when the endpoint is nonfatal)** | The patient dies before the event of interest can occur. | A **competing event**. It is part of the outcome process, not a censoring process. |
| **True competing event (e.g., transplant before waitlist death)** | An event that, by definition, makes the event of interest impossible thereafter. | Same as death above: part of the outcome process. |

: Administrative censoring, LTF, discontinuation, death, and true competing events are five different things {#tbl-cr-vocab}

The single most common conceptual error in applied competing-risks work is treating the bottom two rows like the top two. Death is not a study management problem. It is a clinical event that changes the joint outcome process.

### Notation

Let $T$ be the event time and $J \in \{1, 2\}$ the event type, with $J = 1$ the event of interest and $J = 2$ a competing event. Let $A$ be a baseline treatment, $W$ a vector of baseline covariates, and $C$ a (possibly informative) right-censoring time.

### Cause-specific hazard

For cause $j$,

$$
\lambda_j(t) = \lim_{\Delta t \to 0} \frac{P(t \le T < t+\Delta t,\ J=j \mid T \ge t)}{\Delta t}.
$$

In words: among individuals who are still alive **and** event-free at time $t$, $\lambda_j(t)$ is the instantaneous occurrence rate of cause $j$. It is a rate in the population that has not yet had any event.

### Subdistribution hazard

For the event of interest,

$$
\tilde\lambda_1(t) = \lim_{\Delta t \to 0} \frac{P(t \le T < t+\Delta t,\ J=1 \mid T \ge t \ \text{or}\ (T \le t,\ J \ne 1))}{\Delta t}.
$$

The denominator deliberately keeps subjects who already had the *competing* event in an extended risk set, which is what links $\tilde\lambda_1$ algebraically to the cumulative incidence function. This is mathematically convenient (Fine--Gray regression coefficients map directly to effects on $F_1$), but $\tilde\lambda_1$ is not a physical event rate among truly event-free people.

### Cumulative incidence function (CIF)

Concretely, this is the share of patients who have actually had the event of interest, say relapse, by a given time such as 5 years, counting the real world in which some patients die of other causes first.

$$
F_j(t) = P(T \le t,\ J = j).
$$

This is the probability that cause $j$ has occurred by time $t$. It directly answers the patient-level question. The natural nonparametric estimator is the **Aalen--Johansen** estimator,

$$
\hat F_j(t) = \sum_{t_k \le t} \hat S(t_{k-1})\, \frac{d N_j(t_k)}{Y(t_k)},
$$

where $\hat S$ is the all-cause survival from the Kaplan--Meier-style product of $1 - dN/Y$ over all event types, $dN_j(t_k)$ counts cause-$j$ events at $t_k$, and $Y(t_k)$ is the at-risk size. The crucial feature: $\hat F_j(t)$ uses *all-cause* survival as the carrier, so subjects who die of competing causes are correctly removed from future eligibility for cause $j$ rather than treated as still potentially eligible.

::: {.callout-note}
## Why AJ ≠ 1−KM (in one line)

KM-based "risk" multiplies $1 - dN_j / Y$ over time, treating competing events as if those individuals could still experience cause $j$ later. AJ multiplies the *all-cause* survival by the *cause-specific* increment, so it never re-credits people who have already left the population.
:::

In plain English: at each follow-up time, AJ asks two questions in order. First, what fraction of the original cohort is still alive and event-free, given everything that has happened so far? Second, of those people, what fraction has the event of interest right now? The cumulative incidence by time $t$ is the running sum of those "still around" times "event of interest now" pieces. Note that "still around" gets reduced by competing events as well as by the event of interest, so a patient who died of a non-cancer cause is correctly removed from the pool that could later contribute relapses.

### Plain-language interpretation

- **Cause-specific hazard:** "Among people still alive and event-free, how fast is the event of interest happening right now?"
- **Cumulative incidence:** "By a clinically relevant time point, what proportion have already had the event of interest before any competing event happened first?"
- **Subdistribution hazard:** A mathematical device to model cumulative incidence directly; not a physical rate.

These quantities can move in different directions. A treatment can *lower* the cause-specific hazard of relapse but *raise* non-cancer mortality enough that the *observed risk* of relapse falls mostly because more people die first. Conversely, a therapy can *raise* the relapse hazard but extend life so much that more people live long enough to relapse, *raising* the CIF. Both can be "true." Which one matters depends on the question.

## 3. Estimands and Causal Interpretation {#sec-cr-estimands}

### Total effect on cumulative incidence

For a baseline treatment $A$, the contrast

$$
F_1^{a=1}(t) - F_1^{a=0}(t)
$$

is the total effect on the event of interest by time $t$. It is the cumulative incidence patients actually face under each treatment strategy. In randomized trials with adequate follow-up, this is usually the cleanest causal estimand; in observational data, it requires consistency, conditional exchangeability given measured confounders, and positivity for treatment.

### Cause-specific hazard estimands

The cause-specific hazard ratio is much harder to interpret causally. Even in randomized studies, hazard-based contrasts condition on remaining event-free to time $t$, and treatment may affect *who* remains in that risk set through earlier effects on either cause. So a cause-specific hazard ratio is generally not a marginal causal effect, even under randomization. Use it for etiology, not for the patient-facing risk question.

### Hypothetical (controlled-direct-effect) estimands

If we censor the competing event and try to identify a quantity like

$$
F_1^{a, \overline{c}_2 = 0}(t) = P(T^a \le t,\ J^a = 1 \mid \text{competing event eliminated}),
$$

we are asking what the risk would be in a world where the competing event had been prevented. Identifiability now requires the additional, often heroic, assumption that those artificially censored at the competing event are exchangeable, given measured history, with those who remain uncensored. We also need positivity for the *induced* censoring mechanism. These assumptions are least plausible exactly when competing events are common and clinically important.

### Separable effects

If the underlying scientific question is really *pathway-based* ("What part of the treatment effect on relapse operates through mortality, and what part does not?"), the modern answer is separable effects rather than ad hoc censoring. Conceptually, treatment $A$ is decomposed into two components $A_Y$ (acting on the event of interest) and $A_D$ (acting on the competing event). Identification then requires a treatment-decomposition assumption: that one can intervene on the components separately. This is a much cleaner pathway estimand than "censor the competing event," but it requires that the decomposition is scientifically meaningful (e.g., separating an antineoplastic effect from a cardiotoxic effect).

::: {.callout-important}
## Identifiability summary

| Estimand | Beyond consistency, requires... |
|---|---|
| Cause-specific hazard (descriptive) | Independent right censoring; correct hazard model |
| Total effect on CIF | Conditional exchangeability for $A$; positivity for $A$; independent right censoring given $(A,W)$ |
| Hypothetical CIF (CDE) | All of the above plus exchangeability for the censoring induced by removing the competing event, plus positivity in the induced pseudo-population |
| Separable effects | Total-effect assumptions plus treatment-decomposition assumption (and partial isolation) |

Each row up adds assumptions that are typically harder to defend than the row above it.
:::

## 4. Choosing the Method {#sec-cr-method}

The most important practical point is that "censoring at the competing event" versus "modeling the competing event" is partly a false dichotomy. A cause-specific Cox model *does* censor the competing event in the cause-1 indicator, and that is fine *if* the estimand is a cause-specific hazard. The real mistake is to take that censored analysis and interpret $1-\hat S_{\text{KM}}$, or a single cause-specific hazard ratio, as if it were the observed-world cumulative risk.

If the goal is risk, either estimate cumulative incidence *directly* or model all relevant cause-specific hazards and combine them into cumulative incidence.

### Comparison table

| Approach | Estimand | Main assumptions | Typical bias source | Software | Code skeleton | Verdict |
|---|---|---|---|---|---|---|
| Naive censoring + $1-\hat S_{\text{KM}}$ | Unintended hypothetical risk | Competing event acts like non-informative censoring (rarely credible) | Overestimates observed-world risk | Any survival package | `survfit(Surv(time, event==1) ~ A)` | **Do not use** for observed-world risk |
| Cause-specific Cox | Instantaneous rate of cause 1 among event-free | Independent censoring; correct hazard model; for causal use, conditional exchangeability and positivity | HR misread as risk effect; selection through survivors | R `survival`; Py `scikit-survival`; Jl `LSurvival.jl` | `coxph(Surv(time, event==1) ~ A + W)` | Best for etiologic questions; pair with CIF |
| Nonparametric / standardized CIF (AJ) | Actual probability of event by $t$ | Independent right censoring; correct nuisance models for adjusted versions | Confounded if unadjusted; misspecification if adjusted | R `cmprsk`, `riskRegression`, `adjustedCurves`; Py `scikit-survival`, `lifelines`; Jl `LSurvival.jl` | `cuminc(time, event, group=A)` | Default descriptive target |
| Subdistribution (Fine--Gray) | Effect on subdistribution hazard ↔ CIF | Independent censoring (or covariate-adjusted weights); proportional subdistribution hazards | Coefficients hard to interpret causally; pathway-blending | R `cmprsk::crr`, `riskRegression::FGR`; SAS `PHREG`; Stata `stcrreg` | `crr(time, event, X, failcode=1)` | Useful for prognosis; not for etiologic causal interpretation |
| IPCW after censoring competing event | Hypothetical CDE on CIF | Well-defined hypothetical; exchangeability for induced censoring; positivity | Extreme weights; fragile late in follow-up | R `causalCmprsk`; weighted pooled-logistic pipelines | `glm(N1 ~ A + W + t, weights=1/(gAhat*GChat), family=binomial)` | Use only when the hypothetical estimand is justified |
| TMLE for CIF | Adjusted absolute risk (total or hypothetical) | Causal assumptions for the chosen target; positivity; nuisance robustness | Weak positivity; unstable IPCW component | R `survtmle`, `concrete`; Py `PyTMLE` or custom; Jl `TMLE.jl` + custom | `survtmle(..., method="hazard", returnIC=TRUE)` | **Best advanced choice** for adjusted absolute-risk estimation |

: Method comparison for competing-risks analysis {#tbl-cr-methods}

When proportional subdistribution hazards look dubious, direct CIF or absolute-risk regression (e.g., `timereg`) is preferable to forcing a Fine--Gray model.

### Decision flowchart

```{mermaid}
flowchart TD
    A[Start with the scientific question] --> B{Is the target the actual probability of the event by time t in the world where competing events occur?}
    B -->|Yes| C[Include the competing event in the estimand]
    C --> D{Is the goal mainly descriptive or unadjusted group comparison?}
    D -->|Yes| E[Aalen-Johansen CIF curves]
    D -->|No| F{Primary goal is prognosis or absolute-risk prediction?}
    F -->|Yes| G[Adjusted CIF / Fine-Gray / direct risk regression / TMLE]
    F -->|No| H[Cause-specific hazard models for each cause AND report CIF]
    B -->|No| I{Is a hypothetical intervention preventing the competing event scientifically meaningful and decision-relevant?}
    I -->|No| J[Reconsider the estimand; do not censor the competing event]
    I -->|Yes| K[Hypothetical direct-effect estimand]
    K --> L[IPCW, g-formula, AIPW, or TMLE with explicit competing-event censoring model]
    E --> M[Check curve agreement and report competing-event incidence]
    G --> M
    H --> M
    L --> N[Inspect weights, positivity, and sensitivity to modeling assumptions]
```

## 5. TMLE for Competing Risks in Depth {#sec-cr-tmle}

### Why TMLE here

When the target is an adjusted absolute risk at one or more clinically meaningful horizons, TMLE has four properties that matter in this setting:

1. **It targets the estimand directly,** rather than reporting a regression coefficient and hoping it maps to the question.
2. **It is a plug-in estimator constrained to the parameter space.** Estimated risks stay in $[0,1]$ even with flexible nuisance learners.
3. **It supports data-adaptive nuisance estimation** (e.g., Super Learner) without sacrificing valid inference.
4. **It yields influence-function-based standard errors,** which extend naturally to risk differences, risk ratios, and simultaneous bands over multiple horizons.

Two common discrete-time formulations are used in practice: a cause-specific hazard TMLE and an **iterated-conditional-mean** TMLE. The R package `survtmle` exposes both. Recent work extends TMLE to continuous time and to simultaneous inference over absolute-risk curves through one-step targeting, implemented in R `concrete`.

### Target parameter and identification

In a discrete-time representation over intervals $k = 1, \dots, \tau$, define the cause-specific hazards

$$
h_j(k, a, w) = P(N_{jk} = 1 \mid Y_k = 1, A = a, W = w), \quad j \in \{1, 2\},
$$

where $Y_k$ indicates being uncensored and event-free at the start of interval $k$, and $N_{jk}$ indicates cause-$j$ occurrence in interval $k$. The treatment-specific cumulative incidence for cause 1 at horizon $\tau$ is

$$
\psi_a(\tau) = E_W\!\left[ \sum_{k=1}^{\tau} h_1(k, a, W) \prod_{u=1}^{k-1} \{1 - h_1(u, a, W) - h_2(u, a, W)\} \right].
$$

Read this in pieces. The product is the probability of being event-free through interval $k-1$ under treatment $a$. The factor $h_1(k, a, W)$ is the conditional hazard of the event of interest in interval $k$. Their product, summed over $k$, is the probability of *first* experiencing the event of interest by $\tau$. The outer expectation standardizes over the baseline covariate distribution.

Identification of $\psi_a(\tau)$ as a causal target requires the usual point-treatment assumptions: consistency, conditional exchangeability of $A$ given $W$, positivity, and independent censoring conditional on the measured history. If the target is hypothetical and censors the competing event, the competing event becomes part of the censoring mechanism, and exchangeability and positivity for that *induced* censoring must hold.

### Influence functions and targeted updates

For a fixed treatment $a$ and horizon $\tau$, the efficient influence function (EIF) for the CIF target can be written schematically as

$$
D^*_{a,\tau}(O) = \sum_{k=1}^{\tau} H_{a,\tau,k}(O) \{N_{1k} - h_1(k, A, W)\} + Q_{a,\tau}(W) - \psi_a(\tau).
$$

The first term is a weighted residual for the event-of-interest hazard. The second term recenters the observation around the marginal target. The **clever covariate** $H_{a,\tau,k}(O)$ contains:

- the inverse probability of treatment $g_A(a \mid W)^{-1}$,
- the inverse probability of remaining uncensored up to $k$, $G_C(k \mid A, W)^{-1}$,
- a factor that propagates the contribution of interval $k$ forward to horizon $\tau$ (essentially the remaining event-free survival under treatment $a$).

The targeting step updates the initial nuisance estimates along a least-favorable submodel via a logistic fluctuation:

$$
\text{logit}\{h_1^*(k, A, W)\} = \text{logit}\{\hat h_1(k, A, W)\} + \varepsilon \, H_{a, \tau, k}(O),
$$

where $\varepsilon$ is chosen so that the empirical mean of the estimated influence function is driven (close) to zero. TMLE thus starts with an initial estimate of the hazards or conditional means, then targets it to the exact parameter of interest rather than accepting the initial nuisance fit as final.

In competing-risks settings, IPCW enters TMLE through the clever covariate. This is why TMLE can be viewed as a doubly robust refinement of pure IPCW: it uses censoring weights *and* borrows strength from direct modeling of the event process.

### Pseudo-code

```text
Input:
  O_i = (W_i, A_i, T_i, Delta_i),  i = 1,...,n
  event_of_interest = 1; competing_event = 2; horizon = tau

Step 1. Expand to person-period data for k = 1,...,tau
        Build Y_ik, N1_ik, N2_ik, C_ik

Step 2. Fit treatment model gA_hat(A | W)
        Fit censoring model GC_hat(k | A, W)
        Fit cause-specific hazard models h1_hat(k | A, W), h2_hat(k | A, W)

Step 3. Initial plug-in CIF:
        psi_a_init(tau) = average over W of
          sum_k h1_hat(k,a,W) * prod_{u<k}[1 - h1_hat(u,a,W) - h2_hat(u,a,W)]

Step 4. Construct clever covariate H_{a,tau,k} using:
          inverse treatment probability, inverse censoring survival,
          remaining survival/competing-event-free contribution to tau

Step 5. Targeting update:
          logit(h1_star) = logit(h1_hat) + epsilon * H
        Update other nuisance components along their submodels if needed

Step 6. Recompute targeted CIF psi_a_star(tau)
        Compute estimated influence function D_i_star

Step 7. Report contrasts: risk difference, risk ratio
Step 8. Variance: Var_hat(psi) = (1/n^2) * sum_i (D_i_star)^2
        Build Wald or simultaneous confidence intervals/bands
```

### R implementation: `survtmle` {#sec-cr-r}

```r
# Event coding: 0 = censored, 1 = event of interest, 2 = competing event
library(survtmle)

fit_tmle <- survtmle(
  ftime      = df$time,
  ftype      = df$event,
  trt        = df$A,
  adjustVars = df[c("W1", "W2", "W3")],
  t0         = 365,                      # 1-year horizon
  method     = "hazard",                 # cause-specific hazard TMLE
  SL.trt     = c("SL.glm", "SL.glmnet"),
  SL.ctime   = c("SL.glm", "SL.glmnet"), # censoring nuisance
  SL.ftime   = c("SL.glm", "SL.glmnet"), # cause-specific event nuisances
  returnIC   = TRUE
)

fit_tmle$est                # treatment-specific CIF estimates
sqrt(diag(fit_tmle$var))    # IF-based standard errors
confint(fit_tmle)
```

For continuous-time problems, `concrete` implements one-step TMLE and supports simultaneous bands over the full CIF curve.

### A manual hazard-based prototype in R

This template makes the moving parts visible. It is teaching code, not production code.

```r
# Person-period data: id, t, A, W1, W2, Y, N1, N2, C
# Y = at risk at start of interval; N1 = event of interest; N2 = competing; C = censored

base_df <- unique(long_df[c("id", "A", "W1", "W2")])

# 1. Initial nuisance models -------------------------------------------------
gA_fit <- glm(A ~ W1 + W2, family = binomial(), data = base_df)
gC_fit <- glm(C  ~ A + W1 + W2 + factor(t), family = binomial(),
              data = subset(long_df, Y == 1))
h1_fit <- glm(N1 ~ A + W1 + W2 + factor(t), family = binomial(),
              data = subset(long_df, Y == 1))
h2_fit <- glm(N2 ~ A + W1 + W2 + factor(t), family = binomial(),
              data = subset(long_df, Y == 1))

long_df$gA_hat <- predict(gA_fit, newdata = long_df, type = "response")
long_df$gC_hat <- predict(gC_fit, newdata = long_df, type = "response")
long_df$h1_hat <- predict(h1_fit, newdata = long_df, type = "response")
long_df$h2_hat <- predict(h2_fit, newdata = long_df, type = "response")

# 2. Censoring survival and remaining-survival factors ----------------------
long_df$G_hat        <- censoring_survival(long_df, "gC_hat")
long_df$S_remain_hat <- remaining_survival_to_tau(long_df, "h1_hat", "h2_hat",
                                                  tau = 365)

# 3. Clever covariate for psi_a(tau) ----------------------------------------
a <- 1
long_df$H <- with(long_df,
  as.numeric(A == a) / pmax(gA_hat, 1e-6) /
                       pmax(G_hat,  1e-6) *
                       S_remain_hat
)

# 4. Targeted update of cause-1 hazard --------------------------------------
fluct_fit <- glm(
  N1 ~ -1 + H,
  family = binomial(),
  data   = subset(long_df, Y == 1),
  offset = qlogis(pmin(pmax(h1_hat, 1e-6), 1 - 1e-6))
)
eps <- coef(fluct_fit)[["H"]]

long_df$h1_star <- plogis(
  qlogis(pmin(pmax(long_df$h1_hat, 1e-6), 1 - 1e-6)) + eps * long_df$H
)

# 5. Targeted CIF and EIF-based inference -----------------------------------
psi_hat <- standardized_cif(long_df, h1 = "h1_star", h2 = "h2_hat",
                            tau = 365, a = a)
eif_i   <- eif_cif(long_df, h1 = "h1_star", h2 = "h2_hat",
                   tau = 365, a = a)
se_hat  <- sd(eif_i) / sqrt(length(unique(long_df$id)))
ci_hat  <- psi_hat + c(-1, 1) * 1.96 * se_hat
```

### Python prototype

```python
# Diagnostic: nonparametric competing-risks CIFs
from sksurv.nonparametric import cumulative_incidence_competing_risks

times, cif = cumulative_incidence_competing_risks(
    event     = df["event_code"].to_numpy(),   # 0=censor, 1=interest, 2=competing
    time_exit = df["time"].to_numpy()
)
# cif[1] = event-of-interest CIF; cif[2] = competing-event CIF
```

```python
# Minimal hazard-based competing-risks TMLE prototype
import numpy as np
from scipy.special import expit, logit
from sklearn.linear_model import LogisticRegression

tau, a = 365, 1
long_df, base_df = make_person_period(df, tau=tau)

# 1. Treatment model
gA = LogisticRegression(max_iter=1000).fit(base_df[["W1", "W2"]], base_df["A"])
base_df["gA_hat"] = gA.predict_proba(base_df[["W1", "W2"]])[:, 1]
long_df = long_df.merge(base_df[["id", "gA_hat"]], on="id", how="left")

# 2. Censoring + cause-specific hazards
long_df["gC_hat"] = fit_hazard(long_df[long_df["Y"] == 1], "C",  ["A","W1","W2","t"])
long_df["h1_hat"] = fit_hazard(long_df[long_df["Y"] == 1], "N1", ["A","W1","W2","t"])
long_df["h2_hat"] = fit_hazard(long_df[long_df["Y"] == 1], "N2", ["A","W1","W2","t"])

# 3. Clever covariate
long_df["G_hat"] = censoring_survival(long_df, "gC_hat")
long_df["S_remain_hat"] = remaining_survival_to_tau(long_df, "h1_hat", "h2_hat", tau)
pA = np.where(a == 1, long_df["gA_hat"], 1 - long_df["gA_hat"])
long_df["H"] = ((long_df["A"] == a).astype(float)
                / np.maximum(pA, 1e-6)
                / np.maximum(long_df["G_hat"], 1e-6)
                * long_df["S_remain_hat"])

# 4. One-parameter targeting via Newton updates
mask = long_df["Y"] == 1
offset = logit(np.clip(long_df.loc[mask, "h1_hat"], 1e-6, 1 - 1e-6))
X = long_df.loc[mask, ["H"]].to_numpy()[:, 0]
y = long_df.loc[mask, "N1"].to_numpy()

eps = 0.0
for _ in range(25):
    q_star = expit(offset + eps * X)
    score  = np.sum(X * (y - q_star))
    info   = np.sum((X**2) * q_star * (1 - q_star))
    step   = score / max(info, 1e-8)
    eps   += step
    if abs(step) < 1e-8: break

long_df["h1_star"] = long_df["h1_hat"]
long_df.loc[mask, "h1_star"] = expit(offset + eps * X)

# 5. Targeted CIF and IF-based inference
psi_hat = standardized_cif(long_df, "h1_star", "h2_hat", tau=tau, a=a)
eif_i   = eif_cif(long_df, "h1_star", "h2_hat", tau=tau, a=a)
se_hat  = np.std(eif_i, ddof=1) / np.sqrt(df.shape[0])
ci_hat  = (psi_hat - 1.96 * se_hat, psi_hat + 1.96 * se_hat)
```

### Julia prototype

```julia
using LSurvival

# enter, exit, event with event = 0 censor, 1 interest, 2 competing
aj = aalen_johansen(enter, exit, event)
```

```julia
using DataFrames, MLJ, StatsFuns

τ, a = 365, 1
longdf, basedf = make_person_period(df, τ)

# Treatment model via MLJ
LogisticClassifier = @load LogisticClassifier pkg=MLJLinearModels
machA = machine(LogisticClassifier(), select(basedf, [:W1, :W2]), basedf.A)
fit!(machA)
basedf.gA_hat = predict_mode(machA, select(basedf, [:W1, :W2]))

# Custom hazard fits
longdf.gC_hat = fit_hazard(longdf[longdf.Y .== 1, :], :C,  [:A,:W1,:W2,:t])
longdf.h1_hat = fit_hazard(longdf[longdf.Y .== 1, :], :N1, [:A,:W1,:W2,:t])
longdf.h2_hat = fit_hazard(longdf[longdf.Y .== 1, :], :N2, [:A,:W1,:W2,:t])

longdf.G_hat        = censoring_survival(longdf, :gC_hat)
longdf.S_remain_hat = remaining_survival_to_tau(longdf, :h1_hat, :h2_hat, τ)

longdf.H = [
    (row.A == a ? 1.0 : 0.0) /
    max(row.gA_hat, 1e-6) /
    max(row.G_hat,  1e-6) *
    row.S_remain_hat
    for row in eachrow(longdf)
]

mask   = longdf.Y .== 1
offset = logit.(clamp.(longdf.h1_hat[mask], 1e-6, 1 - 1e-6))
H      = longdf.H[mask]; y = longdf.N1[mask]

ϵ = 0.0
for _ in 1:25
    qstar = logistic.(offset .+ ϵ .* H)
    score = sum(H .* (y .- qstar))
    info  = sum((H .^ 2) .* qstar .* (1 .- qstar))
    step  = score / max(info, 1e-8)
    ϵ    += step
    abs(step) < 1e-8 && break
end

longdf.h1_star = copy(longdf.h1_hat)
longdf.h1_star[mask] .= logistic.(offset .+ ϵ .* H)

ψhat  = standardized_cif(longdf, :h1_star, :h2_hat, τ, a)
eif   = eif_cif(longdf, :h1_star, :h2_hat, τ, a)
sehat = std(eif) / sqrt(nrow(basedf))
cihat = (ψhat - 1.96 * sehat, ψhat + 1.96 * sehat)
```

### Variance estimation and inference

For a scalar target,

$$
\widehat{\mathrm{Var}}(\hat\psi) = \frac{1}{n^2} \sum_{i=1}^n (\hat D_i^*)^2.
$$

For vectors of risks at multiple times, use the empirical covariance matrix of the IF vectors. Risk differences and risk ratios follow from direct IF algebra or the delta method.

Three finite-sample cautions:

1. **Report any weight or propensity truncation explicitly.** If positivity is weak, Wald intervals based on the IF can be optimistic.
2. **With data-adaptive nuisance fits, use cross-validation or cross-fitting.** If you bootstrap, rerun the *entire* nuisance + targeting pipeline inside each bootstrap resample.
3. **For full-curve inference, use one-step continuous-time TMLE** (e.g., `concrete`) to obtain simultaneous bands rather than stitching together unrelated pointwise intervals.

::: {.callout-warning}
## Common mistake #2: black-box trust in TMLE output

A shiny `survtmle` table with a tight confidence interval can hide a positivity catastrophe. Before reporting a TMLE estimate:

- Inspect the distribution of treatment and censoring weights (and any truncation).
- Plot the empirical mean of the estimated influence function across iterations: it should be driven close to zero.
- Compare the targeted CIF to the unadjusted Aalen--Johansen curve. Large differences should be explainable by confounding, not by a fragile nuisance fit.
- Re-run with stricter weight truncation; if results move materially, that is a finding, not a nuisance.
:::

## 6. Applied Example: Heart Failure Hospitalization with Death as a Competing Event {#sec-cr-applied}

This worked example shows how the choice of estimand changes the numeric answer with the same dataset. We simulate a small pharmacoepidemiology cohort in which Drug A reduces HF hospitalization but slightly increases mortality in frailer patients, the exact pattern that makes "censoring at death" misleading.

### Simulate the cohort

```{r}
#| label: sim-cohort
#| message: false
#| warning: false

library(tidyverse)
library(survival)

set.seed(42)

n <- 1500

cohort <- tibble(
  id            = seq_len(n),
  age           = rnorm(n, 70, 8),
  prior_cvd     = rbinom(n, 1, 0.45),
  kidney        = rbinom(n, 1, 0.30),
  frailty       = rnorm(n, 0, 1)
) %>%
  mutate(
    # treatment more likely in less frail patients (mild confounding)
    pA = plogis(-0.2 + 0.02 * (age - 70) - 0.3 * frailty),
    A  = rbinom(n, 1, pA)
  )

# Discrete-time hazards over 24 months
months <- 24
sim_long <- cohort %>%
  crossing(t = seq_len(months)) %>%
  arrange(id, t) %>%
  group_by(id) %>%
  mutate(
    # hospitalization hazard: lower under A, higher with prior CVD/kidney/age
    h_hosp  = plogis(-4.0 + 0.02 * (age - 70) + 0.5 * prior_cvd
                     + 0.4 * kidney + 0.2 * frailty - 0.7 * A),
    # death hazard: SLIGHTLY HIGHER under A in frail patients
    h_death = plogis(-5.0 + 0.04 * (age - 70) + 0.3 * kidney
                     + 0.5 * frailty + 0.4 * A * (frailty > 0)),
    # loss-to-follow-up hazard: small, weakly informative
    h_ltfu  = plogis(-5.5 + 0.1 * frailty)
  ) %>%
  ungroup()

# Walk forward in time and assign the first event
draw_event <- function(df_id) {
  for (k in seq_len(nrow(df_id))) {
    u <- runif(3)
    if (u[1] < df_id$h_hosp[k])  return(tibble(time = k, event = 1))
    if (u[2] < df_id$h_death[k]) return(tibble(time = k, event = 2))
    if (u[3] < df_id$h_ltfu[k])  return(tibble(time = k, event = 3))
  }
  tibble(time = nrow(df_id), event = 0)  # admin censored at end of follow-up
}

events <- sim_long %>%
  group_by(id) %>%
  group_modify(~ draw_event(.x)) %>%
  ungroup()

dat <- cohort %>% select(id, age, prior_cvd, kidney, frailty, A) %>%
  left_join(events, by = "id")

# Coding: 0 = admin censor, 1 = HF hospitalization, 2 = death, 3 = LTFU
dat %>% count(A, event)
```

The event-code table is the first thing to inspect. Death is more common under A in this simulation, hospitalization is less common under A, and LTFU is small.

### Three estimands, three numbers

We now compute the 24-month risk of HF hospitalization three ways: (a) censoring death (wrong for the observed-world question), (b) treating death as a competing event (right for the observed-world question), and (c) using a composite of HF hospitalization or death.

```{r}
#| label: three-estimands
#| message: false
#| warning: false

# (a) Naive Kaplan-Meier: death and LTFU both censored
km_fit <- survfit(
  Surv(time, event == 1) ~ A,
  data = dat %>% mutate(event = if_else(event == 1, 1, 0))
)

km_24 <- summary(km_fit, times = 24, extend = TRUE)
km_risk <- tibble(
  A    = c(0, 1),
  risk = 1 - km_24$surv,
  se   = km_24$std.err
)

# (b) Aalen-Johansen / cumulative incidence: death is a competing event,
#     LTFU and admin censoring are right censoring
aj_dat <- dat %>%
  mutate(
    cr_event = case_when(
      event == 1 ~ 1L,   # HF hospitalization
      event == 2 ~ 2L,   # death (competing)
      TRUE       ~ 0L    # admin censor or LTFU
    )
  )

aj_fit <- survfit(
  Surv(time, factor(cr_event, levels = 0:2)) ~ A,
  data = aj_dat
)

# Pull the cumulative incidence for HF hospitalization at t = 24
aj_24 <- summary(aj_fit, times = 24, extend = TRUE)
aj_risk <- tibble(
  A    = rep(c(0, 1), each = 1),
  # column 2 of pstate is cause 1 (HF hospitalization)
  risk = aj_24$pstate[, 2],
  se   = aj_24$std.err[, 2]
)

# (c) Composite endpoint: hospitalization OR death
comp_fit <- survfit(
  Surv(time, event %in% c(1, 2)) ~ A,
  data = dat
)
comp_24 <- summary(comp_fit, times = 24, extend = TRUE)
comp_risk <- tibble(
  A    = c(0, 1),
  risk = 1 - comp_24$surv,
  se   = comp_24$std.err
)

estimand_tbl <- bind_rows(
  km_risk   %>% mutate(estimand = "(a) KM, death censored"),
  aj_risk   %>% mutate(estimand = "(b) AJ, death as competing event"),
  comp_risk %>% mutate(estimand = "(c) Composite (hosp or death)")
) %>%
  arrange(estimand, A)
estimand_tbl
```

```{r}
#| label: fig-three-estimands
#| fig-cap: "24-month risk of HF hospitalization estimated three ways, by treatment arm (95% CIs). Censoring death (a) inflates the risk relative to the competing-risks Aalen-Johansen estimate (b); the composite endpoint (c) is larger still because it counts a different outcome. The three answers disagree by construction, which is why the estimand must be fixed before the estimator."
#| code-fold: true
ggplot(estimand_tbl, aes(factor(A, labels = c("Drug B (control)", "Drug A")),
                         risk, fill = estimand)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  geom_errorbar(aes(ymin = pmax(risk - 1.96 * se, 0), ymax = risk + 1.96 * se),
                position = position_dodge(0.8), width = 0.2) +
  labs(x = NULL, y = "24-month risk of HF hospitalization", fill = NULL,
       title = "Three estimands, three different numbers") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

The three numbers will not agree, and that disagreement is the whole point. Censoring death (estimand a) inflates the apparent risk of HF hospitalization, especially under Drug A, because frail patients who would have died early are kept "at risk" in the KM denominator as if they could still hospitalize. The AJ number (estimand b) is the patient-facing observed-world risk. The composite (estimand c) is bigger than both because it counts a different outcome.

### Stacked cumulative incidence by treatment

```{r}
#| label: stacked-cif
#| message: false
#| warning: false
#| fig-cap: "Stacked cumulative incidence: HF hospitalization (cause 1) and death (cause 2) by treatment arm."

aj_curve <- broom::tidy(aj_fit) %>%
  filter(state %in% c("1", "2")) %>%
  mutate(
    cause = recode(state, `1` = "HF hospitalization", `2` = "Death"),
    arm   = if_else(strata == "A=0", "Drug B (control)", "Drug A")
  )

aj_curve %>%
  ggplot(aes(time, estimate, fill = cause)) +
  geom_area(position = "stack", alpha = 0.8) +
  facet_wrap(~ arm) +
  labs(x = "Months since time zero", y = "Cumulative incidence",
       fill = NULL) +
  theme_minimal()
```

A reader can now see at a glance that Drug A shifts the death curve up and the HF hospitalization curve down. Reporting only the HF hospitalization arm of the comparison would conceal half of what the drug does.

### Manual cumulative incidence function

It is useful to build a tiny CIF by hand to make the formula concrete. The recipe at each month is:

1. Compute the hospitalization hazard among those still at risk.
2. Compute the death hazard among those still at risk.
3. Update overall survival as $S_k = S_{k-1} (1 - h_{\text{hosp},k} - h_{\text{death},k})$.
4. Add $S_{k-1} \cdot h_{\text{hosp},k}$ to the running cumulative incidence.

In words: cumulative incidence accumulates the chance of having the event of interest at each time, weighted by the chance of still being free of any event just before that time.

```{r}
#| label: manual-cif
#| message: false
#| warning: false

manual_cif <- function(time, event, horizon) {
  n_total <- length(time)                       # everyone is at risk at time zero
  tibble(time = time, event = event) %>%
    arrange(time) %>%
    group_by(time) %>%
    summarise(
      d_hosp  = sum(event == 1),                # hospitalizations at this time
      d_death = sum(event == 2),                # competing deaths at this time
      n_exit  = dplyr::n(),                     # subjects leaving the risk set (event or censor)
      .groups = "drop"
    ) %>%
    mutate(
      # at-risk just before each distinct time = total minus those who exited earlier
      n_at_risk = n_total - cumsum(lag(n_exit, default = 0)),
      h_hosp    = d_hosp  / pmax(n_at_risk, 1),
      h_death   = d_death / pmax(n_at_risk, 1),
      S_prev    = cumprod(lag(1 - h_hosp - h_death, default = 1)),
      cif_hosp  = cumsum(S_prev * h_hosp)
    ) %>%
    filter(time <= horizon) %>%
    select(time, n_at_risk, h_hosp, h_death, S_prev, cif_hosp)
}

dat %>%
  group_by(A) %>%
  group_modify(~ manual_cif(.x$time, .x$event, horizon = 24)) %>%
  filter(time %in% c(6, 12, 18, 24))
```

This hand-rolled estimator is not production code, but it makes the moving parts of AJ visible: the hazard of death enters the running survival factor $S_{k-1}$, which is why competing events shrink the cumulative incidence of the event of interest rather than inflate it the way naive KM does.

### Adjusted absolute risk via TMLE

If the analytic goal is a confounder-adjusted 24-month risk difference, the survival TMLE machinery from [@sec-tte-competing] applies directly. With the event coding above (0 = censor, 1 = HF hospitalization, 2 = death), a minimal call is:

```{r}
#| label: tmle-applied
#| message: false
#| warning: false
#| cache: true

library(survtmle)
library(SuperLearner)

tmle_fit <- survtmle(
  ftime      = dat$time,
  ftype      = ifelse(dat$event == 3, 0, dat$event),  # treat LTFU as censoring
  trt        = dat$A,
  adjustVars = dat %>% select(age, prior_cvd, kidney, frailty),
  t0         = 24,
  method     = "hazard",
  SL.trt     = c("SL.glm", "SL.mean"),
  SL.ctime   = c("SL.glm", "SL.mean"),
  SL.ftime   = c("SL.glm", "SL.mean"),
  returnIC   = TRUE
)

# Treatment-specific 24-month CIF estimates (rows: "<trt> <cause>")
tmle_fit$est
# Influence-function-based standard errors
sqrt(diag(tmle_fit$var))
```

The rows labelled `0 1` and `1 1` are the adjusted 24-month cumulative incidence of HF hospitalization (cause 1) under no treatment and treatment; rows `0 2` and `1 2` are the same for death (cause 2). Reading off the adjusted risk difference for HF hospitalization makes the contrast explicit:

```{r}
#| label: tmle-applied-contrast
est <- tmle_fit$est[, 1]
se  <- sqrt(diag(tmle_fit$var))
# Cause-1 (HF hospitalization) rows are named "0 1" and "1 1"
rd_hosp <- est["1 1"] - est["0 1"]
se_rd   <- sqrt(se["1 1"]^2 + se["0 1"]^2)   # conservative (ignores covariance)
cat(sprintf("Adjusted 24-month HF-hospitalization CIF: A=0 %.3f, A=1 %.3f\n",
            est["0 1"], est["1 1"]))
cat(sprintf("Adjusted risk difference (A=1 - A=0): %.3f (approx 95%% CI %.3f, %.3f)\n",
            rd_hosp, rd_hosp - 1.96 * se_rd, rd_hosp + 1.96 * se_rd))
```

The adjusted TMLE estimate confirms the descriptive Aalen--Johansen picture: Drug A lowers the cumulative incidence of HF hospitalization (cause 1) while raising the cumulative incidence of death (cause 2). The point of using TMLE here is not that it makes competing risks "go away." It targets a specific estimand, the adjusted treatment-specific cumulative incidence of HF hospitalization at 24 months, and gives influence-function-based inference for the contrast.

TMLE does not choose the estimand for you. Software choice is downstream of the estimand. Two analyses can both call `survtmle` and target completely different quantities depending on whether death is coded as a competing event (cause 2) or recoded as censoring. The software will produce a confidence interval either way. Only the analyst can decide which question is being asked.

## 7. Diagnostics and Sensitivity Analyses {#sec-cr-diag}

For descriptive work, the fastest diagnostic is to compare the cumulative incidence curve with $1-\hat S_{\text{KM}}$. A large gap means the competing event is materially important and that naive censoring would be consequential.

For cause-specific Cox models, use Schoenfeld-type checks and time-varying effects when needed. Then compare the model-implied cumulative incidence (combining all cause-specific hazards) against nonparametric AJ; a well-fitting cause-specific model should reproduce the observed CIF reasonably.

For Fine--Gray models, use cumulative-residual diagnostics for proportionality, linearity, and link assumptions. Evaluate calibration and discrimination on the *cumulative-incidence scale*, not just coefficient significance.

For IPCW and TMLE, inspect treatment and censoring weights directly: tails, summary statistics, truncation rules, and support over time. If inference changes materially after reasonable truncation or alternate censoring models, the analysis is leaning on fragile extrapolation.

Sensitivity analyses should target *assumptions*, not just software defaults:

- **Alternative estimands.** Always report at least the CIF of the event of interest *and* the CIF of the competing event. If a hypothetical estimand is primary, also report the total effect on CIF.
- **Alternative censoring models.** Pre-specify at least one alternative specification.
- **Different truncation rules** for IPCW or propensity weights.
- **Nonproportional-hazards relaxations** for Cox or Fine--Gray models.
- **Competing-event misclassification.** Probabilistic re-classification of cause-of-death codes and re-estimation.

## 8. Worked Study Scenarios {#sec-cr-scenarios}

| Setting | Event of interest | Competing event | Recommended primary estimand | Why |
|---|---|---|---|---|
| Adjuvant oncology | Cancer recurrence | Non-cancer death | Total effect on CIF of recurrence | Patient-facing question is "chance of relapse by 5 years"; censoring deaths and reporting $1-\hat S_{\text{KM}}$ would overestimate risk and break arm comparability |
| Aging / dementia cohort | Dementia diagnosis | Death | Total effect on CIF of dementia (with CIF of death reported alongside) | Exposure typically affects both; hazard ratios for dementia alone can be deeply misleading about population burden |
| Transplant waitlist | Death on waitlist | Transplant | Total effect on CIF of waitlist death | Transplant changes what patients experience; a "death risk if no transplant occurred" estimand is a different, narrower question |
| Cardiotoxicity vs. cancer death | Cardiotoxic event | Cancer death | **Separable effects** | Pathway question is the scientific point; treatment plausibly decomposes into antineoplastic and cardiotoxic components |
| Pediatric vaccine RCT | Disease outcome | Withdrawal due to mild AE | Reconsider: probably treatment-policy / total effect | A hypothetical "no withdrawals" world is rarely decision-relevant for the public-health question |

: Example scenarios and recommended primary estimands {#tbl-cr-scenarios}

The third row deserves special note: the hypothetical estimand "death risk in a world without transplant" is not wrong in principle, but it is a narrow auxiliary question. For policy and counseling, the observed-world CIF is the right primary target.

The last row is an example of an estimand that *should be rejected as scientifically weak*: when the competing-event-free hypothetical world has no clean intervention behind it, censoring the competing event is conceptual error dressed as method.

## 9. Reporting Guidance {#sec-cr-report}

A complete competing-risks report should answer four questions in plain language before any numbers appear:

1. **What is the estimand?** (population, treatment conditions, endpoint, summary, intercurrent-event handling)
2. **How are competing events handled in the estimand?** (part of the endpoint, censored as part of a hypothetical, or pathway-decomposed)
3. **What identification assumptions does that choice require?**
4. **What estimator targets that estimand, and what nuisance learners and diagnostics support it?**

In tables and figures, prefer absolute risks and risk differences at clinically meaningful horizons over hazard ratios alone. Always show the competing-event CIF next to the event-of-interest CIF; a treatment that lowers cancer recurrence while raising non-cancer death is not the same drug as one that does both well.

::: {.callout-warning}
## Regulatory writing tip: say exactly how each intercurrent event is handled

Avoid phrases like "adjusted for competing risks" or "competing risks were accounted for." They do not specify an estimand. A regulator (or a careful reviewer) reading those phrases cannot tell whether you targeted the observed-world CIF, a hypothetical world without the competing event, a composite endpoint, or a while-on-treatment quantity. Instead, write something like:

> "Death before HF hospitalization was treated as a competing event (cumulative incidence target). Loss to follow-up and end-of-study were treated as independent right censoring. Treatment discontinuation was handled under a treatment-policy strategy."

One short paragraph that names every intercurrent event and its handling strategy is worth more than any amount of post-hoc methods detail.
:::

### Reusable interpretation template

When the primary estimand is the total effect on the event-of-interest CIF, a reusable paragraph for the methods or results section reads:

> Our primary estimand was the 24-month risk difference for [event of interest] under initiation of [treatment] versus [comparator], treating [competing event] as a competing event and [loss to follow-up] as censoring. This estimand answers the question: among eligible patients at time zero, how would the probability of experiencing [event of interest] before [competing event] by 24 months differ under the two treatment strategies? It does not estimate the risk that would have been observed if [competing event] could be eliminated.

Filling in the bracketed terms forces the analyst to commit to a specific question, a specific competing event, and a specific time horizon, which is exactly the discipline ICH E9(R1) is asking for.

::: {.callout-tip}
## Reporting checklist

- [ ] Estimand named in plain English, before any model.
- [ ] Intercurrent events (including the competing event) explicitly assigned a strategy under ICH E9(R1).
- [ ] CIF of event of interest *and* CIF of competing event reported.
- [ ] If hazard models are reported, they are paired with absolute risks.
- [ ] Adjusted analyses report the nuisance learners, weight summaries, and any truncation.
- [ ] At least one sensitivity analysis varies the estimand, not only the model.
- [ ] Software, package versions, and seeds are recorded.
:::

## 10. Default Analysis Strategy and Decision Checklist {#sec-cr-default}

When in doubt, this is a defensible default for a randomized or observational study with a non-trivial competing event:

1. **State the primary estimand as a treatment-specific CIF at one or two pre-specified horizons,** and pre-specify a contrast (risk difference and/or risk ratio).
2. **Estimate it nonparametrically with Aalen--Johansen** (unadjusted) and with TMLE for the CIF (adjusted), reporting both.
3. **Estimate the cause-specific hazard ratio as a secondary etiologic analysis,** clearly labeled as such.
4. **Report the competing-event CIF alongside the event-of-interest CIF** under each treatment.
5. **Pre-specify sensitivity analyses** that vary the censoring model, weight truncation, and (where relevant) the estimand itself.

::: {.callout-important}
## Decision checklist: should I censor at the competing event?

Answer "yes" only if all four are true:

1. The scientific question is genuinely about a hypothetical world where the competing event is prevented.
2. There is a meaningful, interpretable hypothetical intervention behind that world (not just "imagine they didn't die").
3. The exchangeability assumption for the censoring induced by removing the competing event is defensible given measured covariates.
4. There is enough positivity in the induced pseudo-population that the analysis does not depend on extreme weights.

If any answer is "no," do not censor the competing event. Target the total effect on cumulative incidence instead, and consider separable effects if the underlying motivation is pathway decomposition.
:::

## Glossary {#sec-cr-glossary}

**Aalen--Johansen (AJ) estimator.** Nonparametric estimator of cumulative incidence in competing-risks settings. Uses all-cause survival as the carrier, so subjects experiencing competing events are correctly removed from future eligibility for the event of interest.

**Cause-specific hazard.** Instantaneous occurrence rate of a specific cause among individuals still alive and event-free.

**Clever covariate.** In TMLE, the data-driven covariate used in the targeting step that ensures the estimator solves the efficient-influence-function estimating equation for the chosen target parameter.

**Competing event.** An event whose occurrence makes the event of interest impossible thereafter (e.g., non-cancer death before relapse).

**Controlled direct effect (CDE).** The effect of treatment on the event of interest in a hypothetical world where the competing event has been prevented (or set to a fixed value). Requires strong identification assumptions.

**Cumulative incidence function (CIF).** $F_j(t) = P(T \le t, J = j)$. The probability that cause $j$ has occurred by time $t$. The natural patient-facing risk quantity.

**Efficient influence function (EIF).** The influence function of the most efficient regular asymptotically linear estimator for a target parameter; central to TMLE inference.

**Estimand.** The precise quantity to be estimated. Distinct from the estimator and the estimate. Under ICH E9(R1), specified by population, treatment conditions, endpoint, summary, and intercurrent-event handling.

**Fine--Gray model.** Proportional subdistribution hazards regression. Coefficients are tied algebraically to cumulative incidence.

**Intercurrent event.** Under ICH E9(R1), an event after treatment initiation that affects either the interpretation or the existence of the endpoint. Competing events are intercurrent events.

**IPCW.** Inverse probability of censoring weighting.

**One-step TMLE.** A continuous-time TMLE variant that updates all nuisance components simultaneously along a one-dimensional path; supports simultaneous inference over full risk curves.

**Separable effects.** Decomposition of treatment $A$ into components $A_Y$ and $A_D$ acting on the event of interest and competing event, respectively, enabling pathway-specific causal estimands.

**Subdistribution hazard.** Hazard defined on a risk set that retains subjects who already had a competing event; a mathematical device linking regression to CIF.

**Total effect on CIF.** The contrast $F_1^{a=1}(t) - F_1^{a=0}(t)$. The patient-facing causal effect on actual event probability.

## Key References {#sec-cr-refs}

**Foundations.** Prentice et al. (1978) on competing-risks inference; Aalen and Johansen (1978) on the AJ estimator; Gray (1988) on $K$-sample tests for cumulative incidence; Fine and Gray (1999) on subdistribution hazards.

**Practical reviews.** Putter, Fiocco, and Geskus (2007) on competing risks and multi-state models; Latouche et al. (2013) on reporting Fine--Gray analyses; Andersen and Keiding (2012) on rates vs. risks; Austin and Fine (2017) on practical recommendations.

**Estimands and causal interpretation.** ICH E9(R1) addendum (2019); Young et al. (2020) on a causal framework for competing-events estimands; Stensrud et al. (2022) on separable effects; Rufibach (2019) and Rudolph et al. on E9(R1) in survival contexts.

**TMLE for survival and competing risks.** van der Laan and Rose (2011, 2018) on Targeted Learning; Benkeser, Carone, and Gilbert (2018) on doubly robust competing-risks TMLE; Rytgaard and van der Laan (2023) on continuous-time TMLE; Rytgaard et al. (2021) on one-step TMLE for absolute-risk curves; Smith et al. (2025) on TMLE best-practice guidance.

**Software.** R: `survival`, `cmprsk`, `riskRegression`, `timereg`, `adjustedCurves`, `causalCmprsk`, `survtmle`, `concrete`. Python: `scikit-survival`, `lifelines`, `PyTMLE`. Julia: `LSurvival.jl`, `TMLE.jl`.
